### 2019-05-31 manuscript/outline/03-methods.Rmd
Added bookdown files

# Methods

Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot (Anderson-Teixeira 2015). SCBI (38°53'36.6"N, 78° 08'43.4"W) (Gonzalez-Akre et al 2016) is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park. Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level (Gonzalez-Akre 2016) with a topographic relief of 65m (Bourg et al 2013).

## Data Collection
Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013). In 2016-2017, cores were collected from all dead trees found in the annual mortality census (Gonzalez-Akre et al 2016). They were sanded, measured, and standardized in the same process as seen in Helcoski et al 2019.

Cores were then split into canopy and subcanopy groupings and processed with ARSTAN. Using CRU data **with Valentine's quilts for each grouping**, it was found that all climate variables tested were significant by tree position when using a linear mixed model (**this was done early-on, included here because unsure if will keep for full analysis**).

To determine focus drought years, we used the pointRes package in R (version 3.6.0) to determine resilience metrics. Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the past 5 years. The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA. For ease of research, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.

Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from Jennings 1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed). Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species’ cores. In total, 14 species had cores that were considered complete and usable, as seen in Table 1.

Geographic info

Elevation for the trees was extracted from a USGS DEM in ArcMap. Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.

Growth

As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems. Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,

diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))

First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species. Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness in from the regression equations.

Sapwood ratio also used the bark thickness data, and was also calculated for 2008, where

sapwood_area = total wood area (without bark) - heartwood area, and
sapwood_Ratio = sapwood area / total wood area (without bark)

Once sapwood area for each tree was calculated, log-log regression equations were determined on a per-species basis, which were then used

Height

Height data for trees was obtained from clinometers and rangefinders, from 2012 to 2019. Log=log regression equations were developed for species based on this height data, then heights of the study trees were calculated using these equations. For species that didn't have enough height measurements, heights were calculated from equations derived from all study-focus species.

- sapwood ratio

Traits
Turgor loss point (TLP) was determined per species from leaf trait studies carried out at sCBI in 2018. TLP was determined using __________________ on ### of individuals from each species (https://github.com/EcoClimLab/HydraulicTraits/blob/master/data/SCBI/processed_trait_data/SCBI_LeafTraits_measurements_plus_intermediary.csv).  Ring porosity was qualified from Anderson-Teixeira et al 2015.

After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable. Traits and other geographic data were used as independent variables.

### 2019-06-07 manuscript/03-methods.Rmd
Added more files to make book for article

--- 2019-05-31 manuscript/outline/03-methods.Rmd
@@
 
-Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot (Anderson-Teixeira 2015).
-SCBI (38°53'36.6"N, 78° 08'43.4"W) (Gonzalez-Akre et al 2016) is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level (Gonzalez-Akre 2016) with a topographic relief of 65m (Bourg et al 2013).
+Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot [@andersonteixeira_ctfs-forestgeo:_2015].
+SCBI (38°53'36.6"N, 78° 08'43.4"W) [@gonzalezakre_patterns_2016] is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
+Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m (Bourg et al 2013).
 
@@
 Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013).
-In 2016-2017, cores were collected from all dead trees found in the annual mortality census (Gonzalez-Akre et al 2016).
+In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
 They were sanded, measured, and standardized in the same process as seen in Helcoski et al 2019.
@@
 
-- sapwood ratio
+Height sources include [@stovall_assessing_2018], [@stovall_terrestrial_2018], [@andersonteixeira_ctfs-forestgeo:_2015],
 

### 2019-06-07 manuscript/05-discussion.Rmd
Added more files to make book for article

# Discussion

*1. paragraph summarizing main results--> primary conclusions*

We partially supported the hypothesis that crown exposure makes trees more vulnerable to drought. Co-dominant trees had the highest drought resistance. Dominant trees had lower resistance, likely because they are the most exposed. Other studies have found clear evidence of greater drought sensitivity in trees with exposed crowns (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]). At the same time, intermediate and suppressed trees had even lower resistance. This indicates that other mechanisms such as competition or rooting depth were important. (Also note that our study design was not ideal for testing the role of canopy position. Current canopy position is a conservative separator of canopy position: trees may currently be in more dominant positions than they were at the time, but backwards movement is unlikely. This would bias against finding a signficant effect for H1.2. Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)

Proximity to stream--either vertical (elev) or horizontal (distance)-- did not increase drought resistance; rather, it tended to decrease resistance (H1.3a). This may be because individuals growing further from water are acclimatized to drier conditions. However, the increase in drought resistance with distance from stream was less for small than large trees (H1.3b), indicating a potential importance of root depth/volume in conferring drought resistance.

### 2019-06-07 manuscript/07-references.Rmd
Added more files to make book for article

# References

TRY database:
Kattge, J., S. Diaz, S. Lavorel, I. C. Prentice, P. Leadley, G. Boenisch, E. Garnier, M. Westoby, P. B. Reich, I. J. Wright, J. H. C. Cornelissen, C. Violle, S. P. Harrison, P. M. v. Bodegom, M. Reichstein, B. J. Enquist, N. A. Soudzilovskaia, D. D. Ackerly, M. Anand, O. Atkin, M. Bahn, T. R. Baker, D. Baldocchi, R. Bekker, C. Blanco, B. Blonder, W. J. Bond, R. Bradstock, D. E. Bunker, F. Casanoves, J. Cavender-Bares, J. Q. Chambers, F. S. Chapin, J. Chave, D. Coomes, W. K. Cornwell, J. M. Craine, B. H. Dobrin, L. Duarte, W. Durka, J. Elser, G. Esser, M. Estiarte, W. F. Fagan, J. Fang, F. Fernandez-Mendez, A. Fidelis, B. Finegan, O. Flores, H. Ford, D. Frank, G. T. Freschet, N. M. Fyllas, R. V. Gallagher, W. A. Green, A. G. Gutierrez, T. Hickler, S. Higgins, J. G. Hodgson, A. Jalili, S. Jansen, C. Joly, A. J. Kerkhoff, D. Kirkup, K. Kitajima, M. Kleyer, S. Klotz, J. M. H. Knops, K. Kramer, I. Kuehn, H. Kurokawa, D. Laughlin, T. D. Lee, M. Leishman, F. Lens, T. Lenz, S. L. Lewis, J. Lloyd, J. Llusia, F. Louault, S. Ma, M. D. Mahecha, P. Manning, T. Massad, B. Medlyn, J. Messier, A. T. Moles, S. C. Mueller, K. Nadrowski, S. Naeem, U. Niinemets, S. Noellert, A. Nueske, R. Ogaya, J. Oleksyn, V. G. Onipchenko, Y. Onoda, J. Ordonez, G. Overbeck, W. A. Ozinga, S. Patino, S. Paula, J. G. Pausas, J. Penuelas, O. L. Phillips, V. Pillar, H. Poorter, L. Poorter, P. Poschlod, A. Prinzing, R. Proulx, A. Rammig, S. Reinsch, B. Reu, L. Sack, B. Salgado-Negret, J. Sardans, S. Shiodera, B. Shipley, A. Siefert, E. Sosinski, J.-F. Soussana, E. Swaine, N. Swenson, K. Thompson, P. Thornton, M. Waldram, E. Weiher, M. White, S. White, S. J. Wright, B. Yguel, S. Zaehle, A. E. Zanne, C. Wirth. 2011. TRY - a global database of plant traits. Global Change Biology, 17:2905-2935.

### 2019-06-07 manuscript/index.Rmd
Added more files to make book for article

---
title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
output: html_document
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
github-repo: manuscript/outline
description: "Article for Ian's research during his time in Anderson-Teixeira lab"
---

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

### 2019-06-07 manuscript/bookdown_help.Rmd
Created notes and helpful tips for bookdown

# Basic intro using RMarkdown and/or Bookdown
for journal articles or dissertation

## RMarkdown
- "RMarkdown" = is a formatting style, which is used in .Rmd files. These can be knit separately to make individual PDFs or html files and, if desired, a full article can be written in a single .Rmd file.

See a [guide](https://bookdown.org/yihui/rmarkdown/) to RMarkdown syntax, or see an example of a [github repo](https://github.com/dr-harper/rmarkdown-cookbook) with the files used for bookdown.

## Bookdown
- "Bookdown" = the system for compiling a lot of separate .Rmd or .md files together. This system requires more files (.yml, YAML, .bib, etc).

### File terms
#### .Rmd, .md
.Rmd = RMarkdown file, and .md = Markdown file. These are the files that are each chapter or section in the book, numbered in numerical order: 01-intro.Rmd, 02-Methods.Rmd, etc. The difference between the two is dependent on your content. If you are using figures and inline code, you need .Rmd, but if you're using only text (like a novel), .md works. Each of these chapter files must start with a title header:

- 01-intro.Rmd

    ```
    # Introduction

    Yay a book wow
    ```

#### .yml, .yaml
YAML = YAML Ain't Markup Language. Basically, it's a type of language for writing that is recognized by coding programs (e.g. R code recognizes "#" as a comment - that sort of thing). YAML is used primarily for data usage, and it has a specific style that is apparent in _bookdown.yml files. For more examples, see [here](https://en.wikipedia.org/wiki/YAML).

If you use _bookdown.yml, you can specify the order of your .Rmd files by using
    rmd_files: ["index.Rmd", "abstract.Rmd", "intro.Rmd"]

### File Types
#### index.Rmd
Your first Markdown file for the book should be index.Rmd; this is where you define things about the book. For example,

    ```
    ---
    title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a    temperate broadleaf forest"
    author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil    Pederson, Lawren Sack, Krista Anderson-Teixeira"
    output: html_document
    documentclass: book
    bibliography: [book.bib]
    biblio-style: apalike
    link-citations: yes
    github-repo: manuscript/outline
    description: "Article for Ian's research during his time in Anderson-Teixeira lab"
    ---
    ```
The options to include for index.Rmd are

- title,

- subtitle,

- author (can be list [-] on multiple lines),

- date,

- output,

- documentclass,

- bibliography,

- biblio-style,

- link-citations,

- github-repo,

- description,

- geometry (for specifying margins, very useful for pdf),

- fontsize,

- linestretch,

- toc-depth (how many numbered sections to include in table of contents),

- lof (list of figures [either TRUE or FALSE]),

- lot (list of tables [either TRUE or FALSE]),

- classoption ("oneside" or "twoside").

See [this site](https://eddjberry.netlify.com/post/writing-your-thesis-with-bookdown/) for examples when writing to a pdf for a thesis

 - this site also includes useful formatting for using LaTex to turn off, say, hyphenation across multiple lines, which is explained more in-depth [here](https://bookdown.org/yihui/bookdown/yaml-options.html)

### Writing Advice with Rmarkdown
These come from [here](https://eddjberry.netlify.com/post/writing-your-thesis-with-bookdown/)(also cited in other sections here)

1. Have each sentence on a separate line (this makes it easier to keep track of changes and quickly look over references at line ends).

2. You can keep R-code separate, then use save.image() to save R env, which can then be read into the Rmarkdown env with all the results from the code. Plus, saving that info as a separate R.Data file allows you to use that in other places too.

3. Save figures in separate folders as well.

### Add an image
Adding an image: make sure your working directory is set. Let's say I have an image in the "maps" folder. To add it in the chapter, I simply put in  `![captionofmap](maps/file_name.jpg)` and a blank box is inserted beneath my script. To see what this looks like in the actual book, you can knit the book.

### References with .bib
Citations are can be made in two ways. The easier way is via Zotero.

**As with all edits, please make sure you've fetched origin and your local files are updated before making/committing changes.**

#### Create the citation with Zotero

1. Install [Zotero](https://www.zotero.org/download/) for desktop and the plug-in for Chrome (if not using Chrome, that's ok. It will just take a little bit more effort).

2. Open  Zotero application (not the plugin)
- Click on 'file' -> 'import' -> 'next >'
- Select 'book.bibtex' file (extension is 'bibtex document')
- Make sure to check the 'Place imported collections...' box and click 'next >'
**Important**: Because Zotero automatically saves reference lists you make, you must import to a new collection every time you open, and only work out of that collection. Zotero will say you have duplicates, but that's because it's saving your first list you opened, and it doesn't automatically udpate from Github.
- Once you've imported to a new collection, you can just delete the old collection to avoid confusion when you're working

3. Add a citation either from the plug-in (super easy, just click on the icon when you're on the article or web page you wish to cite) or via manual entry.

4. Review the fields to make sure data is correct. Please remove abstract if automatically populated.

5. When done adding citations, order them by "Creator". 
**IMPORTANT** Select all your items in the collection, right click, and select *Export Items*. Change format to *BibTex* and make sure *Export Notes* is checked.

6. Save the file as **book.bib** and overwrite existing one.
- When looking to see if the file has been saved, the name of the file may show up as 'book.'

7. If you want to see how Bookdown views the file, open book.bib in RStudio. The file will now have your citation looking like this:
      ```
       @article{stovall_assessing_2018,
         title = {Assessing terrestrial laser scanning for developing non-destructive biomass allometry},
	        volume = {427},
	        issn = {0378-1127},
	        url = {http://www.sciencedirect.com/science/article/pii/S0378112718304663},
	        doi = {10.1016/j.foreco.2018.06.004},
	        urldate = {2018-11-26},
	        journal = {Forest Ecology and Management},
	        author = {Stovall, Atticus E. L. and Anderson-Teixeira, Kristina J. and Shugart, Herman H.},
	        month = nov,
	        year = {2018},
	        keywords = {Carbon, Forest structure, Large trees, Quantitative structure models, Sample size, Scaling theory, Terrestrial LiDAR},
	        pages = {217--229},
	        file = {ScienceDirect Full Text PDF:C\:\\Users\\mcgregori\\Zotero\\storage\\DP7LXP7I\\Stovall et al. - 2018 - Assessing terrestrial laser scanning for developin.pdf:application/pdf;ScienceDirect Snapshot:C\:\\Users\\mcgregori\\Zotero\\storage\\I4JV9V4V\\S0378112718304663.html:text/html}
        }
      ```
- for reference, this citation is from [this article](https://www.sciencedirect.com/science/article/pii/S0378112718304663). The `file =` part is Zotero saving a local copy of the article pdf.

#### Create the citation manually

If Zotero isn't working, citations can also be made manually. To do this, open the book.bib file in RStudio, and add the information for your citation. Use the [BibTex documentation] link below for guidance.
- your citation should also look like Step 7 above (the `file =` part is irrelevant).

#### Cite the citation in-text
8. In the citation above, the relevant part for in-text citing is the very first part of the code. Above the `title` is a character string called the "citation key". This is the text you write to tell Bookdown to generate a citation.

9. For example, there is a sentence from this article: "Trees are good." To cite this article, I use [@citation key] (see the [markdown] link below for other variations]. In the .Rmd file, I would write `Trees are good [@stovall_assessing_2018].` When the book is knit or rendered, Bookdown automatically converts this to:

`Trees are good (Stovall, Anderson-Teixeira, and Shugart 2018).` with the following reference appended to the text:

`Stovall, Atticus E. L., Kristina J. Anderson-Teixeira, and Herman H. Shugart. 2018. “Assessing Terrestrial Laser Scanning for Developing Non-Destructive Biomass Allometry.” Forest Ecology and Management 427 (November): 217–29. doi:10.1016/j.foreco.2018.06.004.`

- Helpfully, the year automatically links to the full appended reference.

#### Reference Order

References in each chapter are ordered by appearance in the text.

References for the whole book (in the final book chapter) are ordered alphabetically by author.

#### References for how to make a citation:

Exporting from Zotero: [UniMelbourne](http://unimelb.libguides.com/c.php?g=565734&p=3897111)

Manual citation entry in book.bib: [BibTex documentation](https://en.wikipedia.org/wiki/BibTeX)

Bookdown info for citations: [Yihui](https://bookdown.org/yihui/bookdown/citations.html)

Markdown syntax for in-text citations (and other info): [markdown](https://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

### Updating the book

First and foremost, install Knit and bookdown.

IMPORTANT: If using RProject, you can ONLY render a book if all the .Rmd file components of the book (all your chapters) are in the top folder of the project. You can have the figures and tables in subfolders, but otherwise everything must be in the top source folder.

       install.packages("knitr", "bookdown")
       setwd("localpath/Github/SCBI-Plot-Book")

There are two ways of updating.

#### Updating your chapter via Knitting

The first, and easiest way, is via the "Knit" function.

4. Once knitr is installed, a ball of yarn with a needle will appear on the banner of your .Rmd script in RStudio.

5. When you want to update the chapter (**or, view a preview of what the chapter will look like online**), select "Knit". This will simultaneously save the .Rmd file and update your chapter's html file as well. A blank window will pop up, then select "Open in Browser."

6. Running the Knit function will apply the same code as below in Step 7 with the same outputs. The differences are that Knit will automatically open a prompt to view the updated file, *and* it will only update the specific chapter you're working on.

#### Updating the whole book (including your new chapter)

The second way of updating is by putting together all the elements of the book and updating them all at once (called rendering).

7. Run the following command to put the book together (known as "rendering").

       bookdown::render_book("manuscript/outline/index.Rmd")

8. Your ouput will look something like this:

       processing file: _main.Rmd
        |.................................................................| 100%
        ordinary text without R code
       output file: _main.knit.md

       "C:/Program Files/RStudio/bin/pandoc/pandoc" +RTS -K512m -RTS _main.utf8.md --to html4 --from markdown+autolink_bare_uris+ascii_identifiers+tex_math_single_backslash --output _main.html --smart --email-obfuscation none --standalone --section-divs --table-of-contents --toc-depth 3 --template "C:\Users\mcgregori\Documents\R\win-library\3.5\bookdown\templates\gitbook.html" --highlight-style pygments --number-sections --css style.css --include-in-header "C:\Users\MCGREG~1\AppData\Local\Temp\14\Rtmp8WwPNp\rmarkdown-str397c787bde.html" --mathjax --filter pandoc-citeproc

       Output created: _book/introduction.html
       [1] "localpath/Github/SCBI-Plot-Book/_book/introduction.html"
       Warning message:
       In split_chapters(output, gitbook_page, number_sections, split_by,  :
          You have 7 Rmd input file(s) but only 6 first-level heading(s). Did you forget first-level headings in certain Rmd files?

9. The important part is "Output created." This means the html files have been updated and are in the \_book folder. The warning message refers to the lack of a header (H1 #) in the index.Rmd. This is fine.

10. From your local Github folder on your computer, you can now open the html file associated with your edited chapter and review how it looks.

11. When finished, please remember to commit and push back to Github. When pushing after doing the rendering command, you'll notice there are several changed files that have a .css, .js, and .json extensions. These refer to the elements of the web visualization and as such are necessary to be pushed as well.

#### Live-streaming your book updates
Instead of updating the book constantly using render_book, you can also have your updates be automatically viewed next to you as you write - all you have to do is save the output.

```
serve_book(dir = ".", output_dir = "_book", preview = TRUE, in_session = TRUE, quiet = FALSE, ...)
```

"If it does not really take too much time to recompile the whole book, you may set the argument preview = FALSE, so that every time you update the book, the whole book is recompiled, otherwise only the modified chapters are recompiled via preview_chapter()." see [source](https://bookdown.org/yihui/bookdown/serve-the-book.html)

### 2019-06-10 manuscript/index.Rmd
Added Joseph as author

@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira"
 output: html_document

### 2019-06-10 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R
Removed SLA, changed file-saving locations, did more model runs

@@ -566,24 +566,25 @@ elev <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/S
 
 trees_all$elev.m <- elev$dem_sigeo[match(trees_all$tree, elev$tag)]
 
-##5ci. add in SLA data ####
-traits_sla <- read_excel("data/traits/photosynthesis_traits.xlsx", sheet = "Data")
-traits_sla$species <- paste(traits_sla$Genus, traits_sla$Species)
-SLA <- traits_sla
-SLA$Genus.spp <- paste0(gsub("^(..).*", "\\1", SLA$Genus.spp), 
-                        gsub("^.* (..).*", "\\1", SLA$Genus.spp))
-SLA$Genus.spp <- tolower(SLA$Genus.spp)
-SLA <- SLA[SLA$Genus %in% c("Carya", "Fraxinus", "Fagus", "Juglans", "Liriodendron", "Pinus", "Quercus"), ]
-SLA <- SLA[SLA$Genus.spp %in% neil_sp & !is.na(SLA$SLA), ]
-SLA <- SLA[order(SLA$Genus.spp), ]
-SLA <- SLA[c(10,20,28)]
-unique(SLA$Genus.spp)
-
-mean_SLA <- 
-  group_by(SLA, Genus.spp) %>%
-  summarize(SLA_mean = mean(SLA))
-
-trees_all$SLA_mean <- mean_SLA$SLA_mean[match(trees_all$sp, mean_SLA$Genus.spp)]
+##5ci. add in SLA data (here for reference) ####
+### I initially was going to include SLA but Krista mentioned that SLA is the inverse of LMA, so for the purposes of this modelling, they're equal. We're focused more on having SCBI-specific data when possible, so we're using LMA.
+# traits_sla <- read_excel("data/traits/photosynthesis_traits.xlsx", sheet = "Data")
+# traits_sla$species <- paste(traits_sla$Genus, traits_sla$Species)
+# SLA <- traits_sla
+# SLA$Genus.spp <- paste0(gsub("^(..).*", "\\1", SLA$Genus.spp), 
+#                         gsub("^.* (..).*", "\\1", SLA$Genus.spp))
+# SLA$Genus.spp <- tolower(SLA$Genus.spp)
[…]

### 2019-06-10 manuscript/index.Rmd
Updated citations

--- 2019-06-07 manuscript/index.Rmd
@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira"
 output: html_document
 documentclass: book
-bibliography: [book.bib]
+bibliography: [book.bib, packages.bib]
 biblio-style: apalike
@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2'
 ), 'packages.bib')

### 2019-06-14 manuscript/04-results.Rmd
Updated book and added full .md version for easy editing

# Results

## Biophysical Models
Crown position 
When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought. This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.

From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.

### 2019-06-14 manuscript/index.Rmd
Updated book and added full .md version for easy editing

--- 2019-06-07 manuscript/index.Rmd
@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
-output: html_document
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira"
+output: 
+  bookdown::html_book:
+    theme: united
+  bookdoown::gitbook:
+    highlight: tango
 documentclass: book
-bibliography: [book.bib]
+bibliography: [book.bib, packages.bib]
 biblio-style: apalike
 link-citations: yes
-github-repo: manuscript/outline
+github-repo: manuscript
 description: "Article for Ian's research during his time in Anderson-Teixeira lab"
@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra'
 ), 'packages.bib')

### 2019-06-14 scripts/canopy_position_analysis.R
Changed link, removed HSM and Pmin

@@ -533,7 +533,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
 ##leaf traits gained from this include PLA_dry_percent, LMA_g_per_m2, mean_SPAD, Chl_m2_per_g, and WD [wood density]
 
-leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEOX3ZOPZRRBOJXJXUC5AKN6W"), stringsAsFactors = FALSE)
+leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEMR4YV3HMJYVS3N3325BTQ4U"), stringsAsFactors = FALSE)
 
 leaf_traits <- leaf_traits[, c(1,8,12,24,26,28)]
 
@@ -570,8 +570,9 @@ hydra <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/Hydr
 trees_all$p50.MPa <- hydra$psi_0.5_kl50[match(trees_all$sp, hydra$data.type)]
 trees_all$p80.MPa <- hydra$psi_0.5_kl80[match(trees_all$sp, hydra$data.type)]
 
-trees_all$Pmin.MPa <- 1.1122*trees_all$mean_TLP_Mpa + 0.3849
-trees_all$hsm.MPa <- trees_all$Pmin.MPa - trees_all$p50.MPa
+#after review, we have decided to not focus on pmin or HSM
+# trees_all$Pmin.MPa <- 1.1122*trees_all$mean_TLP_Mpa + 0.3849
+# trees_all$hsm.MPa <- trees_all$Pmin.MPa - trees_all$p50.MPa
 
 # meh <- trees_all[!duplicated(trees_all$sp), ]
 # meh <- meh[, c("sp", "mean_TLP_Mpa", "Pmin.MPa")]

### 2019-06-14 scripts/vertical_height_neon.R
Update vertical_height_neon.R

@@ -13,7 +13,7 @@ library(data.table)
 ##4. Relative humidity: DP1.00098.001, avg = 1min or 30min
 ##5. Shortwave radiation: DP1.00022.001, avg = 1min or 30min
 
-#this command will load the data into R and collapse into one df (within a list).
+#the function "loadByProduct" will load the data into R and collapse into one df (within a list).
 #it will not download/store anything on the computer, but working with large dfs will run slowly. Hence, it is a good idea to look at the 30min avg first.
 
 dp <- data.frame("data" = c("SAAT", "wind", "biotemp", "RH", "SR"),
@@ -28,6 +28,7 @@ dp[] <- lapply(dp, as.character)
 
 #this loop for some reason isn't producing plotly graphs that will work, but everything else runs smoothly
 for (i in seq(along=1:5)){
+  dp$value <- as.character(dp$value)
   value <- dp$value[[i]]
   
   neon_tower <- loadByProduct(dpID=dp$id[[i]], 
@@ -35,8 +36,8 @@ for (i in seq(along=1:5)){
                 package="basic", avg=30, 
                 check.size = FALSE, 
                 #(use TRUE outside loop to see how big the dowloads are)
-                startdate="2018-05",
-                enddate="2018-08")
+                startdate="2017-05",
+                enddate="2017-08")
   
   neon_data <- neon_tower[[1]]
   neon_data_sub <- neon_data[colnames(neon_data) %in% c("verticalPosition", "startDateTime", value)]
@@ -69,9 +70,11 @@ biotemp_plot
 RH_plot
 SR_plot
 
-neon_data_sub$day <- substr(neon_data_sub$startDateTime, 1, nchar(neon_data_sub$startDateTime)-0)
+
 
 #determine threshold for sunny/cloudy day
+neon_data_sub$day <- substr(neon_data_sub$startDateTime, 1, nchar(neon_data_sub$startDateTime)-0)
+
 q <- neon_data_sub %>%
       group_by(day) %>%
       summarize(total_Rad = sum(shortRadMean))

### 2019-06-14 manuscript/_bookdown_files/output_files/McGregor_et_al_2019.tex
Trying to get both a full .md and a html file

\documentclass[]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest},
            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{abstract}{%
\chapter{Abstract}\label{abstract}}

something something something

\hypertarget{introduction}{%
\chapter{Introduction}\label{introduction}}

Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.

Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought. Therefore, in order to understand whole-forest response to drought, we need to know how responses vary by tree size/ species. To do so, there are three fundamental questions that must be addressed:

\emph{First, what drives the observed tendency for large trees to suffer more during drought?}\\
Bennett et al.~2015 showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. It is also expected that roots play a role. These hypotheses need to be tested.

\emph{Second, how do species' traits influence drought response?}
Analyzed drought responses on the species level doesn't get at mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/ traditionally measured traits such as wood density and SLA (Medeiros et al.).

\emph{Third, how do tree size and functional traits interact to influence drought response?}
Its possible that pattern observed by Bennett et al.~could be caused by smaller trees being more drought resistant. Alternatively, larger trees may have more drought-resistant traits.

\emph{Questions \& Hypotheses}

\textbf{1. Why do larger trees suffer greater growth declines during drought?}

1.0. Our forest displays the same trend as most forests globally (Bennett et al.~2015). (Note that Bennett et al.~2015 identified only one study on tree growth responses to drought in the Eastern US temperate deciduous biome. We know little about how tree size shapes drought response in this biome.)

1.1. Height is a strong predictor of drought stress.

\begin{verbatim}
* P1.1-Drought response increases with height at time of drought.
\end{verbatim}

1.2. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.

\begin{verbatim}
* P1.2a- Trees currently in a dominant canopy position suffered more during drought. 
\end{verbatim}

1.3. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.

\begin{verbatim}
 * P1.3a- drought response increases with elevation and/or distance from stream
 * P1.3b- There is a dbh\*elev interaction, elevation (relative to stream?) matters less for big trees with larger (and potentially deeper) root system.
\end{verbatim}

\textbf{2.Do species functional traits predict drought response?}

\begin{verbatim}
 * P2.1- TLP correlates negatively with drought resistance (NEVER tested), 
 * P2.2- diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
\end{verbatim}

\textbf{3.How are functional traits distributed across size classes, and how does this affect size-resistance relationship?} (Reviewers of Bennett et al.~2015 suggested that larger trees may suffer more because larger trees tend to be species with more drought-sensitive traits. \emph{Conversely, larger size classes may have more drought-resistant species because of greater biophysical challenges.}

\begin{verbatim}
 * P3.1a- TLP is lower (larger negative) in smaller/ understory trees
 * P3.1a- diffuse porous species more common in understory 
\end{verbatim}

\emph{For the above, I think it would be more interesting to test the converse prediction, which appears to be true. However, it should sample from tree population based on census data, not cores.}

\begin{verbatim}
 * P3.2- Inclusion of TLP / rp in model eliminates (or significantly reduces) effect of tree size.
\end{verbatim}

\hypertarget{methods}{%
\chapter{Methods}\label{methods}}

Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot \citep{andersonteixeira_ctfs-forestgeo:_2015}. SCBI (38°53'36.6``N, 78° 08'43.4''W) \citep{gonzalezakre_patterns_2016} is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park. Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level \citep{gonzalezakre_patterns_2016} with a topographic relief of 65m (Bourg et al 2013).

\hypertarget{data-collection}{%
\section{Data Collection}\label{data-collection}}

Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013). In 2016-2017, cores were collected from all dead trees found in the annual mortality census \citep{gonzalezakre_patterns_2016}. They were sanded, measured, and standardized in the same process as seen in Helcoski et al 2019.

Cores were then split into canopy and subcanopy groupings and processed with ARSTAN. Using CRU data \textbf{with Valentine's quilts for each grouping}, it was found that all climate variables tested were significant by tree position when using a linear mixed model (\textbf{this was done early-on, included here because unsure if will keep for full analysis}).

To determine focus drought years, we used the pointRes package in R (version 3.5.2) to determine resilience metrics. Specifically, we calculated pointer years (main drought years) if \textgreater50\% of the cored trees experienced \textless30\% growth in a year compared to the past 5 years. The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA. For ease of research, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.

Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species' cores. In total, 14 species had cores that were considered complete and usable, as seen in Table 1.

\begin{longtable}[]{@{}llll@{}}
\toprule
\textbf{Species} & \textbf{n\_cores.canopy} & \textbf{n\_cores.subcanopy} & \textbf{included}\tabularnewline
\midrule
\endhead
\emph{Carya cordiformis} & 2 & 11 & sub-canopy only\tabularnewline
\emph{Carya glabra} & 9 & 22 & yes\tabularnewline
\emph{Carya ovalis} & 9 & 14 & yes\tabularnewline
\emph{Carya tomentosa} & 0 & 13 & sub-canopy only\tabularnewline
\emph{Fagus grandifolia} & 7 & 73 & yes\tabularnewline
\emph{Fraxinus americana} & 20 & 42 & yes\tabularnewline
\emph{Fraxinus nigra} & 1 & 11 & sub-canopy only\tabularnewline
\emph{Juglans nigra} & 22 & 9 & yes\tabularnewline
\emph{Liriodendron tulipifera} & 38 & 60 & yes\tabularnewline
\emph{Pinus strobus} & 14 & 21 & no\tabularnewline
\emph{Quercus alba} & 38 & 23 & yes\tabularnewline
\emph{Quercus prinus} & 27 & 32 & yes\tabularnewline
\emph{Quercus rubra} & 43 & 26 & yes\tabularnewline
\emph{Quercus velutina} & 54 & 23 & yes\tabularnewline
\bottomrule
\end{longtable}

\hypertarget{biophysical-traits}{%
\subsection{Biophysical traits}\label{biophysical-traits}}

Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from \citep{jennings_assessing_1999}, whereby positions were ranked as dominant, codominant, intermediate, or suppressed. For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).

Elevation for the trees was extracted from a USGS DEM in ArcMap. Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.

Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual \citep[NEON]{stovall_assessing_2018}, digital rangefinders \citep{andersonteixeira_ctfs-forestgeo:_2015}, and automatic LiDAR \citep{stovall_terrestrial_2018}. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. The associated errors for using either method were acknowledged \citep{larjavaara_measuring_2013}. Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees. For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.

Equations required the DBH in each individual drought year. As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems. Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,

\texttt{diam\_nobark\_1999\ =\ dbh2008\ -\ 2*(bark.depth2008)\ -\ 2*(sum(ring.width1999:ring.width2008))}

First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species. Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.

\hypertarget{leaf-hydraulic-traits}{%
\subsection{Leaf Hydraulic Traits}\label{leaf-hydraulic-traits}}

Hydraulic traits were collected from SCBI and are summarized in Table 1.

\begin{table}[H]
\centering
\begin{tabular}{l|l|l}
\hline
Trait & Unit & Source\\
\hline
Ring Porosity &  & [@andersonteixeira\_ctfs-forestgeo:\_2015]\\
\hline
Percent Leaf Area & \% & Kunert et al 2019 [in prep]\\
\hline
Leaf Mass Area & g/m2 & Kunert et al 2019 [in prep]\\
\hline
Chlorophyll & m2/g & Kunert et al 2019 [in prep]\\
\hline
Wood density & g/cm3 & Kunert et al 2019 [in prep]\\
\hline
TLP & MPa & Kunert et al 2019 [in prep]\\
\hline
P50 & MPa & Kunert et al 2019 [in prep]\\
\hline
P80 & MPa & Kunert et al 2019 [in prep]\\
\hline
\end{tabular}
\end{table}

After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable. Traits and other geographic data were used as independent variables.

\hypertarget{results}{%
\chapter{Results}\label{results}}

\hypertarget{biophysical-models}{%
\section{Biophysical Models}\label{biophysical-models}}

Crown position
When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought. This follows on work done by \citep{bennett_larger_2015} {[}and others?{]} which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.

From \citep{kannenberg_linking_2019}, species with diffuse porous wood anatomy (\emph{Liriodendron}) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.

\hypertarget{discussion}{%
\chapter{Discussion}\label{discussion}}

\emph{1. paragraph summarizing main results--\textgreater{} primary conclusions}

We partially supported the hypothesis that crown exposure makes trees more vulnerable to drought. Co-dominant trees had the highest drought resistance. Dominant trees had lower resistance, likely because they are the most exposed. Other studies have found clear evidence of greater drought sensitivity in trees with exposed crowns (e.g., \citep{suarez_factors_2004}; \citep{scharnweber_confessions_2019}). At the same time, intermediate and suppressed trees had even lower resistance. This indicates that other mechanisms such as competition or rooting depth were important. (Also note that our study design was not ideal for testing the role of canopy position. Current canopy position is a conservative separator of canopy position: trees may currently be in more dominant positions than they were at the time, but backwards movement is unlikely. This would bias against finding a signficant effect for H1.2. Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)

Proximity to stream--either vertical (elev) or horizontal (distance)-- did not increase drought resistance; rather, it tended to decrease resistance (H1.3a). This may be because individuals growing further from water are acclimatized to drier conditions. However, the increase in drought resistance with distance from stream was less for small than large trees (H1.3b), indicating a potential importance of root depth/volume in conferring drought resistance.

\hypertarget{conclusion}{%
\chapter{Conclusion}\label{conclusion}}

words words words

\hypertarget{references}{%
\chapter{References}\label{references}}

TRY database - ONLY INCLUDE IF USING SLA
Kattge, J., S. Diaz, S. Lavorel, I. C. Prentice, P. Leadley, G. Boenisch, E. Garnier, M. Westoby, P. B. Reich, I. J. Wright, J. H. C. Cornelissen, C. Violle, S. P. Harrison, P. M. v. Bodegom, M. Reichstein, B. J. Enquist, N. A. Soudzilovskaia, D. D. Ackerly, M. Anand, O. Atkin, M. Bahn, T. R. Baker, D. Baldocchi, R. Bekker, C. Blanco, B. Blonder, W. J. Bond, R. Bradstock, D. E. Bunker, F. Casanoves, J. Cavender-Bares, J. Q. Chambers, F. S. Chapin, J. Chave, D. Coomes, W. K. Cornwell, J. M. Craine, B. H. Dobrin, L. Duarte, W. Durka, J. Elser, G. Esser, M. Estiarte, W. F. Fagan, J. Fang, F. Fernandez-Mendez, A. Fidelis, B. Finegan, O. Flores, H. Ford, D. Frank, G. T. Freschet, N. M. Fyllas, R. V. Gallagher, W. A. Green, A. G. Gutierrez, T. Hickler, S. Higgins, J. G. Hodgson, A. Jalili, S. Jansen, C. Joly, A. J. Kerkhoff, D. Kirkup, K. Kitajima, M. Kleyer, S. Klotz, J. M. H. Knops, K. Kramer, I. Kuehn, H. Kurokawa, D. Laughlin, T. D. Lee, M. Leishman, F. Lens, T. Lenz, S. L. Lewis, J. Lloyd, J. Llusia, F. Louault, S. Ma, M. D. Mahecha, P. Manning, T. Massad, B. Medlyn, J. Messier, A. T. Moles, S. C. Mueller, K. Nadrowski, S. Naeem, U. Niinemets, S. Noellert, A. Nueske, R. Ogaya, J. Oleksyn, V. G. Onipchenko, Y. Onoda, J. Ordonez, G. Overbeck, W. A. Ozinga, S. Patino, S. Paula, J. G. Pausas, J. Penuelas, O. L. Phillips, V. Pillar, H. Poorter, L. Poorter, P. Poschlod, A. Prinzing, R. Proulx, A. Rammig, S. Reinsch, B. Reu, L. Sack, B. Salgado-Negret, J. Sardans, S. Shiodera, B. Shipley, A. Siefert, E. Sosinski, J.-F. Soussana, E. Swaine, N. Swenson, K. Thompson, P. Thornton, M. Waldram, E. Weiher, M. White, S. White, S. J. Wright, B. Yguel, S. Zaehle, A. E. Zanne, C. Wirth. 2011. TRY - a global database of plant traits. Global Change Biology, 17:2905-2935.

\bibliography{book.bib,packages.bib}

\end{document}

### 2019-06-14 manuscript/index.Rmd
Trying to get both a full .md and a html file

@@
 output:
-  bookdown::html_book:
-    theme: united
-  bookdoown::gitbook:
-    highlight: tango
+  bookdown::html_book: default
+  bookdown::gitbook: default
 documentclass: book
+site: bookdown::bookdown_site
 bibliography: [book.bib, packages.bib]
@@
 ```
+
+<!-- this code makes the .md file but not a correctly-formatted html -->
+<!-- output:  -->
+<!--   bookdown::html_book: -->
+<!--   bookdown::gitbook: -->
+
+<!-- this code makes a correctly-formatted html file but not a .md file -->
+<!-- output:  -->
+<!--  html_document: -->
+<!--     theme: !expr sample(c("yeti", "united", "lumen"), 1) -->
+<!--  gitbook: -->
+<!--     highlight: tango -->

### 2019-06-14 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -1576,7 +1576,7 @@ aic_top <- var_aic %>%
 #define response and effects
 response <- "resist.value"
 effects <- c("position", "elev.m", "distance.ln.m", "height.ln.m", "(1|sp)")
-# effects <- c("rp", "PLA_dry_percent", "LMA_g_per_m2", "Chl_m2_per_g", "WD_g_per_cm3", "mean_TLP_Mpa", "p50.MPa", "p80.MPa", "hsm.MPa", "year", "(1|sp/tree)")
+# effects <- c("rp", "PLA_dry_percent", "LMA_g_per_m2", "WD_g_per_cm3", "mean_TLP_Mpa", "p50.MPa", "p80.MPa", "hsm.MPa", "year", "(1|sp/tree)")
 
 #create all combinations of random / fixed effects
 effects_comb <- 

### 2019-06-19 manuscript/03-methods.Rmd
Added Helcoski et al 2019 source

@@
 In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
-They were sanded, measured, and standardized in the same process as seen in Helcoski et al 2019.
+They were sanded, measured, and standardized in the same process as seen in [@helcoski_growing_2019].
 
@@
 
-To determine focus drought years, we used the pointRes package in R (version 3.6.0) to determine resilience metrics.
+To determine focus drought years, we used the pointRes package in R (version 3.5.2) to determine resilience metrics.
 Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the past 5 years.
@@
 
-Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from Jennings 1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).
 Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species’ cores.
@@
 
-Geographic info
+|**Species**|**n_cores.canopy**|**n_cores.subcanopy**|**included**
+|----------|-------------|---------------|-----------|
+*Carya cordiformis*|2|11| sub-canopy only
+*Carya glabra*|9|22|yes
+*Carya ovalis*|9|14|yes
+*Carya tomentosa*|0|13|sub-canopy only
+*Fagus grandifolia*|7|73|yes
+*Fraxinus americana*|20|42|yes
+*Fraxinus nigra*|1|11| sub-canopy only
+*Juglans nigra*|22|9|yes
+*Liriodendron tulipifera*|38|60|yes
+*Pinus strobus*|14|21|no
+*Quercus alba*|38|23|yes
+*Quercus prinus*|27|32|yes
+*Quercus rubra*|43|26|yes
+*Quercus velutina*|54|23|yes
+
+``` {r sample, echo=FALSE}
+#numbers for the table above come from this table
+# library(RCurl)
+# library(dplyr)
+# species <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/master/data/core_list_for_neil.csv?token=AJNRBELCYZY576ARLE7BXNK5BEF3O"))
+# 
+# test <- species %>%
+#   group_by(sp, crown.position) %>%
+#   summarize(count=n())
+
+```
+
+### Biophysical traits
+Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).
 
@@
 
-Growth
+Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015], and automatic LiDAR [@stovall_terrestrial_2018].
+Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
+The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
+Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees.
+For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
 
+Equations required the DBH in each individual drought year.
 As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems.
@@
 
-diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))
+`diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))`
 
 First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species.
-Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness in from the regression equations.
+Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.
 
-Sapwood ratio also used the bark thickness data, and was also calculated for 2008, where
+### Leaf Hydraulic Traits
+Hydraulic traits were collected from SCBI and are summarized in Table 1.
 
-sapwood_area = total wood area (without bark) - heartwood area, and
-sapwood_Ratio = sapwood area / total wood area (without bark)
+```{r, echo=FALSE}
+library(knitr)
+library(kableExtra)
 
-Once sapwood area for each tree was calculated, log-log regression equations were determined on a per-species basis, which were then used
+traits_table <- data.frame(
+  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Chlorophyll", "Wood density", "TLP", "P50", "P80"),
+  "Unit" = c("", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"),
+  "Source" = c("[@andersonteixeira_ctfs-forestgeo:_2015]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]"))
 
-Height
+kable(traits_table) %>%
+  kable_styling(bootstrap_options = "striped", full_width = F)
 
-Height data for trees was obtained from clinometers and rangefinders, from 2012 to 2019.
-Log=log regression equations were developed for species based on this height data, then heights of the study trees were calculated using these equations.
-For species that didn't have enough height measurements, heights were calculated from equations derived from all study-focus species.
-
-Height sources include [@stovall_assessing_2018], [@stovall_terrestrial_2018], [@andersonteixeira_ctfs-forestgeo:_2015],
-
-Traits
-Turgor loss point (TLP) was determined per species from leaf trait studies carried out at sCBI in 2018.
-TLP was determined using __________________ on ### of individuals from each species (https://github.com/EcoClimLab/HydraulicTraits/blob/master/data/SCBI/processed_trait_data/SCBI_LeafTraits_measurements_plus_intermediary.csv).
-Ring porosity was qualified from Anderson-Teixeira et al 2015.
+```
 

### 2019-06-25 scripts/canopy_position_analysis.R
Clarified code sections

@@ -184,7 +184,8 @@ library(reshape2)
 #NB ####
 ##to be clear, I wrote this code before I realized that some of the work done in these loops had already been done in the outputs of res.comp (specifically out.select). However, since the code runs well, and I double-checked that it was giving the same outputs as analyzing out.select, I'm keeping it as is.
 
-##4a. canopy ####
+##4a. determine pointer years 
+###canopy ####
 dirs_can <- dir("data/core_files/canopy_cores", pattern = "_canopy.rwl")
 
 #originally, getting the pointer years was done without caco and frni because of too few cores. Now that we have the pointer years, we're adding them back in for the full analysis.
@@ -223,7 +224,7 @@ values <- paste0(sp_can, "_canopy")
 names(widths_can) <- values
 
 
-##4b. subcanopy ####
+###subcanopy ####
 dirs_subcan <- dir("data/core_files/subcanopy_cores", pattern = "_subcanopy.rwl")
 
 #frni_canopy had only 1 core. Originally this was excluded when finding pointer years, but now that we have calculated them, we have added appended the frni_canopy to the frni_subcanopy file and called it "frni_all_drop_subcanopy.rwl"
@@ -263,7 +264,7 @@ widths <- c(widths_can, widths_sub) #combine into one, then delete individual. F
 widths_can <- NULL
 widths_subcan <- NULL
 
-##4c. df for pointer years of all trees combined ####
+### df for pointer years of all trees combined ####
 full_ind <- rbind(canopy_table, subcanopy_table) #full table of indices for canopy and subcanopy cores
 pointers <- full_ind[full_ind$nature == -1, ]
 
@@ -279,7 +280,7 @@ years_bysp <- years_bysp[order(years_bysp$year, years_bysp$sp), ]
 
 #write.csv(pointers, "data/occurrence_of_pointer_yrs.csv", row.names=FALSE)
 
-##4d. resistance metrics for all trees ####
+##4b. resistance metrics for all trees ####
 neil_list <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
 neil_sp <- unique(neil_list$sp)
 
@@ -369,7 +370,7 @@ trees_all$year <- as.numeric(trees_all$year)
 trees_all <- trees_all[!is.na(trees_all$resist.value), ]
 trees_all$year <- as.character(trees_all$year)
 
[…]

### 2019-06-25 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -518,7 +518,6 @@ library(sf) #for mapping
 library(ggthemes) #for removing graticules when making pdf
 library(rgeos) #for distance calculation
 library(RCurl) #for reading in URLs
-library(dplyr)
 library(readxl)
 
 ##5a. add in ring porosity qualifications ####
@@ -1025,7 +1024,6 @@ library(car)
 library(piecewiseSEM) #for R^2 values for all model outputs in a list
 library(MuMIn) #for R^2 values of one model output
 library(stringr)
-library(dplyr)
 
 ##6a. Determine best model to use with AICc ####
 ##6ai. test predictions for paper and put in table

### 2019-06-25 manuscript/03-methods.Rmd
Added images

@@
 SCBI (38°53'36.6"N, 78° 08'43.4"W) [@gonzalezakre_patterns_2016] is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m (Bourg et al 2013).
+Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
+
+![Structure of ForestGEO plot](tables_figures/ForestGEO_plot.jpg)
 
@@
 Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm (Bourg et al 2013).
+Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
 In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
 They were sanded, measured, and standardized in the same process as seen in [@helcoski_growing_2019].
+Cores were then split into canopy and subcanopy groupings as determined in the field.
+Species' inclusion in the study was dependent on the robustness of each species’ cores.
+In total, 14 species had cores that were considered complete and usable, as seen in Table 1.
 
-Cores were then split into canopy and subcanopy groupings and processed with ARSTAN.
 Using CRU data **with Valentine's quilts for each grouping**, it was found that all climate variables tested were significant by tree position when using a linear mixed model (**this was done early-on, included here because unsure if will keep for full analysis**).
 
-To determine focus drought years, we used the pointRes package in R (version 3.5.2) to determine resilience metrics.
-Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the past 5 years.
+To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
+Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
 The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
-For ease of research, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
-
-Individual cores were then grouped by canopy or subcanopy before being processed by ARSTAN, and species inclusion in the study was dependent on the robustness of each species’ cores.
-In total, 14 species had cores that were considered complete and usable, as seen in Table 1.
+Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
 
@@
 |----------|-------------|---------------|-----------|
-*Carya cordiformis*|2|11| sub-canopy only
+*Carya cordiformis*|2|11|yes
 *Carya glabra*|9|22|yes
@@
 *Liriodendron tulipifera*|38|60|yes
-*Pinus strobus*|14|21|no
 *Quercus alba*|38|23|yes
@@
 ### Leaf Hydraulic Traits
-Hydraulic traits were collected from SCBI and are summarized in Table 1.
+Hydraulic traits were collected from SCBI and are summarized in Table 1. Ring porosity qualifications were obtained from [@andersonteixeira_size-related_2015].
+
+Need to say the methodology for all traits used
+1. PLA
+2. LMA
+3. Chlorophyll
+4. Wood density
+5. TLP
+6. P50 = psi_0.5_kl50
+7. P80 = psi_0.5_kl80
 
@@
   "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Chlorophyll", "Wood density", "TLP", "P50", "P80"),
-  "Unit" = c("", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"),
-  "Source" = c("[@andersonteixeira_ctfs-forestgeo:_2015]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]", "Kunert et al 2019 [in prep]"))
+  "Unit" = c("", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"))
 

### 2019-06-25 scripts/plot_time_series.R
update figure Data question #31

@@ -64,7 +64,7 @@ par(mfrow = c(length(SPECIES_IN_ORDER) + 3, 1), mar = c(0,0,0,0), oma = c(4, 6, 
 plot(NULL,
      axes = F,
      ann = F, 
-     xlim = c(1959,2020), ylim = c(100, 160))
+     xlim = c(1959,2015), ylim = c(100, 160))
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
@@ -82,7 +82,7 @@ axis(2, at = c(110, 130, 150), las = 1)
 plot(NULL,
      axes = F,
      ann = F, 
-     xlim = c(1959,2020), ylim = c(40, 180))
+     xlim = c(1959,2015), ylim = c(40, 180))
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
@@ -97,13 +97,13 @@ text(x = 2010, y = 100, "PRE", pos = 4, col = "blue")
 axis(2, at = c(60, 110, 160), las = 1)
 
 mtext(side = 2, text = expression("(mm mo"^-1*")"), line = 3, adj = 0)
-# axis(1, at = c(1700, 2020), labels = F, tck = 0, col = "grey60")
+# axis(1, at = c(1700, 2015), labels = F, tck = 0, col = "grey60")
 
 # pdsi ####
 plot(NULL,
      axes = F,
      ann = F, 
-     xlim = c(1959,2020), ylim = c(-7, 5))
+     xlim = c(1959,2015), ylim = c(-7, 5))
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
@@ -118,7 +118,7 @@ text(x = 2010, y = 0, "PDSI", pos = 4, col = "black")
 axis(2, at = c(-4, 0, 4), las = 1)
 
 # mtext(side = 2, text = expression("(mm mo"^-1*")"), line = 3, adj = -0.5)
-axis(1, at = c(1700, 2020), labels = F, tck = 0, col = "grey60")
+axis(1, at = c(1700, 2015), labels = F, tck = 0, col = "grey60")
 
 # chronologies ####
 for(f in SPECIES_IN_ORDER) {
@@ -140,7 +140,7 @@ for(f in SPECIES_IN_ORDER) {
   plot(NULL,
        axes = F,
        ann = F, 
-       xlim = c(1959,2020), ylim = c(0.5, 1.5))
+       xlim = c(1959,2015), ylim = c(0.5, 1.5))
   
[…]

### 2019-06-25 scripts/plot_time_series.R
change tif to jpg

@@ -54,7 +54,7 @@ pre_drought_years <- drought_years - 5
 
 
 if(save.plots)  {
-  tiff(paste0("manuscript/tables_figures/Time_series_for_each_species.tif"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
+  jpeg(paste0("manuscript/tables_figures/Time_series_for_each_species.jpg"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
 }
 
 par(mfrow = c(length(SPECIES_IN_ORDER) + 3, 1), mar = c(0,0,0,0), oma = c(4, 6, 0, 0))

### 2019-06-27 manuscript/07-supplement.Rmd
Tried to make correct output formats

# Supplementary Information

## p50 and p80

We decided to include values of P50 and P80 in the leaf traits model, defined by [@anderegg_meta-analysis_2016] as the water potentials at which a species loses 50% and 88% [80% by proxy], respectively, of hydraulic conductivity. Values were calculated by (**insert new methods here??**), and were only available for six species (*C. glabra*, *L. tulipifera*, *Q. alba*, *Q. prinus*, *Q. rubra*, and *Q. velutina*). Because of this, the model runs were considered to be incomplete due to the exclusion of the other 8 species. Results revealed neither p50 nor p80 to be significant, thus for the full analysis we decided to drop the two traits in order to include all species in the full analysis.

(see Issue #32)

**other sources???**

### 2019-06-27 manuscript/index.Rmd
Tried to make correct output formats

@@
 output:
-  bookdown::html_book: default
-  bookdown::gitbook: default
+  html_document:
+    fig.width: 12
+    fig.height: 8
+  gitbook:
+    fig_caption: true
 documentclass: book

### 2019-06-28 scripts/canopy_position_analysis.R
Added code to make separate csv files for the processed datasets

@@ -1001,6 +1001,8 @@ trees_all$position_all <- gsub("S", "suppressed", trees_all$position_all)
 #this csv has avg/min/max dbh for each canopy position by sp
 # positionsp <- read.csv("data/core_chronologies_by_crownposition.csv")
 
+##5h1. write trees_all to csv for use in manuscript code ####
+write.csv(trees_all, "manuscript/tables_figures/trees_all.csv", row.names=FALSE)
 ##5i. remove one bad tree & resistance values >2 ####
 ##fram 140939 has been mislabeled. It is recorded as having a small dbh when that is the second stem. In terms of canopy position, though, it fell between time of coring and when positions were recorded, thus we do not know its position.
 trees_all <- trees_all[!trees_all$tree == 140939, ]
@@ -1008,9 +1010,17 @@ trees_all <- trees_all[trees_all$resist.value <=2,]
 
 ##5j. subset for either leaf hydraulic traits or biophysical ####
 trees_all_traits <- trees_all[complete.cases(trees_all), ]
+write.csv(trees_all_traits, "manuscript/tables_figures/trees_all_traits.csv", row.names=FALSE)
 
 trees_all_bio <- trees_all[c("year", "sp", "tree", "position", "resist.value", "elev.m", "distance.ln.m", "height.ln.m", "position_all")]
 trees_all_bio <- trees_all_bio[complete.cases(trees_all_bio), ]
+write.csv(trees_all_bio, "manuscript/tables_figures/trees_all_bio.csv", row.names=FALSE)
+
+#take out p50 and p80, then keep all leaf traits and additionally height plus position
+trees_all_full <- trees_all[!colnames(trees_all) %in% c("p50.MPa", "p80.MPa")]
+trees_all_full <- trees_all_full[, c(1:11,18:20)]
+trees_all_full<- trees_all_full[complete.cases(trees_all_full), ]
+write.csv(trees_all_full, "manuscript/tables_figures/trees_all_full.csv", row.names=FALSE)
 
 ##5k. make subsets for individual years, combine all to list ####
 x1966 <- trees_all_bio[trees_all_bio$year == 1966, ]
@@ -1569,7 +1579,7 @@ write.csv(summary_models, "tables_figures/results_full_models_combined_years.csv
 
[…]

### 2019-06-28 manuscript/03-methods.Rmd
Added images
Added code to make tables

@@
 
-![Structure of ForestGEO plot](tables_figures/ForestGEO_plot.jpg)
+![Structure of ForestGEO plot](tables_figures/ForestGEO_plot.jpg){width=500px}
 
@@
 In total, 14 species had cores that were considered complete and usable, as seen in Table 1.
-
-Using CRU data **with Valentine's quilts for each grouping**, it was found that all climate variables tested were significant by tree position when using a linear mixed model (**this was done early-on, included here because unsure if will keep for full analysis**).
-
-To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
-Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
-Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
 
@@
 
+Using CRU data **with Valentine's quilts for each grouping**, it was found that all climate variables tested were significant by tree position when using a linear mixed model (**this was done early-on, included here because unsure if will keep for full analysis**).
+
+To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
+Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
+The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
+Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
+
+![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
+
 ### Biophysical traits
 Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-For ease of core analysis, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).
+In determining pointer years, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).
+The full set of canopy classes were used for all other analyses.
+
+```{r, Table-cores-position, eval=TRUE, echo=FALSE, message=FALSE}
+library(RCurl)
+library(dplyr)
+library(tidyr)
+library(data.table)
+library(knitr)
+library(kableExtra)
+
+species <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/master/data/core_list_for_neil.csv?token=AJNRBEIAVYFAN2CJB3Y3UAS5D5BNG"), stringsAsFactors = FALSE)
+
+table <- species[, c(1:3,6,17)]
+
+positions <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_dimensions/tree_crowns/cored_dendroband_crown_position_data/dendro_cored_full.csv"))
+
+table$position_all <- positions$crown.position[match(table$tag, positions$tag)]
+table$position_all <- gsub("D", "dominant", table$position_all)
+table$position_all <- gsub("C", "co-dominant", table$position_all)
+table$position_all <- gsub("I", "intermediate", table$position_all)
+table$position_all <- gsub("S", "suppressed", table$position_all)
+
+table1 <- table %>%
+  group_by(sp) %>%
+  summarize(n_cores=n())
+
+table2 <- table %>%
+  group_by(sp, position_all) %>%
+  summarize(count=n())
+
+table2 <- spread(table2, key=position_all, value=count)
+setnames(table2, old="<NA>", new="prior dead")
+
+table2$n_cores <- table1$n_cores[match(table2$sp, table1$sp)]
+table2 <- table2[, c(1,7,3,2,4:6)]
+table2 <- table2[!table2$sp %in% c("frni", "pist"), ]
+
+kable(table2) %>%
+  kable_styling(bootstrap_options = "striped", full_width = F)
+```
 
@@
 
-Need to say the methodology for all traits used
+Need to say the methodology for all traits used (**Nobby making short list**)
 1. PLA
@@
 
-```{r, echo=FALSE}
+```{r, traits-overview, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
@@
 
+trees_all <- read.csv("tables_figures/trees_all.csv", stringsAsFactors = FALSE)
+
+trees_all <- trees_all[!duplicated(trees_all$sp), ]
+
 traits_table <- data.frame(
   "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Chlorophyll", "Wood density", "TLP", "P50", "P80"),
-  "Unit" = c("", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"))
+  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"))
+traits_table$abb <- c("rp", "PLA", "LMA", "Chl", "WD", "TLP", "p50", "p80")
+
+traits_table$mean <- NA
+traits_table$min <- NA
+traits_table$max <- NA
+
+for (i in seq(along=traits_table$abb[2:8])){
+  trait <- traits_table$abb[2:8][i]
+
+  sub <- trees_all[grepl(trait, colnames(trees_all))]
+  traits_table$mean[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+                              mean(sub[, 1], na.rm=TRUE), 
+                              traits_table$mean)
+
+  traits_table$min[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+                              min(sub[, 1], na.rm=TRUE), 
+                              traits_table$min)
+
+  traits_table$max[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+                              max(sub[, 1], na.rm=TRUE), 
+                              traits_table$max)
+}
+
+traits_table[, c("mean", "min", "max")] <- 
+  sapply(traits_table[, c("mean", "min", "max")], function(x)
+  round(x, 2)
+)
+traits_table$abb <- NULL
 
@@
 Traits and other geographic data were used as independent variables.
+
+## Model runs
+Traits' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].
+Specifically, we first determined the most important biophysical traits before including them in the model with the leaf hydraulic traits [**need to say why?**]

### 2019-06-28 manuscript/04-results.Rmd
Added images
Tried to make correct output formats
Added code to make tables

@@
+```{r include=FALSE, cache=FALSE}
+rm(list = ls(all = TRUE))
+```
 # Results
 
-## Biophysical Models
-Crown position 
+## Model runs
 When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought.
 This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.
+Adding in crown position with the leaf hydraulic traits yields a slightly worse predictive model for drought tolerance, with height remaining as the only significant biophysical variable.
+
+``` {r, biophysical-model, eval=TRUE, echo=FALSE, message=FALSE, cache=TRUE}
+library(lme4)
+library(AICcmodavg) #aictab function
+library(car)
+library(data.table)
+library(knitr)
+library(kableExtra)
+
+trees_all_bio <- read.csv("tables_figures/trees_all_bio.csv", stringsAsFactors = FALSE)
+
+response <- "resist.value"
+effects <- c("position_all", "elev.m", "distance.ln.m", "height.ln.m", "year", "(1|sp)")
+
+#create all combinations of random / fixed effects
+effects_comb <- 
+  unlist( sapply( seq_len(length(effects)), 
+                  function(i) {
+                    apply( combn(effects,i), 2, function(x) paste(x, collapse = "+"))
+                  }))
+
+# pair response with effect and sub out combinations that don't include random effects
+#in general, if two variables are >70% correlated, you can toss one of them without significantly affecting the results
+var_comb <- expand.grid(response, effects_comb) 
+var_comb <- var_comb[grepl("1", var_comb$Var2), ] #only keep in fixed/random combos
+var_comb <- var_comb[grepl("year", var_comb$Var2), ] #keep year in for drought sake
+
+# formulas for all combinations. $Var1 is the response, and $Var2 is the effect
+# for good stats, you should have no more total parameters than 1/10th the number of observations in your dataset
+formula_vec <- sprintf("%s ~ %s", var_comb$Var1, var_comb$Var2)
+
+# create list of model outputs
+lmm_all <- lapply(formula_vec, function(x){
+  fit1 <- lmer(x, data = trees_all_bio, REML=FALSE, 
+               control = lmerControl(optimizer ="Nelder_Mead"))
+  return(fit1)
+})
+names(lmm_all) <- formula_vec
+
+var_aic <- aictab(lmm_all, second.ord=TRUE, sort=TRUE) #rank based on AICc
+
+best <- lmm_all[[8]]
+cof_bio <- data.frame("value" = coef(summary(best))[ , "Estimate"])
+cof_bio <- setDT(cof_bio, keep.rownames = TRUE)[]
+setnames(cof_bio, old="rn", new="coefficients")
+
+```
+``` {r, full-model, eval=TRUE, echo=FALSE, message=FALSE, cache=TRUE}
+trees_all_full <- read.csv("tables_figures/trees_all_full.csv", stringsAsFactors = FALSE)
+
+response <- "resist.value"
+effects <- c("rp", "PLA_dry_percent", "LMA_g_per_m2", "WD_g_per_cm3", "mean_TLP_Mpa", "height.ln.m", "position_all", "year", "(1|sp/tree)")
+
+#create all combinations of random / fixed effects
+effects_comb <- 
+  unlist( sapply( seq_len(length(effects)), 
+                  function(i) {
+                    apply( combn(effects,i), 2, function(x) paste(x, collapse = "+"))
+                  }))
+
+# pair response with effect and sub out combinations that don't include random effects
+#in general, if two variables are >70% correlated, you can toss one of them without significantly affecting the results
+var_comb <- expand.grid(response, effects_comb) 
+var_comb <- var_comb[grepl("1", var_comb$Var2), ] #only keep in fixed/random combos
+var_comb <- var_comb[grepl("year", var_comb$Var2), ] #keep year in for drought sake
+
+# formulas for all combinations. $Var1 is the response, and $Var2 is the effect
+# for good stats, you should have no more total parameters than 1/10th the number of observations in your dataset
+formula_vec <- sprintf("%s ~ %s", var_comb$Var1, var_comb$Var2)
+
+# create list of model outputs
+lmm_all <- lapply(formula_vec, function(x){
+  fit1 <- lmer(x, data = trees_all_full, REML=FALSE, 
+               control = lmerControl(optimizer ="Nelder_Mead"))
+  return(fit1)
+})
+names(lmm_all) <- formula_vec
+
+var_aic <- aictab(lmm_all, second.ord=TRUE, sort=TRUE) #rank based on AICc
+
+best <- lmm_all[[121]]
+cof_full <- data.frame("value" = coef(summary(best))[ , "Estimate"])
+cof_full <- setDT(cof_full, keep.rownames = TRUE)[]
+setnames(cof_full, old="rn", new="coefficients")
+```
+```{r, make-tables, eval=TRUE, echo=FALSE, message=FALSE, results = 'asis'}
+library(knitr)
+library(kableExtra)
+
+cof_bio %>%
+  kable("html", align = 'clc', caption = 'Biophysical Model') %>%
+    kable_styling(full_width = F, position = "float_left")
+
+cof_full %>%
+  kable("html", align = 'clc', caption = 'Full Model') %>%
+    kable_styling(full_width = F, position = "right")
+```
 
@@
 This paper mentions it would be good to have this data with respect to latent affects from drought.
+
+### Vertical Profiles
+
+![Climate variables with height](tables_figures/NEON_vertical_profiles.png){width=500px}

### 2019-06-28 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -1665,6 +1665,8 @@ library(tidyr)
 library(grid)
 library(gridExtra)
 
+trees_all_full <- read.csv("manuscript/tables_figures/trees_all_full.csv", stringsAsFactors = FALSE)
+
 graph_traits <- colnames(trees_all_full[, 7:11])
 color <- c("dark green", "blue", "gold", "purple", "magenta")
 
@@ -1680,7 +1682,12 @@ for(i in seq(along=graph_traits)){
   assign(paste0(trait, "_plot"), p)
 }
 
-grid.arrange(PLA_dry_percent_plot, LMA_g_per_m2_plot, Chl_m2_per_g_plot, mean_TLP_Mpa_plot, WD_g_per_cm3_plot, nrow=2, top = textGrob(expression(bold("Hydraulic Traits by Height"))))
+#arrange all graphs together and save image
+png("manuscript/tables_figures/traits_vs_traits.png", width = 1000, height = 1000, pointsize = 18)
+graph <- grid.arrange(PLA_dry_percent_plot, LMA_g_per_m2_plot, Chl_m2_per_g_plot, mean_TLP_Mpa_plot, WD_g_per_cm3_plot, nrow=2, top = textGrob(expression(bold("Hydraulic Traits by Height"))))
+
+dev.off()
+
 
 test
 #

### 2019-06-29 manuscript/index.Rmd
changed author order

--- 2019-06-14 manuscript/index.Rmd
@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
 output:
-  bookdown::html_book: default
-  bookdown::gitbook: default
+  html_document:
+    fig.width: 12
+    fig.height: 8
+  gitbook:
+    fig_caption: true
 documentclass: book

### 2019-06-29 manuscript/03-methods.Rmd
moved map to supplment, edited site description

@@
 # Methods
-
+## Study site
 Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot [@andersonteixeira_ctfs-forestgeo:_2015].
 SCBI (38°53'36.6"N, 78° 08'43.4"W) [@gonzalezakre_patterns_2016] is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Primary forest composition in the plot is oak-hickory, while elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
-
-![Structure of ForestGEO plot](tables_figures/ForestGEO_plot.jpg){width=500px}
+Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
+Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
 

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
edited description of cores collection

@@
 ## Data Collection
+We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP_stem) (Table 1), which together comprise 97% of whole-ecosystem ANPP_stem.
 Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-Live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
+In 2010-2011, cores were collected from randomly selected live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
 In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
-They were sanded, measured, and standardized in the same process as seen in [@helcoski_growing_2019].
-Cores were then split into canopy and subcanopy groupings as determined in the field.
-Species' inclusion in the study was dependent on the robustness of each species’ cores.
-In total, 14 species had cores that were considered complete and usable, as seen in Table 1.
+Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
 

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
Update 03-methods.Rmd
removed  canopy-subcanopy table, description of quilts analysis.

@@
 Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
+The resulting chronologies have been published in association with [@helcoski_growing_2019]: (GitHub URL), (ITRDB).
 
-|**Species**|**n_cores.canopy**|**n_cores.subcanopy**|**included**
-|----------|-------------|---------------|-----------|
-*Carya cordiformis*|2|11|yes
-*Carya glabra*|9|22|yes
-*Carya ovalis*|9|14|yes
-*Carya tomentosa*|0|13|sub-canopy only
-*Fagus grandifolia*|7|73|yes
-*Fraxinus americana*|20|42|yes
-*Fraxinus nigra*|1|11| sub-canopy only
-*Juglans nigra*|22|9|yes
-*Liriodendron tulipifera*|38|60|yes
-*Quercus alba*|38|23|yes
-*Quercus prinus*|27|32|yes
-*Quercus rubra*|43|26|yes
-*Quercus velutina*|54|23|yes
-
-``` {r sample, echo=FALSE}
-#numbers for the table above come from this table
-# library(RCurl)
-# library(dplyr)
-# species <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/master/data/core_list_for_neil.csv?token=AJNRBELCYZY576ARLE7BXNK5BEF3O"))
-# 
-# test <- species %>%
-#   group_by(sp, crown.position) %>%
-#   summarize(count=n())
-
-```
-
-Using CRU data **with Valentine's quilts for each grouping**, it was found that all climate variables tested were significant by tree position when using a linear mixed model (**this was done early-on, included here because unsure if will keep for full analysis**).
+*Table 1. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits* [this should combine the two remaining tables in this section]
 

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
rearrangement

@@
 
-*Table 1. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits* [this should combine the two remaining tables in this section]
-
-To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
-Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
-Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
-
-![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
-
-### Biophysical traits
 Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
@@
 The full set of canopy classes were used for all other analyses.
+
+Elevation for the trees was extracted from a USGS DEM in ArcMap.
+Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.
+
+Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015], and automatic LiDAR [@stovall_terrestrial_2018].
+Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
+The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
+Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees.
+For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
+
+Equations required the DBH in each individual drought year.
+As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems.
+Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,
+
+`diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))`
+
+First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species.
+Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.
+
+Hydraulic traits were collected from SCBI and are summarized in Table 1. Ring porosity qualifications were obtained from [@andersonteixeira_size-related_2015].
+
+Need to say the methodology for all traits used (**Nobby making short list**)
+1. PLA
+2. LMA
+3. Chlorophyll
+4. Wood density
+5. TLP
+6. P50 = psi_0.5_kl50
+7. P80 = psi_0.5_kl80
+
+*Table 1. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits* [this should combine the two remaining tables in this section]
 
@@
 ```
-
-Elevation for the trees was extracted from a USGS DEM in ArcMap.
-Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.
-
-Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015], and automatic LiDAR [@stovall_terrestrial_2018].
-Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
-Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees.
-For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
-
-Equations required the DBH in each individual drought year.
-As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems.
-Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,
-
-`diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))`
-
-First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species.
-Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.
-
-### Leaf Hydraulic Traits
-Hydraulic traits were collected from SCBI and are summarized in Table 1. Ring porosity qualifications were obtained from [@andersonteixeira_size-related_2015].
-
-Need to say the methodology for all traits used (**Nobby making short list**)
-1. PLA
-2. LMA
-3. Chlorophyll
-4. Wood density
-5. TLP
-6. P50 = psi_0.5_kl50
-7. P80 = psi_0.5_kl80
 
@@
 
+## Identification of drought years
+To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
+Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
+The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
+Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
+
+![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
+
 After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable.
@@
 
-## Model runs
+## Analysis
 Traits' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
update on hydraulic traits

@@
 
-Hydraulic traits were collected from SCBI and are summarized in Table 1. Ring porosity qualifications were obtained from [@andersonteixeira_size-related_2015].
-
-Need to say the methodology for all traits used (**Nobby making short list**)
+Hydraulic traits were collected from SCBI and are summarized in Table 1.
+In August 2018, we collected leaf samples from three individuals of each species ...
+(**Nobby's description of methods for the following**)
 1. PLA
 2. LMA
-3. Chlorophyll
 4. Wood density
@@
 
-*Table 1. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits* [this should combine the two remaining tables in this section]
+**Table 1.
+Species analyzed here, listed in descending order of ANPP_stem.
+n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section.
+Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each canopy position) species means for each trait]
 

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
lots of text edits

@@
 ## Study site
-Research was conducted at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA, with cores taken from trees on the 25.6 ha ForestGEO (Global Earth Observatory) study plot [@andersonteixeira_ctfs-forestgeo:_2015].
-SCBI (38°53'36.6"N, 78° 08'43.4"W) [@gonzalezakre_patterns_2016] is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
+Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [@andersonteixeira_ctfs-forestgeo:_2015].
+SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
 Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
@@
 
-## Data Collection
-We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP_stem) (Table 1), which together comprise 97% of whole-ecosystem ANPP_stem.
+## Data collection and preparation
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols...
+[DETAILS].
+Here, we use ...(*very brief summary of which data we use for what purpose-- but some will be better integrated with methods descriptions below*)
+
+We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP_stem) (Table 1), which together comprised 97% of whole-ecosystem ANPP_stem between 2008 and 2013.
 Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
@@
 
-Canopy classes were observed in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-In determining pointer years, these labels were condensed to canopy (including dominant and codominant) and subcanopy (intermediate and suppressed).
-The full set of canopy classes were used for all other analyses.
+Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (*GitHUB URL*).
+Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015] (CITATION NOT APPROPRIATE), and automatic LiDAR [@stovall_terrestrial_2018].
+Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
+The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
+Species-specific height allometries were developed (Table S# - **ADD THIS TABLE TO SI**).
+For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
+
+For each tree, we combined tree-ring records and allometric equations of height and bark thickness to retroactively calculate DBH and estimate height for the years 1950-2009.
+Prior DBH was estimated using the following equation:
+
+`diam_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008)) + 2*(bark.depth_1999)`  ([issue #35](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/35)) (**clean up this equation--put a generic year instead of 1999, use sum symbol, etc.**)
+
+Here, *ring.width* was measured from cores. Bark thickness was estimated from species-specific allometries based on the bark thickness data of [@andersonteixeira_size-related_2015].
+Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#- **create table to give these equations in SI**) and then used these to estimate bark thickness based on DBH.
+
+Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
+While some trees undoubtedly changed position, an analysis of canopy position relative to height (Fig.
+XX) and height change since *1959* indicated that change was likely slow.
+[**work on this-- provide details?**]
 
@@
 Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.
-
-Height data for trees was collected from several researchers at SCBI from 2012 to 2019, with methods including manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015], and automatic LiDAR [@stovall_terrestrial_2018].
-Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
-Log-log, species-specific regression equations were developed, which were used to calculate heights of the study trees.
-For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
-
-Equations required the DBH in each individual drought year.
-As part of the ForestGEO five-year census, there are DBH(diameter at breast height) measurements for all stems.
-Using the data from 2008 and a dataset of bark thickness for the plot collected from previous studies, we devised an equation to retroactively calculate DBH for all trees,
-
-`diam_nobark_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008))`
-
-First, we generated log-log regression equations for bark thickness based on DBH, and then determined mean bark thickness per species.
-Ring widths were determined from the processed cores, and once the diameter without bark was calculated for a certain year, we added the mean bark thickness from the regression equations.
 
@@
 Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
+[**This needs work.
+It's very complex to describe what you did, and in retrospect not well-aligned with our current analysis.
+It would be much more straightforward to simply combine ALL cores in the pointRes package to identify potential pointer years.
+Is this a lot of work?
+In any case, we should find a way to present this analysis in one paragraph (no tables!) of main body text.
+If you need to stick with species/canopy position-level analysis, any tables and detailed description should go in SI**]
 
+**Figure 1.
+Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
 ![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
 
+## Analysis
 After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable.
@@
 
-## Analysis
 Traits' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd
remove issue #35

@@
 
-`diam_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008)) + 2*(bark.depth_1999)`  ([issue #35](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/35)) (**clean up this equation--put a generic year instead of 1999, use sum symbol, etc.**)
+`diam_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008)) + 2*(bark.depth_1999)`  (**clean up this equation--put a generic year instead of 1999, use sum symbol, etc.**)
 

### 2019-06-29 manuscript/03-methods.Rmd
Update 03-methods.Rmd

@@
 
-## Identification of drought years
+## Climate and drought years
+[**add description of climate data used in Fig. 1, NEON vertical profiles**]
+
 To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
@@
 Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from [@helcoski_growing_2019].
 ![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}

### 2019-06-30 manuscript/04-results.Rmd
Updated manuscript
rearranged some content

@@
 
-## Model runs
-When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought.
-This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.
-Adding in crown position with the leaf hydraulic traits yields a slightly worse predictive model for drought tolerance, with height remaining as the only significant biophysical variable.
+## tree and trait distribution within the biophysical setting
+
+**Figure 2. Vertical profiles in microclimate, crown positions, and traits.**[combines the two following figures]
+
+![Climate variables with height](tables_figures/NEON_vertical_profiles.png){width=500px}
+
+![Leaf hydraulic traits with log-values of height](tables_figures/traits_vs_traits.png){width=500px}
+
+[consider an analysis of trait distribution in relation to topography??]
+
+## droughts(?)
+[description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?]
+
+## [1- tree size and exposure]
+We evaluated the predictions outlined in the introduction
+P1.0. Drought resistance decreased with DBH at time of drought.
+P1.1. Drought resistance decreased with height at time of drought.
+P1.2. Trees currently in a dominant crown position suffered more during drought. Intermediate and suppressed trees also suffered more (in model with height)
+P1.3. Resistance was independent of elevation and distance from stream.
 
@@
 ```
-
-From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity.
-This paper mentions it would be good to have this data with respect to latent affects from drought.
-
-### Vertical Profiles
-
-![Climate variables with height](tables_figures/NEON_vertical_profiles.png){width=500px}

### 2019-06-30 manuscript/05-discussion.Rmd
rearranged some content

@@
 *1. paragraph summarizing main results--> primary conclusions*
+When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought.
+This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.
+Adding in crown position with the leaf hydraulic traits yields a slightly worse predictive model for drought tolerance, with height remaining as the only significant biophysical variable.
 
@@
 However, the increase in drought resistance with distance from stream was less for small than large trees (H1.3b), indicating a potential importance of root depth/volume in conferring drought resistance.
+
+#### misc content to integrate
+From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity.
+This paper mentions it would be good to have this data with respect to latent affects from drought.

### 2019-06-30 manuscript/04-results.Rmd
Update 04-results.Rmd
Updated to show r2 values in table with coefficients

@@
 
-## [1- tree size and exposure]
+## 1- tree size and exposure
 We evaluated the predictions outlined in the introduction
@@
 
+[**HYPOTHESIS TESTING TABLE**]
+
+## 2- traits
+
 ``` {r, biophysical-model, eval=TRUE, echo=FALSE, message=FALSE, cache=TRUE}
@@
 library(car)
+library(piecewiseSEM)
 library(data.table)
@@
 trees_all_bio <- read.csv("tables_figures/trees_all_bio.csv", stringsAsFactors = FALSE)
+trees_all_bio$year <- as.character(trees_all_bio$year)
 
@@
 setnames(cof_bio, old="rn", new="coefficients")
-
+r <- rsquared(best)
+r <- r[, c(1,6)]
+r$Response <- as.character(r$Response)
+r[1,1] <- "r^2"
+setnames(r, old=c("Response", "Conditional"), new=c("coefficients", "value"))
+cof_bio <- rbind(cof_bio, r)
 ```
@@
 trees_all_full <- read.csv("tables_figures/trees_all_full.csv", stringsAsFactors = FALSE)
+trees_all_full$year <- as.character(trees_all_full$year)
 
@@
 
-best <- lmm_all[[121]]
+best <- lmm_all[[100]]
 cof_full <- data.frame("value" = coef(summary(best))[ , "Estimate"])
@@
 setnames(cof_full, old="rn", new="coefficients")
+r <- rsquared(best)
+r <- r[, c(1,6)]
+r$Response <- as.character(r$Response)
+r[1,1] <- "r^2"
+setnames(r, old=c("Response", "Conditional"), new=c("coefficients", "value"))
+cof_full <- rbind(cof_full, r)
 ```

### 2019-06-30 manuscript/03-methods.Rmd
Update 03-methods.Rmd

@@
 As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-While some trees undoubtedly changed position, an analysis of canopy position relative to height (Fig.
+While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
 XX) and height change since *1959* indicated that change was likely slow.
@@
 n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section.
-Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each canopy position) species means for each trait]
+Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]
 
@@
 In any case, we should find a way to present this analysis in one paragraph (no tables!) of main body text.
-If you need to stick with species/canopy position-level analysis, any tables and detailed description should go in SI**]
+If you need to stick with species/crown position-level analysis, any tables and detailed description should go in SI**]
 

### 2019-06-30 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -778,6 +778,7 @@ range(bark$predict_barkthick.mm - bark$bark.depth.mm)
 dbh <- trees_all[, c(1:4)]
 scbi.stem1 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem1.csv"))
 dbh$dbh2008.mm <- scbi.stem1$dbh[match(dbh$tree, scbi.stem1$tag)]
+dbh$dbh2008.mm <- as.numeric(dbh$dbh2008.mm)
 
 mean_bark <- aggregate(bark$bark.depth.mm, by=list(bark$species), mean) #mm
 colnames(mean_bark) <- c("sp", "mean_bark_2008.mm")
@@ -1578,7 +1579,7 @@ for (i in seq_along(model_df)){
 write.csv(summary_models, "tables_figures/results_full_models_combined_years.csv", row.names=FALSE)
 
 ##6aii. coefficients ####
-best <- lmm_all[[8]]
+best <- lmm_all[[58]]
 cof <- data.frame("value" = coef(summary(best))[ , "Estimate"])
 
 lm_new <- lm(resist.value ~ dbh_ln*distance_ln.m, data=trees_all, REML=FALSE)
@@ -1593,8 +1594,9 @@ aic_top <- var_aic %>%
 ##6aiii. base code for running multiple models through AICc eval ####
 #define response and effects
 response <- "resist.value"
-effects <- c("position_all", "elev.m", "distance.ln.m", "height.ln.m", "year", "(1|sp)")
+# effects <- c("position_all", "elev.m", "distance.ln.m", "height.ln.m", "year", "(1|sp)")
 # effects <- c("rp", "PLA_dry_percent", "LMA_g_per_m2", "WD_g_per_cm3", "mean_TLP_Mpa", "p50.MPa", "p80.MPa", "year", "(1|sp/tree)")
+effects <- c("rp", "PLA_dry_percent", "LMA_g_per_m2", "WD_g_per_cm3", "mean_TLP_Mpa", "height.ln.m", "year", "(1|sp/tree)")
 
 #create all combinations of random / fixed effects
 effects_comb <- 
@@ -1615,7 +1617,7 @@ formula_vec <- sprintf("%s ~ %s", var_comb$Var1, var_comb$Var2)
 
 # create list of model outputs
[…]

### 2019-06-30 manuscript/03-methods.Rmd
adjusted subheadings

@@
 # Methods
-## Study site
+### Study site
 Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [@andersonteixeira_ctfs-forestgeo:_2015].
@@
 
-## Data collection and preparation
+### Data collection and preparation
 The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols...
@@
 
-## Climate and drought years
+### Climate and drought years
 [**add description of climate data used in Fig. 1, NEON vertical profiles**]
@@
 
-## Analysis
+### Analysis
 After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable.

### 2019-06-30 manuscript/04-results.Rmd
adjusted subheadings

@@
 
-## tree and trait distribution within the biophysical setting
+### tree and trait distribution within the biophysical setting
 
@@
 
-## droughts(?)
+### droughts(?)
 [description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?]
 
-## 1- tree size and exposure
+### 1- tree size and exposure
 We evaluated the predictions outlined in the introduction
@@
 
-## 2- traits
+### 2- traits
 

### 2019-07-01 manuscript/02-intro.Rmd
Update 02-intro.Rmd

# Introduction

Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.

Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS). Therefore, in order to understand whole-forest response to drought, we need to know how responses vary by tree size/ species. To do so, there are three fundamental questions that must be addressed:

*First, what drives the observed tendency for large trees to suffer more during drought?*  
[@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees. It is also expected that roots play a role. These hypotheses need to be tested.

*Second, how do species' traits influence drought response?* 
Analyzed drought responses on the species level doesn't get at mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/ traditionally measured traits such as wood density and SLA (Medeiros et al.).

*Third, how do tree size and functional traits interact to influence drought response?*
Its possible that pattern observed by Bennett et al. could be caused by smaller trees being more drought resistant. Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.

*Questions & Hypotheses*

**1. Why do larger trees suffer greater growth declines during drought?**

1.0. Our forest displays the same trend as most forests globally [@bennett_larger_2015]. (Note that Bennett et al. 2015 identified only one study on tree growth responses to drought in the Eastern US temperate deciduous biome. We know little about how tree size shapes drought response in this biome.)

1.1. Height is a strong predictor of drought stress.

    * P1.1-Drought response increases with height at time of drought.

1.2. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.

    * P1.2a- Trees currently in a dominant canopy position suffered more during drought.

1.3. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.

     * P1.3a- drought response increases with elevation and/or distance from stream
     * P1.3b- There is a dbh\*elev interaction, elevation (relative to stream?) matters less for big trees with larger (and potentially deeper) root system.

**2.Do species functional traits predict drought response?**

     * P2.1- TLP correlates negatively with drought resistance (NEVER tested), 
     * P2.2- diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)

**3.How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**

3.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
3.1_alt. Larger trees suffer more because they have more drought vulnerable traits.

     * P3.1a- TLP is lower (larger negative) in taller/canopy trees
     * P3.1b- diffuse porous species more common in understory

     * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.

### 2019-07-09 manuscript/04-results.Rmd
Update 04-results.Rmd

@@
 ### droughts(?)
-[description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?]
+[description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?.
+See [Slette et al.
+(2019)](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1111/gcb.14747)]
 

### 2019-07-24 manuscript/03-methods.Rmd
Updated references and methods

@@
 
-`diam_1999 = dbh2008 - 2*(bark.depth2008) - 2*(sum(ring.width1999:ring.width2008)) + 2*(bark.depth_1999)`  (**clean up this equation--put a generic year instead of 1999, use sum symbol, etc.**)
+$$
+diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.width1999:ring.width2008) + 2*(bark.depth_YEAR)
+$$
 
@@
 
-To determine focus drought years, we used the pointRes package [@R-pointRes] in R (version 3.5.2) to determine resilience metrics.
-Specifically, we calculated pointer years (main drought years) if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-The top 6 pointer years were compared with PDSI (Palmer Drought Severity Index) values for the region, obtained from NOAA.
-Based on this comparison, we then took the top three pointer years of 1964-1966 (resistance metric averaged due to a single, prolonged drought), 1977, and 1999.
-[**This needs work.
-It's very complex to describe what you did, and in retrospect not well-aligned with our current analysis.
-It would be much more straightforward to simply combine ALL cores in the pointRes package to identify potential pointer years.
-Is this a lot of work?
-In any case, we should find a way to present this analysis in one paragraph (no tables!) of main body text.
+To accurately understand climate sensitivity, this study used a specific definition of drought, which is not a common practice [@slette_how_2019].
+We used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined by [@lloret_components_2011] as the ratio between the performance during and before the disturbance.
+Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
+These were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values for each year, which yielded a set of three periods that were consistently shown as drought: 1964-1966, 1977, and 1999.
+
+[**We should find a way to present this analysis in one paragraph (no tables!) of main body text.
 If you need to stick with species/crown position-level analysis, any tables and detailed description should go in SI**]

### 2019-07-24 scripts/canopy_position_analysis.R
Determined best models and table for presentation

@@ -522,9 +522,9 @@ library(ggthemes) #for removing graticules when making pdf
 library(rgeos) #for distance calculation
 library(RCurl) #for reading in URLs
 library(readxl)
-library(raster)
-library(elevatr)
-library(dynatopmodel)
+library(raster) #for TWI
+library(elevatr) #for TWI
+library(dynatopmodel) #for TWI
 
 ##5a. add in ring porosity qualifications ####
 ring_porosity <- data.frame("sp" = c("cagl",  "caovl", "cato", "fagr", "fram", "juni",  "litu",  "pist",  "qual",  "qupr",  "quru",  "quve", "caco", "frni"), "rp" = c("ring", "ring", "ring", "diffuse", "ring", "semi-ring", "diffuse", NA, "ring", "ring", "ring", "ring", "ring", "ring"))
@@ -1084,125 +1084,7 @@ library(piecewiseSEM) #for R^2 values for all model outputs in a list
 library(MuMIn) #for R^2 values of one model output
 library(stringr)
 
-##6ai. test predictions for paper and put in table (combined dataset only) ####
-##testing each trait
-sum_mod_traits <- data.frame(
-  "prediction" = c(1.1, 2.1, 2.2, 3.1, 3.2, 3.3, 3.4, 4.1, 4.2, 4.3, 4.4, 4.5),
-  "variable" = c("year", "dbh.ln.cm", "height.ln.m","elev.m", "distance.ln.m", "position_all",  "TWI", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
-  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "elevation", "stream.distance", "crown.position", "topographic.wetness.index", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
-  "null_model" = 
-    c("resist.value ~ (1|sp/tree)",
-      "resist.value ~ year+(1|sp/tree)",
-      "resist.value ~ year+(1|sp/tree)",
-      "resist.value ~ height.ln.m+year+(1|sp/tree)",
-      "resist.value ~ height.ln.m+year+(1|sp/tree)",
-      "resist.value ~ height.ln.m+year+(1|sp/tree)",
-      "resist.value ~ height.ln.m+year+(1|sp/tree)",
[…]

### 2019-07-24 scripts/graphs_for_manuscript.R
Update graphs_for_manuscript.R

@@ -133,48 +133,68 @@ for (i in seq(along=2:ncol(leaf_traits))){
 
 
 
-#4b. create height groupings ####
+#4b. create height groupings and graphs ####
 scbi18$height.m <- scbi18_ht$height.m[match(scbi18$stemID, scbi18_ht$stemID)]
 
 range(scbi18$height.m)
-scbi18$bins <- cut(scbi18$height.m, breaks=c(-Inf,10,15,20,25,30,35,40,45, Inf), 
+scbi18$bins <- cut(scbi18$height.m, breaks=c(-Inf,10,15,20,25,30,35,40,45,Inf), 
    labels=c("5-10","10-15", "15-20", "20-25", "25-30", "30-35", "35-40", "40-45", "45-50"))
 
-tlp <- scbi18 %>%
+# tlp ####
+trait_tlp <- scbi18 %>%
    group_by(bins) %>%
    summarize(avg = mean(mean_TLP_Mpa, na.rm=TRUE),
              dev = sd(mean_TLP_Mpa, na.rm=TRUE),
              no_data = (sum(is.na(mean_TLP_Mpa))/length(mean_TLP_Mpa)))
 
-tlp$sdmin <- tlp$avg - tlp$dev
-tlp$sdmax <- tlp$avg + tlp$dev
-
-
-#graph results ####
-plot(scbi18$TWI, scbi18$mean_TLP_Mpa)
+trait_tlp$sdmin <- trait_tlp$avg - trait_tlp$dev
+trait_tlp$sdmax <- trait_tlp$avg + trait_tlp$dev
 
 #TLP by height bins
-ggplot(tlp, aes(x = bins, y = avg)) +
+ggplot(trait_tlp, aes(x = avg, y = bins)) +
    geom_point() +
-   geom_errorbar(aes(x = bins, ymin = sdmin, ymax = sdmax)) +
-   scale_y_continuous(breaks=c(-2.8,-2.6,-2.4,-2.2,-2.0,-1.8), limits=c(-2.8,-1.8)) +
-   xlab("Tree height (m)") +
-   ylab("Mean TLP (MPa)") +
+   ggplot2::geom_errorbarh(aes(y = bins, xmin = sdmin, xmax = sdmax)) +
+   scale_x_continuous(breaks=c(-2.8,-2.6,-2.4,-2.2,-2.0,-1.8), limits=c(-2.8,-1.8)) +
+   ylab("Tree height (m)") +
[…]

### 2019-07-24 manuscript/04-results.Rmd
Updated methods

--- 2019-06-30 manuscript/04-results.Rmd
@@
 ### droughts(?)
-[description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?]
+description of the 3 droughts in terms of timing, duration, severity, proportion of trees showing pointer year..?. See
+
+Proportion of all trees showing growth deficit for that year (meeting our threshold)
+1966: 28.18% (20.88% if doing all trees 1964-1966)
+1977:31.23%
+1999: 26.21%
+
+Proportion of species-canopy classes (e.g. fagr_canopy, quve_subcanopy) that have >50% of their trees showing a pointer year
+1966: 22.73%
+1977: 22.73%
+1999: 13.64%
 

### 2019-07-24 scripts/canopy_position_analysis.R
Updated code for paper figures

@@ -269,9 +269,9 @@ widths_subcan <- NULL
 
 ### df for pointer years of all trees combined ####
 full_ind <- rbind(canopy_table, subcanopy_table) #full table of indices for canopy and subcanopy cores
+
 pointers <- full_ind[full_ind$nature == -1, ]
 
-library(dplyr)
 years_point <- count(pointers, vars=year) #counts the occurrences of each unique year
 colnames(years_point) <- c("yr", "n.pointer")
 years_point <- years_point[order(years_point$n.pointer, decreasing=TRUE), ]
@@ -511,6 +511,29 @@ ggplot(data = pdsi_true) +
 #plot and manually subset to 1950+
 # plot_ly(pdsi_true, x=pdsi_true$year, y=pdsi_true$resid, type="scatter", mode="line")
 
+##4e. Statistics for paper ####
+
+#these numbers are
+#1. subset for year
+#2. proportion of all trees that show growth deficit meeting the threshold we defined
+#3. proportion of species-canopy classes that have >50% trees meeting growth deficit threshold
+
+sum_table <- data.frame("year" = c(1966, 1977, 1999), 
+                        "prop_growth" = NA, 
+                        "prop_pointer" = NA)
+
+for (i in seq(along=sum_table$year)){
+  full_ind_sub <- full_ind[full_ind$year %in% sum_table$year[[i]],  ]
+  
+  sum_table$prop_growth[[i]] <- (sum(full_ind_sub$nb.series*(full_ind_sub$perc.neg/100))/sum(full_ind_sub$nb.series))*100
+  
+  sum_table$prop_growth[[i]] <- round(sum_table$prop_growth[[i]], 2)
+  
+  sum_table$prop_pointer[[i]] <- (nrow(full_ind_sub[full_ind_sub$nature == -1, ])/nrow(full_ind_sub))*100
+  
+  sum_table$prop_pointer[[i]] <- round(sum_table$prop_pointer[[i]], 2)
+}
+
 ##########################################################################################
 #5. add in climate and growth variables ####
 library(ggplot2)
@@ -545,7 +568,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
[…]

### 2019-07-24 scripts/graphs_for_manuscript.R

@@ -23,7 +23,7 @@ scbi.stem3$dbh <- as.numeric(scbi.stem3$dbh)
 
 current_ht <- trees_all[!duplicated(trees_all$tree), ]
 current_ht$year <- 2018
-current_ht <- current_ht[,c(1:4,15:17,19:21)]
+current_ht <- current_ht[,c(1:4,13:15,17:19)]
 
 current_ht$dbh_old.mm <- scbi.stem3$dbh[match(current_ht$tree, scbi.stem3$tag)]
 current_ht$dbh_old.cm <- current_ht$dbh_old.mm/10
@@ -64,7 +64,7 @@ heights <-
   geom_boxplot() +
   ggtitle("Height vs crown position 2018") +
   xlab("crown position") +
-  ylab("height(m)") +
+  ylab("Height (m)") +
   scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits=c(0,60)) +
   theme_minimal()
 
@@ -118,13 +118,11 @@ scbi18 <- sigeo_coords
 scbi18 <- scbi18[, c(2:5,20:21)]
 scbi18_x <- scbi18[, c(5:6)]
 topo <- raster("data/plot_TWI.tif")
-twi <- extract(topo, scbi18_x, method="simple")
+twi <- raster::extract(topo, scbi18_x, method="simple")
 scbi18$TWI <- twi
 
 #add in leaf traits (no rp because it's categorical)
-leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEJQMFQQQHLV5MJVM7K5HCCRC"), stringsAsFactors = FALSE)
-
-leaf_traits <- leaf_traits[, c(1,8,12,26,28)]
+leaf_traits <- trees_all[, c(2,7:10)]
 
 for (i in seq(along=2:ncol(leaf_traits))){
    trait <- colnames(leaf_traits[2:ncol(leaf_traits)])
@@ -151,9 +149,10 @@ trait_tlp$sdmin <- trait_tlp$avg - trait_tlp$dev
 trait_tlp$sdmax <- trait_tlp$avg + trait_tlp$dev
 
 #TLP by height bins
-ggplot(trait_tlp, aes(x = avg, y = bins)) +
+graph_tlp <- 
+   ggplot(trait_tlp, aes(x = avg, y = bins)) +
    geom_point() +
[…]

### 2019-07-24 scripts/vertical_height_neon.R

@@ -157,7 +157,7 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
   }
   
   
-  assign(paste0(dp$data[[i]], "_plot"),
+  graph <-
     data_analy %>%
       arrange(verticalPosition) %>%
       ggplot() +
@@ -166,54 +166,62 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
       geom_point(aes(x = mmin, y = vertPos_jitter, color = month_f), shape=17, position = "jitter") +
       geom_path(aes(x = mmax, y = vertPos_jitter, color = month_f, linetype = "Max"), size = 1) +
       geom_path(aes(x = mmin, y = vertPos_jitter, color = month_f, linetype = "Min"), size = 1) +
-      geom_errorbarh(aes(xmin=mmax-sd_min, xmax=mmax+sd_max, y=vertPos_jitter, color = month_f, linetype = "Max", height=.8)) +
-      geom_errorbarh(aes(xmin=mmin-sd_min, xmax=mmin+sd_max, y=vertPos_jitter, color = month_f, linetype = "Min", height=.8)) +
-      labs(x = dp$xlabs[[i]], y = " ") +
+      ggplot2::geom_errorbarh(aes(xmin=mmax-sd_min, xmax=mmax+sd_max, y=vertPos_jitter, color = month_f, linetype = "Max", height=0.8)) +
+      ggplot2::geom_errorbarh(aes(xmin=mmin-sd_min, xmax=mmin+sd_max, y=vertPos_jitter, color = month_f, linetype = "Min", height=0.8)) +
+      labs(x = dp$xlabs[[i]], y = "Height (m)") +
       scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits=c(0,60)) +
       theme_grey() +
       guides(linetype = guide_legend("Line type"))
-  )
+  
+  if(!i == 1){
+    graph <- graph + theme(axis.title.y = element_blank(), axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  }
+  
+  assign(paste0(dp$data[[i]], "_plot"), graph)
 }
 
-#extract legend
-#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
-g_legend<-function(a.gplot){
-  tmp <- ggplot_gtable(ggplot_build(a.gplot))
[…]

### 2019-07-30 manuscript/index.Rmd
add Atticus as coauthor

--- 2019-06-14 manuscript/index.Rmd
@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Neil Pederson, Lawren Sack, Joseph Zailaa, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
 output:
-  bookdown::html_book: default
-  bookdown::gitbook: default
+  html_document:
+    fig.width: 12
+    fig.height: 8
+  gitbook:
+    fig_caption: true
 documentclass: book

### 2019-07-30 manuscript/01-abstract.Rmd
Update 01-abstract.Rmd
pasted in relevant content from AGU abstract

# Abstract

*content from AGU abstract:*

Here, we integrate forest census data from a 25.6-ha ForestGEO plot in Virginia (USA), tree-ring records for 12 species representing ##% of woody productivity, leaf hydraulic trait measurements, and microhabitat data.

Individual-level growth responses to three droughts (1964-66, 1977, 1999) were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.

We conclude that when droughts occur, large dominant trees, drought sensitive species, and individuals in moist microhabitats are likely to be most strongly affected.

### 2019-07-30 manuscript/03-methods.Rmd
Updated methods
Updated paper

@@
 Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (*GitHUB URL*).
-Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_ctfs-forestgeo:_2015] (CITATION NOT APPROPRIATE), and automatic LiDAR [@stovall_terrestrial_2018].
+Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
 Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
@@
 For each tree, we combined tree-ring records and allometric equations of height and bark thickness to retroactively calculate DBH and estimate height for the years 1950-2009.
-Prior DBH was estimated using the following equation:
+Prior DBH was estimated using the following equation, using 2008 as the earliest year for having reliable DBH measurements:
 
 $$
-diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.width1999:ring.width2008) + 2*(bark.depth_YEAR)
+diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depth_YEAR)
 $$
@@
 
-Elevation for the trees was extracted from a USGS DEM in ArcMap.
-Distance to water was calculated as the shortest distance in meters between each individual tree and the major streams running through the ForestGEO plot.
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.
 
@@
 5. TLP
-6. P50 = psi_0.5_kl50
-7. P80 = psi_0.5_kl80
 
@@
 
-[**We should find a way to present this analysis in one paragraph (no tables!) of main body text.
-If you need to stick with species/crown position-level analysis, any tables and detailed description should go in SI**]
-
 **Figure 1.
@@
 ### Analysis
-After the data was collected, linear mixed models were run with the resistance value of each tree for each pointer year (determined from the pointRes package) as the response variable.
-Traits and other geographic data were used as independent variables.
+Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
+[individual_tested_traits].
+Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable.
+Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions.
+First, we analyzed the combined scenario to determine if "year" was significant.
+Upon establishing this, we tested height and DBH as size parameters.
+Although both were significant, height was kept due to its larger delta AICc compared with the null model.
+We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year.
+This yielded Figure ???
+(cand_full).
+All variables with dAICc >2 were used as candidates for each scenario's best model (figure ????
+(tested_traits_best))
 
-Traits' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].
-Specifically, we first determined the most important biophysical traits before including them in the model with the leaf hydraulic traits [**need to say why?**]
+```{r, individual_tested_traits, eval=TRUE, echo=FALSE, message=FALSE}
+traits_test <- read.csv("tables_figures/tested_traits_all.csv")
+kable(traits_test) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
+```
+
+```{r, all_candidates, eval=TRUE, echo=TRUE, message=FALSE}
+candidates <- read.csv("tables_figures/candidate_traits.csv")
+kable(candidates) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
+```
+
+```{r, tested_traits_best, eval=TRUE, echo=FALSE, message=FALSE}
+traits_best <- read.csv("tables_figures/tested_traits_best.csv")
+kable(traits_best) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
+```

### 2019-07-30 scripts/graphs_for_manuscript.R
Updated graphs for manuscript

@@ -62,13 +62,12 @@ heights <-
   aes(x = position_all, y = height.m, group = position_all) +
   # aes(x=position_all, y=height.m, fill=year) +
   geom_boxplot() +
-  ggtitle("Height vs crown position 2018") +
   xlab("crown position") +
-  ylab("Height (m)") +
+  ylab("Height [m]") +
   scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits=c(0,60)) +
   theme_minimal()
 
-#2a. get height data for all trees >10cm dbh in census ####
+#2b. get height data for all trees >10cm dbh in census ####
 scbi.stem3 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
 
 scbi.stem3$dbh <- as.numeric(scbi.stem3$dbh)
@@ -105,6 +104,8 @@ scbi18_ht$height.m <- exp(scbi18_ht$height.ln.m) #used below in #4
 library(ggstance)
 library(raster)
 library(dplyr)
+library(Hmisc)
+library(magrittr)
 
 #4a. get TWI values for all trees in census, add trait values ####
 scbi.stem3$dbh <- ifelse(is.na(scbi.stem3$dbh), 0, scbi.stem3$dbh)
@@ -120,6 +121,7 @@ scbi18_x <- scbi18[, c(5:6)]
 topo <- raster("data/plot_TWI.tif")
 twi <- raster::extract(topo, scbi18_x, method="simple")
 scbi18$TWI <- twi
+scbi18$TWI.ln <- log(scbi18$TWI)
 
 #add in leaf traits (no rp because it's categorical)
 leaf_traits <- trees_all[, c(2,7:10)]
@@ -133,35 +135,59 @@ for (i in seq(along=2:ncol(leaf_traits))){
 
 #4b. create height groupings and graphs ####
 scbi18$height.m <- scbi18_ht$height.m[match(scbi18$stemID, scbi18_ht$stemID)]
+scbi18$height.ln.m <- log(scbi18$height.m)
 
-range(scbi18$height.m)
+range(scbi18$height.ln.m)
[…]

### 2019-07-30 scripts/vertical_height_neon.R

@@ -26,7 +26,7 @@ library(gridExtra)
 dp <- data.frame("data" = c("SAAT", "wind", "biotemp", "RH", "SR"),
                  "id" = c("DP1.00002.001", "DP1.00001.001", "DP1.00005.001", "DP1.00098.001", "DP1.00014.001"),
                  "value" = c("tempSingleMean", "windSpeedMean", "bioTempMean", "RHMean", "difRadMean"),
-                 "xlabs" = c("Mean Air Temperature (C)", "Mean Windspeed (m/s)", "Mean Infrared Biological Temperature (C)", "Relative Humidity", "Mean shortwave downward radiation (W/m2)"))
+                 "xlabs" = c("Mean Air Temperature [°C]", "Mean Windspeed [m/s]", "Mean Infrared Biological Temperature [°C]", "Relative Humidity [%]", "Mean shortwave downward radiation [W/m^2]"))
 
 
 dp[] <- lapply(dp, as.character)
@@ -168,7 +168,7 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
       geom_path(aes(x = mmin, y = vertPos_jitter, color = month_f, linetype = "Min"), size = 1) +
       ggplot2::geom_errorbarh(aes(xmin=mmax-sd_min, xmax=mmax+sd_max, y=vertPos_jitter, color = month_f, linetype = "Max", height=0.8)) +
       ggplot2::geom_errorbarh(aes(xmin=mmin-sd_min, xmax=mmin+sd_max, y=vertPos_jitter, color = month_f, linetype = "Min", height=0.8)) +
-      labs(x = dp$xlabs[[i]], y = "Height (m)") +
+      labs(x = dp$xlabs[[i]], y = "Height [m]") +
       scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits=c(0,60)) +
       theme_grey() +
       guides(linetype = guide_legend("Line type"))

### 2019-07-30 scripts/canopy_position_analysis.R
Changed TWI to be log(TWI)

@@ -1060,6 +1060,7 @@ twi_trees$TWI <- twi
 
 #6 add to trees_all
 trees_all$TWI <- twi_trees$TWI[match(trees_all$tree, twi_trees$tag)]
+trees_all$TWI.ln <- log(trees_all$TWI)
 
 ##5i1. write trees_all to csv for use in manuscript code ####
 write.csv(trees_all, "manuscript/tables_figures/trees_all.csv", row.names=FALSE)
@@ -1111,8 +1112,8 @@ library(stringr)
 ##6a. test each variable individually for each drought scenario ####
 sum_mod_traits <- data.frame(
   "prediction" = c(1.1, 2.1, 2.2, 3.3, 3.4, 4.1, 4.2, 4.3, 4.4, 4.5),
-  "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all",  "TWI", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
-  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position", "topographic.wetness.index", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
+  "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all",  "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
+  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position", "ln[topographic.wetness.index]", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
   "null_model" = 
     c("resist.value ~ (1|sp/tree)",
       "resist.value ~ year+(1|sp/tree)",
@@ -1270,11 +1271,10 @@ for (i in seq(along=sum_mod_traits[,c(8,11,14,17)])){
 }
 
 write.csv(sum_mod_traits, "manuscript/tables_figures/tested_traits_all.csv", row.names=FALSE)
+write.csv(cand_full, "manuscript/tables_figures/candidate_traits.csv", row.names=FALSE)
 
 ##6b. determine the best full model ####
 best_mod_traits <- data.frame("best_model" = NA,
-                              "coef" = NA,
-                              "r2" = NA,
[…]

### 2019-07-31 manuscript/04-results.Rmd
Updated paper

@@
 
-``` {r, biophysical-model, eval=TRUE, echo=FALSE, message=FALSE, cache=TRUE}
+``` {r, biophysical-model, eval=TRUE, echo=FALSE, message=FALSE}
 library(lme4)

### 2019-07-31 scripts/graphs_for_manuscript.R
Update graphs_for_manuscript.R

@@ -135,147 +135,131 @@ for (i in seq(along=2:ncol(leaf_traits))){
 
 #4b. create height groupings and graphs ####
 scbi18$height.m <- scbi18_ht$height.m[match(scbi18$stemID, scbi18_ht$stemID)]
-scbi18$height.ln.m <- log(scbi18$height.m)
 
-range(scbi18$height.ln.m)
-scbi18$bins <- cut(scbi18$height.m, breaks=c(-Inf,10,15,20,25,30,35,40,45,Inf), 
-   labels=c("5-10","10-15", "15-20", "20-25", "25-30", "30-35", "35-40", "40-45", "45-50"))
-scbi18$bins <- cut(scbi18$height.m, breaks=c(-Inf,7.5,10,12.5,15,17.5,20,22.5,25,27.5,30,32.5,35,37.5,40,42.5,45,47.5,51))
+
+
 
 scbi18$bins <- cut2(scbi18$height.m, g=19)
-
-# tlp ####
-trait_tlp <- scbi18 %>%
-   group_by(bins) %>%
-   dplyr::summarize(avg = mean(mean_TLP_Mpa, na.rm=TRUE),
-             dev = sd(mean_TLP_Mpa, na.rm=TRUE),
-             no_data = (sum(is.na(mean_TLP_Mpa))/length(mean_TLP_Mpa)))
-
-trait_tlp$sdmin <- trait_tlp$avg - trait_tlp$dev
-trait_tlp$sdmax <- trait_tlp$avg + trait_tlp$dev
-trait_tlp$col <- ifelse(trait_tlp$no_data > 0.25, "grey", "black")
-trait_tlp$lab <- seq(5,95,by=5)
-trait_tlp$y <- gsub("^.*,", "", trait_tlp$bins)
-trait_tlp$y <- gsub("[[:punct:]]$", "", trait_tlp$y)
-trait_tlp$y <- as.numeric(trait_tlp$y)
-
-
-#TLP by height bins
-# plot_tlp_ht <- 
-   ggplot(trait_tlp, aes(x = avg, y = lab)) +
-      geom_point(aes(color=col)) +
-      scale_color_manual(values=c("black", "grey")) +
-      ggplot2::geom_errorbarh(aes(y = lab, xmin = sdmin, xmax = sdmax, height=0.25, color=col)) +
[…]

### 2019-08-01 manuscript/02-intro.Rmd
Updated book

@@
+```{r include=FALSE, cache=FALSE}
+rm(list = ls(all = TRUE))
+```
 # Introduction
@@
 Therefore, in order to understand whole-forest response to drought, we need to know how responses vary by tree size/ species.
-To do so, there are three fundamental questions that must be addressed:
+To do so, there are four fundamental questions that must be addressed:
 
-*First, what drives the observed tendency for large trees to suffer more during drought?*  
+*First, how significant is the effect of individual drought years?*
+Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
+This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.
+
+*Second, what drives the observed tendency for large trees to suffer more during drought?*  
 [@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought.
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees.
-It is also expected that roots play a role.
-These hypotheses need to be tested.
+This study proposed three major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees; and (3) greater water availability.
+It is also expected that roots play a role, though these hypotheses still need to be tested.
 
-*Second, how do species' traits influence drought response?* 
-Analyzed drought responses on the species level doesn't get at mechanisms and is not feasible in diverse forests.
+*Third, how do species' traits influence drought response?* 
+Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
 The solution is a trait-based approach.
-Leaf hydraulic traits hold more promise than more commonly/ traditionally measured traits such as wood density and SLA (Medeiros et al.).
+Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA (Medeiros et al.).
 
-*Third, how do tree size and functional traits interact to influence drought response?*
-Its possible that pattern observed by Bennett et al.
-could be caused by smaller trees being more drought resistant.
+*Fourth, how do tree size and functional traits interact to influence drought response?*
+It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant.
 Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.
 
-*Questions & Hypotheses*
+*Hypotheses*
 
-**1. Why do larger trees suffer greater growth declines during drought?**
+**1. How significant are individual drought years on a long time-frame trend?**
+1.1 Individual drought scenario is a strong predictor of drought stress.
 
-1.0.
+    * P1.1 - Drought stress is proportional to the severity of each drought.
+
+**2. Why do larger trees suffer greater growth declines during drought?**
 Our forest displays the same trend as most forests globally [@bennett_larger_2015].
@@
 
-1.1. Height is a strong predictor of drought stress.
+2.1 DBH is a strong predictor of drought stress.
 
-    * P1.1-Drought response increases with height at time of drought.
+    * P2.2-Drought response increases with DBH at time of drought.
 
-1.2. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.
+2.2. Height is a strong predictor of drought stress.
 
-    * P1.2a- Trees currently in a dominant canopy position suffered more during drought.
+    * P2.2-Drought response increases with height at time of drought.
+    * P2.2a - Drought response is better predicted by height than DBH.
 
-1.3. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.
+2.3. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.
 
-     * P1.3a- drought response increases with elevation and/or distance from stream
-     * P1.3b- There is a dbh\*elev interaction, elevation (relative to stream?) matters less for big trees with larger (and potentially deeper) root system.
+    * P3.1- Trees currently in a dominant canopy position suffered more during drought.
 
-**2.Do species functional traits predict drought response?**
+2.4. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.
 
-     * P2.1- TLP correlates negatively with drought resistance (NEVER tested), 
-     * P2.2- diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
+     * P3.2- drought response increases with topographic wetness index
 
-**3.How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**
+**3. Do species functional traits predict drought response?**
 
-3.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
-3.1_alt. Larger trees suffer more because they have more drought vulnerable traits.
+     * P3.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
+     * P3.2 - higher percent leaf area predicts higher drought stress
+     * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [Abrams 1990] [Guerfel et al 2009]
+     * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
+     * P3.5 - higher wood density means more drought resistant
+
+**4. How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**
+
+4.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
+4.1_alt. Larger trees suffer more because they have more drought vulnerable traits.
 
@@
      * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.
+
+**I think for #4 here the focus should be on hypotheses looking at the overall trend compared to the specific scenarios?**
+
+**4. How do droughts vary in their affect on tree resistance? **
+
+    * P4.1 - The combined-years scenario will have more explaining variables than the individual scenarios.
+    * P4.2 - Inclusion of leaf hydraulic traits does not eliminate or significantly reduce effect of tree size.
+    * P4.3 - Drought resistance for each scenario will be explained more by hydraulic traits than by the biophysical environment

### 2019-08-01 manuscript/index.Rmd
Updated book

--- 2019-06-29 manuscript/index.Rmd
@@
 title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
 output:
-  html_document:
-    fig.width: 12
-    fig.height: 8
-  gitbook:
+  pdf_document:
     fig_caption: true
@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel'
 ), 'packages.bib')

### 2019-08-01 scripts/canopy_position_analysis.R
Changed prediction numbers

@@ -1111,7 +1111,7 @@ library(stringr)
 
 ##6a. test each variable individually for each drought scenario ####
 sum_mod_traits <- data.frame(
-  "prediction" = c(1.1, 2.1, 2.2, 3.3, 3.4, 4.1, 4.2, 4.3, 4.4, 4.5),
+  "prediction" = c(1.1, 2.1, 2.2, 2.3, 2.4, 3.1, 3.2, 3.3, 3.4, 3.5),
   "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all",  "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
   "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position", "ln[topographic.wetness.index]", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
   "null_model" = 

### 2019-08-01 manuscript/climate_sensitivity_draft.Rmd
Created 1 .Rmd file for the draft

---
title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [book.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  
- \usepackage{float}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel'
), 'packages.bib')
```

### Abstract
*content from AGU abstract:*
Here, we integrate forest census data from a 25.6-ha ForestGEO plot in Virginia (USA), tree-ring records for 12 species representing ##% of woody productivity, leaf hydraulic trait measurements, and microhabitat data. 
Individual-level growth responses to three droughts (1964-66, 1977, 1999) were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts. 
We conclude that when droughts occur, large dominant trees, drought sensitive species, and individuals in moist microhabitats are likely to be most strongly affected.

### Introduction

Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.

Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS). Therefore, in order to understand whole-forest response to drought, we need to know how responses vary by tree size/ species. To do so, there are four fundamental questions that must be addressed:

*First, how significant is the effect of individual drought years?*
Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same. This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.

*Second, what drives the observed tendency for large trees to suffer more during drought?*  
[@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought. However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed three major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees; and (3) greater water availability. It is also expected that roots play a role, though these hypotheses still need to be tested.

*Third, how do species' traits influence drought response?* 
Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests. The solution is a trait-based approach. Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA (Medeiros et al.).

*Fourth, how do tree size and functional traits interact to influence drought response?*
It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant. Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.

*Hypotheses*

**1. How significant are individual drought years on a long time-frame trend?**
1.1 Individual drought scenario is a strong predictor of drought stress.

    * P1.1 - Drought stress is proportional to the severity of each drought.

**2. Why do larger trees suffer greater growth declines during drought?**
Our forest displays the same trend as most forests globally [@bennett_larger_2015]. (Note that Bennett et al. 2015 identified only one study on tree growth responses to drought in the Eastern US temperate deciduous biome. We know little about how tree size shapes drought response in this biome.)

2.1 DBH is a strong predictor of drought stress.

    * P2.2-Drought response increases with DBH at time of drought.

2.2. Height is a strong predictor of drought stress.

    * P2.2-Drought response increases with height at time of drought.
    * P2.2a - Drought response is better predicted by height than DBH.

2.3. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.

    * P3.1- Trees currently in a dominant canopy position suffered more during drought.

2.4. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.

     * P3.2- drought response increases with topographic wetness index

**3. Do species functional traits predict drought response?**

     * P3.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
     * P3.2 - higher percent leaf area predicts higher drought stress
     * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [Abrams 1990] [Guerfel et al 2009]
     * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
     * P3.5 - higher wood density means more drought resistant

**4. How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**

4.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
4.1_alt. Larger trees suffer more because they have more drought vulnerable traits.

     * P3.1a- TLP is lower (larger negative) in taller/canopy trees
     * P3.1b- diffuse porous species more common in understory

     * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.

**I think for #4 here the focus should be on hypotheses looking at the overall trend compared to the specific scenarios?**

**4. How do droughts vary in their affect on tree resistance? **

    * P4.1 - The combined-years scenario will have more explaining variables than the individual scenarios.
    * P4.2 - Inclusion of leaf hydraulic traits does not eliminate or significantly reduce effect of tree size.
    * P4.3 - Drought resistance for each scenario will be explained more by hydraulic traits than by the biophysical environment

### Methods
*Study site*
Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [@andersonteixeira_ctfs-forestgeo:_2015]. SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.  Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013]. Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).

*Data collection and preparation*
The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols... [DETAILS]. Here, we use ...(*very brief summary of which data we use for what purpose-- but some will be better integrated with methods descriptions below*)

We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP_stem) (Table 1), which together comprised 97% of whole-ecosystem ANPP_stem between 2008 and 2013. Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer. In 2010-2011, cores were collected from randomly selected live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013]. In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies have been published in association with [@helcoski_growing_2019]: (GitHub URL), (ITRDB).

Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (*GitHUB URL*). Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018]. Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights. The associated errors for using either method were acknowledged [@larjavaara_measuring_2013]. Species-specific height allometries were developed (Table S# - **ADD THIS TABLE TO SI**). For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.

For each tree, we combined tree-ring records and allometric equations of height and bark thickness to retroactively calculate DBH and estimate height for the years 1950-2009. Prior DBH was estimated using the following equation, using 2008 as the earliest year for having reliable DBH measurements:

$$ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depth_YEAR) $$

Here, *ring.width* was measured from cores. Bark thickness was estimated from species-specific allometries based on the bark thickness data of [@andersonteixeira_size-related_2015].
Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#- **create table to give these equations in SI**) and then used these to estimate bark thickness based on DBH.

Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed. As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years. While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig. XX) and height change since *1959* indicated that change was likely slow. [**work on this-- provide details?**]

Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.

Hydraulic traits were collected from SCBI and are summarized in Table 1. In August 2018, we collected leaf samples from three individuals of each species ... (**Nobby's description of methods for the following**) 
1. PLA
2. LMA
4. Wood density
5. TLP

**Table 1. Species analyzed here, listed in descending order of ANPP_stem. n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section. Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]

```{r, Table-cores-position, eval=TRUE, echo=FALSE, message=FALSE}
library(RCurl)
library(dplyr)
library(tidyr)
library(data.table)
library(knitr)
library(kableExtra)

species <- read.csv("D:/Dropbox (Smithsonian)/Github_Ian/McGregor_climate-sensitivity-variation/data/core_list_for_neil.csv", stringsAsFactors = FALSE)

table <- species[, c(1:3,6,17)]

positions <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_dimensions/tree_crowns/cored_dendroband_crown_position_data/dendro_cored_full.csv"))

table$position_all <- positions$crown.position[match(table$tag, positions$tag)]
table$position_all <- gsub("D", "dominant", table$position_all)
table$position_all <- gsub("C", "co-dominant", table$position_all)
table$position_all <- gsub("I", "intermediate", table$position_all)
table$position_all <- gsub("S", "suppressed", table$position_all)

table1 <- table %>%
  group_by(sp) %>%
  summarize(n_cores=n())

table2 <- table %>%
  group_by(sp, position_all) %>%
  summarize(count=n())

table2 <- spread(table2, key=position_all, value=count)
setnames(table2, old="<NA>", new="prior dead")

table2$n_cores <- table1$n_cores[match(table2$sp, table1$sp)]
table2 <- table2[, c(1,7,3,2,4:6)]
table2 <- table2[!table2$sp %in% c("frni", "pist"), ]

kable(table2) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```
```{r, traits-overview, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

trees_all <- read.csv("tables_figures/trees_all.csv", stringsAsFactors = FALSE)

trees_all <- trees_all[!duplicated(trees_all$sp), ]

traits_table <- data.frame(
  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Wood density", "TLP"),
  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "g/cm3", "MPa"))
traits_table$abb <- c("rp", "PLA", "LMA", "WD", "TLP")

traits_table$mean <- NA
traits_table$min <- NA
traits_table$max <- NA

for (i in seq(along=traits_table$abb[2:5])){
  trait <- traits_table$abb[2:5][i]

  sub <- trees_all[grepl(trait, colnames(trees_all))]
  traits_table$mean[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                              mean(sub[, 1], na.rm=TRUE), 
                              traits_table$mean)

  traits_table$min[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                              min(sub[, 1], na.rm=TRUE), 
                              traits_table$min)

  traits_table$max[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                              max(sub[, 1], na.rm=TRUE), 
                              traits_table$max)
}

traits_table[, c("mean", "min", "max")] <- 
  sapply(traits_table[, c("mean", "min", "max")], function(x)
  round(x, 2)
)
traits_table$abb <- NULL

kable(traits_table) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

```

*Climate and drought years*
[**add description of climate data used in Fig. 1, NEON vertical profiles**]

To accurately understand climate sensitivity, this study used a specific definition of drought, which is not a common practice [@slette_how_2019]. We used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined by [@lloret_components_2011] as the ratio between the performance during and before the disturbance. Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years. These were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values for each year, which yielded a set of three periods that were consistently shown as drought: 1964-1966, 1977, and 1999.

**Figure 1. Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from [@helcoski_growing_2019].
![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}

### Results
Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

```{r, Table 1, eval=TRUE, echo=FALSE, message=FALSE}
traits_test <- read.csv("tables_figures/tested_traits_all.csv")
kable(traits_test, booktabs=TRUE, format = "latex", caption = "All variables tested for climate sensitivity against appropriate null models, in line with hypothesis predictions. Each variable was tested for each drought scenario excepting \"year\" ") %>%
  kableExtra::landscape()
  # kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "scale_down")

```

```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
candidates <- read.csv("tables_figures/candidate_traits.csv")
kable(candidates, booktabs=TRUE, format = "latex", caption = "Candidate variables to be included in full model, chosen by dAICc>2 when individually tested against a null model") %>%
  kable_styling(bootstrap_options = "striped", latex_options = c("scale_down", "hold_position"))
```

```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
traits_best <- read.csv("tables_figures/tested_traits_best.csv")
kable(traits_best, booktabs=TRUE, format = "latex", caption = "Best full models for each drought scenario") %>%
  kable_styling(bootstrap_options = "striped", latex_options = c("scale_down", "hold_position"))
```

### Discussion

*1. paragraph summarizing main results--> primary conclusions*
When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought. This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses. Adding in crown position with the leaf hydraulic traits yields a slightly worse predictive model for drought tolerance, with height remaining as the only significant biophysical variable.

We partially supported the hypothesis that crown exposure makes trees more vulnerable to drought. Co-dominant trees had the highest drought resistance. Dominant trees had lower resistance, likely because they are the most exposed. Other studies have found clear evidence of greater drought sensitivity in trees with exposed crowns (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]). At the same time, intermediate and suppressed trees had even lower resistance. This indicates that other mechanisms such as competition or rooting depth were important. (Also note that our study design was not ideal for testing the role of canopy position. Current canopy position is a conservative separator of canopy position: trees may currently be in more dominant positions than they were at the time, but backwards movement is unlikely. This would bias against finding a signficant effect for H1.2. Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)

Proximity to stream--either vertical (elev) or horizontal (distance)-- did not increase drought resistance; rather, it tended to decrease resistance (H1.3a). This may be because individuals growing further from water are acclimatized to drier conditions. However, the increase in drought resistance with distance from stream was less for small than large trees (H1.3b), indicating a potential importance of root depth/volume in conferring drought resistance.

**misc content to integrate**
From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity. This paper mentions it would be good to have this data with respect to latent affects from drought.
### Conclusion
words
### Supplementary Information
```{r echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='H'}
  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
```

*p50 and p80*
We decided to include values of P50 and P80 in the leaf traits model, defined by [@anderegg_meta-analysis_2016] as the water potentials at which a species loses 50% and 88% [80% by proxy], respectively, of hydraulic conductivity. Values were calculated by (**insert new methods here??**), and were only available for six species (*C. glabra*, *L. tulipifera*, *Q. alba*, *Q. prinus*, *Q. rubra*, and *Q. velutina*). Because of this, the model runs were considered to be incomplete due to the exclusion of the other 8 species. Results revealed neither p50 nor p80 to be significant, thus for the full analysis we decided to drop the two traits in order to include all species in the full analysis.

(see Issue #32)

**include TWI by trait graphs (2) here**

### 2019-08-01 manuscript/climate_sensitivity_draft.tex
Updated paper

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest},
            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Tree size, exposure, and hydraulic traits interactively shape drought
response in a temperate broadleaf forest}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine
Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack,
Krista Anderson-Teixeira}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{float}
\usepackage{booktabs}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{caption}
\captionsetup[table]{font=small}
\captionsetup[figure]{font=small}
\usepackage{dcolumn}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\subsubsection{Abstract}\label{abstract}}

\emph{content from AGU abstract:} Here, we integrate forest census data
from a 25.6-ha ForestGEO plot in Virginia (USA), tree-ring records for
12 species representing \#\#\% of woody productivity, leaf hydraulic
trait measurements, and microhabitat data. Individual-level growth
responses to three droughts (1964-66, 1977, 1999) were stronger among
taller trees in dominant canopy positions, those in wetter microsites,
and for more drought-sensitive species as assessed by leaf traits
(turgor loss at less negative leaf water potential, greater shrinkage
with leaf dehydration), with substantial variation in the best predictor
variables across given droughts. We conclude that when droughts occur,
large dominant trees, drought sensitive species, and individuals in
moist microhabitats are likely to be most strongly affected.

\hypertarget{introduction}{%
\subsubsection{Introduction}\label{introduction}}

Understanding how and why trees respond to drought is critical to
predicting forest drought responses and climate change feedbacks.

Forests are diverse in terms of tree sizes and functional traits, and it
is known that trees varying in size and functional traits respond
differently to drought (e.g., \citep{bennett_larger_2015}; REFS).
Therefore, in order to understand whole-forest response to drought, we
need to know how responses vary by tree size/ species. To do so, there
are four fundamental questions that must be addressed:

\emph{First, how significant is the effect of individual drought years?}
Droughts are rarely explicitly defined in ecological studies
\citep{slette_how_2019}, yet no two droughts are the same. This study
addressed this by analyzing trees' resistance to drought within and
across three defined drought periods.

\emph{Second, what drives the observed tendency for large trees to
suffer more during drought?}\\
\citep{bennett_larger_2015} showed that in forests globally, large trees
suffer greater growth reductions during drought. However, this analysis
quantified tree size based on DBH, which has no direct mechanistic
meaning. This study proposed three major mechanisms (besides insects):
(1) inherently greater biophysical challenge of being tall; (2) greater
exposure of the crowns of large trees; and (3) greater water
availability. It is also expected that roots play a role, though these
hypotheses still need to be tested.

\emph{Third, how do species' traits influence drought response?}
Analyzing drought responses on the species level does not fully explain
mechanisms and is not feasible in diverse forests. The solution is a
trait-based approach. Leaf hydraulic traits hold more promise than more
commonly/traditionally-measured traits such as wood density and SLA
(Medeiros et al.).

\emph{Fourth, how do tree size and functional traits interact to
influence drought response?} It is possible that the pattern observed by
\citep{bennett_larger_2015} could be caused by smaller trees being more
drought resistant. Alternatively, larger trees may have more
drought-resistant traits as adpatations to greater biophysical
challenges.

\emph{Hypotheses}

\textbf{1. How significant are individual drought years on a long
time-frame trend?} 1.1 Individual drought scenario is a strong predictor
of drought stress.

\begin{verbatim}
* P1.1 - Drought stress is proportional to the severity of each drought.
\end{verbatim}

\textbf{2. Why do larger trees suffer greater growth declines during
drought?} Our forest displays the same trend as most forests globally
\citep{bennett_larger_2015}. (Note that Bennett et al.~2015 identified
only one study on tree growth responses to drought in the Eastern US
temperate deciduous biome. We know little about how tree size shapes
drought response in this biome.)

2.1 DBH is a strong predictor of drought stress.

\begin{verbatim}
* P2.2-Drought response increases with DBH at time of drought.
\end{verbatim}

2.2. Height is a strong predictor of drought stress.

\begin{verbatim}
* P2.2-Drought response increases with height at time of drought.
* P2.2a - Drought response is better predicted by height than DBH.
\end{verbatim}

2.3. Large trees suffer more during drought because of greater exposure
(to radiation, wind, etc.)--either in relation to neighboring trees or
because of position on landscape.

\begin{verbatim}
* P3.1- Trees currently in a dominant canopy position suffered more during drought. 
\end{verbatim}

2.4. Rooting volume/depth relative to water sources are critical in
drought response. Effects of drought on larger trees are mediated by the
fact that large trees have better access to water.

\begin{verbatim}
 * P3.2- drought response increases with topographic wetness index
\end{verbatim}

\textbf{3. Do species functional traits predict drought response?}

\begin{verbatim}
 * P3.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
 * P3.2 - higher percent leaf area predicts higher drought stress
 * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [Abrams 1990] [Guerfel et al 2009]
 * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
 * P3.5 - higher wood density means more drought resistant
\end{verbatim}

\textbf{4. How are functional traits distributed across size classes,
and how does this affect size-resistance relationship?}

4.1. Larger/ more exposed trees have more drought resistant traits,
meaning size effects are buffered by traits\\
4.1\_alt. Larger trees suffer more because they have more drought
vulnerable traits.

\begin{verbatim}
 * P3.1a- TLP is lower (larger negative) in taller/canopy trees
 * P3.1b- diffuse porous species more common in understory

 * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.
\end{verbatim}

\textbf{I think for \#4 here the focus should be on hypotheses looking
at the overall trend compared to the specific scenarios?}

\textbf{4. How do droughts vary in their affect on tree resistance? }

\begin{verbatim}
* P4.1 - The combined-years scenario will have more explaining variables than the individual scenarios.
* P4.2 - Inclusion of leaf hydraulic traits does not eliminate or significantly reduce effect of tree size.
* P4.3 - Drought resistance for each scenario will be explained more by hydraulic traits than by the biophysical environment
\end{verbatim}

\hypertarget{methods}{%
\subsubsection{Methods}\label{methods}}

\emph{Study site} Research was conducted at the 25.6 ha ForestGEO
(Global Earth Observatory) study plot at the Smithsonian Conservation
Biology Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°
08'43.4''W) \citep{andersonteixeira_ctfs-forestgeo:_2015}. SCBI is
located in the central Appalachian Mountains at the northern edge of
Shenandoah National Park. Elevations range from 273-338m above sea level
\citep{gonzalezakre_patterns_2016} with a topographic relief of 65m
\citep{bourg_initial_2013}. Dominant species include \emph{Liriodendron
tulipifera}, oaks (\emph{Quercus} spp.), and hickories (\emph{Carya}
spp.).

\emph{Data collection and preparation} The SCBI ForestGEO plot was
censused in 2008, 2013, and 2018 following standard ForestGEO
protocols\ldots{} {[}DETAILS{]}. Here, we use \ldots(\emph{very brief
summary of which data we use for what purpose-- but some will be better
integrated with methods descriptions below})

We analyzed tree-ring data from the twelve species contributing most to
woody aboveground net primary productivity (ANPP\_stem) (Table 1), which
together comprised 97\% of whole-ecosystem ANPP\_stem between 2008 and
2013. Cores were collected in 2010-2011 or 2016-2017 from a breast
height of 1.3m using a 5mm increment borer. In 2010-2011, cores were
collected from randomly selected live trees were selected at random from
each species in 2010-2011, with at least 30 of those trees having a
diameter at breast height (DBH) of at least 10cm
\citep{bourg_initial_2013}. In 2016-2017, cores were collected from all
dead trees found in the annual mortality census
\citep{gonzalezakre_patterns_2016}. Cores were sanded, measured, and
cross-dated using standard procedures, as detailed in
\citep{helcoski_growing_2019}. The resulting chronologies have been
published in association with \citep{helcoski_growing_2019}: (GitHub
URL), (ITRDB).

Height measurements (n=\# trees) were taken by several researchers
between 2012 to 2019, and are archived in a public GitHub repository
(\emph{GitHUB URL}). Measurement methods included manual
\citep[NEON]{stovall_assessing_2018}, digital rangefinders
\citep{andersonteixeira_size-related_2015}, and automatic LiDAR
\citep{stovall_terrestrial_2018}. Rangefinders either used the tangent
method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon
ForestryPro) for calculating heights. The associated errors for using
either method were acknowledged \citep{larjavaara_measuring_2013}.
Species-specific height allometries were developed (Table S\# -
\textbf{ADD THIS TABLE TO SI}). For species that didn't have enough
height measurements, heights were calculated from equations derived from
all species in the study.

For each tree, we combined tree-ring records and allometric equations of
height and bark thickness to retroactively calculate DBH and estimate
height for the years 1950-2009. Prior DBH was estimated using the
following equation, using 2008 as the earliest year for having reliable
DBH measurements:

\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depth_YEAR) \]

Here, \emph{ring.width} was measured from cores. Bark thickness was
estimated from species-specific allometries based on the bark thickness
data of \citep{andersonteixeira_size-related_2015}. Specifically, we
used linear regression equations on log-transformed data to relate bark
thickness to DBH (Table S\#- \textbf{create table to give these
equations in SI}) and then used these to estimate bark thickness based
on DBH.

Crown positions were recorded in the field during the growing season of
2018 following the crown position protocol from
\citep{jennings_assessing_1999}, whereby positions were ranked as
dominant, codominant, intermediate, or suppressed. As there was no way
to retroactively estimate crown position, we assumed that 2018 crown
position was reflective of each tree's position over the past 60 years.
While some trees undoubtedly changed position, an analysis of crown
position relative to height (Fig. XX) and height change since
\emph{1959} indicated that change was likely slow. {[}\textbf{work on
this-- provide details?}{]}

Topographic wetness index (TWI) was calculated using the
\citep{R-dynatopmodel} package in R.

Hydraulic traits were collected from SCBI and are summarized in Table 1.
In August 2018, we collected leaf samples from three individuals of each
species \ldots{} (\textbf{Nobby's description of methods for the
following}) 1. PLA 2. LMA 4. Wood density 5. TLP

\textbf{Table 1. Species analyzed here, listed in descending order of
ANPP\_stem. n cores and DBH range represented, and species traits}
{[}*This replaces/combines the two remaining tables in this section.
Suggested columns, with those to include only if they fit in
parentheses: species, (stems \textgreater=10 cm per ha in plot),
(ANPP\_stem), n cores, DBH range of cores, (n cores in each crown
position) species means for each trait{]}

\begin{table}[H]
\centering
\begin{tabular}{l|r|r|r|r|r|r}
\hline
sp & n\_cores & dominant & co-dominant & intermediate & suppressed & prior dead\\
\hline
caco & 13 & NA & 2 & 5 & 5 & 1\\
\hline
cagl & 31 & 1 & 8 & 16 & 5 & 1\\
\hline
caovl & 23 & 4 & 5 & 12 & 2 & NA\\
\hline
cato & 13 & NA & NA & 6 & 2 & 5\\
\hline
fagr & 80 & NA & 7 & 48 & 25 & NA\\
\hline
fram & 62 & NA & 17 & 19 & 14 & 12\\
\hline
juni & 31 & NA & 21 & 8 & NA & 2\\
\hline
litu & 98 & 9 & 29 & 25 & 30 & 5\\
\hline
qual & 61 & 4 & 34 & 20 & 3 & NA\\
\hline
qupr & 59 & 1 & 26 & 20 & 12 & NA\\
\hline
quru & 69 & 6 & 36 & 23 & 2 & 2\\
\hline
quve & 77 & 6 & 46 & 22 & 1 & 2\\
\hline
\end{tabular}
\end{table}
\begin{table}[H]
\centering
\begin{tabular}{l|l|r|r|r}
\hline
Trait & Unit & mean & min & max\\
\hline
Ring Porosity & ring, semi-ring, diffuse & NA & NA & NA\\
\hline
Percent Leaf Area & \% & 15.09 & 8.52 & 24.64\\
\hline
Leaf Mass Area & g/m2 & 53.50 & 30.68 & 75.80\\
\hline
Wood density & g/cm3 & 0.70 & 0.40 & 1.09\\
\hline
TLP & MPa & -2.36 & -2.76 & -1.92\\
\hline
\end{tabular}
\end{table}

\emph{Climate and drought years} {[}\textbf{add description of climate
data used in Fig. 1, NEON vertical profiles}{]}

To accurately understand climate sensitivity, this study used a specific
definition of drought, which is not a common practice
\citep{slette_how_2019}. We used the pointRes package \citep{R-pointRes}
in R (version 3.5.3) to determine drought periods based on trees'
drought resistance, which is defined by \citep{lloret_components_2011}
as the ratio between the performance during and before the disturbance.
Candidate drought years were defined if \textgreater50\% of the cored
trees experienced \textless30\% growth in a year compared to the
previous 5 years. These were then cross-validated with the regional
Palmer Drought Severity Index (PDSI) values for each year, which yielded
a set of three periods that were consistently shown as drought:
1964-1966, 1977, and 1999.

\textbf{Figure 1. Time series of peak growing season (May-August)
climate conditions and residual chronologies for each species.} Droughts
analyzed here are indicated by dashed lines, and shading indicates the
pre-drought period used in calculations of the resistance metric. Figure
modified from \citep{helcoski_growing_2019}.
\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Time_series_for_each_species.jpg}

\hypertarget{results}{%
\subsubsection{Results}\label{results}}

Once the data was collected, linear mixed models were run following the
order of the hypotheses as seen in Figure ???
{[}individual\_tested\_traits{]}. Using the \citep{R-pointRes} package,
we set up models with the resistance value as the response variable, and
each prediction's variable as the independent variable. Variables'
importance in predicting drought tolerance was calculated from
mixed-effects models and the lowest AICc
\citep[\citet{R-AICcmodavg}]{R-lme4}.Null models were determined in
order of the predictions. First, we analyzed the combined scenario to
determine if ``year'' was significant. Upon establishing this, we tested
height and DBH as size parameters. Although both were significant,
height was kept due to its larger delta AICc compared with the null
model. We then tested the remaining biophysical and hydraulic traits
individually against a null model containing height and year. This
yielded Figure ??? (cand\_full). All variables with dAICc \textgreater2
were used as candidates for each scenario's best model (figure ????
(tested\_traits\_best))

\begin{landscape}\begin{table}[t]

\caption{\label{tab:Table 1}All variables tested for climate sensitivity against appropriate null models, in line with hypothesis predictions. Each variable was tested for each drought scenario excepting "year" }
\centering
\begin{tabular}{rllllllrllrllrllrll}
\toprule
prediction & variable & variable\_description & null\_model & tested\_model & null\_model\_year & tested\_model\_year & dAIC\_all & coef\_all & coef\_var\_all & dAIC\_1964.1966 & coef\_1964.1966 & coef\_var\_1964.1966 & dAIC\_1977 & coef\_1977 & coef\_var\_1977 & dAIC\_1999 & coef\_1999 & coef\_var\_1999\\
\midrule
1.1 & year & drought.year & resist.value \textasciitilde{} (1|sp/tree) & resist.value \textasciitilde{} (1|sp/tree)+year & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp) & 35.834 & - & year1977 (-0.09), year1999 (-0.081) & 0.000 & NA & NA & 0.000 & NA & NA & 0.000 & NA & NA\\
2.1 & dbh.ln.cm & ln[DBH] & resist.value \textasciitilde{} year+(1|sp/tree) & resist.value \textasciitilde{} year+(1|sp/tree)+dbh.ln.cm & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp)+dbh.ln.cm & 7.411 & - & dbh.ln.cm (-0.035) & 18.276 & - & dbh.ln.cm (-0.082) & -0.947 & - & dbh.ln.cm (-0.021) & -1.918 & + & dbh.ln.cm (0.006)\\
2.2 & height.ln.m & ln[height] & resist.value \textasciitilde{} year+(1|sp/tree) & resist.value \textasciitilde{} year+(1|sp/tree)+height.ln.m & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp)+height.ln.m & 7.591 & - & height.ln.m (-0.058) & 17.788 & - & height.ln.m (-0.133) & -1.077 & - & height.ln.m (-0.032) & -2.022 & + & height.ln.m (0.002)\\
3.3 & position\_all & crown.position & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+position\_all & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+position\_all & 1.682 & - & position\_alldominant (-0.044), position\_allintermediate (-0.042), position\_allsuppressed (-0.053) & -2.548 & - & position\_alldominant (-0.055), position\_allintermediate (0.007), position\_allsuppressed (-0.044) & -0.700 & - & position\_alldominant (-0.076), position\_allintermediate (-0.033), position\_allsuppressed (0.029) & 4.087 & - & position\_alldominant (-0.003), position\_allintermediate (-0.082), position\_allsuppressed (-0.102)\\
3.4 & TWI & topographic.wetness.index & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+TWI & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+TWI & 2.631 & - & TWI (-0.009) & -0.563 & + & TWI (0.009) & 5.264 & - & TWI (-0.02) & 3.037 & - & TWI (-0.015)\\
\addlinespace
4.1 & rp & ring.porosity & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+rp & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+rp & -3.553 & + & rpring (0.04), rpsemi-ring (0.013) & -2.161 & + & rpring (0.101), rpsemi-ring (0.015) & 0.895 & - & rpring (-0.19), rpsemi-ring (-0.147) & 4.083 & + & rpring (0.2), rpsemi-ring (0.151)\\
4.2 & PLA\_dry\_percent & percent.leaf.area & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+PLA\_dry\_percent & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+PLA\_dry\_percent & 4.413 & - & PLA\_dry\_percent (-0.011) & 5.825 & - & PLA\_dry\_percent (-0.016) & -0.190 & - & PLA\_dry\_percent (-0.01) & -0.701 & - & PLA\_dry\_percent (-0.007)\\
4.3 & LMA\_g\_per\_m2 & leaf.mass.area & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+LMA\_g\_per\_m2 & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+LMA\_g\_per\_m2 & -1.895 & + & LMA\_g\_per\_m2 (0.001) & -1.075 & + & LMA\_g\_per\_m2 (0.002) & -1.698 & - & LMA\_g\_per\_m2 (-0.001) & -1.985 & + & LMA\_g\_per\_m2 (0)\\
4.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+mean\_TLP\_Mpa & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+mean\_TLP\_Mpa & 4.580 & - & mean\_TLP\_Mpa (-0.207) & 1.352 & - & mean\_TLP\_Mpa (-0.217) & 1.008 & - & mean\_TLP\_Mpa (-0.236) & 0.132 & - & mean\_TLP\_Mpa (-0.177)\\
4.5 & WD\_g\_per\_cm3 & wood.density & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+WD\_g\_per\_cm3 & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+WD\_g\_per\_cm3 & -2.018 & + & WD\_g\_per\_cm3 (0.005) & -1.960 & - & WD\_g\_per\_cm3 (-0.049) & -1.236 & - & WD\_g\_per\_cm3 (-0.175) & 0.171 & + & WD\_g\_per\_cm3 (0.247)\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(stargazer)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Please cite as:
\end{verbatim}

\begin{verbatim}
##  Hlavac, Marek (2018). stargazer: Well-Formatted Regression and Summary Statistics Tables.
\end{verbatim}

\begin{verbatim}
##  R package version 5.2.2. https://CRAN.R-project.org/package=stargazer
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{stargazer}\NormalTok{(traits_test, }
          \DataTypeTok{float.env =} \StringTok{"sidewaystable"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard
University. E-mail: hlavac at fas.harvard.edu \% Date and time: Thu, Aug
01, 2019 - 12:30:35 PM \% Requires LaTeX packages: rotating

\begin{sidewaystable}[!htbp] \centering 
  \caption{} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
prediction & 10 & 3.360 & 1.182 & 1.100 & 2.475 & 4.275 & 4.500 \\ 
dAIC\_all & 10 & 5.668 & 11.284 & $-$3.553 & $-$1.001 & 6.703 & 35.834 \\ 
dAIC\_1964.1966 & 10 & 3.493 & 8.032 & $-$3 & $-$1.7 & 4.7 & 18 \\ 
dAIC\_1977 & 10 & 0.132 & 2.010 & $-$2 & $-$1.0 & 0.7 & 5 \\ 
dAIC\_1999 & 10 & 0.488 & 2.414 & $-$2 & $-$1.6 & 2.3 & 4 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{sidewaystable}

\begin{table}[!h]

\caption{\label{tab:Table 2}Candidate variables to be included in full model, chosen by dAICc>2 when individually tested against a null model}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{rlll}
\toprule
prediction & variable & variable\_description & top\_model\\
\midrule
1.1 & year & drought.year & all\\
2.2 & height.ln.m & ln[height] & all\\
2.2 & height.ln.m & ln[height] & 1966\\
3.3 & position\_all & crown.position & 1999\\
3.4 & TWI.ln & topographic.wetness.index & all\\
\addlinespace
3.4 & TWI.ln & topographic.wetness.index & 1977\\
3.4 & TWI.ln & topographic.wetness.index & 1999\\
4.1 & rp & ring.porosity & 1999\\
4.2 & PLA\_dry\_percent & percent.leaf.area & all\\
4.2 & PLA\_dry\_percent & percent.leaf.area & 1966\\
\addlinespace
4.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & all\\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[!h]

\caption{\label{tab:Table 3}Best full models for each drought scenario}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{lrl}
\toprule
best\_model & r2 & scenario\\
\midrule
resist.value \textasciitilde{} year+height.ln.m+position\_all+TWI+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.13 & all droughts\\
resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.24 & 1964-1966\\
resist.value \textasciitilde{} TWI+rp+mean\_TLP\_Mpa+(1|sp) & 0.21 & 1977\\
resist.value \textasciitilde{} height.ln.m+position\_all+TWI+rp+PLA\_dry\_percent+(1|sp) & 0.25 & 1999\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{discussion}{%
\subsubsection{Discussion}\label{discussion}}

\emph{1. paragraph summarizing main results--\textgreater{} primary
conclusions} When including only biophysical traits, trees' resistance
value (on a per-species basis) is explained best by crown position and
height, with codominant trees being the most resistant to drought. This
follows on work done by \citep{bennett_larger_2015} {[}and others?{]}
which show that larger trees suffer more during drought, and confirms
that this susceptibility can be seen in tree ring analyses. Adding in
crown position with the leaf hydraulic traits yields a slightly worse
predictive model for drought tolerance, with height remaining as the
only significant biophysical variable.

We partially supported the hypothesis that crown exposure makes trees
more vulnerable to drought. Co-dominant trees had the highest drought
resistance. Dominant trees had lower resistance, likely because they are
the most exposed. Other studies have found clear evidence of greater
drought sensitivity in trees with exposed crowns (e.g.,
\citep{suarez_factors_2004}; \citep{scharnweber_confessions_2019}). At
the same time, intermediate and suppressed trees had even lower
resistance. This indicates that other mechanisms such as competition or
rooting depth were important. (Also note that our study design was not
ideal for testing the role of canopy position. Current canopy position
is a conservative separator of canopy position: trees may currently be
in more dominant positions than they were at the time, but backwards
movement is unlikely. This would bias against finding a signficant
effect for H1.2. Height may be a more reliable predictor of past canopy
position than is current canopy position, and explains a portion of
variation in canopy position.)

Proximity to stream--either vertical (elev) or horizontal (distance)--
did not increase drought resistance; rather, it tended to decrease
resistance (H1.3a). This may be because individuals growing further from
water are acclimatized to drier conditions. However, the increase in
drought resistance with distance from stream was less for small than
large trees (H1.3b), indicating a potential importance of root
depth/volume in conferring drought resistance.

\textbf{misc content to integrate} From \citep{kannenberg_linking_2019},
species with diffuse porous wood anatomy (\emph{Liriodendron}) are more
sensitive to drought, whereas ring-porous are not as affected because
they more easily rebuild structures for hydraulic conductivity. This
paper mentions it would be good to have this data with respect to latent
affects from drought. \#\#\# Conclusion words \#\#\# Supplementary
Information

\begin{figure}[H]
\includegraphics[width=1\linewidth]{tables_figures/ForestGEO_plot} \caption{Map of ForestGEO plot}\label{fig:unnamed-chunk-4}
\end{figure}

\emph{p50 and p80} We decided to include values of P50 and P80 in the
leaf traits model, defined by \citep{anderegg_meta-analysis_2016} as the
water potentials at which a species loses 50\% and 88\% {[}80\% by
proxy{]}, respectively, of hydraulic conductivity. Values were
calculated by (\textbf{insert new methods here??}), and were only
available for six species (\emph{C. glabra}, \emph{L. tulipifera},
\emph{Q. alba}, \emph{Q. prinus}, \emph{Q. rubra}, and \emph{Q.
velutina}). Because of this, the model runs were considered to be
incomplete due to the exclusion of the other 8 species. Results revealed
neither p50 nor p80 to be significant, thus for the full analysis we
decided to drop the two traits in order to include all species in the
full analysis.

(see Issue \#32)

\textbf{include TWI by trait graphs (2) here}

\bibliography{book.bib,packages.bib}

\end{document}

### 2019-08-01 manuscript/climate_sensitivity_draft.Rmd
Re-added pdf

@@
 - \usepackage{float}
+- \usepackage{booktabs}
 - \usepackage{pdflscape}
@@
 - \captionsetup[figure]{font=small} 
+- \usepackage{dcolumn} #for sidewaystable
 ---

### 2019-08-01 manuscript/climate_sensitivity_draft.Rmd
Updated paper

@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
 ), 'packages.bib')
@@
 ### Abstract
-*content from AGU abstract:*
-Here, we integrate forest census data from a 25.6-ha ForestGEO plot in Virginia (USA), tree-ring records for 12 species representing ##% of woody productivity, leaf hydraulic trait measurements, and microhabitat data.
-Individual-level growth responses to three droughts (1964-66, 1977, 1999) were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.
-We conclude that when droughts occur, large dominant trees, drought sensitive species, and individuals in moist microhabitats are likely to be most strongly affected.
+Predicting forest responses to drought is an increasingly critical task under climate change effects.
+Part of the problem is due the lack of a set definition of drought in recent studies, while often the effects of biophysical and hydraulic traits are interpreted separately.
+In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
+Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
+Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested again leaf hyraulic trait measurements and microhabitat biophysical data.
+Individual-level growth responses to three droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.
+We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.
 
@@
 Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.
-
 Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
@@
     * P3.2 - higher percent leaf area predicts higher drought stress
-     * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [Abrams 1990] [Guerfel et al 2009]
+    * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [@abrams_adaptations_1990] [@guerfel_impacts_2009] <- *valid?
+it looks at olive trees*
     * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
@@
 *Data collection and preparation*
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols...
-[DETAILS].
-Here, we use ...(*very brief summary of which data we use for what purpose-- but some will be better integrated with methods descriptions below*)
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems >= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+From this census data, we used measurements of DBH from 2008 to calculate historical DBH (described below), along with tree location in the plot to determine the topographic wetness index.
+*I think that's it, really, in terms of data we're using coming from the censuses.*
 
-We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP_stem) (Table 1), which together comprised 97% of whole-ecosystem ANPP_stem between 2008 and 2013.
+We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity, which together comprised 97% of whole-ecosystem between 2008 and 2013 [@helcoski_growing_2019].
 Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
@@
 Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been published in association with [@helcoski_growing_2019]: (GitHub URL), (ITRDB).
+The resulting chronologies have been [published](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) in association with [@helcoski_growing_2019]: (ITRDB).
+*is this a database?*
 
-Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (*GitHUB URL*).
+Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
 Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
@@
 
-$$ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depth_YEAR) $$
+$$ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) $$
 
@@
 kable(candidates, booktabs=TRUE, format = "latex", caption = "Candidate variables to be included in full model, chosen by dAICc>2 when individually tested against a null model") %>%
-  kable_styling(bootstrap_options = "striped", latex_options = c("scale_down", "hold_position"))
+  kable_styling(bootstrap_options = "striped", latex_options ="hold_position")
 ```
@@
 ### Conclusion
+
 words
+
 ### Supplementary Information
+
 ```{r echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='H'}

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
move comparison of droughts to be the last question/hypothesis, merge drought strength and response questions.

@@
 
-*First, how significant is the effect of individual drought years?*
-Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
-This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.
-
-*Second, what drives the observed tendency for large trees to suffer more during drought?*  
+*First, what drives the observed tendency for large trees to suffer more during drought?*  
 [@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought.
@@
 
-*Third, how do species' traits influence drought response?* 
+*Second, how do species' traits influence drought response?* 
 Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
@@
 
-*Fourth, how do tree size and functional traits interact to influence drought response?*
+*Third, how do tree size and functional traits interact to influence drought response?*
 It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant.
@@
 
+*Fourth, how significant is the effect of individual drought years?*
+Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
+This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.
+
 *Hypotheses*
 
-**1. How significant are individual drought years on a long time-frame trend?**
-1.1 Individual drought scenario is a strong predictor of drought stress.
-
-    * P1.1 - Drought stress is proportional to the severity of each drought.
-
-**2. Why do larger trees suffer greater growth declines during drought?**
+**1. Why do larger trees suffer greater growth declines during drought?**
 Our forest displays the same trend as most forests globally [@bennett_larger_2015].
@@
 
+[differences among droughts:]
 **I think for #4 here the focus should be on hypotheses looking at the overall trend compared to the specific scenarios?**
+
+**1. How significant are individual drought years on a long time-frame trend?**
+1.1 Individual drought scenario is a strong predictor of drought stress.
+
+    * P1.1 - Drought stress is proportional to the severity of each drought.
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
changed title
(not sure we have it right yet)

@@
 ---
-title: "Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest"
+title: "Tree size, microhabitat, and hydraulic traits (interactively) shape drought response in a temperate broadleaf forest"
 author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
working on questions/ hypotheses/ predictions.

@@
 Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.
-Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
-Therefore, in order to understand whole-forest response to drought, we need to know how responses vary by tree size/ species.
-To do so, there are four fundamental questions that must be addressed:
+(1 paragraph on this- I have lots of content on this and will add later.)
+
+[Understanding forest responses to drought requires increased functional understanding of the factors that confir individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
+Therefore, in order to understand whole-forest response to drought, we need a functional understanding of how responses vary by tree size, microhabitat, and species.
+There are 3 fundamental questions that must be addressed:
 
@@
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed three major mechanisms (besides insects): (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees; and (3) greater water availability.
-It is also expected that roots play a role, though these hypotheses still need to be tested.
+This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees.
+Counteracting these effects, (3) the larger root systems of larger trees may confir an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
 
-*Second, how do species' traits influence drought response?* 
+*Second, how do species' traits--alone and in interaction with tree size--influence drought response?* 
 Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
 The solution is a trait-based approach.
-Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA (Medeiros et al.).
+Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA ([Medeiros et al.
+2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
 
-*Third, how do tree size and functional traits interact to influence drought response?*
+Moreover, there may be an interaction between traits and tree size.
 It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant.
@@
 
-*Fourth, how significant is the effect of individual drought years?*
+*Third, are responses similar or variable across individual drought years?*
 Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
@@
 **1. Why do larger trees suffer greater growth declines during drought?**
-Our forest displays the same trend as most forests globally [@bennett_larger_2015].
+
+1.0.
+Consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance.
 (Note that Bennett et al.
@@
 
-2.1 DBH is a strong predictor of drought stress.
+    * P1.0-Drought resistance decreases with DBH at time of drought. (supported most of the time)
 
-    * P2.2-Drought response increases with DBH at time of drought.
+1.1. Drought resistance decreases with tree height.
 
-2.2. Height is a strong predictor of drought stress.
+    * P1.1- Drought resistance decreases with height at time of drought, and height is a better predictor than DBH. (supported most of the time)
 
-    * P2.2-Drought response increases with height at time of drought.
-    * P2.2a - Drought response is better predicted by height than DBH.
+1.2.
+(Beyond the effects of height,) canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
+Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
 
-2.3. Large trees suffer more during drought because of greater exposure (to radiation, wind, etc.)--either in relation to neighboring trees or because of position on landscape.
+    (add predictions for model without height)
+    * P1.2- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
+    * P1.2alt- After accounting for height, subcanopy trees have lowest resistance. (partially supported)
 
-    * P3.1- Trees currently in a dominant canopy position suffered more during drought.
+1.3. During drought, larger trees are at a relatively smaller disadvantage in drier microenvironments where rooting volume/depth is more important for reliable water access.
 
-2.4. Rooting volume/depth relative to water sources are critical in drought response. Effects of drought on larger trees are mediated by the fact that large trees have better access to water.
+  * P1.3- There is a negative interactive effect between tree height and topographic water index.
 
-  * P3.2- drought response increases with topographic wetness index
+**2. Species functional traits predict drought response?**
 
-**3. Do species functional traits predict drought response?**
+   * P2.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
+    * P2.2 - higher percent leaf area predicts higher drought stress
+    * P2.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [@abrams_adaptations_1990] [@guerfel_impacts_2009] <- *valid?
+it looks at olive trees*
+    * P2.4 - TLP correlates negatively with drought resistance (NEVER tested),
+    * P2.5 - higher wood density means more drought resistant
 
-   * P3.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
-    * P3.2 - higher percent leaf area predicts higher drought stress
-    * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [@abrams_adaptations_1990] [@guerfel_impacts_2009] <- *valid?
-it looks at olive trees*
-    * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
-    * P3.5 - higher wood density means more drought resistant
+**3. How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**
 
-**4. How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**
-
-4.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
-4.1_alt. Larger trees suffer more because they have more drought vulnerable traits.
+3.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
+3.1_alt. Larger trees suffer more because they have more drought vulnerable traits.
 
@@
 
-**4. How do droughts vary in their affect on tree resistance? **
+**4. How do droughts vary in their effect on tree resistance? **
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
added references to GitHub issues

@@
 
-    (add predictions for model without height)
+    (add predictions for model without height- ([issue #44](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/44)))
     * P1.2- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
@@
 
-  * P1.3- There is a negative interactive effect between tree height and topographic water index.
+  * P1.3- There is a negative interactive effect between tree height and topographic water index. ([issue #45](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/45))
 
@@
 
-**3. How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**
+**3.
+How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**  ([issue #43](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/43))
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
working on hypotheses/ predictions

@@
 
-**2. Species functional traits predict drought response?**
+**2. Which species functional traits predict drought response (and which are the best predictors)?**
 
-   * P2.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
-    * P2.2 - higher percent leaf area predicts higher drought stress
-    * P2.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [@abrams_adaptations_1990] [@guerfel_impacts_2009] <- *valid?
-it looks at olive trees*
-    * P2.4 - TLP correlates negatively with drought resistance (NEVER tested),
-    * P2.5 - higher wood density means more drought resistant
+2.1.
+Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al.
+2015, Friedrichs et al.
+2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
+However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
+    * P2.1 - wood density correlates positively drought resistant
+    * P2.2 - leaf mass area correlates positively to drought resistance 
+    * P2.3 - diffuse porous species are more sensitive than ring porous  
+    * P2.4 -leaf area shrinkage upon dessication (PLA) is negatively correlated with drought resistance
+    * P2.5 - TLP correlates negatively with drought resistance
+
+ 2.2.
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
working on hypotheses/ predictions

@@
 
-**2. Which species functional traits predict drought response (and which are the best predictors)?**
+**2. How do species' traits--alone and in interaction with tree size--influence drought response?**
 
@@
 2.2.
+Traits vary with microenvironment, with more drought-resistant traits associated to taller trees (and drier topographic positions) ([issue #43](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/43))
 
-**3.
-How are functional traits distributed across size classes, and how does this affect size-resistance relationship?**  ([issue #43](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/43))
+     * P2.2a- traits (all 5 above) correlate with tree height 
+     * (P2.2b- traits (all 5 above) correlate with TWI)
 
-3.1. Larger/ more exposed trees have more drought resistant traits, meaning size effects are buffered by traits  
-3.1_alt. Larger trees suffer more because they have more drought vulnerable traits.
-
-     * P3.1a- TLP is lower (larger negative) in taller/canopy trees
-     * P3.1b- diffuse porous species more common in understory
-
-     * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.
+2.3. Size-dependent drought resistance is driven by--or alternatively, buffered by--functional traits (i.e., larger trees have more drought vulnerable traits).
+    * P3.1 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI).
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
finished (for now) with hypotheses/ predictions.

@@
     * P2.3 - diffuse porous species are more sensitive than ring porous  
-    * P2.4 -leaf area shrinkage upon dessication (PLA) is negatively correlated with drought resistance
-    * P2.5 - TLP correlates negatively with drought resistance
+    * P2.4 -leaf area shrinkage upon dessication (PLA) is negatively correlated with drought resistance (supported as one of the best predictors)
+    * P2.5 - TLP correlates negatively with drought resistance  (supported as one of the best predictors)
 
@@
 
-[differences among droughts:]
-**I think for #4 here the focus should be on hypotheses looking at the overall trend compared to the specific scenarios?**
+**3. Did responses vary across droughts?**
+3.1 Overall community response is proportional to the severity of each drought.
 
-**1. How significant are individual drought years on a long time-frame trend?**
-1.1 Individual drought scenario is a strong predictor of drought stress.
+    * P3.1 -  (How do you define severity? min PDSI? duration? it may be tricky to find the right metric, but I agree that its interesting to compare across droughts)
 
-    * P1.1 - Drought stress is proportional to the severity of each drought.
+3.2. Predictor variables varied across droughts.
 
-**4. How do droughts vary in their effect on tree resistance? **
-
-    * P4.1 - The combined-years scenario will have more explaining variables than the individual scenarios.
-    * P4.2 - Inclusion of leaf hydraulic traits does not eliminate or significantly reduce effect of tree size.
-    * P4.3 - Drought resistance for each scenario will be explained more by hydraulic traits than by the biophysical environment
+    * P3.2 - There will be overall consistnecy in terms of the directions of responses to predictor variables; however, the best set of predictor variables will vary across droughts.
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 1.2.
-(Beyond the effects of height,) canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
+Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-
-    (add predictions for model without height- ([issue #44](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/44)))
-    * P1.2- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
-    * P1.2alt- After accounting for height, subcanopy trees have lowest resistance. (partially supported)
+    * P1.2a- Canopy position is a better predictor of resistance than height, with trees currently in a dominant canopy position having lowest resistance. (rejected)
+    * P1.2b- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
+    * P1.2b_alt- After accounting for height, subcanopy trees have lowest resistance. (partially supported)
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
#43

@@
     * P1.2b- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
-    * P1.2b_alt- After accounting for height, subcanopy trees have lowest resistance. (partially supported)
 
@@
 2.3. Size-dependent drought resistance is driven by--or alternatively, buffered by--functional traits (i.e., larger trees have more drought vulnerable traits).
-    * P3.1 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI).
+    * P2.3 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI).
+That is, the effects of height and and microhabitat will not be negated in the best model including size, microhabitat, and traits.
+(supported)
 

### 2019-08-02 manuscript/climate_sensitivity_draft.Rmd
Updated paper briefly

@@
 
-We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity, which together comprised 97% of whole-ecosystem between 2008 and 2013 [@helcoski_growing_2019].
-Cores were collected in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees were selected at random from each species in 2010-2011, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
+We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019].
+Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
+In 2010-2011, cores were collected from randomly selected live trees were selected at random from each species, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
 In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
 Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been [published](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) in association with [@helcoski_growing_2019]: (ITRDB).
+The resulting chronologies have been [published](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) in association with @helcoski_growing_2019: (ITRDB).
 *is this a database?*
@@
 
+```{r, SI Table 1, eval=TRUE, echo=FALSE, message=FALSE}
+heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
+kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
 **include TWI by trait graphs (2) here**

### 2019-08-02 scripts/canopy_heights.R
Create height_regression table

@@ -3,7 +3,6 @@
 # Developed by: Ian McGregor - mcgregori@si.edu
 # R version 3.5.3 - First created April 2019
 ######################################################
-
 library(data.table)
 library(RCurl)
 library(ggplot2)
@@ -13,20 +12,18 @@ library(stringr)
 #1. prepare SCBI data ####
 heights <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_dimensions/tree_heights/SCBI_tree_heights.csv"), stringsAsFactors = FALSE)
 
-dbh_2008 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem1.csv"))
-dbh_2008$dbh.cm <- dbh_2008$dbh/10 #cm
+dbh_2008 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem1.csv"), stringsAsFactors = FALSE)
+dbh_2008$dbh.cm <- as.numeric(dbh_2008$dbh)/10 #cm
 
-dbh_2013 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem2.csv"))
+dbh_2013 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem2.csv"), stringsAsFactors = FALSE)
 dbh_2013$dbh.cm <- dbh_2013$dbh/10 #cm
 
-dbh_2018 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/scbi.stem3_TEMPORARY.csv"), stringsAsFactors = FALSE)
+dbh_2018 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
 
 # #get stemIDs for each stem (leaving this here if need be)
[…]

### 2019-08-02 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -568,7 +568,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
 ##leaf traits gained from this include PLA_dry_percent, LMA_g_per_m2, Chl_m2_per_g, and WD [wood density]
 
-leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBELAM347ZL7W3JUGZ3S5IHYKE"), stringsAsFactors = FALSE)
+leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEMTPO5NOAZ2XCEGJPS5JWOTK"), stringsAsFactors = FALSE)
 
 leaf_traits <- leaf_traits[, c(1,8,12,26,28)]
 
@@ -1425,11 +1425,11 @@ write.csv(coeff_table, "manuscript/tables_figures/tested_traits_best_coeffs.csv"
 
 ##6c. standalone code to get coefficients and r2 #### 
 ##(for paper, should do ONLY w/REML=TRUE) 
-best <- lmm_all[[46]]
+best <- lmm_all[[58]]
 cof <- data.frame("value" = coef(summary(best))[ , "Estimate"])
 cof$value <- round(cof$value, 3)
 
-r <- rsquared(fit1) #gives R^2 values for models. "Marginal" is the R^2 for just the fixed effects, "Conditional" is the R^2 for everything.
+r <- rsquared(best) #gives R^2 values for models. "Marginal" is the R^2 for just the fixed effects, "Conditional" is the R^2 for everything.
 
 
 ##6d. original table with models based on github issue predictions ####

### 2019-08-03 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
added hypotheses/predictions summary table
relates to issue #46

@@
 
+**Table 1.
+Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC≥2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC.
+Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response.
+(Table 1.xlsx)
+
+(The content below needs to be re-written / mostly removed to go with Table 1.)
+
 **1. Why do larger trees suffer greater growth declines during drought?**
@@
 
-  * P1.3- There is a negative interactive effect between tree height and topographic water index. ([issue #45](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/45))
+  * P1.3- There is a negative interactive effect between tree height and topographic water index.
+(rejected- [issue #45](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/45))
 
@@
 2.3. Size-dependent drought resistance is driven by--or alternatively, buffered by--functional traits (i.e., larger trees have more drought vulnerable traits).
+
   * P2.3 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI).
@@
 3.1 Overall community response is proportional to the severity of each drought.
-
-    * P3.1 -  (How do you define severity? min PDSI? duration? it may be tricky to find the right metric, but I agree that its interesting to compare across droughts)
+(Do you really want to add this hypothesis?
+How do you define severity?
+min PDSI?
+duration?
+it may be tricky to find the right metric, but I agree that its interesting to compare across droughts)
 
@@
 
-**Table 1.
+**Table 2.
 Species analyzed here, listed in descending order of ANPP_stem.

### 2019-08-03 manuscript/climate_sensitivity_draft.Rmd
refined table/ table heading
re-wrote hypothesis text
Following addition of Table 1, it was possible to condense text to one paragraph (last paragraph of intro).

@@
 
+Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
+Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
+
 *Second, how do species' traits--alone and in interaction with tree size--influence drought response?* 
@@
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
+
+Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al.
+2015, Friedrichs et al.
+2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
+However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
 
@@
 Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
-This study addressed this by analyzing trees' resistance to drought within and across three defined drought periods.
 
-*Hypotheses*
+We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature ([@bennett_larger_2015], REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
+
+Here, we combine tree-ring records covering three droughts (1964-66, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses.
+First, we focus on the role of tree size and its interaction with microenvironement.
+We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion (Eastern US temperate deciduous) represented by only one study in Bennett et al (2015) (H1.0).
+We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees indry but not perpetually wet microsites (H1.3).
+Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
+We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point will prove better predictors (P2.1d-e).
+We then test whether these traits correlate with tree height (P2.2), potentially driving the observed tendency for taller trees to suffer more during drought (P2.3).
+Finally, we focused on variability among droughts, asking how  community resistance varied across droughts (H3.1) and whether the factors confirming vulnerability or resistance varied across droughts (H3.2).
 
@@
 Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response.
+"(yes/no) or ("no/yes")" indicates tendancies in opposite directions in univariate tests versus the best full models, respectively.
 (Table 1.xlsx)
-
-(The content below needs to be re-written / mostly removed to go with Table 1.)
-
-**1. Why do larger trees suffer greater growth declines during drought?**
-
-1.0.
-Consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance.
-(Note that Bennett et al.
-2015 identified only one study on tree growth responses to drought in the Eastern US temperate deciduous biome.
-We know little about how tree size shapes drought response in this biome.)
-
-  * P1.0-Drought resistance decreases with DBH at time of drought. (supported most of the time)
-
-1.1. Drought resistance decreases with tree height.
-
-  * P1.1- Drought resistance decreases with height at time of drought, and height is a better predictor than DBH. (supported most of the time)
-
-1.2.
-Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
-Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-  * P1.2a- Canopy position is a better predictor of resistance than height, with trees currently in a dominant canopy position having lowest resistance. (rejected)
-  * P1.2b- After accounting for height, trees currently in a dominant canopy position have lowest resistance. (partially supported)
-
-1.3. During drought, larger trees are at a relatively smaller disadvantage in drier microenvironments where rooting volume/depth is more important for reliable water access.
-
-  * P1.3- There is a negative interactive effect between tree height and topographic water index.
-(rejected- [issue #45](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/45))
-
-**2. How do species' traits--alone and in interaction with tree size--influence drought response?**
-
-2.1.
-Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al.
-2015, Friedrichs et al.
-2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
-However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
-    * P2.1 - wood density correlates positively drought resistant
-    * P2.2 - leaf mass area correlates positively to drought resistance 
-    * P2.3 - diffuse porous species are more sensitive than ring porous  
-    * P2.4 -leaf area shrinkage upon dessication (PLA) is negatively correlated with drought resistance (supported as one of the best predictors)
-    * P2.5 - TLP correlates negatively with drought resistance  (supported as one of the best predictors)
-
-2.2.
-Traits vary with microenvironment, with more drought-resistant traits associated to taller trees (and drier topographic positions) ([issue #43](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/43))
-
-  * P2.2a- traits (all 5 above) correlate with tree height 
-  * (P2.2b- traits (all 5 above) correlate with TWI)
-
-2.3. Size-dependent drought resistance is driven by--or alternatively, buffered by--functional traits (i.e., larger trees have more drought vulnerable traits).
-
-  * P2.3 Inclusion of traits in the statistical model does not eliminate or significantly reduce effect of tree size (or TWI).
-That is, the effects of height and and microhabitat will not be negated in the best model including size, microhabitat, and traits.
-(supported)
-
-**3. Did responses vary across droughts?**
-3.1 Overall community response is proportional to the severity of each drought.
-(Do you really want to add this hypothesis?
-How do you define severity?
-min PDSI?
-duration?
-it may be tricky to find the right metric, but I agree that its interesting to compare across droughts)
-
-3.2. Predictor variables varied across droughts.
-
-  * P3.2 - There will be overall consistnecy in terms of the directions of responses to predictor variables; however, the best set of predictor variables will vary across droughts.
 

### 2019-08-03 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 First, we focus on the role of tree size and its interaction with microenvironement.
-We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion (Eastern US temperate deciduous) represented by only one study in Bennett et al (2015) (H1.0).
+We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in Bennett et al (2015) (H1.0).
 We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees indry but not perpetually wet microsites (H1.3).
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point will prove better predictors (P2.1d-e).
+We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area shrinkage upon dehydration and turgor loss point will prove better predictors (P2.1d-e).
 We then test whether these traits correlate with tree height (P2.2), potentially driving the observed tendency for taller trees to suffer more during drought (P2.3).

### 2019-08-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Predicting forest responses to drought is an increasingly critical task under climate change effects.
-Part of the problem is due the lack of a set definition of drought in recent studies, while often the effects of biophysical and hydraulic traits are interpreted separately.
+Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.
 In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
 Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
-Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested again leaf hyraulic trait measurements and microhabitat biophysical data.
-Individual-level growth responses to three droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.
+Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
+Individual-level growth responses to the three individual droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.
 We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.
@@
 
-[Understanding forest responses to drought requires increased functional understanding of the factors that confir individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
+[Understanding forest responses to drought requires increased functional understanding of the factors that confer individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
 Therefore, in order to understand whole-forest response to drought, we need a functional understanding of how responses vary by tree size, microhabitat, and species.
@@
 *First, what drives the observed tendency for large trees to suffer more during drought?*  
-[@bennett_larger_2015] showed that in forests globally, large trees suffer greater growth reductions during drought.
+@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought.
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
 This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees.
-Counteracting these effects, (3) the larger root systems of larger trees may confir an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Counteracting these effects, (3) the larger root systems of larger trees may confer an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
 
@@
 
-*Second, how do species' traits--alone and in interaction with tree size--influence drought response?* 
+*Second, how do species' traits - alone and in interaction with tree size - influence drought response?* 
 Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
 The solution is a trait-based approach.
-Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and SLA ([Medeiros et al.
+Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and specific leaf area (SLA) ([Medeiros et al.
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
@@
 2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
-However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication and turgor loss point, which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
+However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA) and turgor loss point (TLP), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
 
 Moreover, there may be an interaction between traits and tree size.
-It is possible that the pattern observed by [@bennett_larger_2015] could be caused by smaller trees being more drought resistant.
+It is possible that the pattern observed by @bennett_larger_2015 could be caused by smaller trees being more drought resistant.
 Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.
@@
 
-We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature ([@bennett_larger_2015], REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
+We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature [@bennett_larger_2015] (REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
 
@@
 **Table 1.
-Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC≥2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC.
+Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC>=2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC.
 Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response.
@@
 (Table 1.xlsx)
+
+```{r, Table 1, eval=TRUE, echo=FALSE, message=FALSE}
+library(readxl)
+library(kableExtra)
+Table1 <- read_xlsx("tables_figures/Table1.xlsx")
+Table1 <- sapply(Table1, as.character)
+Table1[is.na(Table1)] <- " "
+
+kable(Table1) %>%
+  kable_styling(bootstrap_options = "striped", full_width = F)
+```
 

### 2019-08-06 manuscript/climate_sensitivity_draft.Rmd
Updated with small changes

@@
 
-```{r, Table 1, eval=TRUE, echo=FALSE, message=FALSE}
-library(readxl)
-library(kableExtra)
-Table1 <- read_xlsx("tables_figures/Table1.xlsx")
-Table1 <- sapply(Table1, as.character)
-Table1[is.na(Table1)] <- " "
-
-kable(Table1) %>%
-  kable_styling(bootstrap_options = "striped", full_width = F)
-```
+![Table 1](tables_figures/Table1.png){width=500px}
 
@@
 Figure modified from [@helcoski_growing_2019].
+
 ![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
@@
 
-```{r, Table 1, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table x, eval=TRUE, echo=FALSE, message=FALSE}
 traits_test <- read.csv("tables_figures/tested_traits_all.csv")
+traits_test$null_model <- gsub("resist.value ~ ", "", traits_test$null_model)
+traits_test$tested_model <- NULL
+traits_test$null_model_year <- gsub("resist.value ~ ", "", traits_test$null_model_year)
+traits_test$tested_model_year <- NULL
+
 kable(traits_test, booktabs=TRUE, format = "latex", caption = "All variables tested for climate sensitivity against appropriate null models, in line with hypothesis predictions.

### 2019-08-06 manuscript/climate_sensitivity_draft.Rmd
Temporary update

@@
 We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in Bennett et al (2015) (H1.0).
-We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees indry but not perpetually wet microsites (H1.3).
+We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees in dry but not perpetually wet microsites (H1.3).
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
@@
 Finally, we focused on variability among droughts, asking how  community resistance varied across droughts (H3.1) and whether the factors confirming vulnerability or resistance varied across droughts (H3.2).
-
-**Table 1.
-Summary of hypotheses, predictions and results** We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction and model dAIC>=2 relative to the approriate null model or to any alternative multivariate model with within 2 dAIC.
-Parentheses indicate that predictions were partially supported (or rejected)-- i.e., with direction of response matching (or contradicting) the prediction but model dAIC<2 relative to the approriate null model, when response partially matched predicted order for categorical variable (canopy position), or when an alternative multivariate model within 2 dAIC had a contradictory response.
-"(yes/no) or ("no/yes")" indicates tendancies in opposite directions in univariate tests versus the best full models, respectively.
-(Table 1.xlsx)
-
-![Table 1](tables_figures/Table1.png){width=500px}
 
@@
 ### Results
+![Table 1](tables_figures/Table1.png){width=500px}
+
+**Table 1. Summary of hypotheses, predictions and results** 
+We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction. 
+H2.3-H3.2 are based on if models containing species traits in H2.1 had dAIC>=2 relative to the appropriate null model or to any alternative multivariate model within 2 dAIC. 
+Parentheses indicate that predictions were partially supported (or rejected).
+In other words, parentheses indicate when the direction of response matched (or contradicted) the prediction in some but not all models with dAIC<2 relative to the appropriate null model.
+With categorical variables such as crown position,  a "(yes)" notation describes when the trend matched the prediction, but it wasn't significant
+A "(yes/no) or ("no/yes")" indicates tendencies in opposite directions in univariate tests versus the best full models, respectively. 
+"(yes)
+(Table 1.xlsx)
+
+trying to test how the negative effect of height is caused by the traits,
+
+(no) means there's a trend in the expected direction but not consistently significant
+
+H2.3
+if the effect is negative and is in every top model (<dAIC2), that would be "yes."
+if the effect is negative mostly
+
 Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???

### 2019-08-06 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -1111,11 +1111,12 @@ library(stringr)
 
 ##6a. test each variable individually for each drought scenario ####
 sum_mod_traits <- data.frame(
-  "prediction" = c(1.1, 2.1, 2.2, 2.3, 2.4, 3.1, 3.2, 3.3, 3.4, 3.5),
-  "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all",  "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
-  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position", "ln[topographic.wetness.index]", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
+  "prediction" = c(3.1, 2.1, 2.2, 2.3, 2.4, 3.1, 3.2, 3.3, 3.4, 3.5),
+  "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all", "TWI.ln*height.ln.m", "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
+  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position", "ln[topographic.wetness.index]*ln[height]", "ln[topographic.wetness.index]", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
   "null_model" = 
     c("resist.value ~ (1|sp/tree)",
+      "resist.value ~ year+(1|sp/tree)",
       "resist.value ~ year+(1|sp/tree)",
       "resist.value ~ year+(1|sp/tree)",
       "resist.value ~ height.ln.m+year+(1|sp/tree)",

### 2019-08-08 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
a bit of work on methods

@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH (described below), along with tree location in the plot to determine the topographic wetness index.
-*I think that's it, really, in terms of data we're using coming from the censuses.*
+Furthermore, we used data for all stems >= 10cm to analyze functional trait composition relative to tree height.
+*I added the last sentence-- not sure if you've described that below yet.
+I don't think you need to note here which data we use, so long as that's described in the portions where you describe the analysis.*
 
@@
 Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees were selected at random from each species, with at least 30 of those trees having a diameter at breast height (DBH) of at least 10cm [@bourg_initial_2013].
-In 2016-2017, cores were collected from all dead trees found in the annual mortality census [@gonzalezakre_patterns_2016].
+In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH >= 10cm [@bourg_initial_2013].
+In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
 Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been [published](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) in association with @helcoski_growing_2019: (ITRDB).
-*is this a database?*
+The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
+*is this a database?
+Yes.
+Ryan submitted the data but I don't think its posted yet.
+We should also cite GitHub/Zenodo here.
+I'll come back to that.
+*
 

### 2019-08-09 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -1199,7 +1199,7 @@ for (i in seq_along(1:12)){
                   coeff <- data.frame(coef(summary(lmm_all[[z]]))[ , "Estimate"]) ##2
                   coeff[,2] <- rownames(coeff)
                   colnames(coeff) <- c("value", "model_var")
-                  coeff$value <- round(coeff$value, 3)
+                  coeff$value <- round(coeff$value, 4)
                   coeff$combo <- paste0(coeff$model_var, " (", coeff$value, ")")
                   coeff$mod <- gsub("coef_", "", column_cof)
                   
@@ -1245,7 +1245,7 @@ for (i in seq_along(1:12)){
                   coeff <- data.frame(coef(summary(lmm_all[[z]]))[ , "Estimate"]) ##2
                   coeff[,2] <- rownames(coeff)
                   colnames(coeff) <- c("value", "model_var")
-                  coeff$value <- round(coeff$value, 3)
+                  coeff$value <- round(coeff$value, 4)
                   coeff$combo <- paste0(coeff$model_var, " (", coeff$value, ")")
                   coeff$mod <- gsub("coef_", "", column_cof)
                   

### 2019-08-09 scripts/tables_for_manuscript.R
Create tables_for_manuscript.R

@@ -1,0 +1,97 @@
+#tables for manuscript
+
+#Table 1 was originally made as .xslx
+#Table 3 was originally made as .xlsx from tested_traits_all.csv
+
+#Table 2 ####
+library(RCurl)
+library(tidyr)
+library(data.table)
+library(dplyr)
+
+species <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+
+table <- species[, c(1:3,6,17)]
+
+positions <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_dimensions/tree_crowns/cored_dendroband_crown_position_data/dendro_cored_full.csv"))
+
+table$position_all <- positions$crown.position[match(table$tag, positions$tag)]
+table$position_all <- gsub("D", "dominant", table$position_all)
+table$position_all <- gsub("C", "co-dominant", table$position_all)
+table$position_all <- gsub("I", "intermediate", table$position_all)
+table$position_all <- gsub("S", "suppressed", table$position_all)
+
+table1 <- table %>%
+  group_by(sp) %>%
+  summarize(n_cores=n())
+
+table2 <- table %>%
+  group_by(sp, position_all) %>%
+  summarize(count=n())
+
+table2 <- spread(table2, key=position_all, value=count)
+setnames(table2, old="<NA>", new="prior dead")
+
+table2$n_cores <- table1$n_cores[match(table2$sp, table1$sp)]
+table2 <- table2[, c(1,7,3,2,4:6)]
+table2 <- table2[!table2$sp %in% c("frni", "pist"), ]
+
+#bring in trait values
+trees_all <- read.csv("manuscript/tables_figures/trees_all.csv", stringsAsFactors = FALSE)
+
+traits <- trees_all[,c(2,6:10)]
+traits <- traits[!duplicated(traits$sp), ]
+
+table2 <- merge(table2, traits, by="sp")
+table2[,c(8:11)] <- round(table2[,c(8:11)], 2)
+
+#bring in mean DBH and DBH range for each species
+
+dbh_cored <- species[,c(1:9)]
+
+detach("package:tidyr", unload = TRUE)
+
+aggregate(dbh_cored[,c(2,9)], by = list(dbh_cored$sp), "mean")
+
+dbh_cored2 <- dbh_cored %>%
+  group_by(sp) %>%
+  summarise(mean = mean(dbh2018)) %>%
[…]

### 2019-08-09 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
Updated rmd file

@@
 
+*Let's put Table 1 here.
+We need to give the hypotheses/ predictions here.
+It seems counterintuitive to put the results right up front like that, but it actually tends to be a higher-impact style of writing, and recommended [here](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619).*
+
 ### Methods
@@
 ```{r, Table-cores-position, eval=TRUE, echo=FALSE, message=FALSE}
-library(RCurl)
-library(dplyr)
-library(tidyr)
-library(data.table)
-library(knitr)
-library(kableExtra)
-
-species <- read.csv("D:/Dropbox (Smithsonian)/Github_Ian/McGregor_climate-sensitivity-variation/data/core_list_for_neil.csv", stringsAsFactors = FALSE)
-
-table <- species[, c(1:3,6,17)]
-
-positions <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_dimensions/tree_crowns/cored_dendroband_crown_position_data/dendro_cored_full.csv"))
-
-table$position_all <- positions$crown.position[match(table$tag, positions$tag)]
-table$position_all <- gsub("D", "dominant", table$position_all)
-table$position_all <- gsub("C", "co-dominant", table$position_all)
-table$position_all <- gsub("I", "intermediate", table$position_all)
-table$position_all <- gsub("S", "suppressed", table$position_all)
-
-table1 <- table %>%
-  group_by(sp) %>%
-  summarize(n_cores=n())
-
-table2 <- table %>%
-  group_by(sp, position_all) %>%
-  summarize(count=n())
-
-table2 <- spread(table2, key=position_all, value=count)
-setnames(table2, old="<NA>", new="prior dead")
-
-table2$n_cores <- table1$n_cores[match(table2$sp, table1$sp)]
-table2 <- table2[, c(1,7,3,2,4:6)]
-table2 <- table2[!table2$sp %in% c("frni", "pist"), ]
 
@@
 ```{r, traits-overview, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-
-trees_all <- read.csv("tables_figures/trees_all.csv", stringsAsFactors = FALSE)
-
-trees_all <- trees_all[!duplicated(trees_all$sp), ]
-
-traits_table <- data.frame(
-  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Wood density", "TLP"),
-  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "g/cm3", "MPa"))
-traits_table$abb <- c("rp", "PLA", "LMA", "WD", "TLP")
-
-traits_table$mean <- NA
-traits_table$min <- NA
-traits_table$max <- NA
-
-for (i in seq(along=traits_table$abb[2:5])){
-  trait <- traits_table$abb[2:5][i]
-
-  sub <- trees_all[grepl(trait, colnames(trees_all))]
-  traits_table$mean[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
-                              mean(sub[, 1], na.rm=TRUE), 
-                              traits_table$mean)
-
-  traits_table$min[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
-                              min(sub[, 1], na.rm=TRUE), 
-                              traits_table$min)
-
-  traits_table$max[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
-                              max(sub[, 1], na.rm=TRUE), 
-                              traits_table$max)
-}
-
-traits_table[, c("mean", "min", "max")] <- 
-  sapply(traits_table[, c("mean", "min", "max")], function(x)
-  round(x, 2)
-)
-traits_table$abb <- NULL
 
@@
 
-```{r, Table x, eval=TRUE, echo=FALSE, message=FALSE}
-traits_test <- read.csv("tables_figures/tested_traits_all.csv")
-traits_test$null_model <- gsub("resist.value ~ ", "", traits_test$null_model)
-traits_test$tested_model <- NULL
-traits_test$null_model_year <- gsub("resist.value ~ ", "", traits_test$null_model_year)
-traits_test$tested_model_year <- NULL
-
-kable(traits_test, booktabs=TRUE, format = "latex", caption = "All variables tested for climate sensitivity against appropriate null models, in line with hypothesis predictions.
-Each variable was tested for each drought scenario excepting \"year\" ") %>%
-  kableExtra::landscape()
-  # kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "scale_down")
-
-```
+![Table 3 (example!)](tables_figures/Table3.png)
 

### 2019-08-12 manuscript/climate_sensitivity_draft.Rmd
Updated manuscript

@@
 ### Introduction
-
 Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.
@@
 It seems counterintuitive to put the results right up front like that, but it actually tends to be a higher-impact style of writing, and recommended [here](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619).*
+
+**Table 1. Summary of hypotheses, predictions and results** 
+![Table 1](tables_figures/Table1.png){width=500px}
+
+We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction. 
+H2.3-H3.2 are based on if models containing species traits in H2.1 had dAIC>=2 relative to the appropriate null model or to any alternative multivariate model within 2 dAIC. 
+Parentheses indicate that predictions were partially supported (or rejected).
+In other words, parentheses indicate when the direction of response matched (or contradicted) the prediction in some but not all models with dAIC<2 relative to the appropriate null model.
+With categorical variables such as crown position,  a "(yes)" notation describes when the trend matched the prediction, but it wasn't significant
+A "(yes/no) or ("no/yes")" indicates tendencies in opposite directions in univariate tests versus the best full models, respectively. 
+"(yes)
 
@@
 In August 2018, we collected leaf samples from three individuals of each species ...
-(**Nobby's description of methods for the following**)
+(**Nobby's description of methods for the following (see word document**)
 1. PLA
@@
 
-```{r, Table-cores-position, eval=TRUE, echo=FALSE, message=FALSE}
-
-kable(table2) %>%
-  kable_styling(bootstrap_options = "striped", full_width = F)
-```
-```{r, traits-overview, eval=TRUE, echo=FALSE, message=FALSE}
-
-kable(traits_table) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
-
+```{r, Table2, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+table2 <- read.csv("tables_figures/Table2.csv")
+kable(table2, booktabs=TRUE, format = "latex", caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
+Units of measurements are in mm (DBH), % (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
 ```
@@
 
-**Figure 1.
-Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-
-![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
-
 ### Results
-![Table 1](tables_figures/Table1.png){width=500px}
-
-**Table 1. Summary of hypotheses, predictions and results** 
-We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction. 
-H2.3-H3.2 are based on if models containing species traits in H2.1 had dAIC>=2 relative to the appropriate null model or to any alternative multivariate model within 2 dAIC. 
-Parentheses indicate that predictions were partially supported (or rejected).
-In other words, parentheses indicate when the direction of response matched (or contradicted) the prediction in some but not all models with dAIC<2 relative to the appropriate null model.
-With categorical variables such as crown position,  a "(yes)" notation describes when the trend matched the prediction, but it wasn't significant
-A "(yes/no) or ("no/yes")" indicates tendencies in opposite directions in univariate tests versus the best full models, respectively. 
-"(yes)
-(Table 1.xlsx)
-
-trying to test how the negative effect of height is caused by the traits,
-
-(no) means there's a trend in the expected direction but not consistently significant
-
-H2.3
-if the effect is negative and is in every top model (<dAIC2), that would be "yes."
-if the effect is negative mostly
-
 Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
@@
 
-![Table 3 (example!)](tables_figures/Table3.png)
+**Describe each individual drought scenario **
+1964-1966
+1977
+1999
 
-```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
-candidates <- read.csv("tables_figures/candidate_traits.csv")
-kable(candidates, booktabs=TRUE, format = "latex", caption = "Candidate variables to be included in full model, chosen by dAICc>2 when individually tested against a null model") %>%
-  kable_styling(bootstrap_options = "striped", latex_options ="hold_position")
-```
+**Figure 1.
+Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from [@helcoski_growing_2019].
 
-```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
-traits_best <- read.csv("tables_figures/tested_traits_best.csv")
-kable(traits_best, booktabs=TRUE, format = "latex", caption = "Best full models for each drought scenario") %>%
-  kable_styling(bootstrap_options = "striped", latex_options = c("scale_down", "hold_position"))
-```
+![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
+
+**Results for first main question: what drives the observed tendency for large trees to suffer more during drought?**
+H1.0, H1.1, H1.2, H1.3
+DBH, height, crown position, and TWI
+
+![Height profile graphs]
+(tables_figures/height graphs ){width=500px}
+
+**Results for second main question: how do species' traits - alone and in interaction with tree size - influence drought response? **
+H2.1, H2.2, H2.3
+Hydraulic traits alone, traits with height
+
+![Table 3 (example!)](tables_figures/Table3.png) #this is tested_traits_all
+
+**Results for third main question: are responses similar or variable across individual drought years?**
+H3.1 and H3.2
+Combining biophysical with hydraulic traits, which come out as candidates for best model?
 
@@
 ### Supplementary Information
-
-```{r echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='H'}
-  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
+```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
+heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
+kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
 ```
@@
 
-(see Issue #32)
-
-```{r, SI Table 1, eval=TRUE, echo=FALSE, message=FALSE}
-heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
-kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
+```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
+tableS2 <- read.csv("tables_figures/candidate_traits.csv", stringsAsFactors = FALSE)
+kable(tableS2, booktabs=TRUE, format = "latex", caption = "Candidate variables for best model") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 
+```{r, Plot, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
+  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
+```
+
+*how do we want to present Table S3? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
+```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
+tableS3 <- read.csv("tables_figures/top_models_dAIC.csv", stringsAsFactors = FALSE)
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Top model variations for each drought scenario, with dAICc values <= 2") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+(see Issue #32)
+
 **include TWI by trait graphs (2) here**

### 2019-08-12 scripts/graphs_for_manuscript.R
Added code for tables and figures

@@ -5,8 +5,9 @@
 ######################################################
 library(ggplot2)
 
-#1 NEON vertical height profiles
+#1. NEON vertical height profiles
 source('scripts/vertical_height_neon.R', echo=TRUE)
+NEON_list <- list(wind_plot, RH_plot, SAAT_plot, biotemp_plot)
 #########################################################################
 #2 height by crown position in 2018 ####
 library(RCurl) #2
@@ -14,7 +15,7 @@ library(tidyr) #2
 library(grid) #2
 library(gridExtra) #2
 
-#2a. heights for all cored trees ####
+##2a. heights for all cored trees ####
 trees_all <- read.csv("manuscript/tables_figures/trees_all.csv", stringsAsFactors = FALSE)
 
 scbi.stem3 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
@@ -51,23 +52,23 @@ current_ht <- current_ht[order(current_ht$tree, current_ht$year), ]
 
 #graphing height by crown position (for paper)
 current_ht <- current_ht[!is.na(current_ht$position_all), ]
-current_ht$position_all <- factor(current_ht$position_all, levels = c("dominant", "co-dominant", "intermediate", "suppressed"))
+current_ht$position_all_abb <- substring(current_ht$position_all, 0, 1)
+current_ht$position_all_abb <- toupper(current_ht$position_all_abb)
+current_ht$position_all_abb <- factor(current_ht$position_all_abb, levels = c("D", "C", "I", "S"))
 
-quantile(current_ht$height.m, c(.95), na.rm=TRUE) #95% quantile = 35.002m
-
-quant <- data.frame(yintercept = 35.0022, Lines = "95th percentile")
 
 heights <-
-  ggplot(data = current_ht) +
-  aes(x = position_all, y = height.m, group = position_all) +
+  ggplot(data = current_ht, aes(x = position_all_abb, y = height.m, group = position_all)) +
   # aes(x=position_all, y=height.m, fill=year) +
   geom_boxplot() +
-  xlab("crown position") +
+  xlab("Crown position") +
   ylab("Height [m]") +
[…]

### 2019-08-12 scripts/tables_for_manuscript.R

@@ -1,8 +1,6 @@
 #tables for manuscript
 
-#Table 1 was originally made as .xslx
-#Table 3 was originally made as .xlsx from tested_traits_all.csv
-
+#Table 1 was originally made as .xslx ####
 #Table 2 ####
 library(RCurl)
 library(tidyr)
@@ -43,55 +41,23 @@ traits <- trees_all[,c(2,6:10)]
 traits <- traits[!duplicated(traits$sp), ]
 
 table2 <- merge(table2, traits, by="sp")
-table2[,c(8:11)] <- round(table2[,c(8:11)], 2)
+table2[,c(9:12)] <- round(table2[,c(9:12)], 2)
 
 #bring in mean DBH and DBH range for each species
 
 dbh_cored <- species[,c(1:9)]
 
-detach("package:tidyr", unload = TRUE)
-
-aggregate(dbh_cored[,c(2,9)], by = list(dbh_cored$sp), "mean")
-
-dbh_cored2 <- dbh_cored %>%
+dbh_cored <- dbh_cored %>%
   group_by(sp) %>%
-  summarise(mean = mean(dbh2018)) %>%
-  summarise(range = max(dbh2018) - min(dbh2018))
+  summarise(mean = mean(dbh2018),
+            range = max(dbh2018) - min(dbh2018))
+dbh_cored$mean <- round(dbh_cored$mean, 2)
 
+table2 <- merge(dbh_cored, table2, by = "sp")
 
+setnames(table2, old=c("mean", "range", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), new=c("mean_DBH", "range_DBH", "RP", "PLA", "LMA", "TLP", "WD"))
 
-trees_all <- read.csv("tables_figures/trees_all.csv", stringsAsFactors = FALSE)
+write.csv(table2, "manuscript/tables_figures/Table2.csv", row.names=FALSE)
 
-trees_all <- trees_all[!duplicated(trees_all$sp), ]
+#Table 3 was originally made as .xlsx from tested_traits_all.csv ####
 
-
-traits_table <- data.frame(
-  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Wood density", "TLP"),
-  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "g/cm3", "MPa"))
[…]

### 2019-08-12 scripts/vertical_height_neon.R

@@ -26,8 +26,9 @@ library(gridExtra)
 dp <- data.frame("data" = c("SAAT", "wind", "biotemp", "RH", "SR"),
                  "id" = c("DP1.00002.001", "DP1.00001.001", "DP1.00005.001", "DP1.00098.001", "DP1.00014.001"),
                  "value" = c("tempSingleMean", "windSpeedMean", "bioTempMean", "RHMean", "difRadMean"),
-                 "xlabs" = c("Mean Air Temperature [°C]", "Mean Windspeed [m/s]", "Mean Infrared Biological Temperature [°C]", "Relative Humidity [%]", "Mean shortwave downward radiation [W/m^2]"))
+                 "xlabs" = c("Mean Air Temperature [°C]", "Wind [m/s]", "Mean Infrared Biological Temperature [°C]", "RH [%]", "Mean shortwave downward radiation [W/m^2]"))
 
+dp <- dp[c(2,4,1,3,5), ]
 
 dp[] <- lapply(dp, as.character)
 dp$value <- as.character(dp$value)
@@ -130,7 +131,7 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
   
   data_analy$vertPos_jitter <- NA
   
-  if(!i == 4){
+  if(!i == 2){
     data_analy$vertPos_jitter <- 
       ifelse(data_analy$month == "May", data_analy$verticalPosition + 0.05,
        ifelse(data_analy$month == "June", data_analy$verticalPosition + 0.25,
@@ -170,11 +171,29 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
       ggplot2::geom_errorbarh(aes(xmin=mmin-sd_min, xmax=mmin+sd_max, y=vertPos_jitter, color = month_f, linetype = "Min", height=0.8)) +
       labs(x = dp$xlabs[[i]], y = "Height [m]") +
       scale_y_continuous(breaks = scales::pretty_breaks(n = 6), limits=c(0,60)) +
-      theme_grey() +
+      theme_bw() +
       guides(linetype = guide_legend("Line type"))
   
   if(!i == 1){
     graph <- graph + theme(axis.title.y = element_blank(), axis.text.y=element_blank(), axis.ticks.y = element_blank())
+  }
+  
+  if(i == 2){
+    graph <- graph + scale_x_continuous(breaks=c(40,60,80,100))
[…]

### 2019-08-12 scripts/canopy_position_analysis.R
Changed PLA to loss and token for leaf traits csv

@@ -568,7 +568,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
 ##leaf traits gained from this include PLA_dry_percent, LMA_g_per_m2, Chl_m2_per_g, and WD [wood density]
 
-leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEMTPO5NOAZ2XCEGJPS5JWOTK"), stringsAsFactors = FALSE)
+leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEJ2XM4T3UWCR6WNPLS5LKJKE"), stringsAsFactors = FALSE)
 
 leaf_traits <- leaf_traits[, c(1,8,12,26,28)]
 
@@ -1113,7 +1113,7 @@ library(stringr)
 sum_mod_traits <- data.frame(
   "prediction" = c(3.1, 2.1, 2.2, 2.3, 1.2, 1.3, 2.4, 3.1, 3.2, 3.3, 3.4, 3.5),
   "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all", "position_all", "height.ln.m*TWI.ln", "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
-  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position alone", "crown.position w/height", "ln[height]*ln[topographic.wetness.index]", "ln[topographic.wetness.index]", "ring.porosity", "percent.leaf.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
+  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position alone", "crown.position w/height", "ln[height]*ln[topographic.wetness.index]", "ln[topographic.wetness.index]", "ring.porosity", "percent.loss.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
   "null_model" = 
     c("resist.value ~ (1|sp/tree)",
       "resist.value ~ year+(1|sp/tree)",

### 2019-08-12 manuscript/climate_sensitivity_draft.Rmd
Updated manuscript

@@
 - \captionsetup[figure]{font=small}
+- \captionsetup[table]{labelformat=empty}
 - \usepackage{dcolumn} #for sidewaystable
@@
 
+![Location of cored trees](tables_figures/Figure3.png){width=500px}
+
 Hydraulic traits were collected from SCBI and are summarized in Table 1.
@@
 
-![Time series of combined tree cores by species](tables_figures/Time_series_for_each_species.jpg){width=500px}
+![Time series of combined tree cores by species](tables_figures/Figure1.jpg){width=500px}
 
@@
 
-![Height profile graphs]
-(tables_figures/height graphs ){width=500px}
+![Height profile graphs](tables_figures/Figure2.png){width=500px}
 
@@
 
+![Figure S1](tables_figures/Table3.png) #this is tested_traits_all
 (see Issue #32)
-
-**include TWI by trait graphs (2) here**

### 2019-08-13 manuscript/climate_sensitivity_draft.Rmd
Changed section names to match Phytologist guidelines

@@
 
-### Abstract
+### Summary
 Predicting forest responses to drought is an increasingly critical task under climate change effects.
@@
 We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.
+
+*The Summary for research papers, which must be usable as a stand- alone document, must not exceed 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion.
+It should not contain citations of other papers.*
 
@@
 
-### Methods
+### Materials and Methods
 *Study site*
@@
 This paper mentions it would be good to have this data with respect to latent affects from drought.
-### Conclusion
+### Acknowledgements
 
+words
+
+### Author Contribution ####
 words

### 2019-08-13 manuscript/climate_sensitivity_draft.Rmd
Updated Figure 2 and manuscript with results/discussion outlines

@@
 ### Results
+Results outline
+
+Descriptions of Droughts
+
+- physical characteristics of drought: duration, min PDSI (and when occurred), see slette et al 2019 to see how they suggest reporting each one
+- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)
+  - make table for this
+  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
+  - this is followed by Figure 2
+
+Hypothesis results
+- one header for each main question
+- present results for each one, walk through via order and logic of Table 1
+- no citations here
+
 Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
@@
 
-![Table 3 (example!)](tables_figures/Table3.png) #this is tested_traits_all
+![Table 4 (example!)](tables_figures/Table4.png) #this is tested_traits_all
 
@@
 ### Discussion
+Discussion outline
+
+- first paragraph is summary of main findings, not re-presenting the results
+  - "we supported this hypothesis, but not this one" (following same order as Table 1 still)
+  - direction of responses seems to be relatively consistent but the individual responses vary
+- tie things in (see here
+  - main thing is that we're now better understanding what confers vulnerability or resilience on trees during drought - how forests respond to future droughts/climate change
+- "we filled the gap" = 1-2 sentences
+- limitations at our site:
+  - aren't able to analyze historical forest community, nor trees that were killed by these droughts [aka we don't have data for individuals that were most severely affected] (though we found that there's little variation in climate sensitivity for trees that were cored dead vs cored alive).
+  - p50/p80
+  - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
+  - We did not use crowding index because it has much more uncertainty
+- limitations for extrapolating other sites:
+  - forests are different from place to place, but we've seen how in forests around the world, forests tend to suffer more. We've identified some facets for why this happens. (cite other studies)
+  - has been observed elsewhere that individuals in more moist habitats are more susceptible to drought (cite other studies)
+  - the species may be different from other sites but the hydraulic traits are general across species;
+- next paragraph saying that our study advances understanding of droughts, e.g. bennett et al 2015 doesn't go into mechanisms but we do
+  - and we show that height is more important than exposure, but doesn't eliminate the effect of it
+  - further contribution is we show that two leaf hydraulic traits - relatively easy to measure - are good predictors of drought response, can be helpful for scaling up (e.g.
+from our site to eastern deciduous biome)
+  - biophysical mechanisms are things that should be seen as universal but relative importance of each can vary within each drought
+  - first study to show how these traits affect woody growth response to drought **confirmed by Lawren**
+  - science is better now.
+by advancing our understanding of the mechanisms for individual-level responses to drought, this opens door for better predictions (elaborate from above).
+"This is absolutely critical to predicting forest responses to future droughts, which we're likely to see more of in the future as a result to climate change.
+Forecasting forest responses to these droughts is a huge and important challenge", science is better.
 
@@
 
-![Figure S1](tables_figures/Table3.png) #this is tested_traits_all
+![Figure S1](tables_figures/FigureS1.png)
 (see Issue #32)

### 2019-08-13 scripts/tables_for_manuscript.R
Split Table 2 into Table 2 and S2 as per #48

@@ -57,7 +57,11 @@ table2 <- merge(dbh_cored, table2, by = "sp")
 
 setnames(table2, old=c("mean", "range", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), new=c("mean_DBH", "range_DBH", "RP", "PLA", "LMA", "TLP", "WD"))
 
+table2_SI <- table2[,c(4:9)]
+table2 <- table2[,-c(4:9)]
+
 write.csv(table2, "manuscript/tables_figures/Table2.csv", row.names=FALSE)
+write.csv(table2_SI, "manuscript/tables_figures/TableS2.csv", row.names=FALSE)
 
 #Table 3 was originally made as .xlsx from tested_traits_all.csv ####
 

### 2019-08-13 scripts/vertical_height_neon.R
Changed variable labels to have subscripts

@@ -185,14 +185,14 @@ for (i in seq(along=dp$value)){ #make 1:5 if using radiation (cloud vs sun thres
   if(i == 3){
     graph <- 
       graph + 
-      labs(x= expression(paste("T_air [",degree,"C]"))) +
+      labs(x= expression(paste("T"["air"], " [",degree,"C]"))) +
       scale_x_continuous(breaks=c(10,20,30), limits=c(5,35))
   }
   
   if(i == 4){
     graph <- 
       graph + 
-      labs(x = expression(paste("T_infrared [",degree,"C]"))) +
+      labs(x = expression(paste("T"["biological"], " [",degree,"C]"))) +
       scale_x_continuous(breaks=c(10,20,30), limits=c(5,35))
   }
   

### 2019-08-13 scripts/graphs_for_manuscript.R
Got heights figure ready as per #37

@@ -8,6 +8,7 @@ library(ggplot2)
 #1. NEON vertical height profiles
 source('scripts/vertical_height_neon.R', echo=TRUE)
 NEON_list <- list(wind_plot, RH_plot, SAAT_plot, biotemp_plot)
+names(NEON_list) <- c("wind_plot", "RH_plot", "SAAT_plot", "biotemp_plot")
 #########################################################################
 #2 height by crown position in 2018 ####
 library(RCurl) #2
@@ -197,11 +198,11 @@ for (i in seq(along=1:4)){
       
       if(i==1){
          q <- q + 
-            xlab("Mean turgor loss point [MPa]") +
+            xlab("TLP [MPa]") +
             scale_x_continuous(breaks=breaks_tlp, limits=limits_tlp)
       } else if(i==2){
          q <- q + 
-            xlab("Percent leaf area [%]") +
+            xlab("PLA [%]") +
             scale_x_continuous(breaks=breaks_pla, limits=limits_pla)
       }
    } else if(grepl("twi", type[[i]])){
@@ -233,6 +234,14 @@ for (i in seq(along=1:4)){
 
 ##3c. Make map of plot using TWI and cored tree locations ####
 species <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+species <- species[!species$sp %in% c("frni", "pist"), ]
+species$sp_fact <- as.factor(species$sp)
+
+color_pallete_function <- colorRampPalette(
+   colors = c("red", "blue", "brown", "purple", "dark green", "light blue"),
+   space = "Lab" # Option used when colors do not represent a quantitative scale
+)
+species_colors <- color_pallete_function(nlevels(species$sp_fact))
 
 cored_points <- SpatialPointsDataFrame(data.frame(species$NAD83_X, species$NAD83_Y), data=species)
 
@@ -241,12 +250,15 @@ plot.new()
 png("manuscript/tables_figures/Figure3.png", width=5, height=7, units="in", res=72)
 plot(topo, axes=FALSE, box=FALSE, 
      legend.args = list(text="Topographic Wetness Index", side=4, font=2, line=2.5, cex=0.8))
-plot(cored_points, pch=20, add=TRUE)
+plot(cored_points, pch=20, col = species_colors[species$sp_fact], add=TRUE)
 dev.off()
 
[…]

### 2019-08-13 scripts/canopy_position_analysis.R
Added stats table with regard to #51

@@ -1100,6 +1100,24 @@ x1999 <- trees_all_sub[trees_all_sub$year == 1999, ]
 
 model_df <- list(trees_all_sub, x1966, x1977, x1999)
 names(model_df) <- c("trees_all_sub", "x1966", "x1977", "x1999")
+
+##5m. get base stats for everything from trees_all_sub ####
+trees_all_sub$dbh.cm <- exp(trees_all_sub$dbh.ln.cm)
+trees_all_sub$height.m <- exp(trees_all_sub$height.ln.m)
+
+stats <- NULL
+for(i in seq(along=trees_all_sub[,c(5,7:10,16,18,19)])){
+  name_trait <- trees_all_sub[,c(5,7:10,16,18,19)][i]
+  tree_stats <- 
+    trees_all_sub %>%
+    summarize(median = median(trees_all_sub[, which(colnames(trees_all_sub) == colnames(name_trait))]),
+              min = min(trees_all_sub[, which(colnames(trees_all_sub) == colnames(name_trait))]),
+              max = max(trees_all_sub[, which(colnames(trees_all_sub) == colnames(name_trait))])
+              )
+  tree_stats$trait <- colnames(name_trait)
+  stats <- rbind(stats, tree_stats)
+}
+
 ##############################################################################################
 #6. mixed effects model for output of #5. ####
 library(lme4)
@@ -1350,52 +1368,41 @@ for (i in seq(along=c(1:4))){
       
       #get all mods <2 dAIC
       var_aic <- var_aic[var_aic$Delta_AICc <= 2, ]
+      var_aic$mod_no <- rownames(var_aic)
       top <- var_aic[,c(1,4)]
       top$Delta_AICc <- round(top$Delta_AICc, 2)
       top$scenario <- mods[[i]]
       top$coef <- NA
       
       for (z in seq(along = lmm_all)){
-        if (names(lmm_all[z]) == var_aic$Modnames[[1]]){
-          
-          #run the best model alone with REML=TRUE
-          fit1 <- lmer(formula_vec[[z]], data = model_df[[j]], REML=TRUE, 
-                      control = lmerControl(optimizer ="Nelder_Mead"))
-          
-          #get coefficients and put in table
-          coeff <- data.frame(coef(summary(fit1))[ , "Estimate"]) ##2
-          coeff[,2] <- rownames(coeff)
-          coeff[,1] <- round(coeff[,1], 3)
-          colnames(coeff) <- c(names(model_df[j]), "model_var")
-
-          #put r2 in table
[…]

### 2019-08-13 scripts/graphs_for_manuscript.R
Changed Figure 3 to S3, added colors #48

@@ -247,7 +247,7 @@ cored_points <- SpatialPointsDataFrame(data.frame(species$NAD83_X, species$NAD83
 
 plot.new()
 
-png("manuscript/tables_figures/Figure3.png", width=5, height=7, units="in", res=72)
+png("manuscript/tables_figures/FigureS3.png", width=5, height=7, units="in", res=300)
 plot(topo, axes=FALSE, box=FALSE, 
      legend.args = list(text="Topographic Wetness Index", side=4, font=2, line=2.5, cex=0.8))
 plot(cored_points, pch=20, col = species_colors[species$sp_fact], add=TRUE)

### 2019-08-14 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
update manuscript, added SI template
changed title

@@
 ---
-title: "Tree size, microhabitat, and hydraulic traits (interactively) shape drought response in a temperate broadleaf forest"
+title: "Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest"
 author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"

### 2019-08-14 manuscript/climate_sensitivity_draft.tex
update manuscript, added SI template
changed title

@@
 \hypersetup{unicode=true,
-            pdftitle={Tree size, exposure, and hydraulic traits interactively shape drought response in a temperate broadleaf forest},
+            pdftitle={Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest},
             pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira},
@@
 \bibliographystyle{apalike}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx,grffile}
@@
 % Create subtitle command for use in maketitle
-\providecommand{\subtitle}[1]{
+\newcommand{\subtitle}[1]{
   \posttitle{
@@
 
-  \title{Tree size, exposure, and hydraulic traits interactively shape drought
-response in a temperate broadleaf forest}
+  \title{Tree size and hydraulic traits shape drought responses in a temperate
+broadleaf forest}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 \captionsetup[figure]{font=small}
+\captionsetup[table]{labelformat=empty}
 \usepackage{dcolumn}
@@
 
-\hypertarget{abstract}{%
-\subsubsection{Abstract}\label{abstract}}
+\subsubsection{Summary}\label{summary}
 
-\emph{content from AGU abstract:} Here, we integrate forest census data
-from a 25.6-ha ForestGEO plot in Virginia (USA), tree-ring records for
-12 species representing \#\#\% of woody productivity, leaf hydraulic
-trait measurements, and microhabitat data. Individual-level growth
-responses to three droughts (1964-66, 1977, 1999) were stronger among
-taller trees in dominant canopy positions, those in wetter microsites,
-and for more drought-sensitive species as assessed by leaf traits
-(turgor loss at less negative leaf water potential, greater shrinkage
-with leaf dehydration), with substantial variation in the best predictor
-variables across given droughts. We conclude that when droughts occur,
-large dominant trees, drought sensitive species, and individuals in
-moist microhabitats are likely to be most strongly affected.
+Predicting forest responses to drought is an increasingly critical task
+under climate change effects. Part of the problem is due to the lack of
+studies analyzing the confluence of leaf hydraulic traits with
+biophysical parameters. In this study, we analyze the interaction
+between these two trait groups using forest census data from a 25.6-ha
+ForestGEO plot in Virginia (USA). Drought periods were defined by both
+Palmer Drought Severity Indices (PDSI) and their identification from
+tree-ring records for 12 species representing 97\% of woody
+productivity. Each drought scenario (1964-66, 1977, 1999), along with
+the overall trend, was then tested against leaf hyraulic trait
+measurements and microhabitat biophysical data. Individual-level growth
+responses to the three individual droughts were stronger among taller
+trees in dominant canopy positions, those in wetter microsites, and for
+more drought-sensitive species as assessed by leaf traits (turgor loss
+at less negative leaf water potential, greater shrinkage with leaf
+dehydration), with substantial variation in the best predictor variables
+across given droughts. We conclude that when droughts occur, large
+dominant trees, drought-sensitive species, and individuals in wetter
+microhabitats are likely to be most strongly affected.
 
-\hypertarget{introduction}{%
-\subsubsection{Introduction}\label{introduction}}
+\emph{The Summary for research papers, which must be usable as a stand-
+alone document, must not exceed 200 words and should be organized using
+four bullet points to indicate: (1) the research conducted, including
+the rationale, (2) methods, (3) key results, and (4) the main
+conclusion, including the key points of discussion. It should not
+contain citations of other papers.}
+
+\subsubsection{Introduction}\label{introduction}
 
 Understanding how and why trees respond to drought is critical to
-predicting forest drought responses and climate change feedbacks.
+predicting forest drought responses and climate change feedbacks. (1
+paragraph on this- I have lots of content on this and will add later.)
 
-Forests are diverse in terms of tree sizes and functional traits, and it
-is known that trees varying in size and functional traits respond
-differently to drought (e.g., \citep{bennett_larger_2015}; REFS).
-Therefore, in order to understand whole-forest response to drought, we
-need to know how responses vary by tree size/ species. To do so, there
-are four fundamental questions that must be addressed:
+{[}Understanding forest responses to drought requires increased
+functional understanding of the factors that confer individual-level
+vulnerability or resistance.{]} Forests are diverse in terms of tree
+sizes and functional traits, and it is known that trees varying in size
+and functional traits respond differently to drought (e.g.,
+\citep{bennett_larger_2015}; REFS). Therefore, in order to understand
+whole-forest response to drought, we need a functional understanding of
+how responses vary by tree size, microhabitat, and species. There are 3
+fundamental questions that must be addressed:
 
-\emph{First, how significant is the effect of individual drought years?}
-Droughts are rarely explicitly defined in ecological studies
-\citep{slette_how_2019}, yet no two droughts are the same. This study
-addressed this by analyzing trees' resistance to drought within and
-across three defined drought periods.
-
-\emph{Second, what drives the observed tendency for large trees to
-suffer more during drought?}\\
-\citep{bennett_larger_2015} showed that in forests globally, large trees
+\emph{First, what drives the observed tendency for large trees to suffer
+more during drought?}\\
+\citet{bennett_larger_2015} showed that in forests globally, large trees
 suffer greater growth reductions during drought. However, this analysis
 quantified tree size based on DBH, which has no direct mechanistic
-meaning. This study proposed three major mechanisms (besides insects):
-(1) inherently greater biophysical challenge of being tall; (2) greater
-exposure of the crowns of large trees; and (3) greater water
-availability. It is also expected that roots play a role, though these
-hypotheses still need to be tested.
+meaning. This study proposed two major mechanisms (besides insects) for
+the observed greater drougth growth reductions of large trees: (1)
+inherently greater biophysical challenge of being tall; (2) greater
+exposure of the crowns of large trees. Counteracting these effects, (3)
+the larger root systems of larger trees may confer an advantage in terms
+of allowing greater access to water, but it appears that this effect is
+usually insufficient to offset the costs of height and/or crown
+exposure.
 
-\emph{Third, how do species' traits influence drought response?}
-Analyzing drought responses on the species level does not fully explain
-mechanisms and is not feasible in diverse forests. The solution is a
-trait-based approach. Leaf hydraulic traits hold more promise than more
-commonly/traditionally-measured traits such as wood density and SLA
-(Medeiros et al.).
+Canopy trees have lower drought resistance because they are exposed to
+higher solar radiation, greater wind speeds, and lower humidity.
+Alternatively, the generally supressed status of subcanopy trees may be
+insufficient to override the benefits of their buffered environment
+during drought.
 
-\emph{Fourth, how do tree size and functional traits interact to
-influence drought response?} It is possible that the pattern observed by
-\citep{bennett_larger_2015} could be caused by smaller trees being more
-drought resistant. Alternatively, larger trees may have more
-drought-resistant traits as adpatations to greater biophysical
-challenges.
+\emph{Second, how do species' traits - alone and in interaction with
+tree size - influence drought response?} Analyzing drought responses on
+the species level does not fully explain mechanisms and is not feasible
+in diverse forests. The solution is a trait-based approach. Leaf
+hydraulic traits hold more promise than more
+commonly/traditionally-measured traits such as wood density and specific
+leaf area (SLA)
+(\href{https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229}{Medeiros
+et al. 2019}).
 
-\emph{Hypotheses}
+Commonly measured traits including wood density (REFS), leaf mass per
+area \citep{abrams_adaptations_1990} \citep{guerfel_impacts_2009}, and
+ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been
+linked to drought responses and likely correlated with drought
+resistance in this forest. However, we hypothesize that leaf hydraulic
+traits such as leaf area shrinkage upon dessication (PLA) and turgor
+loss point (TLP), which are emerging as potentially more informative
+traits but whose effect on drought resistance has never been tested
+(CONFIRM), will prove better predictors.
 
-\textbf{1. How significant are individual drought years on a long
-time-frame trend?} 1.1 Individual drought scenario is a strong predictor
-of drought stress.
+Moreover, there may be an interaction between traits and tree size. It
+is possible that the pattern observed by \citet{bennett_larger_2015}
+could be caused by smaller trees being more drought resistant.
+Alternatively, larger trees may have more drought-resistant traits as
+adpatations to greater biophysical challenges.
 
-\begin{verbatim}
-* P1.1 - Drought stress is proportional to the severity of each drought.
-\end{verbatim}
+\emph{Third, are responses similar or variable across individual drought
+years?} Droughts are rarely explicitly defined in ecological studies
+\citep{slette_how_2019}, yet no two droughts are the same.
 
-\textbf{2. Why do larger trees suffer greater growth declines during
-drought?} Our forest displays the same trend as most forests globally
-\citep{bennett_larger_2015}. (Note that Bennett et al.~2015 identified
-only one study on tree growth responses to drought in the Eastern US
-temperate deciduous biome. We know little about how tree size shapes
-drought response in this biome.)
+We need to understand the the factors confirming drought vulnerability
+or resistance not only for extreme droughts with dramatic impacts on
+tree growth and mortality, which tend to dominate the literature
+\citep{bennett_larger_2015} (REFS), but also for more modest but
+frequent droughts--e.g., those with historical return intervals on the
+order of a decade.
 
-2.1 DBH is a strong predictor of drought stress.
+Here, we combine tree-ring records covering three droughts (1964-66,
+1977, 1999), species functional and hydraulic trait measurments, and
+forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to
+test a series of hypotheses and associated specific predictions (Table
+1) designed to yield functional understanding of how tree size,
+microenrivonment, and species' traits collectively shape drought
+responses. First, we focus on the role of tree size and its interaction
+with microenvironement. We confirm that, consistent with most forests
+globally \citep{bennett_larger_2015}, larger-diameter trees have lower
+drought resistance in this forest, which is in an ecoregion represented
+by only one study in Bennett et al (2015) (H1.0). We then test
+hypotheses designed to disentangle the relative importance of tree
+height (H1.1), crown exposure (H1.2), and root water access, which
+should be greater for larger trees in dry but not perpetually wet
+microsites (H1.3). Second, we focus on the role of species' functional
+and hydraulic traits and their interaction with tree height. We
+hypothesize that drought resistance will follow predicted and observed
+patterns in relation to wood density, specific leaf area, and ring
+porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area
+shrinkage upon dehydration and turgor loss point will prove better
+predictors (P2.1d-e). We then test whether these traits correlate with
+tree height (P2.2), potentially driving the observed tendency for taller
+trees to suffer more during drought (P2.3). Finally, we focused on
+variability among droughts, asking how community resistance varied
+across droughts (H3.1) and whether the factors confirming vulnerability
+or resistance varied across droughts (H3.2).
 
-\begin{verbatim}
-* P2.2-Drought response increases with DBH at time of drought.
-\end{verbatim}
+\emph{Let's put Table 1 here. We need to give the hypotheses/
+predictions here. It seems counterintuitive to put the results right up
+front like that, but it actually tends to be a higher-impact style of
+writing, and recommended
+\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619}{here}.}
 
-2.2. Height is a strong predictor of drought stress.
+\textbf{Table 1. Summary of hypotheses, predictions and results}
+\includegraphics[width=5.20833in]{tables_figures/Table1.png}
 
-\begin{verbatim}
-* P2.2-Drought response increases with height at time of drought.
-* P2.2a - Drought response is better predicted by height than DBH.
-\end{verbatim}
+We count predictions as fully supported (or rejected) when the direction
+of response matches (or contradicts) the prediction. H2.3-H3.2 are based
+on if models containing species traits in H2.1 had dAIC\textgreater{}=2
+relative to the appropriate null model or to any alternative
+multivariate model within 2 dAIC. Parentheses indicate that predictions
+were partially supported (or rejected). In other words, parentheses
+indicate when the direction of response matched (or contradicted) the
+prediction in some but not all models with dAIC\textless{}2 relative to
+the appropriate null model. With categorical variables such as crown
+position, a ``(yes)'' notation describes when the trend matched the
+prediction, but it wasn't significant A ``(yes/no) or (''no/yes``)''
+indicates tendencies in opposite directions in univariate tests versus
+the best full models, respectively. ``(yes)
 
-2.3. Large trees suffer more during drought because of greater exposure
-(to radiation, wind, etc.)--either in relation to neighboring trees or
-because of position on landscape.
-
-\begin{verbatim}
-* P3.1- Trees currently in a dominant canopy position suffered more during drought. 
-\end{verbatim}
-
-2.4. Rooting volume/depth relative to water sources are critical in
-drought response. Effects of drought on larger trees are mediated by the
-fact that large trees have better access to water.
-
-\begin{verbatim}
- * P3.2- drought response increases with topographic wetness index
-\end{verbatim}
-
-\textbf{3. Do species functional traits predict drought response?}
-
-\begin{verbatim}
- * P3.1 - diffuse porous species are more sensitive than ring porous (previously observed in eastern dec forests- Elliot et al. 2015, Friedrichs et al. 2009)
- * P3.2 - higher percent leaf area predicts higher drought stress
- * P3.3 - higher leaf mass area correlates positively to drought resistance (more sclerophyllous vegetation [thick leaves] usually means more adaptation) [Abrams 1990] [Guerfel et al 2009]
- * P3.4 - TLP correlates negatively with drought resistance (NEVER tested),
- * P3.5 - higher wood density means more drought resistant
-\end{verbatim}
-
-\textbf{4. How are functional traits distributed across size classes,
-and how does this affect size-resistance relationship?}
-
-4.1. Larger/ more exposed trees have more drought resistant traits,
-meaning size effects are buffered by traits\\
-4.1\_alt. Larger trees suffer more because they have more drought
-vulnerable traits.
-
-\begin{verbatim}
- * P3.1a- TLP is lower (larger negative) in taller/canopy trees
- * P3.1b- diffuse porous species more common in understory
-
- * P3.2- Inclusion of TLP / rp in model does not eliminate or significantly reduce effect of tree size.
-\end{verbatim}
-
-\textbf{I think for \#4 here the focus should be on hypotheses looking
-at the overall trend compared to the specific scenarios?}
-
-\textbf{4. How do droughts vary in their affect on tree resistance? }
-
-\begin{verbatim}
-* P4.1 - The combined-years scenario will have more explaining variables than the individual scenarios.
-* P4.2 - Inclusion of leaf hydraulic traits does not eliminate or significantly reduce effect of tree size.
-* P4.3 - Drought resistance for each scenario will be explained more by hydraulic traits than by the biophysical environment
-\end{verbatim}
-
-\hypertarget{methods}{%
-\subsubsection{Methods}\label{methods}}
+\subsubsection{Materials and Methods}\label{materials-and-methods}
 
@@
 \emph{Data collection and preparation} The SCBI ForestGEO plot was
-censused in 2008, 2013, and 2018 following standard ForestGEO
-protocols\ldots{} {[}DETAILS{]}. Here, we use \ldots(\emph{very brief
-summary of which data we use for what purpose-- but some will be better
-integrated with methods descriptions below})
+censused in 2008, 2013, and 2018 following standard ForestGEO protocols,
+whereby all free-standing woody stems \textgreater{}= 1cm diameter at
+breast height (DBH) were mapped, tagged, measured at DBH, and identified
+to species \citep{condit_tropical_1998}. From this census data, we used
+measurements of DBH from 2008 to calculate historical DBH (described
+below), along with tree location in the plot to determine the
+topographic wetness index. Furthermore, we used data for all stems
+\textgreater{}= 10cm to analyze functional trait composition relative to
+tree height. \emph{I added the last sentence-- not sure if you've
+described that below yet. I don't think you need to note here which data
+we use, so long as that's described in the portions where you describe
+the analysis.}
 
 We analyzed tree-ring data from the twelve species contributing most to
-woody aboveground net primary productivity (ANPP\_stem) (Table 1), which
-together comprised 97\% of whole-ecosystem ANPP\_stem between 2008 and
-2013. Cores were collected in 2010-2011 or 2016-2017 from a breast
-height of 1.3m using a 5mm increment borer. In 2010-2011, cores were
-collected from randomly selected live trees were selected at random from
-each species in 2010-2011, with at least 30 of those trees having a
-diameter at breast height (DBH) of at least 10cm
+woody aboveground net primary productivity (ANPP), which together
+comprised 97\% of study plot ANPP between 2008 and 2013
+\citep{helcoski_growing_2019}. Cores were obtained in 2010-2011 or
+2016-2017 from a breast height of 1.3m using a 5mm increment borer. In
+2010-2011, cores were collected from randomly selected live trees of
+species with at least 30 individuals of DBH \textgreater{}= 10cm
 \citep{bourg_initial_2013}. In 2016-2017, cores were collected from all
-dead trees found in the annual mortality census
+trees found dead in the annual mortality census
 \citep{gonzalezakre_patterns_2016}. Cores were sanded, measured, and
@@
 \citep{helcoski_growing_2019}. The resulting chronologies have been
-published in association with \citep{helcoski_growing_2019}: (GitHub
-URL), (ITRDB).
+published in association with \citet{helcoski_growing_2019}: (ITRDB;
+GitHub/Zenodo). \emph{is this a database? Yes. Ryan submitted the data
+but I don't think its posted yet. We should also cite GitHub/Zenodo
+here. I'll come back to that. }
 
 Height measurements (n=\# trees) were taken by several researchers
-between 2012 to 2019, and are archived in a public GitHub repository
-(\emph{GitHUB URL}). Measurement methods included manual
+between 2012 to 2019, and are archived in a public
+\href{https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights}{GitHub
+repository}. Measurement methods included manual
 \citep[NEON]{stovall_assessing_2018}, digital rangefinders
@@
 
-\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depth_YEAR) \]
+\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) \]
 
@@
 
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=5.20833in]{tables_figures/Figure3.png}
+\caption{Location of cored trees}
+\end{figure}
+
 Hydraulic traits were collected from SCBI and are summarized in Table 1.
@@
 species \ldots{} (\textbf{Nobby's description of methods for the
-following}) 1. PLA 2. LMA 4. Wood density 5. TLP
+following (see word document}) 1. PLA 2. LMA 4. Wood density 5. TLP
 
-\textbf{Table 1. Species analyzed here, listed in descending order of
+\textbf{Table 2. Species analyzed here, listed in descending order of
 ANPP\_stem. n cores and DBH range represented, and species traits}
@@
 Suggested columns, with those to include only if they fit in
-parentheses: species, (stems \textgreater=10 cm per ha in plot),
+parentheses: species, (stems \textgreater{}=10 cm per ha in plot),
 (ANPP\_stem), n cores, DBH range of cores, (n cores in each crown
@@
 
-\begin{table}[H]
-\centering
-\begin{tabular}{l|r|r|r|r|r|r}
-\hline
-sp & n\_cores & dominant & co-dominant & intermediate & suppressed & prior dead\\
-\hline
-caco & 13 & NA & 2 & 5 & 5 & 1\\
-\hline
-cagl & 31 & 1 & 8 & 16 & 5 & 1\\
-\hline
-caovl & 23 & 4 & 5 & 12 & 2 & NA\\
-\hline
-cato & 13 & NA & NA & 6 & 2 & 5\\
-\hline
-fagr & 80 & NA & 7 & 48 & 25 & NA\\
-\hline
-fram & 62 & NA & 17 & 19 & 14 & 12\\
-\hline
-juni & 31 & NA & 21 & 8 & NA & 2\\
-\hline
-litu & 98 & 9 & 29 & 25 & 30 & 5\\
-\hline
-qual & 61 & 4 & 34 & 20 & 3 & NA\\
-\hline
-qupr & 59 & 1 & 26 & 20 & 12 & NA\\
-\hline
-quru & 69 & 6 & 36 & 23 & 2 & 2\\
-\hline
-quve & 77 & 6 & 46 & 22 & 1 & 2\\
-\hline
+\begin{verbatim}
+## Warning: package 'kableExtra' was built under R version 3.4.4
+\end{verbatim}
+
+\textbackslash{}begin\{table\}{[}!h{]}
+
+\textbackslash{}caption\{\label{tab:Table2}Overview of analyzed species,
+detailing DBH mean and range of cored trees, the number of cores
+represented by each crown position of each species, and mean hydraulic
+trait measurements. Units of measurements are in mm (DBH), \% (PLA),
+g/m2 (LMA), MPa (TLP), and g/cm3 (WD).\} \centering
+
+\begin{tabular}{lrrlrrrr}
+\toprule
+sp & mean\_DBH & range\_DBH & RP & PLA & LMA & TLP & WD\\
+\midrule
+caco & 271.87 & 508.0 & ring & 17.22 & 45.86 & -2.13 & 0.83\\
+cagl & 313.89 & 887.0 & ring & 21.09 & 42.76 & -2.13 & 0.62\\
+caovl & 352.87 & 511.0 & ring & 14.80 & 47.60 & -2.48 & 0.96\\
+cato & 209.74 & 201.1 & ring & 16.56 & 45.36 & -2.20 & 0.83\\
+fagr & 235.11 & 960.0 & diffuse & 9.45 & 30.68 & -2.57 & 0.62\\
+\addlinespace
+fram & 353.63 & 883.3 & ring & 13.06 & 43.28 & -2.10 & 0.56\\
+juni & 481.42 & 628.0 & semi-ring & 24.64 & 72.13 & -2.76 & 1.09\\
+litu & 368.54 & 904.0 & diffuse & 19.56 & 46.92 & -1.92 & 0.40\\
+qual & 471.51 & 677.0 & ring & 8.52 & 75.80 & -2.58 & 0.61\\
+qupr & 422.48 & 767.0 & ring & 11.75 & 71.77 & -2.36 & 0.61\\
+\addlinespace
+quru & 548.79 & 1369.3 & ring & 11.01 & 71.13 & -2.64 & 0.62\\
+quve & 541.38 & 981.8 & ring & 13.42 & 48.69 & -2.39 & 0.65\\
+\bottomrule
 \end{tabular}
-\end{table}
-\begin{table}[H]
-\centering
-\begin{tabular}{l|l|r|r|r}
-\hline
-Trait & Unit & mean & min & max\\
-\hline
-Ring Porosity & ring, semi-ring, diffuse & NA & NA & NA\\
-\hline
-Percent Leaf Area & \% & 15.09 & 8.52 & 24.64\\
-\hline
-Leaf Mass Area & g/m2 & 53.50 & 30.68 & 75.80\\
-\hline
-Wood density & g/cm3 & 0.70 & 0.40 & 1.09\\
-\hline
-TLP & MPa & -2.36 & -2.76 & -1.92\\
-\hline
-\end{tabular}
-\end{table}
+
+\textbackslash{}end\{table\}
 
@@
 as the ratio between the performance during and before the disturbance.
-Candidate drought years were defined if \textgreater50\% of the cored
-trees experienced \textless30\% growth in a year compared to the
+Candidate drought years were defined if \textgreater{}50\% of the cored
+trees experienced \textless{}30\% growth in a year compared to the
 previous 5 years. These were then cross-validated with the regional
@@
 
-\textbf{Figure 1. Time series of peak growing season (May-August)
-climate conditions and residual chronologies for each species.} Droughts
-analyzed here are indicated by dashed lines, and shading indicates the
-pre-drought period used in calculations of the resistance metric. Figure
-modified from \citep{helcoski_growing_2019}.
-\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Time_series_for_each_species.jpg}
+\subsubsection{Results}\label{results}
 
-\hypertarget{results}{%
-\subsubsection{Results}\label{results}}
+Results outline
+
+Descriptions of Droughts
+
+\begin{itemize}
+\tightlist
+\item
+  physical characteristics of drought: duration, min PDSI (and when
+  occurred), see slette et al 2019 to see how they suggest reporting
+  each one
+\item
+  sum nb.series for each scenario, mult by perc.neg (pointers df),
+  define thresholds (e.g.~75\% of all trees experienced the growth
+  reduction of 40\% or more)
+\item
+  make table for this
+\item
+  citing table 3 (the coefficient for individual year) - maybe mention
+  how 77 and 99 have neg coefficients, indicating potential height
+  effect: as time goes on, trees are more affected by drought anyways
+  because they grow taller
+\item
+  this is followed by Figure 2
+\end{itemize}
+
+Hypothesis results - one header for each main question - present results
+for each one, walk through via order and logic of Table 1 - no citations
+here
 
@@
 individually against a null model containing height and year. This
-yielded Figure ??? (cand\_full). All variables with dAICc \textgreater2
-were used as candidates for each scenario's best model (figure ????
-(tested\_traits\_best))
+yielded Figure ??? (cand\_full). All variables with dAICc
+\textgreater{}2 were used as candidates for each scenario's best model
+(figure ???? (tested\_traits\_best))
 
-\begin{landscape}\begin{table}[t]
+\textbf{Describe each individual drought scenario } 1964-1966 1977 1999
 
-\caption{\label{tab:Table 1}All variables tested for climate sensitivity against appropriate null models, in line with hypothesis predictions.
-Each variable was tested for each drought scenario excepting "year" }
+\textbf{Figure 1. Time series of peak growing season (May-August)
+climate conditions and residual chronologies for each species.} Droughts
+analyzed here are indicated by dashed lines, and shading indicates the
+pre-drought period used in calculations of the resistance metric. Figure
+modified from \citep{helcoski_growing_2019}.
+
+\begin{figure}[htbp]
 \centering
-\begin{tabular}{rllllllrllrllrllrll}
-\toprule
-prediction & variable & variable\_description & null\_model & tested\_model & null\_model\_year & tested\_model\_year & dAIC\_all & coef\_all & coef\_var\_all & dAIC\_1964.1966 & coef\_1964.1966 & coef\_var\_1964.1966 & dAIC\_1977 & coef\_1977 & coef\_var\_1977 & dAIC\_1999 & coef\_1999 & coef\_var\_1999\\
-\midrule
-1.1 & year & drought.year & resist.value \textasciitilde{} (1|sp/tree) & resist.value \textasciitilde{} (1|sp/tree)+year & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp) & 35.834 & - & year1977 (-0.09), year1999 (-0.081) & 0.000 & NA & NA & 0.000 & NA & NA & 0.000 & NA & NA\\
-2.1 & dbh.ln.cm & ln[DBH] & resist.value \textasciitilde{} year+(1|sp/tree) & resist.value \textasciitilde{} year+(1|sp/tree)+dbh.ln.cm & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp)+dbh.ln.cm & 7.411 & - & dbh.ln.cm (-0.035) & 18.276 & - & dbh.ln.cm (-0.082) & -0.947 & - & dbh.ln.cm (-0.021) & -1.918 & + & dbh.ln.cm (0.006)\\
-2.2 & height.ln.m & ln[height] & resist.value \textasciitilde{} year+(1|sp/tree) & resist.value \textasciitilde{} year+(1|sp/tree)+height.ln.m & resist.value \textasciitilde{} (1|sp) & resist.value \textasciitilde{} (1|sp)+height.ln.m & 7.591 & - & height.ln.m (-0.058) & 17.788 & - & height.ln.m (-0.133) & -1.077 & - & height.ln.m (-0.032) & -2.022 & + & height.ln.m (0.002)\\
-3.3 & position\_all & crown.position & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+position\_all & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+position\_all & 1.682 & - & position\_alldominant (-0.044), position\_allintermediate (-0.042), position\_allsuppressed (-0.053) & -2.548 & - & position\_alldominant (-0.055), position\_allintermediate (0.007), position\_allsuppressed (-0.044) & -0.700 & - & position\_alldominant (-0.076), position\_allintermediate (-0.033), position\_allsuppressed (0.029) & 4.087 & - & position\_alldominant (-0.003), position\_allintermediate (-0.082), position\_allsuppressed (-0.102)\\
-3.4 & TWI & topographic.wetness.index & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+TWI & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+TWI & 2.631 & - & TWI (-0.009) & -0.563 & + & TWI (0.009) & 5.264 & - & TWI (-0.02) & 3.037 & - & TWI (-0.015)\\
-\addlinespace
-4.1 & rp & ring.porosity & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+rp & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+rp & -3.553 & + & rpring (0.04), rpsemi-ring (0.013) & -2.161 & + & rpring (0.101), rpsemi-ring (0.015) & 0.895 & - & rpring (-0.19), rpsemi-ring (-0.147) & 4.083 & + & rpring (0.2), rpsemi-ring (0.151)\\
-4.2 & PLA\_dry\_percent & percent.leaf.area & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+PLA\_dry\_percent & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+PLA\_dry\_percent & 4.413 & - & PLA\_dry\_percent (-0.011) & 5.825 & - & PLA\_dry\_percent (-0.016) & -0.190 & - & PLA\_dry\_percent (-0.01) & -0.701 & - & PLA\_dry\_percent (-0.007)\\
-4.3 & LMA\_g\_per\_m2 & leaf.mass.area & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+LMA\_g\_per\_m2 & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+LMA\_g\_per\_m2 & -1.895 & + & LMA\_g\_per\_m2 (0.001) & -1.075 & + & LMA\_g\_per\_m2 (0.002) & -1.698 & - & LMA\_g\_per\_m2 (-0.001) & -1.985 & + & LMA\_g\_per\_m2 (0)\\
-4.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+mean\_TLP\_Mpa & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+mean\_TLP\_Mpa & 4.580 & - & mean\_TLP\_Mpa (-0.207) & 1.352 & - & mean\_TLP\_Mpa (-0.217) & 1.008 & - & mean\_TLP\_Mpa (-0.236) & 0.132 & - & mean\_TLP\_Mpa (-0.177)\\
-4.5 & WD\_g\_per\_cm3 & wood.density & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree) & resist.value \textasciitilde{} height.ln.m+year+(1|sp/tree)+WD\_g\_per\_cm3 & resist.value \textasciitilde{} height.ln.m+(1|sp) & resist.value \textasciitilde{} height.ln.m+(1|sp)+WD\_g\_per\_cm3 & -2.018 & + & WD\_g\_per\_cm3 (0.005) & -1.960 & - & WD\_g\_per\_cm3 (-0.049) & -1.236 & - & WD\_g\_per\_cm3 (-0.175) & 0.171 & + & WD\_g\_per\_cm3 (0.247)\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
+\includegraphics[width=5.20833in]{tables_figures/Figure1.jpg}
+\caption{Time series of combined tree cores by species}
+\end{figure}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(stargazer)}
-\end{Highlighting}
-\end{Shaded}
+\textbf{Results for first main question: what drives the observed
+tendency for large trees to suffer more during drought?} H1.0, H1.1,
+H1.2, H1.3 DBH, height, crown position, and TWI
 
-\begin{verbatim}
-## 
-## Please cite as:
-\end{verbatim}
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=5.20833in]{tables_figures/Figure2.png}
+\caption{Height profile graphs}
+\end{figure}
 
-\begin{verbatim}
-##  Hlavac, Marek (2018). stargazer: Well-Formatted Regression and Summary Statistics Tables.
-\end{verbatim}
+\textbf{Results for second main question: how do species' traits - alone
+and in interaction with tree size - influence drought response? } H2.1,
+H2.2, H2.3 Hydraulic traits alone, traits with height
 
-\begin{verbatim}
-##  R package version 5.2.2. https://CRAN.R-project.org/package=stargazer
-\end{verbatim}
+\includegraphics{tables_figures/Table4.png} \#this is
+tested\_traits\_all
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{stargazer}\NormalTok{(traits_test, }
-          \DataTypeTok{float.env =} \StringTok{"sidewaystable"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\textbf{Results for third main question: are responses similar or
+variable across individual drought years?} H3.1 and H3.2 Combining
+biophysical with hydraulic traits, which come out as candidates for best
+model?
 
-\% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard
-University. E-mail: hlavac at fas.harvard.edu \% Date and time: Thu, Aug
-01, 2019 - 12:30:35 PM \% Requires LaTeX packages: rotating
+\subsubsection{Discussion}\label{discussion}
 
-\begin{sidewaystable}[!htbp] \centering 
-  \caption{} 
-  \label{} 
-\begin{tabular}{@{\extracolsep{5pt}}lccccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
-Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
-\hline \\[-1.8ex] 
-prediction & 10 & 3.360 & 1.182 & 1.100 & 2.475 & 4.275 & 4.500 \\ 
-dAIC\_all & 10 & 5.668 & 11.284 & $-$3.553 & $-$1.001 & 6.703 & 35.834 \\ 
-dAIC\_1964.1966 & 10 & 3.493 & 8.032 & $-$3 & $-$1.7 & 4.7 & 18 \\ 
-dAIC\_1977 & 10 & 0.132 & 2.010 & $-$2 & $-$1.0 & 0.7 & 5 \\ 
-dAIC\_1999 & 10 & 0.488 & 2.414 & $-$2 & $-$1.6 & 2.3 & 4 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{sidewaystable}
+Discussion outline
 
-\begin{table}[!h]
-
-\caption{\label{tab:Table 2}Candidate variables to be included in full model, chosen by dAICc>2 when individually tested against a null model}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}{rlll}
-\toprule
-prediction & variable & variable\_description & top\_model\\
-\midrule
-1.1 & year & drought.year & all\\
-2.2 & height.ln.m & ln[height] & all\\
-2.2 & height.ln.m & ln[height] & 1966\\
-3.3 & position\_all & crown.position & 1999\\
-3.4 & TWI.ln & topographic.wetness.index & all\\
-\addlinespace
-3.4 & TWI.ln & topographic.wetness.index & 1977\\
-3.4 & TWI.ln & topographic.wetness.index & 1999\\
-4.1 & rp & ring.porosity & 1999\\
-4.2 & PLA\_dry\_percent & percent.leaf.area & all\\
-4.2 & PLA\_dry\_percent & percent.leaf.area & 1966\\
-\addlinespace
-4.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & all\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\begin{table}[!h]
-
-\caption{\label{tab:Table 3}Best full models for each drought scenario}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}{lrl}
-\toprule
-best\_model & r2 & scenario\\
-\midrule
-resist.value \textasciitilde{} year+height.ln.m+position\_all+TWI+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.13 & all droughts\\
-resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.24 & 1964-1966\\
-resist.value \textasciitilde{} TWI+rp+mean\_TLP\_Mpa+(1|sp) & 0.21 & 1977\\
-resist.value \textasciitilde{} height.ln.m+position\_all+TWI+rp+PLA\_dry\_percent+(1|sp) & 0.25 & 1999\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\hypertarget{discussion}{%
-\subsubsection{Discussion}\label{discussion}}
+\begin{itemize}
+\tightlist
+\item
+  first paragraph is summary of main findings, not re-presenting the
+  results
+\item
+  ``we supported this hypothesis, but not this one'' (following same
+  order as Table 1 still)
+\item
+  direction of responses seems to be relatively consistent but the
+  individual responses vary
+\item
+  tie things in (see here
+\item
+  main thing is that we're now better understanding what confers
+  vulnerability or resilience on trees during drought - how forests
+  respond to future droughts/climate change
+\item
+  ``we filled the gap'' = 1-2 sentences
+\item
+  limitations at our site:
+\item
+  aren't able to analyze historical forest community, nor trees that
+  were killed by these droughts {[}aka we don't have data for
+  individuals that were most severely affected{]} (though we found that
+  there's little variation in climate sensitivity for trees that were
+  cored dead vs cored alive).
+\item
+  p50/p80
+\item
+  We used crown position despite its uncertainty. However, height and
+  crown position change relatively slowly so it shouldn't be that far
+  off.
+\item
+  We did not use crowding index because it has much more uncertainty
+\item
+  limitations for extrapolating other sites:
+\item
+  forests are different from place to place, but we've seen how in
+  forests around the world, forests tend to suffer more. We've
+  identified some facets for why this happens. (cite other studies)
+\item
+  has been observed elsewhere that individuals in more moist habitats
+  are more susceptible to drought (cite other studies)
+\item
+  the species may be different from other sites but the hydraulic traits
+  are general across species;
+\item
+  next paragraph saying that our study advances understanding of
+  droughts, e.g.~bennett et al 2015 doesn't go into mechanisms but we do
+\item
+  and we show that height is more important than exposure, but doesn't
+  eliminate the effect of it
+\item
+  further contribution is we show that two leaf hydraulic traits -
+  relatively easy to measure - are good predictors of drought response,
+  can be helpful for scaling up (e.g.~from our site to eastern deciduous
+  biome)
+\item
+  biophysical mechanisms are things that should be seen as universal but
+  relative importance of each can vary within each drought
+\item
+  first study to show how these traits affect woody growth response to
+  drought \textbf{confirmed by Lawren}
+\item
+  science is better now. by advancing our understanding of the
+  mechanisms for individual-level responses to drought, this opens door
+  for better predictions (elaborate from above). ``This is absolutely
+  critical to predicting forest responses to future droughts, which
+  we're likely to see more of in the future as a result to climate
+  change. Forecasting forest responses to these droughts is a huge and
+  important challenge'', science is better.
+\end{itemize}
 
@@
 paper mentions it would be good to have this data with respect to latent
-affects from drought. \#\#\# Conclusion words \#\#\# Supplementary
-Information
+affects from drought. \#\#\# Acknowledgements
 
-\begin{figure}[H]
-\includegraphics[width=1\linewidth]{tables_figures/ForestGEO_plot} \caption{Map of ForestGEO plot}\label{fig:unnamed-chunk-4}
-\end{figure}
+words
+
+\subsubsection{Author Contribution}\label{author-contribution}
+
+words
+
+\subsubsection{Supplementary
+Information}\label{supplementary-information}
+
+\begin{table}[!h]
+
+\caption{\label{tab:Table S1}Species-specific height regression equations}
+\centering
+\begin{tabular}{llr}
+\toprule
+Species & Equations & r.2\\
+\midrule
+Carya cordiformis & 0.348+0.808*x & 0.879\\
+Carya glabra & 0.681+0.704*x & 0.855\\
+Carya ovalis & 0.621+0.722*x & 0.916\\
+Carya tomentosa & 0.776+0.701*x & 0.894\\
+Fagus grandifolia & 0.708+0.662*x & 0.857\\
+\addlinespace
+Liriodendron tulipifera & 1.32+0.524*x & 0.761\\
+Quercus alba & 1.14+0.548*x & 0.647\\
+Quercus prinus & 0.44+0.751*x & 0.869\\
+Quercus rubra & 1.17+0.533*x & 0.773\\
+all & 0.879+0.634*x & 0.857\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-(see Issue \#32)
+\begin{table}[!h]
 
-\textbf{include TWI by trait graphs (2) here}
+\caption{\label{tab:Table S2}Candidate variables for best model}
+\centering
+\begin{tabular}{rlll}
+\toprule
+prediction & variable & variable\_description & top\_model\\
+\midrule
+1.2 & position\_all & crown.position w/height & 1999\\
+2.2 & height.ln.m & ln[height] & all\\
+2.2 & height.ln.m & ln[height] & 1966\\
+2.3 & position\_all & crown.position alone & 1966\\
+2.4 & TWI.ln & ln[topographic.wetness.index] & all\\
+\addlinespace
+2.4 & TWI.ln & ln[topographic.wetness.index] & 1977\\
+2.4 & TWI.ln & ln[topographic.wetness.index] & 1999\\
+3.1 & year & drought.year & all\\
+3.1 & rp & ring.porosity & 1999\\
+3.2 & PLA\_dry\_percent & percent.loss.area & all\\
+\addlinespace
+3.2 & PLA\_dry\_percent & percent.loss.area & 1966\\
+3.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & all\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{figure}[ht]
+\includegraphics[width=1\linewidth]{tables_figures/ForestGEO_plot} \caption{Map of ForestGEO plot}\label{fig:Plot}
+\end{figure}
+
+\emph{how do we want to present Table S3? Would it be better as an image
+of an excel file, since it's so large? Did we want to keep all
+coefficients here?}
+
+\begin{table}[!h]
+
+\caption{\label{tab:Table S3}Top model variations for each drought scenario, with dAICc values <= 2}
+\centering
+\begin{tabular}{lrll}
+\toprule
+Modnames & Delta\_AICc & scenario & coef\\
+\midrule
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.00 & trees\_all\_sub & (Intercept) (1.101), position\_alldominant (-0.044), position\_allintermediate (-0.039), position\_allsuppressed (-0.049), height.ln.m (-0.092), TWI.ln (-0.058), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.148)\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.84 & trees\_all\_sub & (Intercept) (0.953), height.ln.m (-0.06), TWI.ln (-0.065), year1977 (-0.089), year1999 (-0.076), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.164)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 1.57 & trees\_all\_sub & (Intercept) (1.021), position\_alldominant (-0.046), position\_allintermediate (-0.042), position\_allsuppressed (-0.053), height.ln.m (-0.097), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.009), mean\_TLP\_Mpa (-0.146)\\
+resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1966 & (Intercept) (1.576), height.ln.m (-0.145), rpring (0.091), rpsemi-ring (0.232), PLA\_dry\_percent (-0.021)\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.37 & x1966 & (Intercept) (1.505), height.ln.m (-0.146), TWI.ln (0.043), rpring (0.092), rpsemi-ring (0.227), PLA\_dry\_percent (-0.021)\\
+\addlinespace
+resist.value \textasciitilde{} height.ln.m+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.39 & x1966 & (Intercept) (1.177), height.ln.m (-0.135), PLA\_dry\_percent (-0.014), mean\_TLP\_Mpa (-0.148)\\
+resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.00 & x1977 & (Intercept) (0.386), TWI.ln (-0.136), rpring (-0.233), rpsemi-ring (-0.328), mean\_TLP\_Mpa (-0.37)\\
+resist.value \textasciitilde{} position\_all+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.28 & x1977 & (Intercept) (0.341), position\_alldominant (-0.079), position\_allintermediate (-0.025), position\_allsuppressed (0.039), TWI.ln (-0.128), rpring (-0.23), rpsemi-ring (-0.335), mean\_TLP\_Mpa (-0.385)\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.06 & x1977 & (Intercept) (0.469), height.ln.m (-0.032), TWI.ln (-0.133), rpring (-0.223), rpsemi-ring (-0.318), mean\_TLP\_Mpa (-0.368)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1999 & (Intercept) (1.31), position\_alldominant (0.003), position\_allintermediate (-0.077), position\_allsuppressed (-0.1), height.ln.m (-0.09), TWI.ln (-0.083), rpring (0.191), rpsemi-ring (0.224), PLA\_dry\_percent (-0.009)\\
+\addlinespace
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.05 & x1999 & (Intercept) (0.855), position\_alldominant (0.001), position\_allintermediate (-0.078), position\_allsuppressed (-0.1), height.ln.m (-0.092), TWI.ln (-0.08), rpring (0.183), rpsemi-ring (0.06), mean\_TLP\_Mpa (-0.146)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+(1|sp) & 0.55 & x1999 & (Intercept) (1.19), position\_alldominant (0.002), position\_allintermediate (-0.08), position\_allsuppressed (-0.104), height.ln.m (-0.09), TWI.ln (-0.085), rpring (0.199), rpsemi-ring (0.134)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+mean\_TLP\_Mpa+(1|sp) & 0.94 & x1999 & (Intercept) (0.722), position\_alldominant (-0.002), position\_allintermediate (-0.083), position\_allsuppressed (-0.105), height.ln.m (-0.1), rpring (0.183), rpsemi-ring (0.046), mean\_TLP\_Mpa (-0.154)\\
+resist.value \textasciitilde{} TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.07 & x1999 & (Intercept) (1.044), TWI.ln (-0.097), rpring (0.187), rpsemi-ring (0.265), PLA\_dry\_percent (-0.01)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 1.15 & x1999 & (Intercept) (1.191), position\_alldominant (0.001), position\_allintermediate (-0.081), position\_allsuppressed (-0.106), height.ln.m (-0.098), rpring (0.192), rpsemi-ring (0.215), PLA\_dry\_percent (-0.009)\\
+\addlinespace
+resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.41 & x1999 & (Intercept) (0.516), TWI.ln (-0.094), rpring (0.178), rpsemi-ring (0.074), mean\_TLP\_Mpa (-0.166)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.82 & x1999 & (Intercept) (1.084), position\_alldominant (0.002), position\_allintermediate (-0.077), position\_allsuppressed (-0.099), height.ln.m (-0.091), TWI.ln (-0.081), rpring (0.186), rpsemi-ring (0.147), PLA\_dry\_percent (-0.005), mean\_TLP\_Mpa (-0.076)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+(1|sp) & 1.88 & x1999 & (Intercept) (1.065), position\_alldominant (-0.001), position\_allintermediate (-0.085), position\_allsuppressed (-0.11), height.ln.m (-0.098), rpring (0.201), rpsemi-ring (0.123)\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\includegraphics{tables_figures/FigureS1.png} (see Issue \#32)
 

### 2019-08-14 manuscript/climate_sensitivity_draft.Rmd
Added note for table_variables

@@
 ### Materials and Methods
+
+**table_variables.png** <- where in methods do we want? (following the first citation)
+
 *Study site*

### 2019-08-14 scripts/graphs_for_manuscript.R
Added distr. plot for resistance values

@@ -253,6 +253,13 @@ plot(topo, axes=FALSE, box=FALSE,
 plot(cored_points, pch=20, col = species_colors[species$sp_fact], add=TRUE)
 dev.off()
 
+##density graph of resistance values ####
+ggplot(trees_all_sub) +
+   aes(x = resist.value, fill = year) +
+   geom_density(adjust = 1L) +
+   scale_fill_hue() +
+   theme_minimal()
+
 #######################################################################################
 #4 Export the graphs ####
 library(ggpubr)

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 ---
-title: "Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
+title: "Tree size, (microhabitat)?, and hydraulic traits shape drought responses in a temperate broadleaf forest"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norm Bourg?, Bill McShea?, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
 output:
@@
 *Study site*
-Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [@andersonteixeira_ctfs-forestgeo:_2015].
+Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [[Bourg et al.
+2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
 SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
@@
 
-To accurately understand climate sensitivity, this study used a specific definition of drought, which is not a common practice [@slette_how_2019].
-We used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined by [@lloret_components_2011] as the ratio between the performance during and before the disturbance.
-Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-These were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values for each year, which yielded a set of three periods that were consistently shown as drought: 1964-1966, 1977, and 1999.
+We identified droughts within the time period as 1950-2009, defining drought as events where where peak growing season climatic conditions were anomalously dry relative to the long-term record and where tree growth was substantially reduced for at least some components of the community.
+To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
+Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years for at least three species-canopy class combinations, where canopy class included canopy (dominant and co-domiant) and subcanopy (suppressed and intermediate).
+**[this includes 1991- why did we drop it?
+[issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52)]** Candidate droughts were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values [**source?**] **(and MJJA PET-PRE (droughts identified in Helcoski)?)**.
+**[This paragraph needs work.
+See issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52).
+Put climate criteria first.]**
 
-### Results
-Results outline
-
-Descriptions of Droughts
-
-- physical characteristics of drought: duration, min PDSI (and when occurred), see slette et al 2019 to see how they suggest reporting each one
-- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)
-  - make table for this
-  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
-  - this is followed by Figure 2
-
-Hypothesis results
-- one header for each main question
-- present results for each one, walk through via order and logic of Table 1
-- no citations here
-
+*Analysis* 
 Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
@@
 
-**Describe each individual drought scenario **
-1964-1966
-1977
-1999
+### Results
+
+*Descriptions of Droughts*
+In the 60-year period between 1950 and 2009, there were three droughts that met our definition of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1964-1966, 1977, and 1999 (Fig.
+1).
+We note that 1964 and 1966 were separately identified as drought years and grouped into a single multi-year drought period because 1965 was also anomalously hot and dry, albiet not identified as a pointer year.
+Climatically, these droughts included four of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
+Specifically, 1964, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.87, 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
+The year 2007 also ranked in the top five driest (PET-PRE = 82.13 mm mo<sup>-1</sup>), but it did not have particualarly low PDSI and was not identified as a pointer year.
+**add stats for PDSI, T, precip**
+
+- physical characteristics of drought: duration, min PDSI (and when occurred), see [@slette_how_2019] to see how they suggest reporting each one
+- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)
+
+  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
 

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 **Table 1. Summary of hypotheses, predictions and results**
+
 ![Table 1](tables_figures/Table1.png){width=500px}
@@
 
-*Climate and drought years*
+*Identification of drought years*
 [**add description of climate data used in Fig. 1, NEON vertical profiles**]
 
-We identified droughts within the time period as 1950-2009, defining drought as events where where peak growing season climatic conditions were anomalously dry relative to the long-term record and where tree growth was substantially reduced for at least some components of the community.
+We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where peak growing season climatic conditions were anomalously dry relative to the long-term record and where tree growth was substantially reduced for at least some components of the community.
 To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
@@
 Put climate criteria first.]**
+
+[averaged resistance values for 1964-1966]
 
@@
 *Descriptions of Droughts*
-In the 60-year period between 1950 and 2009, there were three droughts that met our definition of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1964-1966, 1977, and 1999 (Fig.
+In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1964-1966, 1977, and 1999 (Fig.
 1).
-We note that 1964 and 1966 were separately identified as drought years and grouped into a single multi-year drought period because 1965 was also anomalously hot and dry, albiet not identified as a pointer year.
+The years 1964 and 1966 were separately identified as drought years and grouped into a single multi-year drought period because 1965 was also anomalously hot and dry, albiet not identified as a pointer year, and because inclusion of 1964 in the reference period for 1966 would have biased analyses.
 Climatically, these droughts included four of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
 Specifically, 1964, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.87, 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The year 2007 also ranked in the top five driest (PET-PRE = 82.13 mm mo<sup>-1</sup>), but it did not have particualarly low PDSI and was not identified as a pointer year.
+The year 2007 also ranked in the top five driest (PET-PRE = 82.13 mm mo<sup>-1</sup>), but it did not have *particualarly low* PDSI and was not identified as a pointer year.
 **add stats for PDSI, T, precip**
 
-- physical characteristics of drought: duration, min PDSI (and when occurred), see [@slette_how_2019] to see how they suggest reporting each one
-- sum nb.series for each scenario, mult by perc.neg (pointers df), define thresholds (e.g. 75% of all trees experienced the growth reduction of 40% or more)
+Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1964-1966, 1977, and 1999, respectively (Fig.
+1b).
+In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): *28.2%* in 1964-66, 31.2% un 1977, and 26.2% in 1999.
+*Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999.
 
@@
 **Figure 1.
-Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.** Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.
+Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
 Figure modified from [@helcoski_growing_2019].
+(b) density plots of community- wide resitance values for each drought.
 

### 2019-08-15 scripts/plot_time_series.R
Updated time series figure to match colors of density figure (see Figure 1)

@@ -1,5 +1,5 @@
 ######################################################
-# Purpose: Plot figure of time series for manuscript
+# Purpose: Plot figure of ring width time series for McGregor manuscript
 # Developped by: Valentine Herrmann - HerrmannV@si.edu
 # R version 3.5.1 (2018-07-02)
 ######################################################
@@ -68,7 +68,7 @@ plot(NULL,
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
-     xright = drought_years, ytop = par("usr")[4],  col = "grey90", border = "transparent")
+     xright = drought_years, ytop = par("usr")[4],  col = adjustcolor(c("#FF9999", "#FF9999", "#FF9999", "#009900", "#6699CC"), alpha.f=0.5), border = "transparent")
 # abline(v = pre_drought_years, col = "grey", lty = 2) # seq(1959, 2000, by = 20)
 abline(v = drought_years, col = "grey50", lty = 2) # seq(1959, 2000, by = 20)
 
@@ -86,7 +86,7 @@ plot(NULL,
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
-     xright = drought_years, ytop = par("usr")[4],  col = "grey90", border = "transparent")
+     xright = drought_years, ytop = par("usr")[4],  col = adjustcolor(c("#FF9999", "#FF9999", "#FF9999", "#009900", "#6699CC"), alpha.f=0.5), border = "transparent")
 # abline(v = pre_drought_years, col = "grey", lty = 2) # seq(1959, 2000, by = 20)
 abline(v = drought_years, col = "grey50", lty = 2) # seq(1959, 2000, by = 20)
 
@@ -107,7 +107,7 @@ plot(NULL,
 
 
 rect(xleft = pre_drought_years, ybottom = par("usr")[3], 
-     xright = drought_years, ytop = par("usr")[4],  col = "grey90", border = "transparent")
+     xright = drought_years, ytop = par("usr")[4],  col = adjustcolor(c("#FF9999", "#FF9999", "#FF9999", "#009900", "#6699CC"), alpha.f=0.5), border = "transparent")
 # abline(v = pre_drought_years, col = "grey", lty = 2) # seq(1959, 2000, by = 20)
 abline(v = drought_years, col = "grey50", lty = 2) # seq(1959, 2000, by = 20)
 
@@ -143,7 +143,7 @@ for(f in SPECIES_IN_ORDER) {
[…]

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
updated & knit manuscript!
I got it working! issue #30.

@@
 title: "Tree size, (microhabitat)?, and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norm Bourg?, Bill McShea?, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norm Bourg?, Bill McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
 output:
@@
 Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.
-
-![Location of cored trees](tables_figures/Figure3.png){width=500px}
 
@@
 
+![Location of cored trees](tables_figures/FigureS3.png){width=500px}
+
 ![Figure S1](tables_figures/FigureS1.png)

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
updating ms (added Table5 placeholder)

@@
 
+Comparing across droughts, resistance *was highest in 1964-66 drought*, lower in 1999, and lowest in 1977 (Tables 4-5).
+
   - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
@@
 
+![Table 4 - individual variable tests](tables_figures/Table4.png) #this is tested_traits_all
+
+![Table 5 - best full models ](tables_figures/Table5.png)
+
 **Results for first main question: what drives the observed tendency for large trees to suffer more during drought?**
@@
 Hydraulic traits alone, traits with height
-
-![Table 4 (example!)](tables_figures/Table4.png) #this is tested_traits_all
 

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
issue #39

@@
 *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999.
-
-Comparing across droughts, resistance *was highest in 1964-66 drought*, lower in 1999, and lowest in 1977 (Tables 4-5).
-
-  - citing table 3 (the coefficient for individual year) - maybe mention how 77 and 99 have neg coefficients, indicating potential height effect: as time goes on, trees are more affected by drought anyways because they grow taller
+Within the context of mixed effects models, there were signficiant differences among droughts (Table 4), and drought year came out as a factor in all top models (Table 5).
+In all cases, resistance *was highest in 1964-66 drought*, lower in 1999, and lowest in 1977 (Tables 4-5).
 

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
working on manuscript, mainly methods
fixed up equation, more

@@
 
-*Let's put Table 1 here.
-We need to give the hypotheses/ predictions here.
-It seems counterintuitive to put the results right up front like that, but it actually tends to be a higher-impact style of writing, and recommended [here](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619).*
-
-**Table 1. Summary of hypotheses, predictions and results**
-
-![Table 1](tables_figures/Table1.png){width=500px}
+![**Table 1. Summary of hypotheses, predictions and results**](tables_figures/Table1.png){width=500px}
 
@@
 
-**table_variables.png** <- where in methods do we want? (following the first citation)
-
 *Study site*
@@
 *Data collection and preparation*
+
+Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
 The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems >= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH (described below), along with tree location in the plot to determine the topographic wetness index.
-Furthermore, we used data for all stems >= 10cm to analyze functional trait composition relative to tree height.
-*I added the last sentence-- not sure if you've described that below yet.
-I don't think you need to note here which data we use, so long as that's described in the portions where you describe the analysis.*
+From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems >= 10cm to analyze functional trait composition relative to tree height (all analyses described below).
+Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO data portal (www.forestgeo.si.edu).
 
-We analyzed tree-ring data from the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019].
+![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}
+
+We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019].
 Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
@@
 The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-*is this a database?
-Yes.
-Ryan submitted the data but I don't think its posted yet.
+* Ryan submitted the data to ITRDB but I don't think its posted yet.
 We should also cite GitHub/Zenodo here.
@@
 
+For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. Prior DBH was estimated using the following equation:
+
+$$ DBH_Y  = DBH_{2008} - 2*[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}] $$
+
+Here,  *Y*  is denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness.
+Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
+Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#- **create table to give these equations in SI**) and then used these to estimate bark thickness based on DBH.
+
 Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
@@
 Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-The associated errors for using either method were acknowledged [@larjavaara_measuring_2013].
+Both methods are associated with some error [@larjavaara_measuring_2013].
 Species-specific height allometries were developed (Table S# - **ADD THIS TABLE TO SI**).
-For species that didn't have enough height measurements, heights were calculated from equations derived from all species in the study.
-
-For each tree, we combined tree-ring records and allometric equations of height and bark thickness to retroactively calculate DBH and estimate height for the years 1950-2009.
-Prior DBH was estimated using the following equation, using 2008 as the earliest year for having reliable DBH measurements:
-
-$$ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) $$
-
-Here, *ring.width* was measured from cores. Bark thickness was estimated from species-specific allometries based on the bark thickness data of [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#- **create table to give these equations in SI**) and then used these to estimate bark thickness based on DBH.
+For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
 
@@
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [**include a brief explanation of what this is**]
 
-Hydraulic traits were collected from SCBI and are summarized in Table 1.
+Hydraulic traits were collected from SCBI and are summarized in Table 3.
 In August 2018, we collected leaf samples from three individuals of each species ...
@@
 
-**Table 2.
+**Table 3.
 Species analyzed here, listed in descending order of ANPP_stem.
@@
 
-```{r, Table2, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
@@
 
-**Figure 1.
+![**Figure 1.
 ** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.
@@
 Figure modified from [@helcoski_growing_2019].
-(b) density plots of community- wide resitance values for each drought.
-
-![Time series of combined tree cores by species](tables_figures/Figure1.jpg){width=500px}
+(b) density plots of community- wide resitance values for each drought.](tables_figures/Figure1.jpg){width=500px}
 

### 2019-08-15 scripts/canopy_position_analysis.R
Updated scripts with new math

@@ -229,7 +229,6 @@ names(canopy) <- values
 values <- paste0(sp_can, "_canopy")
 names(widths_can) <- values
 
-
 ###subcanopy ####
 dirs_subcan <- dir("data/core_files/subcanopy_cores", pattern = "_subcanopy.rwl")
 dirs_subcan <- dirs_subcan[!dirs_subcan %in% c("pist_drop_subcanopy.rwl", "frni_drop_subcanopy.rwl")]
@@ -262,6 +261,7 @@ values_sub <- paste0(sp_subcan, "_subcan_res")
 names(subcanopy) <- values_sub
 values <- paste0(sp_subcan, "_subcanopy")
 names(widths_sub) <- values
+names(all_bai_subcan) <- values
 
 widths <- c(widths_can, widths_sub) #combine into one, then delete individual. For use in #5d
 widths_can <- NULL
@@ -310,17 +310,39 @@ for (i in seq(along=1:length(tag_n))){
   ind <- can_resist
   ind_neil <- neil_list[neil_list$tag %in% tree_series, ]
   
+  ##for original code here, see below in the subcanopy loop
+  
   ind$year <- years
+  ind$year <- as.numeric(ind$year)
   ind$sp <- unique(ind_neil$sp)
   ind$position <- "canopy"
+
+  ind <- ind[ind$year %in% c(1959:1963,1964:1966,1977,1999), ]
   
-  ind <- ind[ind$year %in% pointer_years, ]
-  
-  ## these three lines of code are for taking the mean of 1964-1966, 
+  ## these lines of code get the mean of 1959:163 for pre-drought, and then do the mean of the drought growth itself (1964:1966). Then you get the resistance value by doing:
+  ###[average growth 1964-66] / [ average growth 1959 - 1963]
   ## since it was a multi-year drought. We're calling it 1966 for simplicity.
-  ind["1966", 1:(ncol(ind) -3)] <- colMeans(ind[c(1:3), 1:(ncol(ind) -3)])
-  ind <- ind[-c(1,2), ]
-  ind[, 1:(ncol(ind) -3)] <- round(ind[, 1:(ncol(ind) -3)], 2)
+  df <- ind[, 1:(ncol(ind) -3)]
+  df_pre <- df[rownames(df) %in% c(1959:1963), ]
+  df_dro <- df[rownames(df) %in% c(1964:1966), ]
+  
+  df_pre <- as.data.frame(t(data.frame(colMeans(df_pre))))
+  colnames(df_pre) <- colnames(df)
+  
[…]

### 2019-08-15 scripts/graphs_for_manuscript.R

@@ -256,11 +256,16 @@ dev.off()
 ##density graph of resistance values ####
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv")
 trees_all_sub$year <- as.character(trees_all_sub$year)
+
+jpeg("manuscript/tables_figures/Figure1a.jpg", res=300, width=150, height=75, units="mm")
 ggplot(trees_all_sub) +
    aes(x = resist.value, fill = year) +
    geom_density(adjust = 1L, alpha=.5) +
-   scale_fill_hue() +
+   geom_vline(xintercept=1) +
+   scale_fill_hue(labels=c("1964-1966", "1977", "1999")) +
+   labs(x="resistance value") +
    theme_minimal()
+dev.off()
 
 #######################################################################################
 #4 Export the graphs ####

### 2019-08-15 manuscript/climate_sensitivity_draft.Rmd
fixed equation
fixed equation
update, knit ms

@@
 
-$$ DBH_Y  = DBH_{2008} - 2*[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}] $$
+$$ DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right] $$
 
@@
 To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
-Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years for at least three species-canopy class combinations, where canopy class included canopy (dominant and co-domiant) and subcanopy (suppressed and intermediate).
-**[this includes 1991- why did we drop it?
+Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years.
+*Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years for at least three species-canopy class combinations, where canopy class included canopy (dominant and co-domiant) and subcanopy (suppressed and intermediate).* **[this includes 1991- why did we drop it?
 [issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52)]** Candidate droughts were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values [**source?**] **(and MJJA PET-PRE (droughts identified in Helcoski)?)**.
@@
 Put climate criteria first.]**
-
-[averaged resistance values for 1964-1966]
 

### 2019-09-03 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 
-Here, we combine tree-ring records covering three droughts (1964-66, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses.
+Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses.
 First, we focus on the role of tree size and its interaction with microenvironement. 
@@
 
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where peak growing season climatic conditions were anomalously dry relative to the long-term record and where tree growth was substantially reduced for at least some components of the community.
+We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period.
 To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
 Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years.
-*Candidate drought years were defined if >50% of the cored trees experienced <30% growth in a year compared to the previous 5 years for at least three species-canopy class combinations, where canopy class included canopy (dominant and co-domiant) and subcanopy (suppressed and intermediate).* **[this includes 1991- why did we drop it?
-[issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52)]** Candidate droughts were then cross-validated with the regional Palmer Drought Severity Index (PDSI) values [**source?**] **(and MJJA PET-PRE (droughts identified in Helcoski)?)**.
-**[This paragraph needs work.
-See issue #52](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/52).
-Put climate criteria first.]**
+Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
+Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site.
+We considered two metrics of moisture deficit: regional Palmer Drought Severity Index (PDSI) values [**source- NOAA-same as Helcoski**] and the difference between potential evapotranspiration (PET) and precipitation (PRE) [**source- same as Helcoski**].
+The driest years were identified through simply ranking of mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
 
@@
 *Descriptions of Droughts*
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1964-1966, 1977, and 1999 (Fig.
+In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig.
 1).
-The years 1964 and 1966 were separately identified as drought years and grouped into a single multi-year drought period because 1965 was also anomalously hot and dry, albiet not identified as a pointer year, and because inclusion of 1964 in the reference period for 1966 would have biased analyses.
-Climatically, these droughts included four of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
-Specifically, 1964, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.87, 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The year 2007 also ranked in the top five driest (PET-PRE = 82.13 mm mo<sup>-1</sup>), but it did not have *particualarly low* PDSI and was not identified as a pointer year.
-**add stats for PDSI, T, precip**
+We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**).
+Rather, reduced growth may be explained in part because gypsy moths (*Lymantria dispar L.*) were present in the region from XXXX-XXXX, causing extensive defoliation of *Quercus* spp.
+(**#% of cores/propoductivity**), with peak impact at SCBI within 3 years prior to 1991 (**Cite Shenanadoah paper, if accepted**).
+Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
+Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
+The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.*
+**The droughts differed in timing/duration/etc.
+..
+The year 1966 was preceded by two relatively dry years...
+1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry.
+**
 

### 2019-09-03 manuscript/climate_sensitivity_draft.Rmd
knit, following updates to years
#54

@@
 Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
-Each drought scenario (1964-66, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
+Each drought scenario (1966, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
 Individual-level growth responses to the three individual droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.

### 2019-09-08 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
a few minor updates

@@
 2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
-However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA) and turgor loss point (TLP), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested (CONFIRM), will prove better predictors.
+However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA; REF) and turgor loss point (TLP; REF), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested, will prove better predictors.
 
@@
 
-We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature [@bennett_larger_2015] (REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
+We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature [@bennett_larger_2015] (*Stovall_new*; REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
 

### 2019-09-09 manuscript/climate_sensitivity_draft.Rmd
added gypsy moth info

@@
 We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**).
-Rather, reduced growth may be explained in part because gypsy moths (*Lymantria dispar L.*) were present in the region from XXXX-XXXX, causing extensive defoliation of *Quercus* spp.
-(**#% of cores/propoductivity**), with peak impact at SCBI within 3 years prior to 1991 (**Cite Shenanadoah paper, if accepted**).
+Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar L.*) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
+(**#% of cores/propoductivity**) (**Cite Shenanadoah paper, if accepted**).
 Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].

### 2019-09-10 manuscript/climate_sensitivity_draft.Rmd
updated tables 1 and 4 with new results, formatted table 5, made some edits to draft

@@
 ---
-title: "Tree size, (microhabitat)?, and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norm Bourg?, Bill McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
+title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
 output:
@@
 *Study site*
-Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78° 08'43.4"W) [[Bourg et al.
+Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78°08'43.4"W) [[Bourg et al.
 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
@@
 The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-* Ryan submitted the data to ITRDB but I don't think its posted yet.
+*Ryan submitted the data to ITRDB but I don't think its posted yet.
 We should also cite GitHub/Zenodo here.
-I'll come back to that.
-*
+I'll come back to that.*
 
@@
 We considered two metrics of moisture deficit: regional Palmer Drought Severity Index (PDSI) values [**source- NOAA-same as Helcoski**] and the difference between potential evapotranspiration (PET) and precipitation (PRE) [**source- same as Helcoski**].
-The driest years were identified through simply ranking of mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
+The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
 
@@
 *Descriptions of Droughts*
+
+![**Figure 1.
+** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.
+Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from [@helcoski_growing_2019].
+(b) density plots of community- wide resitance values for each drought.](tables_figures/Figure1.png){width=500px}
+
 In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig.
@@
 We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**).
-Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar L.*) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(**#% of cores/propoductivity**) (**Cite Shenanadoah paper, if accepted**).
+Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
+(**Cite Shenanadoah paper, if accepted**).
 Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
@@
 
-Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1964-1966, 1977, and 1999, respectively (Fig.
+Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig.
 1b).
-In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): *28.2%* in 1964-66, 31.2% un 1977, and 26.2% in 1999.
+In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999.
 *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999.
-Within the context of mixed effects models, there were signficiant differences among droughts (Table 4), and drought year came out as a factor in all top models (Table 5).
-In all cases, resistance *was highest in 1964-66 drought*, lower in 1999, and lowest in 1977 (Tables 4-5).
-
-![**Figure 1.
-** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.
-Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-(b) density plots of community- wide resitance values for each drought.](tables_figures/Figure1.jpg){width=500px}
+Within the context of mixed effects models, there were no significant differences in resistance across drougth years (Table 4).
 

### 2019-09-14 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
Updated formatting for manuscript

@@
 - \captionsetup[table]{labelformat=empty}
+- \captionsetup[figure]{labelformat=empty}
 - \usepackage{dcolumn} #for sidewaystable
@@
 Each drought scenario (1966, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
-Individual-level growth responses to the three individual droughts  were stronger among taller trees in dominant canopy positions, those in wetter microsites, and for more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration), with substantial variation in the best predictor variables across given droughts.
-We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected.
+Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
+However, there was substantial variation in the best predictor variables across given droughts.
+We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected. **Add discussion points**
 
@@
 *First, what drives the observed tendency for large trees to suffer more during drought?*
+
 @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought.
@@
 *Second, how do species' traits - alone and in interaction with tree size - influence drought response?*
+
 Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
@@
 *Third, are responses similar or variable across individual drought years?*
+
 Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
@@
 
+\clearpage
+
 ### Materials and Methods
+*Study site*
 
-*Study site*
-Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38°53'36.6"N, 78°08'43.4"W) [[Bourg et al.
+Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al.
 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
@@
 Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems >= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems >= 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO data portal (www.forestgeo.si.edu).
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems \>= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems \>= 10cm to analyze functional trait composition relative to tree height (all analyses described below).
+Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO [data portal](www.forestgeo.si.edu).
 
+\begin{figure}[H]
 ![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}
+\end{figure}
 
@@
 
-$$ DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right] $$
+$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$
 
@@
 Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]
-
 ```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
@@
 table2 <- read.csv("tables_figures/Table2.csv")
-kable(table2, booktabs=TRUE, format = "latex", caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
-Units of measurements are in mm (DBH), % (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+
+#the percentage symbol has a double backslash because otherwise this writes out the latex code when making the table
+
+kable(table2, booktabs=TRUE, caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
+Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
 ```
 
+\raggedright
+
 *Identification of drought years*
+
 [**add description of climate data used in Fig. 1, NEON vertical profiles**]
@@
 *Analysis*
+
 Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
@@
 
+\raggedright
+
+\newpage
 ### Results
@@
 *Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999.
-Within the context of mixed effects models, there were no significant differences in resistance across drougth years (Table 4).
+Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).
 
@@
 
+\newpage
 ### Discussion
@@
 This paper mentions it would be good to have this data with respect to latent affects from drought.
+\newpage
 ### Acknowledgements
-
 words
 
-### Author Contribution ####
+### Author Contribution 
 words

### 2019-09-14 scripts/graphs_for_manuscript.R
Removed height labels from NEON graphs #48

@@ -17,7 +17,7 @@ library(grid) #2
 library(gridExtra) #2
 
 ##2a. heights for all cored trees ####
-trees_all <- read.csv("manuscript/tables_figures/trees_all.csv", stringsAsFactors = FALSE)
+trees_all <- read.csv("manuscript/tables_figures/trees_all.csv", stringsAsFactors = FALSE) #these graphs are meant to be for all cored trees, not just the ones being used in analysis
 
 scbi.stem3 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
 
@@ -114,7 +114,7 @@ scbi.stem3$dbh <- ifelse(is.na(scbi.stem3$dbh), 0, scbi.stem3$dbh)
 scbi.stem3 <- scbi.stem3[scbi.stem3$dbh >= 100, ] #>10cm dbh
 
 #get utm coords
-source('D:/Dropbox (Smithsonian)/Github_Ian/SCBI-ForestGEO-Data/R_scripts/SIGEO_plot_grid_UTMcoord.R', echo=TRUE)
+source('E:/Github_SCBI/SCBI-ForestGEO-Data/R_scripts/SIGEO_plot_grid_UTMcoord.R', echo=TRUE)
 
 plot_to_UTM(scbi.stem3)
 scbi18 <- sigeo_coords
@@ -126,7 +126,7 @@ scbi18$TWI <- twi
 scbi18$TWI.ln <- log(scbi18$TWI)
 
 #add in leaf traits (no rp because it's categorical)
-leaf_traits <- trees_all[, c(2,7:10)]
+leaf_traits <- trees_all[, c(4,6:8)]
 
 for (i in seq(along=2:ncol(leaf_traits))){
    trait <- colnames(leaf_traits[2:ncol(leaf_traits)])
@@ -148,8 +148,6 @@ breaks_pla = c(8,12,16,20,24)
 
 limits_tlp = c(-2.8,-1.8)
 limits_pla = c(8,26)
-
-graph_names <- paste0(main$type, "_plot")
 
 for (i in seq(along=1:4)){
    if(!grepl("twi", type[[i]])){
@@ -232,7 +230,13 @@ for (i in seq(along=1:4)){
    assign(paste0("plot_", type[i]), q)
 }
 
-##3c. Make map of plot using TWI and cored tree locations ####
+
[…]

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
-Erika should be coauthor
-drafted first paragraph of intro.

@@
 title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Erika Gonzalez-Akre, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
 output:
@@
 ### Introduction
-Understanding how and why trees respond to drought is critical to predicting forest drought responses and climate change feedbacks.
-(1 paragraph on this- I have lots of content on this and will add later.)
+Forests globally play a critical role in climate regulation ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)), yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change ([Friedlingstein et al.
+2006](https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1)).
+An important aspect of this uncertainty lies in responses to drought (**REF**).
+In many forested regions around the world, the risk of severe drought is increasing ([Trenbert et al.
+2014](https://www.nature.com/articles/nclimate2067)), even in conjunction with increasing precipitation ([IPCC 2014](https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334)).
+Global change-type drought has been affecting forests worldwide ([Allen et al.
+2015](https://doi.org/10.1016/j.foreco.2009.09.001)), and it is expected that future climate change-driven droughts will severely impact forests around the world ([Allen et al.
+2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**).
+Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; Stovall et al.
+2019), resulting in disproportionate impacts on forest C storage ([Meakem et al.
+2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+[Yet, modeling forest responses to drought remains challenging-- in part because some of the mechanisms underlying drought responses remain unclear.]
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
more updates to intro

@@
 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
-[Yet, modeling forest responses to drought remains challenging-- in part because some of the mechanisms underlying drought responses remain unclear.]
-
-[Understanding forest responses to drought requires increased functional understanding of the factors that confer individual-level vulnerability or resistance.] Forests are diverse in terms of tree sizes and functional traits, and it is known that trees varying in size and functional traits respond differently to drought (e.g., [@bennett_larger_2015]; REFS).
-Therefore, in order to understand whole-forest response to drought, we need a functional understanding of how responses vary by tree size, microhabitat, and species.
-There are 3 fundamental questions that must be addressed:
+As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
+Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and whether their influence is consistent across droughts.
 
@@
 
-@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought.
+@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
+2019, **REFS**).
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
deleted obsolete P50/P80 text

@@
 
-*p50 and p80*
-We decided to include values of P50 and P80 in the leaf traits model, defined by [@anderegg_meta-analysis_2016] as the water potentials at which a species loses 50% and 88% [80% by proxy], respectively, of hydraulic conductivity.
-Values were calculated by (**insert new methods here??**), and were only available for six species (*C.
-glabra*, *L.
-tulipifera*, *Q.
-alba*, *Q.
-prinus*, *Q.
-rubra*, and *Q.
-velutina*).
-Because of this, the model runs were considered to be incomplete due to the exclusion of the other 8 species.
-Results revealed neither p50 nor p80 to be significant, thus for the full analysis we decided to drop the two traits in order to include all species in the full analysis.
-
 ```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
-Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and whether their influence is consistent across droughts.
+Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
 
@@
 ### Acknowledgements
-words
+funding: ForestGEO,
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
more work on introduction

@@
 
-*First, what drives the observed tendency for large trees to suffer more during drought?*
-
+One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought?
 @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
@@
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms (besides insects) for the observed greater drougth growth reductions of large trees: (1) inherently greater biophysical challenge of being tall; (2) greater exposure of the crowns of large trees.
-Counteracting these effects, (3) the larger root systems of larger trees may confer an advantage in terms of allowing greater access to water, but it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-
-Canopy trees have lower drought resistance because they are exposed to higher solar radiation, greater wind speeds, and lower humidity.
+This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (**REFS**)--for the observed greater drought growth reductions of large trees.
+First, [: (1)] inherently greater biophysical challenge of being tall].
+Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (**REFS**).
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Elucidating the mechanisms that place larger trees at a relative disadvantage during drought will be critical to better understanding and modeling drought impacts on forests.
 
@@
 We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in Bennett et al (2015) (H1.0).
-We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees in dry but not perpetually wet microsites (H1.3).
+We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (H1.3).
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
more edits to intro

@@
 
-One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought?
-@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
+**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought?** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
 2019, **REFS**).
@@
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-Elucidating the mechanisms that place larger trees at a relative disadvantage during drought will be critical to better understanding and modeling drought impacts on forests.
+A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al.
+2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+It is possible that the pattern observed by @bennett_larger_2015 could be caused by smaller trees being more *drought resistant*.
+Alternatively, larger trees may have more *drought-resistant* [**careful on wording here**] traits as adpatations to greater biophysical challenges.
 
-*Second, how do species' traits - alone and in interaction with tree size - influence drought response?*
+**A second fundamental question regarding forest responses to drought is how species' traits - alone and in interaction with tree size - influence drought response.**
 
@@
 
-Moreover, there may be an interaction between traits and tree size.
-It is possible that the pattern observed by @bennett_larger_2015 could be caused by smaller trees being more drought resistant.
-Alternatively, larger trees may have more drought-resistant traits as adpatations to greater biophysical challenges.
+(interaction between traits and tree size).
 
-*Third, are responses similar or variable across individual drought years?*
+**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing is linked to a diversity of tree repsonses.
+**
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
worked on intro

@@
 
-**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought?** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
+**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
 2019, **REFS**).
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (**REFS**)--for the observed greater drought growth reductions of large trees.
-First, [: (1)] inherently greater biophysical challenge of being tall].
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (**REFS**).
+This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (**REFS(KAT)**)--for the observed greater drought growth reductions of large trees.
+First, taller trees facea greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (), and this may become a greater liability during drought (**REFS(KAT)**).
+Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (**REFS(KAT)**).
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS(KAT)**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
 A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al.
 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
-It is possible that the pattern observed by @bennett_larger_2015 could be caused by smaller trees being more *drought resistant*.
-Alternatively, larger trees may have more *drought-resistant* [**careful on wording here**] traits as adpatations to greater biophysical challenges.
+It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance.
+Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al.
+2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al.
+2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.
 
-**A second fundamental question regarding forest responses to drought is how species' traits - alone and in interaction with tree size - influence drought response.**
-
-Analyzing drought responses on the species level does not fully explain mechanisms and is not feasible in diverse forests.
-The solution is a trait-based approach.
-Leaf hydraulic traits hold more promise than more commonly/traditionally-measured traits such as wood density and specific leaf area (SLA) ([Medeiros et al.
+**A second fundamental question regarding forest responses to drought is how species' traits - alone and in interaction with tree size - influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
+Commonly measured traits including wood density (**symbol**) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al.
+2015, Friedrichs et al.
+2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
+However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication (*PLA*; **REF**) and turgor loss point (*TLP*; **REF**), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as WD and $LMA$ ([Medeiros et al.
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
 
-Commonly measured traits including wood density (REFS), leaf mass per area [@abrams_adaptations_1990] [@guerfel_impacts_2009], and ring porosity (Elliot et al.
-2015, Friedrichs et al.
-2009) have been linked to drought responses and likely correlated with drought resistance in this forest.
  However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA; REF) and turgor loss point (TLP; REF), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested, will prove better predictors.
@@
 
-\begin{figure}[H]
 ![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}
-\end{figure}
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
right-justification of text
finished first draft of intro!

@@
 ```
+\raggedright
 
@@
 
-**A second fundamental question regarding forest responses to drought is how species' traits - alone and in interaction with tree size - influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density (**symbol**) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al.
+**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
+Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al.
 2015, Friedrichs et al.
 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
-However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication (*PLA*; **REF**) and turgor loss point (*TLP*; **REF**), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as WD and $LMA$ ([Medeiros et al.
+However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; **REF**) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs (**REFS**), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al.
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
+Their ability to predict tree performance under drought remains untested.
 
- However, we hypothesize that leaf hydraulic traits such as leaf area shrinkage upon dessication (PLA; REF) and turgor loss point (TLP; REF), which are emerging as potentially more informative traits but whose effect on drought resistance has never been tested, will prove better predictors.
-
-(interaction between traits and tree size).
-
-**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing is linked to a diversity of tree repsonses.
-**
-
-Droughts are rarely explicitly defined in ecological studies [@slette_how_2019], yet no two droughts are the same.
-
-We need to understand the the factors confirming drought vulnerability or resistance not only for extreme droughts with dramatic impacts on tree growth and mortality, which tend to dominate the literature [@bennett_larger_2015] (*Stovall_new*; REFS), but also for more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade.
+**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, yet our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].
+Our knowledge of forest drought responses tends to come disproportionately from extreme droughts with dramatic impacts (*e.g.*, [Allen et al.
+2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**), and our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
+While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
+There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; *Stovall et al.
+2019 (I think!)*).
+[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, it is also likely that tree size and traits interact with drought characteristics to result in differential responses across droughts--but this has yet to be tested.
 
@@
 First, we focus on the role of tree size and its interaction with microenvironement. 
-We confirm that, consistent with most forests globally [@bennett_larger_2015], larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in Bennett et al (2015) (H1.0).
-We then test hypotheses designed to disentangle the relative importance of tree height (H1.1), crown exposure (H1.2), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (H1.3).
+We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
+We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area shrinkage upon dehydration and turgor loss point will prove better predictors (P2.1d-e).
-We then test whether these traits correlate with tree height (P2.2), potentially driving the observed tendency for taller trees to suffer more during drought (P2.3). 
-Finally, we focused on variability among droughts, asking how  community resistance varied across droughts (H3.1) and whether the factors confirming vulnerability or resistance varied across droughts (H3.2).
+We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity , but that leaf hydraulic traits such as leaf area shrinkage upon dehydration and turgor loss point will prove better predictors (*H2.1*).
+We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
+Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).
 
@@
 
+\raggedright
 We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction. 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
replaced Excel /.png version of Table 1 with Markdown!
...and it looks so much better!

@@
 title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Erika Gonzalez-Akre, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
+author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
 output:
@@
 
-![**Table 1. Summary of hypotheses, predictions and results**](tables_figures/Table1.png){width=500px}
-
-\raggedright
-We count predictions as fully supported (or rejected) when the direction of response matches (or contradicts) the prediction. 
-H2.3-H3.2 are based on if models containing species traits in H2.1 had dAIC>=2 relative to the appropriate null model or to any alternative multivariate model within 2 dAIC. 
-Parentheses indicate that predictions were partially supported (or rejected).
-In other words, parentheses indicate when the direction of response matched (or contradicted) the prediction in some but not all models with dAIC<2 relative to the appropriate null model.
-With categorical variables such as crown position,  a "(yes)" notation describes when the trend matched the prediction, but it wasn't significant
-A "(yes/no) or ("no/yes")" indicates tendencies in opposite directions in univariate tests versus the best full models, respectively. 
-"(yes)
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
+Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable).
+Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 9) %>%
+  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 4, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.3. Drought resistance is linked to rooting volume in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.1. Species traits predict drought resistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H2.2. More drought-resistant traits associated to taller trees.", 12, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.1. Responses varied by drought.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H3.2. Predictor variables varied across droughts.", 19, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
+  column_spec(1, width = "15cm") %>%
+  column_spec(2:5, width = "1cm") %>%
+  column_spec(6, width = "2cm") %>%
+  kableExtra::landscape()
+```
 
@@
 
-![**Table 2. Summary of hypotheses, predictions and results** ](tables_figures/table_variables.png){width=500px}
+![**Table 2. Summary of variables** ](tables_figures/table_variables.png){width=500px}
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
worked on methods

@@
   pack_rows("H1.2. Drought resistance decreases with crown exposure.", 4, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Drought resistance is linked to rooting volume in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.3. Small trees with lower root volume suffer proportionally more in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H2.1. Species traits predict drought resistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. More drought-resistant traits associated to taller trees.", 12, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 12, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
 
-Research was conducted at the 25.6 ha ForestGEO (Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al.
+Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al.
 2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
@@
 Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
-Dominant species include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
+Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
 
@@
 Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems \>= 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems \>= 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**???**), were obtained through the ForestGEO [data portal](www.forestgeo.si.edu).
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\le$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
+Census data, which were last updated in 2018 (**confirm**), are available through the [ForestGEO data portal](www.forestgeo.si.edu).
 
@@
 Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH >= 10cm [@bourg_initial_2013].
+In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013].
 In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
@@
 
-For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. Prior DBH was estimated using the following equation:
+For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
+Prior DBH was estimated using the following equation:
 
@@
 
-Here,  *Y*  is denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness.
+Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#- **create table to give these equations in SI**) and then used these to estimate bark thickness based on DBH.
+Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#) and then used these to estimate bark thickness based on DBH.
 
@@
 Both methods are associated with some error [@larjavaara_measuring_2013].
-Species-specific height allometries were developed (Table S# - **ADD THIS TABLE TO SI**).
+Species-specific height allometries were developed (Table S#).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
@@
 While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-XX) and height change since *1959* indicated that change was likely slow.
-[**work on this-- provide details?**]
+2) and height change since *1959* indicated that change was likely slow.
+Specifically, [**Ian, please provide details--e.g., average rate of height growth**]
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [**include a brief explanation of what this is**]
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is**]
 
 Hydraulic traits were collected from SCBI and are summarized in Table 3.
-In August 2018, we collected leaf samples from three individuals of each species ...
-(**Nobby's description of methods for the following (see word document**)
-1. PLA
-2. LMA
-4. Wood density
-5. TLP
+In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
+These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis.
+Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
+Leaf area was calculated from scanned images using an R script (**details**).
+$LMA$ was calculated as the ratio of leaf dry mass to fresh area.
+$PLA$ was calculated as the percent loss of area between fresh and dry leaves.
+$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
+We used the rapid determination method of ([Bartlett et al., 2012]) to estimate the turgor loss point ($\pi_{tlp}$).
+Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
+Osmotic potential ($\pi_{osm}$ given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm}^{-0.631}$ ([Bartlett et al., 2012]).
+We also characterized hydraulic vulnerability curves for the # most productive species, but because P50 and P80 [**define**] did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
+Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
 
-**Table 3.
-Species analyzed here, listed in descending order of ANPP_stem.
-n cores and DBH range represented, and species traits** [*This replaces/combines the two remaining tables in this section.
-Suggested columns, with those to include only if they fit in parentheses: species, (stems >=10 cm per ha in plot), (ANPP_stem), n cores, DBH range of cores, (n cores in each crown position) species means for each trait]
 ```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
@@
 library(kableExtra)
-table2 <- read.csv("tables_figures/Table2.csv")
+table2 <- read.csv("tables_figures/species_table.csv")
 
@@
 
-kable(table2, booktabs=TRUE, caption = "Overview of analyzed species, detailing DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
+kable(table2, booktabs=TRUE, caption = "**Table 3.
+Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
 Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
coded Table 2

@@
 
-![**Table 2. Summary of variables** ](tables_figures/table_variables.png){width=500px}
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
+  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Independent variables", 2, 21, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
+  #kable_styling("striped", full_width = F) %>%
+  #add_indent(c(6,7,9,13, 15,16,17,18,19,20,21))
+  kable_styling()
+```
 

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
clean up (mainly Table 1)

@@
   pack_rows("H1.2. Drought resistance decreases with crown exposure.", 4, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Small trees with lower root volume suffer proportionally more in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H2.1. Species traits predict drought resistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
@@
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(1, width = "15cm") %>%
+  column_spec(1, width = "12cm") %>%
   column_spec(2:5, width = "1cm") %>%
@@
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
-Osmotic potential ($\pi_{osm}$ given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm}^{-0.631}$ ([Bartlett et al., 2012]).
+Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ ([Bartlett et al., 2012]).
 We also characterized hydraulic vulnerability curves for the # most productive species, but because P50 and P80 [**define**] did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
@@
 
-\raggedright
+(**add description of climate data used in Fig. 1, NEON vertical profiles**)
 
 *Identification of drought years*
-
-[**add description of climate data used in Fig. 1, NEON vertical profiles**]
 
@@
 
-*Analysis*
+*Analysis*  (**this needs work**)
 
-Once the data was collected, linear mixed models were run following the order of the hypotheses as seen in Figure ???
+Linear mixed models were run following the order of the hypotheses as seen in Figure ???
 [individual_tested_traits].
-Using the [@R-pointRes] package, we set up models with the resistance value as the response variable, and each prediction's variable as the independent variable.
+Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
 Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions.

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
worked on table 2
still can't figure out how to correctly represent math symbols
issue #48

@@
 Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 9) %>%
   pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   column_spec(6, width = "2cm") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
   kableExtra::landscape()
@@
 hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = TRUE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
+  column_spec(4, width = "7cm") %>%
   pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 21, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
-  #kable_styling("striped", full_width = F) %>%
-  #add_indent(c(6,7,9,13, 15,16,17,18,19,20,21))
+  pack_rows("Independent variables", 2, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()
@@
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is**]
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is (plus citations)**]
 
@@
 In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999.
-*Some* trees exhibited increased growth: (resistance > 1.0): #% in 1964-66, #% un 1977, and #% in 1999.
+*Some* trees exhibited increased growth: (resistance > 1.0): #% in 1966, #% un 1977, and #% in 1999.
 Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
minor fixes

@@
   pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%

### 2019-09-18 manuscript/climate_sensitivity_draft.Rmd
specified climate data source

@@
 Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site.
-We considered two metrics of moisture deficit: regional Palmer Drought Severity Index (PDSI) values [**source- NOAA-same as Helcoski**] and the difference between potential evapotranspiration (PET) and precipitation (PRE) [**source- same as Helcoski**].
+We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE).
+These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris etal., 2014).
 The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
added references to intro

@@
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (**REFS(KAT)**)--for the observed greater drought growth reductions of large trees.
-First, taller trees facea greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (), and this may become a greater liability during drought (**REFS(KAT)**).
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (**REFS(KAT)**).
+This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E.
+M., Hicke, J.
+A.
+& Meddens, A.
+J.
+H.
+Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States.
+Glob.
+Change Biol.
+17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees.
+First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N.
+G., Bond, B.
+J., Hill, L., Ryan, M.
+G.
+& Whitehead, D.
+in Size and age related changes in tree structure and function (eds Meinzer, F.C.
+& Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N.
+G.
+& Allen, C.
+D.
+Darcy’s law predicts widespread forest mortality under climate warming.
+Nat.
+Clim.
+Change 5, 669–672 (2015); Ryan, M.
+G., Phillips, N.
+& Bond, B.
+J.
+The hydraulic limitation hypothesis revisited.
+Plant Cell Environ.
+29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al.
+2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)).
+Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@suarez_factors_2004]; [@scharnweber_confessions_2019].
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS(KAT)**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
 A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al.
@@
 2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
-However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; **REF**) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs (**REFS**), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al.
+However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al.
+2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al.
+2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al.
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
worked on results

@@
 
-*Descriptions of Droughts*
+*Focal droughts and their characteristics*
 
 ![**Figure 1.
-** (a) Time series of peak growing season (May-August) climate conditions and residual chronologies for each species.
-Droughts analyzed here are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
 Figure modified from [@helcoski_growing_2019].
-(b) density plots of community- wide resitance values for each drought.](tables_figures/Figure1.png){width=500px}
+Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
 
@@
 Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(**Cite Shenanadoah paper, if accepted**).
+(*Cite Shenanadoah paper, if accepted*).
 Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
@@
 1b).
-In each drought, roughly 30% of the cored trees suffered growth reductions of 30% or more (resistance <= 0.7): #% in 1966, #2% un 1977, and #% in 1999.
-*Some* trees exhibited increased growth: (resistance > 1.0): #% in 1966, #% un 1977, and #% in 1999.
+In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999.
+*Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999.
 Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).
@@
 
-**Results for first main question: what drives the observed tendency for large trees to suffer more during drought?**
+*Tree size and drought resistance*
+
+Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
+The same held true for height as a univariate predictor (*H1.1*; Tables 1, 4).
+When combined with other predictor variariables in our multivariate models, the top models usually included an effect of height, and its coefficient was consistently negative, as predicted (Tables 1, 5).
+We note that a non-signficant positive correlation between $H$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
 H1.0, H1.1, H1.2, H1.3
@@
 
-![Height profile graphs](tables_figures/Figure2.png){width=500px}
+![**Figure 2.
+Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity (RH), (c) air temperature, and (d) biological temperature averaged over each month of the peak growing season (May-August) from YEAR-YEAR.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$.
+In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
+In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
 

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
removed H1.1b
not significant, not an appropriate test (because height is calculated from DBH).

@@
   pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 4, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.1. Species traits predict drought resistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 12, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Responses varied by drought.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. Predictor variables varied across droughts.", 19, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.1. Responses varied by drought.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H3.2. Predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
worked on results

@@
 
+*Tree size and drought resistance*
+
+Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
+The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4).
+When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
+We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
+
+In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
+2).
+Specifically, [**add content on vertical profiles of climate variables**]
+
 ![Table 4 - individual variable tests](tables_figures/Table4.png) #this is tested_traits_all
@@
 
-*Tree size and drought resistance*
+Crown position was generally correlated with $H$, but with substantial variation (Fig.
+2e).
+Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
+When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$.
+When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$.
+Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
+Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
+Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.
 
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
-The same held true for height as a univariate predictor (*H1.1*; Tables 1, 4).
-When combined with other predictor variariables in our multivariate models, the top models usually included an effect of height, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $H$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
-H1.0, H1.1, H1.2, H1.3
-DBH, height, crown position, and TWI
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
+*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
 
@@
 
-**Results for second main question: how do species' traits - alone and in interaction with tree size - influence drought response? **
+*Species' traits and drought resistance*
 H2.1, H2.2, H2.3

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
references, minor edits to text
-reviewed some other notes/files, added notes and references as appropriate (see also refs to review/cite in issue #56 )
-added pdfs for some references

@@
 **One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
-2019, **REFS**).
+2019, Hacket-Pain et al.
+2016 (DOI 10.1007/s10342-016-0982-7) **REFS**).
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
@@
 2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
-Their ability to predict tree performance under drought remains untested.
+There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al.
+2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested.
+(**But see D'Orangeville et al.
+2018--except that it may not be a well-constructed test for traits**)
 
-**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, yet our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].
-Our knowledge of forest drought responses tends to come disproportionately from extreme droughts with dramatic impacts (*e.g.*, [Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**), and our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
+**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand ([D'Orangeville et al.
+2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096)).
+However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
+(**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011];  D'Orangeville et al.
+2018 **REFS**), these don't test for differential trait effects across droughts (D'Orangeville et al.
+2018) and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
+The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [Allen et al.
+2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**).
+Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
 While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
@@
 2019 (I think!)*).
-[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, it is also likely that tree size and traits interact with drought characteristics to result in differential responses across droughts--but this has yet to be tested.
+[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
+(**PELTIER**)
 
@@
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ ([Bartlett et al., 2012]).
-We also characterized hydraulic vulnerability curves for the # most productive species, but because P50 and P80 [**define**] did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
+We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
 Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
@@
   - aren't able to analyze historical forest community, nor trees that were killed by these droughts [aka we don't have data for individuals that were most severely affected] (though we found that there's little variation in climate sensitivity for trees that were cored dead vs cored alive).
-  - p50/p80
+  - we characterize hydraulic traits at the species level, missing known variation with height ([Couvreur et al.
+2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU)).
+thus, *H2.2* is not a great test.
   - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
@@
   - and we show that height is more important than exposure, but doesn't eliminate the effect of it
+  - we see no indication that rooting volume gives a relative advantage to large trees.
+Root systems are important, but [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071) show that "the limiting physiological processes for transpiration shifted from operating at the plant–soil interface to the plant–atmosphere interface with progressing forest succession.""
   - further contribution is we show that two leaf hydraulic traits - relatively easy to measure - are good predictors of drought response, can be helpful for scaling up (e.g.
@@
 This paper mentions it would be good to have this data with respect to latent affects from drought.
-\newpage
+
+our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]
+
+No previous study has compared drought responses derived from tree-rings for so many species at one site. By comparing species at the same site, can control for other factors.
+
 ### Acknowledgements
@@
 
+\newpage
 ### Supplementary Information

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
worked on results, minor adjustments to table

@@
 **A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and ring porosity (Elliot et al.
+Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture (Elliot et al.
 2015, Friedrichs et al.
@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density, specific leaf area, and ring porosity , but that leaf hydraulic traits such as leaf area shrinkage upon dehydration and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
@@
   pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Responses varied by drought.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. Predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.1. Overall community drought resistance differed across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H3.2. The direction of responses to predictor variables was consistent across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.3. The strength of responses to predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
 *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999.
-Within the context of mixed effects models, there were no significant differences in resistance across drought years (Table 4).
 
@@
 
-In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-2).
-Specifically, [**add content on vertical profiles of climate variables**]
-
 ![Table 4 - individual variable tests](tables_figures/Table4.png) #this is tested_traits_all
@@
 Crown position was generally correlated with $H$, but with substantial variation (Fig.
-2e).
+2**e**).
 Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
-*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
+In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
+2).
+Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
+2a).
+Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig.
+2b).
+Air temperature did not vary across the vertical profile (Fig.
+2c).
+[**biological temperature?- probably drop**]
 
 ![**Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity (RH), (c) air temperature, and (d) biological temperature averaged over each month of the peak growing season (May-August) from YEAR-YEAR.
+Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), (c) air temperature ($T_{air}$), and (d) biological temperature ($T_{biological}$)averaged over each month of the peak growing season (May-August) from YEAR-YEAR.
 In these plots, heights are slightly offset for visualization purposes.
@@
 
-*Species' traits and drought resistance*
-H2.1, H2.2, H2.3
-Hydraulic traits alone, traits with height
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
+*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
 
-**Results for third main question: are responses similar or variable across individual drought years?**
-H3.1 and H3.2
-Combining biophysical with hydraulic traits, which come out as candidates for best model?
+*Species' traits, height, and drought resistance*
+
+We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
+Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.23; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
+In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
+Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts.
+$\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4).
+Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
+Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.
+
+We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
+In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha **Ian, confirm**), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**).
+However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$.
+Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
+2**f-g**).
+
+We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is *not* driven by more drought-sensitive traits in larger trees, but rather by size itself (Tables 1,5).
+As discussed above, there was little meaningful variation in traits with height at the community level.
+When $H$ and hydraulic traits were considered together in multivariate models, the effect of $H$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $H$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
+
+*Responses across droughts*
+
+We reject *H3.1*.
+Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
+This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig.
+1b).
+
+We support Despite the broad similarity in the extent to which the three droughts affected the community as a whole, there was variation as to which variables best predicted drought responses (*H3.2b*), and occasionally in the direction of these responses (*H3.2b*).
 

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
finished drafting results

@@
   pack_rows("H3.1. Overall community drought resistance differed across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. The direction of responses to predictor variables was consistent across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.3. The strength of responses to predictor variables varied across droughts.", 18, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.2. The direction of responses to predictor variables differ across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.3. The strength of responses to predictor variables varied across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
 
-We reject *H3.1*.
+We reject the hypothesis (*H3.1*) that overall community responses varied across droughts.
 Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
@@
 
-We support Despite the broad similarity in the extent to which the three droughts affected the community as a whole, there was variation as to which variables best predicted drought responses (*H3.2b*), and occasionally in the direction of these responses (*H3.2b*).
+We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts.
+In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
+However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $\pi_{tlp}$ in the multivariate model for the 1966 drought; Tables 4, 5).
+These differences may very well be random, as opposed to statistically meaningful.
+Among the univariate models, there was no instance predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
+Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4).
+The difference most likely to be real is that ring and semi-ring porous species had lower resistance than diffuse porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
+
+We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
+For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $CP$ (lowest $R$ among supressed trees) and $XP$ (lowest $R$ among diffuse-porous trees) were strongest in 1999 (Tables 4,5).
 

### 2019-09-19 manuscript/climate_sensitivity_draft.Rmd
minor updates- table 1

@@
   pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.1. Drought resistance decreases with tree height.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Drought resistance decreases with crown exposure.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
coded table 5
#48

@@
   pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Overall community drought resistance differed across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. The direction of responses to predictor variables differ across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.3. The strength of responses to predictor variables varied across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
 
-![Table 5 - best full models ](tables_figures/Table5.png)
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Multivariate models**", format = 'latex', escape = TRUE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
+  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  kable_styling()
+```
 

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
coded table 4

@@
 
-![Table 4 - individual variable tests](tables_figures/Table4.png) #this is tested_traits_all
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypothesis_table <- read.csv("tables_figures/univariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  kable_styling()
+```
 

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
minor fixes Table 4

@@
   add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
   kable_styling()

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
worked on abstract and discussion

@@
 
-### Summary
-Predicting forest responses to drought is an increasingly critical task under climate change effects.
-Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.
-In this study, we analyze the interaction between these two trait groups using forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
-Drought periods were defined by both Palmer Drought Severity Indices (PDSI) and their identification from tree-ring records for 12 species representing 97% of woody productivity.
-Each drought scenario (1966, 1977, 1999), along with the overall trend, was then tested against leaf hyraulic trait measurements and microhabitat biophysical data.
-Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
+### Summary (**needs work**)
+- Predicting forest responses to drought is an increasingly critical task under climate change effects.
+[Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.]
+
+- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60 year period from 1950 and 2009.
+
+- Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
 However, there was substantial variation in the best predictor variables across given droughts.
-We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats are likely to be most strongly affected. **Add discussion points**
+
+- We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
 
@@
 
+\newpage
 ### Introduction
@@
 
-\raggedright
-
-\newpage
 ### Results
@@
 
+### Discussion
+
+Our results reveal how tree size, microhabitat, and hydraulic traits shape tree growth responses across three droughts in a temperate deciduous forest.
+The tendencey for larger trees to suffer more was driven primarily by their height, with a marginal effect of crown exposure.
+
 \newpage
-### Discussion
 Discussion outline

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

@@
 Our results reveal how tree size, microhabitat, and hydraulic traits shape tree growth responses across three droughts in a temperate deciduous forest.
-The tendencey for larger trees to suffer more was driven primarily by their height, with a marginal effect of crown exposure.
+The tendencey for larger trees to suffer more, which has been observed in forests around the world (**REFS**), was driven primarily by their height, with a marginal effect of crown exposure.
+We find on evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more, and larger rooting volume provided no advange in the drier microenvironments.
+Together, these findings demonstrate that the height is the primary mechanism through which tree size mediates drought response.
 

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

--- 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
@@
 2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)).
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@suarez_factors_2004]; [@scharnweber_confessions_2019].
+Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
@@
 As discussed above, there was little meaningful variation in traits with height at the community level.
-When $H$ and hydraulic traits were considered together in multivariate models, the effect of $H$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $H$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
+When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $ln[H]$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
 
@@
 
-Our results reveal how tree size, microhabitat, and hydraulic traits shape tree growth responses across three droughts in a temperate deciduous forest.
-The tendencey for larger trees to suffer more was driven primarily by their height, with a marginal effect of crown exposure.
+Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
+The tendencey for larger trees to suffer more, observed here as in forests around the world (**REFS**), was driven primarily by their height.
+There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **).
+There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [Zuleta et al.](https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1&isAllowed=y)), and larger rooting volume provided no advantage in the drier microenvironments.
+The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
+
+Together, these findings demonstrate that the height is the primary mechanism through which tree size mediates drought response.
+
+This study
 
@@
   - "we filled the gap" = 1-2 sentences
+
 - limitations at our site:
@@
   - we characterize hydraulic traits at the species level, missing known variation with height ([Couvreur et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU)).
-thus, *H2.2* is not a great test.
+2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al.
+2001](https://watermark.silverchair.com/21-12-13-951.pdf)).
+thus, *H2.2* is not a great test, and there may be interactions between height and traits that we're not capturing.
+What we show is that the canopy isn't dominated by more dorught sensitive species.
   - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
   - We did not use crowding index because it has much more uncertainty
+
 - limitations for extrapolating other sites:
@@
   - the species may be different from other sites but the hydraulic traits are general across species;
+
 - next paragraph saying that our study advances understanding of droughts, e.g. bennett et al 2015 doesn't go into mechanisms but we do
@@
   - first study to show how these traits affect woody growth response to drought **confirmed by Lawren**
+
 - science is better now.
@@
 Forecasting forest responses to these droughts is a huge and important challenge", science is better.
+
+-vertical profile in CO2!!  [Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS
 

### 2019-09-20 manuscript/climate_sensitivity_draft.Rmd
worked on discussion, messed up table 5

@@
 hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Multivariate models**", format = 'latex', escape = TRUE) %>%
+kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5.
+Summary of R^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 ```
+
+\raggedright
 
@@
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
+Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
+These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought...
 
-Together, these findings demonstrate that the height is the primary mechanism through which tree size mediates drought response.
+The droughts considered here were of similar severity (Fig.
+1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
+1a, @helcoski_growing_2019).
+Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature.
+While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
+1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
+It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5).
+Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
+These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [Allen et al.
+2015](https://doi.org/10.1016/j.foreco.2009.09.001)), many of which have shaped our understanding about the role of tree size (Bennett et al.
+2015; Stovall et al.
+2019) and--to some extent-- traits (**REFS**).
+Nevertheless, our results are consistent with findings from more extreme droughts.
 
-This study
+Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with tree height.
+To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**).
+They also face different microenvironments, which are partially related to canopy position.
+Even under non-drought conditions, evaporative demand increases with tree height (Fig.
+1a-b; see also **REFS**.
+**The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Furthermore, daytime $CO_2$ concentrations tend to be lower in the canopy than in the understory **(Fig.
+2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al.
+2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake.
+**[McDowell-vertical profiles in isotopes?*]**.
+Taller trees are more likely to be in co-dominant or domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
+2**e?**) ([Mueller-Landau et al.
+2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)).
+**[discuss $CP$] **
 
 \newpage
-Discussion outline
 
-- first paragraph is summary of main findings, not re-presenting the results
-  - "we supported this hypothesis, but not this one" (following same order as Table 1 still)
-  - direction of responses seems to be relatively consistent but the individual responses vary
-- tie things in (see here
-  - main thing is that we're now better understanding what confers vulnerability or resilience on trees during drought - how forests respond to future droughts/climate change
-  - "we filled the gap" = 1-2 sentences
+### new discussion outline
 
-- limitations at our site:
-  - aren't able to analyze historical forest community, nor trees that were killed by these droughts [aka we don't have data for individuals that were most severely affected] (though we found that there's little variation in climate sensitivity for trees that were cored dead vs cored alive).
+1. summary of main findings
+
+2. droughts were fairly week
+
+3. height & canopy position - biophysical factors
+  - and we show that height is more important than exposure, but doesn't eliminate the effect of it
+  - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
+  - We did not use crowding index because it has more uncertainty
+  - we see no indication that rooting volume gives a relative advantage to large trees.
+Root systems are important, but [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071) show that "the limiting physiological processes for transpiration shifted from operating at the plant–soil interface to the plant–atmosphere interface with progressing forest succession.""
+
+4. tree size - biological factors
+- Beyond microenvironment, multiple tree characteristics and scale with tree size (e.g., [Anderson-Teixeira et al.
+2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470)) and vary with vertical position in a tree (**REFS**)...
  - we characterize hydraulic traits at the species level, missing known variation with height ([Couvreur et al.
@@
 What we show is that the canopy isn't dominated by more dorught sensitive species.
-  - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
-  - We did not use crowding index because it has much more uncertainty
+ - "Together, these findings demonstrate that the height is the primary mechanism through which tree size mediates drought response.""
 
-- limitations for extrapolating other sites:
-  - forests are different from place to place, but we've seen how in forests around the world, forests tend to suffer more. We've identified some facets for why this happens. (cite other studies)
-  - has been observed elsewhere that individuals in more moist habitats are more susceptible to drought (cite other studies)
-  - the species may be different from other sites but the hydraulic traits are general across species;
-
-- next paragraph saying that our study advances understanding of droughts, e.g. bennett et al 2015 doesn't go into mechanisms but we do
-  - and we show that height is more important than exposure, but doesn't eliminate the effect of it
-  - we see no indication that rooting volume gives a relative advantage to large trees.
-Root systems are important, but [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071) show that "the limiting physiological processes for transpiration shifted from operating at the plant–soil interface to the plant–atmosphere interface with progressing forest succession.""
+5. we advance knowledge regarding hydraulic traits
+- the fact that  $PLA_{dry}$ and $\pi_{tlp}$ predict drought response is physiologically logical (**REFS**) but scientifically novel in that this is the first study to demonstrate that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought response.
+(*confirmed by Lawren*)
    - further contribution is we show that two leaf hydraulic traits - relatively easy to measure - are good predictors of drought response, can be helpful for scaling up (e.g.
 from our site to eastern deciduous biome)
-  - biophysical mechanisms are things that should be seen as universal but relative importance of each can vary within each drought
-  - first study to show how these traits affect woody growth response to drought **confirmed by Lawren**
+  - tree species are variable around the world and very diverse in the tropics, and hydraulic traits give us the opportunity the generalize this response;
 
-- science is better now.
-by advancing our understanding of the mechanisms for individual-level responses to drought, this opens door for better predictions (elaborate from above).
+6. we advance knowledge regarding differences across droughts
+  - biophysical mechanisms are things that should be seen as universal but relative importance of each can vary within each drought (or across sites)
+
+7. science is better
+- by advancing our understanding of the mechanisms for individual-level responses to drought, this opens door for better predictions (elaborate from above).
 "This is absolutely critical to predicting forest responses to future droughts, which we're likely to see more of in the future as a result to climate change.
-Forecasting forest responses to these droughts is a huge and important challenge", science is better.
-
--vertical profile in CO2!!  [Koike et al. 2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS
+Forecasting forest responses to these droughts is a huge and important challenge",
+- furthermore, we have deepened understading of which trees will suffer , and there are implications.
+- big trees
 
 *1. paragraph summarizing main results--> primary conclusions*
-When including only biophysical traits, trees' resistance value (on a per-species basis) is explained best by crown position and height, with codominant trees being the most resistant to drought.
-This follows on work done by [@bennett_larger_2015] [and others?] which show that larger trees suffer more during drought, and confirms that this susceptibility can be seen in tree ring analyses.
-Adding in crown position with the leaf hydraulic traits yields a slightly worse predictive model for drought tolerance, with height remaining as the only significant biophysical variable.
 
@@
 Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)
-
-Proximity to stream--either vertical (elev) or horizontal (distance)-- did not increase drought resistance; rather, it tended to decrease resistance (H1.3a).
-This may be because individuals growing further from water are acclimatized to drier conditions.
-However, the increase in drought resistance with distance from stream was less for small than large trees (H1.3b), indicating a potential importance of root depth/volume in conferring drought resistance.
 

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

@@
 **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime $CO_2$ concentrations tend to be lower in the canopy than in the understory **(Fig.
+Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig.
 2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al.
@@
 **[McDowell-vertical profiles in isotopes?*]**.
-Taller trees are more likely to be in co-dominant or domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
+Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
 2**e?**) ([Mueller-Landau et al.
 2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)).
-**[discuss $CP$] **
+We show that height is a far stronger predictor than crown position (Tables 1,4,5).
+Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
+However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**).
+The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
 
@@
 3. height & canopy position - biophysical factors
-  - and we show that height is more important than exposure, but doesn't eliminate the effect of it
-  - We used crown position despite its uncertainty. However, height and crown position change relatively slowly so it shouldn't be that far off.
-  - We did not use crowding index because it has more uncertainty
-  - we see no indication that rooting volume gives a relative advantage to large trees.
-Root systems are important, but [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071) show that "the limiting physiological processes for transpiration shifted from operating at the plant–soil interface to the plant–atmosphere interface with progressing forest succession.""
 

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
worked on discussion + word count

@@
 
+### Title page
+
+text | word count | other | n
+---------------------------------------------|---------------------|-----------------------------|--------------------
+Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
+Summary|  	## (limit 200) | 	No. of Tables | 	5  (**check if there's a figures/tables limit for regular research papers**- I don't see one)
+Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
+Materials and Methods |	currently ~1442	||	 
+Results |	currently ~1437	 ||
+Discussion |	currently ~1339	+ ~200 unwritten (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
+Acknowledgements | 	~75 unwritten	||
+
+\newpage
+
 ### Summary (**needs work**)
@@
 It should not contain citations of other papers.*
+
+*Key words*: [5-8] *Five to eight key words (in alphabetical order) .
+Words that are in the title can, and should, be among these.
+Very short phrases and scientific names with their common equivalents (e.g.
+Nicotiana tabacum (tobacco)) are acceptable.*
 
@@
 
-Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with tree height.
+Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height.
 To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**).
@@
 
+Beyond microenvironment, multiple tree characteristics covary with height.
+To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ...
+(e.g., [Anderson-Teixeira et al.
+2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470); **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**).
+These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al.
+2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
+However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
+Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ...
+(**describe differences**) ([Couvreur et al.
+2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al.
+2001](https://watermark.silverchair.com/21-12-13-951.pdf)).
+Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
+On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height.
+In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
+Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses (*Couvreur et al.
+2018*).
+
+Our finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+(**REFS**).
+It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species.
+The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).
+
 \newpage
@@
 4. tree size - biological factors
-- Beyond microenvironment, multiple tree characteristics and scale with tree size (e.g., [Anderson-Teixeira et al.
-2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470)) and vary with vertical position in a tree (**REFS**)...
- - we characterize hydraulic traits at the species level, missing known variation with height ([Couvreur et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al.
-2001](https://watermark.silverchair.com/21-12-13-951.pdf)).
-thus, *H2.2* is not a great test, and there may be interactions between height and traits that we're not capturing.
-What we show is that the canopy isn't dominated by more dorught sensitive species.
- - "Together, these findings demonstrate that the height is the primary mechanism through which tree size mediates drought response.""
 
 5. we advance knowledge regarding hydraulic traits
-- the fact that  $PLA_{dry}$ and $\pi_{tlp}$ predict drought response is physiologically logical (**REFS**) but scientifically novel in that this is the first study to demonstrate that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought response.
-(*confirmed by Lawren*)
-   - further contribution is we show that two leaf hydraulic traits - relatively easy to measure - are good predictors of drought response, can be helpful for scaling up (e.g.
-from our site to eastern deciduous biome)
-  - tree species are variable around the world and very diverse in the tropics, and hydraulic traits give us the opportunity the generalize this response;
 
-6. we advance knowledge regarding differences across droughts
+6. we advance knowledge regarding differences across droughts ??
   - biophysical mechanisms are things that should be seen as universal but relative importance of each can vary within each drought (or across sites)

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
changed title, added key words

@@
 ---
-title: "Tree size, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
+title: "Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
 author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
@@
 
-*Key words*: [5-8] *Five to eight key words (in alphabetical order) .
+*Key words*:  canopy position; drought; hydraulic traits; microhabitat; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
 Words that are in the title can, and should, be among these.

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
minor edits

@@
 
-*Key words*:  canopy position; drought; hydraulic traits; microhabitat; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
+*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
 Words that are in the title can, and should, be among these.

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

@@
 
+Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
+[**DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought.
+...**] Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
+
+Our study reveals that tree height, microhabitat, and species' hydraulic traits interactively shape tree growth responses to drought.
+
 \newpage
@@
 
-6. we advance knowledge regarding differences across droughts ??
-  - biophysical mechanisms are things that should be seen as universal but relative importance of each can vary within each drought (or across sites)
+6. we advance knowledge regarding differences across droughts
 

### 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
discussion

@@
 
-Our study reveals that tree height, microhabitat, and species' hydraulic traits interactively shape tree growth responses to drought.
+As climate change drives increasing drought in many of the world's forests (*REFS*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
+Tall trees with exposed crowns are suffering disroportionately,
 
@@
 - big trees
+- traits/ successional status?
 

### 2019-09-22 manuscript/climate_sensitivity_draft.Rmd
finished drafting discussion!

--- 2019-09-21 manuscript/climate_sensitivity_draft.Rmd
@@
 Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	## (limit 200) | 	No. of Tables | 	5  (**check if there's a figures/tables limit for regular research papers**- I don't see one)
+Summary|  	## (limit 200) | 	No. of Tables | 	5  
 Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
@@
 **A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture (Elliot et al.
+Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al.
 2015, Friedrichs et al.
@@
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
 The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
-These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought...
+These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.
 
@@
 They also face different microenvironments, which are partially related to canopy position.
-Even under non-drought conditions, evaporative demand increases with tree height (Fig.
-1a-b; see also **REFS**.
+Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig.
+1a-b; **REFS- Jarvis 1984?**; [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)).
 **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
@@
 The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
+Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
 
@@
 
-Our finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
+Concerted measurement of leaf hydraulic traits of emerging importance (**REFS--KAT/NOBBY/LAWREN**) allowed novel insights into the role of hydraulic traits in shaping drought response.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
 (**REFS**).
@@
 [**DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought.
-...**] Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
+...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).
+Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].
+Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
 
-Our study reveals that tree height, microhabitat, and species' hydraulic traits interactively shape tree growth responses to drought.
-
-\newpage
-
-### new discussion outline
-
-1. summary of main findings
-
-2. droughts were fairly week
-
-3. height & canopy position - biophysical factors
-
-4. tree size - biological factors
-
-5. we advance knowledge regarding hydraulic traits
-
-6. we advance knowledge regarding differences across droughts
-
-7. science is better
-- by advancing our understanding of the mechanisms for individual-level responses to drought, this opens door for better predictions (elaborate from above).
-"This is absolutely critical to predicting forest responses to future droughts, which we're likely to see more of in the future as a result to climate change.
-Forecasting forest responses to these droughts is a huge and important challenge",
-- furthermore, we have deepened understading of which trees will suffer , and there are implications.
-- big trees
-
-*1. paragraph summarizing main results--> primary conclusions*
-
-We partially supported the hypothesis that crown exposure makes trees more vulnerable to drought.
-Co-dominant trees had the highest drought resistance.
-Dominant trees had lower resistance, likely because they are the most exposed.
-Other studies have found clear evidence of greater drought sensitivity in trees with exposed crowns (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-At the same time, intermediate and suppressed trees had even lower resistance.
-This indicates that other mechanisms such as competition or rooting depth were important.
-(Also note that our study design was not ideal for testing the role of canopy position.
-Current canopy position is a conservative separator of canopy position: trees may currently be in more dominant positions than they were at the time, but backwards movement is unlikely.
-This would bias against finding a signficant effect for H1.2.
-Height may be a more reliable predictor of past canopy position than is current canopy position, and explains a portion of variation in canopy position.)
-
-**misc content to integrate**
-From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity.
-This paper mentions it would be good to have this data with respect to latent affects from drought.
-
-our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019]
-
-No previous study has compared drought responses derived from tree-rings for so many species at one site. By comparing species at the same site, can control for other factors.
+As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
+Large trees have been suffering disrpoportionately in forests around the world (*REFS*), and we here show that this is primarily driven by their height, with some contributions from canopy position.
+The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
+Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
+The later successional species at our site (*Fagus grandifolia*, ...
+) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
+In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
 

### 2019-09-22 manuscript/climate_sensitivity_draft.Rmd
minor changes

@@
 Results |	currently ~1437	 ||
-Discussion |	currently ~1339	+ ~200 unwritten (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
+Discussion |	currently ~1881 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
 Acknowledgements | 	~75 unwritten	||
@@
 ### Summary (**needs work**)
-- Predicting forest responses to drought is an increasingly critical task under climate change effects.
-[Part of the problem is due to the lack of studies analyzing the confluence of leaf hydraulic traits with biophysical parameters.]
+- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important.
+Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.
 
@@
 Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-[**DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought.
+[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought.
 ...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).

### 2019-09-27 manuscript/climate_sensitivity_draft.Rmd
updated with Stovall reference

@@
 2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**).
-Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; Stovall et al.
-2019), resulting in disproportionate impacts on forest C storage ([Meakem et al.
+Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [Stovall et al.
+2019]( https://doi.org/10.1038/s41467-019-12380-6)), resulting in disproportionate impacts on forest C storage ([Meakem et al.
 2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
@@
 While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
-There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; *Stovall et al.
-2019 (I think!)*).
+There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; Stovall et al.
+2019).
 [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
@@
 Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- e.g., 1966 was preceded by two dry years, and height mattered most in this drought.
+[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase (Stovall et al.
+2019).
 ...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).

### 2019-10-06 manuscript/climate_sensitivity_draft.Rmd
Tried updating symbols

@@
 hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypothesis_table$symbol[14] <- "X"
+hypothesis_table$symbol[17] <- "T"
+
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = TRUE) %>%
@@
 ```
+Caption: T = $\pi_{tlp}$, X = ${XP}$
 

### 2019-10-06 manuscript/climate_sensitivity_draft.tex
Tried updating symbols

@@
 \hypersetup{unicode=true,
-            pdftitle={Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest},
-            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira},
+            pdftitle={Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest},
+            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira},
             pdfborder={0 0 0},
@@
 \bibliographystyle{apalike}
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 % Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
+\providecommand{\subtitle}[1]{
   \posttitle{
@@
 
-  \title{Tree size and hydraulic traits shape drought responses in a temperate
-broadleaf forest}
+  \title{Tree height, microhabitat, and hydraulic traits shape drought responses
+in a temperate broadleaf forest}
     \pretitle{\vspace{\droptitle}\centering\huge}
   \posttitle{\par}
-    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine
-Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack,
-Krista Anderson-Teixeira}
+    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika
+Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall,
+Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina
+Anderson-Teixeira}
     \preauthor{\centering\large\emph}
@@
 \captionsetup[table]{labelformat=empty}
+\captionsetup[figure]{labelformat=empty}
 \usepackage{dcolumn}
@@
 
-\subsubsection{Summary}\label{summary}
-
-Predicting forest responses to drought is an increasingly critical task
-under climate change effects. Part of the problem is due to the lack of
-studies analyzing the confluence of leaf hydraulic traits with
-biophysical parameters. In this study, we analyze the interaction
-between these two trait groups using forest census data from a 25.6-ha
-ForestGEO plot in Virginia (USA). Drought periods were defined by both
-Palmer Drought Severity Indices (PDSI) and their identification from
-tree-ring records for 12 species representing 97\% of woody
-productivity. Each drought scenario (1964-66, 1977, 1999), along with
-the overall trend, was then tested against leaf hyraulic trait
-measurements and microhabitat biophysical data. Individual-level growth
-responses to the three individual droughts were stronger among taller
-trees in dominant canopy positions, those in wetter microsites, and for
-more drought-sensitive species as assessed by leaf traits (turgor loss
-at less negative leaf water potential, greater shrinkage with leaf
-dehydration), with substantial variation in the best predictor variables
-across given droughts. We conclude that when droughts occur, large
-dominant trees, drought-sensitive species, and individuals in wetter
-microhabitats are likely to be most strongly affected.
+\raggedright
+
+\hypertarget{title-page}{%
+\subsubsection{Title page}\label{title-page}}
+
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.35\columnwidth}\raggedright
+text\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright
+word count\strut
+\end{minipage} & \begin{minipage}[b]{0.22\columnwidth}\raggedright
+other\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\raggedright
+n\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Total word count (excluding summary, references and legends)\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+currently \textasciitilde{}6000 (strict limit 6,500)\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No.~of figures\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+2 (both colour)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Summary\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+\#\# (limit 200)\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No.~of Tables\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+5\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Introduction\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+currently \textasciitilde{}1536\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No of Supporting Information files\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\#\#\#.\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Materials and Methods\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+currently \textasciitilde{}1442\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Results\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+currently \textasciitilde{}1437\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Discussion\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+currently \textasciitilde{}1881 (limit 30\% of total (not strict), or
+1950 if manuscript reaches word limit)\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Acknowledgements\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+\textasciitilde{}75 unwritten\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
+
+\newpage
+
+\hypertarget{summary-needs-work}{%
+\subsubsection{\texorpdfstring{Summary (\textbf{needs
+work})}{Summary (needs work)}}\label{summary-needs-work}}
+
+\begin{itemize}
+\item
+  As the climate changes, driving increased drought in many forested
+  regions around the world, mechanistic understanding of factors
+  conferring drought vulnerability and resistance in trees is
+  increasingly important. Yet it remains unclear how tree size and
+  species' traits interactively shape tree growth responses across
+  droughts.
+\item
+  In this study, we analyze tree-ring records for 12 species
+  representing 97\% of woody productivity in the 25.6-ha ForestGEO plot
+  in Virginia (USA) to determine how tree size, microhabitat, and
+  species' traits interactively shape drought responses across the three
+  strongest droughts over the 60 year period from 1950 and 2009.
+\item
+  Individual-level growth responses to the three individual droughts
+  were stronger in three cases: taller trees in dominant canopy
+  positions, trees in wetter microsites, and more drought-sensitive
+  species as assessed by leaf traits (turgor loss at less negative leaf
+  water potential, greater shrinkage with leaf dehydration). However,
+  there was substantial variation in the best predictor variables across
+  given droughts.
+\item
+  We conclude that when droughts occur, large dominant trees,
+  drought-sensitive species, and individuals in wetter microhabitats
+  tend to be most strongly affected.
+\end{itemize}
 
@@
 
-\subsubsection{Introduction}\label{introduction}
-
-Understanding how and why trees respond to drought is critical to
-predicting forest drought responses and climate change feedbacks. (1
-paragraph on this- I have lots of content on this and will add later.)
-
-{[}Understanding forest responses to drought requires increased
-functional understanding of the factors that confer individual-level
-vulnerability or resistance.{]} Forests are diverse in terms of tree
-sizes and functional traits, and it is known that trees varying in size
-and functional traits respond differently to drought (e.g.,
-\citep{bennett_larger_2015}; REFS). Therefore, in order to understand
-whole-forest response to drought, we need a functional understanding of
-how responses vary by tree size, microhabitat, and species. There are 3
-fundamental questions that must be addressed:
-
-\emph{First, what drives the observed tendency for large trees to suffer
-more during drought?}\\
-\citet{bennett_larger_2015} showed that in forests globally, large trees
-suffer greater growth reductions during drought. However, this analysis
-quantified tree size based on DBH, which has no direct mechanistic
-meaning. This study proposed two major mechanisms (besides insects) for
-the observed greater drougth growth reductions of large trees: (1)
-inherently greater biophysical challenge of being tall; (2) greater
-exposure of the crowns of large trees. Counteracting these effects, (3)
-the larger root systems of larger trees may confer an advantage in terms
-of allowing greater access to water, but it appears that this effect is
-usually insufficient to offset the costs of height and/or crown
-exposure.
-
-Canopy trees have lower drought resistance because they are exposed to
-higher solar radiation, greater wind speeds, and lower humidity.
-Alternatively, the generally supressed status of subcanopy trees may be
-insufficient to override the benefits of their buffered environment
-during drought.
-
-\emph{Second, how do species' traits - alone and in interaction with
-tree size - influence drought response?} Analyzing drought responses on
-the species level does not fully explain mechanisms and is not feasible
-in diverse forests. The solution is a trait-based approach. Leaf
-hydraulic traits hold more promise than more
-commonly/traditionally-measured traits such as wood density and specific
-leaf area (SLA)
+\emph{Key words}: canopy position; drought; Forest Global Earth
+Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous
+forest; tree growth; tree height; tree-ring {[}5-8{]} \emph{Five to
+eight key words (in alphabetical order) . Words that are in the title
+can, and should, be among these. Very short phrases and scientific names
+with their common equivalents (e.g.~Nicotiana tabacum (tobacco)) are
+acceptable.}
+
+\newpage
+
+\hypertarget{introduction}{%
+\subsubsection{Introduction}\label{introduction}}
+
+Forests globally play a critical role in climate regulation
+(\href{https://science.sciencemag.org/content/320/5882/1444}{Bonan
+2008}), yet there remains enormous uncertainty as to how the terrestrial
+carbon (C) sink, which is dominated by forests, will respond to climate
+change
+(\href{https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1}{Friedlingstein
+et al.~2006}). An important aspect of this uncertainty lies in responses
+to drought (\textbf{REF}). In many forested regions around the world,
+the risk of severe drought is increasing
+(\href{https://www.nature.com/articles/nclimate2067}{Trenbert et
+al.~2014}), even in conjunction with increasing precipitation
+(\href{https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334}{IPCC
+2014}). Global change-type drought has been affecting forests worldwide
+(\href{https://doi.org/10.1016/j.foreco.2009.09.001}{Allen et
+al.~2015}), and it is expected that future climate change-driven
+droughts will severely impact forests around the world
+(\href{https://doi.org/10.1016/j.foreco.2009.09.001}{Allen et al.~2015};
+\textbf{REFS}). Larger trees tend to suffer more (e.g.,
+\citep{bennett_larger_2015};
+\href{https://doi.org/10.1038/s41467-019-12380-6}{Stovall et al.~2019}),
+resulting in disproportionate impacts on forest C storage
+(\href{https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633}{Meakem
+et al.~2018}). As a result, forest drought responses stand to strongly
+impact forest feedbacks to climate change (\textbf{REFS}), yet accurate
+characterization of drought responses remains a modeling challenge
+(\textbf{REFS})-- in part because some of the mechanisms underlying
+drought responses remain unclear. Understanding forest responses to
+drought requires increased functional understanding of how tree size,
+microhabitat, and species' traits jointly confer individual-level
+vulnerability or resistance, and the extent to which their influence is
+consistent across droughts.
+
+\textbf{One fundamental question regarding forest responses to drought
+is what drives the observed tendency for large trees to suffer more
+during drought.} \citet{bennett_larger_2015} showed that in forests
+globally, large trees suffer greater growth reductions during drought,
+and numerous subsequent studies have reinforced this finding (Stovall et
+al.~2019, Hacket-Pain et al.~2016 (DOI 10.1007/s10342-016-0982-7)
+\textbf{REFS}). However, this analysis quantified tree size based on
+DBH, which has no direct mechanistic meaning. This study proposed two
+major mechanisms--besides the tendency for bark beetles to
+preferentially attack larger trees (\emph{Pfeifer, E. M., Hicke, J. A.
+\& Meddens, A. J. H. Observations and modeling of aboveground tree
+carbon stocks and fluxes following a bark beetle outbreak in the western
+United States. Glob. Change Biol. 17, 339--350 (2011).})--for the
+observed greater drought growth reductions of large trees. First, taller
+trees face greater biophysical challenge of lifting water greater
+distances against the effects of gravity and friction (* McDowell, N.
+G., Bond, B. J., Hill, L., Ryan, M. G. \& Whitehead, D. in Size and age
+related changes in tree structure and function (eds Meinzer, F.C. \&
+Niinemets, U.) 255--286 (Springer Publishing, 2011); McDowell, N. G. \&
+Allen, C. D. Darcy's law predicts widespread forest mortality under
+climate warming. Nat. Clim. Change 5, 669--672 (2015); Ryan, M. G.,
+Phillips, N. \& Bond, B. J. The hydraulic limitation hypothesis
+revisited. Plant Cell Environ. 29, 367--381 (2006).*), and this may
+become a greater liability during drought
+(\href{https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation}{Zhang
+et al.~2009}). Second, larger trees may have lower drought resistance
+because they are more often in the canopy, where they are exposed to
+higher solar radiation, greater wind speeds, lower humidity, and lower
+\(CO_2\) concentrations (\textbf{REFS-KAT}). Alternatively, the
+generally supressed status of subcanopy trees may be insufficient to
+override the benefits of their buffered environment during drought.
+Potentially counteracting the biophysical challenges faced by large
+trees, their larger root systems may confer an advantage in terms of
+allowing greater access to water (\textbf{REFS?}); however, it appears
+that this effect is usually insufficient to offset the costs of height
+and/or crown exposure \citep{bennett_larger_2015}. A final mechanism
+that could mediate tree size-related responses to drought is how
+hydraulic traits are distributed with respect to size
+(\href{https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633}{Meakem
+et al.~2018}). It is possible that the pattern observed by
+\citet{bennett_larger_2015} could be caused if the larger size classes
+were dominated by species less adapted to handle drought, be it through
+avoidance, resistance, or resiliance. Alternatively, larger size classes
+may be dominated by species that are better adapted to inherently
+greater biophysical challenges--as is the case in tropical moist forests
+of Panama, where larger size classes contain greater proportions of
+deciduous species
+(\href{https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572}{Condit
+et al.~2000};
+\href{https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633}{Meakem
+et al.~2018}). Understanding the mechanisms underlying the tendency for
+larger trees to suffer more during drought will require sorting out the
+interactive effects of height, canopy position, root water acess, and
+species' traits.
+
+\textbf{A second fundamental question regarding forest responses to
+drought is how species' traits -- alone and in interaction with tree
+size -- influence drought response.} To link drought responss to
+fundamental physiological characteristics, and because measuring and
+modeling drought responses of every species is infeasible in diverse
+forests, it is important to understand how traits shape drought
+responses. Commonly measured traits including wood density (\(WD\))
+(\textbf{REFS}), leaf mass per area (\(LMA\))
+\citep{abrams_adaptations_1990, guerfel_impacts_2009}, and xylem
+architecture \citep{kannenberg_linking_2019}(Elliot et al.~2015,
+Friedrichs et al.~2009) have been linked to drought responses in
+temperate deciduous forests--as well as in other forest biomes
+(\textbf{REFS}). However, these traits have less direct linkage to plant
+hydraulic function than leaf hydraulic traits such as leaf area
+shrinkage upon dessication (\(PLA_{dry}\); \emph{Scoffoni et al.~2013-
+DOI: 10.1104/pp.113.221424}) and turgor loss point
+(\(\pi_{tlp}\))--i.e., the water potential at which leaf wilting occurs
+(\href{https://www.pnas.org/content/113/46/13098.short}{Bartlett et
+al.~2016}), which are emerging as traits with potential to explain
+greater variation in plant distribution and function than the more
+commonly-measured traits such as \(WD\) and \(LMA\)
 (\href{https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229}{Medeiros
-et al. 2019}).
-
-Commonly measured traits including wood density (REFS), leaf mass per
-area \citep{abrams_adaptations_1990} \citep{guerfel_impacts_2009}, and
-ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been
-linked to drought responses and likely correlated with drought
-resistance in this forest. However, we hypothesize that leaf hydraulic
-traits such as leaf area shrinkage upon dessication (PLA) and turgor
-loss point (TLP), which are emerging as potentially more informative
-traits but whose effect on drought resistance has never been tested
-(CONFIRM), will prove better predictors.
-
-Moreover, there may be an interaction between traits and tree size. It
-is possible that the pattern observed by \citet{bennett_larger_2015}
-could be caused by smaller trees being more drought resistant.
-Alternatively, larger trees may have more drought-resistant traits as
-adpatations to greater biophysical challenges.
-
-\emph{Third, are responses similar or variable across individual drought
-years?} Droughts are rarely explicitly defined in ecological studies
-\citep{slette_how_2019}, yet no two droughts are the same.
-
-We need to understand the the factors confirming drought vulnerability
-or resistance not only for extreme droughts with dramatic impacts on
-tree growth and mortality, which tend to dominate the literature
-\citep{bennett_larger_2015} (REFS), but also for more modest but
-frequent droughts--e.g., those with historical return intervals on the
-order of a decade.
-
-Here, we combine tree-ring records covering three droughts (1964-66,
-1977, 1999), species functional and hydraulic trait measurments, and
-forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to
-test a series of hypotheses and associated specific predictions (Table
-1) designed to yield functional understanding of how tree size,
+et al.~2019}). There is also evidence that ecosystems with high
+hydraulic trait diversity--but not traits like \(LMA\) and \(WD\)-- show
+more modest ecosystem flux variation in response to drought
+(\href{https://doi.org/10.1038/s41586-018-0539-7}{Anderegg et al.~2018
+(\textbf{but review this pub})}), but the ability of hydraulic traits to
+tree performance under drought remains untested. (\textbf{But see
+D'Orangeville et al.~2018--except that it may not be a well-constructed
+test for traits})
+
+\textbf{A final fundamental question regarding forest responses to
+drought is whether tree size and species' traits have similar influence
+across droughts, or whether drought variability in factors such as
+severity, duration, and timing interact with tree size and traits such
+that different components of the community respond differently to
+different droughts.} No two droughts are the same, and tree growth
+responses vary with drought characteristics such as timing and
+atmospheric demand
+(\href{https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096}{D'Orangeville
+et al.~2018}). However, we are not aware of any studies that compare how
+tree size and species' traits mediate growth responses across droughts.
+(\textbf{BUT ARE WE MISSING SOMETHING??}) While tree-ring studies
+provide long-term records of tree responses to multiple droughts (e.g.,
+\citep{lloret_components_2011}; D'Orangeville et al.~2018
+\textbf{REFS}), these don't test for differential trait effects across
+droughts (D'Orangeville et al.~2018) and generally focus on
+species-level responses, which preclude consideration of the roles of
+tree size and microenvironment. The ecological studies that have shaped
+our understanding of the role of tree size and microenvironment in
+forest drought responses generally examine only a single drought and
+tend to focus disproportionately on extreme droughts with dramatic
+impacts (\emph{e.g.},
+\href{https://doi.org/10.1016/j.foreco.2009.09.001}{Allen et al.~2015};
+\citep{bennett_larger_2015}; Stovall; \textbf{MORE REFS}). Thus, our
+knowledge of forest responses to more modest but frequent
+droughts--e.g., those with historical return intervals on the order of a
+decade--remains more limited. While the tendency for larger trees to
+suffer more definitely predominates \citep{bennett_larger_2015}, there
+are exceptions (e.g., \textbf{REFS}). There is also evidence that the
+degree to which larger trees suffer more increases with the severity of
+drought conditions (\citep{bennett_larger_2015}; Stovall et al.~2019).
+{[}\emph{Are there any studies showing interactions of drought type with
+traits?}{]} Thus, while we expect many of the fundamental mechanisms
+shaping drought responses to be universal, we have little undertanding
+of how tree size and traits interact with drought characteristics to
+result in differential responses across droughts. (\textbf{PELTIER})
+
+Here, we combine tree-ring records covering three droughts (1966, 1977,
+1999), species functional and hydraulic trait measurments, and forest
+census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a
+series of hypotheses and associated specific predictions (Table 1)
+designed to yield functional understanding of how tree size,
 microenrivonment, and species' traits collectively shape drought
@@
 with microenvironement. We confirm that, consistent with most forests
-globally \citep{bennett_larger_2015}, larger-diameter trees have lower
-drought resistance in this forest, which is in an ecoregion represented
-by only one study in Bennett et al (2015) (H1.0). We then test
-hypotheses designed to disentangle the relative importance of tree
-height (H1.1), crown exposure (H1.2), and root water access, which
-should be greater for larger trees in dry but not perpetually wet
-microsites (H1.3). Second, we focus on the role of species' functional
-and hydraulic traits and their interaction with tree height. We
-hypothesize that drought resistance will follow predicted and observed
-patterns in relation to wood density, specific leaf area, and ring
-porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area
-shrinkage upon dehydration and turgor loss point will prove better
-predictors (P2.1d-e). We then test whether these traits correlate with
-tree height (P2.2), potentially driving the observed tendency for taller
-trees to suffer more during drought (P2.3). Finally, we focused on
-variability among droughts, asking how community resistance varied
-across droughts (H3.1) and whether the factors confirming vulnerability
-or resistance varied across droughts (H3.2).
-
-\emph{Let's put Table 1 here. We need to give the hypotheses/
-predictions here. It seems counterintuitive to put the results right up
-front like that, but it actually tends to be a higher-impact style of
-writing, and recommended
-\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619}{here}.}
-
-\textbf{Table 1. Summary of hypotheses, predictions and results}
-\includegraphics[width=5.20833in]{tables_figures/Table1.png}
-
-We count predictions as fully supported (or rejected) when the direction
-of response matches (or contradicts) the prediction. H2.3-H3.2 are based
-on if models containing species traits in H2.1 had dAIC\textgreater{}=2
-relative to the appropriate null model or to any alternative
-multivariate model within 2 dAIC. Parentheses indicate that predictions
-were partially supported (or rejected). In other words, parentheses
-indicate when the direction of response matched (or contradicted) the
-prediction in some but not all models with dAIC\textless{}2 relative to
-the appropriate null model. With categorical variables such as crown
-position, a ``(yes)'' notation describes when the trend matched the
-prediction, but it wasn't significant A ``(yes/no) or (''no/yes``)''
-indicates tendencies in opposite directions in univariate tests versus
-the best full models, respectively. ``(yes)
-
-\subsubsection{Materials and Methods}\label{materials-and-methods}
-
-\emph{Study site} Research was conducted at the 25.6 ha ForestGEO
-(Global Earth Observatory) study plot at the Smithsonian Conservation
-Biology Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°
-08'43.4''W) \citep{andersonteixeira_ctfs-forestgeo:_2015}. SCBI is
-located in the central Appalachian Mountains at the northern edge of
-Shenandoah National Park. Elevations range from 273-338m above sea level
+globally, larger-diameter trees have lower drought resistance in this
+forest, which is in an ecoregion represented by only one study in
+\citep{bennett_larger_2015} (\emph{H1.0}). We then test hypotheses
+designed to disentangle the relative importance of tree height
+(\emph{H1.1}), crown exposure (\emph{H1.2}), and root water access,
+which should be greater for larger trees in dry but not in perpetually
+wet microsites (\emph{H1.3}). Second, we focus on the role of species'
+functional and hydraulic traits and their interaction with tree height.
+We hypothesize that drought resistance will follow predicted and
+observed patterns in relation to wood density and specific leaf area,
+but that hydraulic traits including xylem architecture (\emph{i.e.},
+ring, semi-ring, or diffuse porous), leaf area shrinkage upon
+dehydration, and turgor loss point will prove better predictors
+(\emph{H2.1}). We then test whether these traits correlate with tree
+height (\emph{H2.2}), potentially driving the observed tendency for
+taller trees to suffer more during drought (\emph{H2.3}). Finally, we
+focused on variability among droughts, asking how community resistance
+varied across droughts (\emph{H3.1}) and whether the factors confirming
+vulnerability or resistance varied across droughts (\emph{H3.2}).
+
+\begin{landscape}\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-3}**Table 1.
+Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable).
+Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.}
+\centering
+\resizebox{\linewidth}{!}{
+\fontsize{8.5}{10.5}\selectfont
+\begin{tabular}{>{\raggedright\arraybackslash}p{12cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{1cm}>{\raggedright\arraybackslash}p{2cm}}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Prediction supported?} & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){2-5}
+Hypotheses \& Specific Predictions & Overall & 1966 & 1977 & 1999 & Results\\
+\midrule
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H1.0. Larger-diameter trees have lower drought resistance.}}\\
+\hspace{1em}1.0- Drought resistance decreases with DBH. & yes & yes & (yes) & (no) & Table 4\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H1.1. Tall trees have lower drought resistance.}}\\
+\hspace{1em}1.1 - Drought resistance decreases with height. & yes & yes & (yes) & (no/yes) & Tables 4, 5\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H1.2. Trees with more exposed crowns have lower drought resistance .}}\\
+\hspace{1em}1.2a - When CP is considered alone, dominant trees have lowest R. & (yes) & yes & (yes) & (no) & Table 4\\
+\hspace{1em}1.2b - In models with CP and H, dominant trees have lowest R. & (no) & (no) & (yes) & (no) & Tables 4, 5\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H1.3. Small trees (lower root volume) suffer more in drier microhabitats.}}\\
+\hspace{1em}1.3 - There is a negative interactive effect between H and TWI. & (no) & (no) & (no) & (no) & Table 4\\
+\addlinespace[1em]
+\hline
+\multicolumn{1}{l}{\textbf{H2.1. Species traits predict drought resistance.}}\\
+\hspace{1em}2.1a - WD correlates positively to drought resistant. & (no) & (no) & (no) & (yes) & Table 4\\
+\hspace{1em}2.1b - LMA correlates positively to drought resistance. & (yes) & (yes) & (no) & (yes) & Table 4\\
+\hspace{1em}2.1c - Diffuse porous species have lower drought resistance. & (yes) & (yes) & (no) & yes & Tables 4, 5\\
+\hspace{1em}2.1d  - PLA correlates negatively with drought resistance. & yes & yes & (yes) & (yes) & Tables 4, 5\\
+\hspace{1em}2.1e - TLP correlates negatively with drought resistance. & (yes) & (yes/no) & (yes) & (yes) & Tables 4, 5\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H2.2. Taller trees have more drought-resistant traits.}}\\
+\hspace{1em}2.2a - Community mean WD correlates positively to H. & no & - & - & - & Table S\#\\
+\hspace{1em}2.2b - Community mean LMA correlates positively to H. & yes & - & - & - & Table S\#\\
+\hspace{1em}2.2c - Community fraction of diffuse porous species decreases with H. & no & - & - & - & Table S\#\\
+\hspace{1em}2.2d - Community mean PLA correlates negatively to H. & no & - & - & - & Table S\#\\
+\hspace{1em}2.2e - Community mean TLP correlates negatively to H. & no & - & - & - & Table S\#\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H2.3. Size-dependent drought resistance is not driven by functional traits.}}\\
+\hspace{1em}2.3 - Effect of H is negative when traits are included in the statistical model. & yes & yes & (yes) & (yes) & Table 5\\
+\addlinespace[1em]
+\hline
+\multicolumn{1}{l}{\textbf{H3.1. Community drought resistance differs across droughts.}}\\
+\hspace{1em}3.1 - Drought year explains variation in drought resistance. & no & - & - & - & Table 4\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H3.2. The direction of responses to predictor variables differs across droughts.}}\\
+\hspace{1em}3.2 - Directions of responses to best predictor variables differ across droughts. & rarely & - & - & - & Tables 4,5\\
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{H3.3. The strength of responses to predictor variables vary across droughts.}}\\
+\hspace{1em}3.3 - Best predictor variables differ across droughts. & yes & - & - & - & Table 5\\
+\bottomrule
+\end{tabular}}
+\end{table}
+\end{landscape}
+
+\clearpage
+
+\hypertarget{materials-and-methods}{%
+\subsubsection{Materials and Methods}\label{materials-and-methods}}
+
+\emph{Study site}
+
+Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth
+Observatory) study plot at the Smithsonian Conservation Biology
+Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°08'43.4''W)
+\citep[\href{https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1}{Bourg
+et al.~2013};][]{andersonteixeira_ctfs-forestgeo:_2015}. SCBI is located
+in the central Appalachian Mountains at the northern edge of Shenandoah
+National Park. Elevations range from 273-338m above sea level
 \citep{gonzalezakre_patterns_2016} with a topographic relief of 65m
-\citep{bourg_initial_2013}. Dominant species include \emph{Liriodendron
-tulipifera}, oaks (\emph{Quercus} spp.), and hickories (\emph{Carya}
-spp.).
-
-\emph{Data collection and preparation} The SCBI ForestGEO plot was
-censused in 2008, 2013, and 2018 following standard ForestGEO protocols,
-whereby all free-standing woody stems \textgreater{}= 1cm diameter at
-breast height (DBH) were mapped, tagged, measured at DBH, and identified
-to species \citep{condit_tropical_1998}. From this census data, we used
-measurements of DBH from 2008 to calculate historical DBH (described
-below), along with tree location in the plot to determine the
-topographic wetness index. Furthermore, we used data for all stems
-\textgreater{}= 10cm to analyze functional trait composition relative to
-tree height. \emph{I added the last sentence-- not sure if you've
-described that below yet. I don't think you need to note here which data
-we use, so long as that's described in the portions where you describe
-the analysis.}
-
-We analyzed tree-ring data from the twelve species contributing most to
-woody aboveground net primary productivity (ANPP), which together
-comprised 97\% of study plot ANPP between 2008 and 2013
-\citep{helcoski_growing_2019}. Cores were obtained in 2010-2011 or
-2016-2017 from a breast height of 1.3m using a 5mm increment borer. In
-2010-2011, cores were collected from randomly selected live trees of
-species with at least 30 individuals of DBH \textgreater{}= 10cm
+\citep{bourg_initial_2013}. Dominant tree taxa include
+\emph{Liriodendron tulipifera}, oaks (\emph{Quercus} spp.), and
+hickories (\emph{Carya} spp.).
+
+\emph{Data collection and preparation}
+
+Within or just outside the ForestGEO plot, we collected data on a suite
+of variables including tree size, microenvironemnt, and species traits
+(Table 2). The SCBI ForestGEO plot was censused in 2008, 2013, and 2018
+following standard ForestGEO protocols, whereby all free-standing woody
+stems \(\le\) 1cm diameter at breast height (DBH) were mapped, tagged,
+measured at DBH, and identified to species \citep{condit_tropical_1998}.
+From this census data, we used measurements of DBH from 2008 to
+calculate historical DBH, tree location in the plot to determine the
+topographic wetness index, and data for all stems \(\ge\) 10cm to
+analyze functional trait composition relative to tree height (all
+analyses described below). Census data, which were last updated in 2018
+(\textbf{confirm}), are available through the
+\href{www.forestgeo.si.edu}{ForestGEO data portal}.
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-4}**Table 2. Summary of variables**}
+\centering
+\resizebox{\linewidth}{!}{
+\fontsize{12}{14}\selectfont
+\begin{tabular}{lll>{\raggedright\arraybackslash}p{7cm}llllll}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{3}{c}{observed values} & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){7-9}
+variable & symbol & units & description & category & n & median & min & max & ln-transformed?\\
+\midrule
+\addlinespace[1em]
+\multicolumn{1}{l}{\textbf{Dependent variable}}\\
+\hspace{1em}drought resistance & R & - & ratio of growth during drought year to mean growth of the 5 years prior. & - & 1596 & 0.87 & 0 & 1.99 & no\\
+\addlinespace[.4em]
+\multicolumn{1}{l}{\textbf{Independent variables}}\\
+\hspace{1em}drought year & Y & - & year of drought & 1966 & 478 & - & - & - & -\\
+\hspace{1em} &  &  &  & 1977 & 547 & - & - & - & -\\
+\hspace{1em} &  &  &  & 1999 & 571 & - & - & - & -\\
+\addlinespace[.4em]
+\multicolumn{1}{l}{\textit{tree size}}\\
+\hspace{1em}\hspace{1em}diameter breast height & DBH & cm & DBH in drought year & - & all & 31.92 & 3.92 & 134.19 & yes\\
+\hspace{1em}\hspace{1em}height & H & m & H in drought year & - & all & 20.21 & 4.76 & 43.87 & yes\\
+\addlinespace[.4em]
+\multicolumn{1}{l}{\textit{microhabitat}}\\
+\hspace{1em}\hspace{1em}crown position & CP & - & 2018 crown position & dominant (D) & 31 & - & - & - & -\\
+\hspace{1em}\hspace{1em} &  &  &  & co-dominant (C) & 231 & - & - & - & -\\
+\hspace{1em}\hspace{1em} &  &  &  & intermediate (I) & 224 & - & - & - & -\\
+\hspace{1em}\hspace{1em} &  &  &  & suppressed (S) & 101 & - & - & - & -\\
+\hspace{1em}\hspace{1em}topographic wetness index & TWI & - & steady-state wetness index based on slope and upstream contributing area & - & all & 5.66 & 0 & 16 & yes\\
+\addlinespace[.4em]
+\multicolumn{1}{l}{\textit{species' traits}}\\
+\hspace{1em}\hspace{1em}wood density & WD & g cm-3 & dry mass of a unit volume of fresh wood & - & all & 0.62 & 0.4 & 1.09 & no\\
+\hspace{1em}\hspace{1em}leaf mass per area & LMA & kg m-2 & ratio of leaf dry mass to fresh leaf area & - & all & 48.69 & 30.68 & 75.8 & no\\
+\hspace{1em}\hspace{1em}xylem porosity & X & - & vessel arrangement in xylem & ring & 408 & - &  &  & -\\
+\hspace{1em}\hspace{1em} &  &  &  & semi-ring & 31 & - &  &  & -\\
+\hspace{1em}\hspace{1em} &  &  &  & diffuse & 178 & - &  &  & -\\
+\hspace{1em}\hspace{1em}turgor loss point & T & MPa & water potential at which leaves wilt & - & all & -2.39 & -2.76 & -1.92 & no\\
+\hspace{1em}\hspace{1em}percent loss area & PLA & \% & percent loss of leaf area upon dessication & - & all & 13.06 & 8.52 & 24.64 & no\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+Caption: T = \(\pi_{tlp}\), X = \({XP}\)
+
+We analyzed tree-ring data from 571 trees representing the twelve
+species contributing most to woody aboveground net primary productivity
+(ANPP), which together comprised 97\% of study plot ANPP between 2008
+and 2013 \citep{helcoski_growing_2019}. Cores were obtained in 2010-2011
+or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
+In 2010-2011, cores were collected from randomly selected live trees of
+species with at least 30 individuals of DBH \(\ge\) 10cm
 \citep{bourg_initial_2013}. In 2016-2017, cores were collected from all
@@
 published in association with \citet{helcoski_growing_2019}: (ITRDB;
-GitHub/Zenodo). \emph{is this a database? Yes. Ryan submitted the data
-but I don't think its posted yet. We should also cite GitHub/Zenodo
-here. I'll come back to that. }
+GitHub/Zenodo). \emph{Ryan submitted the data to ITRDB but I don't think
+its posted yet. We should also cite GitHub/Zenodo here. I'll come back
+to that.}
+
+For each tree, we combined tree-ring records and allometric equations of
+bark thickness to retroactively calculate DBH for the years 1950-2009.
+Prior DBH was estimated using the following equation:
+
+\(DBH_Y = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]\)
+
+Here, \emph{Y} denotes the year of interest, \(r_{ring}\) denotes ring
+width derived from cores, and \(r_{bark}\) denotes bark thickness. Bark
+thickness was estimated from species-specific allometries based on the
+bark thickness data from the site
+\citep{andersonteixeira_size-related_2015}. Specifically, we used linear
+regression equations on log-transformed data to relate bark thickness to
+DBH (Table S\#) and then used these to estimate bark thickness based on
+DBH.
 
@@
 method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon
-ForestryPro) for calculating heights. The associated errors for using
-either method were acknowledged \citep{larjavaara_measuring_2013}.
-Species-specific height allometries were developed (Table S\# -
-\textbf{ADD THIS TABLE TO SI}). For species that didn't have enough
-height measurements, heights were calculated from equations derived from
-all species in the study.
-
-For each tree, we combined tree-ring records and allometric equations of
-height and bark thickness to retroactively calculate DBH and estimate
-height for the years 1950-2009. Prior DBH was estimated using the
-following equation, using 2008 as the earliest year for having reliable
-DBH measurements:
-
-\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) \]
-
-Here, \emph{ring.width} was measured from cores. Bark thickness was
-estimated from species-specific allometries based on the bark thickness
-data of \citep{andersonteixeira_size-related_2015}. Specifically, we
-used linear regression equations on log-transformed data to relate bark
-thickness to DBH (Table S\#- \textbf{create table to give these
-equations in SI}) and then used these to estimate bark thickness based
-on DBH.
+ForestryPro) for calculating heights. Both methods are associated with
+some error \citep{larjavaara_measuring_2013}. Species-specific height
+allometries were developed (Table S\#). For species with insufficient
+height data to create reliable species-specific allometries, heights
+were calculated from equations derived from all species in the study.
 
@@
 While some trees undoubtedly changed position, an analysis of crown
-position relative to height (Fig. XX) and height change since
-\emph{1959} indicated that change was likely slow. {[}\textbf{work on
-this-- provide details?}{]}
+position relative to height (Fig. 2) and height change since \emph{1959}
+indicated that change was likely slow. Specifically, {[}\textbf{Ian,
+please provide details--e.g., average rate of height growth}{]}
 
 Topographic wetness index (TWI) was calculated using the
-\citep{R-dynatopmodel} package in R.
-
-\begin{figure}[htbp]
+\citep{R-dynatopmodel} package in R. {[}** Ian, include a brief
+explanation of what this is (plus citations)**{]}
+
+Hydraulic traits were collected from SCBI and are summarized in Table 3.
+In August 2018, we sampled small sun-exposed branches from three
+individuals of each species in and around the ForestGEO plot. These were
+covered with opaque plastic bags, re-cut under water, and re-hydrated
+overnight before further analysis. Rehydrated leaves (n=3 per
+indivdiual) were scanned, weighed, dried at 60\(^\circ\) C for \(\ge\)
+48 hours, and then re-scanned and weighed. Leaf area was calculated from
+scanned images using an R script (\textbf{details}). \(LMA\) was
+calculated as the ratio of leaf dry mass to fresh area. \(PLA\) was
+calculated as the percent loss of area between fresh and dry leaves.
+\(WD\) was calculated for \textasciitilde{}1cm diameter stem samples
+(bark and pith removed) as the ratio of dry weight to volume. We used
+the rapid determination method of ({[}Bartlett et al., 2012{]}) to
+estimate the turgor loss point (\(\pi_{tlp}\)). Briefly, two 4mm
+diameter leaf discs were cut from each leaf, tightly wrapped in foil,
+submerged in liquid nitrogen, perforated 10-15 times with a dissection
+needle, and then measured using a vapour pressure osmometer (VAPRO 5520,
+Wescor, Logan, UT, USA). Osmotic potential (\(\pi_{osm}\)) given by the
+osmometer was used to estimate (\(\pi_{tlp}\)) using the equation
+\(\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}\) ({[}Bartlett et al., 2012{]}).
+We also characterized hydraulic vulnerability curves for the \# most
+productive species, but because the water potentials at which 50\% and
+80\% of conductivity is lost, \(P50\) and \(P80\), did not come out as
+top predictors in preliminary analyses and their inclusion limited the
+set of species that could be included in the full analysis, these traits
+were dropped from further consideration. Data and R scripts for
+hydraulic traits are available at {[}\textbf{create new public GitHub
+repo for hydraulic traits, archive in Zenodo, give DOI}{]}.
+
+\begin{table}[t]
+
+\caption{\label{tab:Table3}**Table 3.
+Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
+Units of measurements are in mm (DBH), \% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).}
 \centering
-\includegraphics[width=5.20833in]{tables_figures/Figure3.png}
-\caption{Location of cored trees}
-\end{figure}
-
-Hydraulic traits were collected from SCBI and are summarized in Table 1.
-In August 2018, we collected leaf samples from three individuals of each
-species \ldots{} (\textbf{Nobby's description of methods for the
-following (see word document}) 1. PLA 2. LMA 4. Wood density 5. TLP
-
-\textbf{Table 2. Species analyzed here, listed in descending order of
-ANPP\_stem. n cores and DBH range represented, and species traits}
-{[}*This replaces/combines the two remaining tables in this section.
-Suggested columns, with those to include only if they fit in
-parentheses: species, (stems \textgreater{}=10 cm per ha in plot),
-(ANPP\_stem), n cores, DBH range of cores, (n cores in each crown
-position) species means for each trait{]}
-
-\begin{verbatim}
-## Warning: package 'kableExtra' was built under R version 3.4.4
-\end{verbatim}
-
-\textbackslash{}begin\{table\}{[}!h{]}
-
-\textbackslash{}caption\{\label{tab:Table2}Overview of analyzed species,
-detailing DBH mean and range of cored trees, the number of cores
-represented by each crown position of each species, and mean hydraulic
-trait measurements. Units of measurements are in mm (DBH), \% (PLA),
-g/m2 (LMA), MPa (TLP), and g/cm3 (WD).\} \centering
-
 \begin{tabular}{lrrlrrrr}
@@
 \end{tabular}
-
-\textbackslash{}end\{table\}
-
-\emph{Climate and drought years} {[}\textbf{add description of climate
-data used in Fig. 1, NEON vertical profiles}{]}
-
-To accurately understand climate sensitivity, this study used a specific
-definition of drought, which is not a common practice
-\citep{slette_how_2019}. We used the pointRes package \citep{R-pointRes}
-in R (version 3.5.3) to determine drought periods based on trees'
-drought resistance, which is defined by \citep{lloret_components_2011}
-as the ratio between the performance during and before the disturbance.
-Candidate drought years were defined if \textgreater{}50\% of the cored
-trees experienced \textless{}30\% growth in a year compared to the
-previous 5 years. These were then cross-validated with the regional
-Palmer Drought Severity Index (PDSI) values for each year, which yielded
-a set of three periods that were consistently shown as drought:
-1964-1966, 1977, and 1999.
-
-\subsubsection{Results}\label{results}
-
-Results outline
-
-Descriptions of Droughts
-
-\begin{itemize}
-\tightlist
-\item
-  physical characteristics of drought: duration, min PDSI (and when
-  occurred), see slette et al 2019 to see how they suggest reporting
-  each one
-\item
-  sum nb.series for each scenario, mult by perc.neg (pointers df),
-  define thresholds (e.g.~75\% of all trees experienced the growth
-  reduction of 40\% or more)
-\item
-  make table for this
-\item
-  citing table 3 (the coefficient for individual year) - maybe mention
-  how 77 and 99 have neg coefficients, indicating potential height
-  effect: as time goes on, trees are more affected by drought anyways
-  because they grow taller
-\item
-  this is followed by Figure 2
-\end{itemize}
-
-Hypothesis results - one header for each main question - present results
-for each one, walk through via order and logic of Table 1 - no citations
-here
-
-Once the data was collected, linear mixed models were run following the
-order of the hypotheses as seen in Figure ???
-{[}individual\_tested\_traits{]}. Using the \citep{R-pointRes} package,
-we set up models with the resistance value as the response variable, and
-each prediction's variable as the independent variable. Variables'
+\end{table}
+
+(\textbf{add description of climate data used in Fig. 1, NEON vertical
+profiles})
+
+\emph{Identification of drought years}
+
+We identified droughts within the time period 1950-2009, defining
+drought \citep{slette_how_2019} as events where tree growth was
+substantially reduced and where peak growing season climatic conditions
+were among the driest of the time period. To identify years with
+widespread reductions in tree growth, we used the pointRes package
+\citep{R-pointRes} in R (version 3.5.3) to determine drought periods
+based on trees' drought resistance, which is defined as the ratio
+between the performance during and before the disturbance
+\citep{lloret_components_2011}. Specifically, we looked at the ratio
+between annual basal area increment (BAI) in the year of the drought to
+average annual BAI in the 5 preceding years. Candidate drought years
+were defined if \textgreater{}25\% of the cored trees experienced
+\textless{}30\% growth in a year compared to the previous 5 years.
+Separately, we identified the years with driest conditions during
+May-August, which stood out in the analysis of
+\citep{helcoski_growing_2019} as the months (of the current year) to
+which annual growth was most sensitive for trees at this site. We
+considered two metrics of moisture deficit: NOAA Divisional Data's
+Palmer Drought Severity Index (PDSI) and the difference between
+potential evapotranspiration (PET) and precipitation (PRE). These data
+were obtained from the ForestGEO Climate Data Portal
+(\url{https://github.com/forestgeo/Climate}) in August 2018, with
+monthly PET and PRE sourced from Climatic Research Unit high-resolution
+gridded dataset (CRU TS v.4.01; Harris etal., 2014). The driest years
+were identified through simply ranking mean May-August PDSI or
+{[}PET-PRE{]} for the time period from driest to wettest.
+
+\emph{Analysis} (\textbf{this needs work})
+
+Linear mixed models were run following the order of the hypotheses as
+seen in Figure ??? {[}individual\_tested\_traits{]}. Using the
+\citep{R-pointRes} package, we set up models with the resistance value
+as the response variable (excluding outliers with values of \(R\ge 2\)),
+and each prediction's variable as the independent variable. Variables'
 importance in predicting drought tolerance was calculated from
@@
 
-\textbf{Describe each individual drought scenario } 1964-1966 1977 1999
-
-\textbf{Figure 1. Time series of peak growing season (May-August)
-climate conditions and residual chronologies for each species.} Droughts
-analyzed here are indicated by dashed lines, and shading indicates the
-pre-drought period used in calculations of the resistance metric. Figure
-modified from \citep{helcoski_growing_2019}.
-
-\begin{figure}[htbp]
+\hypertarget{results}{%
+\subsubsection{Results}\label{results}}
+
+\emph{Focal droughts and their characteristics}
+
+\begin{figure}
 \centering
-\includegraphics[width=5.20833in]{tables_figures/Figure1.jpg}
-\caption{Time series of combined tree cores by species}
+\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure1.png}
+\caption{\textbf{Figure 1. Climate and species-level growth responses
+over our study period, highlighting the three focal drougths (a) and
+community-wide responses} Time series plot (a) shows peak growing season
+(May-August) climate conditions and residual chronologies for each
+species. Focal droughts are indicated by dashed lines, and shading
+indicates the pre-drought period used in calculations of the resistance
+metric. Figure modified from \citep{helcoski_growing_2019}. Density
+plots (b) show community- wide resistance values for each drought.}
 \end{figure}
 
-\textbf{Results for first main question: what drives the observed
-tendency for large trees to suffer more during drought?} H1.0, H1.1,
-H1.2, H1.3 DBH, height, crown position, and TWI
-
-\begin{figure}[htbp]
+In the 60-year period between 1950 and 2009, there were three droughts
+that met our criteria of anomalously dry climatic conditions coupled
+with substantial reductions in tree growth for at least some portions of
+the community: 1966, 1977, and 1999 (Fig. 1). We excluded one year
+(1991) meeting the growth reduction criteria (26.5\% of trees
+experienced \textgreater{}30\% growth reduction, mean resistance=
+-13.8\%) because this year was not among the strongest droughts of the
+study period (\textbf{DETAILS}). Rather, the severity of growth
+reduction may be explained by defoliation by gypsy moths
+(\emph{Lymantria dispar} L.) from approximately 1988-1995, which most
+stronly impacted \emph{Quercus} spp. (\emph{Cite Shenanadoah paper, if
+accepted}). Climatically, these droughts included three of the five
+years between 1950 and 2009 with greatest moisture deficit (PET-PRE)
+during the peak growing season months of May-August, which are the
+months to which annual tree growth at this site is most sensitive
+\citep{helcoski_growing_2019}. Specifically, 1966, 1977, and 1999 had
+mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo-1, respectively. The
+years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87
+and 82.13 mm mo-1), but \emph{were not among the lowest in terms of PDSI
+and were not identified as a pointer yeasr.} \textbf{The droughts
+differed in timing/duration/etc. .. The year 1966 was preceded by two
+relatively dry years\ldots{} 1964 among five driest in terms of
+May-August {[}PET-PRE{]}, 1965 also anomalously hot and dry. }
+
+Community-level tree growth responses to these droughts were modest,
+with modal resistance values of \#, \#, and \# for 1966, 1977, and 1999,
+respectively (Fig. 1b). In each drought, roughly 30\% of the cored trees
+suffered \(\ge\) 30\% growth reductions (\(R\) \(\le\) 0.7): \#\% in
+1966, \#2\% un 1977, and \#\% in 1999. \emph{Some} trees exhibited
+increased growth: (\(R\) \textgreater{} 1.0): \#\% in 1966, \#\% un
+1977, and \#\% in 1999.
+
+\emph{Tree size and drought resistance}
+
+Overall, our analysis confirmed the tendency for larger-diameter trees
+to show greater reductions in growth during drought
+\citep{bennett_larger_2015} (\emph{H1.0}), although there was no
+signficant effect for 1977 or 1999 individually (Tables 1, 4). The same
+held true for \(ln[H]\) as a univariate predictor (\emph{H1.1}; Tables
+1, 4). When combined with other predictor variariables in our
+multivariate models, the top models usually included an effect of
+\(ln[H]\), and its coefficient was consistently negative, as predicted
+(Tables 1, 5). We note that a non-signficant positive correlation
+between \(ln[H]\) and \(R\) for 1999 became negative in the context of
+the multivariate models, again supporting \emph{H1.1} (Table 1).
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-5}**Table 4. Univariate models**}
 \centering
-\includegraphics[width=5.20833in]{tables_figures/Figure2.png}
-\caption{Height profile graphs}
+\resizebox{\linewidth}{!}{
+\fontsize{12}{14}\selectfont
+\begin{tabular}{llllrllllll}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{all droughts} & \multicolumn{2}{c}{1966} & \multicolumn{2}{c}{1977} & \multicolumn{2}{c}{1999} \\
+\cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9} \cmidrule(l{3pt}r{3pt}){10-11}
+ï»¿variable & category & null variables & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients\\
+\midrule
+drought year & 1966 &  & -2.42 & 0.0000 & - & - & - & - & - & -\\
+ & 1977 &  & - & -0.0209 & - & - & - & - & - & -\\
+ & 1999 &  & - & -0.0105 & - & - & - & - & - & -\\
+ln[DBH] &  & Y & 8.17 & -0.0385 & 15.32 & -0.0888 & -0.87 & -0.0214 & -1.93 & 0.0057\\
+ln[height] &  & Y & 8.8 & -0.0648 & 15.27 & -0.1443 & -0.98 & -0.0335 & -2.03 & 0.0018\\
+\addlinespace
+crown position & D & Y & -2.96 & -0.0461 & 3.25 & -0.0509 & 0.66 & -0.0759 & 0.38 & -0.0103\\
+(alone) & C &  & - & 0.0000 & - & 0 & - & 0 & - & 0\\
+ & I &  & - & -0.0063 & - & 0.0732 & - & -0.0298 & - & -0.0563\\
+ & S &  & - & 0.0122 & - & 0.0526 & - & 0.0432 & - & -0.0483\\
+crown position & D & ln[H]+Y & 0.55 & -0.0364 & -1.41 & -0.0359 & -0.24 & -0.074 & 3.99 & -0.0027\\
+\addlinespace
+(with height) & C &  & - & 0.0000 & - & 0 & - & 0 & - & 0\\
+ & I &  & - & -0.0406 & - & 0.0177 & - & -0.0363 & - & -0.0823\\
+ & S &  & - & -0.0586 & - & -0.0654 & - & 0.03 & - & -0.1011\\
+ln[TWI] &  & ln[H]+Y & 5.33 & -0.0886 & -1.96 & -0.0168 & 5.06 & -0.1406 & 2.72 & -0.1025\\
+ln[TWI]*ln[H] &  & ln[H]+ln[T]+Y & -1.18 & 0.0677 & -1.75 & 0.0749 & -1.86 & 0.0533 & -1.79 & 0.0566\\
+\addlinespace
+wood density &  & ln[H]+Y & -1.89 & -0.0498 & -1.1 & -0.2161 & -1.19 & -0.1827 & 0.23 & 0.2512\\
+leaf mass per area &  & ln[H]+Y & -2 & 0.0003 & -1.89 & 0.0011 & -1.74 & -0.0014 & -1.99 & 0.0005\\
+ring porosity & ring & ln[H]+Y & -2.4 & 0.0574 & 1.78 & 0.151 & 0.59 & -0.1879 & 4.25 & 0.2025\\
+ & semi-ring &  & - & -0.0335 & - & -0.1324 & - & -0.1426 & - & 0.1516\\
+ & diffuse &  & - & 0.0000 & - & 0 & - & 0 & - & 0\\
+\addlinespace
+turgor loss point &  & ln[H]+Y & 1.22 & -0.1675 & -1.78 & -0.0845 & 1.14 & -0.2432 & 0.06 & -0.1749\\
+percent loss area &  & ln[H]+Y & 7.89 & -0.0138 & 9.82 & -0.0244 & -0.07 & -0.0104 & -0.73 & -0.0074\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\begin{table}[!h]
+
+\caption{\label{tab:unnamed-chunk-6}**Table 5.
+Summary of R^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model ($dAICc<2$).}
+\centering
+\resizebox{\linewidth}{!}{
+\fontsize{12}{14}\selectfont
+\begin{tabular}{lrrrlllllllllll}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{4}{c}{crown position} & \multicolumn{1}{c}{ } & \multicolumn{3}{c}{xylem architecture} & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){11-13}
+ï»¿drought & dAICc & R2 & Intercept & ln[H] & D & C & I & S & ln[TWI] & diffuse & semi-ring & ring & PLA & TLP\\
+\midrule
+\addlinespace[.4em]
+\multicolumn{1}{l}{\textbf{}}\\
+\hspace{1em}all & 0.000 & 0.11 & 1.108 & -0.062 & - & - & - & - & -0.085 & - & - & - & -0.012 & -0.105\\
+\hspace{1em} & 0.338 & 0.10 & 1.494 & -0.097 & -0.036 & 0 & -0.038 & -0.056 & -0.078 & - & - & - & -0.013 & -\\
+\hspace{1em} & 0.403 & 0.12 & 1.263 & -0.096 & -0.035 & 0 & -0.036 & -0.053 & -0.078 & - & - & - & -0.012 & -0.087\\
+\hspace{1em} & 0.515 & 0.12 & 1.393 & -0.064 & - & - & - & - & -0.087 & 0 & 0.147 & 0.048 & -0.017 & -\\
+\hspace{1em} & 0.571 & 0.10 & 1.375 & -0.061 & - & - & - & - & -0.085 & - & - & - & -0.013 & -\\
+\hspace{1em} & 0.903 & 0.12 & 1.492 & -0.098 & -0.034 & 0 & -0.036 & -0.053 & -0.079 & 0 & 0.122 & 0.049 & -0.016 & -\\
+\addlinespace[.4em]
+\hline
+\multicolumn{1}{l}{\textbf{}}\\
+\hspace{1em}1996 & 0.000 & 0.26 & 2.271 & -0.153 & - & - & - & - & - & 0 & 0.35 & 0.152 & -0.035 & 0.25\\
+\hspace{1em} & 0.537 & 0.26 & 1.537 & -0.154 & - & - & - & - & - & 0 & 0.098 & 0.133 & -0.023 & -\\
+\hspace{1em} & 1.093 & 0.27 & 2.389 & -0.177 & -0.038 & 0 & 0.016 & -0.068 & - & 0 & 0.352 & 0.153 & -0.035 & 0.27\\
+\hspace{1em} & 1.434 & 0.24 & 1.629 & -0.143 & - & - & - & - & - & - & - & - & -0.024 & -\\
+\hspace{1em} & 1.982 & 0.26 & 2.307 & -0.152 & - & - & - & - & -0.016 & 0 & 0.356 & 0.152 & -0.035 & 0.254\\
+\addlinespace[.4em]
+\hline
+\multicolumn{1}{l}{\textbf{}}\\
+\hspace{1em}1977 & 0.000 & 0.22 & 0.346 & - & -0.074 & 0 & -0.027 & 0.042 & -0.131 & 0 & -0.331 & -0.23 & - & -0.384\\
+\hspace{1em} & 0.090 & 0.21 & 0.393 & - & - & - & - & - & -0.14 & 0 & -0.324 & -0.234 & - & -0.369\\
+\hspace{1em} & 1.116 & 0.21 & 0.460 & -0.026 & - & - & - & - & -0.137 & 0 & -0.316 & -0.225 & - & -0.367\\
+\hspace{1em} & 1.914 & 0.22 & 0.367 & -0.006 & -0.073 & 0 & -0.029 & 0.037 & -0.13 & 0 & -0.33 & -0.229 & - & -0.383\\
+\addlinespace[.4em]
+\hline
+\multicolumn{1}{l}{\textbf{}}\\
+\hspace{1em}1999 & 0.000 & 0.25 & 1.284 & -0.081 & 0.003 & 0 & -0.077 & -0.095 & -0.087 & 0 & 0.22 & 0.193 & -0.008 & -\\
+\hspace{1em} & 0.085 & 0.25 & 0.844 & -0.082 & 0.001 & 0 & -0.078 & -0.095 & -0.085 & 0 & 0.062 & 0.185 & - & -0.142\\
+\hspace{1em} & 0.462 & 0.23 & 1.174 & -0.083 & 0.002 & 0 & -0.079 & -0.099 & -0.088 & 0 & 0.135 & 0.2 & - & -\\
+\hspace{1em} & 0.956 & 0.24 & 1.042 & - & - & - & - & - & -0.1 & 0 & 0.261 & 0.191 & -0.01 & -\\
+\hspace{1em} & 1.029 & 0.24 & 0.700 & -0.09 & -0.002 & 0 & -0.082 & -0.101 & - & 0 & 0.046 & 0.185 & - & -0.151\\
+\hspace{1em} & 1.193 & 0.24 & 1.159 & -0.089 & 0 & 0 & -0.081 & -0.101 & - & 0 & 0.21 & 0.194 & -0.008 & -\\
+\hspace{1em} & 1.326 & 0.24 & 0.533 & - & - & - & - & - & -0.098 & 0 & 0.077 & 0.181 & - & -0.161\\
+\hspace{1em} & 1.849 & 0.26 & 1.046 & -0.079 & 0.002 & 0 & -0.077 & -0.094 & -0.086 & 0 & 0.14 & 0.188 & -0.005 & -0.078\\
+\hspace{1em} & 1.851 & 0.22 & 1.045 & -0.092 & -0.001 & 0 & -0.084 & -0.105 & - & 0 & 0.123 & 0.201 & - & -\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\raggedright
+
+Crown position was generally correlated with \(H\), but with substantial
+variation (Fig. 2\textbf{e}). Crown position was a much poorer predictor
+of \(R\) than was \(H\) (Table 4), lending little overall support to
+\emph{H1.2} (Table 1). When considered alone, \(CP\) had a significant
+influence only in the 1977 drought, during which trees with dominant
+\(CP\) had the lowest \(R\). When considered in conjunction with \(H\),
+\(CP\) came out as a significant predictor only for the 1999 drought,
+during which suppressed trees had the lowest \(R\). Crown position was
+included in \textbf{roughly half} of the top models, with mixed results
+as to how \(R\) varied with \(CP\) (Table 5). Most commonly in these
+multivariate models, as in the univariate models (Table 4), the
+resistance of dominant trees was less than that of co-dominant trees but
+higher that of suppressed trees. Thus, \(CP\) was sometimes a useful
+predictor of \(R\), but overall had a weak effect relative to that of
+\(H\).
+
+In the non-drought years for which we have vertical profiles in climate
+data (\textbf{YEARS}), taller trees--or those in dominant crown
+positions-- were generally exposed to higher evaporative demand during
+the peak growing season months (May-August; Fig. 2). Specifically,
+maximum daily wind speeds were signficantly higher above the top of the
+canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative
+humidity was also somewhat lower during June-August, ranging from
+\textasciitilde{}50-80 above the canopy and \textasciitilde{}60-90\% in
+the understory (Fig. 2b). Air temperature did not vary across the
+vertical profile (Fig. 2c). {[}\textbf{biological temperature?- probably
+drop}{]}
+
+\begin{figure}
+\centering
+\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure2.png}
+\caption{\textbf{Figure 2. Height profiles in growing season climatic
+conditions, tree heights by crown position, and leaf hydraulic traits}
+The top row shows averages (\(\pm\) SD) of daily maxima and minima of
+(a) wind speed, (b) relative humidity (\(RH\)), (c) air temperature
+(\(T_{air}\)), and (d) biological temperature
+(\(T_{biological}\))averaged over each month of the peak growing season
+(May-August) from YEAR-YEAR. In these plots, heights are slightly offset
+for visualization purposes. Also shown are (e) 2018 tree heights by
+canopy position (see Table 2 for codes) and vertical profiles in (f)
+\(PLA_{dry}\) and (g) \(\pi_{tlp}\). In (f-g), values are community-wide
+averages across height bins (plotted at upper end of height bin), with
+grey indicating bins for which species-level trait measurements are
+available for \textless{}75\% of individuals. In all plots, the dahsed
+horizontal line indicates the 95th percentile of tree heigts in the
+ForestGEO plot.}
 \end{figure}
 
-\textbf{Results for second main question: how do species' traits - alone
-and in interaction with tree size - influence drought response? } H2.1,
-H2.2, H2.3 Hydraulic traits alone, traits with height
-
-\includegraphics{tables_figures/Table4.png} \#this is
-tested\_traits\_all
-
-\textbf{Results for third main question: are responses similar or
-variable across individual drought years?} H3.1 and H3.2 Combining
-biophysical with hydraulic traits, which come out as candidates for best
-model?
-
-\subsubsection{Discussion}\label{discussion}
-
-Discussion outline
-
-\begin{itemize}
-\tightlist
-\item
-  first paragraph is summary of main findings, not re-presenting the
-  results
-\item
-  ``we supported this hypothesis, but not this one'' (following same
-  order as Table 1 still)
-\item
-  direction of responses seems to be relatively consistent but the
-  individual responses vary
-\item
-  tie things in (see here
-\item
-  main thing is that we're now better understanding what confers
-  vulnerability or resilience on trees during drought - how forests
-  respond to future droughts/climate change
-\item
-  ``we filled the gap'' = 1-2 sentences
-\item
-  limitations at our site:
-\item
-  aren't able to analyze historical forest community, nor trees that
-  were killed by these droughts {[}aka we don't have data for
-  individuals that were most severely affected{]} (though we found that
-  there's little variation in climate sensitivity for trees that were
-  cored dead vs cored alive).
-\item
-  p50/p80
-\item
-  We used crown position despite its uncertainty. However, height and
-  crown position change relatively slowly so it shouldn't be that far
-  off.
-\item
-  We did not use crowding index because it has much more uncertainty
-\item
-  limitations for extrapolating other sites:
-\item
-  forests are different from place to place, but we've seen how in
-  forests around the world, forests tend to suffer more. We've
-  identified some facets for why this happens. (cite other studies)
-\item
-  has been observed elsewhere that individuals in more moist habitats
-  are more susceptible to drought (cite other studies)
-\item
-  the species may be different from other sites but the hydraulic traits
-  are general across species;
-\item
-  next paragraph saying that our study advances understanding of
-  droughts, e.g.~bennett et al 2015 doesn't go into mechanisms but we do
-\item
-  and we show that height is more important than exposure, but doesn't
-  eliminate the effect of it
-\item
-  further contribution is we show that two leaf hydraulic traits -
-  relatively easy to measure - are good predictors of drought response,
-  can be helpful for scaling up (e.g.~from our site to eastern deciduous
-  biome)
-\item
-  biophysical mechanisms are things that should be seen as universal but
-  relative importance of each can vary within each drought
-\item
-  first study to show how these traits affect woody growth response to
-  drought \textbf{confirmed by Lawren}
-\item
-  science is better now. by advancing our understanding of the
-  mechanisms for individual-level responses to drought, this opens door
-  for better predictions (elaborate from above). ``This is absolutely
-  critical to predicting forest responses to future droughts, which
-  we're likely to see more of in the future as a result to climate
-  change. Forecasting forest responses to these droughts is a huge and
-  important challenge'', science is better.
-\end{itemize}
-
-\emph{1. paragraph summarizing main results--\textgreater{} primary
-conclusions} When including only biophysical traits, trees' resistance
-value (on a per-species basis) is explained best by crown position and
-height, with codominant trees being the most resistant to drought. This
-follows on work done by \citep{bennett_larger_2015} {[}and others?{]}
-which show that larger trees suffer more during drought, and confirms
-that this susceptibility can be seen in tree ring analyses. Adding in
-crown position with the leaf hydraulic traits yields a slightly worse
-predictive model for drought tolerance, with height remaining as the
-only significant biophysical variable.
-
-We partially supported the hypothesis that crown exposure makes trees
-more vulnerable to drought. Co-dominant trees had the highest drought
-resistance. Dominant trees had lower resistance, likely because they are
-the most exposed. Other studies have found clear evidence of greater
-drought sensitivity in trees with exposed crowns (e.g.,
-\citep{suarez_factors_2004}; \citep{scharnweber_confessions_2019}). At
-the same time, intermediate and suppressed trees had even lower
-resistance. This indicates that other mechanisms such as competition or
-rooting depth were important. (Also note that our study design was not
-ideal for testing the role of canopy position. Current canopy position
-is a conservative separator of canopy position: trees may currently be
-in more dominant positions than they were at the time, but backwards
-movement is unlikely. This would bias against finding a signficant
-effect for H1.2. Height may be a more reliable predictor of past canopy
-position than is current canopy position, and explains a portion of
-variation in canopy position.)
-
-Proximity to stream--either vertical (elev) or horizontal (distance)--
-did not increase drought resistance; rather, it tended to decrease
-resistance (H1.3a). This may be because individuals growing further from
-water are acclimatized to drier conditions. However, the increase in
-drought resistance with distance from stream was less for small than
-large trees (H1.3b), indicating a potential importance of root
-depth/volume in conferring drought resistance.
-
-\textbf{misc content to integrate} From \citep{kannenberg_linking_2019},
-species with diffuse porous wood anatomy (\emph{Liriodendron}) are more
-sensitive to drought, whereas ring-porous are not as affected because
-they more easily rebuild structures for hydraulic conductivity. This
-paper mentions it would be good to have this data with respect to latent
-affects from drought. \#\#\# Acknowledgements
+Resistance was negatively correlated with \(ln[TWI]\) (Tables 4-5),
+negating the idea that trees in moist microsites would suffer less
+during drought. Nevertheless, we tested for a negative
+\(ln[H] *ln[TWI]\) interaction (\emph{H1.3}), which could indicate that
+smaller trees (with smaller rooting volume) have a greater tendency to
+suffer more in drier microenvironments with greater depth to the water
+table. \emph{H1.3} was rejected; the \(ln[H] *ln[TWI]\) interaction was
+never signficant and had a consistently positive coefficient (Table 4).
+
+\emph{Species' traits, height, and drought resistance}
+
+We partially support \emph{H2.1}: Species' hydraulic traits --\(XP\),
+\(PLA_{dry}\), and \(\pi_{tlp}\)--were useful in explaining variation in
+drought responses, whereas \(LMA\) and \(WD\) were not (Tables 1,4,5).
+Specifically, \(LMA\) and \(WD\) never significantly associated with
+\(R\) in the univariate models (all dAIC \(\le\) 0.23; Table 4), and
+therefore these were excluded as candidate variables for the full
+multivariate models. In contrast, \(XP\), \(PLA_{dry}\), and
+\(\pi_{tlp}\) all explained at least modest amounts of variation (dAIC
+\textgreater{} 1.0) in at least one drought (Table 4). Of these,
+\(PLA_{dry}\) was the strongest predictor, with consistently negative
+coefficients across all droughts. \(\pi_{tlp}\) never came out as
+significant (dAIC \(\ge\) 2) in the univariate models, but had a
+consistently negative coefficient (Table 4). Whereas ring-porous species
+had highest \(R\) overall and in the 1966 and 1999 droughts, they had
+lower \(R\) in 1977. Results were similar in the context of multivariate
+models (Table 5), except that \(\pi_{tlp}\) had a positive coefficient
+in the 1966 models in which it was included.
+
+We reject \emph{H2.2}, finding no evidence that taller trees tend to
+have traits associated with greater drought resistance. In part because
+of the large sample size (n=\# trees--all individuals of our 12 focal
+species \(\ge\) 10 cm in 25.6 ha \textbf{Ian, confirm}), there were very
+significant (p\textless{}0.0001) correlations of \(H\) with all species'
+traits (\textbf{SI Table?}). However, the correlation only matched the
+predicted direction (\emph{i.e.}, more drought-resistant traits
+associated to taller trees) in the case of \(LMA\), which did not
+correlate with \(R\). Furthermore, although correlations were
+statistically signifcant, trait variation within each height class
+overwhelmed any vertical trends (Fig. 2\textbf{f-g}).
+
+We support the hypothesis (\emph{H2.3}) that the observed tendency for
+larger trees to have greater growth reductions during drought (lower
+\(R\)) is \emph{not} driven by more drought-sensitive traits in larger
+trees, but rather by size itself (Tables 1,5). As discussed above, there
+was little meaningful variation in traits with height at the community
+level. When \(ln[H]\) and hydraulic traits were considered together in
+multivariate models, the effect of \(ln[H]\) on \(R\) was consistently
+negative (Table 5)--reversing a non-signficant positive \(ln[H]\)-\(R\)
+correlation in the univariate model for the 1999 drought (Table 4).
+
+\emph{Responses across droughts}
+
+We reject the hypothesis (\emph{H3.1}) that overall community responses
+varied across droughts. Within the context of mixed effects models,
+there were no significant differences in \(R\) across drought years
+(Table 4). This is consistent with the observation that the distribution
+of \(R\) values was similar across droughts (Fig. 1b).
+
+We mostly reject the hyopthesis (\emph{H3.2}) that directions of
+responses varied across droughts. In the majority of cases, response
+directions were consistent across droughts in both univariate and
+multivariate models (Tables 1,4,5). However, there were a few
+exceptions--most commonly in the categorical variables (\(CP\) and
+\(XP\)) but also for \(\pi_{tlp}\) in the multivariate model for the
+1966 drought; Tables 4, 5). These differences may very well be random,
+as opposed to statistically meaningful. Among the univariate models,
+there was no instance predictor variables signficantly improved the
+models of two different droughts (dAIC \(\ge\) 2), but with contrasting
+coefficients (Table 4). Among the multivariate models, \(CP\) was not
+consistently in the top models for any drought (Table 5), and
+\(\pi_{tlp}\) only appeared with a positive coefficient in two of five
+models for the 1966 drought (contrasting with a negative coefficient in
+the univariate model; Table 4). The difference most likely to be real is
+that ring and semi-ring porous species had lower resistance than diffuse
+porous species in the 1977 drought, contrasting with higher resistance
+in 1966 and 1999 (Tables 4,5), but note that \(XP\) was not a signficant
+predictor on its own for the 1977 drought.
+
+We support the hypothesis (\emph{H3.3}) that the strength of predictor
+variables was different across the droughts (Tables 1,4,5). For
+instance, \(ln[H]\) and \(PLA_{dry}\) had much stronger negative effects
+in 1966 than in the other two years, \(ln[TWI]\) had the strongest
+negative effect in 1977, and \(CP\) (lowest \(R\) among supressed trees)
+and \(XP\) (lowest \(R\) among diffuse-porous trees) were strongest in
+1999 (Tables 4,5).
+
+\hypertarget{discussion}{%
+\subsubsection{Discussion}\label{discussion}}
+
+Our results reveal how tree size, microhabitat, and hydraulic traits
+shaped tree growth responses across three droughts in a temperate
+deciduous forest (Table 1). The tendencey for larger trees to suffer
+more, observed here as in forests around the world (\textbf{REFS}), was
+driven primarily by their height. There was a marginal additional effect
+of crown exposure, with the most exposed and the most suppressed trees
+suffering most--consistent with observations of both greater drought
+sensitivity of exposed trees (e.g., \citep{suarez_factors_2004};
+\citep{scharnweber_confessions_2019}; MORE??) and greater sensitivity of
+suppressed and crowded individuals (\textbf{REFS- }). There was no
+evidence that root water access increased drougth resistance; in
+contrast, trees in wetter topographic positions suffered more
+(consistent with
+\href{https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1\&isAllowed=y}{Zuleta
+et al.}), and larger rooting volume provided no advantage in the drier
+microenvironments. The lower drought resistance of larger trees was not
+driven by any tendency for the canopy to be dominated by more
+drought-sensitive species. Drought-sensitive species were not linked to
+\(LMA\) and \(WD\) but were predicted by hydraulic traits
+(\(PLA_{dry}\), \(\pi_{tlp}\), and \(XP\)), which is physiologically
+logical (\textbf{REFS}\citep{kannenberg_linking_2019}) but
+scientifically novel in that \(PLA_{dry}\) and \(\pi_{tlp}\) have not
+previously been linked to drought growth responses. The direction of
+these responses was mostly consistent across droughts, indicating that
+they were driven by fundamental physiological mechanisms; however, the
+strengths of each predictor varied across droughts, indicating that
+specific drought characteristics interact with tree size,
+microenvironment, and traits to shape which individuals suffer most.
+These findings significantly advance our knowledge as to the factors
+that confer vulnerability or resilience on trees during drought.
+
+The droughts considered here were of similar severity (Fig. 1b) and
+fairly moderate; droughts of this magnitude have occurred with an
+average frequency of approximately one per 10-15 years (Fig. 1a,
+\citet{helcoski_growing_2019}). Therefore, we excpect that most species
+are adapted, and individual trees acclimatized, to survive droughts of
+this nature. While the majority of trees experienced reduced growth, a
+substantial portion had increased growth (Fig. 1b), underlining the fact
+that these droughts did not induce extreme stress on the entire forest.
+It is likely for this reason, combined with the fact that many factors
+other than climate affect tree growth in closed-canopy forests, that our
+best models characterize only a modest amount of variation: 10-12\% for
+all droughts combined, and 21-27\% for each individual drought (Table
+5). Methodologically, the moderate nature of these drougths is an
+advantage because our analysis considers only trees that survived all of
+these droughts, and we lack information on the trees that were killed.
+These are likely to be relatively modest in number, and local forest
+monitoring data stretching back to the late 1980s confirms that the 1999
+drought did not trigger major declines in tree abundance or biomass
+(\emph{Anderson-Teixeira et al., in review}). Thus, the droughts
+considered here are substantially weaker than those that have triggered
+massive tree die-off (e.g.,
+\href{https://doi.org/10.1016/j.foreco.2009.09.001}{Allen et al.~2015}),
+many of which have shaped our understanding about the role of tree size
+(Bennett et al.~2015; Stovall et al.~2019) and--to some extent-- traits
+(\textbf{REFS}). Nevertheless, our results are consistent with findings
+from more extreme droughts.
+
+Our analysis indicates that height is the primary factor through which
+tree size mediates drought response; however, specific identification of
+the primary physiological mechanism(s) is complicated by the fact that
+multiple drought-relevant environmental parameters and tree
+characteristics covary with height. To begin with, taller trees face
+inherent biophysical challenges in lifting water a greater distance
+against the effects of gravity and friction (\textbf{REFS}). They also
+face different microenvironments, which are partially related to canopy
+position. Even under non-drought conditions, evaporative demand
+increases with tree height, which are more closely coupled to the
+atmosphere (Fig. 1a-b; \textbf{REFS- Jarvis 1984?};
+\href{https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071}{Bretfeld
+et al}). \textbf{The vegetation reaches higher
+temperatures\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58}{??}},
+which would occur more during a drought when solar radiation tends to be
+higher and less water is available for evaporative cooling of the
+leaves. Furthermore, daytime CO\_2 concentrations tend to be lower in
+the canopy than in the understory \textbf{(Fig.
+2\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58}{??}}
+(\href{https://watermark.silverchair.com/21-12-13-951.pdf}{Koike et
+al.~2001}; MORE REFS), implying greater water costs of \(CO_2\) uptake.
+\textbf{{[}McDowell-vertical profiles in isotopes?*{]}}. Taller trees
+are more likely to be in domiant canopy positions, although signficant
+decoupling can be introduced by the configuration of neighboring trees
+(Fig. 2\textbf{e?})
+(\href{https://www.ncbi.nlm.nih.gov/pubmed/16643303}{Mueller-Landau et
+al.~2016}). We show that height is a far stronger predictor than crown
+position (Tables 1,4,5). Our analysis does have the limitation that
+canopy positions were recored in 2018 and undoubtedly changed for some
+trees since the 1960s, and we note that \(CP\) became an increasingly
+poor predictor moving from 1999 back to 1966 (Table 4). However, rates
+of change are relatively slow, and the fact that some trees inevitably
+did grow towards more dominant positions would bias against the
+acceptance of (\emph{H1.2}), implying that dominant crown positions did
+indeed reduce \(R\), which makes sense in light of the vertical
+enviromnental gradients described above and agrees with previous studies
+showing greater drought sensitivity in more exposed trees
+(\citep{suarez_factors_2004}; \citep{scharnweber_confessions_2019};
+MORE??). It is safe to assume that currently suppressed trees have
+always been suppressed, and their relatively low \(R\) (after correcting
+for height effects) is real, which is consistent with analyses showing
+that suppressed--and particularly crowded--trees can suffer
+distproportionately during drought (\textbf{REFS}). The observed
+height-sensitivity of \(R\), together with the apparent lack of
+importance of root water access (\emph{H1.3}), agrees with the concept
+that physiological limitations to transpiration under drought shift from
+root water access to the plant-atmosphere intreface as forests age
+(\href{https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071}{Bretfeld
+et al}), such that tall trees--particularly the very tallest--are the
+most sensitive in mature forests. Additional research comparing drought
+responses of young and old forest stands, along with and short and tall
+isolated trees, would be valuable for more clearly disentanging the
+roles of tree height and crown exposure.
+
+Beyond microenvironment, multiple tree characteristics covary with
+height. To begin with, multiple tree characteristics that may influence
+drought resistance scale with \(DBH\)--or, similarly, \(H\)--including
+crown volume, rooting volume, sapwood area, transpiration \ldots{}
+(e.g.,
+\href{https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470}{Anderson-Teixeira
+et al.~2015\_scaling}; \textbf{REFS--KAT: Sperry scaling model; roots
+reference; BAAD database?}). These work in combination to shape
+whole-plant hydraulic function (\textbf{REFS-KAT: Sperry scaling model,
+check McDowell scaling paper, Couvreur et al.~2018; others?}), and we
+cannot expect to statistically identify a single driver that fully
+encapsulates size-related variation in trees. However, height is
+arguably the most biophysically meaningful tree-size variable shaping
+drought response. Beyond whole-plant characteristics, leaf traits vary
+with height on the tree--sometimes quite dramatically, with upper-canopy
+leaves \ldots{} (\textbf{describe differences})
+(\href{https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA\%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU}{Couvreur
+et al.~2018};
+\href{https://watermark.silverchair.com/21-12-13-951.pdf}{Koike et
+al.~2001}). Our analysis focuses on species-level comparisons and does
+not characterize the role of variation with height. On an individual
+level, average leaf characteristics of an individual tree would scale
+with its crown height, with taller individuals having on average more
+drought-resistant traits--assuming trends for \(PLA_{dry}\) and
+\(\pi_{tlp}\), which have not been characterized (\emph{\textbf{LAWREN,
+IS THIS TRUE?}), follow the general pattern of increasing drought
+resistance with height. In not accounting for this, our analysis is most
+likely biased in a conservative direction when assessing }H2.3* (Table
+1)--\emph{i.e.}, accounting traits on an individual level should result
+in a stronger negative effect of \(ln[H]\). Further characterization of
+leaf hydraulic traits in relation to height and crown exposure would be
+valuable for enhancing our understanding of the interactive effects of
+tree height and traits on drought responses (\emph{Couvreur et
+al.~2018}).
+
+The development of tree-ring chronologies for all dominant tree species
+at our site \citep{helcoski_growing_2019} made it possible to compare
+historical drought responses across 12 species and their associated
+traits at a single site for the first time (\textbf{verify- Neil,
+Alan}). Concerted measurement of leaf hydraulic traits of emerging
+importance (\textbf{REFS--KAT/NOBBY/LAWREN}) allowed novel insights into
+the role of hydraulic traits in shaping drought response. The finding
+that \(PLA_{dry}\) and \(\pi_{tlp}\) are useful for predicting drought
+responses is consistent with studies demonstrating that these are
+physiolgoically meaningful traits linked to species distribution along
+moisture gradients (\textbf{REFS--KAT/NOBBY/LAWREN}),
+\textbf{MORE--KAT/NOBBY/LAWREN}\ldots{} (\textbf{REFS}). It is
+scientifically exciting in that it indicates that these traits, which
+can be measured relatively easily \(PLA_{dry}\) and \(\pi_{tlp}\)
+(\textbf{REFS--KAT/NOBBY/LAWREN}), hold promise for predicting drought
+growth responses across species. The importance of linking species'
+traits to drought responses increases with tree species diversity;
+whereas it is feasible to study drought responses for all dominant
+species in most boreal and temperate forests (e.g., this study;
+\textbf{REFS}), this becomes difficult to impossible for diverse
+tropical forests, where linking hydraulic traits to drought responses
+would be invaluable for forecasting how little-known species and whole
+forests will respond to future droughts
+(\textbf{REFS--KAT/NOBBY/LAWREN}).
+
+Although the physiological mechanisms discussed above lead to generally
+consistent directions of growth responses to tree height and hydaulic
+traits across droughts, indicating the universality of the underlying
+mechanisms, the relative importantance of the drivers varied widely
+across droughts, indicating an interaction with drought characteristics
+(Tables 4-5). {[}\textbf{BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS--
+For example, 1966 was preceded by two dry years, and height had the
+strongest effect in this drought, consistent with the finding that
+height becomes a stronger predictor of mortality as drought duration and
+intensity increase (Stovall et al.~2019). \ldots{}}{]} Of course, site
+characteristics also define the nature of droughts, and comparisons of
+size and trait effects across sites would be of great value to
+elucidating the mechanisms through which these shape drought resistance
+(\textbf{check for refs}). Unfortunately, our ability to compare forest
+responses across droughts is hampered by the fact that droughts are
+rarely explicitly defined in ecological studies \citep{slette_how_2019}.
+Further research will be required to understand how drought
+characteristics interact with driver variables to shape tree growth
+responses.
+
+As climate change drives increasing drought in many of the world's
+forests (\emph{REFS-KAT}), the fate of forests and their climate
+feedbacks will be shaped by the biophysical and physiological drivers
+observed here. Large trees have been suffering disrpoportionately in
+forests around the world (\emph{REFS}), and we here show that this is
+primarily driven by their height, with some contributions from canopy
+position. The distinction is important because it suggests that height
+\emph{per se} makes trees vulnerable, even if their crowns are somewhat
+protected by neighbors, whereas solitary trees or the dominant trees in
+young regrowth forests should be less vulnerable. Considering just
+height and crown position, this would suggest that mature forests would
+be more vulnerable to drought than young forests with short trees;
+however, root water access may limit the young forests (Bretfield et
+al.), and species traits often shift as forests age, with early
+successional species tending to have lower wood densities and higher
+hydraulic conductivities that facilitate rapid growth (\textbf{REFS in
+Bretfield-beginning of discussion}). The later successional species at
+our site (\emph{Fagus grandifolia}, \ldots{} ) have a mix of traits
+conferring drought tolerance and resistance (Table 3), and further
+research on how hydraulic traits and drought vulnerability change over
+the course of succession would be valuable for getting at the very
+signficant question of whether old-growth forests are more vulnerable to
+drought, and to which types of drought. In the meantime, the results of
+this study advance our knowledge of the factors conferring drought
+vulnerability and resistance in a mature forest, opening the door for
+more accurate forecasting of forest responses to future drought.
+
+\hypertarget{acknowledgements}{%
+\subsubsection{Acknowledgements}\label{acknowledgements}}
+
+funding: ForestGEO,
+
+\hypertarget{author-contribution}{%
+\subsubsection{Author Contribution}\label{author-contribution}}
 
@@
 
-\subsubsection{Author Contribution}\label{author-contribution}
-
-words
-
+\newpage
+
+\hypertarget{supplementary-information}{%
 \subsubsection{Supplementary
-Information}\label{supplementary-information}
+Information}\label{supplementary-information}}
 
@@
 
-\emph{p50 and p80} We decided to include values of P50 and P80 in the
-leaf traits model, defined by \citep{anderegg_meta-analysis_2016} as the
-water potentials at which a species loses 50\% and 88\% {[}80\% by
-proxy{]}, respectively, of hydraulic conductivity. Values were
-calculated by (\textbf{insert new methods here??}), and were only
-available for six species (\emph{C. glabra}, \emph{L. tulipifera},
-\emph{Q. alba}, \emph{Q. prinus}, \emph{Q. rubra}, and \emph{Q.
-velutina}). Because of this, the model runs were considered to be
-incomplete due to the exclusion of the other 8 species. Results revealed
-neither p50 nor p80 to be significant, thus for the full analysis we
-decided to drop the two traits in order to include all species in the
-full analysis.
-
 \begin{table}[!h]
@@
 2.4 & TWI.ln & ln[topographic.wetness.index] & 1999\\
-3.1 & year & drought.year & all\\
+3.1 & rp & ring.porosity & 1966\\
 3.1 & rp & ring.porosity & 1999\\
@@
 3.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & all\\
+3.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & 1977\\
 \bottomrule
@@
 \midrule
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.00 & trees\_all\_sub & (Intercept) (1.101), position\_alldominant (-0.044), position\_allintermediate (-0.039), position\_allsuppressed (-0.049), height.ln.m (-0.092), TWI.ln (-0.058), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.148)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.84 & trees\_all\_sub & (Intercept) (0.953), height.ln.m (-0.06), TWI.ln (-0.065), year1977 (-0.089), year1999 (-0.076), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.164)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 1.57 & trees\_all\_sub & (Intercept) (1.021), position\_alldominant (-0.046), position\_allintermediate (-0.042), position\_allsuppressed (-0.053), height.ln.m (-0.097), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.009), mean\_TLP\_Mpa (-0.146)\\
-resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1966 & (Intercept) (1.576), height.ln.m (-0.145), rpring (0.091), rpsemi-ring (0.232), PLA\_dry\_percent (-0.021)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.37 & x1966 & (Intercept) (1.505), height.ln.m (-0.146), TWI.ln (0.043), rpring (0.092), rpsemi-ring (0.227), PLA\_dry\_percent (-0.021)\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.00 & trees\_all\_sub & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+PLA\_dry\_percent+(1|sp/tree) & 0.34 & trees\_all\_sub & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.40 & trees\_all\_sub & NA\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp/tree) & 0.52 & trees\_all\_sub & NA\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+PLA\_dry\_percent+(1|sp/tree) & 0.57 & trees\_all\_sub & NA\\
 \addlinespace
-resist.value \textasciitilde{} height.ln.m+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.39 & x1966 & (Intercept) (1.177), height.ln.m (-0.135), PLA\_dry\_percent (-0.014), mean\_TLP\_Mpa (-0.148)\\
-resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.00 & x1977 & (Intercept) (0.386), TWI.ln (-0.136), rpring (-0.233), rpsemi-ring (-0.328), mean\_TLP\_Mpa (-0.37)\\
-resist.value \textasciitilde{} position\_all+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.28 & x1977 & (Intercept) (0.341), position\_alldominant (-0.079), position\_allintermediate (-0.025), position\_allsuppressed (0.039), TWI.ln (-0.128), rpring (-0.23), rpsemi-ring (-0.335), mean\_TLP\_Mpa (-0.385)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.06 & x1977 & (Intercept) (0.469), height.ln.m (-0.032), TWI.ln (-0.133), rpring (-0.223), rpsemi-ring (-0.318), mean\_TLP\_Mpa (-0.368)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1999 & (Intercept) (1.31), position\_alldominant (0.003), position\_allintermediate (-0.077), position\_allsuppressed (-0.1), height.ln.m (-0.09), TWI.ln (-0.083), rpring (0.191), rpsemi-ring (0.224), PLA\_dry\_percent (-0.009)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp/tree) & 0.90 & trees\_all\_sub & NA\\
+resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 0.00 & x1966 & NA\\
+resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.54 & x1966 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.09 & x1966 & NA\\
+resist.value \textasciitilde{} height.ln.m+PLA\_dry\_percent+(1|sp) & 1.43 & x1966 & NA\\
 \addlinespace
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.05 & x1999 & (Intercept) (0.855), position\_alldominant (0.001), position\_allintermediate (-0.078), position\_allsuppressed (-0.1), height.ln.m (-0.092), TWI.ln (-0.08), rpring (0.183), rpsemi-ring (0.06), mean\_TLP\_Mpa (-0.146)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+(1|sp) & 0.55 & x1999 & (Intercept) (1.19), position\_alldominant (0.002), position\_allintermediate (-0.08), position\_allsuppressed (-0.104), height.ln.m (-0.09), TWI.ln (-0.085), rpring (0.199), rpsemi-ring (0.134)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+mean\_TLP\_Mpa+(1|sp) & 0.94 & x1999 & (Intercept) (0.722), position\_alldominant (-0.002), position\_allintermediate (-0.083), position\_allsuppressed (-0.105), height.ln.m (-0.1), rpring (0.183), rpsemi-ring (0.046), mean\_TLP\_Mpa (-0.154)\\
-resist.value \textasciitilde{} TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.07 & x1999 & (Intercept) (1.044), TWI.ln (-0.097), rpring (0.187), rpsemi-ring (0.265), PLA\_dry\_percent (-0.01)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 1.15 & x1999 & (Intercept) (1.191), position\_alldominant (0.001), position\_allintermediate (-0.081), position\_allsuppressed (-0.106), height.ln.m (-0.098), rpring (0.192), rpsemi-ring (0.215), PLA\_dry\_percent (-0.009)\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.98 & x1966 & NA\\
+resist.value \textasciitilde{} position\_all+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.00 & x1977 & NA\\
+resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.09 & x1977 & NA\\
+resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.12 & x1977 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.91 & x1977 & NA\\
 \addlinespace
-resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.41 & x1999 & (Intercept) (0.516), TWI.ln (-0.094), rpring (0.178), rpsemi-ring (0.074), mean\_TLP\_Mpa (-0.166)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.82 & x1999 & (Intercept) (1.084), position\_alldominant (0.002), position\_allintermediate (-0.077), position\_allsuppressed (-0.099), height.ln.m (-0.091), TWI.ln (-0.081), rpring (0.186), rpsemi-ring (0.147), PLA\_dry\_percent (-0.005), mean\_TLP\_Mpa (-0.076)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+(1|sp) & 1.88 & x1999 & (Intercept) (1.065), position\_alldominant (-0.001), position\_allintermediate (-0.085), position\_allsuppressed (-0.11), height.ln.m (-0.098), rpring (0.201), rpsemi-ring (0.123)\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1999 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.09 & x1999 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+(1|sp) & 0.46 & x1999 & NA\\
+resist.value \textasciitilde{} TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 0.96 & x1999 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+mean\_TLP\_Mpa+(1|sp) & 1.03 & x1999 & NA\\
+\addlinespace
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 1.19 & x1999 & NA\\
+resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.33 & x1999 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.85 & x1999 & NA\\
+resist.value \textasciitilde{} position\_all+height.ln.m+rp+(1|sp) & 1.85 & x1999 & NA\\
 \bottomrule
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=5.20833in,height=\textheight]{tables_figures/FigureS3.png}
+\caption{Location of cored trees}
+\end{figure}
+
 \includegraphics{tables_figures/FigureS1.png} (see Issue \#32)

### 2019-10-08 manuscript/climate_sensitivity_draft.Rmd
quick fix on table 2
No longer trying to format in LaTEX. We will need to change this at the proof stage.

--- 2019-09-27 manuscript/climate_sensitivity_draft.Rmd
@@
 hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = TRUE) %>%

### 2019-10-12 test.Rmd
Made sample for question on rstudio community

---
title: "Test"
output: pdf_document
---

``` {r} 
dat <- data.frame(variable = c(rep("turgor loss point", 10), rep("volume", 10)),
                    unit = c(rep("$\\pi_{tlp}$", 10), rep("$cm^3$", 10)),
                    value = c(round(runif(10), 2), round(runif(10,20,30),2)))

kableExtra::kable(dat, booktabs = TRUE, format = 'latex', escape=FALSE)
```
Printing in the table leaves the raw symbology, but if I want to discuss the use of the volume ($cm^3$) and the turgor loss point ($\pi_{tlp}$) in the text itself, the symbols render without a problem.

### 2019-10-12 manuscript/climate_sensitivity_draft.Rmd
Tried fixing errors with latex

@@
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
+```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
+```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = TRUE) %>%
+kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 library(kableExtra)
-table2 <- read.csv("tables_figures/species_table.csv")
+table3 <- read.csv("tables_figures/species_table.csv")
 
@@
 
-kable(table2, booktabs=TRUE, caption = "**Table 3.
+kable(table3, booktabs=TRUE, caption = "**Table 3.
 Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
@@
 This yielded Figure ???
-(cand_full).
-All variables with dAICc >2 were used as candidates for each scenario's best model (figure ????
-(tested_traits_best))
+(candfull).
+All variables with dAICc > 2 were used as candidates for each scenario's best model (figure ????
+(tested traits best))
 
@@
 Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(*Cite Shenanadoah paper, if accepted*).
+(*Cite Shenandoah paper, if accepted*).
 Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
@@
 In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999.
-*Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999.
+*Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.
 
@@
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
+Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2019-10-12 manuscript/climate_sensitivity_draft.tex
Tried fixing errors with latex

@@
 \maketitle
-
-\raggedright
 
@@
 
-\caption{\label{tab:unnamed-chunk-3}**Table 1.
+\caption{\label{tab:Table 1}**Table 1.
 Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable).
@@
 
-\caption{\label{tab:unnamed-chunk-4}**Table 2. Summary of variables**}
+\caption{\label{tab:Table 2}**Table 2. Summary of variables**}
 \centering
@@
 \hspace{1em}\hspace{1em}leaf mass per area & LMA & kg m-2 & ratio of leaf dry mass to fresh leaf area & - & all & 48.69 & 30.68 & 75.8 & no\\
-\hspace{1em}\hspace{1em}xylem porosity & X & - & vessel arrangement in xylem & ring & 408 & - &  &  & -\\
+\hspace{1em}\hspace{1em}xylem porosity & XP & - & vessel arrangement in xylem & ring & 408 & - &  &  & -\\
 \hspace{1em}\hspace{1em} &  &  &  & semi-ring & 31 & - &  &  & -\\
 \hspace{1em}\hspace{1em} &  &  &  & diffuse & 178 & - &  &  & -\\
-\hspace{1em}\hspace{1em}turgor loss point & T & MPa & water potential at which leaves wilt & - & all & -2.39 & -2.76 & -1.92 & no\\
-\hspace{1em}\hspace{1em}percent loss area & PLA & \% & percent loss of leaf area upon dessication & - & all & 13.06 & 8.52 & 24.64 & no\\
+\hspace{1em}\hspace{1em}turgor loss point & $\pi_{tlp}$ & MPa & water potential at which leaves wilt & - & all & -2.39 & -2.76 & -1.92 & no\\
+\hspace{1em}\hspace{1em}percent loss area & PLA & % & percent loss of leaf area upon dessication & - & all & 13.06 & 8.52 & 24.64 & no\\
 \bottomrule
@@
 \end{table}
-
-Caption: T = \(\pi_{tlp}\), X = \({XP}\)
 
@@
 individually against a null model containing height and year. This
-yielded Figure ??? (cand\_full). All variables with dAICc
-\textgreater{}2 were used as candidates for each scenario's best model
-(figure ???? (tested\_traits\_best))
+yielded Figure ??? (candfull). All variables with dAICc \textgreater{} 2
+were used as candidates for each scenario's best model (figure ????
+(tested traits best))
 
@@
 (\emph{Lymantria dispar} L.) from approximately 1988-1995, which most
-stronly impacted \emph{Quercus} spp. (\emph{Cite Shenanadoah paper, if
+stronly impacted \emph{Quercus} spp. (\emph{Cite Shenandoah paper, if
 accepted}). Climatically, these droughts included three of the five
@@
 1966, \#2\% un 1977, and \#\% in 1999. \emph{Some} trees exhibited
-increased growth: (\(R\) \textgreater{} 1.0): \#\% in 1966, \#\% un
+increased growth: (\(R\) \textgreater{} 1.0): \#\% in 1966, \#\% in
 1977, and \#\% in 1999.
@@
 
-\caption{\label{tab:unnamed-chunk-5}**Table 4. Univariate models**}
+\caption{\label{tab:unnamed-chunk-3}**Table 4. Univariate models**}
 \centering
@@
 
-\caption{\label{tab:unnamed-chunk-6}**Table 5.
-Summary of R^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model ($dAICc<2$).}
+\caption{\label{tab:unnamed-chunk-4}**Table 5.
+Summary of R\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).}
 \centering
@@
 \end{table}
-
-\raggedright
 

### 2019-10-16 manuscript/climate_sensitivity_SI.Rmd
Moved SI to another .rmd file

--- 2019-10-12 manuscript/climate_sensitivity_draft.Rmd
@@
 ---
-title: "Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
+title: "Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
 author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
@@
   pdf_document:
-    citation_package: natbib
     fig_caption: yes
-bibliography: [book.bib,  packages.bib]
 biblio-style: apalike
@@
 header-includes:
-- \usepackage{float}
 - \usepackage{booktabs}
-- \usepackage{pdflscape}
-- \newcommand{\blandscape}{\begin{landscape}}
-- \newcommand{\elandscape}{\end{landscape}}
-- \usepackage{caption}
-- \captionsetup[table]{font=small}
-- \captionsetup[figure]{font=small}
-- \captionsetup[table]{labelformat=empty}
-- \captionsetup[figure]{labelformat=empty}
 - \usepackage{dcolumn} #for sidewaystable
@@
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
-), 'packages.bib')
-```
-\raggedright
-
-### Title page
-
-text | word count | other | n
----------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	## (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
-Materials and Methods |	currently ~1442	||	 
-Results |	currently ~1437	 ||
-Discussion |	currently ~1881 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
-Acknowledgements | 	~75 unwritten	||
-
-\newpage
-
-### Summary (**needs work**)
-- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important.
-Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.
-
-- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60 year period from 1950 and 2009.
-
-- Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
-However, there was substantial variation in the best predictor variables across given droughts.
-
-- We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
-
-*The Summary for research papers, which must be usable as a stand- alone document, must not exceed 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion.
-It should not contain citations of other papers.*
-
-*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.
-Very short phrases and scientific names with their common equivalents (e.g.
-Nicotiana tabacum (tobacco)) are acceptable.*
-
-\newpage
-### Introduction
-Forests globally play a critical role in climate regulation ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)), yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change ([Friedlingstein et al.
-2006](https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1)).
-An important aspect of this uncertainty lies in responses to drought (**REF**).
-In many forested regions around the world, the risk of severe drought is increasing ([Trenbert et al.
-2014](https://www.nature.com/articles/nclimate2067)), even in conjunction with increasing precipitation ([IPCC 2014](https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334)).
-Global change-type drought has been affecting forests worldwide ([Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001)), and it is expected that future climate change-driven droughts will severely impact forests around the world ([Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**).
-Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [Stovall et al.
-2019]( https://doi.org/10.1038/s41467-019-12380-6)), resulting in disproportionate impacts on forest C storage ([Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
-As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
-Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
-
-**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
-2019, Hacket-Pain et al.
-2016 (DOI 10.1007/s10342-016-0982-7) **REFS**).
-However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E.
-M., Hicke, J.
-A.
-& Meddens, A.
-J.
-H.
-Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States.
-Glob.
-Change Biol.
-17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N.
-G., Bond, B.
-J., Hill, L., Ryan, M.
-G.
-& Whitehead, D.
-in Size and age related changes in tree structure and function (eds Meinzer, F.C.
-& Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N.
-G.
-& Allen, C.
-D.
-Darcy’s law predicts widespread forest mortality under climate warming.
-Nat.
-Clim.
-Change 5, 669–672 (2015); Ryan, M.
-G., Phillips, N.
-& Bond, B.
-J.
-The hydraulic limitation hypothesis revisited.
-Plant Cell Environ.
-29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al.
-2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)).
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
-Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
-It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance.
-Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al.
-2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
-Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.
-
-**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al.
-2015, Friedrichs et al.
-2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
-However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al.
-2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al.
-2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
-There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al.
-2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested.
-(**But see D'Orangeville et al.
-2018--except that it may not be a well-constructed test for traits**)
-
-**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand ([D'Orangeville et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096)).
-However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-(**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011];  D'Orangeville et al.
-2018 **REFS**), these don't test for differential trait effects across droughts (D'Orangeville et al.
-2018) and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**).
-Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
-While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
-There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; Stovall et al.
-2019).
-[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
-(**PELTIER**)
-
-Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses.
-First, we focus on the role of tree size and its interaction with microenvironement. 
-We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
-We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
-Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
-We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
-Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).
-
-```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
+### Supplementary Information
+```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable).
-Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
-  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(1, width = "12cm") %>%
-  column_spec(2:5, width = "1cm") %>%
-  column_spec(6, width = "2cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kableExtra::landscape()
-```
 
-\clearpage
-
-### Materials and Methods
-*Study site*
-
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al.
-2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
-SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
-Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
-
-*Data collection and preparation*
-
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\le$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**confirm**), are available through the [ForestGEO data portal](www.forestgeo.si.edu).
-
-```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-
-kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
-  column_spec(4, width = "7cm") %>%
-  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  kable_styling()
-```
-
-We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019].
-Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013].
-In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
-Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-*Ryan submitted the data to ITRDB but I don't think its posted yet.
-We should also cite GitHub/Zenodo here.
-I'll come back to that.*
-
-For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
-Prior DBH was estimated using the following equation:
-
-$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$
-
-Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
-Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#) and then used these to estimate bark thickness based on DBH.
-
-Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
-Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
-Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-Both methods are associated with some error [@larjavaara_measuring_2013].
-Species-specific height allometries were developed (Table S#).
-For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
-
-Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2) and height change since *1959* indicated that change was likely slow.
-Specifically, [**Ian, please provide details--e.g., average rate of height growth**]
-
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is (plus citations)**]
-
-Hydraulic traits were collected from SCBI and are summarized in Table 3.
-In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
-These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis.
-Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using an R script (**details**).
-$LMA$ was calculated as the ratio of leaf dry mass to fresh area.
-$PLA$ was calculated as the percent loss of area between fresh and dry leaves.
-$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-We used the rapid determination method of ([Bartlett et al., 2012]) to estimate the turgor loss point ($\pi_{tlp}$).
-Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
-Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ ([Bartlett et al., 2012]).
-We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
-Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
-
-```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table3 <- read.csv("tables_figures/species_table.csv")
-
-#the percentage symbol has a double backslash because otherwise this writes out the latex code when making the table
-
-kable(table3, booktabs=TRUE, caption = "**Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
-Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
-```
-
-(**add description of climate data used in Fig. 1, NEON vertical profiles**)
-
-*Identification of drought years*
-
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period.
-To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
-Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years.
-Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site.
-We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE).
-These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris etal., 2014).
-The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
-
-*Analysis*  (**this needs work**)
-
-Linear mixed models were run following the order of the hypotheses as seen in Figure ???
-[individual_tested_traits].
-Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
-Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions.
-First, we analyzed the combined scenario to determine if "year" was significant.
-Upon establishing this, we tested height and DBH as size parameters.
-Although both were significant, height was kept due to its larger delta AICc compared with the null model.
-We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year.
-This yielded Figure ???
-(candfull).
-All variables with dAICc > 2 were used as candidates for each scenario's best model (figure ????
-(tested traits best))
-
-### Results
-
-*Focal droughts and their characteristics*
-
-![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
-
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig.
-1).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**).
-Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(*Cite Shenandoah paper, if accepted*).
-Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
-Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.*
-**The droughts differed in timing/duration/etc.
-..
-The year 1966 was preceded by two relatively dry years...
-1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry.
-**
-
-Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig.
-1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999.
-*Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.
-
-*Tree size and drought resistance*
-
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4).
-When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/univariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
-  kable_styling()
-```
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
-  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  kable_styling()
-```
-
-\raggedright
-
-Crown position was generally correlated with $H$, but with substantial variation (Fig.
-2**e**).
-Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$.
-When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$.
-Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
-Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
-Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.
-
-In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-2).
-Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-2a).
-Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig.
-2b).
-Air temperature did not vary across the vertical profile (Fig.
-2c).
-[**biological temperature?- probably drop**]
-
-![**Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), (c) air temperature ($T_{air}$), and (d) biological temperature ($T_{biological}$)averaged over each month of the peak growing season (May-August) from YEAR-YEAR.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$.
-In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
-In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
-
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
-*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
-
-*Species' traits, height, and drought resistance*
-
-We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.23; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts.
-$\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4).
-Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
-Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.
-
-We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha **Ian, confirm**), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**).
-However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$.
-Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
-2**f-g**).
-
-We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is *not* driven by more drought-sensitive traits in larger trees, but rather by size itself (Tables 1,5).
-As discussed above, there was little meaningful variation in traits with height at the community level.
-When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $ln[H]$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
-
-*Responses across droughts*
-
-We reject the hypothesis (*H3.1*) that overall community responses varied across droughts.
-Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
-This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig.
-1b).
-
-We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts.
-In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
-However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $\pi_{tlp}$ in the multivariate model for the 1966 drought; Tables 4, 5).
-These differences may very well be random, as opposed to statistically meaningful.
-Among the univariate models, there was no instance predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
-Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-The difference most likely to be real is that ring and semi-ring porous species had lower resistance than diffuse porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
-
-We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
-For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $CP$ (lowest $R$ among supressed trees) and $XP$ (lowest $R$ among diffuse-porous trees) were strongest in 1999 (Tables 4,5).
-
-### Discussion
-
-Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
-The tendencey for larger trees to suffer more, observed here as in forests around the world (**REFS**), was driven primarily by their height.
-There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **).
-There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [Zuleta et al.](https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1&isAllowed=y)), and larger rooting volume provided no advantage in the drier microenvironments.
-The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
-The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
-These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.
-
-The droughts considered here were of similar severity (Fig.
-1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
-1a, @helcoski_growing_2019).
-Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature.
-While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
-1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5).
-Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
-These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001)), many of which have shaped our understanding about the role of tree size (Bennett et al.
-2015; Stovall et al.
-2019) and--to some extent-- traits (**REFS**).
-Nevertheless, our results are consistent with findings from more extreme droughts.
-
-Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height.
-To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**).
-They also face different microenvironments, which are partially related to canopy position.
-Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig.
-1a-b; **REFS- Jarvis 1984?**; [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)).
-**The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig.
-2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al.
-2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake.
-**[McDowell-vertical profiles in isotopes?*]**.
-Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
-2**e?**) ([Mueller-Landau et al.
-2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)).
-We show that height is a far stronger predictor than crown position (Tables 1,4,5).
-Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
-However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**).
-The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
-Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
-
-Beyond microenvironment, multiple tree characteristics covary with height.
-To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ...
-(e.g., [Anderson-Teixeira et al.
-2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470); **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**).
-These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al.
-2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
-However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
-Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ...
-(**describe differences**) ([Couvreur et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al.
-2001](https://watermark.silverchair.com/21-12-13-951.pdf)).
-Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
-On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height.
-In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
-Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses (*Couvreur et al.
-2018*).
-
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
-Concerted measurement of leaf hydraulic traits of emerging importance (**REFS--KAT/NOBBY/LAWREN**) allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
-(**REFS**).
-It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).
-
-Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase (Stovall et al.
-2019).
-...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).
-Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].
-Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
-
-As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been suffering disrpoportionately in forests around the world (*REFS*), and we here show that this is primarily driven by their height, with some contributions from canopy position.
-The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
-The later successional species at our site (*Fagus grandifolia*, ...
-) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
-In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
-
-### Acknowledgements
-funding: ForestGEO,
-
-### Author Contribution 
-words
-
-\newpage
-### Supplementary Information
-```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
 heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)

### 2019-10-16 manuscript/climate_sensitivity_draft.Rmd
Added all references from introduction, and updated text with citations

@@
 ### Introduction
-Forests globally play a critical role in climate regulation ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)), yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change ([Friedlingstein et al.
-2006](https://journals.ametsoc.org/doi/full/10.1175/JCLI3800.1)).
+Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies in responses to drought (**REF**).
-In many forested regions around the world, the risk of severe drought is increasing ([Trenbert et al.
-2014](https://www.nature.com/articles/nclimate2067)), even in conjunction with increasing precipitation ([IPCC 2014](https://www.cambridge.org/core/books/climate-change-2014-impacts-adaptation-and-vulnerability-part-b-regional-aspects/036A899BD52861D61B0D519C5F2B9334)).
-Global change-type drought has been affecting forests worldwide ([Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001)), and it is expected that future climate change-driven droughts will severely impact forests around the world ([Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001); **REFS**).
-Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [Stovall et al.
-2019]( https://doi.org/10.1038/s41467-019-12380-6)), resulting in disproportionate impacts on forest C storage ([Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**).
+Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [@stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
 As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
@@
 
-**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (Stovall et al.
-2019, Hacket-Pain et al.
-2016 (DOI 10.1007/s10342-016-0982-7) **REFS**).
+**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**).
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees (*Pfeifer, E.
-M., Hicke, J.
-A.
-& Meddens, A.
-J.
-H.
-Observations and modeling of aboveground tree carbon stocks and fluxes following a bark beetle outbreak in the western United States.
-Glob.
-Change Biol.
-17, 339–350 (2011).*)--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction (* McDowell, N.
-G., Bond, B.
-J., Hill, L., Ryan, M.
-G.
-& Whitehead, D.
-in Size and age related changes in tree structure and function (eds Meinzer, F.C.
-& Niinemets, U.) 255–286 (Springer Publishing, 2011); McDowell, N.
-G.
-& Allen, C.
-D.
-Darcy’s law predicts widespread forest mortality under climate warming.
-Nat.
-Clim.
-Change 5, 669–672 (2015); Ryan, M.
-G., Phillips, N.
-& Bond, B.
-J.
-The hydraulic limitation hypothesis revisited.
-Plant Cell Environ.
-29, 367–381 (2006).*), and this may become a greater liability during drought ([Zhang et al.
-2009](https://www.researchgate.net/publication/26322653_Size-dependent_mortality_in_a_Neotropical_savanna_tree_The_role_of_height-related_adjustments_in_hydraulic_architecture_and_carbon_allocation)).
+This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
+First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009].
 Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
@@
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size ([Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
 It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance.
-Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species ([Condit et al.
-2000](https://doi-org.smithsonian.idm.oclc.org/10.2307/3236572); [Meakem et al.
-2018](https://nph.onlinelibrary.wiley.com/doi/10.1111/nph.14633)).
+Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
 Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.
@@
 **A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019](Elliot et al.
+Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019] *(Elliot et al.
 2015, Friedrichs et al.
-2009) have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
-However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; *Scoffoni et al.
-2013- DOI: 10.1104/pp.113.221424*) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs ([Bartlett et al.
-2016](https://www.pnas.org/content/113/46/13098.short)), which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ ([Medeiros et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229)).
-There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought ([Anderegg et al.
-2018 (**but review this pub**)](https://doi.org/10.1038/s41586-018-0539-7)), but the ability of hydraulic traits to tree performance under drought remains untested.
+2009) unsure which ones these are* have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
+However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
+There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought [@anderegg_hydraulic_2018] **review this pub**, but the ability of hydraulic traits to tree performance under drought remains untested.
 (**But see D'Orangeville et al.
@@
 
-**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand ([D'Orangeville et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/gcb.14096)).
+**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
 However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-(**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011];  D'Orangeville et al.
-2018 **REFS**), these don't test for differential trait effects across droughts (D'Orangeville et al.
-2018) and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001); [@bennett_larger_2015]; Stovall; **MORE REFS**).
+(**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
+The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; Stovall; **MORE REFS**).
 Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
 While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
-There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions ([@bennett_larger_2015]; Stovall et al.
-2019).
+There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
 [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
@@
 \newpage
-### Supplementary Information
-```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
-heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
-kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
-tableS2 <- read.csv("tables_figures/candidate_traits.csv", stringsAsFactors = FALSE)
-kable(tableS2, booktabs=TRUE, format = "latex", caption = "Candidate variables for best model") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-```{r, Plot, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
-  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
-```
-
-*how do we want to present Table S3? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
-```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/top_models_dAIC.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Top model variations for each drought scenario, with dAICc values <= 2") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-![Location of cored trees](tables_figures/FigureS3.png){width=500px}
-
-![Figure S1](tables_figures/FigureS1.png)
-(see Issue #32)

### 2019-10-19 manuscript/climate_sensitivity_draft.Rmd
Updated some text

@@
 Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\le$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
@@
 
+\noindent\hrulefill 
 ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-Height measurements (n=# trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
+Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
 Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
@@
 Both methods are associated with some error [@larjavaara_measuring_2013].
-Species-specific height allometries were developed (Table S#).
+Species-specific height allometries were developed (Table S1).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
 
-Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from [@jennings_assessing_1999], whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
 As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
 While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2) and height change since *1959* indicated that change was likely slow.
-Specifically, [**Ian, please provide details--e.g., average rate of height growth**]
+2) and height change since the beginning of the study period indicated that change was likely slow.
+Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, and ~1m from 1977 to 1999.
 

### 2019-10-19 scripts/canopy_heights.R
Updated heights script and regression equations

@@ -20,18 +20,6 @@ dbh_2013$dbh.cm <- dbh_2013$dbh/10 #cm
 
 dbh_2018 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
 
-# #get stemIDs for each stem (leaving this here if need be)
-# heights$stemID <- ifelse(is.na(heights$stemID), dbh_2013$stemID[match(paste(heights$tag, heights$stemtag), paste(dbh_2013$tag, dbh_2013$StemTag))], heights$stemID)
-# 
-# heights$tag <- ifelse(is.na(heights$tag), dbh_2013$tag[match(heights$stemID, dbh_2013$stemID)], heights$tag)
-
-#separate stemID in 2018 data
-dbh_2018$tag <- gsub("_.*$", "", dbh_2018$Tree_ID_Num)
-dbh_2018$stemtag <- gsub("[[:digit:]]*_", "", dbh_2018$Tree_ID_Num)
-dbh_2018$tag <- as.numeric(as.character(dbh_2018$tag))
-dbh_2018$stemtag <- as.numeric(as.character(dbh_2018$stemtag))
-
-
 heights$dbh_regr.cm <- paste0(heights$DBH.2008.cm, heights$DBH.2013.cm, heights$DBH.TLS.2015.cm, heights$DBH.2018.cm)
 heights$dbh_regr.cm <- gsub("NA", "", heights$dbh_regr.cm)
 heights$dbh_regr.cm <- as.numeric(heights$dbh_regr.cm)
@@ -49,7 +37,7 @@ heights_nodup <- heights[!heights$stemID %in% dup, ]
 
 heights_recent <- NULL
 for (i in seq(along=unique(heights_dup$stemID))){
-  sub <- heights_dup[heights_dup$stemID == unique(heights_dup$stemID)[[i]], ]
+  sub <- heights_dup[heights_dup$stemID %in% unique(heights_dup$stemID)[[i]], ]
   
   sub <- sub[sub$height.year == max(sub$height.year), ]
   
@@ -57,7 +45,7 @@ for (i in seq(along=unique(heights_dup$stemID))){
 }
 
 heights_all <- rbind(heights_nodup, heights_recent)
-heights_all <- heights_all[c(1:4,6,8,15:17,24:25)]
[…]

### 2019-10-19 scripts/canopy_position_analysis.R
Made loop for easier height changes

@@ -674,7 +674,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
 ##leaf traits gained from this include PLA_dry_percent, LMA_g_per_m2, Chl_m2_per_g, and WD [wood density]
 
-leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEN4O7QACR7L5X4AHK25OZHZM"), stringsAsFactors = FALSE)
+leaf_traits <- read.csv(text=getURL("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEPDTUEGGQUCPMMGPQ25WRPM4"), stringsAsFactors = FALSE)
 
 leaf_traits <- leaf_traits[, c(1,8,12,26,28)]
 
@@ -1120,20 +1120,29 @@ trees_all$sap_ratio <- dbh$sap_ratio[match(trees_all$tree, dbh$tree) & match(tre
 ##5g. add in tree heights ####
 ## taken from the canopy_heights script
 #the full equation is using all points for which we have data to create the equation, despite that for several species we don't have enough data to get a sp-specific equation
-trees_all$height.ln.m <- ifelse(trees_all$sp == "caco", (0.348+0.808*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "cagl", (0.681+0.704*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "caovl", (0.621+0.722*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "cato", (0.776+0.701*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "fagr", (0.708+0.662*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "fram", (0.715+0.619*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "juni", (1.22+0.49*trees_all$dbh.ln.cm),
-                      ifelse(trees_all$sp == "litu", (1.32+0.524*trees_all$dbh.ln.cm),
[…]

### 2019-10-20 manuscript/climate_sensitivity_draft.Rmd
Added more references and updated text

@@
 
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [[Bourg et al.
-2013](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/13-0010.1); @andersonteixeira_ctfs-forestgeo:_2015].
+Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
 SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
@@
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R. [** Ian, include a brief explanation of what this is (plus citations)**]
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.
+Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
+TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on run-off tendencies of the plot.
+The calculation of TWI here performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
 
@@
 $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-We used the rapid determination method of ([Bartlett et al., 2012]) to estimate the turgor loss point ($\pi_{tlp}$).
+We used the rapid determination method of @bartlett_rapid_2012 to estimate the turgor loss point ($\pi_{tlp}$).
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
-Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ ([Bartlett et al., 2012]).
+Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
 We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
@@
 Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
-Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions.
+Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].
+Null models were determined in order of the predictions.
 First, we analyzed the combined scenario to determine if "year" was significant.
@@
 2c).
-[**biological temperature?- probably drop**]
 
@@
 We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha **Ian, confirm**), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**).
+In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**).
 However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$.
@@
 There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **).
-There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [Zuleta et al.](https://repository.si.edu/bitstream/handle/10088/32814/Zuleta-et-al_2017_Ecology.pdf?sequence=1&isAllowed=y)), and larger rooting volume provided no advantage in the drier microenvironments.
+There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
@@
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [Allen et al.
-2015](https://doi.org/10.1016/j.foreco.2009.09.001)), many of which have shaped our understanding about the role of tree size (Bennett et al.
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size (Bennett et al.
 2015; Stovall et al.
@@
 Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig.
-1a-b; **REFS- Jarvis 1984?**; [Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)).
+1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
 **The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
@@
 Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
-2**e?**) ([Mueller-Landau et al.
-2016](https://www.ncbi.nlm.nih.gov/pubmed/16643303)).
+2**e?**) [@muller-landau_testing_2006].
 We show that height is a far stronger predictor than crown position (Tables 1,4,5).
@@
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**).
-The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age ([Bretfeld et al](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15071)), such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
+The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.

### 2019-10-20 manuscript/climate_sensitivity_SI.Rmd

@@
 heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
+heights[,2] <- NULL
 kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%

### 2019-10-20 manuscript/climate_sensitivity_draft.Rmd

@@
 
-(**add description of climate data used in Fig. 1, NEON vertical profiles**)
+To determine a vertical height climate profile, data was measured by a National Ecological Observation Network (NEON) eddy-covariance (?) flux tower located within a 1km from the study area.
+In total, wind, relative humidity, and air temperature data were collected for the years 2016-2018 [@noauthor_national_nodate].
+After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
+The resultant plots (Figure 2) illustrate the maximum and minimum of these mean values, with error represented by the standard deviation.
 
@@
 We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE).
-These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris etal., 2014).
+These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
 The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
@@
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size (Bennett et al.
-2015; Stovall et al.
-2019) and--to some extent-- traits (**REFS**).
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent-- traits (**REFS**).
 Nevertheless, our results are consistent with findings from more extreme droughts.
@@
 Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig.
-2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** ([Koike et al.
-2001](https://watermark.silverchair.com/21-12-13-951.pdf); MORE REFS), implying greater water costs of $CO_2$ uptake.
+2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake.
 **[McDowell-vertical profiles in isotopes?*]**.
@@
 To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ...
-(e.g., [Anderson-Teixeira et al.
-2015_scaling](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.12470); **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**).
+(e.g., [@andersonteixeira_size-related_2015; **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)].
 These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al.
@@
 Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ...
-(**describe differences**) ([Couvreur et al.
-2018](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/pce.13322?casa_token=92wZmmL9EhQAAAAA%3A2cq5yAR0_pyM9TVBa5o_DKkhDuHwIUFryxZ43zsdTgQyEa8f0ZfiEHZ5SHV7fFB49UkzXNvL7nMwabU); [Koike et al.
-2001](https://watermark.silverchair.com/21-12-13-951.pdf)).
+(**describe differences**) [@couvreur_water_2018; @koike_leaf_2001].
 Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
@@
 In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
-Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses (*Couvreur et al.
-2018*).
+Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].
 
@@
 Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase (Stovall et al.
-2019).
-...**] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).
+[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase** [@stovall_tree_2019].
+...] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).
 Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].

### 2019-10-26 manuscript/climate_sensitivity_draft.Rmd
Updated file names

@@
 A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
-It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resiliance.
+It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience.
 Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
@@
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
@@
 library(kableExtra)
-table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 
@@
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to DBH (Table S#) and then used these to estimate bark thickness based on DBH.
+Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
+Additional species were used here than used in the final analysis to establish base averages.
+(**do we want to keep this or take out juni/fram from table?**)
 
@@
 Both methods are associated with some error [@larjavaara_measuring_2013].
-Species-specific height allometries were developed (Table S1).
+Species-specific height allometries were developed (Table S2).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
@@
 library(kableExtra)
-table3 <- read.csv("tables_figures/species_table.csv")
+table3 <- read.csv("tables_figures/table3_species_table.csv")
 
@@
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/univariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
+univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+kable(univar_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5.
+multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+kable(multivar_table, booktabs = TRUE, caption = "**Table 5.
 Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
@@
 **[McDowell-vertical profiles in isotopes?*]**.
-Taller trees are more likely to be in domiant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
+Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
 2**e?**) [@muller-landau_testing_2006].

### 2019-10-26 scripts/canopy_position_analysis.R

@@ -725,7 +725,7 @@ elev <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/S
 
 trees_all$elev.m <- elev$dem_sigeo[match(trees_all$tree, elev$tag)]
 
-# ##5d. add in distance to water (here for reference) ####
+##5d. add in distance to water (here for reference) ####
 # ## mapping code here is taken from survey_maps.R in Dendrobands Rscripts folder.
 # 
 # ## I have not found a way to make this not involve personal directories without moving all the data to my folder, which I'm hesitant about doing due to data redundancy.
@@ -868,7 +868,7 @@ bark$diam_nobark_2008.mm <- bark$DBH.mm.2008 - 2*bark$bark.depth.mm
 #3. Fit a linear model, and use model to predict log(bark.depth.mm)
 source_gist("524eade46135f6348140")
 ggplot(data = bark, aes(x = log(diam_nobark_2008.mm), y = log(bark.depth.mm))) +
-  stat_smooth_func(geom="text",method="lm",hjust=0.16, vjust=-1,parse=TRUE) +
+  stat_smooth_func(geom="text",method="lm",hjust=0.16, vjust=-0.5,parse=TRUE) +
   geom_smooth(method="lm", se=FALSE, color="black") +
   geom_point(color = "#0c4c8a") +
   theme_minimal() +
@@ -882,19 +882,20 @@ ggplot(data = bark, aes(x = log(diam_nobark_2008.mm), y = log(bark.depth.mm))) +
   theme_minimal()
 
 #no total regression equation at bottom because all species are accounted for in dataset.
+##these equations are also found in tableS1_bark_regression.csv
 bark$predict_barkthick.ln.mm <- NA
 bark$predict_barkthick.ln.mm <- 
-                  ifelse(bark$species == "caco", -1.56+0.416*log(bark$diam_nobark_2008.mm),
-                  ifelse(bark$species == "cagl", -0.393+0.268*log(bark$diam_nobark_2008.mm),
-                  ifelse(bark$species == "caovl", -2.18+0.651*log(bark$diam_nobark_2008.mm),
-                  ifelse(bark$species == "cato", -0.477+0.301*log(bark$diam_nobark_2008.mm),
[…]

### 2019-10-26 scripts/graphs_for_manuscript.R

@@ -250,7 +250,7 @@ plot_pla_twi
 plot_tlp_twi
 
 traits <- ggarrange(plot_tlp_twi, plot_pla_twi, nrow=1, ncol=2)
-ggsave("manuscript/tables_figures/FigureS1.png", width=5, height=7, units="in", traits)
+ggsave("manuscript/tables_figures/figureS3_traits_with_twi.png", width=5, height=7, units="in", traits)
 
 ##4b. Export map of plot using TWI and cored tree locations ####
 species <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
@@ -267,7 +267,7 @@ cored_points <- SpatialPointsDataFrame(data.frame(species$NAD83_X, species$NAD83
 
 plot.new()
 
-png("manuscript/tables_figures/FigureS3.png", width=5, height=7, units="in", res=300)
+png("manuscript/tables_figures/figureS2_location_cored_trees.png", width=5, height=7, units="in", res=300)
 levelplot(topo, margin=FALSE, scales=list(draw=FALSE), 
           key=list(space="right",
                    text=list(lab=levels(cored_points@data$sp_fact)),

### 2019-10-26 manuscript/climate_sensitivity_SI.Rmd
Updated file names and added bark regression

--- 2019-10-16 manuscript/climate_sensitivity_SI.Rmd
@@
 
-heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
+bark <- read.csv("tables_figures/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
+bark[,2] <- NULL
+kable(bark, booktabs=TRUE, format = "latex", caption = "Species-specific bark thickness regression equations") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+
+heights <- read.csv("tables_figures/tableS2_height_regression.csv", stringsAsFactors = FALSE)
+heights[,2] <- NULL
 kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
@@
 
-```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
-tableS2 <- read.csv("tables_figures/candidate_traits.csv", stringsAsFactors = FALSE)
+```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
+tableS2 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
 kable(tableS2, booktabs=TRUE, format = "latex", caption = "Candidate variables for best model") %>%
@@
 
-```{r, Plot, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
-  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
+```{r Figure S1, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
+  knitr::include_graphics("tables_figures/figureS1_ForestGEO_plot.jpg")
 ```
@@
 ```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/top_models_dAIC.csv", stringsAsFactors = FALSE)
+tableS3 <- read.csv("tables_figures/tableS3_top_models_dAIC.csv", stringsAsFactors = FALSE)
 kable(tableS3, booktabs=TRUE, format = "latex", caption = "Top model variations for each drought scenario, with dAICc values <= 2") %>%
@@
 
-![Location of cored trees](tables_figures/FigureS3.png){width=500px}
+![Location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
 
-![Figure S1](tables_figures/FigureS1.png)
+![Figure S3](tables_figures/figureS3_traits_with_twi.png)
 (see Issue #32)

### 2019-10-27 manuscript/climate_sensitivity_draft.Rmd
Updated pdf (without table2)

@@
 
-Linear mixed models were run following the order of the hypotheses as seen in Figure ???
-[individual_tested_traits].
-Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
-Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].
+Linear mixed models were run following the order of the hypotheses as seen in ???
+**If we decide to include this table, it is "tested_traits_all.csv", but I'm not sure if we want to.**.
+Using the pointRes package [@R-pointRes], we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
+Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [lme4: @R-lme4; AICcmodavg: @R-AICcmodavg].
 Null models were determined in order of the predictions.
@@
 We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year.
-This yielded Figure ???
-(candfull).
-All variables with dAICc > 2 were used as candidates for each scenario's best model (figure ????
-(tested traits best))
+All variables with dAICc > 2 were used as candidates for each scenario's best model (Figure S3).
 
@@
 1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999.
+In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% in 1977, and #% in 1999.
 *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.
@@
 
-In the non-drought years for which we have vertical profiles in climate data (**YEARS**), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
+In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
 2).
@@
 ![**Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), (c) air temperature ($T_{air}$), and (d) biological temperature ($T_{biological}$)averaged over each month of the peak growing season (May-August) from YEAR-YEAR.
+Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
 In these plots, heights are slightly offset for visualization purposes.
-Also shown are (e) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (f) $PLA_{dry}$ and (g) $\pi_{tlp}$.
-In (f-g), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
+Also shown are (d) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (e) $PLA_{dry}$ and (f) $\pi_{tlp}$.
+In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
 In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
@@
 Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
-2**f-g**).
+2**e-f**).
 

### 2019-10-27 manuscript/climate_sensitivity_SI.Rmd
Updated SI with new file names

--- 2019-10-16 manuscript/climate_sensitivity_SI.Rmd
@@
 - \usepackage{booktabs}
-- \usepackage{dcolumn} #for sidewaystable
+- \usepackage{caption}
+- \captionsetup[figure]{labelformat=empty}
 ---
@@
 ### Supplementary Information
-```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
+```{r Table S1, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
@@
 
-heights <- read.csv("tables_figures/height_regression.csv", stringsAsFactors = FALSE)
-kable(heights, booktabs=TRUE, format = "latex", caption = "Species-specific height regression equations") %>%
+bark <- read.csv("tables_figures/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
+bark[,2] <- NULL
+kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 
-```{r, Table S2, eval=TRUE, echo=FALSE, message=FALSE}
-tableS2 <- read.csv("tables_figures/candidate_traits.csv", stringsAsFactors = FALSE)
-kable(tableS2, booktabs=TRUE, format = "latex", caption = "Candidate variables for best model") %>%
+```{r Table S2, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+
+heights <- read.csv("tables_figures/tableS2_height_regression.csv", stringsAsFactors = FALSE)
+heights[,2] <- NULL
+kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 
-```{r, Plot, echo=FALSE, out.width='100%', fig.cap = "Map of ForestGEO plot", fig.pos='ht'}
-  knitr::include_graphics("tables_figures/ForestGEO_plot.jpg")
-```
-
-*how do we want to present Table S3? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
 ```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/top_models_dAIC.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Top model variations for each drought scenario, with dAICc values <= 2") %>%
+tableS3 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Candidate variables for best model") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 
-![Location of cored trees](tables_figures/FigureS3.png){width=500px}
+```{r Figure S1, echo=FALSE, out.width='100%', fig.cap = "Figure S1: Map of ForestGEO plot", fig.pos='ht'}
+  knitr::include_graphics("tables_figures/figureS1_ForestGEO_plot.jpg")
+```
 
-![Figure S1](tables_figures/FigureS1.png)
+\newpage
+*how do we want to present Table S4? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
+```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+tableS4 <- read.csv("tables_figures/tableS4_top_models_dAIC.csv", stringsAsFactors = FALSE)
+kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Top model variations for each drought scenario, with dAICc values <= 2") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+![Figure S2: Location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
+
+\newpage
+![Figure S3: Traits with TWI](tables_figures/figureS3_traits_with_twi.png)
 (see Issue #32)

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
changed title
took out "microhabitat", as its effect is really weak, added "across droughts"

@@
 ---
-title: "Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
+title: "Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest"
 author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
@@
 
-- Individual-level growth responses to the three individual droughts  were stronger in three cases: taller trees in dominant canopy positions, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
+- Individual-level growth responses to the three individual droughts were stronger in three cases: taller trees, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
 However, there was substantial variation in the best predictor variables across given droughts.
 
-- We conclude that when droughts occur, large dominant trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
+- We conclude that when droughts occur, tall trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
 
@@
   pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**confirm**), are available through the [ForestGEO data portal](www.forestgeo.si.edu).
+Census data, which were last updated in 2018 (**confirm**), are available through the ForestGEO data portal(www.forestgeo.si.edu).
 
 \noindent\hrulefill 
-```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
+# ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
   pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 12, 14, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 15, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
   kable_styling()

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
worked on references

@@
 
-**One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.** @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**).
+One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
+@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**).
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
@@
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water (**REFS?**); however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
 A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
@@
 
-**A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.** To link drought responss to fundamental physiological characteristics, and because measuring and modeling drought responses of every species is infeasible in diverse forests, it is important to understand how traits shape drought responses.
-Commonly measured traits including wood density ($WD$) (**REFS**), leaf mass per area ($LMA$) [@abrams_adaptations_1990; @guerfel_impacts_2009], and xylem architecture [@kannenberg_linking_2019] *(Elliot et al.
-2015, Friedrichs et al.
-2009) unsure which ones these are* have been linked to drought responses in temperate deciduous forests--as well as in other forest biomes (**REFS**).
-However, these traits have less direct linkage to plant hydraulic function than leaf hydraulic traits such as leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], which are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
-There is also evidence that ecosystems with high hydraulic trait diversity--but not traits like $LMA$ and $WD$-- show more modest ecosystem flux variation in response to drought [@anderegg_hydraulic_2018] **review this pub**, but the ability of hydraulic traits to tree performance under drought remains untested.
-(**But see D'Orangeville et al.
-2018--except that it may not be a well-constructed test for traits**)
+A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.
+Xylem architecture plays a role, with diffuse porous species tending to be more drought sensitive than ring-porous species [@kannenberg_linking_2019] *(Elliot et al.
+2015-doi: 10.1111/gcb.13045, Friedrichs et al.
+2009-doi:10.1007/s00468-009-0315-2)*, but does not differentiate species beyond broad classes.
+To link species' level drought responses to fundamental physiological characteristics, it is important to understand how traits shape drought responses.
+Commonly measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009]*(Hoffmann et al.
+2011- doi: 10.1111/j.1365-2486.2011.02401.x)* and other forest biomes around the world (*Greenwood et al.
+2007- doi: 10.1111/ele.12748*).
+However, the direction of response is not always consistent; for example, wood density correlated negatively with drought tree drought performance  in a broadleaf deciduous forest in the southeast US *(Hoffmann et al.
+2011)* but positively across a global scale (*Greenwood et al.
+2007- doi: 10.1111/ele.12748*).
+Leaf hydraulic traits with more direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
+The ability of these hydraulic traits to tree performance under drought remains untested.
 
-**A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.** No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
+A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.
+No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
 However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
@@
 First, we focus on the role of tree size and its interaction with microenvironement. 
-We confirm that, consistent with most forests globally, larger-diameter trees have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
+We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
 We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow predicted and observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
changed predicted direction of WD response
following Hoffmann et al. 2011, who found negative response in NC.
WD never had a sig effect, so this has no impact on results/interpretation.

@@
 2007- doi: 10.1111/ele.12748*).
-However, the direction of response is not always consistent; for example, wood density correlated negatively with drought tree drought performance  in a broadleaf deciduous forest in the southeast US *(Hoffmann et al.
+However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeast US *(Hoffmann et al.
 2011)* but positively across a global scale (*Greenwood et al.

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 2009-doi:10.1007/s00468-009-0315-2)*, but does not differentiate species beyond broad classes.
-To link species' level drought responses to fundamental physiological characteristics, it is important to understand how traits shape drought responses.
 Commonly measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009]*(Hoffmann et al.
@@
 2007- doi: 10.1111/ele.12748*).
-However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeast US *(Hoffmann et al.
-2011)* but positively across a global scale (*Greenwood et al.
-2007- doi: 10.1111/ele.12748*).
-Leaf hydraulic traits with more direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
+However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeastern United States *(Hoffmann et al.
+2011)* but positively or with no effect at a global scale (*Greenwood et al.
+2007- doi: 10.1111/ele.12748; Anderegg et al.
+2016- doi: 10.1073/pnas.1525678113*).
+Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance (*Anderegg et al.
+2016*) but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses highly diverse forests (*e.g.*, in the tropics).
+More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
 The ability of these hydraulic traits to tree performance under drought remains untested.
@@
 However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-(**BUT ARE WE MISSING SOMETHING??**) While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; Stovall; **MORE REFS**).
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
+The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019] *Anderegg et al.
+2016*).
 Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
@@
 [*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
-(**PELTIER**)
 
@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density and specific leaf area, but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; *Hoffmann et al.
+2011*) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**confirm**), are available through the ForestGEO data portal(www.forestgeo.si.edu).
+Census data, which were last updated in 2018 (**Ian, confirm**), are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
 \noindent\hrulefill 
-# ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
+ ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 
-We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity (ANPP), which together comprised 97% of study plot ANPP between 2008 and 2013 [@helcoski_growing_2019].
+We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019].
 Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
@@
 
-$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$
+$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$$
 
-Here, *Y* denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
+Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
 Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
-Additional species were used here than used in the final analysis to establish base averages.
-(**do we want to keep this or take out juni/fram from table?**)
 
@@
 Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
-Rangefinders either used the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-Both methods are associated with some error [@larjavaara_measuring_2013].
+Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
+Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
 Species-specific height allometries were developed (Table S2).
-For species with insufficient height data to create reliable species-specific allometries, heights were calculated from equations derived from all species in the study.
+For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
 

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
working on Table 3

@@
 \noindent\hrulefill 
- ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
+ #```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 
-Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol from @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
 As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
@@
 TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on run-off tendencies of the plot.
-The calculation of TWI here performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
+*Based on our observations in the plot* (**Ian, is this what you mean?**), the calculation of TWI here performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
 
-Hydraulic traits were collected from SCBI and are summarized in Table 3.
+Hydraulic traits were collected at SCBI (Table 3).
 In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
@@
 
-#the percentage symbol has a double backslash because otherwise this writes out the latex code when making the table
+kable(table3, booktabs=TRUE, caption = "**Table 3.
+Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
 
-kable(table3, booktabs=TRUE, caption = "**Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.
-Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
 ```
 
-To determine a vertical height climate profile, data was measured by a National Ecological Observation Network (NEON) eddy-covariance (?) flux tower located within a 1km from the study area.
+To determine a vertical height climate profile, data was measured by a National Ecological Observation Network (NEON) tower located <1km from the study area.
 In total, wind, relative humidity, and air temperature data were collected for the years 2016-2018 [@noauthor_national_nodate].
 After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
-The resultant plots (Figure 2) illustrate the maximum and minimum of these mean values, with error represented by the standard deviation.
 

### 2019-10-29 manuscript/climate_sensitivity_draft.Rmd
various updates

@@
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable).
-Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
+Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and all top multivariate models (when applicable).
+Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
   pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H2.2. At the community level, taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2.3. When traits are accounted for, taller trees still have lower drought resistance.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H3.1. Resistance differs across the droughts considered here.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
   pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   kable_styling(bootstrap_options = "striped", full_width = FALSE)
-
 ```
 
-To determine a vertical height climate profile, data was measured by a National Ecological Observation Network (NEON) tower located <1km from the study area.
-In total, wind, relative humidity, and air temperature data were collected for the years 2016-2018 [@noauthor_national_nodate].
+To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area.
+We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_nodate].
 After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
@@
 
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events where tree growth was substantially reduced and where peak growing season climatic conditions were among the driest of the time period.
-To identify years with widespread reductions in tree growth, we used the pointRes package [@R-pointRes] in R (version 3.5.3) to determine drought periods based on trees' drought resistance, which is defined as the ratio between the performance during and before the disturbance [@lloret_components_2011].
-Specifically, we looked at the ratio between annual basal area increment (BAI) in the year of the drought to average annual BAI in the 5 preceding years.
-Candidate drought years were defined if >25% of the cored trees experienced <30% growth in a year compared to the previous 5 years.
-Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the months (of the current year) to which annual growth was most sensitive for trees at this site.
-We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between potential evapotranspiration (PET) and precipitation (PRE).
+We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
+Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R (version 3.5.3).
+Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site.
+We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE).
 These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
@@
 
-*Analysis*  (**this needs work**)
+*Analysis*
 
-Linear mixed models were run following the order of the hypotheses as seen in ???
-**If we decide to include this table, it is "tested_traits_all.csv", but I'm not sure if we want to.**.
-Using the pointRes package [@R-pointRes], we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable.
-Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [lme4: @R-lme4; AICcmodavg: @R-AICcmodavg].
-Null models were determined in order of the predictions.
-First, we analyzed the combined scenario to determine if "year" was significant.
-Upon establishing this, we tested height and DBH as size parameters.
-Although both were significant, height was kept due to its larger delta AICc compared with the null model.
-We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year.
-All variables with dAICc > 2 were used as candidates for each scenario's best model (Figure S3).
+For each drought period, we calculated drought *resistance* ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011].
+Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable.
+The general statistical model for hypothesis testing was a mixed effects model (*lme4* package in R: [@R-lme4]) with $R$ as the response variable, tree *nested within species* (**IAN, CONFIRM**) as a random effect, and one or more independent variables as fixed effects.
+We used AICc (*AICcmodavg* package: [@R-AICcmodavg]) to assess model fit.
+Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
+We first tested the predictor variables idependently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
+Because height had a substantial influence on $R$, we included it in the null model for testing other variables (Table 4).
+
+Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R.
+All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S3).
+We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
+These were counted as the "top multivariate models".
+When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models.
+If the variable appeared in only some of the models, we considered this partial support.
 

### 2019-10-30 manuscript/climate_sensitivity_draft.Rmd
Updated with percentages for densities

@@
 \noindent\hrulefill 
- #```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
+ ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+
+kable(table2, booktabs=TRUE)
+
+```
 
@@
   kable_styling()
-```
 
@@
 1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% in 1977, and #% in 1999.
-*Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.
+In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
+*Some* trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
 

### 2019-10-30 scripts/graphs_for_manuscript.R

@@ -363,6 +363,22 @@ library(cowplot)
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv")
 trees_all_sub$year <- as.character(trees_all_sub$year)
 
+#In each drought, roughly 30% of the cored trees
+#Find n trees that have resistance <= 0.7 per year: #% in 1966, #2% in 1977, and #% in 1999. #Do same thing for those that have resistance >1 for each year
+library(data.table)
+test <- as.data.table(trees_all_sub)
+
+test_reduct <- test[resist.value<=0.7,.(reduct = length(resist.value)), by=.(year)]
+by_year <- test[, .(total=length(resist.value)), by=.(year)]
+test_reduct$total_re <- by_year$total
+test_reduct$perc_re <- test_reduct$reduct/test_reduct$total_re
+
+test_growth <- test[resist.value>1,.(growth = length(resist.value)), by=.(year)]
+test_growth$total_gr <- by_year$total
+test_growth$perc_gr <- test_growth$growth/test_growth$total_gr
+
+test_full <- merge(test_reduct, test_growth)
+
 #save as png so that way you're arranging only png
 png("manuscript/tables_figures/density_plot.png", res = 300, width = 150, height = 50, units = "mm", pointsize = 10)
 ggplot(trees_all_sub) +

### 2019-11-01 manuscript/climate_sensitivity_SI.Rmd
Added tableS5 and renamed file

@@
 *how do we want to present Table S4? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
-```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+```{r Table S4, eval=TRUE, echo=FALSE, message=FALSE}
 tableS4 <- read.csv("tables_figures/tableS4_top_models_dAIC.csv", stringsAsFactors = FALSE)
@@
 
+```{r Table S5, eval=TRUE, echo=FALSE, message=FALSE}
+tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)
+
+tableS5 <- tableS5[,-1]
+names(tableS5) <- c("variable", "model", "coefficient", "p-value")
+tableS5[,c(3)] <- round(tableS5[,c(3)], 2)
+
+tableS5$variable <- c("WD", "LMA", "ring porosity", "PLA", "TLP")
+tableS5$model <- c("WD~ln[H]","LMA~ln[H]","ring porosity~ln[H]", "PLA~ln[H]", "PLA~ln[H]")
+
+kable(tableS5, booktabs=TRUE, format = "latex", caption = "Importance of height compared to each trait") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+\newpage
 ![Figure S2: Location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
 
-\newpage
 ![Figure S3: Traits with TWI](tables_figures/figureS3_traits_with_twi.png)
-(see Issue #32)

### 2019-11-01 scripts/canopy_position_analysis.R

@@ -1761,7 +1761,7 @@ for(i in seq(along=trait)){
   trait_ht[,5][[i]] <- lm_anova$`Pr(>F)`[1]
 }
 
-write.csv(trait_ht, "manuscript/tables_figures/tested_traits_height.csv", row.names=FALSE)
+write.csv(trait_ht, "manuscript/tables_figures/tableS5_tested_traits_height.csv", row.names=FALSE)
 
 ##6d. original table with models based on github issue predictions ####
 summary_models <- data.frame(

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Updated draft

@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**Ian, confirm**), are available through the ForestGEO data portal (www.forestgeo.si.edu).
+Census data, which were last updated in 2018 (**the main census data was taken in 2018, but otherwise we had the final update from Suzanne in 2019.
+Did you want 2019 here?**), are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
@@
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
-TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on run-off tendencies of the plot.
-*Based on our observations in the plot* (**Ian, is this what you mean?**), the calculation of TWI here performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
+TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on areas where run-off is more likely.
+From our observations in the plot, the calculation of TWI performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
 
@@
 Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using an R script (**details**).
+Leaf area was calculated from scanned images using an R script (**details - Valentine?
+Nobby?**).
 $LMA$ was calculated as the ratio of leaf dry mass to fresh area.
@@
 To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_nodate].
+We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018].
 After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
@@
 Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model (*lme4* package in R: [@R-lme4]) with $R$ as the response variable, tree *nested within species* (**IAN, CONFIRM**) as a random effect, and one or more independent variables as fixed effects.
+The general statistical model for hypothesis testing was a mixed effects model (*lme4* package in R: [@R-lme4]) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
 We used AICc (*AICcmodavg* package: [@R-AICcmodavg]) to assess model fit.
@@
 We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (**SI Table?**).
+In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5).
 However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$.

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
update results related to CP

@@
 Crown position was generally correlated with $H$, but with substantial variation (Fig.
-2**e**).
+2d).
 Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-When considered alone, $CP$ had a significant influence only in the 1977 drought, during which trees with dominant $CP$ had the lowest $R$.
-When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed trees had the lowest $R$.
-Crown position was included in **roughly half** of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
+When considered alone, $CP$ had a significant influence only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$.
+When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$.
+Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
 Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
-In all plots, the dahsed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
 
@@
 
-We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.23; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
+We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
+Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S3).
 In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (n=# trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5).
-However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the case of $LMA$, which did not correlate with $R$.
+In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5).
+However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$.
 Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
-2**e-f**).
+2e-f).
 
-We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is *not* driven by more drought-sensitive traits in larger trees, but rather by size itself (Tables 1,5).
+We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
 As discussed above, there was little meaningful variation in traits with height at the community level.

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 These differences may very well be random, as opposed to statistically meaningful.
-Among the univariate models, there was no instance predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
+Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
 Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4).

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
-The tendencey for larger trees to suffer more, observed here as in forests around the world (**REFS**), was driven primarily by their height.
-There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??) and greater sensitivity of suppressed and crowded individuals (**REFS- **).
+The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height.
+There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**).
 There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
 The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
working on references

@@
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical (**REFS**[@kannenberg_linking_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
 The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
@@
 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 10-12% for all droughts combined, and 21-27% for each individual drought (Table 5).
-Methodologically, the moderate nature of these drougths is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
+It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5).
+Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent-- traits (**REFS**).
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (*Greenwood et al.
+2007- doi: 10.1111/ele.12748; Anderegg et al.
+2016- doi: 10.1073/pnas.1525678113*).
 Nevertheless, our results are consistent with findings from more extreme droughts.
@@
 Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height.
-To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (**REFS**).
+To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (*Ryan et al.
+2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al.
+2011](https://www.springer.com/gp/book/9789400712416)*).
 They also face different microenvironments, which are partially related to canopy position.
@@
 1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
-**The vegetation reaches higher temperatures[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)**, which would occur more during a drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime CO_2 concentrations tend to be lower in the canopy than in the understory **(Fig.
-2[??](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/58)** [@koike_leaf_2001; MORE REFS], implying greater water costs of $CO_2$ uptake.
-**[McDowell-vertical profiles in isotopes?*]**.
+Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake.
 Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
-2**e?**) [@muller-landau_testing_2006].
+2d) [@muller-landau_testing_2006].
 We show that height is a far stronger predictor than crown position (Tables 1,4,5).
 Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
-However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of (*H1.2*), implying that dominant crown positions did indeed reduce $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]; MORE??).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer distproportionately during drought (**REFS**).
+However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*.
+The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**).
 The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
worked on discsussion/ refs

@@
 2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al.
-2011](https://www.springer.com/gp/book/9789400712416)*).
+2011](https://www.springer.com/gp/book/9789400712416);*)[@couvreur_water_2018.
 They also face different microenvironments, which are partially related to canopy position.
@@
 Beyond microenvironment, multiple tree characteristics covary with height.
-To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including crown volume, rooting volume, sapwood area, transpiration ...
-(e.g., [@andersonteixeira_size-related_2015; **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)].
+To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015;  **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)].
 These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al.
-2018; others?**), and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
+2018; others?**)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
 However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
-Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves ...
-(**describe differences**) [@couvreur_water_2018; @koike_leaf_2001].
+Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001]*[McDowell et al.
+2011](https://www.springer.com/gp/book/9789400712416)*.
 Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
@@
 As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been suffering disrpoportionately in forests around the world (*REFS*), and we here show that this is primarily driven by their height, with some contributions from canopy position.
+Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.

### 2019-11-01 manuscript/climate_sensitivity_SI.Rmd
worked on references/discussion, minor text changes

--- 2019-10-27 manuscript/climate_sensitivity_SI.Rmd
@@
 - \usepackage{caption}
+- \captionsetup[figure]{labelformat=empty}
+- \usepackage{caption}
+- \captionsetup[table]{font=small}
+- \captionsetup[figure]{font=small}
+- \captionsetup[table]{labelformat=empty}
 - \captionsetup[figure]{labelformat=empty}
@@
 
-```{r Figure S1, echo=FALSE, out.width='100%', fig.cap = "Figure S1: Map of ForestGEO plot", fig.pos='ht'}
-  knitr::include_graphics("tables_figures/figureS1_ForestGEO_plot.jpg")
+\newpage
+
+```{r Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)
+
+tableS5 <- tableS5[,-1]
+names(tableS5) <- c("variable", "model", "coefficient", "p-value")
+tableS5[,c(3)] <- round(tableS5[,c(3)], 2)
+
+tableS5$variable <- c("WD", "LMA", "ring porosity", "PLA", "TLP")
+tableS5$model <- c("WD~ln[H]","LMA~ln[H]","ring porosity~ln[H]", "PLA~ln[H]", "PLA~ln[H]")
+
+kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S4. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
 ```
@@
 \newpage
-*how do we want to present Table S4? Would it be better as an image of an excel file, since it's so large? Did we want to keep all coefficients here?*
-```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
-tableS4 <- read.csv("tables_figures/tableS4_top_models_dAIC.csv", stringsAsFactors = FALSE)
-kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Top model variations for each drought scenario, with dAICc values <= 2") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-![Figure S2: Location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
-
-\newpage
-![Figure S3: Traits with TWI](tables_figures/figureS3_traits_with_twi.png)
-(see Issue #32)
+![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
worked on references/discussion, minor text changes

@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2018 (**the main census data was taken in 2018, but otherwise we had the final update from Suzanne in 2019.
-Did you want 2019 here?**), are available through the ForestGEO data portal (www.forestgeo.si.edu).
+Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
@@
 
-We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019].
+We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
 Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
@@
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R.
+Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R (Figure S1).
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
@@
 We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5).
+In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).
 However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$.
@@
 Beyond microenvironment, multiple tree characteristics covary with height.
-To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015;  **REFS--KAT: Sperry scaling model; roots reference; BAAD database?**)].
-These work in combination to shape whole-plant hydraulic function (**REFS-KAT: Sperry scaling model, check McDowell scaling paper, Couvreur et al.
-2018; others?**)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
+To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015]; *Chen et al.
+2018: DOI: 10.1111/ele.13193.*).
+These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?** *Sperry et al.
+2012 (doi: 10.1111/j.1365-2435.2012.02022.x)*)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
 However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
@@
 The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
-Concerted measurement of leaf hydraulic traits of emerging importance (**REFS--KAT/NOBBY/LAWREN**) allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients (**REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019](**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
 (**REFS**).
-It is scientifically exciting in that it indicates that these traits, which can be measured relatively easily $PLA_{dry}$ and $\pi_{tlp}$ (**REFS--KAT/NOBBY/LAWREN**), hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study; **REFS**), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS--KAT/NOBBY/LAWREN**).
+It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
+The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
 
@@
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (Bretfield et al.), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
+Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (*Bretfield et al.*), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
 The later successional species at our site (*Fagus grandifolia*, ...

### 2019-11-01 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (*Bretfield et al.*), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
-The later successional species at our site (*Fagus grandifolia*, ...
-) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
+Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
 In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

### 2019-11-02 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 1).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**).
+We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the top ten driest of the study period in terms of May-August [PET-PRE] or PDSI.
 Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
@@
 The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.*
+
 **The droughts differed in timing/duration/etc.

### 2019-11-02 manuscript/climate_sensitivity_SI.Rmd
added Table S3- drought stats

@@
 ```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Candidate variables for best model") %>%
+tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other years referenced in the manuscript") %>%
+  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("others", 13, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 17, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 21, 24, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 
-```{r Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+tableS4 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
+kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Candidate variables for best model") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+\newpage
+
+```{r Table S5, eval=TRUE, echo=FALSE, message=FALSE}
 tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)

### 2019-11-02 manuscript/climate_sensitivity_draft.Rmd
added Table S3- drought stats

@@
 Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R.
-All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S3).
+All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S4).
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
@@
 1).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the top ten driest of the study period in terms of May-August [PET-PRE] or PDSI.
+We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
 Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
@@
 Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.*
+The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were not identified as a pointer yeasr (Table S3)
 
@@
 We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S3).
+Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S4).
 In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
@@
 We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).
+In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5).
 However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$.

### 2019-11-02 manuscript/climate_sensitivity_draft.Rmd
Added references, fixed typos, added correct formatting to have r-packages be cited

@@
     fig_caption: yes
-bibliography: [book.bib,  packages.bib]
-biblio-style: apalike
+bibliography: ["book.bib",  "packages.bib"]
+biblio-style: "apalike"
 csl: apa.csl
@@
 A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.
-Xylem architecture plays a role, with diffuse porous species tending to be more drought sensitive than ring-porous species [@kannenberg_linking_2019] *(Elliot et al.
-2015-doi: 10.1111/gcb.13045, Friedrichs et al.
-2009-doi:10.1007/s00468-009-0315-2)*, but does not differentiate species beyond broad classes.
-Commonly measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009]*(Hoffmann et al.
-2011- doi: 10.1111/j.1365-2486.2011.02401.x)* and other forest biomes around the world (*Greenwood et al.
-2007- doi: 10.1111/ele.12748*).
-However, the direction of response is not always consistent; wood density correlated negatively with drought tree drought performance in a broadleaf deciduous forest in the southeastern United States *(Hoffmann et al.
-2011)* but positively or with no effect at a global scale (*Greenwood et al.
-2007- doi: 10.1111/ele.12748; Anderegg et al.
-2016- doi: 10.1073/pnas.1525678113*).
-Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance (*Anderegg et al.
-2016*) but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses highly diverse forests (*e.g.*, in the tropics).
-More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016], are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
-The ability of these hydraulic traits to tree performance under drought remains untested.
+Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes.
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
+However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018).
+Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
+The ability of these hydraulic traits to explain tree performance under drought remains untested.
 
@@
 While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019] *Anderegg et al.
-2016*).
+The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
 Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
-While the tendency for larger trees to suffer more definitely predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
+While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
 There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
@@
 
-Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurments, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenrivonment, and species' traits collectively shape drought responses.
-First, we focus on the role of tree size and its interaction with microenvironement. 
+Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses.
+First, we focus on the role of tree size and its interaction with microenvironment. 
 We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; *Hoffmann et al.
-2011*) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
@@
 
-Topographic wetness index (TWI) was calculated using the [@R-dynatopmodel] package in R (Figure S1).
+Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel].
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
@@
 
-For each drought period, we calculated drought *resistance* ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011].
+For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011].
 Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model (*lme4* package in R: [@R-lme4]) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-We used AICc (*AICcmodavg* package: [@R-AICcmodavg]) to assess model fit.
+The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model fit.
 Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
-We first tested the predictor variables idependently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
+We first tested the predictor variables independently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
 Because height had a substantial influence on $R$, we included it in the null model for testing other variables (Table 4).
 
-Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R.
-All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S4).
+Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R$.
+All variables with dAICc > 2.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S4).
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
@@
 Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (*Greenwood et al.
-2007- doi: 10.1111/ele.12748; Anderegg et al.
-2016- doi: 10.1073/pnas.1525678113*).
+2007- doi: @greenwood_tree_2017; @anderegg_meta-analysis_2016).
 Nevertheless, our results are consistent with findings from more extreme droughts.
@@
 Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height.
-To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction (*Ryan et al.
-2006-  DOI:10.1111/j.1365-3040.2005.01478.x; McDowell & Allen 2015: DOI: 10.1038/NCLIMATE2641; [McDowell et al.
-2011](https://www.springer.com/gp/book/9789400712416);*)[@couvreur_water_2018.
+To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
 They also face different microenvironments, which are partially related to canopy position.
@@
 Beyond microenvironment, multiple tree characteristics covary with height.
-To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., [@andersonteixeira_size-related_2015]; *Chen et al.
-2018: DOI: 10.1111/ele.13193.*).
-These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?** *Sperry et al.
-2012 (doi: 10.1111/j.1365-2435.2012.02022.x)*)[@couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
+To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., @andersonteixeira_size-related_2015; @chen_allometry_2019).
+These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?**; @sperry_species-level_2012; @couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
 However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
-Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001]*[McDowell et al.
-2011](https://www.springer.com/gp/book/9789400712416)*.
+Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011].
 Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
@@
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019](**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
 (**REFS**).

### 2019-11-03 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
issue #60

@@
 As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
+*While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
 2) and height change since the beginning of the study period indicated that change was likely slow.
-Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, and ~1m from 1977 to 1999.
+Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, and ~1m from 1977 to 1999.* (see issue [#60](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/60)).
 
@@
 Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
-However, rates of change are relatively slow, and the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*.
+However, the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*.
 The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).

### 2019-11-04 manuscript/climate_sensitivity_draft.Rmd
worked on discussion related to tree height

@@
 This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006], and this may become a greater liability during drought [@zhang_size-dependent_2009].
+First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009].
 Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
@@
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (*Greenwood et al.
-2007- doi: @greenwood_tree_2017; @anderegg_meta-analysis_2016).
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
 Nevertheless, our results are consistent with findings from more extreme droughts.
 
-Our analysis indicates that height is the primary factor through which tree size mediates drought response; however, specific identification of the primary physiological mechanism(s) is complicated by the fact that multiple drought-relevant environmental parameters and tree characteristics covary with height.
-To begin with, taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-They also face different microenvironments, which are partially related to canopy position.
-Even under non-drought conditions, evaporative demand increases with tree height, which are more closely coupled to the atmosphere  (Fig.
-1a-b; **REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
+Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response.
+Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
+Vertical gradients in stem and leaf hydraulic traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges.
+Taller trees also face different microenvironments (Fig.
+2a-b), in part because they are more likely to be in dominant canopy positions (Fig.
+2d).
+Even under non-drought conditions, evaporative demand increases with tree height, and tall trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
 Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime CO_2_ concentrations tend to be lower in the canopy than in the understory [@koike_leaf_2001], implying greater water costs of $CO_2$ uptake.
-Taller trees are more likely to be in dominant canopy positions, although signficant decoupling can be introduced by the configuration of neighboring trees (Fig.
-2d) [@muller-landau_testing_2006].
-We show that height is a far stronger predictor than crown position (Tables 1,4,5).
+Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
+Correlation between height and canopy position (Fig.
+2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment.
+However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig.
+2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5).
 Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
-However, the fact that some trees inevitably did grow towards more dominant positions would bias against the acceptance of *H1.2*.
+However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*.
 The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
@@
 
-Beyond microenvironment, multiple tree characteristics covary with height.
-To begin with, multiple tree characteristics that may influence drought resistance scale with $DBH$--or, similarly, $H$--including leaf area and crown volume, fine root biomass and rooting volume, sapwood area, and transpiration (e.g., @andersonteixeira_size-related_2015; @chen_allometry_2019).
-These work in combination to shape whole-plant hydraulic function (**REFS-KAT: check McDowell scaling paper,others?**; @sperry_species-level_2012; @couvreur_water_2018], and we cannot expect to statistically identify a single driver that fully encapsulates size-related variation in trees.
-However, height is arguably the most biophysically meaningful tree-size variable shaping drought response.
-Beyond whole-plant characteristics, leaf traits vary with height on the tree--sometimes quite dramatically, with upper-canopy leaves tending to be smaller and thicker (higher LMA), with more negative $P50$ and lower hydraulic conductivity [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011].
-Our analysis focuses on species-level comparisons and does not characterize the role of variation with height.
-On an individual level, average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits--assuming trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height.
-In not accounting for this, our analysis is most likely biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
-Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses [@couvreur_water_2018].
-
 The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ are useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
 (**REFS**).
@@
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
+
+Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height.
+As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits.
+If trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
+Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
 

### 2019-11-04 manuscript/climate_sensitivity_SI.Rmd
Updated caption to reflect Table S5
created and added Fig. S2- drought plot

@@
 
-kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S4. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
+kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S5. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 ![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
+
+\newpage
+![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/drought_plot.jpg){width=600px}

### 2019-11-04 manuscript/climate_sensitivity_draft.Rmd
created and added Fig. S2- drought plot

@@
 
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig.
-1).
+In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Figs.
+1, S2).
 We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).

### 2019-11-04 scripts/drought_plot.m

@@ -1,0 +1,43 @@
+%script to plot drought conditions
+%by Kristina Anderson-Teixeira
+%~~~~~~~~
+clf; clear; close all;
+
+cd '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/data/climate'
+NOAA_table = readtable('Formated_NOAA_PDSI_Northern_Virginia_1895_2017.csv');
+NOAA_table2 = readmatrix('Formated_NOAA_PDSI_Northern_Virginia_1895_2017.csv');
+year=NOAA_table2(:,1);
+month=NOAA_table2(:,2);
+date=datenum([year month 1+0*month]);
+PDSI=NOAA_table2(:,5);
+
+dr=[1966 1977 1999];
+colors= [ 1 .4 .7; .2 .7 .2 ; .4 .4 1];
+for d=1:3
+    fyr=dr(d);
+    index=find(year>(fyr-3) & year<(fyr+1));
+    date_rel=date(index)-datenum([fyr 1 1]).* (1+0*(index));
+    plot(date_rel,PDSI(index), 'LineWidth',2, 'Color', colors(d,:)); hold on;
+end
+
+xlim([-729 365]);
+set(gca, 'XTick', (-729 : 365/4 : 365 ));
+datetick('x','mmm','keepticks')
+
+xline(121,'-r',{'peak growing season','of drought year'},'LineWidth',4);
+xline(243,'-r','LineWidth',4);
+xline(0,'--');
+xline(-365,'--');
+xline(-730,'--');
+
+xlabel('month')
+ylabel('PDSI')
+legend({'1966' '1977' '1999'},'Location','Southwest')
+legend('Boxoff')
+
+
+shg
+
+cd '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/manuscript/tables_figures'
+print('drought_plot', '-djpeg')
+

### 2019-11-04 scripts/drought_plot.m
adjusted plotting
adjusted to show monthly value on day 15, as opposed to day 1, of each mont

@@ -8,7 +8,7 @@ NOAA_table = readtable('Formated_NOAA_PDSI_Northern_Virginia_1895_2017.csv');
 NOAA_table2 = readmatrix('Formated_NOAA_PDSI_Northern_Virginia_1895_2017.csv');
 year=NOAA_table2(:,1);
 month=NOAA_table2(:,2);
-date=datenum([year month 1+0*month]);
+date=datenum([year month 15+0*month]);
 PDSI=NOAA_table2(:,5);
 
 dr=[1966 1977 1999];

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
update draft with droughts comparison

@@
 Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**).
-Larger trees tend to suffer more (e.g., [@bennett_larger_2015]; [@stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
+Larger trees tend to suffer more (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
 As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
@@
 
-**The droughts differed in timing/duration/etc.
-..
-The year 1966 was preceded by two relatively dry years...
-1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry.
-**
+The droughts differed in intensity and prior onset (Table S3, Fig.
+S2).
+In terms of intensity during the peak growing season of the focal year, the 1999 drought was the most intense (lowest PDSI) during May-July.
+It was surpassed in intensity in August by the 1966 drought, which was otherwise the second most intense during the peak growing season.
+The 1977 was the least intense throughout the growing season.
+All droughts increased in intensity from May through August, but differed in the months previous.
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
+Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
+The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
 
-Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig.
+Community-level tree growth responses to these droughts were modest, with modal resistance values of **#, #, and #** for 1966, 1977, and 1999, respectively (Fig.
 1b).
 In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
-*Some* trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
+Some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
 
@@
 As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits.
-If trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
+If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
 Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
@@
 Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-[**BRIEFLY DISCUSS DIFFERENCES ACROSS DROUGHTS-- For example, 1966 was preceded by two dry years, and height had the strongest effect in this drought, consistent with the finding that height becomes a stronger predictor of mortality as drought duration and intensity increase** [@stovall_tree_2019].
-...] Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites would be of great value to elucidating the mechanisms through which these shape drought resistance (**check for refs**).
-Unfortunately, our ability to compare forest responses across droughts is hampered by the fact that droughts are rarely explicitly defined in ecological studies [@slette_how_2019].
-Further research will be required to understand how drought characteristics interact with driver variables to shape tree growth responses.
+Although there were not signficant differences in $R$ among the drought years, $R$ tended to be somewhat lower in 1966 (Fig.
+1, Table 4).
+Height had the strongest negative effect in this drought (Tables 4-5), consistent with the finding that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
+In 1977, which was the weakest drought (Fig.
+S3, Table S2), the only factor useful for predicting $R$ in the univariate models was microhabitat moisture ($TWI$, Table 4).
+Somewhat surprisingly, but consistent with the weakening drought sensitivity observed by [@helcoski_growing_2019], the highest overall $R$ was observed during the 1999 drought, which was the most intense and second-longest in duration.
+For this drought, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.
+Thus, responses differed with the nature of each drought.
+Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.
 

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
made Table 2 render correctly (but with "TLP" instead of symbol)

@@
 
-\noindent\hrulefill 
  ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
@@
   table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-
-kable(table2, booktabs=TRUE)
-
-```
-
-kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  kable(table2, booktabs = TRUE, caption = "**Table 2x. Summary of variables**") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
   add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
@@
 kable_styling()
+```
 

### 2019-11-05 manuscript/climate_sensitivity_SI.Rmd
removed table s4 (not necessary)

@@
 
-```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
-tableS4 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
-kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Candidate variables for best model") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-\newpage
-
 ```{r Table S5, eval=TRUE, echo=FALSE, message=FALSE}
@@
 
-kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S5. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
+kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S4. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
removed table s4 (not necessary)

@@
 ### Title page
+
+**Title:**
+
+**Authors:**
 
@@
 Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R$.
-All variables with dAICc > 2.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S4).
+All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model.
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
@@
 We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S4).
+Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
 In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
@@
 1, Table 4).
-Height had the strongest negative effect in this drought (Tables 4-5), consistent with the finding that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
+Height and dominant canopy position had the stronger negative effects in this drought than in the others (Tables 4-5), consistent with the finding that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
 In 1977, which was the weakest drought (Fig.

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
fixing up Table 1
when possible, write out variable names so that table makes sense stand-alone (and to avoid symbol formatting problems!)

issue #60

@@
 Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
-  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.0. Larger-diameter trees have lower drought resistance (R).", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   column_spec(2:5, width = "1cm") %>%
-  column_spec(6, width = "2cm") %>%
+  column_spec(6, width = "2.4cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%

### 2019-11-05 manuscript/tables&figures.Rmd
created seperate document with tables & figures

--- 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
@@
 ```
-\raggedright
-
-### Title page
-
-**Title:**
-
-**Authors:**
-
-text | word count | other | n
----------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	## (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
-Materials and Methods |	currently ~1442	||	 
-Results |	currently ~1437	 ||
-Discussion |	currently ~1881 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
-Acknowledgements | 	~75 unwritten	||
-
-\newpage
-
-### Summary (**needs work**)
-- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important.
-Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.
-
-- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60 year period from 1950 and 2009.
-
-- Individual-level growth responses to the three individual droughts were stronger in three cases: taller trees, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
-However, there was substantial variation in the best predictor variables across given droughts.
-
-- We conclude that when droughts occur, tall trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
-
-*The Summary for research papers, which must be usable as a stand- alone document, must not exceed 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion.
-It should not contain citations of other papers.*
-
-*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.
-Very short phrases and scientific names with their common equivalents (e.g.
-Nicotiana tabacum (tobacco)) are acceptable.*
-
-\newpage
-### Introduction
-Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
-An important aspect of this uncertainty lies in responses to drought (**REF**).
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**).
-Larger trees tend to suffer more (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
-As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
-Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
-
-One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
-@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**).
-However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009].
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
-Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
-It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience.
-Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
-Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.
-
-A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.
-Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes.
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018).
-Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
-The ability of these hydraulic traits to explain tree performance under drought remains untested.
-
-A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.
-No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
-However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
-Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
-While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
-There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
-[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
-
-Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses.
-First, we focus on the role of tree size and its interaction with microenvironment. 
-We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
-We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
-Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
-We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
-Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).
 
@@
 ```
-
 \clearpage
-
-### Materials and Methods
-*Study site*
-
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
-SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
-Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
-
-*Data collection and preparation*
-
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
@@
 ```
-
-We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
-Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013].
-In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
-Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-*Ryan submitted the data to ITRDB but I don't think its posted yet.
-We should also cite GitHub/Zenodo here.
-I'll come back to that.*
-
-For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
-Prior DBH was estimated using the following equation:
-
-$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$$
-
-Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
-Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
-
-Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
-Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
-Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
-Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
-Species-specific height allometries were developed (Table S2).
-For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
-
-Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-*While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2) and height change since the beginning of the study period indicated that change was likely slow.
-Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, and ~1m from 1977 to 1999.* (see issue [#60](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/60)).
-
-Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel].
-Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
-TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on areas where run-off is more likely.
-From our observations in the plot, the calculation of TWI performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
-
-Hydraulic traits were collected at SCBI (Table 3).
-In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
-These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis.
-Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using an R script (**details - Valentine?
-Nobby?**).
-$LMA$ was calculated as the ratio of leaf dry mass to fresh area.
-$PLA$ was calculated as the percent loss of area between fresh and dry leaves.
-$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-We used the rapid determination method of @bartlett_rapid_2012 to estimate the turgor loss point ($\pi_{tlp}$).
-Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
-Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
-We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
-Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
+\clearpage
 
@@
 
-To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018].
-After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
-
-*Identification of drought years*
-
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
-Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R (version 3.5.3).
-Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site.
-We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE).
-These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
-The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
-
-*Analysis*
-
-For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011].
-Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model fit.
-Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
-We first tested the predictor variables independently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
-Because height had a substantial influence on $R$, we included it in the null model for testing other variables (Table 4).
-
-Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R$.
-All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model.
-We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
-These were counted as the "top multivariate models".
-When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models.
-If the variable appeared in only some of the models, we considered this partial support.
-
-### Results
-
-*Focal droughts and their characteristics*
-
-![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
-
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Figs.
-1, S2).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
-Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(*Cite Shenandoah paper, if accepted*).
-Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
-Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were not identified as a pointer yeasr (Table S3)
-
-The droughts differed in intensity and prior onset (Table S3, Fig.
-S2).
-In terms of intensity during the peak growing season of the focal year, the 1999 drought was the most intense (lowest PDSI) during May-July.
-It was surpassed in intensity in August by the 1966 drought, which was otherwise the second most intense during the peak growing season.
-The 1977 was the least intense throughout the growing season.
-All droughts increased in intensity from May through August, but differed in the months previous.
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
-Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
-The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
-
-Community-level tree growth responses to these droughts were modest, with modal resistance values of **#, #, and #** for 1966, 1977, and 1999, respectively (Fig.
-1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
-Some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
-
-*Tree size and drought resistance*
-
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4).
-When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
-
+\clearpage
 ```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
+\clearpage
 ```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-\raggedright
-
-Crown position was generally correlated with $H$, but with substantial variation (Fig.
-2d).
-Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-When considered alone, $CP$ had a significant influence only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$.
-When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$.
-Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
-Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
-Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.
-
-In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-2).
-Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-2a).
-Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig.
-2b).
-Air temperature did not vary across the vertical profile (Fig.
-2c).
+![**Figure 1.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from [@helcoski_growing_2019].
+Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
 
@@
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
-
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
-*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
-
-*Species' traits, height, and drought resistance*
-
-We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts.
-$\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4).
-Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
-Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.
-
-We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).
-However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$.
-Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
-2e-f).
-
-We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
-As discussed above, there was little meaningful variation in traits with height at the community level.
-When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $ln[H]$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
-
-*Responses across droughts*
-
-We reject the hypothesis (*H3.1*) that overall community responses varied across droughts.
-Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
-This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig.
-1b).
-
-We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts.
-In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
-However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $\pi_{tlp}$ in the multivariate model for the 1966 drought; Tables 4, 5).
-These differences may very well be random, as opposed to statistically meaningful.
-Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
-Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-The difference most likely to be real is that ring and semi-ring porous species had lower resistance than diffuse porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
-
-We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
-For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $CP$ (lowest $R$ among supressed trees) and $XP$ (lowest $R$ among diffuse-porous trees) were strongest in 1999 (Tables 4,5).
-
-### Discussion
-
-Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
-The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height.
-There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**).
-There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
-The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
-The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
-These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.
-
-The droughts considered here were of similar severity (Fig.
-1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
-1a, @helcoski_growing_2019).
-Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature.
-While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
-1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5).
-Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
-These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
-Nevertheless, our results are consistent with findings from more extreme droughts.
-
-Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response.
-Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf hydraulic traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges.
-Taller trees also face different microenvironments (Fig.
-2a-b), in part because they are more likely to be in dominant canopy positions (Fig.
-2d).
-Even under non-drought conditions, evaporative demand increases with tree height, and tall trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
-Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
-Correlation between height and canopy position (Fig.
-2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment.
-However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig.
-2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5).
-Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
-However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*.
-The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**).
-The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
-Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
-
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
-Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
-(**REFS**).
-It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
-
-Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height.
-As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits.
-If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
-Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
-
-Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-Although there were not signficant differences in $R$ among the drought years, $R$ tended to be somewhat lower in 1966 (Fig.
-1, Table 4).
-Height and dominant canopy position had the stronger negative effects in this drought than in the others (Tables 4-5), consistent with the finding that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
-In 1977, which was the weakest drought (Fig.
-S3, Table S2), the only factor useful for predicting $R$ in the univariate models was microhabitat moisture ($TWI$, Table 4).
-Somewhat surprisingly, but consistent with the weakening drought sensitivity observed by [@helcoski_growing_2019], the highest overall $R$ was observed during the 1999 drought, which was the most intense and second-longest in duration.
-For this drought, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.
-Thus, responses differed with the nature of each drought.
-Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.
-
-As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position.
-The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (*Bretfield et al.*), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
-Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
-In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
-
-### Acknowledgements
-funding: ForestGEO,
-
-### Author Contribution 
-words
-
-\newpage

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
worked on table 3
removed tables & figures from main document

--- 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
@@
 
-```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and all top multivariate models (when applicable).
-Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
-  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. At the community level, taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.3. When traits are accounted for, taller trees still have lower drought resistance.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Resistance differs across the droughts considered here.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(1, width = "12cm") %>%
-  column_spec(2:5, width = "1cm") %>%
-  column_spec(6, width = "2cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kableExtra::landscape()
-```
-
-\clearpage
-
 ### Materials and Methods
@@
 Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).
-
- ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
-  library(knitr)
-  library(kableExtra)
-  table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-  kable(table2, booktabs = TRUE, caption = "**Table 2x. Summary of variables**") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
-  column_spec(4, width = "7cm") %>%
-  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-kable_styling()
-```
 
@@
 
-```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table3 <- read.csv("tables_figures/table3_species_table.csv")
-
-kable(table3, booktabs=TRUE, caption = "**Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
-```
-
 To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area.
@@
 
-![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
-
 In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Figs.
@@
 
-```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(univar_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
-  kable_styling()
-```
-
-```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(multivar_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
-  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 12, 14, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 15, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  kable_styling()
-```
-
-\raggedright
-
 Crown position was generally correlated with $H$, but with substantial variation (Fig.
@@
 2c).
-
-![**Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown are (d) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (e) $PLA_{dry}$ and (f) $\pi_{tlp}$.
-In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
 

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
1.5 line spacing

@@
 header-includes:
+- \usepackage{setspace}\onehalfspacing
 - \usepackage{float}

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
update word count

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  currently ~6000 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
+Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
 Summary|  	## (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1536 |	No of Supporting Information files | 	###.
-Materials and Methods |	currently ~1442	||	 
-Results |	currently ~1437	 ||
-Discussion |	currently ~1881 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
+Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
+Materials and Methods |	currently ~1604	||	 
+Results |	currently ~1558	 ||
+Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
 Acknowledgements | 	~75 unwritten	||

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
redid title page, drafted acknowledgements & author contributions

@@
 ---
-title: "Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
+title: ""
+author: ""
 output:
@@
 
-### Title page
-
-**Title:**
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
 
 **Authors:** 
+Ian McGregor^1,2^
+Ryan Helcoski^1^
+Norbert Kunert^1,3^
+Alan Tepley^1,4^
+Erika Gonzalez-Akre^1^
+Valentine Herrmann^1^
+Joseph Zailaa^1,5^
+Atticus Stovall^1,6^
+Norman Bourg^1^?
+William McShea^1^?
+Neil Pederson^7^
+Lawren Sack^8^
+Kristina J. Anderson-Teixeira^1,3^
+
+**Author Affiliations:**
+
+1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
+2.
+3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+4.
+5.
+6.
+7.
+8.
 
@@
 Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
-Acknowledgements | 	~75 unwritten	||
+Acknowledgements | currently ~73	||
 
@@
 ### Acknowledgements
-funding: ForestGEO,
+Thanks to all researchers who helped to collect the data used here, in particular Jennifer C.
+McGarvey, Jonathan R.
+Thomspon, and Victoria Meakem for original collection and processing of cores.
+** others??** This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper.
+accurate here?** **coauthors, other funding?**)
 
 ### Author Contribution 
-words
+KAT and IM designed the research.
+Tree-ring chronologies were developed by RH under guidance of AT and NP.
+Trait data was collected by IM, JZ under guidance of NK and LS.
+Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM.
+Data analyses were performed by IM under guidance of KAT and VH.
+KAT and IM interpreted the results.
+IM and KAT wrote the first draft of manuscript, and all authors (RH, KAT, AT, NP, JM, VM, JRT and VH) contributed to revisions.
 

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
worked on abstract

@@
 Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	## (limit 200) | 	No. of Tables | 	5  
+Summary|  	currently 198 (limit 200) | 	No. of Tables | 	5  
 Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
@@
 
-### Summary (**needs work**)
+### Summary
 - As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important.
@@
 
-- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60 year period from 1950 and 2009.
+- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60-year period from 1950 and 2009.
 
-- Individual-level growth responses to the three individual droughts were stronger in three cases: taller trees, trees in wetter microsites, and more drought-sensitive species as assessed by leaf traits (turgor loss at less negative leaf water potential, greater shrinkage with leaf dehydration).
-However, there was substantial variation in the best predictor variables across given droughts.
+- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance.
+*Resistance was lower among diffuse-porous species and those whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential.* However, there was substantial variation in the best predictor variables across given droughts.
 
-- We conclude that when droughts occur, tall trees, drought-sensitive species, and individuals in wetter microhabitats tend to be most strongly affected.
+- We conclude that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
 

### 2019-11-05 manuscript/climate_sensitivity_SI.Rmd
removed title pages on figures & SI

@@
 ---
-title: "Supplementary Information: \nTree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest"
-author: "Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Erika Gonzalez-Akre, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Norman Bourg?, William McShea?, Neil Pederson, Lawren Sack, Kristina Anderson-Teixeira"
+title: ""
+author: ""
 output:
@@
 \newpage
+\clearpage
 

### 2019-11-05 manuscript/tables&figures.Rmd
adjustments to Fig. 1 caption
removed title pages on figures & SI

@@
   table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-  kable(table2, booktabs = TRUE, caption = "**Table 2x. Summary of variables**") %>%
+  kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
@@
 
+\newpage
+
 ![**Figure 1.
@@
 Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from [@helcoski_growing_2019].
-Density plots (b) show community- wide resistance values for each drought.](tables_figures/Figure1.png){width=500px}
+Figure modified from Helcoski *et al.* (2019).
+Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}
 

### 2019-11-05 scripts/canopy_position_analysis.R
Update canopy_position_analysis.R

@@ -1656,7 +1656,6 @@ ord_lab <- c("trees_all", "x1966", "x1977", "x1999")
 
 coeff_list1 <- list()
 for(q in seq(along=ord_lab)){
-  
   coeff_list_temp <- coeff_list[grepl(ord_lab[[q]], names(coeff_list))]
   coeff_list_temp <- 
     coeff_list_temp[

### 2019-11-05 manuscript/tables&figures.Rmd

@@
   pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("species' traits", 12, 17, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()
@@
 ```
+*Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories.
 

### 2019-11-05 manuscript/tables&figures.Rmd
update table 2
update table 5

@@
 
-```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
+  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 2," ", " ")) %>%
   pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 12, 14, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 15, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 7, 9, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 10, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 16, 21, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
   kable_styling()

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
update summary and results

@@
 Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	currently 198 (limit 200) | 	No. of Tables | 	5  
+Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
 Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
@@
 - Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance.
-*Resistance was lower among diffuse-porous species and those whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential.* However, there was substantial variation in the best predictor variables across given droughts.
+Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential.
+However, there was substantial variation in the best predictor variables across given droughts.
 
@@
 
-*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
+*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest;  tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
 Words that are in the title can, and should, be among these.
@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring, semi-ring, or diffuse porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
@@
 In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts.
+Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts.
 $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4).
 Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
-Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.
+Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.
 
@@
 In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
-However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $\pi_{tlp}$ in the multivariate model for the 1966 drought; Tables 4, 5).
+However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5).
 These differences may very well be random, as opposed to statistically meaningful.
 Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
-Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $\pi_{tlp}$ only appeared with a positive coefficient in two of five models for the 1966 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-The difference most likely to be real is that ring and semi-ring porous species had lower resistance than diffuse porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
+Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest of six models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4).
+The difference most likely to be real is that ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
 
 We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
-For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $CP$ (lowest $R$ among supressed trees) and $XP$ (lowest $R$ among diffuse-porous trees) were strongest in 1999 (Tables 4,5).
+For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
 

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
update conclusions, knit

@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011) and specific leaf area (positive effect), but that hydraulic traits including xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), leaf area shrinkage upon dehydration, and turgor loss point will prove better predictors (*H2.1*).
+We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011), specific leaf area (positive effect), and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
@@
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought-sensitive species were not linked to $LMA$ and $WD$ but were predicted--at least some of the time--by hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$, and $XP$), which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
-The direction of these responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
+Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture.
+However, drought resistance was correlated with the hydraulic leaf traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
 These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.
@@
 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-27% for each individual drought (Table 5).
+It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
 Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
@@
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses--with both appearing in the top overall model and two of the three top models for each drought (Table 5)--is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
 (**REFS**).

### 2019-11-05 manuscript/climate_sensitivity_draft.Rmd
added site climate description
misc cleanup

@@
 **Authors:** 
-Ian McGregor^1,2^
+Ian R. McGregor^1,2^
 Ryan Helcoski^1^
 Norbert Kunert^1,3^
-Alan Tepley^1,4^
-Erika Gonzalez-Akre^1^
+Alan J. Tepley^1,4^
+Erika B. Gonzalez-Akre^1^
 Valentine Herrmann^1^
@@
 Atticus Stovall^1,6^
-Norman Bourg^1^?
-William McShea^1^?
+Norman A. Bourg^1^?
+William J. McShea^1^?
 Neil Pederson^7^
 Lawren Sack^8^
-Kristina J. Anderson-Teixeira^1,3^
+Kristina J. Anderson-Teixeira^1,3^*
 
@@
 
+*corresponding author: teixeirak@si.edu; +1 540 635 6546
+
+**coauthors, fill in middle initials and affiliations, if missing***
+
 text | word count | other | n
@@
 Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1339 |	No of Supporting Information files | 	###.
+Introduction|	currently ~1339 |	No of Supporting Information files | 	6
 Materials and Methods |	currently ~1604	||	 
@@
 
-- In this study, we analyze tree-ring records for 12 species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over the 60-year period from 1950 and 2009.
+- In this study, we analyze tree-ring records for twelve species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over a 60-year period (1950 - 2009).
 
 - Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance.
-Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative leaf water potential.
+Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative water potential.
 However, there was substantial variation in the best predictor variables across given droughts.
 
-- We conclude that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
-
-*The Summary for research papers, which must be usable as a stand- alone document, must not exceed 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion.
-It should not contain citations of other papers.*
+- We conclude that height--as opposed to canopy position, root water access, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
 
 *Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.
-Very short phrases and scientific names with their common equivalents (e.g.
-Nicotiana tabacum (tobacco)) are acceptable.*
+Words that are in the title can, and should, be among these.*
 
@@
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
-An important aspect of this uncertainty lies in responses to drought (**REF**).
+An important aspect of this uncertainty lies in responses to drought (*[Kennedy et al.
+2019](https://doi.org/10.1029/2018MS001500)*).
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
@@
 Larger trees tend to suffer more (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
-As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (**REFS**), yet accurate characterization of drought responses remains a modeling challenge (**REFS**)-- in part because some of the mechanisms underlying drought responses remain unclear.
+As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (*e.g., McDowell et al.
+2016(doi:10.1038/nclimate2873); Charney et al.
+2016 (doi: 10.1111/ele.12650)*), yet accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al.
+2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear.
 Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
@@
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
-@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding [@stovall_tree_2019; @hacket-pain_consistent_2016] **REFS**).
+@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
 However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
@@
 First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009].
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower $CO_2$ concentrations (**REFS-KAT**).
+Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**).
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
@@
 Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
-While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (e.g., **REFS**).
+While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (*e.g.*, **REFS??**).
 There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
@@
 Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
+Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
 Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
@@
 The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-*Ryan submitted the data to ITRDB but I don't think its posted yet.
+**Ryan, have you heard back from ITRDB?
 We should also cite GitHub/Zenodo here.
-I'll come back to that.*
+**
 
@@
 Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
-These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
+These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*).
 Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
@@
 The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS**).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
 The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
@@
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses--with both appearing in the top overall model and two of the three top models for each drought (Table 5)--is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**), **MORE--KAT/NOBBY/LAWREN**...
-(**REFS**).
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses--with both appearing in the top overall model and two of the three top models for each drought (Table 5)--is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**).
 It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
@@
 In 1977, which was the weakest drought (Fig.
-S3, Table S2), the only factor useful for predicting $R$ in the univariate models was microhabitat moisture ($TWI$, Table 4).
+S3, Table S2), ring-porous species had lower $R$, contrasting with their higher $R$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
 Somewhat surprisingly, but consistent with the weakening drought sensitivity observed by [@helcoski_growing_2019], the highest overall $R$ was observed during the 1999 drought, which was the most intense and second-longest in duration.
@@
 
-As climate change drives increasing drought in many of the world's forests (*REFS-KAT*), the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
+As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
 Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests (*Bretfield et al.*), and species traits often shift as forests age, with early successional species tending to have lower wood densities and higher hydraulic conductivities that facilitate rapid growth (**REFS in Bretfield-beginning of discussion**).
-Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether old-growth forests are more vulnerable to drought, and to which types of drought.
+Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
+Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether and how drought tolerance changes as forests age.
 In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
@@
 ### Author Contribution 
-KAT and IM designed the research.
+KAT, IM, and AT designed the research.
 Tree-ring chronologies were developed by RH under guidance of AT and NP.
@@
 KAT and IM interpreted the results.
-IM and KAT wrote the first draft of manuscript, and all authors (RH, KAT, AT, NP, JM, VM, JRT and VH) contributed to revisions.
-
-\newpage
+IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
+**coauthors, please check/edit**

### 2019-11-05 manuscript/tables&figures.Rmd
misc cleanup

@@
 kable(multivar_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
+Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AICc, and we show all models whose AICc value falls within 2.0 (dAICc<2) of the best model (bold).", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
   pack_rows("", 16, 21, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  row_spec(1, bold = T) %>%
+  row_spec(7, bold = T) %>%
+  row_spec(10, bold = T) %>%
+  row_spec(16, bold = T) %>%
   kable_styling()

### 2019-11-07 manuscript/climate_sensitivity_draft.Rmd
added mean R values for drought, adjusted discussion of drought strength
finishes issue #60

@@
 
-Community-level tree growth responses to these droughts were modest, with modal resistance values of **#, #, and #** for 1966, 1977, and 1999, respectively (Fig.
+Community-level tree growth responses to these droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999, respectively (Fig.
 1b).
@@
 1, Table 4).
-Height and dominant canopy position had the stronger negative effects in this drought than in the others (Tables 4-5), consistent with the finding that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
-In 1977, which was the weakest drought (Fig.
-S3, Table S2), ring-porous species had lower $R$, contrasting with their higher $R$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
-Somewhat surprisingly, but consistent with the weakening drought sensitivity observed by [@helcoski_growing_2019], the highest overall $R$ was observed during the 1999 drought, which was the most intense and second-longest in duration.
-For this drought, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.
+Height and dominant canopy position had the stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the fact that this was the longest-lasting drought and and that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
+In 1977, which was the weakest drought in terms of PDSI (Fig.
+S2, Table S3), ring-porous species had lower $R$, contrasting with their higher $R$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
+In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.
 Thus, responses differed with the nature of each drought.

### 2019-11-07 manuscript/climate_sensitivity_draft.Rmd
update height change

@@
 As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-*While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2) and height change since the beginning of the study period indicated that change was likely slow.
-Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, and ~1m from 1977 to 1999.* (see issue [#60](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/60)).
+While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
+2) and height change since the beginning of the study period indicated that change was likely slow (Fig.
+S3).
+Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, ~1m from 1977 to 1999, and **# (IAN)** from 1999 to 2018.
 

### 2019-11-07 manuscript/climate_sensitivity_draft.Rmd
incorporated (most of) JZ's edits

@@
 4.
-5.
+5. Biological Sciences Department; California State University; Los Angeles, CA, USA
 6.
@@
 ### Summary
-- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought vulnerability and resistance in trees is increasingly important.
-Yet it remains unclear how tree size and species' traits interactively shape tree growth responses across droughts.
+- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
+Yet it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.
 
@@
 - Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance.
-Resistance was lower among species whose leaves shrunk more upon dehydration and lost turgor at less negative water potential.
+Resistance was lower among species whose leaves had higher percent loss of area upon dehydration and lost turgor at less negative water potential.
 However, there was substantial variation in the best predictor variables across given droughts.
@@
 2019](https://doi.org/10.1029/2018MS001500)*).
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], even in conjunction with increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Global change-type drought has been affecting forests worldwide [@allen_global_2010], and it is expected that future climate change-driven droughts will severely impact forests around the world [@allen_global_2010]; **REFS**).
-Larger trees tend to suffer more (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010]; **REFS**).
+During drought, larger trees generally suffer greater reductions in growth and increases in mortality (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
 As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (*e.g., McDowell et al.
 2016(doi:10.1038/nclimate2873); Charney et al.
-2016 (doi: 10.1111/ele.12650)*), yet accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al.
-2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear.
+2016 (doi: 10.1111/ele.12650)*).
+Yet, accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al.
+2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear (**REF**).
 Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
@@
 @bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
-However, this analysis quantified tree size based on DBH, which has no direct mechanistic meaning.
+However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic meaning.
 This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], and this may become a greater liability during drought [@zhang_size-dependent_2009].
+First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], which becomes a greater liability during drought [@zhang_size-dependent_2009].
 Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**).
 Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems may confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
 A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
-It is possible that the pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience.
+The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience.
 Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
@@
 Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured traits such as $WD$ and $LMA$ [@medeiros_extensive_2019].
-The ability of these hydraulic traits to explain tree performance under drought remains untested.
+More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the osmotic potential at turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
+The ability of $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
 
-A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether drought variability in factors such as severity, duration, and timing interact with tree size and traits such that different components of the community respond differently to different droughts.
+A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether variability in factors such as drought severity, duration, and timing interact with tree size and traits such that components of the community respond differently across droughts.
 No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
 However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across droughts [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
 The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
@@
 In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
-These were covered with opaque plastic bags, re-cut under water, and re-hydrated overnight before further analysis.
+Samples were re-cut under water by at least two notes and re-hydrated overnight in buckets covered with opaque plastic bags before measurements.
 Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using an R script (**details - Valentine?
-Nobby?**).
+Leaf area was calculated from scanned images using the Leafarea R package (**citation**).
 $LMA$ was calculated as the ratio of leaf dry mass to fresh area.
@@
 $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-We used the rapid determination method of @bartlett_rapid_2012 to estimate the turgor loss point ($\pi_{tlp}$).
+We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
-We also characterized hydraulic vulnerability curves for the # most productive species, but because the water potentials at which 50% and 80% of conductivity is lost, $P50$ and $P80$, did not come out as top predictors in preliminary analyses and their inclusion limited the set of species that could be included in the full analysis, these traits were dropped from further consideration.
 Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
 
-To characterize how environmental conditions vary with height, data were obtained from the a National Ecological Observation Network (NEON) tower located <1km from the study area.
+To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
 We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018].
@@
 Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts.
-$\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4).
+$\pi_{tlp}$ never came out as significant (dAIC $\le$ 2) in the univariate models but had a consistently negative coefficient (Table 4).
 Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
 Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.
+$PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).
 
@@
 In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).
-However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$.
+However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ and $LMA$, which were not useful predictors of $R$.
 Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
@@
 Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture.
-However, drought resistance was correlated with the hydraulic leaf traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
 The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
-These findings significantly advance our knowledge as to the factors that confer vulnerability or resilience on trees during drought.
+These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
 
@@
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
+Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
 Nevertheless, our results are consistent with findings from more extreme droughts.
@@
 Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf hydraulic traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges.
+Vertical gradients in stem and leaf traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges.
 Taller trees also face different microenvironments (Fig.
@@
 2d).
-Even under non-drought conditions, evaporative demand increases with tree height, and tall trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
-Exposed canopy leaves reach higher temperatures (**REFS**), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Even under non-drought conditions, evaporative demand increases with tree height in taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
+Exposed canopy leaves reach higher temperatures (*[Smith & Nobel 1977](https://doi.org/10.1093/jxb/28.1.169)*), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
 Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
@@
 The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
-Additional research comparing drought responses of young and old forest stands, along with and short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
+Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
 
@@
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses--with both appearing in the top overall model and two of the three top models for each drought (Table 5)--is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**).
-It is scientifically exciting in that it indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
+It is scientifically exciting in that this finding indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
@@
 Thomspon, and Victoria Meakem for original collection and processing of cores.
+Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements and to Edward Brzostek's lab for collaboration on leaf sampling.
 ** others??** This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper.

### 2019-11-08 manuscript/climate_sensitivity_draft.Rmd
added citation

@@
 Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using the Leafarea R package (**citation**).
+Leaf area was calculated from scanned images using the Leafarea R package (*[Katabuchi, M.
+2015.
+LeafArea: an R package for rapid digital image analysis of leaf area.
+Ecological Reserach 30:1073-1077](https://cran.r-project.org/web/packages/LeafArea/citation.html)*).
 $LMA$ was calculated as the ratio of leaf dry mass to fresh area.

### 2019-11-09 manuscript/climate_sensitivity_draft.Rmd
Formatted LeafArea reference

--- 2019-11-07 manuscript/climate_sensitivity_draft.Rmd
@@
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
+  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr', 'LeafArea'
 ), 'packages.bib')
@@
 S3).
-Specifically, average tree height growth was confined to <0.5m from 1966 to 1977, ~1m from 1977 to 1999, and **# (IAN)** from 1999 to 2018.
+Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
 
@@
 Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
-Leaf area was calculated from scanned images using the Leafarea R package (**citation**).
+Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea].
 $LMA$ was calculated as the ratio of leaf dry mass to fresh area.

### 2019-11-09 scripts/graphs_for_manuscript.R
Determined avg growth by year and plots for scbi plot book

@@ -172,15 +172,18 @@ test <- meh[year %in% c("1999", "2018"), .(shrunk = height.m[1] - height.m[2]),
             ][, .(perc= sum(shrunk >0, na.rm=TRUE)/sum(shrunk>0 | shrunk <=0, na.rm=TRUE)), 
                   by=.(position_all_abb)]
 
-#dbh growth showing almost no negative growth from 1999 to 2018 (this means height equations are wrong)
-test <- meh[year %in% c("1999", "2018"), .(shrunk = dbh_old.cm[1] - dbh_old.cm[2]), 
-            by=.(tree, position_all_abb)
-            ][, .(perc= sum(shrunk >0, na.rm=TRUE)/sum(shrunk>0 | shrunk <=0, na.rm=TRUE)), 
-              by=.(position_all_abb)]
+#avg growth between each of the scenario years
+meh$group <- cut(meh$year, breaks=3)
+test66 <- meh[year %in% c("1966", "1977"), .(gro = height.m[2] - height.m[1]), by=tree
+            ][, .(avg_gro = mean(gro, na.rm=TRUE))]
+test77 <- meh[year %in% c("1977", "1999"), .(gro = height.m[2] - height.m[1]), by=tree
+              ][, .(avg_gro = mean(gro, na.rm=TRUE))]
+test99 <- meh[year %in% c("1999", "2018"), .(gro = height.m[2] - height.m[1]), by=tree
+              ][, .(avg_gro = mean(gro, na.rm=TRUE))]
+avg_growth <- data.frame(g66_77 = test66$avg_gro,
+                         g77_99 = test77$avg_gro,
+                         g99_18 = test99$avg_gro)
 
-test <- meh[year %in% c("1999", "2018"), .(shrunk = dbh_old.cm[1] - dbh_old.cm[2]), 
-            by=.(tree, position_all_abb)
-            ][shrunk>0]
 #######################################################################################
 #3 Get PLA and TLP as function of TWI and height ####
 library(ggstance)
@@ -356,6 +359,25 @@ levelplot(topo, margin=FALSE, scales=list(draw=FALSE),
    layer(sp.points(cored_points, pch=20, col = species_colors[species$sp_fact]))
 dev.off()
 
+#export TWI plot for SCBI plot book
[…]

### 2019-11-13 manuscript/climate_sensitivity_draft.Rmd
Added citations plus some of Nobby's corrections

@@
 ### Summary
-- As the climate changes, driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
-Yet it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.
+- As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
+However, it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.
 
-- In this study, we analyze tree-ring records for twelve species representing 97% of woody productivity in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over a 60-year period (1950 - 2009).
+- In this study, we analyze tree-ring records for twelve species of an oak-hickory forest -  representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over a 60-year period (1950 - 2009).
 
@@
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
-An important aspect of this uncertainty lies in responses to drought (*[Kennedy et al.
-2019](https://doi.org/10.1029/2018MS001500)*).
+An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010]; **REFS**).
-During drought, larger trees generally suffer greater reductions in growth and increases in mortality (e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
-As a result, forest drought responses stand to strongly impact forest feedbacks to climate change (*e.g., McDowell et al.
-2016(doi:10.1038/nclimate2873); Charney et al.
-2016 (doi: 10.1111/ele.12650)*).
-Yet, accurate characterization of drought responses remains a modeling challenge (*[Kennedy et al.
-2019](https://doi.org/10.1029/2018MS001500)*)-- in part because some of the mechanisms underlying drought responses remain unclear (**REF**).
+Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010; **REFS**].
+During drought, larger trees generally suffer greater reductions in growth and increases in mortality [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
+As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016].
+Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**).
 Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
@@
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
-@bennett_larger_2015 showed that in forests globally, large trees suffer greater growth reductions during drought, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
+@bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
 However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic meaning.

### 2019-11-16 manuscript/climate_sensitivity_draft.Rmd
loaded Bill's feedback
plust created folder, added Bill & Norm as definite authors
Consolidate author feedback - up to end of 2nd paragraph of intro
#65

Also consolidated all author feedback into one document (minus Zailaa)

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest
 
 **Authors:** 
-Ian R. McGregor^1,2^
-Ryan Helcoski^1^
-Norbert Kunert^1,3^
-Alan J. Tepley^1,4^
-Erika B. Gonzalez-Akre^1^
-Valentine Herrmann^1^
-Joseph Zailaa^1,5^
-Atticus Stovall^1,6^
-Norman A. Bourg^1^?
-William J. McShea^1^?
-Neil Pederson^7^
-Lawren Sack^8^
+Ian R. McGregor^1,2^,
+Ryan Helcoski^1^,
+Norbert Kunert^1,3^,
+Alan J. Tepley^1,4^,
+Erika B. Gonzalez-Akre^1^,
+Valentine Herrmann^1^,
+Joseph Zailaa^1,5^,
+Atticus Stovall^1,6^,
+Norman A. Bourg^1^,
+William J. McShea^1^,
+Neil Pederson^7^,
+Lawren Sack^8^,
 Kristina J. Anderson-Teixeira^1,3^*
@@
 3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
-4.
+4. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
 5. Biological Sciences Department; California State University; Los Angeles, CA, USA
@@
 
-text | word count | other | n
+Text | word count | other | n
 ---------------------------------------------|---------------------|-----------------------------|--------------------
@@
 
-- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance.
-Resistance was lower among species whose leaves had higher percent loss of area upon dehydration and lost turgor at less negative water potential.
-However, there was substantial variation in the best predictor variables across given droughts.
+- Individual-level drought resistance decreased with tree height, whereas crown position and root water access explained little variation.
+Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential.
+However, there was substantial variation in the best predictor variables across the three drought periods.
+**I think Alan's comments here warrant discussion.
+I believe the answer to his second comment is yes.
+In addition, Erika suggests spelling out the best predictor traits despite the point below.
+Thoughts?**
 
@@
 *Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.*
+Words that are in the title can, and should, be among these.* **Erika is of the opinion that broadleaf should not be included here, instead using "deciduous".
+She also added deciduous to the title.
+I'm unsure if this is redundant**
 
@@
 ### Introduction
+**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010; **REFS**].
-During drought, larger trees generally suffer greater reductions in growth and increases in mortality [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] **Neil adds in a number of refs here, unsure if we should just include all of them**, sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@allen_global_2010; **REFS**] **Alan suggests not citing Allen2010 here and instead gives a list of other sources.
+Erika also suggests 2 papers**.
+During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
 As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016].
@@
 @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
-However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic meaning.
-This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
-First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], which becomes a greater liability during drought [@zhang_size-dependent_2009].
-Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**).
-Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought.
+However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic response.
+That study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper.
+Bill says this sentence is confusing but I believe it is fine**.
+First, they found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
+Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
+Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018].
-The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience.
-Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
-Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.
+**For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
+The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things.
+Alan points out we should define these terms here since is the first time they're used.
+Unsure what your thoughts are on that**.
+Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
+Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
 

### 2019-11-16 manuscript/climate_sensitivity_draft.Rmd
Consolidated comments up to end of 3rd paragraph of intro
#65

@@
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] **Neil adds in a number of refs here, unsure if we should just include all of them**, sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@allen_global_2010; **REFS**] **Alan suggests not citing Allen2010 here and instead gives a list of other sources.
-Erika also suggests 2 papers**.
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017].
 During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
@@
 @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
-However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic response.
-That study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper.
-Bill says this sentence is confusing but I believe it is fine**.
-First, they found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
+However, they quantified tree size based on DBH, which has no direct mechanistic response.
+Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper; I'm unsure how you want to consolidate that with this sentence.**.
+First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
 Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
@@
 
-A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response.
-Xylem architecture plays a role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but does not differentiate species beyond broad classes.
+A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response [@elliott_forest_2015; @martinbenito_convergence_2015].
+Separating species by ring porosity provides a first step toward identifying sensitivity, with diffuse-porous species tending to be more sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but additional insight similar to @roskilly_conflicting_2019 is needed to better predict sensitivity at species or individual levels.
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
 However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018).
-Hydraulic traits including water potentials at which percent loss conductivity passes a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function, including leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the osmotic potential at turgor loss point ($\pi_{tlp}$)--i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016] -- are emerging as traits with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
-The ability of $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
+Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
+These include traits such as the fraction of conductivity lost to embolism, is another relevant metric (percent loss conductivity) [**REF??**], leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]), and the leaf water potential at turgor loss point ($\pi_{tlp}$), i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
 

### 2019-11-17 manuscript/climate_sensitivity_draft.Rmd
Consolidate author feedback to end of methods
#65

NOT including Lawren's yet

@@
 
-- Individual-level drought resistance decreased with tree height, whereas crown position and root water access explained little variation.
+- Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation.
 Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential.
 However, there was substantial variation in the best predictor variables across the three drought periods.
-**I think Alan's comments here warrant discussion.
-I believe the answer to his second comment is yes.
-In addition, Erika suggests spelling out the best predictor traits despite the point below.
-Thoughts?**
+**See comment 1 and 2**
 
-- We conclude that height--as opposed to canopy position, root water access, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
+- We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
 
 *Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.* **Erika is of the opinion that broadleaf should not be included here, instead using "deciduous".
-She also added deciduous to the title.
-I'm unsure if this is redundant**
+Words that are in the title can, and should, be among these.* **See comment 3 **
 
@@
 ### Introduction
-**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
+**See comment 4**
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
@@
 However, they quantified tree size based on DBH, which has no direct mechanistic response.
-Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper; I'm unsure how you want to consolidate that with this sentence.**.
+Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **see comment 5**.
 First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
@@
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-**For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
-The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things.
-Alan points out we should define these terms here since is the first time they're used.
-Unsure what your thoughts are on that**.
+**see comment 6** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
+The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **see comment 7**.
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
-Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
+Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
 
@@
 
-A final fundamental question regarding forest responses to drought is whether tree size and species' traits have similar influence across droughts, or whether variability in factors such as drought severity, duration, and timing interact with tree size and traits such that components of the community respond differently across droughts.
-No two droughts are the same, and tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018].
-However, we are not aware of any studies that compare how tree size and species' traits mediate growth responses across droughts.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), these don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally focus on species-level responses, which preclude consideration of the roles of tree size and microenvironment.
-The ecological studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
-Thus, our knowledge of forest responses to more modest but frequent droughts--e.g., those with historical return intervals on the order of a decade--remains more limited.
-While the tendency for larger trees to suffer more certainly predominates [@bennett_larger_2015], there are exceptions (*e.g.*, **REFS??**).
-There is also evidence that the degree to which larger trees suffer more increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
-[*Are there any studies showing interactions of drought type with traits?*] Thus, while we expect many of the fundamental mechanisms shaping drought responses to be universal, we have little undertanding of how tree size and traits interact with drought characteristics to result in differential responses across droughts.
+A final central question regarding forest responses to drought is whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
+Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they *don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally* **see comment 8** focus on species-level responses, and do not consider the roles of tree size and microenvironment.
+The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
+Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited.
+While the tendency for larger trees to experience more severe drought certainly predominates the literature [@bennett_larger_2015], there are exceptions [*e.g.*, @suarez_factors_2004; @zang_size-dependent_2012; @mcdowell_mechanisms_2008].
+There is also evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
+[*Are there any studies showing interactions of drought type with traits?*] **see comment 9** Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.
 
-Here, we combine tree-ring records covering three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to test a series of hypotheses and associated specific predictions (Table 1) designed to yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses.
+To yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
 First, we focus on the role of tree size and its interaction with microenvironment.
-We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion represented by only one study in [@bennett_larger_2015] (*H1.0*).
-We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and root water access, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
+We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion **see comment 10** represented by only one study in [@bennett_larger_2015] (*H1.0*).
+We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
+**see comments 11,12**.
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will follow observed patterns in relation to wood density (negative effect; @hoffmann_hydraulic_2011), specific leaf area (positive effect), and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*).
-We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought (*H2.3*). 
+We hypothesize that drought resistance will be negatively affected by wood density (following observed patterns by **e.g.** @hoffmann_hydraulic_2011), and positively affected by specific leaf area and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*).
+We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*).
 Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).
@@
 SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Elevations range from 273-338m above sea level [@gonzalezakre_patterns_2016] with a topographic relief of 65m [@bourg_initial_2013].
+Elevations range from 273-338m above sea level with a topographic relief of 65m [@bourg_initial_2013].
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
-Dominant tree taxa include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
+Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
+**see comment 13**
 
@@
 
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironemnt, and species traits (Table 2).
-The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH, tree location in the plot to determine the topographic wetness index, and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
+All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
+
+Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment, and species traits (Table 2).
+Destructive sampling was only carried out within the plot.
+**see comment 14** The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
 Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
-We analyzed tree-ring data from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
-Cores were obtained in 2010-2011 or 2016-2017 from a breast height of 1.3m using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals of DBH $\ge$ 10cm [@bourg_initial_2013].
+We analyzed tree-ring data (cambial growth increment) **see comment 15** from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
+Cores were collected at breast heigh (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer.
+In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013].
 In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
@@
 
-For each tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
+For each cored tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
 Prior DBH was estimated using the following equation:
@@
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter without bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
+Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter inside bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
 
 Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
-Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic LiDAR [@stovall_terrestrial_2018].
+Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
 Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
@@
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
+**see comments 16,17**
 
 Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-As there was no way to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
+As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
 While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2) and height change since the beginning of the study period indicated that change was likely slow (Fig.
+2d) and height variance since the beginning of the study period indicated that change was likely slow (Fig.
 S3).
-Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
+**see comment 18,19** Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
 
@@
 
-Hydraulic traits were collected at SCBI (Table 3).
-In August 2018, we sampled small sun-exposed branches from three individuals of each species in and around the ForestGEO plot.
-Samples were re-cut under water by at least two notes and re-hydrated overnight in buckets covered with opaque plastic bags before measurements.
+Hydraulic traits were collected at SCBI (Table 3) in August 2018.
+We sampled small sun-exposed branches **see comment 20** from three individuals of each species in and around the ForestGEO plot.
+Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken.
 Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
@@
 $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
+**see comment 21** We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
@@
 Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
+**see comment 22**
 
 To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile, for the years 2016-2018 [@noauthor_national_2018].
+We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile **see comment 23**, for the years 2016-2018 [@noauthor_national_2018].
 After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
+The range of these means were then aggregated at a month scale.
 
@@
 We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
-Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R (version 3.5.3).
+Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
+**see comment 24**
 Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site.
@@
 These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
-The driest years were identified through simply ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
+The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
+**see comment 25**
 
-*Analysis*
+*Statistical Analysis*
 
-For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to that of the five previous years [@lloret_components_2011].
-Analyses focused on testing the predictions presented in Table 1, most of which consider $R$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model fit.
+For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
+Analyses focused on testing the predictions presented in Table 1, with either $R$ or $H$ as the response variable.
+**see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.
+
 Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
-We first tested the predictor variables independently, counting a variable as useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
-Because height had a substantial influence on $R$, we included it in the null model for testing other variables (Table 4).
+In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $R$ given height's substantial influence.
+Variables were considered useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
 
-Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R$.
-All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model.
+Once these beset predictor variables were determined, we calculated the best overall models for predicting $R$ across all traits, each individual drought year, and all years combined.
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
@@
 If the variable appeared in only some of the models, we considered this partial support.
+**see comment 27**
 
@@
 There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**).
-There was no evidence that root water access increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
+There was no evidence that soil water availability increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
 The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
@@
 
-Our analysis indicates that height--as opposed to canopy position or root water access--is the primary factor through which tree size mediates drought response.
+Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response.
 Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
@@
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
-The observed height-sensitivity of $R$, together with the apparent lack of importance of root water access (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from root water access to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
+The observed height-sensitivity of $R$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
@@
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
+Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
 Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether and how drought tolerance changes as forests age.

### 2019-11-19 manuscript/climate_sensitivity_draft.Rmd
Updated with author feedback to end of 2nd paragraph in results
#65

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest **see comment 28**
 
@@
 However, it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.
+**see comment 29**
 
-- In this study, we analyze tree-ring records for twelve species of an oak-hickory forest -  representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
 - Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation.
-Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential.
+Resistance was greater both among species whose leaves had a lower turgor loss point, and whose leaves experienced less shrinkage upon dessication.
 However, there was substantial variation in the best predictor variables across the three drought periods.
@@
 - We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
+**see comment 30**
 
@@
 ### Introduction
-**See comment 4**
+**See comment 4,31**
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
@@
 
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Figs.
+**Nobby questions using "useful predictor" as a term, instead using more jargon I think**
+In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs.
 1, S2).
 We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
-Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp.
-(*Cite Shenandoah paper, if accepted*).
-Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019].
+Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
+[*Cite Shenandoah paper, if accepted*; @twery_effects_1991].
+
+Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
 Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were not identified as a pointer yeasr (Table S3)
+The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
+**see comment 32**
 

### 2019-11-20 manuscript/climate_sensitivity_draft.Rmd
Added feedback up to beginning of last 3 pgrph of results
#65

@@
 
-Height measurements (n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
+Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
 Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
@@
 Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
-Species-specific height allometries were developed (Table S2).
+Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
@@
 
-Crown positions were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+Crown positions ($CP$) were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
 As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
@@
 Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
+Specifically, $R$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth.
 **see comment 24**
@@
 S2).
-In terms of intensity during the peak growing season of the focal year, the 1999 drought was the most intense (lowest PDSI) during May-July.
-It was surpassed in intensity in August by the 1966 drought, which was otherwise the second most intense during the peak growing season.
-The 1977 was the least intense throughout the growing season.
-All droughts increased in intensity from May through August, but differed in the months previous.
+1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season.
+The 1977 drought was the least intense throughout the growing season.
+All droughts increased in intensity from May through August, but differed in the previous months.
 The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
@@
 1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
-Some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
+All droughts yielded about 5% more trees experiencing negative growth ($R$ $\le$ 0.7) as defined by our drought threshold, specifically, 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
+However, some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
+**see comment 33**
 
@@
 
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no signficant effect for 1977 or 1999 individually (Tables 1, 4).
+Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
 The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4).
@@
 Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-When considered alone, $CP$ had a significant influence only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$.
-When considered in conjunction with $H$, $CP$ came out as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$.
+**see comment 34** When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$.
+When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$.
 Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
 Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
-Thus, $CP$ was sometimes a useful predictor of $R$, but overall had a weak effect relative to that of $H$.
+Although $CP$ was a useful predictor of $R$, it overall had a weak effect relative to that of $H$.
 
@@
 2).
-Specifically, maximum daily wind speeds were signficantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
+Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
 2a).
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) have a greater tendency to suffer more in drier microenvironments with greater depth to the water table.
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought.
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
 *H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
@@
 
-We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5).
-Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-Of these, $PLA_{dry}$ was a strong predictor for the 1966 drought and overall, with consistently negative coefficients across all droughts.
-$\pi_{tlp}$ never came out as significant (dAIC $\le$ 2) in the univariate models but had a consistently negative coefficient (Table 4).
+Our results partially support Hypothesis *H2.1*, where species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Tables 1,4,5).
+Therefore, these were excluded as candidate variables for the full multivariate models.
+In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
+$PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients.
+$\pi_{tlp}$ was not significant in the univariate models but had a consistently negative coefficient (Table 4).
 Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
@@
 
-We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance.
-In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).
-However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ and $LMA$, which were not useful predictors of $R$.
-Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig.
+The results reject Hypothesis *H2.2*, finding no evidence that taller trees had traits associated with greater drought resistance.
+In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $R$.
+Furthermore, although correlations were statistically significant, trait variability within each height class overwhelmed any vertical trends (Fig.
 2e-f).
 
-We support the hypothesis (*H2.3*) that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
+Our findings support the hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
 As discussed above, there was little meaningful variation in traits with height at the community level.
-When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5)--reversing a non-signficant positive $ln[H]$-$R$ correlation in the univariate model for the 1999 drought (Table 4).
+When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5).
+Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$R$ correlation (Table 4).
 

### 2019-11-20 scripts/graphs_for_manuscript.R
Fixed a reference number

@@ -131,7 +131,7 @@ for(w in seq(along=height_regr$sp)){
    }
 }
 
-scbi18_ht$height.m <- exp(scbi18_ht$height.ln.m) #used below in #4
+scbi18_ht$height.m <- exp(scbi18_ht$height.ln.m) #used below in #3
 
 ##2c. combine heights to make plot with all years ####
 

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
Updated with author feedback to end of 1st paragraph of discussion
#65

@@
 
-- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microhabitat, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
@@
 Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**).
-Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
+Understanding forest responses to drought requires increased functional understanding of how tree size, microenvironment, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
 
@@
 
-To yield functional understanding of how tree size, microenvironment, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
+To yield functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
 First, we focus on the role of tree size and its interaction with microenvironment.
@@
 
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment, and species traits (Table 2).
+Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 2).
 Destructive sampling was only carried out within the plot.
@@
 
-Our findings support the hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
+Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
 As discussed above, there was little meaningful variation in traits with height at the community level.
@@
 
-We reject the hypothesis (*H3.1*) that overall community responses varied across droughts.
+Our results reject the hypothesis that overall community responses varied across droughts (*H3.1*).
 Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
-This is consistent with the observation that the distribution of $R$ values was similar across droughts (Fig.
-1b).
+**see comment 36**
 
-We mostly reject the hyopthesis (*H3.2*) that directions of responses varied across droughts.
+Our findings give marginal support for Hypothesis *H3.2*, whereby we predicted directions of responses varied across droughts.
 In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
-However, there were a few exceptions--most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5).
-These differences may very well be random, as opposed to statistically meaningful.
-Among the univariate models, there was no instance where predictor variables signficantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
-Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest of six models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-The difference most likely to be real is that ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5), but note that $XP$ was not a signficant predictor on its own for the 1977 drought.
+However, there were a few exceptions-most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5).
+These differences may be spurious as opposed to statistically meaningful.
+Among the univariate models, there was no instance where predictor variables significantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
+Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest predictor model of the six total models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4).
+It is probable that in the case of ring porosity, ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5).
+Note, however, that $XP$ was not a signficant predictor on its own for the 1977 drought.
 
-We support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
+Further, our findings support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
 For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
+**see comment 37**
 
@@
 
-Our results reveal how tree size, microhabitat, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
-The tendencey for larger trees to suffer more, observed here as in forests around the world [@bennett_larger_2015], was driven primarily by their height.
-There was a marginal additional effect of crown exposure, with the most exposed and the most suppressed trees suffering most--consistent with observations of both greater drought sensitivity of exposed trees (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and greater sensitivity of suppressed and crowded individuals (**REFS**).
-There was no evidence that soil water availability increased drougth resistance; in contrast, trees in wetter topographic positions suffered more (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
-The lower drought resistance of larger trees was not driven by any tendency for the canopy to be dominated by more drought-sensitive species.
-Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture.
-However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts, which is physiologically logical ([@scoffoni_leaf_2014];[@bartlett_correlations_2016]; [@medeiros_extensive_2019]) but scientifically novel in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
-The direction of responses was mostly consistent across droughts, indicating that they were driven by fundamental physiological mechanisms; however, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals suffer most.
+Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
+The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
+There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees.
+This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**).
+**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
+**see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy.
+Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture.
+However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019].
+This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms.
+However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
@@
 1a, @helcoski_growing_2019).
-Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature.
+Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances.
 While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
 1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, combined with the fact that many factors other than climate affect tree growth in closed-canopy forests, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
+It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
 Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
worked on abstract, methods

@@
 - As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
-However, it remains unclear how tree size and species' traits interactively shape tree growth responses during droughts.
-**see comment 29**
+The dendrochronological record provides a window through which we can understand how tree size and species’ traits interactively shape tree growth responses during droughts.
 
@@
 
-- Individual-level drought resistance decreased with tree height, whereas crown position and soil water availability explained little variation.
-Resistance was greater both among species whose leaves had a lower turgor loss point, and whose leaves experienced less shrinkage upon dessication.
+- Individual-level drought resistance decreased with tree height, which was the dominant size-related variable affecting drought response.
+Resistance was greater both among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon dessication.
 However, there was substantial variation in the best predictor variables across the three drought periods.
-**See comment 1 and 2**
 
-- We conclude that height--as opposed to canopy position, soil water availability, or species' traits--is the primary factor through which tree size mediates drought response, that leaf hydraulic traits are useful indicators of drought sensitivity, and that directions but not strengths of responses are consistent across droughts.
-**see comment 30**
+- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree ring record spanning historical droughts.
+Thus, these traits can be useful for predicting future droughts under climate change.
 
@@
 Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
-**see comment 13**
 
@@
 Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 2).
-Destructive sampling was only carried out within the plot.
-**see comment 14** The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
+The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
@@
 
-We analyzed tree-ring data (cambial growth increment) **see comment 15** from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
+We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
 Cores were collected at breast heigh (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer.
@@
 
-Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
+Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
 Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
@@
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
-**see comments 16,17**
 
 Crown positions ($CP$) were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
-As it was unfeasible to retroactively estimate crown position, we assumed that 2018 crown position was reflective of each tree's position over the past 60 years.
-While some trees undoubtedly changed position, an analysis of crown position relative to height (Fig.
-2d) and height variance since the beginning of the study period indicated that change was likely slow (Fig.
+While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
+An analysis of crown position relative to height (Fig.
+2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions, although dominant and co-dominant trees were similar in height (Fig.
 S3).
-**see comment 18,19** Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
+Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
+
+While some canopy positions undoubtedly shifted in the # years between the 1966 drought and our observations in 2018, change was relatively slow, and bias would more likely result in type x error with regards to testing our hypothesis.
 
@@
 We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile **see comment 23**, for the years 2016-2018 [@noauthor_national_2018].
-After filtering for missing and outlier values, the data was consolidated to represent the mean values per sensor height per day.
+After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
 The range of these means were then aggregated at a month scale.

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
manuscript updates, changed R symbol to Rt
issue #65

@@
 $PLA$ was calculated as the percent loss of area between fresh and dry leaves.
-$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume.
-**see comment 21** We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
+$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.
+We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
-Data and R scripts for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
-**see comment 22**
+Data for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
 
@@
 We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
-Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($R$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
-Specifically, $R$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth.
+Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
+Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth.
 **see comment 24**
@@
 
-For each drought period, we calculated drought resistance ($R$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
-Analyses focused on testing the predictions presented in Table 1, with either $R$ or $H$ as the response variable.
-**see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $R$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
+Analyses focused on testing the predictions presented in Table 1, with either $Rt$ or $H$ as the response variable.
+**see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
 We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.
@@
 Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
-In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $R$ given height's substantial influence.
-Variables were considered useful for predicting $R$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
+In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
+Variables were considered useful for predicting $Rt$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
 
-Once these beset predictor variables were determined, we calculated the best overall models for predicting $R$ across all traits, each individual drought year, and all years combined.
+Once these beset predictor variables were determined, we calculated the best overall models for predicting $Rt$ across all traits, each individual drought year, and all years combined.
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
@@
 1b).
-All droughts yielded about 5% more trees experiencing negative growth ($R$ $\le$ 0.7) as defined by our drought threshold, specifically, 29.43% in 1966, 31.86% in 1977, and 26.81% in 1999.
-However, some trees exhibited increased growth: ($R$ > 1.0): 26.31% in 1966, 21.92% in 1977, and 25.57% in 1999.
+All droughts yielded about 5% more trees experiencing negative growth ($Rt$ $\le$ 0.7) as defined by our drought threshold, specifically, 29% in 1966, 32% in 1977, and 27% in 1999.
+However, some trees exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
 **see comment 33**
@@
 When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $ln[H]$ and $R$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
+We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
 
@@
 2d).
-Crown position was a much poorer predictor of $R$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-**see comment 34** When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $R$.
-When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $R$.
-Crown position was included in almost half of the top models, with mixed results as to how $R$ varied with $CP$ (Table 5).
+Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
+**see comment 34** When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
+When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$.
+Crown position was included in almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5).
 Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
-Although $CP$ was a useful predictor of $R$, it overall had a weak effect relative to that of $H$.
+Although $CP$ was a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.
 
@@
 
-Our results partially support Hypothesis *H2.1*, where species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Tables 1,4,5).
+Our results partially support Hypothesis *H2.1*, where species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $Rt$ in the univariate models (all dAIC $\le$ 0.22; Tables 1,4,5).
 Therefore, these were excluded as candidate variables for the full multivariate models.
@@
 $\pi_{tlp}$ was not significant in the univariate models but had a consistently negative coefficient (Table 4).
-Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977.
+Whereas ring-porous species had highest $Rt$ overall and in the 1966 and 1999 droughts, they had lower $Rt$ in 1977.
 Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.
@@
 The results reject Hypothesis *H2.2*, finding no evidence that taller trees had traits associated with greater drought resistance.
-In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $R$.
+In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $Rt$.
 Furthermore, although correlations were statistically significant, trait variability within each height class overwhelmed any vertical trends (Fig.
@@
 
-Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $R$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
+Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
 As discussed above, there was little meaningful variation in traits with height at the community level.
-When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $R$ was consistently negative (Table 5).
-Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$R$ correlation (Table 4).
+When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5).
+Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$Rt$ correlation (Table 4).
 
@@
 Our results reject the hypothesis that overall community responses varied across droughts (*H3.1*).
-Within the context of mixed effects models, there were no significant differences in $R$ across drought years (Table 4).
+Within the context of mixed effects models, there were no significant differences in $Rt$ across drought years (Table 4).
 **see comment 36**
@@
 Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest predictor model of the six total models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-It is probable that in the case of ring porosity, ring-porous species had lower $R$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5).
+It is probable that in the case of ring porosity, ring-porous species had lower $Rt$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5).
 Note, however, that $XP$ was not a signficant predictor on its own for the 1977 drought.
@@
 Further, our findings support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
-For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $R$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
+For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
 **see comment 37**
@@
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*.
-The implication is that dominant crown positions did have a marginally negative influence on $R$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $R$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
-The observed height-sensitivity of $R$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
+The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
+The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
@@
 Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-Although there were not signficant differences in $R$ among the drought years, $R$ tended to be somewhat lower in 1966 (Fig.
+Although there were not signficant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig.
 1, Table 4).
@@
 In 1977, which was the weakest drought in terms of PDSI (Fig.
-S2, Table S3), ring-porous species had lower $R$, contrasting with their higher $R$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
+S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
 In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
-Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
-**see comment 32**
+Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively.
+The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
 
@@
 However, some trees exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
-**see comment 33**
 
@@
 Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
-**see comment 34** When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
+When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
 When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$.

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
working on issue #65
#65

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest **see comment 28**
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest **see comment KAT**
 
@@
 However, they quantified tree size based on DBH, which has no direct mechanistic response.
-Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **see comment 5**.
+Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
 First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
@@
 Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015].
-**see comment 6** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
-The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **see comment 7**.
+Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
+2019](https://www.pnas.org/content/pnas/early/2019/06/14/1900734116.full.pdf)].
+The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
@@
 Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they *don't test for differential trait effects across periods of water shortage [@dorangeville_drought_2018] and generally* **see comment 8** focus on species-level responses, and do not consider the roles of tree size and microenvironment.
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
 The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
@@
 There is also evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
-[*Are there any studies showing interactions of drought type with traits?*] **see comment 9** Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.
+Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.
 
@@
 First, we focus on the role of tree size and its interaction with microenvironment.
-We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in an ecoregion **see comment 10** represented by only one study in [@bennett_larger_2015] (*H1.0*).
+We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015] (*H1.0*).
 We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).

### 2019-11-21 manuscript/climate_sensitivity_draft.Rmd
worked on intro comments
#65
comments 23-24
issue #65

@@
 Joseph Zailaa^1,5^,
-Atticus Stovall^1,6^,
+Atticus Stovall^1,6,7^,
 Norman A. Bourg^1^,
 William J. McShea^1^,
-Neil Pederson^7^,
-Lawren Sack^8^,
+Neil Pederson^8^,
+Lawren Sack^9^,
 Kristina J. Anderson-Teixeira^1,3^*
@@
 5. Biological Sciences Department; California State University; Los Angeles, CA, USA
-6.
-7. 
+6. Department of Environmental Sciences, University of Virginia, Charlottesville, VA 22903, USA
+7. NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD, USA
 8. 
+9.
 
@@
 We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
-**see comments 11,12**.
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
@@
 Hydraulic traits were collected at SCBI (Table 3) in August 2018.
-We sampled small sun-exposed branches **see comment 20** from three individuals of each species in and around the ForestGEO plot.
+We sampled small sun-exposed branches within eight meters of the ground from three individuals of each species in and around the ForestGEO plot.
 Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken.
-Rehydrated leaves (n=3 per indivdiual) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
+Rehydrated leaves taken towards the apical end of the branch (n=3 per indivdiual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
 Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea].
@@
 To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile **see comment 23**, for the years 2016-2018 [@noauthor_national_2018].
+We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile spanning from 7.2 m height to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
 After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
@@
 We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
-Specifically, candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
+Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community.
+Candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
 Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth.
-**see comment 24**
 Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site.

### 2019-11-22 manuscript/climate_sensitivity_draft.Rmd
worked on #65

@@
 Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited.
-While the tendency for larger trees to experience more severe drought certainly predominates the literature [@bennett_larger_2015], there are exceptions [*e.g.*, @suarez_factors_2004; @zang_size-dependent_2012; @mcdowell_mechanisms_2008].
-There is also evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019].
+There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts.
 Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.
@@
 We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
-Cores were collected at breast heigh (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer.
+Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer.
 In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013].
@@
 
-Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
-Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
+Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).**Ian, I think we need to include a little more about the NEON data here.** Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
 Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
@@
 
-Crown positions ($CP$) were recorded in the field during the growing season of 2018 following the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+Crown positions ($CP$) were recorded for all cored trees that remained standing during the growing season of 2018.
+We followed the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
 While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
@@
 The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
-**see comment 25**
 
@@
 For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
+Thus, $Rt$<1 indicates reduced growth under drought.
 Analyses focused on testing the predictions presented in Table 1, with either $Rt$ or $H$ as the response variable.
-**see comment 26** The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
 We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.
@@
 This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**).
-**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017]), and larger rooting volume provided no advantage in the drier microenvironments.
+**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volume provided no advantage in the drier microenvironments.
 **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy.

### 2019-11-22 manuscript/climate_sensitivity_draft.Rmd
Added comments for discussion

@@
 The droughts considered here were of similar severity (Fig.
-1b) and fairly moderate; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
+1b) and fairly moderate **see comment 41**; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
 1a, @helcoski_growing_2019).
 Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances.
-While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
-1b), underlining the fact that these droughts did not induce extreme stress on the entire forest.
+**see comment 42,43** While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
+1b), **see comment 44** supporting the observation that these droughts did not induce extreme stress on the entire forest.
 It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
-Methodologically, the moderate nature of these droughts is an advantage because our analysis considers only trees that survived all of these droughts, and we lack information on the trees that were killed.
+The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008.
 These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*).
-Thus, the droughts considered here are substantially weaker than those that have triggered massive tree die-off (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and--to some extent--traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
-Nevertheless, our results are consistent with findings from more extreme droughts.
+Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
+This is important because while our results are consistent with findings from more extreme, multiannual droughts, these extremes have not occurred in the Eastern United States similar to the West, and therefore we do not know exactly how forests would respond under higher mortality conditions [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016].
+Given this consistency with our results, however, we believe this is an indication of a more intensive response.
 
@@
 Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker (higher LMA) leaves, more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges.
+Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
+Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
 Taller trees also face different microenvironments (Fig.
@@
 2d).
-Even under non-drought conditions, evaporative demand increases with tree height in taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018]).
-Exposed canopy leaves reach higher temperatures (*[Smith & Nobel 1977](https://doi.org/10.1093/jxb/28.1.169)*), particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime CO_2_ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
+Even under non-drought conditions, evaporative demand increases with tree height while taller trees are more closely coupled to the atmosphere **see comment 45** (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
+Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Furthermore, daytime $CO_2$ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
 Correlation between height and canopy position (Fig.
 2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment.
-However, signficant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig.
-2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position (Tables 1,4,5).
-Our analysis does have the limitation that canopy positions were recored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving from 1999 back to 1966 (Table 4).
+However, significant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig.
+2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position for the three droughts evaluated in our study area (Tables 1,4,5).
+
+Our analysis does have the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*.
-The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical enviromnental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
+The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
-The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere intreface as forests age [@bretfeld_plant_2018], such that tall trees--particularly the very tallest--are the most sensitive in mature forests.
-Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentanging the roles of tree height and crown exposure.
+**see comment 46** The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
+Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
 
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time (**verify- Neil, Alan**).
-Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses--with both appearing in the top overall model and two of the three top models for each drought (Table 5)--is consistent with studies demonstrating that these are physiolgoically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019] (**MORE REFS--KAT/NOBBY/LAWREN**).
-It is scientifically exciting in that this finding indicates that $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests, where linking hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
+The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
+At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.
+Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiolgically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015].
+Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
+The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests.
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
 
+**see comment 48**
 Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height.
 As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits.
-If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
+If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**; **see comment 49**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
 Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
 
-Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydaulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importantance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-Although there were not signficant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig.
+Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
+Although there were not significant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig.
 1, Table 4).
-Height and dominant canopy position had the stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the fact that this was the longest-lasting drought and and that height becomes a stronger predictor of mortality as the severity of the drought increases [@stovall_tree_2019].
-In 1977, which was the weakest drought in terms of PDSI (Fig.
+Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].
+**see comment 50** In 1977, which was the weakest drought in terms of PDSI (Fig.
 S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
-In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, xylem architecture was the most useful predictor, with diffuse-porous species suffering most.
+In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted.
 Thus, responses differed with the nature of each drought.

### 2019-11-22 manuscript/climate_sensitivity_draft.Rmd
Updated with coauthor feedback to end of paper
#65

@@
 This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**).
-**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volume provided no advantage in the drier microenvironments.
+**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments.
 **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy.
@@
 
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
+The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
 At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.
@@
 
-**see comment 48**
+**see comment 48; the first paragraph was condensed here, but unsure where this should be put now**
 Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height.
-As noted above, leaves found higher on a tree tend to have more hydraulically conservative traits, and therefore we would expect that average leaf characteristics of an individual tree would scale with its crown height, with taller individuals having on average more drought-resistant traits.
-If vertical trends for $PLA_{dry}$ and $\pi_{tlp}$, which have not been characterized (***LAWREN, IS THIS TRUE?**; **see comment 49**), follow the general pattern of increasing leaf drought resistance with height, our analysis would be biased in a conservative direction when assessing *H2.3* (Table 1)--*i.e.*, accounting traits on an individual level should result in a stronger negative effect of $ln[H]$.
 Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
@@
 Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].
-**see comment 50** In 1977, which was the weakest drought in terms of PDSI (Fig.
+**see comment 49** In 1977, which was the weakest drought in terms of PDSI (Fig.
 S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
-In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted.
+In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted (Fig.
+S2).
 Thus, responses differed with the nature of each drought.
@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been suffering disrpoportionately in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we here show that this is primarily driven by their height, with some contributions from canopy position.
+Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for getting at the very signficant question of whether and how drought tolerance changes as forests age.
+Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
+Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+@rodriguez-caton_long-term_2015].
 In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
 
-### Acknowledgements
+### Acknowledgements **see comment 51**
 Thanks to all researchers who helped to collect the data used here, in particular Jennifer C.
@@
 Thomspon, and Victoria Meakem for original collection and processing of cores.
-Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements and to Edward Brzostek's lab for collaboration on leaf sampling.
-** others??** This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper.
+Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
+This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper.
 accurate here?** **coauthors, other funding?**)

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
author affiliations

@@
 Neil Pederson^8^,
-Lawren Sack^9^,
+Lawren Sack^9,10^,
 Kristina J. Anderson-Teixeira^1,3^*
@@
 
-1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
-2.
+1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA 22630, USA
+2. **IAN- current affliliation**
 3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 4. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
-5. Biological Sciences Department; California State University; Los Angeles, CA, USA
+5. Biological Sciences Department; California State University; Los Angeles, CA 90032, USA
 6. Department of Environmental Sciences, University of Virginia, Charlottesville, VA 22903, USA
-7. NASA Goddard Space Flight Center, 8800 Greenbelt Rd., Greenbelt, MD, USA
-8. 
-9.
+7. NASA Goddard Space Flight Center; Greenbelt, MD 20771, USA
+8. Harvard Forest, Petersham, MA 01366, USA
+9. Department of Ecology and Evolutionary Biology; University of California, Los Angeles; Los Angeles, CA 90095, USA
+10. Institute of the Environment and Sustainability; University of California, Los Angeles; Los Angeles, CA 90095, USA
 

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
title & keywords

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest **see comment KAT**
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
 
@@
 
-**coauthors, fill in middle initials and affiliations, if missing***
-
 Text | word count | other | n
@@
 
-*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.* **See comment 3 **
+*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; annual growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
+Words that are in the title can, and should, be among these.*
 

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
working on intro, Lawren's comments

@@
 - Individual-level drought resistance decreased with tree height, which was the dominant size-related variable affecting drought response.
-Resistance was greater both among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon dessication.
+Resistance was greater  among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon dessication.
 However, there was substantial variation in the best predictor variables across the three drought periods.
@@
 - We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree ring record spanning historical droughts.
-Thus, these traits can be useful for predicting future droughts under climate change.
+Thus, these traits can be useful for predicting future drought responses under climate change.
 
-*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; annual growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
+*Key words*: annual growth; canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
 Words that are in the title can, and should, be among these.*
@@
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017].
-During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018].
-As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016].
-Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**).
-Understanding forest responses to drought requires increased functional understanding of how tree size, microenvironment, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.
+Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**too many refs here.
+reduce to 2-3**.
+However, the traits that predict tree growth and mortality responses to drought have been under debate, and are difficult to resolve with the limited available census data, which only rarely includes extreme drought.
+The aim of this study was to test how tree size and traits shaped growth responses to historical droughts in the tree-ring record.
 
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
-@bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016].
-However, they quantified tree size based on DBH, which has no direct mechanistic response.
-Their study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees.
-First, @bennett_larger_2015 found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
+@bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016].
+However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern.
+First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
 Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
@@
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
+A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
 2019](https://www.pnas.org/content/pnas/early/2019/06/14/1900734116.full.pdf)].
@@
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
-Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
+Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
 

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 reduce to 2-3**.
-However, the traits that predict tree growth and mortality responses to drought have been under debate, and are difficult to resolve with the limited available census data, which only rarely includes extreme drought.
+However, it is difficult to resolve factors conferring drought resistance in trees with the limited available census data, which only rarely includes extreme drought.
 The aim of this study was to test how tree size and traits shaped growth responses to historical droughts in the tree-ring record.

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
working on LS comments

@@
 First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
-Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
+Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**) **Lawren doesn't think CO2 would matter**.
 Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
+*A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
 2019](https://www.pnas.org/content/pnas/early/2019/06/14/1900734116.full.pdf)].
-The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
+**LS:—larger trees may have larger xylem conduits, which in turn may have less resistance to air seeding and subsequent embolism (Mark Olsen argument, right?) ** The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
 Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
+*
 
-A second fundamental question regarding forest responses to drought is how species' traits -- alone and in interaction with tree size -- influence drought response [@elliott_forest_2015; @martinbenito_convergence_2015].
-Separating species by ring porosity provides a first step toward identifying sensitivity, with diffuse-porous species tending to be more sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but additional insight similar to @roskilly_conflicting_2019 is needed to better predict sensitivity at species or individual levels.
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018).
-Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin tend to be more successful at predicting drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+Debates have also arisen regarding the traits that would influence tree growth responses to drought.
+Xylem architecture, for example, plays an important role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017], **but not in others (REFS)**.
+Indeed, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018] **and their role may be due to indirect correlations with other traits within life-history strategies (refs)**.
+Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
+Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
-These include traits such as the fraction of conductivity lost to embolism, is another relevant metric (percent loss conductivity) [**REF??**], leaf area shrinkage upon dessication ($PLA_{dry}$; [@scoffoni_leaf_2014]), and the leaf water potential at turgor loss point ($\pi_{tlp}$), i.e., the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
+
+**KAT is here with integrating LS comments**
 

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Xylem architecture, for example, plays an important role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017], **but not in others (REFS)**.
-Indeed, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018] **and their role may be due to indirect correlations with other traits within life-history strategies (refs)**.
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
+However, these traits have failed to explain drought responses in other cases [@anderegg_hydraulic_2018] (**REFS**), and the direction of response is not always consistent; for instance, higher wood density is associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Thus, their role omay be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
worked on treatment of traits in intro

--- 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
@@
 Debates have also arisen regarding the traits that would influence tree growth responses to drought.
-Xylem architecture, for example, plays an important role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017], **but not in others (REFS)**.
-Indeed, the direction of response is not always consistent; wood density correlated negatively with drought tree performance in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011] but positively or with no effect at a global scale [@greenwood_tree_2017; @anderegg_hydraulic_2018] **and their role may be due to indirect correlations with other traits within life-history strategies (refs)**.
+*Xylem architecture, for example, plays an important role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].* Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
+However, these traits have failed to explain drought responses in other cases (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We hypothesize that drought resistance will be negatively affected by wood density (following observed patterns by **e.g.** @hoffmann_hydraulic_2011), and positively affected by specific leaf area and xylem architecture (*i.e.*, ring or diffuse/ semi-ring porous), but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ will prove better predictors (*H2.1*).
+We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
+We test predictions that drought resistance are **negatively** correlated with wood density (following patterns observed by @hoffmann_hydraulic_2011) and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors (*H2.1*).
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*).

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
knit ms

@@
 Debates have also arisen regarding the traits that would influence tree growth responses to drought.
-*Xylem architecture, for example, plays an important role, with diffuse porous species tending to be more drought-sensitive than ring-porous species [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].* Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
 However, these traits have failed to explain drought responses in other cases (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
minor change

@@
 Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
-We test predictions that drought resistance are **negatively** correlated with wood density (following patterns observed by @hoffmann_hydraulic_2011) and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors (*H2.1*).
+We test predictions that drought resistance is **negatively** correlated with wood density (following patterns observed by @hoffmann_hydraulic_2011) and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors (*H2.1*).
+We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
 We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*).

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
minor updates

@@
 Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
-This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, *and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.* (**Neil, this is copied from Ryan's paper.
-accurate here?** **coauthors, other funding?**)
+This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.
 
@@
 IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
-**coauthors, please check/edit**

### 2019-12-06 manuscript/climate_sensitivity_draft.Rmd
update draft

@@
 **See comment 4,31**
+
 Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**too many refs here.
-reduce to 2-3**.
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**{too many refs here.
+reduce to 2-3}**.
 However, it is difficult to resolve factors conferring drought resistance in trees with the limited available census data, which only rarely includes extreme drought.
@@
 2019](https://www.pnas.org/content/pnas/early/2019/06/14/1900734116.full.pdf)].
-**LS:—larger trees may have larger xylem conduits, which in turn may have less resistance to air seeding and subsequent embolism (Mark Olsen argument, right?) ** The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
+**{LS:—larger trees may have larger xylem conduits, which in turn may have less resistance to air seeding and subsequent embolism (Mark Olsen argument, right?) }** The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
 Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
-*
+* {see Lawren's comment on order of presentation}
 
@@
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, these traits have failed to explain drought responses in other cases (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+However, in other cases these traits have failed to explain drought responses (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].
@@
 
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] **see comment 47** made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
+The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
 At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.

### 2019-12-08 manuscript/climate_sensitivity_draft.Rmd
Added snippet on NEON height measurements

@@
 1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA 22630, USA
-2. **IAN- current affliliation**
+2. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA
 3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
@@
 
-Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public GitHub repository (https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).**Ian, I think we need to include a little more about the NEON data here.** Measurement methods included manual [@stovall_assessing_2018, NEON], digital rangefinders [@andersonteixeira_size-related_2015], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
+Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
+Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
 Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
 Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
+Measurements from the National Ecological Observatory Network (NEON) were collected nearby the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
 Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$).

### 2019-12-09 manuscript/climate_sensitivity_draft.Rmd
removed CO2 height profile comment
Koike et al showed constant CO2 across the heights of tree crowns; higher CO2 is closer to the ground.

@@
 First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
-Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**) **Lawren doesn't think CO2 would matter**.
+Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
 Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
@@
 Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Furthermore, daytime $CO_2$ concentrations tend to decrease with height [@koike_leaf_2001], implying that water costs of $CO_2$ uptake increase with height.
 Correlation between height and canopy position (Fig.

### 2019-12-14 manuscript/climate_sensitivity_draft.Rmd
added ref

@@
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019] [Dai et al.
+2018](https://link.springer.com/article/10.1007/s40641-018-0101-6) **{too many}**, often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
 Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**{too many refs here.

### 2020-01-07 manuscript/tables&figures.Rmd
manuscript updates, changed R symbol to Rt
issue #65
removed H2.2, 2.3, 3.1, 3.2, 3.3 from Table 1
issue #67

@@
 Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
-  pack_rows("H1.0. Larger-diameter trees have lower drought resistance (R).", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1.0. Larger-diameter trees have lower drought resistance (Rt).", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2.2. At the community level, taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.3. When traits are accounted for, taller trees still have lower drought resistance.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.1. Resistance differs across the droughts considered here.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-01-07 manuscript/tables&figures.Rmd
worked on Table 1/ hypotheses
#67

@@
 Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
-  pack_rows("H1.0. Larger-diameter trees have lower drought resistance (Rt).", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1a. Larger, taller trees have lower drought resistance (Rt).", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1b. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1c. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("H2a. conventional traits....", 6, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2b. leaf hydraulic traits....", 9, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
   pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 17, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("species' traits", 12, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()

### 2020-01-07 manuscript/tables&figures.Rmd
working on table 1/hypotheses
issue #67

This still needs work

@@
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1a. Larger, taller trees have lower drought resistance (Rt).", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1b. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1c. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H1a. Larger, taller trees have lower drought resistance (Rt).", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("H1b. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("H1c. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2a. conventional traits....", 6, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H2b. leaf hydraulic traits....", 9, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("H2a. conventional traits....", 6, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("H2b. leaf hydraulic traits....", 9, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-01-07 manuscript/tables&figures.Rmd
small fix table 2

@@
   pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()

### 2020-01-07 manuscript/tables&figures.Rmd
working on table 1
#67

@@
 Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
-  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("H1a. Larger, taller trees have lower drought resistance (Rt).", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("H1b. Trees with more exposed crowns have lower drought resistance.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("H1c. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("H2a. conventional traits....", 6, 7, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("H2b. leaf hydraulic traits....", 9, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Hypothesis 1: Tree height is the dominant size-related variable affecting drought resistance (Rt)", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("1a. Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("1b. Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("1c. Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Hypothesis 2: Species' traits--particularly leaf hydraulic traits--predict Rt", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
   column_spec(1, width = "12cm") %>%
-  column_spec(2:5, width = "1cm") %>%
+  column_spec(2:5, width = "1.2cm") %>%
   column_spec(6, width = "2.4cm") %>%

### 2020-01-07 manuscript/climate_sensitivity_draft.Rmd
removed "interactively" shape from abstract

@@
 - As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
-The dendrochronological record provides a window through which we can understand how tree size and species’ traits interactively shape tree growth responses during droughts.
+The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape tree growth responses during droughts.
 
-- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits interactively shape drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 

### 2020-01-23 manuscript/climate_sensitivity_draft.Rmd
add Zuleta ref as counter-example to height trend

@@
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
-@bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016].
+@bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017].
 However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern.

### 2020-01-31 manuscript/tables&figures.Rmd
update/simplify Table 1
- simplified criteria for indicating support/ lack thereof
- updated to include limited set of models in Table 5 (issue #70)

@@
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and all top multivariate models (when applicable).
-Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction but that the effect was not significant in all models.") %>%
+Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
   pack_rows("Hypothesis 1: Tree height is the dominant size-related variable affecting drought resistance (Rt)", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%

### 2020-01-31 manuscript/tables&figures.Rmd
update Table 1
altered presenation of hypotheses and sub-hypotheses (issue #67), visual cleanup

@@
 Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
-  pack_rows("Hypothesis 1: Tree height is the dominant size-related variable affecting drought resistance (Rt)", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("1a. Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
@@
   pack_rows("1c. Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Hypothesis 2: Species' traits--particularly leaf hydraulic traits--predict Rt", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("2. Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-01-31 manuscript/tables&figures.Rmd
update Table 5 to dAICc<1
issue #70

@@
 kable(multivar_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AICc, and we show all models whose AICc value falls within 2.0 (dAICc<2) of the best model (bold).", format = 'latex', escape = TRUE) %>%
+Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
   add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 2," ", " ")) %>%
-  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 7, 9, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 10, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
-  pack_rows("", 16, 21, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
+  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 6, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 10, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(7, bold = T) %>%
+  row_spec(5, bold = T) %>%
+  row_spec(6, bold = T) %>%
   row_spec(10, bold = T) %>%
-  row_spec(16, bold = T) %>%
   kable_styling()

### 2020-01-31 manuscript/tables&figures.Rmd
update Table 2
removed reported range of values. This simplifies the table.

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
+  column_spec(2, italic = T) %>% 
   column_spec(4, width = "7cm") %>%

### 2020-01-31 manuscript/climate_sensitivity_draft.Rmd
update related to issue #70
worked on intro

@@
 First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
+To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019].
+Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}.
 Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
-Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
+Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
 Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-*A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019 [Roskilly et al.
-2019](https://www.pnas.org/content/pnas/early/2019/06/14/1900734116.full.pdf)].
-**{LS:—larger trees may have larger xylem conduits, which in turn may have less resistance to air seeding and subsequent embolism (Mark Olsen argument, right?) }** The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought.
+A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
+Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought.
 Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
 Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
-* {see Lawren's comment on order of presentation}
+{**see Lawren's comment on order of presentation**}
 
@@
 Whereas ring-porous species had highest $Rt$ overall and in the 1966 and 1999 droughts, they had lower $Rt$ in 1977.
-Results were similar in the context of multivariate models (Table 5), except that $PLA_{dry}$ had a small positive coefficient in the one 1999 model in which it was included.
+Results were similar in the context of multivariate models (Table 5).
 $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).

### 2020-02-03 manuscript/climate_sensitivity_draft.Rmd
worked on intro
#65
This now integrates Lawren's feedback

@@
 ### Introduction
-**See comment 4,31**
 
@@
 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6) **{too many}**, often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**{too many refs here.
+Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**{too many refs here.
 reduce to 2-3}**.
-However, it is difficult to resolve factors conferring drought resistance in trees with the limited available census data, which only rarely includes extreme drought.
-The aim of this study was to test how tree size and traits shaped growth responses to historical droughts in the tree-ring record.
+Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
+However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and are not necessarily associated with ecological data.
+The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.
 
@@
 Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
-{**see Lawren's comment on order of presentation**}
+{**ask Lawren to check revisions to this paragraph**}
 
@@
 
-**KAT is here with integrating LS comments**
-
-A final central question regarding forest responses to drought is whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
+The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
 Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
 While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
-The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
+The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
 Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited.
 There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts.
-Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics.
+Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.
 
@@
 First, we focus on the role of tree size and its interaction with microenvironment.
-We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015] (*H1.0*).
-We then test hypotheses designed to disentangle the relative importance of tree height (*H1.1*), crown exposure (*H1.2*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1.3*).
-Second, we focus on the role of species' functional and hydraulic traits and their interaction with tree height.
-We test predictions that drought resistance is **negatively** correlated with wood density (following patterns observed by @hoffmann_hydraulic_2011) and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors (*H2.1*).
+We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015].
+We then test hypotheses designed to disentangle the relative importance of tree height (*H1a*), crown exposure (*H1b*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1c*).
+Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits–particularly leaf hydraulic traits–predict Rt (*H2*).
+We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
 We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
-We then test whether these traits correlate with tree height (*H2.2*), potentially driving the observed tendency for taller trees to suffer more during drought after accounting for trait effects (*H2.3*).
-Finally, we focused on variability among droughts, asking how community resistance varied across droughts (*H3.1*) and whether the factors confirming vulnerability or resistance varied across droughts (*H3.2*).
 
@@
 
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
+Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo_2015].
 SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.

### 2020-02-03 manuscript/climate_sensitivity_draft.Rmd
editing methods & results

@@
 Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies have been published in association with @helcoski_growing_2019: (ITRDB; GitHub/Zenodo).
-**Ryan, have you heard back from ITRDB?
-We should also cite GitHub/Zenodo here.
-**
+The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).
 
@@
 
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI- ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).
+
 ### Results
@@
 
-*Tree size and drought resistance*
+*Tree size, microenvironment, and drought resistance*
 
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015] (*H1.0*), although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ as a univariate predictor (*H1.1*; Tables 1, 4).
+Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015], although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
+The same held true for $ln[H]$ as a univariate predictor (*H1a*; Tables 1, 4).
 When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1.1* (Table 1).
+We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1a* (Table 1).
 
@@
 2d).
-Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1.2* (Table 1).
+Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1b* (Table 1).
 When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
 When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$.
-Crown position was included in almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5).
-Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees.
-Although $CP$ was a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.
+Crown position was included in **almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5).
+Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees**.
+Although $CP$ was at times a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.
 
@@
 Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1.3*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
-*H1.3* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1c*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
+*H1c* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
 
-*Species' traits, height, and drought resistance*
+*Species' traits and drought resistance*
 
@@
 
-The results reject Hypothesis *H2.2*, finding no evidence that taller trees had traits associated with greater drought resistance.
-In part because of the large sample size (n=7549 trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4).**see comment 35** $WD$ and $LMA$ were the only traits where the correlation matched the predicted direction (*i.e.*, evidence that more drought-resistant traits are associated to taller trees), but ultimately neither of these traits were useful predictors of $Rt$.
-Furthermore, although correlations were statistically significant, trait variability within each height class overwhelmed any vertical trends (Fig.
-2e-f).
+*Combined effects of tree size and traits*
 
-Our findings support Hypothesis *H2.3* that the observed tendency for larger trees to have greater growth reductions during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Tables 1,5).
-As discussed above, there was little meaningful variation in traits with height at the community level.
+*Our findings indicate that the greater growth reductions in large trees during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Table 5).
+Tthere was little meaningful variation in traits with height at the community level (**Fig.
+S?**).
 When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5).
-Simultaneously in the 1999 univariate model, this co-consideration resulted in a non-signficant positive $ln[H]$-$Rt$ correlation (Table 4).
+*
 
-*Responses across droughts*
-
-Our results reject the hypothesis that overall community responses varied across droughts (*H3.1*).
-Within the context of mixed effects models, there were no significant differences in $Rt$ across drought years (Table 4).
-**see comment 36**
+*Responses across droughts* {**remove this section, but first see if there's anything that should be preserved.**}
 
@@
 
-Further, our findings support the hypothesis (*H3.3*) that the strength of predictor variables was different across the droughts (Tables 1,4,5).
  For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5).

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on LS comments
#65

@@
 
-Crown positions ($CP$) were recorded for all cored trees that remained standing during the growing season of 2018.
-We followed the crown position protocol of @jennings_assessing_1999, whereby positions were ranked as dominant, codominant, intermediate, or suppressed.
+Crown position ($CP$)--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
 While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
@@
 In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
-Variables were considered useful for predicting $Rt$ when AICc was reduced by $\ge$ 2 relative to the appropriate null model lacking that variable (Table 4).
+Variables were considered useful for predicting $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
 
@@
 
-*Focal droughts and their characteristics*
+*Focal droughts and their characteristics* {**Lawren thinks most of this (1st 3 par) belong in the methods, that par 3 might not be really useful**}
 
@@
 All droughts yielded about 5% more trees experiencing negative growth ($Rt$ $\le$ 0.7) as defined by our drought threshold, specifically, 29% in 1966, 32% in 1977, and 27% in 1999.
-However, some trees exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
+However, some individuals exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
 
@@
 
-Overall, our analysis confirmed the tendency for larger-diameter trees to show greater reductions in growth during drought [@bennett_larger_2015], although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
+Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
 The same held true for $ln[H]$ as a univariate predictor (*H1a*; Tables 1, 4).
-When combined with other predictor variariables in our multivariate models, the top models usually included an effect of $ln[H]$, and its coefficient was consistently negative, as predicted (Tables 1, 5).
-We note that a non-signficant positive correlation between $ln[H]$ and $Rt$ for 1999 became negative in the context of the multivariate models, again supporting *H1a* (Table 1).
+When combined with other predictor variables in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).
 
-Crown position was generally correlated with $H$, but with substantial variation (Fig.
+Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > supressed), but with substantial variation (Fig.
 2d).
@@
 
-Our results partially support Hypothesis *H2.1*, where species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $Rt$ in the univariate models (all dAIC $\le$ 0.22; Tables 1,4,5).
-Therefore, these were excluded as candidate variables for the full multivariate models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
+Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $Rt$ in the univariate models (Tables 1,4,5).
+$XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
 $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients.
-$\pi_{tlp}$ was not significant in the univariate models but had a consistently negative coefficient (Table 4).
+$\pi_{tlp}$ had no significant effect in the univariate models, but coefficients were consistently negative (Table 4).
 Whereas ring-porous species had highest $Rt$ overall and in the 1966 and 1999 droughts, they had lower $Rt$ in 1977.
-Results were similar in the context of multivariate models (Table 5).
-$PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).
+Results were similar in the context of multivariate models (Table 5), and $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).
 

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
remove hypothesis numbering
per LS suggestion
#65

@@
 First, we focus on the role of tree size and its interaction with microenvironment.
-We confirm that, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015].
-We then test hypotheses designed to disentangle the relative importance of tree height (*H1a*), crown exposure (*H1b*), and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites (*H1c*).
-Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits–particularly leaf hydraulic traits–predict Rt (*H2*).
+We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015].
+We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
+Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits–particularly leaf hydraulic traits–predict Rt.
 We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
@@
 Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
-In order to determine the relative importance of the traits alone (H2.1), we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
+In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
 Variables were considered useful for predicting $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
@@
 Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ as a univariate predictor (*H1a*; Tables 1, 4).
+The same held true for $ln[H]$ as a univariate predictor (Tables 1, 4).
 When combined with other predictor variables in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).
@@
 2d).
-Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to *H1b* (Table 1).
+Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to the hypothesis that trees with more exposed crowns have lower $Rt$ (Table 1).
 When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
@@
 Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction (*H1c*), which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
-*H1c* was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
+This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
 
@@
 
-Our findings give marginal support for Hypothesis *H3.2*, whereby we predicted directions of responses varied across droughts.
 In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
@@
 Our analysis does have the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
-However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of *H1.2*.
+However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
-**see comment 46** The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability (*H1.3*), agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
+**see comment 46** The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

### 2020-02-04 manuscript/tables&figures.Rmd
remove hypothesis numbering
per LS suggestion
#65

@@
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("1a. Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("1b. Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("1c. Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("2. Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-02-04 manuscript/tables&figures.Rmd
fixed error in table 1

@@
 kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
+Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') or rejected ('no') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
workign on results, discussion

@@
 Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
-[*Cite Shenandoah paper, if accepted*; @twery_effects_1991].
+[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
 
@@
 The same held true for $ln[H]$ as a univariate predictor (Tables 1, 4).
-When combined with other predictor variables in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).
+When combined with other predictor variables, including species' traits, in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).
 
@@
 
-Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $Rt$ in the univariate models (Tables 1,4,5).
-$XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-$PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients.
+Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses (Tables 1,4,5).
+In the *single-variable tests*, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from multivariable models.
+In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
+$PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4).
+Similarly $PLA_{dry}$ was consistently included, with negative coefficient, in multivariable models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
 $\pi_{tlp}$ had no significant effect in the univariate models, but coefficients were consistently negative (Table 4).
-Whereas ring-porous species had highest $Rt$ overall and in the 1966 and 1999 droughts, they had lower $Rt$ in 1977.
-Results were similar in the context of multivariate models (Table 5), and $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).
-
-*Combined effects of tree size and traits*
-
-*Our findings indicate that the greater growth reductions in large trees during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Table 5).
-Tthere was little meaningful variation in traits with height at the community level (**Fig.
-S?**).
-When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5).
-*
-
-*Responses across droughts* {**remove this section, but first see if there's anything that should be preserved.**}
-
-In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5).
-However, there were a few exceptions-most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5).
-These differences may be spurious as opposed to statistically meaningful.
-Among the univariate models, there was no instance where predictor variables significantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4).
-Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest predictor model of the six total models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4).
-It is probable that in the case of ring porosity, ring-porous species had lower $Rt$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5).
-Note, however, that $XP$ was not a signficant predictor on its own for the 1977 drought.
-
- For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5).
-**see comment 37**
+It was, however, included in the top multivariate model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+Xylem porosity was not signficant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
 
@@
 
-Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
+Tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
@@
 The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008.
-These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*).
+These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
 Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
@@
 
+Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].
+**see comment 49**
+
 Our analysis does have the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
@@
 The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
-At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.
-Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiolgically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015].
+{*At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.*}  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] {**Rosas et al NP about Spanish tree distributions; Fletcher et al AJB about Ceanothus**}.
 Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--Nate Swenson papers?
+there was a recent review in TREE**).
 
-**see comment 48; the first paragraph was condensed here, but unsure where this should be put now**
-Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height.
-Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.
-
-Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-Although there were not significant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig.
-1, Table 4).
-Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].
-**see comment 49** In 1977, which was the weakest drought in terms of PDSI (Fig.
-S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015.
-In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted (Fig.
-S2).
-Thus, responses differed with the nature of each drought.
-Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.
+*Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
+*
 
@@
 Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on discussion

@@
 
-Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].
-**see comment 49**
+*Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].* **see comment 49**
 
@@
 
-The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time.
-{*At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.*}  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] {**Rosas et al NP about Spanish tree distributions; Fletcher et al AJB about Ceanothus**}.
+The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015].
+Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
+2019, DOI: 10.1111/nph.15684*, [Fletcher et al.
+2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)).
 Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
@@
 
-*Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5).
-*
-
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
@@
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
+Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

@@
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
+Previous studies have demostrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
 2019, DOI: 10.1111/nph.15684*, [Fletcher et al.
-2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)).
-Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species.
+2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses.
+Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant that drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al.
+2016, DOI: 10.1111/gcb.13731*).
+Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species.
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--Nate Swenson papers?
-there was a recent review in TREE**).
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al.
+2016, DOI: 10.1111/gcb.13731*).
 

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on discussion

@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, which is consistent with analyses showing that suppressed--and particularly crowded--trees can suffer disproportionately during drought (**REFS- Alan?**).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (**REFS- Alan?**).
 **see comment 46** The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
moved drought descriptions to methods

@@
 
+*
+In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs.
+1, S2).
+We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
+Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
+[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
+
+Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
+Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively.
+The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
+
+The droughts differed in intensity and prior onset (Table S3, Fig.
+S2).
+1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season.
+The 1977 drought was the least intense throughout the growing season.
+All droughts increased in intensity from May through August, but differed in the previous months.
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
+Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
+The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.*
+
 *Statistical Analysis*
@@
 
-*Focal droughts and their characteristics* {**Lawren thinks most of this (1st 3 par) belong in the methods, that par 3 might not be really useful**}
-
-**Nobby questions using "useful predictor" as a term, instead using more jargon I think**
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs.
-1, S2).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
-Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
-[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
-
-Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
-Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
-
-The droughts differed in intensity and prior onset (Table S3, Fig.
-S2).
-1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season.
-The 1977 drought was the least intense throughout the growing season.
-All droughts increased in intensity from May through August, but differed in the previous months.
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
-Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
-The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
+*Community-level drought responses*
 
@@
 1b).
-All droughts yielded about 5% more trees experiencing negative growth ($Rt$ $\le$ 0.7) as defined by our drought threshold, specifically, 29% in 1966, 32% in 1977, and 27% in 1999.
+All droughts yielded *about 5% more trees experiencing negative growth ($Rt$ $\le$ 0.7) as defined by our drought threshold*, specifically, 29% in 1966, 32% in 1977, and 27% in 1999.
 However, some individuals exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
@@
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (**REFS- Alan?**).
-**see comment 46** The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
+The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
reverted confusing change introduced during incorporation of coauthor feedback

@@
 
-Community-level tree growth responses to these droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999, respectively (Fig.
+Community-level tree growth responses to all three droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig.
 1b).
-All droughts yielded *about 5% more trees experiencing negative growth ($Rt$ $\le$ 0.7) as defined by our drought threshold*, specifically, 29% in 1966, 32% in 1977, and 27% in 1999.
-However, some individuals exhibited increased growth: ($Rt$ > 1.0): 26% in 1966, 22% in 1977, and 26% in 1999.
+In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
+However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.
 

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
worked on drought descriptions

@@
 
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with widespread reductions in tree growth and anomalously dry peak growing season climatic conditions.
+We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth.
 Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community.
-Candidate drought years were defined as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011], and identified using the pointRes package [@R-pointRes] in R.
-Specifically, $Rt$ values $\le$ 0.7 indicated growth reductions, whereas values $\ge$ 1.0 indicated increased growth.
-Separately, we identified the years with driest conditions during May-August, which stood out in the analysis of [@helcoski_growing_2019] as the current-year months to which annual growth was most sensitive for trees at this site.
+
+We identified the years with driest conditions during May-August, which stood out in the analysis of @helcoski_growing_2019 as the current-year months to which annual growth was most sensitive for trees at this site.
 We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE).
-These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
+{**ADD SPEI**} These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
 The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
-
-*
-In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs.
-1, S2).
-We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3).
-Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
-[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
-
 Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
@@
 
-The droughts differed in intensity and prior onset (Table S3, Fig.
-S2).
+We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
+$Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate increased growth.
+Pointer years were identified using the pointRes package [@R-pointRes] in R.
+Four years met our criteria: 1966, 1977, 1991, and 1999.
+We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period.
+Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
+[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
+
+Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs.
+1, S2, Table S3).
+The droughts differed in intensity and prior onset (Fig.
+S2, Table S3).
 1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season.
@@
 Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
-The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.*
+The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
 

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
added Sohn ref

@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (**REFS- Alan?**).
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (*[Sohn et al.
+2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
 The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on text

@@
 
-One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought.
+One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought.
 @bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017].
@@
 1b).
-In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
+In each drought, roughly 30% of the cored trees experienced $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
 However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would suffer less during drought.
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would be less affected by drought.
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
@@
 
-Tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
+Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
-There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees.
-This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**).
-**see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments.
-**see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy.
-Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture.
-However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019].
+There was a marginal additional effect of crown exposure, with lowest $Rt$ among the most exposed (dominant) and suppressed trees.
+There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019].
 This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
 The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms.
-However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
 
-The droughts considered here were of similar severity (Fig.
-1b) and fairly moderate **see comment 41**; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig.
-1a, @helcoski_growing_2019).
-Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances.
-**see comment 42,43** While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
-1b), **see comment 44** supporting the observation that these droughts did not induce extreme stress on the entire forest.
-It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
-The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008.
-These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).
-Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016).
-This is important because while our results are consistent with findings from more extreme, multiannual droughts, these extremes have not occurred in the Eastern United States similar to the West, and therefore we do not know exactly how forests would respond under higher mortality conditions [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016].
-Given this consistency with our results, however, we believe this is an indication of a more intensive response.
+The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig.
+1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig.
+1b).
+While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
+1b), potentially due to decreased leaf area of competitors during the drought.
+It is likely because of the moderate severity of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
+While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]).
+Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts.
 

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
working on text/ coauthor feedback
issue #65

@@
 In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
-Variables were considered useful for predicting $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
+Variables were considered to have signficant influence on $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
 
-Once these beset predictor variables were determined, we calculated the best overall models for predicting $Rt$ across all traits, each individual drought year, and all years combined.
-We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc).
+We then determined the best multivariate models for predicting $Rt$ for each individual drought year, and for all years combined.
+Candidate variables were identified as those with dAICc $\ge$ 1.0 in at least one drought (Table 4).
+We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the top model (that with lowest AICc).
 These were counted as the "top multivariate models".
@@
 If the variable appeared in only some of the models, we considered this partial support.
-**see comment 27**
+
+**here and throughout, need clear and consistent way for referring to "univariate" and multivariate" models-- our usage was a sort of semantic shortcut and never technically correct**
 
@@
 
+**This paragraph needs work-- it is repetitive with intro**
 Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response.
@@
 2d).
-Even under non-drought conditions, evaporative demand increases with tree height while taller trees are more closely coupled to the atmosphere **see comment 45** (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
+Even under non-drought conditions, evaporative demand increases with tree height, and taller trees are more closely coupled to the atmosphere **see comment 45** (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
 Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
@@
 
-*Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019].* **see comment 49**
+*Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].*
 
-Our analysis does have the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
+Our analysis has the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
-The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees ([@suarez_factors_2004]; [@scharnweber_confessions_2019]).
+The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
 It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (*[Sohn et al.
@@
 
-### Acknowledgements **see comment 51**
+### Acknowledgements **See Bourg et al. (2013) for recommended acknowledgement language to add here for the initial census**
 Thanks to all researchers who helped to collect the data used here, in particular Jennifer C.

### 2020-02-07 manuscript/tables&figures.Rmd
fixing up Table 4
issue #74

@@
 univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(univar_table, booktabs = TRUE, caption = "**Table 4. Univariate models**", format = 'latex', escape = TRUE) %>%
+kable(univar_table, booktabs = TRUE, caption = "Table 4.
+Single-variable tests of hypothesized drivers of drought resistance.
+Models including each variable were compared to corresponding null models.
+dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
   add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work

### 2020-02-07 manuscript/tables&figures.Rmd
cleanup of Table 4 and others
#74

@@
 hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('yes') or rejected ('no') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
+kable(hypothesis_table, booktabs = TRUE, caption = "Table 1.
+Summary of hypotheses, corresponding specific predictions, and results.
+We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-  kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**") %>%
+  kable(table2, booktabs = TRUE, caption = "Table 2. Summary of variables.") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
@@
 
-kable(table3, booktabs=TRUE, caption = "**Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
+kable(table3, booktabs=TRUE, caption = "Table 3.
+Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
+Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%

### 2020-02-07 manuscript/tables&figures.Rmd
highlighted dAICc values >1 and >2 in Table 4

@@
 ```
+
 \clearpage
@@
 ```
+*dAICc > 1: variable qualified for inclusion in full model
+
+**dAICc > 2: statistically signficant, variable qualified for inclusion in full model
 

### 2020-02-08 manuscript/climate_sensitivity_draft.Rmd
fixed a couple errors
fixed univariate/ multivariate language

@@
 Thus, $Rt$<1 indicates reduced growth under drought.
-Analyses focused on testing the predictions presented in Table 1, with either $Rt$ or $H$ as the response variable.
+Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
 The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
@@
 
-We then determined the best multivariate models for predicting $Rt$ for each individual drought year, and for all years combined.
-Candidate variables were identified as those with dAICc $\ge$ 1.0 in at least one drought (Table 4).
-We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the top model (that with lowest AICc).
-These were counted as the "top multivariate models".
-When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models.
-If the variable appeared in only some of the models, we considered this partial support.
-
-**here and throughout, need clear and consistent way for referring to "univariate" and multivariate" models-- our usage was a sort of semantic shortcut and never technically correct**
+We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
+Candidate variabels were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought (Table 4).
+We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
+When a variable appeared in all top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the full models.
+If the variable appeared in only some of the models, we considered this partial support/rejection.
 
@@
 Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ as a univariate predictor (Tables 1, 4).
-When combined with other predictor variables, including species' traits, in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).
+The same held true for $ln[H]$ in single-variable tests (Tables 1, 4).
+When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two 1999 models (Tables 1, 5).
 
@@
 2d).
-Crown position was a much poorer predictor of $Rt$ than was $H$ (Table 4), lending little overall support to the hypothesis that trees with more exposed crowns have lower $Rt$ (Table 1).
+Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that trees with more exposed crowns have lower $Rt$ (Table 1).
 When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
-When considered in conjunction with $H$, $CP$ resulted as a significant predictor only for the 1999 drought, during which suppressed and then intermediate trees had the lowest $Rt$.
-Crown position was included in **almost half of the top models, with mixed results as to how $Rt$ varied with $CP$ (Table 5).
-Most commonly in these multivariate models, as in the univariate models (Table 4), the resistance of dominant trees was less than that of co-dominant trees but higher that of suppressed trees**.
-Although $CP$ was at times a useful predictor of $Rt$, it overall had a weak effect relative to that of $H$.
+When $H$ was included in the model, $CP$ was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
+Crown position was included in some of the full models (Table 5).
+In 1977, where $H$ was not included in the full model, dominant trees had the lowet $Rt$, and suppressed the highest.
+In contrast, in full models including both $H$ and $CP$ (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
 
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the hypothesis that trees in moist microsites would be less affected by drought.
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments that have a deeper water table.
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would be less affected by drought.
+Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
 This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
@@
 Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses (Tables 1,4,5).
-In the *single-variable tests*, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from multivariable models.
+In the single-variable tests, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
 In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
 $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4).
-Similarly $PLA_{dry}$ was consistently included, with negative coefficient, in multivariable models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
-$\pi_{tlp}$ had no significant effect in the univariate models, but coefficients were consistently negative (Table 4).
-It was, however, included in the top multivariate model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+Similarly $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
+$\pi_{tlp}$ did not come out signficant in any single-variable tests, but coefficients were consistently negative (Table 4).
+It was, however, included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
 Xylem porosity was not signficant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

### 2020-02-08 manuscript/tables&figures.Rmd
fixed univariate/ multivariate language

@@
 Summary of hypotheses, corresponding specific predictions, and results.
-We count predictions as fully supported ('yes') when the response is signficant in univariate models and included in all top multivariate models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but  the effect was not significant in all models.") %>%
+We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.") %>%
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
 multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(multivar_table, booktabs = TRUE, caption = "**Table 5.
-Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
+kable(multivar_table, booktabs = TRUE, caption = "Table 5.
+Summary of top full models for each drought instance.
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-02-08 manuscript/climate_sensitivity_draft.Rmd
working on discussion

@@
 **This paragraph needs work-- it is repetitive with intro**
-Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response.
+Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response, and that its effect holds when also accounting for species' traits.
 Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
 Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
-Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
-Taller trees also face different microenvironments (Fig.
+*Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].* Taller trees also face different microenvironments (Fig.
 2a-b), in part because they are more likely to be in dominant canopy positions (Fig.
 2d).
-Even under non-drought conditions, evaporative demand increases with tree height, and taller trees are more closely coupled to the atmosphere **see comment 45** (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
+Even under non-drought conditions, evaporative demand increases with tree height, and taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
 Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
@@
 *Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].*
+*In this section, be sure to acknowledge collinearity between CP and H*
 
-Our analysis has the limitation that canopy positions were re-cored in 2018 and undoubtedly changed for some trees since the 1960s, and we note that $CP$ became an increasingly poor predictor moving back in time from 1999 to 1966 (Table 4).
+Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts.
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
@@
 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
-The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall trees (dominant) are the most sensitive in mature forests.
+The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

### 2020-02-09 manuscript/climate_sensitivity_draft.Rmd
discussion

@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after correcting for height effects) is real, likely as a result of competition (*[Sohn et al.
+It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).

### 2020-02-10 manuscript/climate_sensitivity_draft.Rmd
working on discussion

@@
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought.
-@bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017].
-However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern.
+Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016), although there are exceptions [@zuleta_drought-induced_2017].
+It has yet to be resolved which of several potential underlying mechanisms drive this pattern.
 First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
@@
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
-There was a marginal additional effect of crown exposure, with lowest $Rt$ among the most exposed (dominant) and suppressed trees.
-There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+There was a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
+There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+The negative effect of height on $Rt$ held when also accounting for species' traits.
 Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019].
@@
 1b), potentially due to decreased leaf area of competitors during the drought.
-It is likely because of the moderate severity of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
-While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
+*Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), potentially because dry conditions in the two preceding years (Fig.
+S2) severely stressed a larger portion of trees (Fig.
+1b) and increased the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].*
+**While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]).
 Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts.
+**
 
-**This paragraph needs work-- it is repetitive with intro**
-Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response, and that its effect holds when also accounting for species' traits.
-Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
-*Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].* Taller trees also face different microenvironments (Fig.
-2a-b), in part because they are more likely to be in dominant canopy positions (Fig.
-2d).
-Even under non-drought conditions, evaporative demand increases with tree height, and taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]).
-Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
-Correlation between height and canopy position (Fig.
-2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment.
-However, significant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig.
-2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position for the three droughts evaluated in our study area (Tables 1,4,5).
-
-*Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].*
-*In this section, be sure to acknowledge collinearity between CP and H*
+Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5).
+This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@couvreur_water_2018; @roskilly_conflicting_2019] **REFS**.
+However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig.
+**2d**) makes it impossible to confidently partition causality.
+Taller trees are more likely to be in dominant canopy positions (Fig.
+**2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
+**2a-b**).
+Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
+2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
 
@@
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
+
+**orphan text**
+Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
+Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
 

### 2020-02-11 manuscript/tables&figures.Rmd
Changed file names per #74
Changed file names to be single variable and multivariable, as opposed to what we were using before ("univariate" and "multivariate")

@@
 library(kableExtra)
-univar_table <- read.csv("tables_figures/table4_univariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(univar_table, booktabs = TRUE, caption = "Table 4.
+singlevar_table <- read.csv("tables_figures/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
 Single-variable tests of hypothesized drivers of drought resistance.
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/table5_multivariate_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+multivar_table <- read.csv("tables_figures/table5_multivariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
 kable(multivar_table, booktabs = TRUE, caption = "Table 5.

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
worked on discussion

@@
 {**ask Lawren to check revisions to this paragraph**}
+
+**the following text needs to be integrated into the paragraph above (KAT)**
+Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
+Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
 
@@
 The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms.
-However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
@@
 1b).
-While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig.
+These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
+Of the droughts considered here, the 1966 drought, which was prededed by two years of dry conditions (Fig.
+S2), severely stressed a larger portion of trees (Fig.
+1b).
+It may be noteable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought strength [@bennett_larger_2015; @stovall_tree_2019].
+Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
 1b), potentially due to decreased leaf area of competitors during the drought.
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
-*Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), potentially because dry conditions in the two preceding years (Fig.
-S2) severely stressed a larger portion of trees (Fig.
-1b) and increased the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].*
-**While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]).
-Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts.
-**
 
@@
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
-
-**orphan text**
-Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
-Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
 

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
worked on intro

@@
 One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought.
-Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016), although there are exceptions [@zuleta_drought-induced_2017].
+Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
 It has yet to be resolved which of several potential underlying mechanisms drive this pattern.
-First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009].
-To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019].
-Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}.
-Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**).
+First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018].
+Meanwhile, tall trees require greater hydraulic efficiency, such that xylem conduit diameter increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
+Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
+Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
@@
 A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
-Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought.
-Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018].
 Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
-{**ask Lawren to check revisions to this paragraph**}
-
-**the following text needs to be integrated into the paragraph above (KAT)**
-Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018].
-Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
 
@@
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits have failed to explain drought responses (**REFS**), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+However, in other cases these traits have failed to explain drought responses, and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].
@@
 Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., [@lloret_components_2011;  @dorangeville_drought_2018] **REFS**), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
 The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
@@
 Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5).
-This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@couvreur_water_2018; @roskilly_conflicting_2019] **REFS**.
+This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
 However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig.

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
reviewed some references
#56

@@
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits have failed to explain drought responses, and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al.
+2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
reviewed wording on soil water availability
#66

@@
 A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
-Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.
+Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
 
@@
 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
-The observed height-sensitivity of $Rt$, together with the apparent lack of importance of soil water availability, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
+The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
@@
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
-This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
+This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
 Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.

### 2020-02-11 manuscript/climate_sensitivity_draft.Rmd
issue #68

@@
 We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE).
-{**ADD SPEI**} These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
+These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
 The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.

### 2020-02-12 scripts/plot_time_series.R
Added "Figure1a" to filename

@@ -54,7 +54,7 @@ pre_drought_years <- drought_years - 5
 
 # create plot ####
 if(save.plots)  {
-  png(paste0("manuscript/tables_figures/Time_series_for_each_species.png"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
+  png(paste0("manuscript/tables_figures/Figure1a_Time_series_for_each_species.png"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
 }
 
 # plot parameters ####

### 2020-02-12 scripts/vertical_height_neon.R
Changed "wind" to "wind speed" as per #62

@@ -26,7 +26,7 @@ library(ggplot2)
 dp <- data.frame("data" = c("SAAT", "wind", "biotemp", "RH", "SR"),
                  "id" = c("DP1.00002.001", "DP1.00001.001", "DP1.00005.001", "DP1.00098.001", "DP1.00014.001"),
                  "value" = c("tempSingleMean", "windSpeedMean", "bioTempMean", "RHMean", "difRadMean"),
-                 "xlabs" = c("Mean Air Temperature [°C]", "Wind [m/s]", "Mean Infrared Biological Temperature [°C]", "RH [%]", "Mean shortwave downward radiation [W/m^2]"))
+                 "xlabs" = c("Mean Air Temperature [°C]", "Wind speed [m/s]", "Mean Infrared Biological Temperature [°C]", "RH [%]", "Mean shortwave downward radiation [W/m^2]"))
 
 dp <- dp[c(2,4,1,3,5), ]
 

### 2020-02-12 scripts/graphs_for_manuscript.R
Reformatted script for clarity
Reordered sections and moved code containing the TLP and PLA plots to the bottom (since in #62 we decided to remove plots e and f from Figure 2).

@@ -5,22 +5,123 @@
 ######################################################
 library(ggplot2)
 
-#1. NEON vertical height profiles
+#1 Figure 1: Distribution of resistance values and time series
+## necessary packages ####
+library(ggplot2)
+library(png)
+library(grid)
+library(gridExtra)
+library(cowplot)
+
+##1a make density plot ####
+trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv")
+trees_all_sub$year <- as.character(trees_all_sub$year)
+
+#In each drought, roughly 30% of the cored trees
+#Find n trees that have resistance <= 0.7 per year: #% in 1966, #2% in 1977, and #% in 1999. #Do same thing for those that have resistance >1 for each year
+library(data.table)
+test <- as.data.table(trees_all_sub)
+
+test_reduct <- test[resist.value<=0.7,.(reduct = length(resist.value)), by=.(year)]
+by_year <- test[, .(total=length(resist.value)), by=.(year)]
+test_reduct$total_re <- by_year$total
+test_reduct$perc_re <- test_reduct$reduct/test_reduct$total_re
+
+test_growth <- test[resist.value>1,.(growth = length(resist.value)), by=.(year)]
+test_growth$total_gr <- by_year$total
+test_growth$perc_gr <- test_growth$growth/test_growth$total_gr
+
+test_full <- merge(test_reduct, test_growth)
+
+test[,.(ran = range(resist.value, na.rm=TRUE)), by=.(year)]
+test[,.(mean = mean(resist.value, na.rm=TRUE)), by=.(year)]
+
+Mode <- function(x) {
+   ux <- unique(x)
+   ux[which.max(tabulate(match(x, ux)))]
+}
+
+apply(d, 2, Mode)
+test[,.(mode = Mode(resist.value)), by=.(year)]
+
+#save as png so that way you're arranging only png
+png("manuscript/tables_figures/Figure1b_density_plot.png", res = 300, width = 150, height = 50, units = "mm", pointsize = 10)
+ggplot(trees_all_sub) +
+   aes(x = resist.value, fill = year) +
+   geom_density(adjust = 1L, alpha=.5) +
+   geom_vline(xintercept=1) +
+   scale_fill_hue(labels=c("1966", "1977", "1999")) +
+   labs(x="resistance value") +
+   theme_minimal()
+dev.off()
+
[…]

### 2020-02-12 manuscript/climate_sensitivity_SI.Rmd
insert Fig. S3
Updated file names to match figures

@@
 \newpage
-![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS2_location_cored_trees.png){width=500px}
+![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS1_location_cored_trees.png){width=500px}
 
 \newpage
-![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/drought_plot.jpg){width=600px}
+![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/figureS2_drought_plot.jpg){width=600px}
+
+\newpage
+![Figure S3: Changes in height by canopy position](tables_figures/figureS3_height_plot_analysis.png){width=600px}

### 2020-02-20 scripts/graphs_for_manuscript.R
Fixed issues with code and updated Figure 2 as per #62
This also addresses #76

@@ -68,11 +68,13 @@ dev.off()
 library(ggpubr)
 library(extrafont)
 library(rasterVis)
+library(RCurl)
+library(stringi)
 
 loadfonts(device="win") #to get TNR
 
-quantile(current_ht$height.m, c(.95), na.rm=TRUE) #95% quantile = 35.002m
-quant <- data.frame(yintercept = 35.0022, Lines = "95th percentile")
+quantile(current_ht$height.m, c(.95), na.rm=TRUE) #95% quantile = 38.34479m
+quant <- data.frame(yintercept = 38.34479, Lines = "95th percentile")
 
 ##2a. Get NEON plots ####
 source('scripts/vertical_height_neon.R', echo=TRUE)
@@ -91,12 +93,12 @@ for (i in seq(along=1:3)){
       geom_hline(aes(yintercept = yintercept), linetype = "longdash", quant) +
       annotate(geom="text", x=NEON_order_x[[i]], y=NEON_order_y[[i]], 
                label = NEON_order[[i]], fontface="bold", size=7)
-   if(i==3){
+   if(i==1){
       NEON_list[[i]] <-
          NEON_list[[i]] +
          theme(legend.title = element_blank(),
                legend.box = "vertical",
-               legend.position = c(0.8, 0.85),
+               legend.position = c(0.8, 0.25),
                legend.background = element_rect(fill=alpha("white", 0.01)),
                legend.text=element_text(size=12),
                legend.key.size=unit(4,"mm"),
@@ -104,13 +106,13 @@ for (i in seq(along=1:3)){
          )
    }
    
-   if(!i==3){
+   if(!i==1){
       NEON_list[[i]] <-
          NEON_list[[i]] +
          theme(legend.position = "none")
    }
    
-   if(!i==1){
+   if(i==2){
       NEON_list[[i]] <- 
          NEON_list[[i]] +
          theme(axis.title.y = element_blank(),
@@ -119,7 +121,7 @@ for (i in seq(along=1:3)){
    }
 }
 
-NEON <- ggarrange(NEON_list$wind_plot, NEON_list$RH_plot, NEON_list$SAAT_plot,  nrow=1, ncol=3, align="h")
+NEON <- ggarrange(NEON_list$wind_plot, NEON_list$RH_plot, NEON_list$SAAT_plot,  nrow=2, ncol=2, align="h")
 
 ##2c. Get the boxplots for 2018 heights (this is also prep for #3) ####
[…]

### 2020-02-20 scripts/vertical_height_neon.R

@@ -23,7 +23,7 @@ library(ggplot2)
 #the function "loadByProduct" will load the data into R and collapse into one df (within a list).
 #it will not download/store anything on the computer, but working with large dfs will run slowly. Hence, it is a good idea to look at the 30min avg first.
 
-dp <- data.frame("data" = c("SAAT", "wind", "biotemp", "RH", "SR"),
+dp <- data.frame("data" = c("SAAT", "2DWSD", "biotemp", "RH", "SR"),
                  "id" = c("DP1.00002.001", "DP1.00001.001", "DP1.00005.001", "DP1.00098.001", "DP1.00014.001"),
                  "value" = c("tempSingleMean", "windSpeedMean", "bioTempMean", "RHMean", "difRadMean"),
                  "xlabs" = c("Mean Air Temperature [°C]", "Wind speed [m/s]", "Mean Infrared Biological Temperature [°C]", "RH [%]", "Mean shortwave downward radiation [W/m^2]"))
@@ -46,15 +46,36 @@ for (i in seq(along=dp$value[1:3])){ #4 is biotemp and 5 is radiation (cloud vs 
   
   for (j in seq(along=years)){
     if (value != "RHMean" | j != 1){
-      neon_tower <- loadByProduct(dpID=dp$id[[i]], 
-                                  site=c("SCBI"),
-                                  package="basic", avg=30, 
-                                  check.size = FALSE, 
-                                  #(use TRUE outside loop to see how big the dowloads are)
-                                  startdate=paste0(years[[j]], "-05"),
-                                  enddate=paste0(years[[j]], "-08"))
       
-      neon_data <- neon_tower[[1]]
+      #this first if...else statement is due to an error with API. I've been told by 
+      #NEON (Claire Lunch) that it has been fixed and will be updated with the CRAN
+      #update by end March 2020.
+      #To change this back with that new update, simply have the "else" condition be the 
+      #code for everything.
+      if((j==2) | (i==3 & j==1)){ 
+        neon_tower <- loadByProduct(dpID=dp$id[[i]], 
+                                    site=c("SCBI"),
+                                    package="basic", 
+                                    check.size = FALSE, 
+                                    #(use TRUE outside loop to see how big the dowloads are)
+                                    startdate=paste0(years[[j]], "-05"),
[…]

### 2020-02-20 manuscript/tables&figures.Rmd
options(knitr.table.format = "latex")
this may be key to getting symbols in tables, issue #75

missed pushing changes to this file in the last commit
Made adjustments to tables-figures

@@
 options(tinytex.verbose = TRUE)
+options(knitr.table.format = "latex")
 ```
@@
 table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-  kable(table2, booktabs = TRUE, caption = "Table 2. Summary of variables.") %>%
+kable(table2, booktabs = TRUE, caption = "Table 2. Summary of variables.", 
+        escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
-  column_spec(2, italic = T) %>% 
+  #column_spec(2, italic = T) %>% 
   column_spec(4, width = "7cm") %>%

### 2020-02-20 manuscript/tables-figures.tex
Made adjustments to tables-figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{float}
\usepackage{booktabs}
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{caption}
\captionsetup[table]{font=small}
\captionsetup[figure]{font=small}
\captionsetup[table]{labelformat=empty}
\captionsetup[figure]{labelformat=empty}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[]{natbib}
\bibliographystyle{apalike}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\begin{landscape}\begin{table}[!h]

\caption{\label{tab:Table 1}Table 1. Summary of hypotheses, corresponding specific predictions, and results. We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{8.5}{10.5}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{12cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{2.4cm}}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Prediction supported?} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-5}
Hypotheses \& Specific Predictions & Overall & 1966 & 1977 & 1999 & Results\\
\midrule
\addlinespace[1em]
\multicolumn{1}{l}{\textbf{Tree size and microenvironment}}\\
\addlinespace[1em]
\multicolumn{1}{l}{\textit{Larger, taller trees have lower Rt.}}\\
\hspace{1em}\hspace{1em}Rt decreases with stem diameter (DBH). & yes & yes & - & - & Table 4\\
\hspace{1em}\hspace{1em}Rt decreases with height (H). & yes & yes & - & (yes) & Tables 4, 5\\
\addlinespace[1em]
\multicolumn{1}{l}{\textit{Trees with more exposed crowns have lower Rt.}}\\
\hspace{1em}\hspace{1em}Dominant trees have lowest Rt. & - & yes & (yes) & - & Tables 4, 5\\
\hspace{1em}\hspace{1em}Correcting for H, dominant trees have lowest Rt. & (no) &  & - & (no) & Tables 4, 5\\
\addlinespace[1em]
\multicolumn{1}{l}{\textit{Small trees (lower root volume) in drier microhabitats have lower Rt.}}\\
\hspace{1em}\hspace{1em}There is a negative interactive effect between H and topographic wetness index. & - & - & - & - & Table 4\\
\addlinespace[1em]
\multicolumn{1}{l}{\textbf{Species traits}}\\
\addlinespace[1em]
\multicolumn{1}{l}{\textit{Species' traits--particularly leaf hydraulic traits--predict Rt.}}\\
\hspace{1em}\hspace{1em}Wood density correlates (positively or negatively) to Rt. & - & - & - & - & Table 4\\
\hspace{1em}\hspace{1em}Leaf mass per area correlates positively to Rt. & - & - & - & - & Table 4\\
\hspace{1em}\hspace{1em}Ring-porous species have higher Rt than diffuse- or semi-ring- porous. & - & yes & (no) & yes & Tables 4, 5\\
\hspace{1em}\hspace{1em}Percent loss leaf area upon desiccation correlates negatively with Rt. & yes & yes & (yes) & - & Tables 4, 5\\
\hspace{1em}\hspace{1em}Water potential at turgor loss correlates negatively with Rt. & (yes) & - & (yes) & (yes) & Tables 4, 5\\
\bottomrule
\end{tabular}}
\end{table}
\end{landscape}

\clearpage

\begin{table}[!h]

\caption{\label{tab:Table 2}Table 2. Summary of variables.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lll>{\raggedright\arraybackslash}p{7cm}ll}
\toprule
variable & symbol & units & description & category & n\\
\midrule
\addlinespace[1em]
\multicolumn{1}{l}{\textbf{Dependent variable}}\\
\hspace{1em}drought resistance & Rt & - & ratio of growth during drought year to mean growth of the 5 years prior. & - & 1596\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textbf{Independent variables}}\\
\hspace{1em}drought year & Y & - & year of drought & 1966 & 478\\
\hspace{1em} &  &  &  & 1977 & 547\\
\hspace{1em} &  &  &  & 1999 & 571\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textit{tree size}}\\
\hspace{1em}\hspace{1em}diameter breast height & DBH & cm & DBH in drought year & - & all\\
\hspace{1em}\hspace{1em}height & H & m & estimated H in drought year & - & all\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textit{microhabitat}}\\
\hspace{1em}\hspace{1em}crown position & CP & - & 2018 crown position & dominant (D) & 31\\
\hspace{1em}\hspace{1em} &  &  &  & co-dominant (C) & 231\\
\hspace{1em}\hspace{1em} &  &  &  & intermediate (I) & 224\\
\hspace{1em}\hspace{1em} &  &  &  & suppressed (S) & 101\\
\hspace{1em}\hspace{1em}topographic wetness index & TWI & - & steady-state wetness index based on slope and upstream contributing area & - & all\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textit{species' traits}}\\
\hspace{1em}\hspace{1em}wood density & WD & g cm-3 & dry mass of a unit volume of fresh wood & - & all\\
\hspace{1em}\hspace{1em}leaf mass per area & LMA & kg m-2 & ratio of leaf dry mass to fresh leaf area & - & all\\
\hspace{1em}\hspace{1em}xylem porosity & XP & - & vessel arrangement in xylem & ring (R) & 408\\
\hspace{1em}\hspace{1em} &  &  &  & semi-ring (SR) & 31\\
\hspace{1em}\hspace{1em} &  &  &  & diffuse (D) & 178\\
\hspace{1em}\hspace{1em}turgor loss point & $\psi_{tlp}$ & MPa & water potential at which leaves wilt & - & all\\
\hspace{1em}\hspace{1em}percent loss area & PLA & % & percent loss of leaf area upon dessication & - & all\\
\bottomrule
\end{tabular}}
\end{table}
\clearpage

\begin{table}[!h]

\caption{\label{tab:Table 3}Table 3. Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.}
\centering
\resizebox{\linewidth}{!}{
\fontsize{8.5}{10.5}\selectfont
\begin{tabular}[t]{lrrrrrrlrr}
\toprule
species & percent.ANPP & n.cores & mean.DBH\_cm & DBH.range\_cm & WD\_g.per.cm3 & LMA\_g.per.cm2 & xylem.porosity & TLP\_Mpa & PLA\_percent\\
\midrule
Liriodendron tulipifera & 47.1 & 109 & 36.9 & 90.4 & 0.40 & 46.92 & diffuse & -1.92 & 19.56\\
Quercus alba & 10.7 & 66 & 47.2 & 67.7 & 0.61 & 75.80 & ring & -2.58 & 8.52\\
Quercus rubra & 10.1 & 71 & 54.9 & 136.9 & 0.62 & 71.13 & ring & -2.64 & 11.01\\
Quercus velutina & 7.8 & 83 & 54.1 & 98.2 & 0.65 & 48.69 & ring & -2.39 & 13.42\\
Quercus montana & 4.8 & 67 & 42.2 & 76.7 & 0.61 & 71.77 & ring & -2.36 & 11.75\\
\addlinespace
Fraxinus americana & 3.8 & 69 & 35.4 & 88.3 & 0.56 & 43.28 & ring & -2.10 & 13.06\\
Carya glabra & 3.7 & 39 & 31.4 & 88.7 & 0.62 & 42.76 & ring & -2.13 & 21.09\\
Juglans nigra & 2.1 & 31 & 48.1 & 62.8 & 1.09 & 72.13 & semi-ring* & -2.76 & 24.64\\
Carya cordiformis & 2.0 & 17 & 27.2 & 50.8 & 0.83 & 45.86 & ring & -2.13 & 17.22\\
Carya tomentosa & 2.0 & 18 & 21.0 & 20.1 & 0.83 & 45.36 & ring & -2.20 & 16.56\\
\addlinespace
Fagus grandifolia & 1.5 & 81 & 23.5 & 96.0 & 0.62 & 30.68 & diffuse & -2.57 & 9.45\\
Carya ovalis & 1.1 & 24 & 35.3 & 51.1 & 0.96 & 47.60 & ring & -2.48 & 14.80\\
\bottomrule
\end{tabular}}
\end{table}

*Semi-ring porosity is intermediate between ring and diffuse. We group
it with diffuse-porous species for more even division of species between
categories.

\clearpage
\begin{table}[!h]

\caption{\label{tab:table 4}Table 4. Single-variable tests of hypothesized drivers of drought resistance. Models including each variable were compared to corresponding null models. dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model}
\centering
\resizebox{\linewidth}{!}{
\fontsize{12}{14}\selectfont
\begin{tabular}[t]{llllrlrlrlr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{all droughts} & \multicolumn{2}{c}{1966} & \multicolumn{2}{c}{1977} & \multicolumn{2}{c}{1999} \\
\cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9} \cmidrule(l{3pt}r{3pt}){10-11}
﻿variable & category & null model variables & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients\\
\midrule
\addlinespace[1em]
\multicolumn{1}{l}{\textbf{Tree size and microenvironment}}\\
\hspace{1em}ln[DBH] &  & (year) & 8.17** & -0.0385 & 15.32** & -0.0888 & -0.87 & -0.0214 & -1.93 & 0.0057\\
\hspace{1em}ln[H] &  & (year) & 8.17** & -0.0600 & 15.32** & -0.1383 & -0.87 & -0.0334 & -1.93 & 0.0089\\
\hspace{1em}crown position & D & (year) & -2.96 & -0.0461 & 3.25** & -0.0509 & 0.66 & -0.0759 & 0.38 & -0.0103\\
\hspace{1em}(alone) & C &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
\hspace{1em} & I &  & - & -0.0063 & - & 0.0732 & - & -0.0298 & - & -0.0563\\
\hspace{1em} & S &  & - & 0.0122 & - & 0.0526 & - & 0.0432 & - & -0.0483\\
\hspace{1em}crown position & D & ln[H] (+year) & 0.57 & -0.0347 & -1.84 & -0.0328 & -0.23 & -0.0730 & 3.04** & -0.0024\\
\hspace{1em}(with height) & C &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
\hspace{1em} & I &  & - & -0.0425 & - & 0.0139 & - & -0.0388 & - & -0.0810\\
\hspace{1em} & S &  & - & -0.0582 & - & -0.0662 & - & 0.0258 & - & -0.0956\\
\hspace{1em}ln[TWI] &  & ln[H] (+year) & 5.34** & -0.0890 & -1.96 & -0.0171 & 5.05** & -0.1404 & 2.8** & -0.1033\\
\hspace{1em}ln[TWI]*ln[H] &  & ln[H]+ln[TWI] (+year) & -0.83 & 0.0797 & -1.58 & 0.0927 & -1.47 & 0.0861 & -1.9 & 0.0414\\
\addlinespace[1em]
\multicolumn{1}{l}{\textbf{Species traits}}\\
\hspace{1em}wood density &  & ln[H] (+year) & -1.91 & -0.0479 & -1.24 & -0.2089 & -1.22 & -0.1812 & 0.22 & 0.2502\\
\hspace{1em}leaf mass per area &  & ln[H] (+year) & -1.99 & 0.0003 & -1.88 & 0.0012 & -1.76 & -0.0013 & -2 & 0.0004\\
\hspace{1em}xylem porosity & R & ln[H] (+year) & -0.71 & 0.0660 & 2.305** & 0.1888 & 1.399* & -0.1452 & 3.765** & 0.1544\\
\hspace{1em} & D/SR &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
\hspace{1em}turgor loss point &  & ln[H] (+year) & 1.33* & -0.1777 & -1.64 & -0.1078 & 1.26* & -0.2500 & 0.016 & -0.1732\\
\hspace{1em}percent loss area &  & ln[H] (+year) & 7.17** & -0.0140 & 9.18** & -0.0249 & -0.05 & -0.0105 & -0.716 & -0.0074\\
\bottomrule
\end{tabular}}
\end{table}

*dAICc \textgreater{} 1: variable qualified for inclusion in full model

**dAICc \textgreater{} 2: statistically signficant, variable qualified
for inclusion in full model

\clearpage
\begin{table}[!h]

\caption{\label{tab:table 5}Table 5. Summary of top full models for each drought instance. Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).}
\centering
\resizebox{\linewidth}{!}{
\fontsize{12}{14}\selectfont
\begin{tabular}[t]{lrrrllllllllll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{4}{c}{crown position} & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{xylem architecture} & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){11-12}
﻿drought & dAICc & R2 & Intercept & ln[H] & D & C & I & S & ln[TWI] & D/SR & R & PLA & TLP\\
\midrule
\addlinespace[.4em]
\multicolumn{1}{l}{\textbf{}}\\
\textbf{\hspace{1em}all} & \textbf{0.000} & \textbf{0.12} & \textbf{1.077} & \textbf{-0.057} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.086} & \textbf{-} & \textbf{-} & \textbf{-0.012} & \textbf{-0.113}\\
\hspace{1em} & 0.586 & 0.11 & 1.365 & -0.055 & - & - & - & - & -0.086 & - & - & -0.013 & -\\
\hspace{1em} & 0.726 & 0.12 & 1.220 & -0.089 & -0.034 & 0 & -0.037 & -0.051 & -0.079 & - & - & -0.012 & -0.101\\
\hspace{1em} & 0.813 & 0.11 & 1.481 & -0.089 & -0.034 & 0 & -0.039 & -0.054 & -0.079 & - & - & -0.014 & -\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textbf{}}\\
\textbf{\hspace{1em}1966} & \textbf{0.000} & \textbf{0.25} & \textbf{1.503} & \textbf{-0.141} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{0} & \textbf{0.11} & \textbf{-0.021} & \textbf{-}\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textbf{}}\\
\textbf{\hspace{1em}1977} & \textbf{0.000} & \textbf{0.21} & \textbf{1.136} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.145} & \textbf{0} & \textbf{-0.205} & \textbf{-0.015} & \textbf{-0.13}\\
\hspace{1em} & 0.040 & 0.21 & 1.490 & - & - & - & - & - & -0.145 & 0 & -0.22 & -0.017 & -\\
\hspace{1em} & 0.505 & 0.22 & 1.089 & - & -0.069 & 0 & -0.025 & 0.043 & -0.137 & 0 & -0.199 & -0.014 & -0.143\\
\hspace{1em} & 0.818 & 0.22 & 1.481 & - & -0.07 & 0 & -0.027 & 0.038 & -0.136 & 0 & -0.216 & -0.017 & -\\
\addlinespace[.4em]
\multicolumn{1}{l}{\textbf{}}\\
\textbf{\hspace{1em}1999} & \textbf{0.000} & \textbf{0.23} & \textbf{0.464} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.095} & \textbf{0} & \textbf{0.16} & \textbf{-} & \textbf{-0.197}\\
\hspace{1em} & 0.019 & 0.24 & 0.725 & -0.068 & 0 & 0 & -0.077 & -0.09 & -0.084 & 0 & 0.167 & - & -0.183\\
\bottomrule
\end{tabular}}
\end{table}

\newpage

\begin{figure}
\centering
\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure1.png}
\caption{\textbf{Figure 1. Climate and species-level growth responses
over our study period, highlighting the three focal drougths (a) and
community-wide responses} Time series plot (a) shows peak growing season
(May-August) climate conditions and residual chronologies for each
species. Focal droughts are indicated by dashed lines, and shading
indicates the pre-drought period used in calculations of the resistance
metric. Figure modified from Helcoski \emph{et al.} (2019). Density
plots (b) show the distribution of resistance values for each drought.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure2.png}
\caption{\textbf{Figure 2. Height profiles in growing season climatic
conditions, tree heights by crown position, and leaf hydraulic traits}
The top row shows averages (\(\pm\) SD) of daily maxima and minima of
(a) wind speed, (b) relative humidity (\(RH\)), and (c) air temperature
(\(T_{air}\)) averaged over each month of the peak growing season
(May-August) from 2016-2018. In these plots, heights are slightly offset
for visualization purposes. Also shown are (d) 2018 tree heights by
canopy position (see Table 2 for codes) and vertical profiles in (e)
\(PLA_{dry}\) and (f) \(\pi_{tlp}\). In (e-f), values are community-wide
averages across height bins (plotted at upper end of height bin), with
grey indicating bins for which species-level trait measurements are
available for \textless75\% of individuals. In all plots, the dashed
horizontal line indicates the 95th percentile of tree heigts in the
ForestGEO plot.}
\end{figure}

  \bibliography{book.bib,packages.bib}

\end{document}

### 2020-02-20 manuscript/test.Rmd
Made adjustments to tables-figures

--- 2020-02-20 manuscript/tables&figures.Rmd
@@
 ---
-title: ""
-author: ""
-output:
-  pdf_document:
-    citation_package: natbib
-    fig_caption: yes
-bibliography: ["book.bib",  "packages.bib"]
-biblio-style: "apalike"
-csl: apa.csl
+title: "test"
+output: pdf_document
 header-includes: 
@@
 
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
-), 'packages.bib')
-```
-
-```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.
-We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.") %>%
-  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(1, width = "12cm") %>%
-  column_spec(2:5, width = "1.2cm") %>%
-  column_spec(6, width = "2.4cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kableExtra::landscape()
-```
-
-\clearpage
-
 ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
@@
         escape=FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
-  #column_spec(2, italic = T) %>% 
-  column_spec(4, width = "7cm") %>%
-  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()
 ```
-\clearpage
-
-```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table3 <- read.csv("tables_figures/table3_species_table.csv")
-
-kable(table3, booktabs=TRUE, caption = "Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
-Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
-```
-*Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories.
-
-\clearpage
-```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-singlevar_table <- read.csv("tables_figures/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
-kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
-Single-variable tests of hypothesized drivers of drought resistance.
-Models including each variable were compared to corresponding null models.
-dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
-  kable_styling()
-```
-*dAICc > 1: variable qualified for inclusion in full model
-
-**dAICc > 2: statistically signficant, variable qualified for inclusion in full model
-
-\clearpage
-```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-multivar_table <- read.csv("tables_figures/table5_multivariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(multivar_table, booktabs = TRUE, caption = "Table 5.
-Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 2," ", " ")) %>%
-  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 6, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 10, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  row_spec(1, bold = T) %>%
-  row_spec(5, bold = T) %>%
-  row_spec(6, bold = T) %>%
-  row_spec(10, bold = T) %>%
-  kable_styling()
-```
-
-\newpage
-
-![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}
-
-![**Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown are (d) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (e) $PLA_{dry}$ and (f) $\pi_{tlp}$.
-In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}

### 2020-02-20 manuscript/test.tex
Made adjustments to tables-figures

--- 2020-02-20 manuscript/tables-figures.tex
@@
 \hypersetup{
+  pdftitle={test},
   hidelinks,
@@
 \usepackage{booktabs}
-\usepackage{pdflscape}
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-\usepackage{caption}
-\captionsetup[table]{font=small}
-\captionsetup[figure]{font=small}
-\captionsetup[table]{labelformat=empty}
-\captionsetup[figure]{labelformat=empty}
-\usepackage{dcolumn}
 \usepackage{booktabs}
@@
 \usepackage{xcolor}
-\usepackage[]{natbib}
-\bibliographystyle{apalike}
 
+\title{test}
 \author{}
@@
 \begin{document}
+\maketitle
 
-\begin{landscape}\begin{table}[!h]
-
-\caption{\label{tab:Table 1}Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.
-We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.}
-\centering
-\resizebox{\linewidth}{!}{
-\fontsize{8.5}{10.5}\selectfont
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{12cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{1.2cm}>{\raggedright\arraybackslash}p{2.4cm}}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{4}{c}{Prediction supported?} & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){2-5}
-Hypotheses \& Specific Predictions & Overall & 1966 & 1977 & 1999 & Results\\
-\midrule
-\addlinespace[1em]
-\multicolumn{1}{l}{\textbf{Tree size and microenvironment}}\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textit{Larger, taller trees have lower Rt.}}\\
-\hspace{1em}\hspace{1em}Rt decreases with stem diameter (DBH). & yes & yes & - & - & Table 4\\
-\hspace{1em}\hspace{1em}Rt decreases with height (H). & yes & yes & - & (yes) & Tables 4, 5\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textit{Trees with more exposed crowns have lower Rt.}}\\
-\hspace{1em}\hspace{1em}Dominant trees have lowest Rt. & - & yes & (yes) & - & Tables 4, 5\\
-\hspace{1em}\hspace{1em}Correcting for H, dominant trees have lowest Rt. & (no) &  & - & (no) & Tables 4, 5\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textit{Small trees (lower root volume) in drier microhabitats have lower Rt.}}\\
-\hspace{1em}\hspace{1em}There is a negative interactive effect between H and topographic wetness index. & - & - & - & - & Table 4\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textbf{Species traits}}\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textit{Species' traits--particularly leaf hydraulic traits--predict Rt.}}\\
-\hspace{1em}\hspace{1em}Wood density correlates (positively or negatively) to Rt. & - & - & - & - & Table 4\\
-\hspace{1em}\hspace{1em}Leaf mass per area correlates positively to Rt. & - & - & - & - & Table 4\\
-\hspace{1em}\hspace{1em}Ring-porous species have higher Rt than diffuse- or semi-ring- porous. & - & yes & (no) & yes & Tables 4, 5\\
-\hspace{1em}\hspace{1em}Percent loss leaf area upon desiccation correlates negatively with Rt. & yes & yes & (yes) & - & Tables 4, 5\\
-\hspace{1em}\hspace{1em}Water potential at turgor loss correlates negatively with Rt. & (yes) & - & (yes) & (yes) & Tables 4, 5\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\end{landscape}
-
-\clearpage
-
-\begin{table}[!h]
+\begin{table}
 
@@
 \centering
-\resizebox{\linewidth}{!}{
-\fontsize{10}{12}\selectfont
-\begin{tabular}[t]{lll>{\raggedright\arraybackslash}p{7cm}ll}
+\begin{tabular}[t]{llllll}
 \toprule
@@
 \midrule
-\addlinespace[1em]
-\multicolumn{1}{l}{\textbf{Dependent variable}}\\
-\hspace{1em}drought resistance & Rt & - & ratio of growth during drought year to mean growth of the 5 years prior. & - & 1596\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textbf{Independent variables}}\\
-\hspace{1em}drought year & Y & - & year of drought & 1966 & 478\\
-\hspace{1em} &  &  &  & 1977 & 547\\
-\hspace{1em} &  &  &  & 1999 & 571\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textit{tree size}}\\
-\hspace{1em}\hspace{1em}diameter breast height & DBH & cm & DBH in drought year & - & all\\
-\hspace{1em}\hspace{1em}height & H & m & estimated H in drought year & - & all\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textit{microhabitat}}\\
-\hspace{1em}\hspace{1em}crown position & CP & - & 2018 crown position & dominant (D) & 31\\
-\hspace{1em}\hspace{1em} &  &  &  & co-dominant (C) & 231\\
-\hspace{1em}\hspace{1em} &  &  &  & intermediate (I) & 224\\
-\hspace{1em}\hspace{1em} &  &  &  & suppressed (S) & 101\\
-\hspace{1em}\hspace{1em}topographic wetness index & TWI & - & steady-state wetness index based on slope and upstream contributing area & - & all\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textit{species' traits}}\\
-\hspace{1em}\hspace{1em}wood density & WD & g cm-3 & dry mass of a unit volume of fresh wood & - & all\\
-\hspace{1em}\hspace{1em}leaf mass per area & LMA & kg m-2 & ratio of leaf dry mass to fresh leaf area & - & all\\
-\hspace{1em}\hspace{1em}xylem porosity & XP & - & vessel arrangement in xylem & ring (R) & 408\\
-\hspace{1em}\hspace{1em} &  &  &  & semi-ring (SR) & 31\\
-\hspace{1em}\hspace{1em} &  &  &  & diffuse (D) & 178\\
-\hspace{1em}\hspace{1em}turgor loss point & $\psi_{tlp}$ & MPa & water potential at which leaves wilt & - & all\\
-\hspace{1em}\hspace{1em}percent loss area & PLA & % & percent loss of leaf area upon dessication & - & all\\
+drought resistance & Rt & - & ratio of growth during drought year to mean growth of the 5 years prior. & - & 1596\\
+drought year & Y & - & year of drought & 1966 & 478\\
+ &  &  &  & 1977 & 547\\
+ &  &  &  & 1999 & 571\\
+diameter breast height & DBH & cm & DBH in drought year & - & all\\
+\addlinespace
+height & H & m & estimated H in drought year & - & all\\
+crown position & CP & - & 2018 crown position & dominant (D) & 31\\
+ &  &  &  & co-dominant (C) & 231\\
+ &  &  &  & intermediate (I) & 224\\
+ &  &  &  & suppressed (S) & 101\\
+\addlinespace
+topographic wetness index & TWI & - & steady-state wetness index based on slope and upstream contributing area & - & all\\
+wood density & WD & g cm-3 & dry mass of a unit volume of fresh wood & - & all\\
+leaf mass per area & LMA & kg m-2 & ratio of leaf dry mass to fresh leaf area & - & all\\
+xylem porosity & XP & - & vessel arrangement in xylem & ring (R) & 408\\
+ &  &  &  & semi-ring (SR) & 31\\
+\addlinespace
+ &  &  &  & diffuse (D) & 178\\
+turgor loss point & $\psi_{tlp}$ & MPa & water potential at which leaves wilt & - & all\\
+percent loss area & PLA & % & percent loss of leaf area upon dessication & - & all\\
 \bottomrule
-\end{tabular}}
-\end{table}
-\clearpage
-
-\begin{table}[!h]
-
-\caption{\label{tab:Table 3}Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
-Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.}
-\centering
-\resizebox{\linewidth}{!}{
-\fontsize{8.5}{10.5}\selectfont
-\begin{tabular}[t]{lrrrrrrlrr}
-\toprule
-species & percent.ANPP & n.cores & mean.DBH\_cm & DBH.range\_cm & WD\_g.per.cm3 & LMA\_g.per.cm2 & xylem.porosity & TLP\_Mpa & PLA\_percent\\
-\midrule
-Liriodendron tulipifera & 47.1 & 109 & 36.9 & 90.4 & 0.40 & 46.92 & diffuse & -1.92 & 19.56\\
-Quercus alba & 10.7 & 66 & 47.2 & 67.7 & 0.61 & 75.80 & ring & -2.58 & 8.52\\
-Quercus rubra & 10.1 & 71 & 54.9 & 136.9 & 0.62 & 71.13 & ring & -2.64 & 11.01\\
-Quercus velutina & 7.8 & 83 & 54.1 & 98.2 & 0.65 & 48.69 & ring & -2.39 & 13.42\\
-Quercus montana & 4.8 & 67 & 42.2 & 76.7 & 0.61 & 71.77 & ring & -2.36 & 11.75\\
-\addlinespace
-Fraxinus americana & 3.8 & 69 & 35.4 & 88.3 & 0.56 & 43.28 & ring & -2.10 & 13.06\\
-Carya glabra & 3.7 & 39 & 31.4 & 88.7 & 0.62 & 42.76 & ring & -2.13 & 21.09\\
-Juglans nigra & 2.1 & 31 & 48.1 & 62.8 & 1.09 & 72.13 & semi-ring* & -2.76 & 24.64\\
-Carya cordiformis & 2.0 & 17 & 27.2 & 50.8 & 0.83 & 45.86 & ring & -2.13 & 17.22\\
-Carya tomentosa & 2.0 & 18 & 21.0 & 20.1 & 0.83 & 45.36 & ring & -2.20 & 16.56\\
-\addlinespace
-Fagus grandifolia & 1.5 & 81 & 23.5 & 96.0 & 0.62 & 30.68 & diffuse & -2.57 & 9.45\\
-Carya ovalis & 1.1 & 24 & 35.3 & 51.1 & 0.96 & 47.60 & ring & -2.48 & 14.80\\
-\bottomrule
-\end{tabular}}
+\end{tabular}
 \end{table}
 
-*Semi-ring porosity is intermediate between ring and diffuse. We group
-it with diffuse-porous species for more even division of species between
-categories.
-
-\clearpage
-\begin{table}[!h]
-
-\caption{\label{tab:table 4}Table 4.
-Single-variable tests of hypothesized drivers of drought resistance.
-Models including each variable were compared to corresponding null models.
-dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model}
-\centering
-\resizebox{\linewidth}{!}{
-\fontsize{12}{14}\selectfont
-\begin{tabular}[t]{llllrlrlrlr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{all droughts} & \multicolumn{2}{c}{1966} & \multicolumn{2}{c}{1977} & \multicolumn{2}{c}{1999} \\
-\cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9} \cmidrule(l{3pt}r{3pt}){10-11}
-﻿variable & category & null model variables & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients & dAICc & coefficients\\
-\midrule
-\addlinespace[1em]
-\multicolumn{1}{l}{\textbf{Tree size and microenvironment}}\\
-\hspace{1em}ln[DBH] &  & (year) & 8.17** & -0.0385 & 15.32** & -0.0888 & -0.87 & -0.0214 & -1.93 & 0.0057\\
-\hspace{1em}ln[H] &  & (year) & 8.17** & -0.0600 & 15.32** & -0.1383 & -0.87 & -0.0334 & -1.93 & 0.0089\\
-\hspace{1em}crown position & D & (year) & -2.96 & -0.0461 & 3.25** & -0.0509 & 0.66 & -0.0759 & 0.38 & -0.0103\\
-\hspace{1em}(alone) & C &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
-\hspace{1em} & I &  & - & -0.0063 & - & 0.0732 & - & -0.0298 & - & -0.0563\\
-\hspace{1em} & S &  & - & 0.0122 & - & 0.0526 & - & 0.0432 & - & -0.0483\\
-\hspace{1em}crown position & D & ln[H] (+year) & 0.57 & -0.0347 & -1.84 & -0.0328 & -0.23 & -0.0730 & 3.04** & -0.0024\\
-\hspace{1em}(with height) & C &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
-\hspace{1em} & I &  & - & -0.0425 & - & 0.0139 & - & -0.0388 & - & -0.0810\\
-\hspace{1em} & S &  & - & -0.0582 & - & -0.0662 & - & 0.0258 & - & -0.0956\\
-\hspace{1em}ln[TWI] &  & ln[H] (+year) & 5.34** & -0.0890 & -1.96 & -0.0171 & 5.05** & -0.1404 & 2.8** & -0.1033\\
-\hspace{1em}ln[TWI]*ln[H] &  & ln[H]+ln[TWI] (+year) & -0.83 & 0.0797 & -1.58 & 0.0927 & -1.47 & 0.0861 & -1.9 & 0.0414\\
-\addlinespace[1em]
-\multicolumn{1}{l}{\textbf{Species traits}}\\
-\hspace{1em}wood density &  & ln[H] (+year) & -1.91 & -0.0479 & -1.24 & -0.2089 & -1.22 & -0.1812 & 0.22 & 0.2502\\
-\hspace{1em}leaf mass per area &  & ln[H] (+year) & -1.99 & 0.0003 & -1.88 & 0.0012 & -1.76 & -0.0013 & -2 & 0.0004\\
-\hspace{1em}xylem porosity & R & ln[H] (+year) & -0.71 & 0.0660 & 2.305** & 0.1888 & 1.399* & -0.1452 & 3.765** & 0.1544\\
-\hspace{1em} & D/SR &  & - & 0.0000 & - & 0.0000 & - & 0.0000 & - & 0.0000\\
-\hspace{1em}turgor loss point &  & ln[H] (+year) & 1.33* & -0.1777 & -1.64 & -0.1078 & 1.26* & -0.2500 & 0.016 & -0.1732\\
-\hspace{1em}percent loss area &  & ln[H] (+year) & 7.17** & -0.0140 & 9.18** & -0.0249 & -0.05 & -0.0105 & -0.716 & -0.0074\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-*dAICc \textgreater{} 1: variable qualified for inclusion in full model
-
-**dAICc \textgreater{} 2: statistically signficant, variable qualified
-for inclusion in full model
-
-\clearpage
-\begin{table}[!h]
-
-\caption{\label{tab:table 5}Table 5.
-Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).}
-\centering
-\resizebox{\linewidth}{!}{
-\fontsize{12}{14}\selectfont
-\begin{tabular}[t]{lrrrllllllllll}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } & \multicolumn{4}{c}{crown position} & \multicolumn{1}{c}{ } & \multicolumn{2}{c}{xylem architecture} & \multicolumn{1}{c}{ } & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){6-9} \cmidrule(l{3pt}r{3pt}){11-12}
-﻿drought & dAICc & R2 & Intercept & ln[H] & D & C & I & S & ln[TWI] & D/SR & R & PLA & TLP\\
-\midrule
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textbf{}}\\
-\textbf{\hspace{1em}all} & \textbf{0.000} & \textbf{0.12} & \textbf{1.077} & \textbf{-0.057} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.086} & \textbf{-} & \textbf{-} & \textbf{-0.012} & \textbf{-0.113}\\
-\hspace{1em} & 0.586 & 0.11 & 1.365 & -0.055 & - & - & - & - & -0.086 & - & - & -0.013 & -\\
-\hspace{1em} & 0.726 & 0.12 & 1.220 & -0.089 & -0.034 & 0 & -0.037 & -0.051 & -0.079 & - & - & -0.012 & -0.101\\
-\hspace{1em} & 0.813 & 0.11 & 1.481 & -0.089 & -0.034 & 0 & -0.039 & -0.054 & -0.079 & - & - & -0.014 & -\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textbf{}}\\
-\textbf{\hspace{1em}1966} & \textbf{0.000} & \textbf{0.25} & \textbf{1.503} & \textbf{-0.141} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{0} & \textbf{0.11} & \textbf{-0.021} & \textbf{-}\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textbf{}}\\
-\textbf{\hspace{1em}1977} & \textbf{0.000} & \textbf{0.21} & \textbf{1.136} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.145} & \textbf{0} & \textbf{-0.205} & \textbf{-0.015} & \textbf{-0.13}\\
-\hspace{1em} & 0.040 & 0.21 & 1.490 & - & - & - & - & - & -0.145 & 0 & -0.22 & -0.017 & -\\
-\hspace{1em} & 0.505 & 0.22 & 1.089 & - & -0.069 & 0 & -0.025 & 0.043 & -0.137 & 0 & -0.199 & -0.014 & -0.143\\
-\hspace{1em} & 0.818 & 0.22 & 1.481 & - & -0.07 & 0 & -0.027 & 0.038 & -0.136 & 0 & -0.216 & -0.017 & -\\
-\addlinespace[.4em]
-\multicolumn{1}{l}{\textbf{}}\\
-\textbf{\hspace{1em}1999} & \textbf{0.000} & \textbf{0.23} & \textbf{0.464} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-} & \textbf{-0.095} & \textbf{0} & \textbf{0.16} & \textbf{-} & \textbf{-0.197}\\
-\hspace{1em} & 0.019 & 0.24 & 0.725 & -0.068 & 0 & 0 & -0.077 & -0.09 & -0.084 & 0 & 0.167 & - & -0.183\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\newpage
-
-\begin{figure}
-\centering
-\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure1.png}
-\caption{\textbf{Figure 1. Climate and species-level growth responses
-over our study period, highlighting the three focal drougths (a) and
-community-wide responses} Time series plot (a) shows peak growing season
-(May-August) climate conditions and residual chronologies for each
-species. Focal droughts are indicated by dashed lines, and shading
-indicates the pre-drought period used in calculations of the resistance
-metric. Figure modified from Helcoski \emph{et al.} (2019). Density
-plots (b) show the distribution of resistance values for each drought.}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=5.20833in,height=\textheight]{tables_figures/Figure2.png}
-\caption{\textbf{Figure 2. Height profiles in growing season climatic
-conditions, tree heights by crown position, and leaf hydraulic traits}
-The top row shows averages (\(\pm\) SD) of daily maxima and minima of
-(a) wind speed, (b) relative humidity (\(RH\)), and (c) air temperature
-(\(T_{air}\)) averaged over each month of the peak growing season
-(May-August) from 2016-2018. In these plots, heights are slightly offset
-for visualization purposes. Also shown are (d) 2018 tree heights by
-canopy position (see Table 2 for codes) and vertical profiles in (e)
-\(PLA_{dry}\) and (f) \(\pi_{tlp}\). In (e-f), values are community-wide
-averages across height bins (plotted at upper end of height bin), with
-grey indicating bins for which species-level trait measurements are
-available for \textless75\% of individuals. In all plots, the dashed
-horizontal line indicates the 95th percentile of tree heigts in the
-ForestGEO plot.}
-\end{figure}
-
-  \bibliography{book.bib,packages.bib}
-
 \end{document}

### 2020-02-20 manuscript/tables&figures.Rmd
updated caption on Fig. 2

@@
 In these plots, heights are slightly offset for visualization purposes.
-Also shown are (d) 2018 tree heights by canopy position (see Table 2 for codes) and vertical profiles in (e) $PLA_{dry}$ and (f) $\pi_{tlp}$.
-In (e-f), values are community-wide averages across height bins (plotted at upper end of height bin), with grey indicating bins for which species-level trait measurements are available for <75% of individuals.
+Also shown is (d) 2018 tree heights by canopy position (see Table 2 for codes).
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
minor updates

@@
 
-While some canopy positions undoubtedly shifted in the # years between the 1966 drought and our observations in 2018, change was relatively slow, and bias would more likely result in type x error with regards to testing our hypothesis.
+While some canopy positions undoubtedly shifted in the 52 years between the 1966 drought and our observations in 2018, change was relatively slow, and bias would more likely result in type x error with regards to testing our hypothesis.
 
@@
 However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig.
-**2d**) makes it impossible to confidently partition causality.
+2d) makes it impossible to confidently partition causality.
 Taller trees are more likely to be in dominant canopy positions (Fig.
-**2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
-**2a-b**).
+2d) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
+2a-b).
 Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
update discussion on CP error

@@
 An analysis of crown position relative to height (Fig.
-2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions, although dominant and co-dominant trees were similar in height (Fig.
-S3).
-Specifically, average tree height growth was confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
-
-While some canopy positions undoubtedly shifted in the 52 years between the 1966 drought and our observations in 2018, change was relatively slow, and bias would more likely result in type x error with regards to testing our hypothesis.
+2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions (Fig.
+S3), with average tree height growth confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
+However, dominant and co-dominant trees were similar in height (Figs.
+2d, S3).
+While some canopy positions undoubtedly shifted in the 52 years between the 1966 drought and our observations in 2018, the resulting bias would be more likely result in type II error with regards to testing the hypothesis that dominant trees had lower resistance, meaning that our test is conservative.
 
@@
 Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
-[*Anderson-Teixeira et al., in review*; @twery_effects_1991].
+[@twery_effects_1991].
 

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
cut some references from intro
there were too many references for a couple statements in 1st par

@@
 
-*Key words*: annual growth; canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) .
-Words that are in the title can, and should, be among these.*
+*Key words*: annual growth; canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring
 
@@
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @cook_global_2014; @cook_unprecedented_2015; @mankin_mid-latitude_2019] [Dai et al.
-2018](https://link.springer.com/article/10.1007/s40641-018-0101-6) **{too many}**, often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@anderegg_consequences_2013; @adams_temperature_2009; @breda_temperate_2006; @hartmann_research_2018; @allen_underestimation_2015; @abrams_interdisciplinary_2016; @adams_multi-species_2017]**{too many refs here.
-reduce to 2-3}**.
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] [Dai et al.
+2018](https://link.springer.com/article/10.1007/s40641-018-0101-6), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
 Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 - We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree ring record spanning historical droughts.
-Thus, these traits can be useful for predicting future drought responses under climate change.
+Thus, these factors can be useful for predicting future drought responses under climate change.
 
@@
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] [Dai et al.
-2018](https://link.springer.com/article/10.1007/s40641-018-0101-6), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al.
+2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
 Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
@@
 It has yet to be resolved which of several potential underlying mechanisms drive this pattern.
-First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018].
+First, tree height may be a primary driver.
+Taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018].
 Meanwhile, tall trees require greater hydraulic efficiency, such that xylem conduit diameter increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
 Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
+Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
minor edits

@@
 While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
-The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, [@allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
+The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
 Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited.
@@
 First, we focus on the role of tree size and its interaction with microenvironment.
-We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in [@bennett_larger_2015].
+We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in @bennett_larger_2015.
 We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
-Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits–particularly leaf hydraulic traits–predict Rt.
+Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$.
 We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
@@
 
-Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019, and are archived in a public [GitHub repository](https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights).
+Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019.
 Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
@@
 2d, S3).
-While some canopy positions undoubtedly shifted in the 52 years between the 1966 drought and our observations in 2018, the resulting bias would be more likely result in type II error with regards to testing the hypothesis that dominant trees had lower resistance, meaning that our test is conservative.
 

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
edititing manuscript

@@
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
-Data for hydraulic traits are available at [**create new public GitHub repo for hydraulic traits, archive in Zenodo, give DOI**].
 
@@
 
-We identified the years with driest conditions during May-August, which stood out in the analysis of @helcoski_growing_2019 as the current-year months to which annual growth was most sensitive for trees at this site.
-We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration (PET) and precipitation (PRE).
+We identified the years with driest conditions during May-August (MJJA), which stood out in the analysis of @helcoski_growing_2019 as the current-year months to which annual growth was most sensitive for trees at this site.
+We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration ($PET$) and precipitation ($PRE$).
 These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
-The driest years were identified through ranking mean May-August PDSI or [PET-PRE] for the time period from driest to wettest.
-Climatically, these drought years included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August (MJJA), which for this site are when annual tree growth is most sensitive [@helcoski_growing_2019].
-Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83, 87, and 80 mm mo<sup>-1</sup>, respectively.
-The years 1964 and 2007 also ranked in the top five driest (PET-PRE =84 and 82 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
+The driest years were identified through ranking mean MJJA $PET-PRE$ or $PDSI$ for the time period from driest to wettest.
+Three of the five years between 1950 and 2009 with greatest moisture deficit ($PET-PRE$) during MJJA consistently ranked as the three driest in terms of $PDSI$: 1966, 1977, and 1999, which had mean MJJA $PET-PRE$ of 83, 87, and 80 mm mo<sup>-1</sup>, respectively (Table S3).
+The years 1964 and 2007 also ranked among the five lowest $PET-PRE$ (84 and 82 mm mo<sup>-1</sup>, respectively), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
 
-We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment (BAI) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
+We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
 $Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate increased growth.
@@
 Four years met our criteria: 1966, 1977, 1991, and 1999.
-We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period.
+We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period (Table S3).
 Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
@@
 
-Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs.
-1, S2, Table S3).
-The droughts differed in intensity and prior onset (Fig.
-S2, Table S3).
-1999 for example was the most intense (having the lowest PDSI) during May-July, while 1966 had a more intense August PDSI and overall had the second strongest drought during the growing season.
-The 1977 drought was the least intense throughout the growing season.
-All droughts increased in intensity from May through August, but differed in the previous months.
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall.
-Similarly, the 1999 drought was preceded by severe to extreme drought starting the previous fall, but before that conditions were wetter than average until the previous June.
-The 1977 drought was preceded by 2.5 years of near-normal conditions, and was therefore the mildest of the three droughts by any measure.
+Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and prior onset (Fig. S2, Table S3). 
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the minimum growing season $PDSI$ (-4.82) among of any of the three droughts.
+The 1977 drought was the least intense throughout the growing season, and was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
+The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July.
 

### 2020-02-20 manuscript/climate_sensitivity_draft.Rmd
spell check and updated word count

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1339 |	No of Supporting Information files | 	6
-Materials and Methods |	currently ~1604	||	 
-Results |	currently ~1558	 ||
-Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
-Acknowledgements | currently ~73	||
+Total word count (excluding summary, references and legends) |  currently ~5855 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
+Summary|  	currently 194 (limit 200) | 	No. of Tables | 	5  
+Introduction|	currently ~1468 |	No of Supporting Information files | 	6
+Materials and Methods |	currently ~2162	||	 
+Results |	currently ~674	 ||
+Discussion |	currently ~1551 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
+Acknowledgements | currently ~124	||
 
@@
 - Individual-level drought resistance decreased with tree height, which was the dominant size-related variable affecting drought response.
-Resistance was greater  among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon dessication.
+Resistance was greater  among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon desiccation.
 However, there was substantial variation in the best predictor variables across the three drought periods.
@@
 Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken.
-Rehydrated leaves taken towards the apical end of the branch (n=3 per indivdiual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
+Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.
 Leaf area was calculated from scanned images using the LeafArea R package [@R-LeafArea].
@@
 In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
-Variables were considered to have signficant influence on $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
+Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
 
 We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variabels were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought (Table 4).
+Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought (Table 4).
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
@@
 
-Community-level tree growth responses to all three droughts were modest, with mean resistance values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig.
+Community-level tree growth responses to all three droughts were modest, with mean $Rt$ values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig.
 1b).
@@
 
-Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
+Larger-diameter trees showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
 The same held true for $ln[H]$ in single-variable tests (Tables 1, 4).
@@
 
-Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > supressed), but with substantial variation (Fig.
+Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
 2d).
@@
 Crown position was included in some of the full models (Table 5).
-In 1977, where $H$ was not included in the full model, dominant trees had the lowet $Rt$, and suppressed the highest.
+In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest.
 In contrast, in full models including both $H$ and $CP$ (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
@@
 2a).
-Relative humidity was also somewhat lower during June-August, ranging from ~50-80 above the canopy and ~60-90% in the understory (Fig.
+Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
 2b).
@@
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never signficant and had a consistently positive coefficient (Table 4).
+This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).
 
@@
 
-Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses (Tables 1,4,5).
+Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
 In the single-variable tests, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
@@
 $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4).
-Similarly $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
-$\pi_{tlp}$ did not come out signficant in any single-variable tests, but coefficients were consistently negative (Table 4).
-It was, however, included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
-Xylem porosity was not signficant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
+Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
+$\pi_{tlp}$ did not come out significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
 
@@
 
-Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1).
+Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
@@
 1b).
-These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
-Of the droughts considered here, the 1966 drought, which was prededed by two years of dry conditions (Fig.
+These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts ("megadroughts") of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
+Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
 S2), severely stressed a larger portion of trees (Fig.
 1b).
-It may be noteable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought strength [@bennett_larger_2015; @stovall_tree_2019].
+It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought strength [@bennett_larger_2015; @stovall_tree_2019].
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
@@
 This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
-However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig.
+However, this result must be interpreted with some caution, given that collinearity between the two variables (Fig.
 2d) makes it impossible to confidently partition causality.
@@
 2a-b).
-Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
 However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees have always been suppressed, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
+It is safe to assume that currently suppressed trees have always been in the understory, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
@@
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
-Previous studies have demostrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
+Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
 2019, DOI: 10.1111/nph.15684*, [Fletcher et al.
@@
 
-### Acknowledgements **See Bourg et al. (2013) for recommended acknowledgement language to add here for the initial census**
-Thanks to all researchers who helped to collect the data used here, in particular Jennifer C.
+### Acknowledgements 
+We especially thank the numerous researchers who helped to collect the data used here, in particular Jennifer C.
 McGarvey, Jonathan R.
@@
 Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
-This study was funded by the Smithsonian's Forest Global Earth Observatory (ForestGEO), a Virginia Native Plant Society grant to KAT and AJT, and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.
+Funding for the establishment of the SCBI ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led Forest Global Earth Observatory (ForestGEO), the Smithsonian Institution, and the HSBC Climate Partnership.
+This study was funded by ForestGEO, a Virginia Native Plant Society grant to KAT and AJT, and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.
 

### 2020-02-20 manuscript/tables&figures.Rmd
updated table 2 symbols
#75

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 10, protect_latex = TRUE) %>%
-  #column_spec(2, italic = T) %>% 
   column_spec(4, width = "7cm") %>%

### 2020-02-20 manuscript/tables&figures.Rmd
tried and failed to use latex on Table 1 variables
#75
saving version of Table 1 with latex styling

@@
 We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
   column_spec(6, width = "2.4cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
   kableExtra::landscape()

### 2020-02-20 manuscript/climate_sensitivity_SI.Rmd
Made adjustments to tables-figures
SI cleanup (not yet complete)

@@
 
+\newpage
 ```{r Table S2, eval=TRUE, echo=FALSE, message=FALSE}
@@
 
+\newpage
 ```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
 tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other years referenced in the manuscript") %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other candidate drought years") %>%
   pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
@@
   pack_rows("", 21, 24, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-\newpage
-\clearpage
-
-```{r Table S5, eval=TRUE, echo=FALSE, message=FALSE}
-tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)
-
-tableS5 <- tableS5[,-1]
-names(tableS5) <- c("variable", "model", "coefficient", "p-value")
-tableS5[,c(3)] <- round(tableS5[,c(3)], 2)
-
-tableS5$variable <- c("WD", "LMA", "ring porosity", "PLA", "TLP")
-tableS5$model <- c("WD~ln[H]","LMA~ln[H]","ring porosity~ln[H]", "PLA~ln[H]", "PLA~ln[H]")
-
-kable(tableS5, booktabs=TRUE, format = "latex", caption = "Table S4. Correlation of species' traits with tree height across all individuals in the ForestGEO plot") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")

### 2020-02-29 manuscript/tables&figures.Rmd
fix to table 4 format

@@
   add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
   #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work

### 2020-02-29 manuscript/climate_sensitivity_draft.Rmd
updated table 2 symbols
#75
checked references were showing properly

@@
 
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo_2015].
+Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
 SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
@@
 $LMA$ was calculated as the ratio of leaf dry mass to fresh area.
-$PLA$ was calculated as the percent loss of area between fresh and dry leaves.
+$PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves.
 $WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.

### 2020-02-29 scripts/climate_sensitivity_analysis.R

@@ -434,8 +434,9 @@ trees_all_sub$dbh.cm <- exp(trees_all_sub$dbh.ln.cm)
 trees_all_sub$height.m <- exp(trees_all_sub$height.ln.m)
 
 stats <- NULL
-for(i in seq(along=trees_all_sub[,c(5,7:10,16,18,19)])){
-  name_trait <- trees_all_sub[,c(5,7:10,16,18,19)][i]
+for(i in seq(along=trees_all_sub[,c(5:9,16)])){
+  for(j in seq)
+  name_trait <- trees_all_sub[,c(5:9,16)][i]
   tree_stats <- 
     trees_all_sub %>%
     summarize(median = median(trees_all_sub[, which(colnames(trees_all_sub) == colnames(name_trait))]),

### 2020-02-29 manuscript/tables&figures.Rmd
Updated table3 as per #62

@@
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table_latex.csv", stringsAsFactors = FALSE, check.names = FALSE)
 kable(hypothesis_table, booktabs = TRUE, caption = "Table 1.

### 2020-02-29 scripts/tables_for_manuscript.R

@@ -1,7 +1,6 @@
 #tables for manuscript
 
-#Table 1 was originally made as .xslx ####
-#Table 2 ####
+#Table 3 ####
 library(RCurl)
 library(tidyr)
 library(data.table)
@@ -33,15 +32,15 @@ setnames(table2, old="<NA>", new="prior dead")
 table2$n_cores <- table1$n_cores[match(table2$sp, table1$sp)]
 table2 <- table2[, c(1,7,3,2,4:6)]
 table2 <- table2[!table2$sp %in% c("frni", "pist"), ]
+table1 <- table1[!table1$sp %in% c("frni", "pist"), ]
 
 #bring in trait values
 trees_all <- read.csv("manuscript/tables_figures/trees_all.csv", stringsAsFactors = FALSE)
 
-traits <- trees_all[,c(2,6:10)]
+traits <- trees_all[,c(4:9)]
 traits <- traits[!duplicated(traits$sp), ]
 
 table2 <- merge(table2, traits, by="sp")
-table2[,c(9:12)] <- round(table2[,c(9:12)], 2)
 
 #bring in mean DBH and DBH range for each species
 
@@ -50,18 +49,41 @@ dbh_cored <- species[,c(1:9)]
 dbh_cored <- dbh_cored %>%
   group_by(sp) %>%
   summarise(mean = mean(dbh2018),
-            range = max(dbh2018) - min(dbh2018))
+            min = min(dbh2018),
+            max = max(dbh2018))
 dbh_cored$mean <- round(dbh_cored$mean, 2)
+dbh_cored$range <- paste0(dbh_cored$min, " - ", dbh_cored$max)
+dbh_cored[,c("min", "max")] <- NULL
+
+dbh_cored <- dbh_cored[dbh_cored$sp != c("frni", "pist"), ]
 
 table2 <- merge(dbh_cored, table2, by = "sp")
+table2_SI <- table2[,c(5:9)]
+table2 <- table2[,-c(5:9)]
 
-setnames(table2, old=c("mean", "range", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), new=c("mean_DBH", "range_DBH", "RP", "PLA", "LMA", "TLP", "WD"))
+table2[,c(6:9)] <- round(table2[,c(6:9)], 2)
 
-table2_SI <- table2[,c(4:9)]
-table2 <- table2[,-c(4:9)]
+setnames(table2, 
[…]

### 2020-02-29 scripts/tables_for_manuscript.R
Added sd for traits #62

@@ -73,14 +73,29 @@ table2$percent.ANPP <- c(2, 3.7, 1.1,
                          2.1, 47.1, 10.7,
                          4.8, 10.1, 7.8)
 
+
+
+#get standard deviation
+org_traits <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEKK2MN6MSMV5QFIKS26MQDCM", stringsAsFactors = FALSE)
+
+table2$`$PLA$ (\\%)_sd` <- org_traits$PLA_dry_percent_sd[match(table2$sp, org_traits$sp)]
+table2$`$LMA$ ($\frac{g}{cm^2}$)_sd` <- org_traits$LMA_g_per_m2_sd[match(table2$sp, org_traits$sp)]
+table2$`$\\pi_{tlp}$ (Mpa)_sd` <- org_traits$mean_TLP_Mpa_sd[match(table2$sp, org_traits$sp)]
+table2$`$WD$ ($\frac{g}{cm^3}$)_sd` <- org_traits$WD_g_per_cm3_sd[match(table2$sp, org_traits$sp)]
+
+
 table2$sp <- c("Carya cordiformis (CACO)", "Carya glabra (CAGL)", "Carya ovalis (CAOVL)",
                "Carya tomentosa (CATO)", "Fagus grandifolia (FAGR)", "Fraxinus americana (FRAM)	",
                "Juglans nigra (JUNI)", "Liriodendron tulipifera (LITU)", "Quercus alba (QUAL)", 
                "Quercus montana (QUPR)", "Quercus rubra (QURU)", "Quercus velutina (QUVE)")
 
-table2 <- table2[,c(1,10,4,2:3,5:9)]
+
+table2 <- table2[,c(1,10,4,2:3,5,6,11,7,12,8,13,9,14)]
 table2 <- table2[base::order(table2$percent.ANPP, decreasing=TRUE), ]
 table2[8,6] <- "semi-ring*" #change juni ring porosity
+
+
+
 
 
 write.csv(table2, "manuscript/tables_figures/table3_species_table.csv", row.names=FALSE)

### 2020-03-02 manuscript/tables&figures.Rmd
working on tables

--- 2020-02-29 manuscript/tables&figures.Rmd
@@
 hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, booktabs = TRUE, caption = "Table 1.
+kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1.
 Summary of hypotheses, corresponding specific predictions, and results.
@@
 table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(table2, booktabs = TRUE, caption = "Table 2. Summary of variables.", 
+kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2. Summary of variables.", 
         escape=FALSE) %>%
@@
 
-kable(table3, booktabs=TRUE, caption = "Table 3.
+kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
 Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
@@
 Models including each variable were compared to corresponding null models.
-dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = TRUE) %>%
+dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-03-02 manuscript/tables&figures.Rmd
fixed Table 3 problem

@@
 library(kableExtra)
-table3 <- read.csv("tables_figures/table3_species_table.csv")
-
+table3 <- read.csv("tables_figures/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table3[8,6] <- paste0(table3[8,6], footnote_marker_symbol(1))
 kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
 Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
-Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE)
+Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.", escape=FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
+  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
+kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse.
+We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
 ```
-*Semi-ring porosity is intermediate between ring and diffuse. We group it with diffuse-porous species for more even division of species between categories.
 

### 2020-03-02 manuscript/tables&figures.Rmd
working on table 3

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," ", "DBH (cm)" = 2, " ")) %>%
   #kable_styling(bootstrap_options = "striped", full_width = FALSE)

### 2020-03-02 manuscript/tables&figures.Rmd
finished fixing up Table 3

--- 2020-03-02 manuscript/tables&figures.Rmd
@@
 table3 <- read.csv("tables_figures/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-table3[8,6] <- paste0(table3[8,6], footnote_marker_symbol(1))
+table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
 kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
   #kable_styling(bootstrap_options = "striped", full_width = FALSE)

### 2020-03-02 manuscript/tables&figures.Rmd
update table 3 header

@@
 kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
-Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.
-Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements.", escape=FALSE) %>%
+Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%

### 2020-03-02 manuscript/tables&figures.Rmd
symbols table 5

@@
 Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-03-02 manuscript/climate_sensitivity_draft.Rmd
#80

@@
 ### Summary
-- As climate change is driving increased drought in many forested regions around the world, mechanistic understanding of factors conferring drought resistance in trees is increasingly important.
+- As climate change is driving increased drought frequency and severity in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important.
 The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape tree growth responses during droughts.
 
-- We analyzed tree-ring records for twelve species of an oak-hickory forest - representing 97% of woody productivity - in the 25.6-ha ForestGEO plot in Virginia (USA) to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA).
+We sought to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
@@
 
-- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree ring record spanning historical droughts.
+- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree-ring record spanning historical droughts.
 Thus, these factors can be useful for predicting future drought responses under climate change.
@@
 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
-Droughts, exasperated by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
+Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
 Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and are not necessarily associated with ecological data.
+However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data.
 The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.
 
-One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought.
-Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
-It has yet to be resolved which of several potential underlying mechanisms drive this pattern.
+One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more severely affected by drought.
+Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
+It has yet to be resolved which of the potential underlying mechanisms drive this pattern.
 First, tree height may be a primary driver.
 Taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018].
-Meanwhile, tall trees require greater hydraulic efficiency, such that xylem conduit diameter increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), more negative water potential inducing 50% loss of hydraulic conductivity ($P50$), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018].
+Meanwhile, tall trees require greater hydraulic efficiency, such that the xylem conduit diameter at the base of the tree increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
 Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
+Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
-Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
 A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
-Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
+Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
 
-Debates have also arisen regarding the traits that would influence tree growth responses to drought.
+Debates have also arisen regarding the traits influencing tree growth responses to drought.
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
 However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017] but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
-Thus, their role may be due to indirect correlations with other traits within life-history strategies [@hoffmann_hydraulic_2011].
+2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
-Hydraulic traits including water potentials at which percent loss conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].

### 2020-03-02 manuscript/climate_sensitivity_draft.Rmd
#80

@@
 Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More rapidly measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
+More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
 These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].

### 2020-03-02 manuscript/climate_sensitivity_draft.Rmd
#80

@@
 
-The long time frame captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
-Tree growth responses vary with drought characteristics such as timing and atmospheric demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
+The long timeframe captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
+Tree growth responses vary with drought characteristics such as the timing of onset, antecedent moisture conditions, drought duration, and temperature-driven evaporative demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
 While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
-The ecological field-baesd studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
-Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals on the order of one to two decades - remains limited.
-There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be less in weaker--but more common--droughts.
+The ecological field-based studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
+Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited.
+There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be weaker in less severe--but more common--droughts.
 Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.
 
-To yield functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a 25.6-ha ForestGEO plot in Virginia (USA).
+Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA).
 First, we focus on the role of tree size and its interaction with microenvironment.
@@
 
-Research was conducted at the 25.6 ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
-SCBI is located in the central Appalachian Mountains at the northern edge of Shenandoah National Park.
-Elevations range from 273-338m above sea level with a topographic relief of 65m [@bourg_initial_2013].
+Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
+SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park.
+Elevations range from 273 to 338m above sea level with a topographic relief of 65m [@bourg_initial_2013].
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
@@
 
-All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
-
 Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 2).
@@
 From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data, which were last updated in 2019, are available through the ForestGEO data portal (www.forestgeo.si.edu).
+Census data are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
-We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species contributing most to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
-Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017 using a 5mm increment borer.
-In 2010-2011, cores were collected from randomly selected live trees of species with at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013].
-In 2016-2017, cores were collected from all trees found dead in the annual mortality census [@gonzalezakre_patterns_2016].
-Cores were sanded, measured, and cross-dated using standard procedures, as detailed in [@helcoski_growing_2019].
+We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
+Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017.
+In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013].
+In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016].
+Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019].
 The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).
@@
 
-$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}:r_{ring,2008}) - r_{bark,Y} + r_{bark,2008}\right]$$
+$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}) - r_{bark,Y} + r_{bark,2008}\right]$$
 
@@
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression equations on log-transformed data to relate bark thickness to diameter inside bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
+Specifically, we used linear regression on log-transformed data to relate bark thickness to diameter inside bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
 
-Height measurements ($H$, n=1518 trees) were taken by several researchers between 2012 to 2019.
+Height measurements ($H$, n=1,518 trees) were taken by several researchers for a variety of purposes between 2012 to 2019.
 Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
@@
 Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
-Measurements from the National Ecological Observatory Network (NEON) were collected nearby the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
+Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
 Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$).
-For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed using all height measurements.
+For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.
 
 Crown position ($CP$)--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
-While some trees undoubtedly changed position in the 52 years between the 1966 drought and our observations in 2018, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
+While some trees undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
 An analysis of crown position relative to height (Fig.
-2d) and height changes since the beginning of the study period indicated that changes were fairly small relative to differences among canopy positions (Fig.
+2d) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig.
 S3), with average tree height growth confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
@@
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
-TWI calculation depends on an input of a digital elevation model (DEM), and from this yields a quantitative assessment defined by how "wet" an area is, based on areas where run-off is more likely.
-From our observations in the plot, the calculation of TWI performed comparatively better at categorizing wet areas than the calculation of a distance matrix from a stream shapefile.
+TWI calculation depends on an input of a digital elevation model (DEM; **Ian, specify resolution**), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
+From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from a stream shapefile.
 
-Hydraulic traits were collected at SCBI (Table 3) in August 2018.
-We sampled small sun-exposed branches within eight meters of the ground from three individuals of each species in and around the ForestGEO plot.
+Hydraulic traits were collected in August 2018 (Table 3).
+We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot.
 Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken.
@@
 To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile spanning from 7.2 m height to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
+We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
 After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
@@
 
+All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI- ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).

### 2020-03-02 manuscript/climate_sensitivity_draft.Rmd
#80

@@
 
-*Identification of drought years*
+*Identifying drought years*
 
 We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth.
-Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of observed growth declines and that our focus remained on droughts that substantially impacted the forest community.
+Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of the observed growth declines and that our focus remained on droughts that substantially impacted the forest community.
 
-We identified the years with driest conditions during May-August (MJJA), which stood out in the analysis of @helcoski_growing_2019 as the current-year months to which annual growth was most sensitive for trees at this site.
+We identified the years with driest conditions during May-August (MJJA), which stood out in the analysis of @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
 We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration ($PET$) and precipitation ($PRE$).
 These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
-The driest years were identified through ranking mean MJJA $PET-PRE$ or $PDSI$ for the time period from driest to wettest.
-Three of the five years between 1950 and 2009 with greatest moisture deficit ($PET-PRE$) during MJJA consistently ranked as the three driest in terms of $PDSI$: 1966, 1977, and 1999, which had mean MJJA $PET-PRE$ of 83, 87, and 80 mm mo<sup>-1</sup>, respectively (Table S3).
-The years 1964 and 2007 also ranked among the five lowest $PET-PRE$ (84 and 82 mm mo<sup>-1</sup>, respectively), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
+The driest years were identified through by ranking mean MJJA PET-PRE or PDSI for the time period from driest to wettest.
+Three of the five years with greatest MJJA moisture deficit (PET-PRE) between 1950 and 2009 with greatest moisture deficit ($PET-PRE$) during MJJA consistently ranked as the three driest in terms of $PDSI$: 1966, 1977, and 1999, which had mean MJJA $PET-PRE$ of 83, 87, and 80 mm mo<sup>-1</sup>, respectively (Table S3).
+The years 1964 and 2007 also ranked among the top five in terms of moisture deficit lowest ($PET-PRE$ of 84 and 82 mm mo<sup>-1</sup>, respectively), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
 

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
finished incorporating Bill's & Alan's comments (...wjm_AT)

@@
 We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
-$Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate increased growth.
+$Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate above-average growth.
 Pointer years were identified using the pointRes package [@R-pointRes] in R.
@@
 We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period (Table S3).
-Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp.
-[@twery_effects_1991].
+Rather, the severity of growth reduction could probably be explained in part by defoliation by the gypsy moth (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp.
+in the area from approximately 1988 through 1995 [@twery_effects_1991].
 
-Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and prior onset (Fig. S2, Table S3). 
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the minimum growing season $PDSI$ (-4.82) among of any of the three droughts.
-The 1977 drought was the least intense throughout the growing season, and was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
-The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July.
+Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
+The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
+The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).
 
@@
 
-For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
 Thus, $Rt$<1 indicates reduced growth under drought.
@@
 
-Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently.
+Models were run for all drought years combined (with year as a fixed effect) and for each drought year individually.
 In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
-Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).
+Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 units relative to the corresponding null model lacking that variable (Table 4).
 
 We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought (Table 4).
+Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought year (Table 4).
 We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
-When a variable appeared in all top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the full models.
+When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models.
 If the variable appeared in only some of the models, we considered this partial support/rejection.
@@
 All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI- ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).
 
@@
 
-Community-level tree growth responses to all three droughts were modest, with mean $Rt$ values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig.
+Community-level tree growth responses to all three droughts were modest, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
 1b).
@@
 
-Larger-diameter trees showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4).
+Larger-diameter trees showed greater reductions in growth when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4).
 The same held true for $ln[H]$ in single-variable tests (Tables 1, 4).
-When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two 1999 models (Tables 1, 5).
+When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).
 
@@
 2d).
-Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that trees with more exposed crowns have lower $Rt$ (Table 1).
+Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
 When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
@@
 
-In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
+In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
 2).
@@
 
-Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
-In the single-variable tests, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4).
-$PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4).
+Hydraulic traits, including xylem porosity ($XP$), $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
+In the single-variable tests, $LMA$ and $WD$ were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
+In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
+$PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4).
 Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
-$\pi_{tlp}$ did not come out significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+$\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
 Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
@@
 Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1).
-The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019].
-There was a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
+The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019].
+We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held when also accounting for species' traits.
-Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019].
+The negative effect of height on $Rt$ held after accounting for species' traits.
+Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs.
+diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.
 This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
@@
 However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
-These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.
+These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al.
+2019, doi: 10.1093/treephys/tpz062**).
 
-The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig.
+The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig.
@@
 1b).
-It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought strength [@bennett_larger_2015; @stovall_tree_2019].
+It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees have always been in the understory, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
+It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
@@
 2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses.
-Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant that drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al.
+Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al.
 2016, DOI: 10.1111/gcb.13731*).
@@
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al.
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al.
 2016, DOI: 10.1111/gcb.13731*).
@@
 Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position.
-The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable.
+The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable.
 This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+Early- to mid- successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].
@@
 McGarvey, Jonathan R.
-Thomspon, and Victoria Meakem for original collection and processing of cores.
+Thompson, and Victoria Meakem for original collection and processing of cores.
 Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
@@
 ### Author Contribution 
-KAT, IM, and AT designed the research.
-Tree-ring chronologies were developed by RH under guidance of AT and NP.
+KAT, IM, and AJT designed the research.
+Tree-ring chronologies were developed by RH under guidance of AJT and NP.
 Trait data was collected by IM, JZ under guidance of NK and LS.

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
finished integrating comments in ...wjmEGAnp_AES

@@
 Joseph Zailaa^1,5^,
-Atticus Stovall^1,6,7^,
+Atticus E.L. Stovall^1,6,7^,
 Norman A. Bourg^1^,
@@
 
-Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
+Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al.
-2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015].
+2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
 Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
 Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-However, it has proved difficult to resolve factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data.
+However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data.
 The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011] and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits have failed to link to drought tolerance ([Maréchaux et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), and the direction of response is not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017].
+However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al.
+2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
@@
 Tree growth responses vary with drought characteristics such as the timing of onset, antecedent moisture conditions, drought duration, and temperature-driven evaporative demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
+While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018)([**Truettner et al**](https://doi.org/10.1016/j.foreco.2018.01.044)), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
 The ecological field-based studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
-Thus, our knowledge of forest responses to more modest but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited.
+Thus, our knowledge of forest responses to more moderate but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited.
 There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be weaker in less severe--but more common--droughts.
@@
 
-Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA).
+Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three drought (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA).
 First, we focus on the role of tree size and its interaction with microenvironment.
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), negating the idea that trees in moist microsites would be less affected by drought.
+Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), rejecting the idea that trees in moist microsites would be less affected by drought.
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected; the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).
+This hypothesis was rejected as the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).
 
@@
 
-Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1).
+Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019].
@@
 This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
-The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms.
+The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms.
 However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
@@
 1b).
-These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts ("megadroughts") of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
+These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
 Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
@@
 1b).
-It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
+It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
@@
 2a-b).
-Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
 However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
+Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
 

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
removed XP acronymn
#82

@@
 
-Hydraulic traits, including xylem porosity ($XP$), $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
+Hydraulic traits, including xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
 In the single-variable tests, $LMA$ and $WD$ were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
-In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
+In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
 $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4).

### 2020-03-03 manuscript/tables&figures.Rmd
removed XP acronymn
#82

--- 2020-03-02 manuscript/tables&figures.Rmd
@@
 Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = TRUE) %>%
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 2," ", " ")) %>%
+  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem porosity" = 2," ", " ")) %>%
   pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
removed CP acronymn
#82

@@
 
-Crown position ($CP$)--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
+Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
 While some trees undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
@@
 Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
-When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$.
-When $H$ was included in the model, $CP$ was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
+When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown positiono had the lowest $Rt$.
+When $H$ was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
 Crown position was included in some of the full models (Table 5).
 In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest.
-In contrast, in full models including both $H$ and $CP$ (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
+In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
 

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
reduced use of acronymns in text

@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017].
+Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017].
 However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher $WD$ has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
@@
 Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured $WD$ and $LMA$ [@medeiros_extensive_2019].
+More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured wood density and $LMA$ [@medeiros_extensive_2019].
 These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
@@
 
-Height measurements ($H$, n=1,518 trees) were taken by several researchers for a variety of purposes between 2012 to 2019.
+Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees).
 Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
@@
 $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves.
-$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.
+wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.
 We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
@@
 
-Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
+Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
 2d).
-Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
+Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
 When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown positiono had the lowest $Rt$.
-When $H$ was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
+When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
 Crown position was included in some of the full models (Table 5).
-In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest.
+In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest.
 In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
@@
 Hydraulic traits, including xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
-In the single-variable tests, $LMA$ and $WD$ were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
+In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
 In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
@@
 The negative effect of height on $Rt$ held after accounting for species' traits.
-Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs.
+Drought sensitivity was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.

### 2020-03-03 manuscript/tables&figures.Rmd
reduced use of acronymns in text

@@
 kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
-Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
+Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
+Variable abbreviations are as in Table 2.", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
@@
 Models including each variable were compared to corresponding null models.
-dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model", format = 'latex', escape = FALSE) %>%
+dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model).
+Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-03-03 manuscript/climate_sensitivity_SI.Rmd
SI cleanup
#82

@@
 options(knitr.table.format = "latex")
+options(knitr.kable.NA = '')
 ```
@@
 ### Supplementary Information
+## Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
+
+Ian R. McGregor,
+Ryan Helcoski,
+Norbert Kunert,
+Alan J. Tepley,
+Erika B. Gonzalez-Akre,
+Valentine Herrmann,
+Joseph Zailaa,
+Atticus E.L. Stovall,
+Norman A. Bourg,
+William J. McShea,
+Neil Pederson,
+Lawren Sack,
+Kristina J. Anderson-Teixeira
+
+\listoftables
+\listoffigures
+
+\newpage
 ```{r Table S1, eval=TRUE, echo=FALSE, message=FALSE}
@@
 kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other candidate drought years") %>%
-  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
   pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("others", 13, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("others", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
   pack_rows("", 17, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
@@
 \newpage
-![Figure S3: Changes in height by canopy position](tables_figures/figureS3_height_plot_analysis.png){width=600px}
+![Figure S3: Height by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/figureS3_height_plot_analysis.png){width=600px}

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
SI cleanup
#82

@@
 IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
+
+### Supplementary Information
+Table S1: Species-specific bark thickness regression equations
+
+Table S2: Species-specific height regression equations
+
+Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other
+candidate drought years
+
+Figure S1: Map of ForestGEO plot showing TWI and location of cored trees
+
+Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to
+each focal drought
+
+Figure S3: Height by canopy position across the three focal droughts and in the year of
+measurement (2018)

### 2020-03-04 manuscript/climate_sensitivity_draft.Rmd
integrated LS comments on intro. cut out 4th paragraph
LS recommended shortening/ streamlining. I decided 4th par really didn't add much, so cut.

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest (**[issue #84](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/84))
 
@@
 
-- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA).
-We sought to determine how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
@@
 Thus, these factors can be useful for predicting future drought responses under climate change.
+(**LS suggests take out "influence growth responses during drought, and".
+I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal**)
 
@@
 2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
-Droughts, intensified by climate change, have been affecting forests worldwide [@allen_global_2010], and are expected to continue as one of the most important drivers of forest change in the future [@allen_underestimation_2015].
-Understanding and modeling forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which rarely represent many species and size classes at a single site and rarely sampled in association with detailed ecological data.
-The aim of this study was to pair tree-ring and forest plot data to test how tree size and traits shaped growth responses to historical droughts.
+Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015].
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
+However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.
 
-One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more severely affected by drought.
+Many studies have shown that within species, large trees tend to be more affected by drought.
 Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
@@
 First, tree height may be a primary driver.
-Taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), more negative water potential inducing 50% loss of hydraulic conductivity ($P50$), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--make it biophysically possible for trees to become tall [@couvreur_water_2018].
-Meanwhile, tall trees require greater hydraulic efficiency, such that the xylem conduit diameter at the base of the tree increases with tree height within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
+Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity,more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Indeed, tall trees require wood of greater hydraulic efficiency in their basal portions, such that xylem conduit diameters are wider in taller trees within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
 Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because they tend to have more exposed crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
+Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
 Large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
-A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019].
+Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019].
 Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
+Differences have been observed between ring- and diffuse-porous species--with ring-porous species showing higher drought tolerance [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but this classification does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance drought tolerance ([Maréchaux et al.
+However, in other cases these traits could not explain drought tolerance ([Maréchaux et al.
 2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
@@
 Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function are emerging with potential to explain greater variation in plant distribution and function than the more commonly-measured wood density and $LMA$ [@medeiros_extensive_2019].
+More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
 These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
@@
 
-The long timeframe captured in tree-ring data enables us to address the question of whether tree size and species' traits have similar influence across different drought events, or whether that influence is more strongly predicted by community-level responses to variable drought severity, duration, and timing based on tree size and traits.
-Tree growth responses vary with drought characteristics such as the timing of onset, antecedent moisture conditions, drought duration, and temperature-driven evaporative demand [@dorangeville_drought_2018], but the question of how tree size and species' traits impact growth responses across droughts still remains.
-While tree-ring studies provide long-term records of tree responses to multiple droughts (e.g., @lloret_components_2011;  @dorangeville_drought_2018)([**Truettner et al**](https://doi.org/10.1016/j.foreco.2018.01.044)), they generally focus on species-level responses, and do not consider the roles of tree size and microenvironment.
-The ecological field-based studies that have shaped our understanding of the role of tree size and microenvironment in forest drought responses generally examine only a single drought and tend to focus disproportionately on the most extreme droughts with dramatic impacts (*e.g.*, @allen_global_2010; @bennett_larger_2015; @stovall_tree_2019; @anderegg_meta-analysis_2016).
-Thus, our knowledge of forest responses to more moderate but frequent droughts - e.g., those with historical return intervals of one to two decades - remains limited.
-There is evidence that the degree to which larger trees are more impacted by drought increases with the severity of drought conditions [@bennett_larger_2015; @stovall_tree_2019], so the influence of tree size may be weaker in less severe--but more common--droughts.
-Thus, while we expect many of the mechanisms shaping drought responses to be universal, we have little understanding of how tree size and traits interact with drought characteristics, and the extent to which their influence is consistent across droughts.
-
-Here, to yield a functional understanding of how tree size, microenvironment characteristics, and species' traits collectively shape drought responses, we test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three drought (1966, 1977, 1999), species functional and hydraulic trait measurements, and forest census data from a large forest dynamics plot in Virginia (USA).
+Here, we examine how tree size, microenvironment characteristics, and species' traits collectively shape drought responses.
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three drought (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia (USA).
 First, we focus on the role of tree size and its interaction with microenvironment.
@@
 Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$.
-We test predictions that drought resistance is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with specific leaf area, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
-We expect that ring-porous species would have greater drought resistance than semi-ring and diffuse-porous species, as has been previously observed [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009].
+We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
 

### 2020-03-04 manuscript/climate_sensitivity_draft.Rmd
worked on LS comments
got up to (not including) "Identifying drought years". comments that have been dealt with are highlighted in Word file.

#80

@@
 Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
-Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H]$).
+Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] ~ ln[DBH]$).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.
@@
 Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
-While some trees undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (i.e., type I error unlikely; type II error possible).
+While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative.
 An analysis of crown position relative to height (Fig.
@@
 TWI calculation depends on an input of a digital elevation model (DEM; **Ian, specify resolution**), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
-From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from a stream shapefile.
+From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.
 
@@
 We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot.
-Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets (opaque plastic bags) before measurements were taken.
+Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken.
 Rehydrated leaves taken towards the apical end of the branch (n=3 per individual: small, medium, and large) were scanned, weighed, dried at 60$^\circ$ C for $\ge$ 48 hours, and then re-scanned and weighed.

### 2020-03-05 manuscript/climate_sensitivity_SI.Rmd
removed discussion of MJJA PET-PRE as drought metric
this added complexity but did not change the content

@@
 tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other candidate drought years") %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for years with widespread growth reduction") %>%
   pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
@@
   pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("others", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
-  pack_rows("", 17, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 21, 24, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("other", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")

### 2020-03-05 manuscript/climate_sensitivity_draft.Rmd
removed discussion of MJJA PET-PRE as drought metric
this added complexity but did not change the content

@@
 
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth.
-Simultaneous consideration of both meteorological conditions and tree growth ensured that drought was the primary driver of the observed growth declines and that our focus remained on droughts that substantially impacted the forest community.
-
-We identified the years with driest conditions during May-August (MJJA), which stood out in the analysis of @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
-We considered two metrics of moisture deficit: NOAA Divisional Data’s Palmer Drought Severity Index (PDSI) and the difference between monthly potential evapotranspiration ($PET$) and precipitation ($PRE$).
-These data were obtained from the ForestGEO Climate Data Portal (https://github.com/forestgeo/Climate) in August 2018, with monthly PET and PRE sourced from Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; @harris_updated_2014).
-The driest years were identified through by ranking mean MJJA PET-PRE or PDSI for the time period from driest to wettest.
-Three of the five years with greatest MJJA moisture deficit (PET-PRE) between 1950 and 2009 with greatest moisture deficit ($PET-PRE$) during MJJA consistently ranked as the three driest in terms of $PDSI$: 1966, 1977, and 1999, which had mean MJJA $PET-PRE$ of 83, 87, and 80 mm mo<sup>-1</sup>, respectively (Table S3).
-The years 1964 and 2007 also ranked among the top five in terms of moisture deficit lowest ($PET-PRE$ of 84 and 82 mm mo<sup>-1</sup>, respectively), but were not among the lowest in terms of PDSI and were thus not identified as candidate years for inclusion as top drought years (Table S3).
+We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community.
+We identified three drought years: 1966, 1977, and 1999 (Figs.
+1, S2, Table S3).
+These were the years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017.
 
@@
 
-Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
+The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
 The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
@@
 
-Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other
-candidate drought years
+Table S3: Palmer drought severity index (PDSI) by month for years with widespread growth reduction
 

### 2020-03-05 manuscript/climate_sensitivity_draft.Rmd
condensed/ cleaned-up text on identifying drought
#80

@@
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017.
-
-We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
-$Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate above-average growth.
+These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
+$Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
 Pointer years were identified using the pointRes package [@R-pointRes] in R.
-Four years met our criteria: 1966, 1977, 1991, and 1999.
-We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period (Table S3).
-Rather, the severity of growth reduction could probably be explained in part by defoliation by the gypsy moth (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp.
+In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3).
+Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp.
 in the area from approximately 1988 through 1995 [@twery_effects_1991].

### 2020-03-05 manuscript/climate_sensitivity_draft.Rmd
integrated LS results comments
#80

@@
 
-Community-level tree growth responses to all three droughts were modest, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
+At the community-level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
 1b).
-In each drought, roughly 30% of the cored trees experienced $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
-However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.
+In each drought, roughly 30% of the cored trees had $Rt \le 0.7$: 29% in 1966, 32% in 1977, and 27% in 1999.
+However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
 
@@
 
-Larger-diameter trees showed greater reductions in growth when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4).
+Larger-diameter trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4).
 The same held true for $ln[H]$ in single-variable tests (Tables 1, 4).
@@
 2d).
+When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$.
 Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
-When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown positiono had the lowest $Rt$.
 When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
 Crown position was included in some of the full models (Table 5).
-In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest.
+In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest.
 In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
@@
 
-Resistance was negatively correlated with $ln[TWI]$ (Tables 4-5), rejecting the idea that trees in moist microsites would be less affected by drought.
+$Rt$ was negatively correlated with $ln[TWI]$ (Tables 4-5), rejecting the idea that trees in moist microsites would be less affected by drought.
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected as the $ln[H] *ln[TWI]$ interaction was never significant and had a consistently positive coefficient (Table 4).
+This hypothesis was rejected as the $ln[H] *ln[TWI]$ interaction was never significant (Table 4).
 

### 2020-03-05 manuscript/climate_sensitivity_draft.Rmd
finished integrating LS comments
#80

@@
 We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
+The negative effect of height on $Rt$ held after accounting for species' traits.
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held after accounting for species' traits.
 Drought sensitivity was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.
-This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses.
+This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
 The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms.
@@
 The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
-1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig.
+1a, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig.
 1b).
@@
 1b).
-It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
+The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
@@
 This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
-However, this result must be interpreted with some caution, given that collinearity between the two variables (Fig.
+However, the collinearity between the two variables (Fig.
 2d) makes it impossible to confidently partition causality.
@@
 Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests.
+The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
 Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al.

### 2020-03-05 manuscript/climate_sensitivity_draft.Rmd
integrated JZ comments
closes #80

@@
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
-Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin have enabled prediction of drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
@@
 SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park.
-Elevations range from 273 to 338m above sea level with a topographic relief of 65m [@bourg_initial_2013].
+Elevations range from 273 to 338 m above sea level with a topographic relief of 65m [@bourg_initial_2013].
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
@@
 Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017.
-In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10cm DBH [@bourg_initial_2013].
+In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013].
 In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016].
@@
 2d) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig.
-S3), with average tree height growth confined to ~0.82m from 1966 to 1977, ~1.45m from 1977 to 1999, and ~1.97m from 1999 to 2018.
+S3), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018.
 However, dominant and co-dominant trees were similar in height (Figs.
@@
 wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.
-We used the rapid determination method of @bartlett_rapid_2012 to estimate water potential at turgor loss point ($\pi_{tlp}$).
+We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$).
 Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
@@
 To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used data on wind speed, relative humidity, and air temperature, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
+We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
 After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
@@
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-Drought sensitivity was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.
+Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
+diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.
 This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
@@
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
-The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
+The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
 It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
@@
 Thompson, and Victoria Meakem for original collection and processing of cores.
-Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
+Thanks also to Camila D.
+Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
 Funding for the establishment of the SCBI ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led Forest Global Earth Observatory (ForestGEO), the Smithsonian Institution, and the HSBC Climate Partnership.

### 2020-03-07 manuscript/climate_sensitivity_draft.Rmd
Added citations for #78 and #77

@@
 # automatically create a bib database for R packages
+library(LeafArea)
 knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr', 'LeafArea'
-), 'packages.bib')
+  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea'
+), file = 'packages.bib')
 ```
@@
 (**LS suggests take out "influence growth responses during drought, and".
-I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal**)
+I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal** - *I agree with this*)
 
@@
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [Dai et al.
-2018](https://link.springer.com/article/10.1007/s40641-018-0101-6)), often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [@dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015].
@@
 Differences have been observed between ring- and diffuse-porous species--with ring-porous species showing higher drought tolerance [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but this classification does not resolve differences among the many species within each category.
-Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011](**Martin-Benito & Pederson 2015, doi:10.1111/jbi.12462**) and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance ([Maréchaux et al.
-2019](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2745.13321)), or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
+However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
@@
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
-TWI calculation depends on an input of a digital elevation model (DEM; **Ian, specify resolution**), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
+TWI calculation depends on an input of a digital elevation model (DEM; ~3.7 m resolution from the elevatr package [@R-elevatr]), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
 From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.
@@
 
-To characterize how environmental conditions vary with height, data were obtained from the National Ecological Observation Network (NEON) tower located <1km from the study area.
-We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@noauthor_national_2018].
+To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area.
+We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
 After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
@@
 All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively **krista just to remind you this TBD is here**).
 
@@
 However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
-These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al.
-2019, doi: 10.1093/treephys/tpz062**).
+These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
 
@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al.
-2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*).
+It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016].
 The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
-Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al.
-2019, DOI: 10.1111/nph.15684*, [Fletcher et al.
-2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses.
-Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al.
-2016, DOI: 10.1111/gcb.13731*).
+Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015; @rosas_adjustments_2019; @fletcher_evolution_2018], and our findings indicate that these traits also influence drought responses.
+Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017].
 Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species.
 The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al.
-2016, DOI: 10.1111/gcb.13731*).
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
 

### 2020-03-09 manuscript/climate_sensitivity_SI.Rmd
misc cleanup

@@
 tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for years with widespread growth reduction") %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts") %>%
   pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
misc cleanup

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest (**[issue #84](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/84))
+**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
 
@@
 
-- We conclude that hydraulic traits and tree height influence growth responses during drought, and can explain variation in the tree-ring record spanning historical droughts.
+- We conclude that hydraulic traits and tree height influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
 Thus, these factors can be useful for predicting future drought responses under climate change.
-(**LS suggests take out "influence growth responses during drought, and".
-I agree this conclusion is a bit repetitive/could be shortened, but would target the "explain variation..." for removal** - *I agree with this*)
 
@@
 All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base].
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively **krista just to remind you this TBD is here**).
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively.
 
@@
 
-Table S3: Palmer drought severity index (PDSI) by month for years with widespread growth reduction
+Table S3: Palmer drought severity index (PDSI) by month for focal droughts
 

### 2020-03-09 manuscript/climate_sensitivity_SI.Rmd
misc cleanup to conform to journal guidelines
#82

@@
 - \usepackage{caption}
+- \usepackage{setspace}\onehalfspacing
 - \captionsetup[figure]{labelformat=empty}

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
misc cleanup to conform to journal guidelines
#82

@@
 - \usepackage{setspace}\onehalfspacing
+- \usepackage{lineno}
+- \linenumbers
 - \usepackage{float}
@@
 
-We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1).
+We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig.
+**S1**).
 Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017.
@@
 An analysis of crown position relative to height (Fig.
-2d) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig.
-S3), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018.
+**2d**) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig.
+**S3**), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018.
 However, dominant and co-dominant trees were similar in height (Figs.
-2d, S3).
+**2d**, **S3**).
 
-Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel].
+Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Fig.
+**S1**) [@R-dynatopmodel].
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
@@
 We identified three drought years: 1966, 1977, and 1999 (Figs.
-1, S2, Table S3).
+**1**, **S2**, Table S3).
 These were the years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
@@
 
-The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
+The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
 The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
@@
 At the community-level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
-1b).
+**1b**).
 In each drought, roughly 30% of the cored trees had $Rt \le 0.7$: 29% in 1966, 32% in 1977, and 27% in 1999.
@@
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
-2d).
+**2d**).
 When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$.
@@
 In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-2).
+**2**).
 Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-2a).
+**2a**).
 Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
-2b).
+**2b**).
 Air temperature did not vary across the vertical profile (Fig.
-2c).
+**2c**).
 
@@
 The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
-1a, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig.
-1b).
+**1a**, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig.
+**1b**).
 These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
 Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
-S2), severely stressed a larger portion of trees (Fig.
-1b).
+**S2**), severely stressed a larger portion of trees (Fig.
+**1b**).
 The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), potentially due to decreased leaf area of competitors during the drought.
+**1b**), potentially due to decreased leaf area of competitors during the drought.
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
@@
 However, the collinearity between the two variables (Fig.
-2d) makes it impossible to confidently partition causality.
+**2d**) makes it impossible to confidently partition causality.
 Taller trees are more likely to be in dominant canopy positions (Fig.
-2d) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
-2a-b).
+**2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
+**2a-b**).
 Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
 However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
-2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
+**2d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
 Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

### 2020-03-09 manuscript/tables&figures.Rmd
added climate ref
misc cleanup to conform to journal guidelines
#82

@@
 - \usepackage{pdflscape}
+- \usepackage{setspace}\onehalfspacing
 - \newcommand{\blandscape}{\begin{landscape}}
@@
 \newpage
+### Figure Legends
+
+**Figure 1.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
+2014).
+Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from Helcoski *et al.* (2019).
+Density plots (**b**) show the distribution of resistance values for each drought.
+
+**Figure 2.
+Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (**d**) 2018 tree heights by canopy position (see Table 2 for codes).
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
+
+\newpage
 
@@
 Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
+2014).
 Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
minor changes in final review

@@
 
-Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
+Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
-In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014]; [@dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
+In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015].
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-However, it has proved difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.
+However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.
 
@@
 Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
-It has yet to be resolved which of the potential underlying mechanisms drive this pattern.
+It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response.
 First, tree height may be a primary driver.
@@
 Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity,more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Indeed, tall trees require wood of greater hydraulic efficiency in their basal portions, such that xylem conduit diameters are wider in taller trees within and across species [@olson_plant_2018; @liu_hydraulic_2019], making large trees more vulnerable to embolism during drought [@olson_plant_2018].
-Traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
+Indeed, tall trees require xylem of greater hydraulic efficiency in their basal portions, such that xylem conduit diameters are wider in taller trees within and across species [@olson_plant_2018; @liu_hydraulic_2019].
+Wider xylem conduits make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
 Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
-Large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Third, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019].
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Differences have been observed between ring- and diffuse-porous species--with ring-porous species showing higher drought tolerance [@kannenberg_linking_2019; @elliott_forest_2015; @friedrichs_species-specific_2009], but this classification does not resolve differences among the many species within each category.
+It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent; for instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
-Thus, their perceived influence on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
+However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent.
+For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
@@
 More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation ($PLA_{dry}$; [@scoffoni_leaf_2014]) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+These include leaf area shrinkage upon desiccation ($PLA_{dry}$) [@scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
@@
 Here, we examine how tree size, microenvironment characteristics, and species' traits collectively shape drought responses.
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three drought (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia (USA).
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
 First, we focus on the role of tree size and its interaction with microenvironment.
-We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in @bennett_larger_2015.
+We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of @bennett_larger_2015.
 We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
 Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$.
-We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, but that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
+We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
 
@@
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).
+The resulting chronologies were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.
 
@@
 Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees).
-Measurement methods included manual [@stovall_assessing_2018; @neon_national_2018], digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018], and automatic, ground-based LiDAR [@stovall_terrestrial_2018].
+Measurement methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018].
 Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
@@
 We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$).
-Briefly, two 4mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
+Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
@@
 We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
-After filtering for missing and outlier values, the data were consolidated to represent the mean values per sensor height per day.
-The range of these means were then aggregated at a month scale.
+After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.
 
@@
 **1**, **S2**, Table S3).
-These were the years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+These were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017.
 These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
-$Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Pointer years were identified using the pointRes package [@R-pointRes] in R.
+Pointer years were identified using the pointRes package in R [@R-pointRes].
 In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3).
@@
 
-For each drought year, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011].
-Thus, $Rt$<1 indicates reduced growth under drought.
+For each drought year, we calculated drought resistance ($Rt$) as the ratio of $BAI$ during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
 Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.
+The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+Mixed effects models were implemented in the lme4 package in R [@R-lme4].
+We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit, implmented in the AICcmodavg package in R [@R-AICcmodavg].
 
-Models were run for all drought years combined (with year as a fixed effect) and for each drought year individually.
-In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence.
+Models were run for all drought years combined and for each drought year individually.
+In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence).
 Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 units relative to the corresponding null model lacking that variable (Table 4).

### 2020-03-09 manuscript/tables&figures.Rmd
minor changes in final review

@@
 **Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
 In these plots, heights are slightly offset for visualization purposes.

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
working on final review

@@
 
-At the community-level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
+At the community level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
 **1b**).
-In each drought, roughly 30% of the cored trees had $Rt \le 0.7$: 29% in 1966, 32% in 1977, and 27% in 1999.
+In each drought, roughly 30% of the cored trees had growth reductions of >30% ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
@@
 
-Hydraulic traits, including xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5).
+THe leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5).
 In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
 In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
+Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
 $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4).
@@
 $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
-Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
 

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 
-THe leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5).
+The leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5).
 In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
Changed scoffoni citation #78

@@
 More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation ($PLA_{dry}$) [@scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+These include leaf area shrinkage upon desiccation ($PLA_{dry}$; @scoffoni_leaf_2014) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
finished final pass

--- 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
@@
 The negative effect of height on $Rt$ held after accounting for species' traits.
-There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
 Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
@@
 The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
-**1a**, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig.
+**1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Fig.
 **1b**).
@@
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016].
+It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016].
 The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
-Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015; @rosas_adjustments_2019; @fletcher_evolution_2018], and our findings indicate that these traits also influence drought responses.
+Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses.
 Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017].
-Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species.
-The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
+Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests.
+The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
 
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position.
+Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable.

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
redid word count
#82

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  currently ~5855 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
-Summary|  	currently 194 (limit 200) | 	No. of Tables | 	5  
-Introduction|	currently ~1468 |	No of Supporting Information files | 	6
-Materials and Methods |	currently ~2162	||	 
-Results |	currently ~674	 ||
-Discussion |	currently ~1551 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
-Acknowledgements | currently ~124	||
+Total word count (excluding summary, references and legends) |  5,365	 |	No. of figures | 	2 (both colour)
+Summary|  	198 | 	No. of Tables | 	5  
+Introduction|	1,034 |	No of Supporting Information files | 	6
+Materials and Methods |	1,945	||	 
+Results |	697	 ||
+Discussion |	1467	||
+Acknowledgements | 125	||
 

### 2020-03-21 non_publication/manuscript/original_bookdown_files/03-methods.Rmd
moved to non_publication folder (iii)
Files that are not being used in the final manuscript were moved into a non_publication folder as a holding step. Next step is dependent on each file, either delete, keep here or consolidate further.

--- 2019-07-30 manuscript/03-methods.Rmd
@@
+```{r include=FALSE, cache=FALSE}
+rm(list = ls(all = TRUE))
+```
 # Methods
@@
 
-species <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/master/data/core_list_for_neil.csv?token=AJNRBEIAVYFAN2CJB3Y3UAS5D5BNG"), stringsAsFactors = FALSE)
+species <- read.csv("D:/Dropbox (Smithsonian)/Github_Ian/McGregor_climate-sensitivity-variation/data/core_list_for_neil.csv", stringsAsFactors = FALSE)
 
@@
 traits_table <- data.frame(
-  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Chlorophyll", "Wood density", "TLP", "P50", "P80"),
-  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "m2/g", "g/cm3", "MPa", "MPa", "MPa"))
-traits_table$abb <- c("rp", "PLA", "LMA", "Chl", "WD", "TLP", "p50", "p80")
+  "Trait" = c("Ring Porosity", "Percent Leaf Area", "Leaf Mass Area", "Wood density", "TLP"),
+  "Unit" = c("ring, semi-ring, diffuse", "%", "g/m2", "g/cm3", "MPa"))
+traits_table$abb <- c("rp", "PLA", "LMA", "WD", "TLP")
 
@@
 
-for (i in seq(along=traits_table$abb[2:8])){
-  trait <- traits_table$abb[2:8][i]
+for (i in seq(along=traits_table$abb[2:5])){
+  trait <- traits_table$abb[2:5][i]
 
   sub <- trees_all[grepl(trait, colnames(trees_all))]
-  traits_table$mean[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+  traits_table$mean[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                               mean(sub[, 1], na.rm=TRUE), 
@@
 
-  traits_table$min[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+  traits_table$min[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                               min(sub[, 1], na.rm=TRUE), 
@@
 
-  traits_table$max[2:8][i] <- ifelse(grepl(trait, colnames(sub)), 
+  traits_table$max[2:5][i] <- ifelse(grepl(trait, colnames(sub)), 
                               max(sub[, 1], na.rm=TRUE), 
@@
 kable(traits_table) %>%
-  kable_styling(bootstrap_options = "striped", full_width = F)
+  kable_styling(bootstrap_options = "striped", full_width = FALSE)
 
@@
 
-```{r, all_candidates, eval=TRUE, echo=TRUE, message=FALSE}
+```{r, all_candidates, eval=TRUE, echo=FALSE, message=FALSE}
 candidates <- read.csv("tables_figures/candidate_traits.csv")

### 2020-03-21 non_publication/manuscript/original_bookdown_files/05-discussion.Rmd
moved to non_publication folder (iii)
Files that are not being used in the final manuscript were moved into a non_publication folder as a holding step. Next step is dependent on each file, either delete, keep here or consolidate further.

--- 2019-06-30 manuscript/05-discussion.Rmd
@@
 
-#### misc content to integrate
+### misc content to integrate
 From [@kannenberg_linking_2019], species with diffuse porous wood anatomy (*Liriodendron*) are more sensitive to drought, whereas ring-porous are not as affected because they more easily rebuild structures for hydraulic conductivity.

### 2020-03-21 non_publication/manuscript/original_bookdown_files/archive_methods.Rmd
moved to non_publication folder (iii)
Files that are not being used in the final manuscript were moved into a non_publication folder as a holding step. Next step is dependent on each file, either delete, keep here or consolidate further.

---
title: "archive_methods"
output: html_document
---
This document is meant to keep track of methods, even though we're not necessarily putting them in the paper.

## Sapwood ratio
Sapwood ratio also used the bark thickness data, and was also calculated for 2008, where
```
sapwood_area = total wood area (without bark) - heartwood area, and
sapwood_Ratio = sapwood area / total wood area (without bark)
```
Once sapwood area for each tree was calculated, log-log regression equations were determined on a per-species basis, which were then used

### 2020-03-21 non_publication/scripts/archive_graphs_for_manuscript.R

@@ -1,0 +1,478 @@
+###########################################################
+# Purpose: Archive figure and table script from "graphs_for_manuscript.R"
+# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
+# First written: July 2019
+###########################################################
+#5. Archive: Original Figure 2 with PLA and TLP plotted against height
+# These PLA and TLP plots were originally plots e and f of Figure 2. We have chosen
+# to not include these.
+
+## necessary packages ####
+library(ggstance)
+library(raster)
+library(dplyr)
+library(Hmisc)
+library(magrittr)
+
+
+##5a. get height data for all trees >10cm dbh in census ####
+scbi.stem3 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
+
+scbi.stem3$dbh <- as.numeric(scbi.stem3$dbh)
+scbi.stem3$dbh <- ifelse(is.na(scbi.stem3$dbh), 0, scbi.stem3$dbh)
+
+scbi18_ht <- scbi.stem3
+scbi18_ht <- scbi18_ht[scbi18_ht$dbh>=100, ]
+scbi18_ht <- scbi18_ht[,c(2:5,11,14:15)]
+
+scbi18_ht$dbh_old.cm <- scbi18_ht$dbh/10
+scbi18_ht$dbh.ln.cm <- log(scbi18_ht$dbh_old.cm)
+
+#linear log-log regression
+#use height_regr from above
+
+scbi18_ht$height.ln.m <- NA
+for(w in seq(along=height_regr$sp)){
+  sp_foc <- height_regr$sp[[w]]
+  ht_eq <- height_regr[height_regr$sp == sp_foc, ]
+  num <- gsub("\\*x", "", ht_eq$Equations)
+  num1 <- as.numeric(stri_extract_first_regex(num, "[[:digit:]].[[:digit:]]+"))
+  num2 <- as.numeric(stri_extract_last_regex(num, "[[:digit:]].[[:digit:]]+"))
+  
+  if(sp_foc %in% scbi18_ht[,"sp"]){
+    scbi18_ht$height.ln.m <- 
+      ifelse(scbi18_ht$sp == sp_foc,
+             num1 + num2*scbi18_ht$dbh.ln.cm,
[…]

### 2020-03-21 manuscript/climate_sensitivity_SI.Rmd
updated scripts with new file/folder locations

@@
 
-bark <- read.csv("tables_figures/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
+bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
 bark[,2] <- NULL
@@
 
-heights <- read.csv("tables_figures/tableS2_height_regression.csv", stringsAsFactors = FALSE)
+heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
 heights[,2] <- NULL
@@
 ```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
+tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
 kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts") %>%
@@
 \newpage
-![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/figureS1_location_cored_trees.png){width=500px}
+![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/publication/figureS1_location_cored_trees.png){width=500px}
 
 \newpage
-![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/figureS2_drought_plot.jpg){width=600px}
+![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/publication/figureS2_drought_plot.jpg){width=600px}
 
 \newpage
-![Figure S3: Height by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/figureS3_height_plot_analysis.png){width=600px}
+![Figure S3: Height by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}

### 2020-03-21 manuscript/tables&figures.Rmd
updated scripts with new file/folder locations

@@
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1.
@@
 library(kableExtra)
-table2 <- read.csv("tables_figures/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table2 <- read.csv("tables_figures/publication/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2. Summary of variables.", 
@@
 library(kableExtra)
-table3 <- read.csv("tables_figures/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table3 <- read.csv("tables_figures/publication/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
 table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
 kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/table5_multivariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
 kable(multivar_table, booktabs = TRUE, caption = "Table 5.
@@
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}
+Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png){width=500px}
 
@@
 Also shown is (d) 2018 tree heights by canopy position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/Figure2.png){width=500px}
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2.png){width=500px}

### 2020-03-21 non_publication/scripts/archive_climate_sensitivity.R

@@ -1,6 +1,6 @@
 ######################################################
 # Purpose: Original code for McGregor et al, kept here for reference
-# Created by: Ian McGregor - (mcgregorian93@gmail.com)
+# Created by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.2 - First created February 2020
 ######################################################
 
@@ -230,7 +230,7 @@ years_bysp <- years_bysp[order(years_bysp$year, years_bysp$sp), ]
 #write.csv(pointers, "data/occurrence_of_pointer_yrs.csv", row.names=FALSE)
 
 ##4b. resistance metrics for all trees ####
-neil_list <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+neil_list <- read.csv("data/core_files/core_list.csv", stringsAsFactors = FALSE)
 neil_sp <- unique(neil_list$sp)
 
 neil_list$tag <- paste0("X", neil_list$tag) #to match the colnames of can_resist below

### 2020-03-21 scripts/canopy_heights.R

@@ -1,6 +1,6 @@
 ######################################################
 # Purpose: Calculate past heights of different species in ForestGEO from regression equations
-# Developed by: Ian McGregor - mcgregori@si.edu
+# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.3 - First created April 2019
 ######################################################
 library(data.table)
@@ -150,7 +150,7 @@ check <- heights_all[is.na(heights_all$dbh) | heights_all$dbh ==0, ]
 
 #4. make regression equations ####
 #bring in list of cored species we're using
-neil_list <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+neil_list <- read.csv("data/core_files/core_list.csv", stringsAsFactors = FALSE)
 neil_sp <- unique(neil_list$sp)
 
 paper_heights <- heights_all[heights_all$sp %in% neil_sp, ]
@@ -195,7 +195,7 @@ regr <- data.frame("Species" = c("Carya cordiformis", "Carya glabra", "Carya ova
                              0.719,0.886, 0.755, 
                              0.857))
 
-write.csv(regr, "manuscript/tables_figures/tableS2_height_regression.csv", row.names=FALSE)
+write.csv(regr, "manuscript/tables_figures/publication/tableS2_height_regression.csv", row.names=FALSE)
 
 #########################################################################
 #create plots showing height diff by year (stand-alone) ####

### 2020-03-21 scripts/climate_sensitivity_analysis.R

@@ -1,6 +1,6 @@
 ######################################################
 # Purpose: Analysis of tree cores with relation to tree characteristics, in order to determine causes of drought susceptibility (using ForestGEO cores)
-# Developed by: Ian McGregor - mcgregori@si.edu
+# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.2 - First created February 2019, streamlined Feb. 2020
 ######################################################
 
@@ -45,7 +45,7 @@ pointers <- pointers[order(pointers$nb.series, decreasing=TRUE), ]
 rownames(pointers) <- 1:nrow(pointers)
 
 #get specific resist values for all trees
-neil_list <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+neil_list <- read.csv("data/core_files/core_list.csv", stringsAsFactors = FALSE)
 neil_sp <- unique(neil_list$sp)
 
 neil_list$tag <- paste0("X", neil_list$tag) #to match the colnames of can_resist below
@@ -318,7 +318,7 @@ trees_all$dbh.ln.cm <- log(trees_all$dbh_old.cm)
 ##2e. add in tree heights ####
 ## taken from the canopy_heights script
 #the full equation is using all points for which we have data to create the equation, despite that for several species we don't have enough data to get a sp-specific equation
-height_regr <- read.csv("manuscript/tables_figures/tableS2_height_regression.csv", stringsAsFactors = FALSE)
+height_regr <- read.csv("manuscript/tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
 
 height_regr$Equations <- gsub("^.*= ", "", height_regr$Equations)
 height_regr$Equations <- gsub("[[:alpha:]].*$", "x", height_regr$Equations)
@@ -375,7 +375,7 @@ trees_all$position_all <- gsub("S", "suppressed", trees_all$position_all)
 # trees_all$illum <- as.character(trees_all$illum)
 
 #this csv has avg/min/max dbh for each canopy position by sp
-# positionsp <- read.csv("data/core_chronologies_by_crownposition.csv")
+# positionsp <- read.csv("data/core_files/core_chronologies_by_crownposition.csv")
 
[…]

### 2020-03-21 scripts/graphs_for_manuscript.R

@@ -1,6 +1,6 @@
 ######################################################
 # Purpose: Create conglomerate graphs for McGregor et al 2019
-# Developed by: Ian McGregor - mcgregori@si.edu
+# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.3 - First created July 2019
 ######################################################
 library(ggplot2)
@@ -45,7 +45,7 @@ apply(d, 2, Mode)
 test[,.(mode = Mode(resist.value)), by=.(year)]
 
 #save as png so that way you're arranging only png
-png("manuscript/tables_figures/Figure1b_density_plot.png", res = 300, width = 150, height = 50, units = "mm", pointsize = 10)
+png("manuscript/tables_figures/publication/Figure1b_density_plot.png", res = 300, width = 150, height = 50, units = "mm", pointsize = 10)
 ggplot(trees_all_sub) +
    aes(x = resist.value, fill = year) +
    geom_density(adjust = 1L, alpha=.5) +
@@ -55,10 +55,10 @@ ggplot(trees_all_sub) +
    theme_minimal()
 dev.off()
 
-plot1 <- readPNG("manuscript/tables_figures/Figure1a_Time_series_for_each_species.png")
-plot2 <- readPNG("manuscript/tables_figures/Figure1b_density_plot.png")
+plot1 <- readPNG("manuscript/tables_figures/publication/Figure1a_Time_series_for_each_species.png")
+plot2 <- readPNG("manuscript/tables_figures/publication/Figure1b_density_plot.png")
 
-png("manuscript/tables_figures/Figure1.png", res=300, height=200, width=150, units="mm", pointsize=10)
+png("manuscript/tables_figures/publication/Figure1.png", res=300, height=200, width=150, units="mm", pointsize=10)
 plot_grid(rasterGrob(plot1), rasterGrob(plot2), align = "v", nrow = 2, rel_heights = c(3/4, 1/4), axis = "b")
 dev.off()
 
@@ -156,7 +156,7 @@ current_ht$dbh.ln.cm <- log(current_ht$dbh_old.cm)
[…]

### 2020-03-21 scripts/plot_time_series.R

@@ -54,7 +54,7 @@ pre_drought_years <- drought_years - 5
 
 # create plot ####
 if(save.plots)  {
-  png(paste0("manuscript/tables_figures/Figure1a_Time_series_for_each_species.png"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
+  png(paste0("manuscript/tables_figures/publication/Figure1a_Time_series_for_each_species.png"), res = 300, width = 150, height = 150, units = "mm", pointsize = 10)
 }
 
 # plot parameters ####

### 2020-03-21 scripts/tables_for_manuscript.R

@@ -1,12 +1,17 @@
-#tables for manuscript
-
+###############################################################
+# Purpose: Other tables for McGregor et al 2020
+# Developed by Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
+# First created November 2019
+################################################################
 #Table 3 ####
+
+# necessary packages
 library(RCurl)
 library(tidyr)
 library(data.table)
 library(dplyr)
 
-species <- read.csv("data/core_list_for_neil.csv", stringsAsFactors = FALSE)
+species <- read.csv("data/core_files/core_list.csv", stringsAsFactors = FALSE)
 
 table <- species[, c(1:3,6,17)]
 
@@ -98,7 +103,7 @@ table2[8,6] <- "semi-ring*" #change juni ring porosity
 
 
 
-write.csv(table2, "manuscript/tables_figures/table3_species_table.csv", row.names=FALSE)
-write.csv(table2_SI, "manuscript/tables_figures/TableS2.csv", row.names=FALSE)
+write.csv(table2, "manuscript/tables_figures/publication/table3_species_table.csv", row.names=FALSE)
+write.csv(table2_SI, "manuscript/tables_figures/publication/TableS2.csv", row.names=FALSE)
 
 

### 2020-03-21 scripts/vertical_height_neon.R

@@ -1,6 +1,6 @@
 ######################################################
 # Purpose: Create vertical profiles of climate variables for ForestGEO plot using NEON tower data
-# Developed by: Ian McGregor - mcgregori@si.edu
+# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # R version 3.5.3 - First created May 2019
 ######################################################
 library(neonUtilities)

### 2020-04-30 manuscript/2020_4_30_NewPhyt_review1/4_30_decision&reviews.Rmd
Create 4_30_decision&reviews.Rmd

External Email - Exercise Caution

NPH-MS-2020-32742 Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
by McGregor, Ian; Helcoski, Ryan; Kunert, Norbert; Tepley, Alan; Gonzalez-Akre, Erika; Herrmann, Valentine; Zailaa, Joseph; Stovall, Atticus; Bourg, Norman; McShea, William; Pederson, Neil; Sack, Lawren; Anderson-Teixeira, Kristina

Dear Dr Anderson-Teixeira

thank you for your patience. As you can see the reviewers had very divergent opinions on your manuscript. You address an exciting and urgently needed topic. This has been highlighted by the reviewers and I fully agree. However, there are worries about the conceptual foundation, statistical data analyses, data interpretation, etc, in addition to many other points which preclude acceptance of this work.
Therefore, I regret to inform you that your current study cannot be published in the New Phytologist. If you wish to re-work this paper, I invite a resubmission. If you decide to submit elsewhere, I am sure the careful comments of the reviewer will be useful  for you.

Thank you for giving us the opportunity to consider your work for the New Phytologist.

Sincerely,

Andrea Polle
Editor, New Phytologist

*****************************************************************
RESUBMISSION INSTRUCTIONS

Your manuscript has been rejected with encouragement to resubmit; please note that all resubmitted manuscripts are treated as new submissions and undergo the full review system. Also, New Phytologist does not usually permit multiple rounds of resubmission.

If you decide to resubmit your manuscript please click the below link
*** PLEASE NOTE: This is a two-step process. After clicking on the link, you will be directed to a webpage to confirm. ***

https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fmc.manuscriptcentral.com%2Fnewphytologist%3FURL_MASK%3D5bbd712fd98f48f5ac7ab9463d79d2c3&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=3oAHQBW9wA%2FeYQk1B1IjI5zwpJQiAzELTGj%2B0EyaM3o%3D&amp;reserved=0

Alternatively go to https://nam02.safelinks.protection.outlook.com/?url=http%3A%2F%2Fmc.manuscriptcentral.com%2Fnewphytologist&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=1PW5%2B7EmlwMEdX0WS%2FKLCZnsxc2SoOzj2GucLadhNlM%3D&amp;reserved=0 and log into your Author Center. Click on 'Manuscripts with Decisions', and then click on 'Create a Resubmission' located next to the ID number of this manuscript.

When preparing your resubmission please can you make sure that your comments for the attention of the Editor/Managing Editor, but not referees, are included in your 'Cover letter'. Your comments which specifically address the referees' reports, and may therefore need to be viewed by new referees, should be made in a separate Word file titled 'Response to referees' and uploaded when the actual manuscript files are uploaded. This will usually be a point by point response to the referees' reports.

In any correspondence regarding this manuscript, please include the manuscript reference number and copy the correspondence to New Phytologist Central Office (np-managinged@lancaster.ac.uk).
*****************************************************************

Decision: Reject, resubmission encouraged

Referee: 1

Comments to the Author
The study addresses the relationship between drought resistance (Rt) calculated from radial growth, and different traits and microenvironmental characteristics on 12 tree species from one forest in the Eastern US. Both the dataset and the study are meaningful, and the relationship found between Rt and height of biological importance. However, I think there are a number of issues that need to be worked out.
- It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought. This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript). In the paper, the authors use Rt similarly to most studies in the literature, which is fine. Yet, Rt (and particularly the other resilience indices based on radial growth) does not only depend on the individual response to drought events, but also on stand dynamics around target trees. For this reason, if stands around trees are not comparable or show changes after some disturbance (e.g. the drought events analysed) differently, this may bias comparisons of Resilience indices (although it is true that Rt should be less influenced than the other indices). The problem is that there is not such stand competition data for old droughts. This does not invalidate the study but I think the authors should discuss the issue, particularly for the analysis and comparisons of single droughts. I would suggest including also other Resilience indices, not only Rt, to enrich the discussion.
- Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis. Then I missed a further discussion on the role of individual species (e.g. see my comment below on Rt>1).
I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.
Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models. LL ratio tests or deviance-anova tests can be used in the case of nested models with LMM or GLMM respectively. Along the text: dAIC is generally referred as ΔAIC, please change accordingly.
Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included. And I suggest discussing only the model with all years together. This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation. This would help to ensure a better use of the Resilience indices.
The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis. If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.
- Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al. included). Droughts must be selected based only on some climatic/drought index (PDSI, SPEI,…) and independently of growth. Otherwise the analysis is biased towards years with low growth, and different questions would arise with a difficult answer. If the authors select only those droughts where there is a growth decrease: is that decrease evident in all species and similar across species? What happens with more intense droughts that those selected (if any), why they do not force a decrease in growth if drought is a main forcing agent? I see that 1966 and 1999 are the driest (as in PDSI) in Figure 1, but there is at least one year around 1955 that it is as dry or drier than 1977. In line 204 it is stated that they are the driest. If this is the case, the analysis should be sound but it should be deleted what is stated in lines 202-203 where authors say that they selected years also based on low growth. In fact, that should be a result after having selected the drought years only based on dryness (as in lines 208-209). Line 211 then they search for pointer years… a little bit messy. Please, explain better, and it should follow: 1) selection of dry years without regarding growth data; 2) characterise growth those dry years; 3) If the authors want to compare with other pointer years, fine, but that should be a third independent step properly justified and explained. Following this rationale, then in line 221 it should be difficult that Rt>1, had we selected those years with a negative growth anomaly. Please, explain clearly and consistently.
Other minor comments:
- Line 113: correct ‘positively’.
- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.
- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This should be rather straightforward and the authors would increase accuracy across species in height estimation.
- Lines 166-167: please explain further the hypothesis and why.
- Line 222: what about climate in years before 1991?
- Line 231: define AICc.
- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).
- Line 241-242: enough to say once that models were selected based on AIC (line 237-238) and then one criterion, why dAIC<1 here and dAIC<2 before? The classic assumption is 2 (e.g. Burnham & Anderson 2004). See also my other comment above.
- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.
- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?
- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.
- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest. They are indeed interesting to characterise the forest, but they do not add to individual tree variability. You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.
- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?
-Line 331-333: not so sure about this. It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).
- Line 348: non-drought.
- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.

Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
Table 3: the authors could show some test to see if the traits presented are different among species.

Table S1 and S2: correct R2.
Table S3: please, explain what ‘rank’ means.
Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time. The question is whether there was any stand dynamics involved in those height series that had to be included in the models.

Referee: 2

Comments to the Author
This manuscript examines a truly urgent and important issue. The general issue is the clear vulnerability of forests to ongoing climate change induced drought worldwide. The issue is particularly vexing because, while it seems clear that plant hydraulics is involved in vulnerability to drought, the exact mechanisms connecting hydraulics to vulnerability are not clear, and just what indices biologists should use to gauge vulnerability is as a result unclear. Within the broad pattern of global forest decline with climate change induced drought is the particularly mystifying and horrific preferential loss of the largest individuals all over the world. In addition to their esthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.

The issue of the preferential death of large trees is not only important because of its effects on forest function and culture. It is important because it seems to require novel theory. Under current expectations, the largest individuals, with their greater root reach, etc., should be more rather than less resistant. That they might not be demands a novel explanatory framework. Yet the very existence of the tendency is debated, with some studies reporting greater mortality among small rather than large individuals (Camarero et al., 2015; Colangelo et al., 2017). These studies, however, were not designed to take into account the microsite variation that presumably should lead to marked differences in survivorship even within the same height class with drought. One approach, which we have used (Soriano et al., in press; Olson et al., 2018) is to conduct experiments under controlled conditions to detect the effect of height, and our results are consistent with the greater vulnerability of taller individuals. However, under natural conditions, a study that takes this variation into account is needed to test the notion that taller individuals are preferentially vulnerable is needed, and this study provides exactly such a design.

I found this paper not only to address a very important point but also to be exceptionally well structured and written. My main suggestion is that it could use a bit more emphasis on its importance, and I would be interested in hearing a bit more about what the authors feel their results imply regarding the cause of the preferential vulnerability of large individuals. I give line-by-line suggestions (and sometimes just reactions):

first paragraph—this setup is very compelling, and the point is urgently true.
66- “in their basal portions” seems fine for this level of intro, but note that in angiosperms, taller trees also have wider vessels at the stem tip (and thus throughout the conductive system) (Zach et al., 2010; Olson et al., 2014, 2018). This seems plausibly the result of selection keeping conduit carbon costs down with height growth. As a tree grows taller, if terminal twig vessel diameters become wider, then given a constant rate of tip-to-base vessel widening, vessel diameters will be wider along the entire stem. Vessels that are wider throughout would have lower per-conduit resistance than those that, all else being equal, are narrower. So, conceivably as a tree gets taller, if vessels widen at the twig tip, then lower per-conduit resistance could mean that the same leaf area is supplied by fewer vessels. This could potentially contribute to keeping per-leaf area carbon costs constant with height growth. This prediction seems confirmed by the experiment of (Echeverría et al., 2019), who found that as Moringa trees grew taller, there were wider vessels at the stem tip and fewer vessels per unit leaf area, but that conductance per unit leaf area should remain the same. Terminal twig tracheids don’t seem to widen in gymnosperms, e.g. (Williams et al., 2019). (Sack et al., 2012) have shown very clearly that petiole base vessel diameter scales predictably with leaf size, as would be expected; it is not clear how leaf size variation is involved in terminal twig vessel diameter scaling with height, though in wide samplings leaf size doesn’t seem to scale with height (Jensen & Zwieniecki, 2013).

67 Appealing to the vulnerability-diameter link might get blowback from some readers because there is as yet no satisfactory theory linking vulnerability to diameter in an entirely satisfactory way, so maybe just say “wider conduits are likely” or that they are “plausibly” or “possibly” more vulnerable, or that many “data are consistent” with such a link. My view is that there very well might not be a link, but in the absence of a link between vulnerability and conduit diameter, it is very hard to explain a very large number of well-established observations. These include the following: why growth rings always proceed from wide to narrow conduits as drought ensues; why maximum community conduit diameter is narrower in drylands than in wet places; why lianas have wider vessel diameter variances (i.e. not only wider but also numerous narrower vessels) than self-supporting plants (Rosell & Olson, 2014); why plants virtually always produce vessels narrower than the maximum possible; when internal helical sculpture in conduits varies in a growth ring it is always associated with narrow latewood vessels; when simple and scalariform plates vary across growth rings sclariform plates are always found in the narrow latewood vessels; and many other observations. If there is no link between drought induced embolism vulnerability and conduit diameter, and there might not be, all of these observations, which are currently simultaneously and parsimoniously explained by the link, will require alternative explanations. As a result, while there is as yet no well accepted mechanistic link between vulnerability and diameter, it seems to me irresponsible to reject the link without providing a satisfactory and similarly parsimonious and general replacement explanation for all of these phenomena. So, I think the authors are justified in cautiously invoking the link here.

69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability. The generalization that on a global level large trees seem disproportionately vulnerable is often greeted with the skeptical observation that in specific cases a small tree died next to a large one that survived. However, the prediction is that *all else being equal* large individuals should be more vulnerable than smaller *conspecifics*. There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.

73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.

73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.

77 In the concluding sentence, “greater growth reductions” came as a surprise because I thought that the focus was greater vulnerability/ mortality. Also, there are convincing theoretical and empirical reasons to think that large trees don’t slow down e.g. (Sillett et al., 2010; Selaya & Anten, 2010; Michaletz et al., 2014).

96 this all seems compelling. To make the point stronger, either justify that the leaf traits are causally more important, or that, while they make up part of a unified hydraulic system all of whose components are important, leaf traits are simply easier to measure and this justifies their study. The latter position seems most plausible to me.

102 great questions, though within-species comparisons are likely to be most informative here, and your mixed model design is in accord with this; just being cautious because sometimes it’s seductive to regard selection acting between rather than within species

105 this sounds great, but was a little unclear. When you say “first, we test…” I thought that you would examine a relationship between microsite and some tree characteristic, e.g. that the largest trees are in moist microsites. But the next sentence didn’t refer to a tree-microsite correlation but vulnerability-size. It would be good to clarify the prose here because these are very important passages.  The third sentence did seem to be about tree-microsite correlations.

112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from. Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007). They are virtually absent in drylands, especially warm tropical drylands. Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).

113 These correlations seem reasonable, e.g. see (Méndez-Alonzo et al., 2012)

124 and Table 3. This seems like a good palette of species because they are very different in their hydraulic construction. In addition to libriform fibers, oaks have very abundant vasicentric tracheids, especially in latewood. This means that the oaks will likely maintain some conductance even under highly negative tensions. This is not by virtue of their ring porosity (note that most oaks globally aren’t ring porous!), but because of the high proportion of vasicentric tracheids in their latewood. Liriodendron (diffuse porous, scalariform perforation plates) and Carya (likely ring porous at your locality; simple perforation plates) both have libriform fibers and lack vasicentric tracheids. See the data in (Olson et al., in press). There is relatively low family-level diversity represented here; our phylogenetic analyses (Olson et al., in press) suggest that there is very little phylogenetic signal in the conduit diameter-stem length relationship, so this could help justify if necessary not including phylogeny in the analyses.

127 “tree size”= height?

138 you might discuss the adequacy of the height range studied. If it is true that conduit diameter predicts vulnerability, then the greatest differences in vulnerability should be over the height range over which conduit diameter changes most rapidly. The relationship between conduit diameter and plant height turns out to follow a power-law like distribution, see for example the SI in (Olson et al., 2018). This means that differences in vulnerability should be very marked across smaller size classes (say below 7 m or so and very marked below 3 m) and be less and less visible across larger plants. So, studies of changes in conduit diameter/ vulnerability with height would ideally include plants smaller than those included here, so maybe comment on how the height range might affect the ability to detect the patterns of interest. It strikes me that it is appropriate because it is the height range over which the putative preferential susceptibility of large trees to drought is being observed.

173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.

209 It couldn’t hurt to define basal area increment

227 the nesting seems appropriate and sufficient for taking species/ “phylogeny” into account

260 amazing

275—some people think that the increase in terminal twig vessels with height growth keeps pace with the changes in VPD from understory to canopy

281 nice

301 this intro paragraph to the discussion ranks very well the results from most significant to least

327 I agree that this is the point that most deserves discussion in paragraph 2

334 You might go into detail regarding what you think this says about the causal hypothesis behind the global decline of large individuals. To me it helps reject notions that somehow larger trees are more exposed to wind throw or other damaging agents. It is consistent with the hypothesis that larger individuals have wider conduits, and (assuming that wider conduits are more vulnerable to drought induced embolism) thus larger individuals are (all else being equal…) more vulnerable to embolism than smaller conspecifics.

336 beautifully said “impossible for trees to e ciently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”

349 fascinating; it seems plausible that current climate changed induced large-individual mortality involves trees growing to conduit diameters, and thus heights, permitted under previously more even and moister climates, and now finding themselves in more extreme climates that make their heights and conduits excessively wide given current conditions.

350 if you wanted to expand your discussion of causal mechanisms or the implications of your findings, I think you could jettison this paragraph. I admire your thoroughness in offering caveats, but I agree that this does not seem like a serious problem for your study.

382 an effort that seems in harmony with these concerns is (Christoffersen et al., 2016)

385 very nice paragraph

In conclusion, this ms addresses a very important issue with a well-designed study and appropriate analyses. It is well written and interesting and I hope my comments are useful.

Mark Olson molson@ib.unam.mx

Camarero JJ, Gazol A, Sangüesa-Barreda G, Oliva J, Vicente-Serrano SM. 2015. To die or not to die: early warnings of tree dieback in response to a severe drought (D Gibson, Ed.). Journal of Ecology 103: 44–57.
Christoffersen BO, Gloor M, Fauset S, Fyllas NM, Galbraith DR, Baker TR, Kruijt B, Rowland L, Fisher RA, Binks OJ, et al. 2016. Linking hydraulic traits to tropical forest function in a size-structured and trait-driven model (TFS v.1-Hydro). Geoscientific Model Development 9: 4227–4255.
Colangelo M, Camarero JJ, Borghetti M, Gazol A, Gentilesca T, Ripullone F. 2017. Size Matters a Lot: Drought-Affected Italian Oaks Are Smaller and Show Lower Growth Prior to Tree Death. Frontiers in Plant Science 8.
Echeverría A, Anfodillo T, Soriano D, Rosell JA, Olson ME. 2019. Constant theoretical conductance, changes in vessel diameter and number with height growth in Moringa oleifera. Journal of Experimental Botany.
Jensen KH, Zwieniecki MA. 2013. Physical Limits to Leaf Size in Tall Trees. Physical Review Letters 110.
Méndez-Alonzo R, Paz H, Zuluaga RC, Rosell JA, Olson ME. 2012. Coordinated evolution of leaf and stem economics in tropical dry forest trees. Ecology 93: 2397–2406.
Michaletz ST, Cheng D, Kerkhoff AJ, Enquist BJ. 2014. Convergence of terrestrial plant production across global climate gradients. Nature 512: 39–43.
Olson ME, Anfodillo T, Rosell JA, Petit G, Crivellaro A, Isnard S, León-Gómez C, Alvarado-Cárdenas LO, Castorena M. 2014. Universal hydraulics of the flowering plants: vessel diameter scales with stem length across angiosperm lineages, habits and climates. Ecology Letters 17: 988–997.
Olson ME, Rosell JA, Martínez-Pérez C, León-Gómez C, Fajardo A, Isnard S, Cervantes-Alcayde M-A, Echeverría A, Figueroa-Abúndiz, Segovia-Rivas A, et al. in press. Xylem vessel diameter-shoot length scaling: ecological significance of porosity types and other traits. Ecological Monographs.
Olson ME, Soriano D, Rosell JA, Anfodillo T, Donoghue MJ, Edwards EJ, León-Gómez C, Dawson T, Camarero Martínez JJ, Castorena M, et al. 2018. Plant height and hydraulic vulnerability to drought and cold. Proceedings of the National Academy of Sciences 115: 7551–7556.
Rosell JA, Olson ME. 2014. Do lianas really have wide vessels? Vessel diameter–stem length scaling in non-self-supporting plants. Perspectives in Plant Ecology, Evolution and Systematics 16: 288–295.
Sack L, Scoffoni C, McKown AD, Rawls M, Havran JC, Tran H, Tran T. 2012. Developmentally based scaling of leaf venation architecture explains global ecological patterns. Nature Communications 3: ncomms1835.
Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional trees have similar lifetime carbon gain in tropical secondary forest. Ecology 91: 1102–1113.
Sillett SC, Van Pelt R, Koch GW, Ambrose AR, Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production through old age in tall trees. Forest Ecology and Management 259: 976–994.
Soriano D, Echeverría A, Anfodillo T, Rosell JA, Olson ME. in press. Hydraulic traits vary as the result of tip-to-base conduit widening in vascular plants. Journal of Experimental Botany.
Wheeler EA, Baas P, Rodgers S. 2007. Variations In Dicot Wood Anatomy: A Global Analysis Based on the Insidewood Database. IAWA Journal 28: 229–258.
Williams CB, Anfodillo T, Crivellaro A, Lazzarin M, Dawson TE, Koch GW. 2019. Axial variation of xylem conduits in the Earth’s tallest trees. Trees 33: 1299–1311.
Zach A, Schuldt B, Brix S, Culmsee H, Culmsee H. 2010. Vessel diameter and xylem hydraulic conductivity increase with tree height in tropical rainforest trees in Sulawesi, Indonesia. Flora 205: 506–512.

Referee: 3

Comments to the Author
McGregor and colleagues use a combination of long-term tree records and functional trait data to explore the key drivers that shape the growth resistance of different temperate tree species to extreme drought. They find that that leaf hydraulic traits and tree size are strong predictors of growth responses to drought – with taller trees most affected.

I enjoyed the paper very much. It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology. I could definitely see the paper being of interest to a broad readership. However, I do have some important questions and suggestions about the analysis which I think need more careful thought. Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.

Main comments
L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!). But doesn’t this introduce a bias in your analysis? Trees that died may have had very different responses to drought that those which survived. Do you get similar results if you analyse trees that lived and died separately?

L165-167: I think this needs to be explained more clearly. Why are type I errors unlikely and type II error possible? I also think there are real issue with the analysis of crown position. Presumably on average the crown illumination index of trees would have increased over the past 50 or so years (from suppressed to dominant) as trees grew and neighbouring trees died. So I’m not convinced you can include this in your models going back half a century. It’s just not sensible to assume that crown position would not have changed over such a long time period.

L217-226: this is problematic because drought resistance (Rt) as defined by Lloret et al. (2011) assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time. But this is rarely/never the case with tree growth, which shows clear temporal trends related to tree size/ontogeny. If you don’t account for these directional trends then your estimates of Rt will be biased. For instance, if a tree is on a long-term trajectory of decreasing BAI over time, when you calculate Rt you will overestimate the impact of the drought (and vice versa for trees on an upward growth trajectory). So you need to first detrend the tree ring series to remove long-term growth trends (unless you can show that your trees exhibited no long-term temporal trends in BAI). The fact that pre-drought conditions were different for the three droughts is also a problem, but not much you can do about this except acknowledge as you have done.

L260-261: This is rather strange. So for the same individual tree, size was important in determining response to drought in one year but not in the others? And that’s for the years further back in time when the tree would have been smallest? Couldn’t that be interpreted as trees becoming less susceptible to drought as they grow larger/older? It’s also a little concerning that results are very different across the three droughts. Of these, the 1999 drought is the one I would tend to trust most as stand structure and composition would have been most similar to today.

Other comments
L42-54: I think it would be worth making the link between growth resilience and probability of mortality. Below are links to a couple of papers that have recently shown that trees that are less resilient to drought are also the ones most likely to die.
https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41467-020-14300-5.epdf%3Fshared_access_token%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw%253D%253D&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=6%2B7kb11Bzq4LivdOhnR4frg1MEFGaX%2FiIMUqJBG9gsM%3D&amp;reserved=0
https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0

L80-101 & L112-115: This is a nice summary of how traits can help us understand species responses to drought. The only thing that is a little strange is that you start by discussing commonly measured traits that have been used to explore susceptibility to drought (e.g., WD and LMA). Then you say that the low predictive power of these traits is because they are poor surrogates of what we should be measuring – i.e. hydraulic traits. But then you end by saying that these are too hard/expensive to measure, so you’re going to look at other surrogates.

L102 & L106: the introduction gives a nice overview of size and trait influences on resistance to drought. But the third focus of your study – microenvironment characteristics – gets little or no attention. In fact it’s not very clear what you mean by ‘microenvironment’ for most of the paper.

L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?

L173: this is the first mention of what I’m guessing is your ‘microenvironment’ descriptor. Worth briefly mentioning why you are measuring TWI.

L230: Somewhere here it would be good to remind the reader what these explanatory variables are.

L234-237: Not very clear how these models were constructed. Why did you fit each predictor independently to test its importance as opposed to standardizing them and having them together in a multiple regression and then compare their coefficients? And why was height retained in all models? Testing one predictor at a time might be sensible if you have lots of collinearity between predictors. But it could also mean you miss important conditional effects (e.g., the effect of variable x only emerges when you first account for the effect of variable y).

L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular). So given that you’ve already mentioned how you defined drought in the methods maybe you don’t need to repeat this here?

L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?

L287: specify how they were related to drought. Positive? Negative?

Tables and Figures: there are a large number of tables. I wonder if Table 2 and 3 would best fit in supporting information (with key information integrated directly into the text). I also feel that it would be nice to represent some of the key results (e.g., Rt and its relationship to height and leaf hydraulic traits) in a figure.

Referee: 4

Comments to the Author
In this paper, McGregor et al. study the effect of tree height, crown position, ring-porosity, wood density, LMA and two leaf water relations traits (percent loss of leaf area upon desiccation, PLA; and the turgor loss point, TLP) on the drought resistance of 12 temperate species from an oak-hickory forest of northern Virginia (USA). Drought resistance is assessed using series of radial growth data and for the three most severe droughts in the period 1950 - 2009 in the study area. The authors find consistent negative effects of tree height on drought resistance; effects of crown position, ring-porosity and water relation traits for some droughts; and no effect of wood density and LMA. The paper is well written and nicely organized around clear hypotheses. The results are intriguing but, in my opinion, the experimental design is not strong enough to clearly advance our understanding of the determinants of tree drought responses.

Firstly, the way drought resistance (Rt) is estimated, as the ratio of BAI during drought to the mean BAI over the five years preceding the drought can be problematic, as growth dynamics during this pre-drought period were quite variable across species (Figure 1; incidentally, why do you show ring width indices in this figure if you then use BAI?). This approach has also advantages, particularly when applied in a resilience framework (as in the original Lloret et al. (2011)). However, if you focus specificaly on resistance it seems that directly modelling the growth reduction caused by the drought (e.g., using intervention time series analysis or other methods) would be a more robust approach. At the very least, if you continue using your simpler approach I would strongly suggest including some measure of pre-drought growth in the models to account for differences in growth dynamics among species. It would also nice to clarify why you do not assess recovery and resilience, unlike Lloret et al. (2011).

Secondly, the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance, and it mixes different types of hypotheses operating at different levels. On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses. On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height. In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability. In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results. Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.

Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required. The paper assumes that all species were subjected to the same drought levels. However, this is unlikely to be true, as different species are likely to occupy different microhabitats in the plot, and hence be subjected to different drought conditions. The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.

Minor or more technical suggestions:
L59-79: the arguments presented in this paragraph mix different scales, from within-individual to within- or among-communities. I think the arguments could be better structured to reflect the sources of variability involved and, ideally, focus on the scale that will be addressed in the paper.

L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.

L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.

L97: ‘greater variation’ that what?

L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…

L137-140: mixing cores from dying and surviving individuals could be problematic. Did you find any association between survival and previous drought resistance? Was mortality different across species?

L160: you mean a log-log regression, right?

L181: N=3 individuals per species is really quite low. In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.

L188: ‘Wood’.

L189: fresh volume?

L223: clarify that the -4.53 value corresponds to July.

L228-234: please clarify model structure. I understand you included tree as a random factor only when the three droughts were analyzed together, right? It would be nice to know what was the contribution of random factors to model fit (only one R2 is shown in Table 5, and it is unclear whether it corresponds to marginal or conditional values).

L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.

L287-299: it would be very useful to use scatterplots to show the effects of traits on Rt in a figure.

L291-294: but this is problematic considering that you only had two diffuse-porous species.

L297: TLP was significant for 1966 according to Table 4.

L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.

L305-306: not for the 1977 and 1999 droughts, according to Table 5.

L314-316: or simply that your sample size was rather low. See also main comments above.

L329-331: do you have any information on the demographic effects (e.g., mortality) of the three droughts studied?

Table 3: are DBH units correct? I suspect they should be mm instead of cm.

Figure 1: please explain all acronyms.

Figure 2: where are the leaf hydraulic traits shown (cf. caption)?

Table S1: why only 11 species? Some of the fits are really low…

Table S3: what does ‘rank’ refer to?

Figure S1: please add spatial scale to the figure.

Figure S3: please specify that height values are estimated from reconstructed DBH.

### 2020-05-01 manuscript/response_to_reviews.Rmd
created response_to_reviews
@mcgregorian1 , FYI

---
title: "Response to reviews"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Referee: 1

Comments to the Author

The study addresses the relationship between drought resistance (Rt) calculated from radial growth, and different traits and microenvironmental characteristics on 12 tree species from one forest in the Eastern US. Both the dataset and the study are meaningful, and the relationship found between Rt and height of biological importance. However, I think there are a number of issues that need to be worked out.

- It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought. This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript). In the paper, the authors use Rt similarly to most studies in the literature, which is fine. Yet, Rt (and particularly the other resilience indices based on radial growth) does not only depend on the individual response to drought events, but also on stand dynamics around target trees. For this reason, if stands around trees are not comparable or show changes after some disturbance (e.g. the drought events analysed) differently, this may bias comparisons of Resilience indices (although it is true that Rt should be less influenced than the other indices). The problem is that there is not such stand competition data for old droughts. This does not invalidate the study but I think the authors should discuss the issue, particularly for the analysis and comparisons of single droughts. I would suggest including also other Resilience indices, not only Rt, to enrich the discussion.

- Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis. Then I missed a further discussion on the role of individual species (e.g. see my comment below on Rt>1).

I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.

Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models. LL ratio tests or deviance-anova tests can be used in the case of nested models with LMM or GLMM respectively. Along the text: dAIC is generally referred as $\Delta$AIC, please change accordingly.

Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included. And I suggest discussing only the model with all years together. This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation. This would help to ensure a better use of the Resilience indices.

The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis. If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.

- Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al. included). Droughts must be selected based only on some climatic/drought index (PDSI, SPEI,…) and independently of growth. Otherwise the analysis is biased towards years with low growth, and different questions would arise with a difficult answer. If the authors select only those droughts where there is a growth decrease: is that decrease evident in all species and similar across species? What happens with more intense droughts that those selected (if any), why they do not force a decrease in growth if drought is a main forcing agent? I see that 1966 and 1999 are the driest (as in PDSI) in Figure 1, but there is at least one year around 1955 that it is as dry or drier than 1977. In line 204 it is stated that they are the driest. If this is the case, the analysis should be sound but it should be deleted what is stated in lines 202-203 where authors say that they selected years also based on low growth. In fact, that should be a result after having selected the drought years only based on dryness (as in lines 208-209). Line 211 then they search for pointer years… a little bit messy. Please, explain better, and it should follow: 1) selection of dry years without regarding growth data; 2) characterise growth those dry years; 3) If the authors want to compare with other pointer years, fine, but that should be a third independent step properly justified and explained. Following this rationale, then in line 221 it should be difficult that Rt>1, had we selected those years with a negative growth anomaly. Please, explain clearly and consistently.

Other minor comments:

- Line 113: correct ‘positively’.

- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.

- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
should be rather straightforward and the authors would increase accuracy across species in height estimation.

- Lines 166-167: please explain further the hypothesis and why.

- Line 222: what about climate in years before 1991?

- Line 231: define AICc.

- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).

- Line 241-242: enough to say once that models were selected based on AIC (line 237-238) and then one criterion, why dAIC<1 here and dAIC<2 before? The classic assumption is 2 (e.g. Burnham & Anderson 2004). See also my other comment above.

- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.

- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?

- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.

- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest. They are indeed interesting to characterise the forest, but they do not add to individual tree variability. You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.

- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?

-Line 331-333: not so sure about this. It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.

- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).

- Line 348: non-drought.

- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.

- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.

Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.

Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.

Table 3: the authors could show some test to see if the traits presented are different among species.

Table S1 and S2: correct R2.

Table S3: please, explain what ‘rank’ means.

Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time. The question is whether there was any stand dynamics involved in those height series that had to be included in the models.

## Referee: 2

Comments to the Author

This manuscript examines a truly urgent and important issue. The general issue is the clear vulnerability of forests to ongoing climate change induced drought worldwide. The issue is particularly vexing because, while it seems clear that plant hydraulics is involved in vulnerability to drought, the exact mechanisms connecting hydraulics to vulnerability are not clear, and just what indices biologists should use to gauge vulnerability is as a result unclear. Within the broad pattern of global forest decline with climate change induced drought is the particularly mystifying and horrific preferential loss of the largest individuals all over the world. In addition to their esthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.

The issue of the preferential death of large trees is not only important because of its effects on forest function and culture. It is important because it seems to require novel theory. Under current expectations, the largest individuals, with their greater root reach, etc., should be more rather than less resistant. That they might not be demands a novel explanatory framework. Yet the very existence of the tendency is debated, with some studies reporting greater mortality among small rather than large individuals (Camarero et al., 2015; Colangelo et al., 2017). These studies, however, were not designed to take into account the microsite variation that presumably should lead to marked differences in survivorship even within the same height class with drought. One approach, which we have used (Soriano et al., in press; Olson et al., 2018) is to conduct experiments under controlled conditions to detect the effect of height, and our results are consistent with the greater vulnerability of taller individuals. However, under natural conditions, a study that takes this variation into account is needed to test the notion that taller individuals are preferentially vulnerable is needed, and this study provides exactly such a design.

I found this paper not only to address a very important point but also to be exceptionally well structured and written. My main suggestion is that it could use a bit more emphasis on its importance, and I would be interested in hearing a bit more about what the authors feel their results imply regarding the cause of the preferential vulnerability of large individuals. I give line-by-line suggestions (and sometimes just reactions):

first paragraph—this setup is very compelling, and the point is urgently true.

66- “in their basal portions” seems fine for this level of intro, but note that in angiosperms, taller trees also have wider vessels at the stem tip (and thus throughout the conductive system) (Zach et al., 2010; Olson et al., 2014, 2018). This seems plausibly the result of selection keeping conduit carbon costs down with height growth. As a tree grows taller, if terminal twig vessel diameters become wider, then given a constant rate of tip-to-base vessel widening, vessel diameters will be wider along the entire stem. Vessels that are wider throughout would have lower per-conduit resistance than those that, all else being equal, are narrower. So, conceivably as a tree gets taller, if vessels widen at the twig tip, then lower per-conduit resistance could mean that the same leaf area is supplied by fewer vessels. This could potentially contribute to keeping per-leaf area carbon costs constant with height growth. This prediction seems confirmed by the experiment of (Echeverría et al., 2019), who found that as Moringa trees grew taller, there were wider vessels at the stem tip and fewer vessels per unit leaf area, but that conductance per unit leaf area should remain the same. Terminal twig tracheids don’t seem to widen in gymnosperms, e.g. (Williams et al., 2019). (Sack et al., 2012) have shown very clearly that petiole base vessel diameter scales predictably with leaf size, as would be expected; it is not clear how leaf size variation is involved in terminal twig vessel diameter scaling with height, though in wide samplings leaf size doesn’t seem to scale with height (Jensen & Zwieniecki, 2013).

67 Appealing to the vulnerability-diameter link might get blowback from some readers because there is as yet no satisfactory theory linking vulnerability to diameter in an entirely satisfactory way, so maybe just say “wider conduits are likely” or that they are “plausibly” or “possibly” more vulnerable, or that many “data are consistent” with such a link. My view is that there very well might not be a link, but in the absence of a link between vulnerability and conduit diameter, it is very hard to explain a very large number of well-established observations. These include the following: why growth rings always proceed from wide to narrow conduits as drought ensues; why maximum community conduit diameter is narrower in drylands than in wet places; why lianas have wider vessel diameter variances (i.e. not only wider but also numerous narrower vessels) than self-supporting plants (Rosell & Olson, 2014); why plants virtually always produce vessels narrower than the maximum possible; when internal helical sculpture in conduits varies in a growth ring it is always associated with narrow latewood vessels; when simple and scalariform plates vary across growth rings sclariform plates are always found in the narrow latewood vessels; and many other observations. If there is no link between drought induced embolism vulnerability and conduit diameter, and there might not be, all of these observations, which are currently simultaneously and parsimoniously explained by the link, will require alternative explanations. As a result, while there is as yet no well accepted mechanistic link between vulnerability and diameter, it seems to me irresponsible to reject the link without providing a satisfactory and similarly parsimonious and general replacement explanation for all of these phenomena. So, I think the authors are justified in cautiously invoking the link here.

69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability. The generalization that on a global level large trees seem disproportionately vulnerable is often greeted with the skeptical observation that in specific cases a small tree died next to a large one that survived. However, the prediction is that *all else being equal* large individuals should be more vulnerable than smaller *conspecifics*. There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.

73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.

73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.

77 In the concluding sentence, “greater growth reductions” came as a surprise because I thought that the focus was greater vulnerability/ mortality. Also, there are convincing theoretical and empirical reasons to think that large trees don’t slow down e.g. (Sillett et al., 2010; Selaya & Anten, 2010; Michaletz et al., 2014).

96 this all seems compelling. To make the point stronger, either justify that the leaf traits are causally more important, or that, while they make up part of a unified hydraulic system all of whose components are important, leaf traits are simply easier to measure and this justifies their study. The latter position seems most plausible to me.

102 great questions, though within-species comparisons are likely to be most informative here, and your mixed model design is in accord with this; just being cautious because sometimes it’s seductive to regard selection acting between rather than within species

105 this sounds great, but was a little unclear. When you say “first, we test…” I thought that you would examine a relationship between microsite and some tree characteristic, e.g. that the largest trees are in moist microsites. But the next sentence didn’t refer to a tree-microsite correlation but vulnerability-size. It would be good to clarify the prose here because these are very important passages.  The third sentence did seem to be about tree-microsite correlations.

112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from. Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007). They are virtually absent in drylands, especially warm tropical drylands. Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).

113 These correlations seem reasonable, e.g. see (Méndez-Alonzo et al., 2012)

124 and Table 3. This seems like a good palette of species because they are very different in their hydraulic construction. In addition to libriform fibers, oaks have very abundant vasicentric tracheids, especially in latewood. This means that the oaks will likely maintain some conductance even under highly negative tensions. This is not by virtue of their ring porosity (note that most oaks globally aren’t ring porous!), but because of the high proportion of vasicentric tracheids in their latewood. Liriodendron (diffuse porous, scalariform perforation plates) and Carya (likely ring porous at your locality; simple perforation plates) both have libriform fibers and lack vasicentric tracheids. See the data in (Olson et al., in press). There is relatively low family-level diversity represented here; our phylogenetic analyses (Olson et al., in press) suggest that there is very little phylogenetic signal in the conduit diameter-stem length relationship, so this could help justify if necessary not including phylogeny in the analyses.

127 “tree size”= height?

138 you might discuss the adequacy of the height range studied. If it is true that conduit diameter predicts vulnerability, then the greatest differences in vulnerability should be over the height range over which conduit diameter changes most rapidly. The relationship between conduit diameter and plant height turns out to follow a power-law like distribution, see for example the SI in (Olson et al., 2018). This means that differences in vulnerability should be very marked across smaller size classes (say below 7 m or so and very marked below 3 m) and be less and less visible across larger plants. So, studies of changes in conduit diameter/ vulnerability with height would ideally include plants smaller than those included here, so maybe comment on how the height range might affect the ability to detect the patterns of interest. It strikes me that it is appropriate because it is the height range over which the putative preferential susceptibility of large trees to drought is being observed.

173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.

209 It couldn’t hurt to define basal area increment

227 the nesting seems appropriate and sufficient for taking species/ “phylogeny” into account

260 amazing

275—some people think that the increase in terminal twig vessels with height growth keeps pace with the changes in VPD from understory to canopy

281 nice

301 this intro paragraph to the discussion ranks very well the results from most significant to least

327 I agree that this is the point that most deserves discussion in paragraph 2

334 You might go into detail regarding what you think this says about the causal hypothesis behind the global decline of large individuals. To me it helps reject notions that somehow larger trees are more exposed to wind throw or other damaging agents. It is consistent with the hypothesis that larger individuals have wider conduits, and (assuming that wider conduits are more vulnerable to drought induced embolism) thus larger individuals are (all else being equal…) more vulnerable to embolism than smaller conspecifics.

336 beautifully said “impossible for trees to e ciently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”

349 fascinating; it seems plausible that current climate changed induced large-individual mortality involves trees growing to conduit diameters, and thus heights, permitted under previously more even and moister climates, and now finding themselves in more extreme climates that make their heights and conduits excessively wide given current conditions.

350 if you wanted to expand your discussion of causal mechanisms or the implications of your findings, I think you could jettison this paragraph. I admire your thoroughness in offering caveats, but I agree that this does not seem like a serious problem for your study.

382 an effort that seems in harmony with these concerns is (Christoffersen et al., 2016)

385 very nice paragraph

In conclusion, this ms addresses a very important issue with a well-designed study and appropriate analyses. It is well written and interesting and I hope my comments are useful.

Mark Olson molson@ib.unam.mx

Camarero JJ, Gazol A, Sangüesa-Barreda G, Oliva J, Vicente-Serrano SM. 2015. To die or not to die: early warnings of tree dieback in response to a severe drought (D Gibson, Ed.). Journal of Ecology 103: 44–57.

Christoffersen BO, Gloor M, Fauset S, Fyllas NM, Galbraith DR, Baker TR, Kruijt B, Rowland L, Fisher RA, Binks OJ, et al. 2016. Linking hydraulic traits to tropical forest function in a size-structured and trait-driven model (TFS v.1-Hydro). Geoscientific Model Development 9: 4227–4255.

Colangelo M, Camarero JJ, Borghetti M, Gazol A, Gentilesca T, Ripullone F. 2017. Size Matters a Lot: Drought-Affected Italian Oaks Are Smaller and Show Lower Growth Prior to Tree Death. Frontiers in Plant Science 8.

Echeverría A, Anfodillo T, Soriano D, Rosell JA, Olson ME. 2019. Constant theoretical conductance, changes in vessel diameter and number with height growth in Moringa oleifera. Journal of Experimental Botany.

Jensen KH, Zwieniecki MA. 2013. Physical Limits to Leaf Size in Tall Trees. Physical Review Letters 110.

Méndez-Alonzo R, Paz H, Zuluaga RC, Rosell JA, Olson ME. 2012. Coordinated evolution of leaf and stem economics in tropical dry forest trees. Ecology 93: 2397–2406.

Michaletz ST, Cheng D, Kerkhoff AJ, Enquist BJ. 2014. Convergence of terrestrial plant production across global climate gradients. Nature 512: 39–43.

Olson ME, Anfodillo T, Rosell JA, Petit G, Crivellaro A, Isnard S, León-Gómez C, Alvarado-Cárdenas LO, Castorena M. 2014. Universal hydraulics of the flowering plants: vessel diameter scales with stem length across angiosperm lineages, habits and climates. Ecology Letters 17: 988–997.

Olson ME, Rosell JA, Martínez-Pérez C, León-Gómez C, Fajardo A, Isnard S, Cervantes-Alcayde M-A, Echeverría A, Figueroa-Abúndiz, Segovia-Rivas A, et al. in press. Xylem vessel diameter-shoot length scaling: ecological significance of porosity types and other traits. Ecological Monographs.

Olson ME, Soriano D, Rosell JA, Anfodillo T, Donoghue MJ, Edwards EJ, León-Gómez C, Dawson T, Camarero Martínez JJ, Castorena M, et al. 2018. Plant height and hydraulic vulnerability to drought and cold. Proceedings of the National Academy of Sciences 115: 7551–7556.

Rosell JA, Olson ME. 2014. Do lianas really have wide vessels? Vessel diameter–stem length scaling in non-self-supporting plants. Perspectives in Plant Ecology, Evolution and Systematics 16: 288–295.

Sack L, Scoffoni C, McKown AD, Rawls M, Havran JC, Tran H, Tran T. 2012. Developmentally based scaling of leaf venation architecture explains global ecological patterns. Nature Communications 3: ncomms1835.

Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional trees have similar lifetime carbon gain in tropical secondary forest. Ecology 91: 1102–1113.
Sillett SC, Van Pelt R, Koch GW, Ambrose AR, Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production through old age in tall trees. Forest Ecology and Management 259: 976–994.

Soriano D, Echeverría A, Anfodillo T, Rosell JA, Olson ME. in press. Hydraulic traits vary as the result of tip-to-base conduit widening in vascular plants. Journal of Experimental Botany.

Wheeler EA, Baas P, Rodgers S. 2007. Variations In Dicot Wood Anatomy: A Global Analysis Based on the Insidewood Database. IAWA Journal 28: 229–258.

Williams CB, Anfodillo T, Crivellaro A, Lazzarin M, Dawson TE, Koch GW. 2019. Axial variation of xylem conduits in the Earth’s tallest trees. Trees 33: 1299–1311.

Zach A, Schuldt B, Brix S, Culmsee H, Culmsee H. 2010. Vessel diameter and xylem hydraulic conductivity increase with tree height in tropical rainforest trees in Sulawesi, Indonesia. Flora 205: 506–512.

## Referee: 3

Comments to the Author

McGregor and colleagues use a combination of long-term tree records and functional trait data to explore the key drivers that shape the growth resistance of different temperate tree species to extreme drought. They find that that leaf hydraulic traits and tree size are strong predictors of growth responses to drought – with taller trees most affected.

I enjoyed the paper very much. It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology. I could definitely see the paper being of interest to a broad readership. However, I do have some important questions and suggestions about the analysis which I think need more careful thought. Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.

Main comments

L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!). But doesn’t this introduce a bias in your analysis? Trees that died may have had very different responses to drought that those which survived. Do you get similar results if you analyse trees that lived and died separately?

L165-167: I think this needs to be explained more clearly. Why are type I errors unlikely and type II error possible? I also think there are real issue with the analysis of crown position. Presumably on average the crown illumination index of trees would have increased over the past 50 or so years (from suppressed to dominant) as trees grew and neighbouring trees died. So I’m not convinced you can include this in your models going back half a century. It’s just not sensible to assume that crown position would not have changed over such a long time period.

L217-226: this is problematic because drought resistance (Rt) as defined by Lloret et al. (2011) assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time. But this is rarely/never the case with tree growth, which shows clear temporal trends related to tree size/ontogeny. If you don’t account for these directional trends then your estimates of Rt will be biased. For instance, if a tree is on a long-term trajectory of decreasing BAI over time, when you calculate Rt you will overestimate the impact of the drought (and vice versa for trees on an upward growth trajectory). So you need to first detrend the tree ring series to remove long-term growth trends (unless you can show that your trees exhibited no long-term temporal trends in BAI). The fact that pre-drought conditions were different for the three droughts is also a problem, but not much you can do about this except acknowledge as you have done.

L260-261: This is rather strange. So for the same individual tree, size was important in determining response to drought in one year but not in the others? And that’s for the years further back in time when the tree would have been smallest? Couldn’t that be interpreted as trees becoming less susceptible to drought as they grow larger/older? It’s also a little concerning that results are very different across the three droughts. Of these, the 1999 drought is the one I would tend to trust most as stand structure and composition would have been most similar to today.

Other comments

L42-54: I think it would be worth making the link between growth resilience and probability of mortality. Below are links to a couple of papers that have recently shown that trees that are less resilient to drought are also the ones most likely to die.
https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41467-020-14300-5.epdf%3Fshared_access_token%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw%253D%253D&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=6%2B7kb11Bzq4LivdOhnR4frg1MEFGaX%2FiIMUqJBG9gsM%3D&amp;reserved=0
https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0

L80-101 & L112-115: This is a nice summary of how traits can help us understand species responses to drought. The only thing that is a little strange is that you start by discussing commonly measured traits that have been used to explore susceptibility to drought (e.g., WD and LMA). Then you say that the low predictive power of these traits is because they are poor surrogates of what we should be measuring – i.e. hydraulic traits. But then you end by saying that these are too hard/expensive to measure, so you’re going to look at other surrogates.

L102 & L106: the introduction gives a nice overview of size and trait influences on resistance to drought. But the third focus of your study – microenvironment characteristics – gets little or no attention. In fact it’s not very clear what you mean by ‘microenvironment’ for most of the paper.

L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?

L173: this is the first mention of what I’m guessing is your ‘microenvironment’ descriptor. Worth briefly mentioning why you are measuring TWI.

L230: Somewhere here it would be good to remind the reader what these explanatory variables are.

L234-237: Not very clear how these models were constructed. Why did you fit each predictor independently to test its importance as opposed to standardizing them and having them together in a multiple regression and then compare their coefficients? And why was height retained in all models? Testing one predictor at a time might be sensible if you have lots of collinearity between predictors. But it could also mean you miss important conditional effects (e.g., the effect of variable x only emerges when you first account for the effect of variable y).

L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular). So given that you’ve already mentioned how you defined drought in the methods maybe you don’t need to repeat this here?

L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?

L287: specify how they were related to drought. Positive? Negative?

Tables and Figures: there are a large number of tables. I wonder if Table 2 and 3 would best fit in supporting information (with key information integrated directly into the text). I also feel that it would be nice to represent some of the key results (e.g., Rt and its relationship to height and leaf hydraulic traits) in a figure.

## Referee: 4

Comments to the Author

In this paper, McGregor et al. study the effect of tree height, crown position, ring-porosity, wood density, LMA and two leaf water relations traits (percent loss of leaf area upon desiccation, PLA; and the turgor loss point, TLP) on the drought resistance of 12 temperate species from an oak-hickory forest of northern Virginia (USA). Drought resistance is assessed using series of radial growth data and for the three most severe droughts in the period 1950 - 2009 in the study area. The authors find consistent negative effects of tree height on drought resistance; effects of crown position, ring-porosity and water relation traits for some droughts; and no effect of wood density and LMA. The paper is well written and nicely organized around clear hypotheses. The results are intriguing but, in my opinion, the experimental design is not strong enough to clearly advance our understanding of the determinants of tree drought responses.

Firstly, the way drought resistance (Rt) is estimated, as the ratio of BAI during drought to the mean BAI over the five years preceding the drought can be problematic, as growth dynamics during this pre-drought period were quite variable across species (Figure 1; incidentally, why do you show ring width indices in this figure if you then use BAI?). This approach has also advantages, particularly when applied in a resilience framework (as in the original Lloret et al. (2011)). However, if you focus specificaly on resistance it seems that directly modelling the growth reduction caused by the drought (e.g., using intervention time series analysis or other methods) would be a more robust approach. At the very least, if you continue using your simpler approach I would strongly suggest including some measure of pre-drought growth in the models to account for differences in growth dynamics among species. It would also nice to clarify why you do not assess recovery and resilience, unlike Lloret et al. (2011).

Secondly, the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance, and it mixes different types of hypotheses operating at different levels. On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses. On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height. In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability. In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results. Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.

Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required. The paper assumes that all species were subjected to the same drought levels. However, this is unlikely to be true, as different species are likely to occupy different microhabitats in the plot, and hence be subjected to different drought conditions. The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.

Minor or more technical suggestions:

L59-79: the arguments presented in this paragraph mix different scales, from within-individual to within- or among-communities. I think the arguments could be better structured to reflect the sources of variability involved and, ideally, focus on the scale that will be addressed in the paper.

L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.

L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.

L97: ‘greater variation’ that what?

L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…

L137-140: mixing cores from dying and surviving individuals could be problematic. Did you find any association between survival and previous drought resistance? Was mortality different across species?

L160: you mean a log-log regression, right?

L181: N=3 individuals per species is really quite low. In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.

L188: ‘Wood’.

L189: fresh volume?

L223: clarify that the -4.53 value corresponds to July.

L228-234: please clarify model structure. I understand you included tree as a random factor only when the three droughts were analyzed together, right? It would be nice to know what was the contribution of random factors to model fit (only one R2 is shown in Table 5, and it is unclear whether it corresponds to marginal or conditional values).

L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.

L287-299: it would be very useful to use scatterplots to show the effects of traits on Rt in a figure.

L291-294: but this is problematic considering that you only had two diffuse-porous species.

L297: TLP was significant for 1966 according to Table 4.

L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.

L305-306: not for the 1977 and 1999 droughts, according to Table 5.

L314-316: or simply that your sample size was rather low. See also main comments above.

L329-331: do you have any information on the demographic effects (e.g., mortality) of the three droughts studied?

Table 3: are DBH units correct? I suspect they should be mm instead of cm.

Figure 1: please explain all acronyms.

Figure 2: where are the leaf hydraulic traits shown (cf. caption)?

Table S1: why only 11 species? Some of the fits are really low…

Table S3: what does ‘rank’ refer to?

Figure S1: please add spatial scale to the figure.

Figure S3: please specify that height values are estimated from reconstructed DBH.

### 2020-05-01 manuscript/response_to_reviews.Rmd
started assessing reviewer comments
I think I highlighted the most important issues through the GitHub issues that I opened (#71, #74, #92, #93, #94).

@@
 ```
+
+*italics - temporary comments for points that remain to be addressed*
+
+**bold - response**
 
@@
 
+[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
+
+[*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
+
 - Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis.
@@
 
+[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
+
 I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.
+
+[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
 
@@
 Along the text: dAIC is generally referred as $\Delta$AIC, please change accordingly.
+
+[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
+
+*dAIC: minor fix*
 
@@
 
+*This is an interesting suggestion.
+Depending on where we end up going with some other issues, it may make sense to de-emphasize the individual models--perhaps move Table 5 to SI and replace with figure ([issue #71](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71))?
+However, I think we could also just make an argument for leaving this as is.*
+
+related to [*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
 The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis.
 If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.
+
+[*issue #74*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/74)
 
@@
 Please, explain clearly and consistently.
+
+*Ugghh, this doesn't even change anything--we'd still be analyzing the same 3 droughts!
+The question is whether to remove the growth reduction criteria.
+I personally think that analyzing 2 criteria is a strength, as our goal was to identify droughts that were both climatologically and biologically signficant.
+In any case, this is just an issue of changing the wording.*
 
@@
 
+*this is true of crown position, not height*
+
 - Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest.
@@
 
+[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
 - Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?
 
--Line 331-333: not so sure about this.
+- Line 331-333: not so sure about this.
 It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
@@
 I give line-by-line suggestions (and sometimes just reactions):
+
+*What a nice review! We can easily add some language on the importance.*
 
@@
 
+*good point. hypothesis refers to temperate deciduous biome.*
+
 113 These correlations seem reasonable, e.g. see (Méndez-Alonzo et al., 2012)
@@
 There is relatively low family-level diversity represented here; our phylogenetic analyses (Olson et al., in press) suggest that there is very little phylogenetic signal in the conduit diameter-stem length relationship, so this could help justify if necessary not including phylogeny in the analyses.
+
+*this is new terminology to me. Look into Olson et al. in press. Would these wood traits be worth including as covariates?*
 
@@
 
+**Thank you.**
+
 275—some people think that the increase in terminal twig vessels with height growth keeps pace with the changes in VPD from understory to canopy
@@
 
+**Thank you.**
+
 301 this intro paragraph to the discussion ranks very well the results from most significant to least
+
+**Thank you.**
 
@@
 336 beautifully said “impossible for trees to e ciently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”
+
+**Thank you.**
 
@@
 
+**Thank you.**
+
 In conclusion, this ms addresses a very important issue with a well-designed study and appropriate analyses. It is well written and interesting and I hope my comments are useful.
+
+**Thank you.**
 
@@
 
+[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
+
 L260-261: This is rather strange.
@@
 
+[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
 ## Referee: 4
@@
 (2011).
+
+[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
+
+[*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
 
@@
 
+[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
 L291-294: but this is problematic considering that you only had two diffuse-porous species.

### 2020-05-17 manuscript/response_to_reviews.Rmd
Added comments to review reponses

@@
 In any case, this is just an issue of changing the wording.*
+
+**IM - I kind of understand what they meant, but more in terms of a logic flow position. I agree that this can pretty much be fixed**
 
@@
 
+**It seems from these comments they think this shouldn't be published at all due to its limited scope and not enough explanatory power.
+The resulting study that they suggest, if completely thorough, would take a long time to create.
+I'm not sure how to address this comment, other than to make a note in the manuscript acknowledging that a more thorough study would be great but this is what we found with what we had.**
+
 Minor or more technical suggestions:

### 2020-05-19 manuscript/response_to_reviews.Rmd
address comments on resilience
#93

--- 2020-05-01 manuscript/response_to_reviews.Rmd
@@
 
-- It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought.
+It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought.
 This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript).
@@
 
+**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
+While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
+(2) [DeSoto et al.
+(2020, Nature)](https://www.nature.com/articles/s41467-020-14300-5) used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
+In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
+In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
+These findings would argue that resistance is a more important variable for angiosperms.
+(3) As the reviewer notes, resilience indices would tend to be more sensitive to a tree's neighborhood and prior conditions.
+We now state the logic for focusing on $Rt$ in the methods section:**
+
 [*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
 
-- Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis.
+Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis.
 Then I missed a further discussion on the role of individual species (e.g.
@@
 
-- Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al.
+Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al.
 included).
@@
 In any case, this is just an issue of changing the wording.*
+
+*IM - I kind of understand what they meant, but more in terms of a logic flow position. I agree that this can pretty much be fixed*
 
@@
 
+**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
+While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
+(2) [DeSoto et al.
+(2020, Nature)](https://www.nature.com/articles/s41467-020-14300-5) used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
+In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
+In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
+These findings would argue that resistance is a more important variable for angiosperms.
+(3) As the reviewer notes, resilience indices would tend to be more sensitive to a tree's neighborhood and prior conditions.
+We now state the logic for focusing on $Rt$ in the methods section:**
+
 [*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
@@
 The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.
+
+*It seems from these comments they think this shouldn't be published at all due to its limited scope and not enough explanatory power.
+The resulting study that they suggest, if completely thorough, would take a long time to create.
+I'm not sure how to address this comment, other than to make a note in the manuscript acknowledging that a more thorough study would be great but this is what we found with what we had.*
+
+*I agree.... I wouldn't change analysis for this. We can come back to wording a response later.*
 

### 2020-05-19 manuscript/climate_sensitivity_draft.Rmd
address comments on resilience
#93

@@
 Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
+We focus exclusively on $Rt$, and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, (2) the findings of *[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms, and (3) *$Rt$ would be less sensitive than resilience to a tree's neighborhood and prior conditions*.
 The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.

### 2020-05-19 manuscript/response_to_reviews.Rmd
response on resilience metrics
#93

@@
 These findings would argue that resistance is a more important variable for angiosperms.
-(3) As the reviewer notes, resilience indices would tend to be more sensitive to a tree's neighborhood and prior conditions.
+(3) As the reviewer notes, resilience indices would probably tend to be more sensitive to a tree's neighborhood and prior conditions.
 We now state the logic for focusing on $Rt$ in the methods section:**
 
-[*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
+**"We focus exclusively on Rt, and not on the resilience metrics described in Lloret et al.
+(2011), because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of DeSoto et al.
+(2020) suggest that Rt is a more important drought response metric for angiosperms."**
 
@@
 These findings would argue that resistance is a more important variable for angiosperms.
-(3) As the reviewer notes, resilience indices would tend to be more sensitive to a tree's neighborhood and prior conditions.
 We now state the logic for focusing on $Rt$ in the methods section:**
 
-[*issue #93*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/93)
+**"We focus exclusively on Rt, and not on the resilience metrics described in Lloret et al.
+(2011), because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of DeSoto et al.
+(2020) suggest that Rt is a more important drought response metric for angiosperms."**
 

### 2020-05-19 manuscript/climate_sensitivity_draft.Rmd
response on resilience metrics
#93

--- 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
@@
 Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
+We focus exclusively on $Rt$, and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
 The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.

### 2020-06-12 manuscript/tables&figures.Rmd
fig legend fix

@@
 **Figure 2.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
 In these plots, heights are slightly offset for visualization purposes.

### 2020-06-13 manuscript/tables&figures.Rmd
Added tables and updated rmd for doing arima ratio instead of difference

@@
 
+<!-- \clearpage -->
+<!-- ```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE} -->
+<!-- # *****this is for if doing arimaratio!!!!***** -->
+<!-- library(knitr) -->
+<!-- library(kableExtra) -->
+<!-- multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+-->
+<!-- kable(multivar_table, booktabs = TRUE, caption = "Table 5.
+Summary of top full models for each drought instance.
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>% -->
+<!--   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>% -->
+<!--   add_header_above(c(" "," "," "," "," ", " ", " ", "xylem porosity" = 2," ", " ")) %>% -->
+<!--   pack_rows("", 1, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
+<!--   pack_rows("", 3, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
+<!--   pack_rows("", 6, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
+<!--   pack_rows("", 8, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
+<!--   row_spec(1, bold = T) %>% -->
+<!--   row_spec(3, bold = T) %>% -->
+<!--   row_spec(6, bold = T) %>% -->
+<!--   row_spec(8, bold = T) %>% -->
+<!--   kable_styling() -->
+<!-- ``` -->
+
 \newpage

### 2020-06-13 scripts/climate_sensitivity_analysis.R

@@ -34,6 +34,9 @@ library(reshape2)
 rings <- read.rwl("data/core_files/all_species_except_FRNI_PIST.rwl") #read in rwl file
 widths <- rings #for consistency with original code
 area <- bai.in(rings) #convert to bai.in
+
+## IF DO ARIMA SKIP DOWN NOW.
+
 resil_metrics <- res.comp(area, nb.yrs=5, res.thresh.neg = 30, series.thresh = 25) #get resilience metrics
 
 resil_pointers <- data.frame(resil_metrics$out)
@@ -78,6 +81,7 @@ change <- change[change$year %in% c("1966","1977","1999"), ] #for models
 
 #this is trees_all
 trees_all <- change[complete.cases(change), ]
+trees_all <- trees_all[trees_all$resist.value <=2,] #constrain resistance values
 
 #
 
@@ -122,9 +126,10 @@ for(j in seq(along=trees)){
       future <- forecast(arima_dt, h=1, level=c(99.5))
       # plot(future)
       
-      #3. Get the difference between the observed and predicted
+      #3. Get the ratio between the observed and predicted
       areaobs <- area_dt[area_dt$year == droughts[i], ]
-      diff <- round(areaobs$val - as.numeric(future$mean),2)
+      diff <- round(areaobs$val / as.numeric(future$mean),2) #ratio
+      # diff <- round(areaobs$val / as.numeric(future$mean),2) #difference
     } else {diff <- NA}
     
     inddrt <- data.frame(year = droughts[i],
@@ -182,6 +187,9 @@ ind$sp <- ind_neil$sp[match(ind$tree, ind_neil$tag)]
 
 #this is trees_all
 trees_all <- ind[complete.cases(ind), ]
+trees_all <- trees_all[trees_all$resist.value != Inf &
+                         trees_all$resist.value < 4 &
+                         trees_all$resist.value >=0, ]
 
 #
 ##1b. summary stat: determine proportion of resistance values per sp ####
@@ -240,7 +248,7 @@ ggplot(data = rp_test) +
 #this comes from the hydraulic traits repo, "SCBI_all_traits_table_species_level.csv"
 ##leaf traits gained from this include PLA_dry_percent, LMA_g_per_m2, Chl_m2_per_g, and WD [wood density]
 
[…]

### 2020-06-14 scripts/climate_sensitivity_analysis.R
Added comparison of Rt and arima_ratio values

@@ -588,8 +588,8 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 ##start here if just re-running model runs ####
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
 # trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arima.csv", stringsAsFactors = FALSE)
-# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv", 
-#                           stringsAsFactors = FALSE)
+trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv",
+                          stringsAsFactors = FALSE)
 x1966 <- trees_all_sub[trees_all_sub$year == 1966, ]
 x1977 <- trees_all_sub[trees_all_sub$year == 1977, ]
 x1999 <- trees_all_sub[trees_all_sub$year == 1999, ]
@@ -1096,6 +1096,49 @@ for(i in seq(along=patterns)){
 
 write.csv(coeff_new, "manuscript/tables_figures/tested_traits_best_coeff.csv", row.names=FALSE)
 
+
+# ##3d. compare Rt values with arima_ratio ####
+# arima_ratio <- read.csv("trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
+# 
+# rt <- read.csv("trees_all_sub.csv", stringsAsFactors = FALSE)
+# 
+# x1966 <- rt[rt$year == 1966, ]
+# x1977 <- rt[rt$year == 1977, ]
+# x1999 <- rt[rt$year == 1999, ]
+# 
+# arima1966 <- arima_ratio[arima_ratio$year == 1966, ]
+# arima1977 <- arima_ratio[arima_ratio$year == 1977, ]
+# arima1999 <- arima_ratio[arima_ratio$year == 1999, ]
+# 
+# layout(matrix(1:4, nrow=2, byrow=TRUE))
+# compare <- rt[,c("year", "tree", "resist.value")]
+# compare$arimart <- arima_ratio$resist.value[
+#   match(paste0(compare$year, compare$tree), 
+#         paste0(arima_ratio$year, arima_ratio$tree))]
+# plot(compare$resist.value, compare$arimart, main="All years",
+#      xlab="rt", ylab="ARIMA")
+# abline(coef=c(0,1), col="red")
+# 
+# compare66 <- x1966[,c("tree", "resist.value")]
[…]

### 2020-06-14 manuscript/tables&figures.Rmd
Got results using new base model method, for both arima_ratio and Rt

@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
 kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
@@
   add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
-  pack_rows("Species traits", 13, 18, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
   #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
 kable(multivar_table, booktabs = TRUE, caption = "Table 5.
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem porosity" = 2," ", " ")) %>%
-  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 6, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 10, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
+                     "xylem porosity" = 2," ")) %>%
+  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 4, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(5, bold = T) %>%
-  row_spec(6, bold = T) %>%
-  row_spec(10, bold = T) %>%
+  row_spec(2, bold = T) %>%
+  row_spec(3, bold = T) %>%
+  row_spec(4, bold = T) %>%
   kable_styling()

### 2020-06-14 scripts/climate_sensitivity_analysis.R

@@ -586,7 +586,7 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #3. mixed effects model for output of #2.
 
 ##start here if just re-running model runs ####
-trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
+# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
 # trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arima.csv", stringsAsFactors = FALSE)
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv",
                           stringsAsFactors = FALSE)
@@ -606,8 +606,9 @@ library(MuMIn) #for R^2 values of one model output
 library(stringr)
 library(purrr)
 library(dplyr)
+library(data.table)
 
-##3a. test each variable individually (expand for fuller explanation) ####
+##3a reform. test each variable individually (expand for fuller explanation) ####
 # this code chunk tests each variable individually for each drought scenario,
 # using a predefined null model and test model (i.e. all test models have height).
 # After these are tested from a defined data frame (sum_mod_traits), the "candidate"
@@ -615,22 +616,15 @@ library(dplyr)
 # top models. These candidates are then used in ##6b.
 
 sum_mod_traits <- data.frame(
-  "prediction" = c(3.1, 2.1, 2.2, 2.3, 1.2, 1.3, 2.4, 3.1, 3.2, 3.3, 3.4, 3.5),
-  "variable" = c("year", "dbh.ln.cm", "height.ln.m", "position_all", "position_all", "height.ln.m*TWI.ln", "TWI.ln", "rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
-  "variable_description" = c("drought.year", "ln[DBH]", "ln[height]", "crown.position alone", "crown.position w/height", "ln[height]*ln[topographic.wetness.index]", "ln[topographic.wetness.index]", "ring.porosity", "percent.loss.area", "leaf.mass.area", "mean.turgor.loss.point", "wood.density"),
+  "prediction" = c(3.1, 3.2, 3.3, 3.4, 3.5),
+  "variable" = c("rp", "PLA_dry_percent", "LMA_g_per_m2", "mean_TLP_Mpa", "WD_g_per_cm3"), 
[…]

### 2020-06-28 manuscript/response_to_reviews.Rmd
update response to reviews
drafted reviewer responses on ARIMA
We still need to add specifics once the text/ SI content is finalized.
#92

@@
 
+**To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA).
+Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
+In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
+We now include the following content on ARIMA in the manuscript:**
+
+(*insert text*)
+
 **Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
@@
 
-**"We focus exclusively on Rt, and not on the resilience metrics described in Lloret et al.
-(2011), because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of DeSoto et al.
-(2020) suggest that Rt is a more important drought response metric for angiosperms."**
+**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
+"**
 
@@
 
-[*issue #74*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/74)
+[*issue #95*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/95)
 
@@
 
+**We do not agree that the $Rt$ metric "assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time." If this was true, $Rt$ would be calculated by dividing growth (BAI) in the drought year by the long-term mean growth increment.
+Instead, $Rt$ is calculated by dividing growth in the focal year by the mean growth over a narrow window of preceding years (5 years in this study).
+Even though there may be a long-term directional trend in BAI, if you were to calculate the trend in BAI over each 5-year increment, the trend could go in any direction.
+To account for such trends, we tried an intervention time series analysis (ARIMA).
+Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
+In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
+We now include the following content on ARIMA in the manuscript:**
+
+(*insert text*)
+
 L260-261: This is rather strange.
@@
 
+*add specifics (changes to text; SI table/fig #s) to text below*
+
+**We tried an intervention time series analysis (ARIMA).
+This yielded similar results, which we now present in the Supplementary Info.
+The original $Rt$ and $Rt_{ARIMA}$ were fairly well correlated, but with some scatter (see figs below).
+To get a sense of which metric was performing better, we visually reviewed the records of the 18 trees with largest differences between $Rt$ and $Rt_{ARIMA}$.
+$Rt$ estimates were better for the majority of these, and generally were not hugely unrealistic, whereas $Rt_{ARIMA}$ could be way off (e.g., for several trees that were undergoing growth declines prior to drought and hit a minima at or near the drought, $Rt_{ARIMA}$ was much greater than 1, suggesting a substantial growth increase during the drought.).
+Thus, we decided to stick with $Rt$ as the primary metric.**
+
+![](tables_figures/Rt_arima_comparison_all_1966.png)
+![](tables_figures/Rt_arima_comparison_1977_1999.png)
+
 **Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.

### 2020-06-28 manuscript/climate_sensitivity_draft.Rmd
added text on ARIMA (needs some details)
#92

--- 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
@@
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
+Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA) that predicted drought-year growth based on trends over the past **10** years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
+We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
+
 Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.

### 2020-06-28 manuscript/response_to_reviews.Rmd
update response to reviews

@@
 In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
+While this unquestionably leads to biased estimates for some trees, we would not expect it to bias our analysis; rather, it is one of the reasons that the amount of variation that can be explained by analyses such as these is relatively low.
 We now include the following content on ARIMA in the manuscript:**
@@
 see my comment below on Rt>1).
+
+**We switched to use of Generalized linear mixed models. This did not change the outcome of any analyses.**
 

### 2020-06-28 manuscript/response_to_reviews.Rmd
Added text and code for #92, plus added full R package citations

@@
 
-**We tried an intervention time series analysis (ARIMA).
-This yielded similar results, which we now present in the Supplementary Info.
+**In response to Reviewer 4's comments, we tried an intervention time series analysis using the 'auto.arima' function from the 'forecast' package [@R-forecast].
+Specifically, we modeled BAI  based on the timeseries for the 10 years prior to the drought year (itself acting as the intervention).
+We then recorded the ratio between the mean predicted BAI and the observed BAI during the drought year, and compared these results to $Rt$.
+These yielded similar outcomes, which we now present in the Supplementary Info.
 The original $Rt$ and $Rt_{ARIMA}$ were fairly well correlated, but with some scatter (see figs below).
 To get a sense of which metric was performing better, we visually reviewed the records of the 18 trees with largest differences between $Rt$ and $Rt_{ARIMA}$.
-$Rt$ estimates were better for the majority of these, and generally were not hugely unrealistic, whereas $Rt_{ARIMA}$ could be way off (e.g., for several trees that were undergoing growth declines prior to drought and hit a minima at or near the drought, $Rt_{ARIMA}$ was much greater than 1, suggesting a substantial growth increase during the drought.).
+$Rt$ estimates were better for the majority of these, and generally were not hugely unrealistic, whereas $Rt_{ARIMA}$ could be way off (e.g., for several trees that were undergoing growth declines prior to drought and hit minima at or near the drought, $Rt_{ARIMA}$ was much greater than 1, suggesting a substantial growth increase during the drought.).
 Thus, we decided to stick with $Rt$ as the primary metric.**
 
-![](tables_figures/Rt_arima_comparison_all_1966.png)
-![](tables_figures/Rt_arima_comparison_1977_1999.png)
+![Figure S4](tables_figures/publication/figureS4_Rt_arima_comparison.png)
 

### 2020-06-28 manuscript/climate_sensitivity_SI.Rmd
Added rt-arima comparison figure
Added text and code for #92, plus added full R package citations

@@
     fig_caption: yes
+bibliography: ["packages.bib"]
 biblio-style: apalike
-csl: apa.csl
 header-includes:
@@
 \newpage
+While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes.
+As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others.
+While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.
+
+R-packages not already cited in the main manuscript include the following, listed alphabetically by corresponding package name:
+
+[@R-base; @R-broom; @R-car; @R-cowplot; @R-data.table; @R-devtools; @R-dplR; @R-dplyr; @R-extrafont; @R-ggplot2; @R-ggpubr; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lubridate; @R-MuMIn; @R-piecewiseSEM; @R-png; @R-purrr; @R-raster; @R-rasterVis; @R-RCurl; @R-readxl; @R-reshape2; @R-rgdal; @R-rgeos; @R-rmarkdown; @R-sf; @R-stringi; @R-stringr; @R-tidyr]
+
+\newpage
 ```{r Table S1, eval=TRUE, echo=FALSE, message=FALSE}
@@
 \newpage
+```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
+tableS4 <- read.csv("tables_figures/publication/tableS4_Rt_arima.csv", stringsAsFactors = FALSE)
+tableS4 <- tableS4[1:21, ]
+colnames(tableS4) <- c("Year", "Tree", "Rt", "Rt_ARIMA")
+kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Comparison of Rt and ARIMA Rt results, showing a sample of 7 trees. Full table can be found at THIS WEBSITE.") %>%
+  # pack_rows("", 1, 3, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 4, 6, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 7, 9, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 10, 12, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 13, 15, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 16, 18, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("", 19, 21, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+\newpage
 ![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/publication/figureS1_location_cored_trees.png){width=500px}
@@
 ![Figure S3: Height by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
+
+\newpage
+![Figure S4: Comparison of Rt and ARIMA results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)

### 2020-06-28 manuscript/climate_sensitivity_draft.Rmd
Added text and code for #92, plus added full R package citations

@@
 knitr::write_bib(c(
-  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea'
+  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea', 'forecast'
 ), file = 'packages.bib')
@@
 
-To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area.
+To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities].
 We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
@@
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA) that predicted drought-year growth based on trends over the past **10** years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past **10** years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
@@
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively.
+
+Other R-packages aside from those already listed were very helpful in conducting analyses. These are listed in the Supplementary Information.
 

### 2020-06-28 scripts/climate_sensitivity_analysis.R

@@ -129,7 +129,7 @@ for(j in seq(along=trees)){
       #3. Get the ratio between the observed and predicted
       areaobs <- area_dt[area_dt$year == droughts[i], ]
       diff <- round(areaobs$val / as.numeric(future$mean),2) #ratio
-      # diff <- round(areaobs$val / as.numeric(future$mean),2) #difference
+      # diff <- round(areaobs$val - as.numeric(future$mean),2) #difference
     } else {diff <- NA}
     
     inddrt <- data.frame(year = droughts[i],
@@ -137,37 +137,37 @@ for(j in seq(along=trees)){
                        diff = diff)
     indtree <- rbind(indtree, inddrt)
     
-    # Plotting arima results ####
-    # #3.1. Prepare data for plotting
-    # orig <- data.frame(year = seq(droughts[i]-10, droughts[i]-1, by=1),
-    #                    val = NA,
-    #                    diff=NA)
-    # areaobs <- rbind(orig[,1:2], areaobs)
-    # pred <- rbind(orig, pred)
-    
-    # #4. Plot / characterize change
-    # plot(area_dt$year, area_dt$val, ylim=c(min(pred$diff, na.rm=TRUE), 
-    #                                        max(area_dt$val, na.rm=TRUE)),
-    #      xlab="", ylab="")
-    # lines(area_dt$year, area_dt$val)
-    # points(areaobs$year, areaobs$val, col="blue") #observed
-    # lines(areaobs$year, areaobs$val, col="blue")
-    # points(pred$year, pred$val, col="red")        #predicted
-    # lines(pred$year, pred$val, col="red")
-    # points(pred$year, pred$diff, col="orange")    #difference
-    # lines(pred$year, pred$diff, col="orange")
-    # abline(v=droughts[i], lty=2)
-    # abline(h=mean(area_dt$val[area_dt$year<droughts[i]], na.rm=TRUE), 
-    #        lty=3, col="brown")
-    # 
-    # legend("bottomleft", 
-    #        legend=c("ARIMA data", "observed", "predicted", "difference"),
-    #        col=c("black", "blue", "red", "orange"),
-    #        pch=1,
-    #        cex=0.55, pt.cex=1, bty="n")
-    # title(main=paste0(droughts[i], " drought \nMagnitude mean BAI change = ", 
[…]

### 2020-07-04 manuscript/response_to_reviews.Rmd
Added content based on Reviewer 1 comments.

@@
 
-*dAIC: minor fix*
+**I have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **
 
@@
 
-*IM - I kind of understand what they meant, but more in terms of a logic flow position. I agree that this can pretty much be fixed*
+*IM - I've updated the language in the manuscript, I think it's better.*
 
@@
 
-- Line 113: correct ‘positively’.
+- Line 113: correct ‘positively’. **done**
 
-- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.
+- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’. **done**
 
@@
 should be rather straightforward and the authors would increase accuracy across species in height estimation.
+  - *IM I'm confused about how to answer this.
+There's no direct response to the first question, as the measurements I got (along with Atticus and previous measurements) were all lumped into the pot to make the regression equations, which we explain.
+Otherwise, I did go out and confirm measurements.
+I guess my only response (and question to you @teixeirak) is it worth explaining how disparate our data was and that's why we have different methods?
+Along with me going and getting new ones?*
 
-- Lines 166-167: please explain further the hypothesis and why.
+- Lines 166-167: please explain further the hypothesis and why. *IM this is now line 170. I'm a little confused myself.*
 
-- Line 222: what about climate in years before 1991?
+- Line 222: what about climate in years before 1991? *IM I don't get this. At the beginning of the paragraph we say we looked at droughts 1950-2009...*
 
-- Line 231: define AICc.
+- Line 231: define AICc. **This is done (with added reference)**
 
+- *IM I'm a bit confused by the next three comments.*
 - Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).
@@
 *this is true of crown position, not height*
+- *IM how do we want to officially respond to this?*
 
@@
 
+**IM I have done this (with arima ratio), see below. Assuming this should be SI plot?**
+![Rt (Arima ratio) across species](tables_figures/rt_across_species.png)
+
 - Line 331-333: not so sure about this.
 It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
+  - *IM Krista do you know more about this?*
 
-- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).
+- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated). *IM does this need a response?*
 
-- Line 348: non-drought.
+- Line 348: non-drought. *IM do we have the original submitted draft? I'm not sure what this is referring to*
 
 - Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
+  - *IM well, in our original course of thinking about it as based on Lloret, yes, Rt is a measure of a tree's particular resistance, and it fluctuates from year to year.
+Maybe the reviewer would have preferred if we had said that lower radial growth shows the tree was more vulnerable *in that year*?
+And then we reiterate that the focal drought years all had low Rt?*
+  - *That being said, we're now using arima ratio and all values of Rt are positive. Regardless, I don't see where we make the full claim the reviewer is referring to.*
 
 - Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.
+  - *this explains the previous comment*
 
 Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
+- *IM I added in a thing to the caption*
 
 Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
+- *IM Krista, isn't the number of cores synonymous for us with the sample size for each species?*
 
 Table 3: the authors could show some test to see if the traits presented are different among species.
+- *IM Krista, do you think this is necessary?
+This would require extra stats to say something like "x traits did not differ significantly across species." I'm not sure how that result would affect our conclusions/writing.*
 
-Table S1 and S2: correct R2.
+Table S1 and S2: correct R2. *IM I added the correct Latex notation but it's not rendering it when I knit the file...*
 
-Table S3: please, explain what ‘rank’ means.
+Table S3: please, explain what ‘rank’ means. *IM Krista, do you remember? Is this from the pointer year exercise? Or was this ranking the months in the context of each year's PDSI?*
 
@@
 The question is whether there was any stand dynamics involved in those height series that had to be included in the models.
+- *IM seems like this comment is meant for the paper than for Figure S3. What do you think?*
 

### 2020-07-04 manuscript/climate_sensitivity_SI.Rmd
Added content based on Reviewer 1 comments.

@@
 bark[,2] <- NULL
+colnames(bark) <- c("Species", "Equations", "$R^{2}$")
 kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
@@
 heights[,2] <- NULL
+colnames(heights) <- c("Species", "Equations", "$R^{2}$")
 kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
@@
 tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts") %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. Rank refers to ") %>%
   pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%

### 2020-07-04 manuscript/climate_sensitivity_draft.Rmd
Added content based on Reviewer 1 comments.

@@
 
-We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig.
+We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig.
 **S1**).
@@
 Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
-Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] ~ ln[DBH]$).
+Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] \sim ln[DBH]$).
 For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.
@@
 We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community.
-We identified three drought years: 1966, 1977, and 1999 (Figs.
-**1**, **S2**, Table S3).
-These were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+Using the pointRes package in R [@R-pointRes], we identified three drought years - 1966, 1977, and 1999 (Figs.
+**1**, **S2**, Table S3) - for two reasons.
+First, these were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017.
-These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011].
-Pointer years were identified using the pointRes package in R [@R-pointRes].
+Second, these were also years with widespread tree growth reduction, here defined as when >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years.
+This follows the drought resistance ($Rt$) metric of [@lloret_components_2011].
 In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3).
@@
 Mixed effects models were implemented in the lme4 package in R [@R-lme4].
-We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit, implmented in the AICcmodavg package in R [@R-AICcmodavg].
+We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implmented in the AICcmodavg package in R [@R-AICcmodavg].
+AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016]
 
@@
 In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence).
-Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 units relative to the corresponding null model lacking that variable (Table 4).
+Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 1 unit relative to the corresponding null model lacking that variable (Table 4).
 
 We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought year (Table 4).
-We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
+Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
 When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models.
@@
 In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
-In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4).
+In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table 4).
 Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

### 2020-07-04 manuscript/tables&figures.Rmd
Added content based on Reviewer 1 comments.

@@
 Summary of hypotheses, corresponding specific predictions, and results.
-We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.") %>%
+We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.
+A dash symbol indicates that the response was not significant in any of the top full models.") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
@@
 Models including each variable were compared to corresponding null models.
-dAIC is the AICc of the null model minus that of the model including the variable (thus, dAICc>2 indicates that the variable significantly improves the model).
+Delta AICc is the AICc of the null model minus that of the model including the variable (thus, Delta AICc > 1 indicates that the variable significantly improves the model).
 Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
@@
 ```
-*dAICc > 1: variable qualified for inclusion in full model
-
-**dAICc > 2: statistically signficant, variable qualified for inclusion in full model
+*$\Delta$AICc > 1: variable qualified for inclusion in full model
 
@@
 Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (Delta AICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-07-04 scripts/climate_sensitivity_analysis.R

@@ -788,7 +788,7 @@ write.csv(cand_full, "manuscript/tables_figures/publication/S3_candidate_traits_
 # the best full model. In other words, we have seen which variables across all
 # scenarios have the most influence. Now, we test only those variables against
 # each other, again across all scenarios. The loop populates the table (best_mod_traits)
-# with the results AND identifies the top models for each scenario that are <=2 dAIC 
+# with the results AND identifies the top models for each scenario that are <=1 dAIC 
 # (top_models) from the best (best_mod_traits).
 
 best_mod_traits <- data.frame("best_model" = NA,
@@ -1096,6 +1096,14 @@ write.csv(res_full, "manuscript/tables_figures/publication/tableS7_top_residual_
 
 
 #
+##3e. Compare Rt values across species ####
+library(ggplot2)
+png("manuscript/tables_figures/rt_across_species.png", width=960)
+ggplot(trees_all_subar) +
+  aes(x = sp, y = resist.value) +
+  geom_boxplot(fill = "#0c4c8a") +
+  theme_minimal()
+dev.off()
 ##3a original. test each variable individually (expand for fuller explanation) ####
 # this code chunk tests each variable individually for each drought scenario,
 # using a predefined null model and test model (i.e. all test models have height).

### 2020-07-05 manuscript/response_to_reviews.Rmd
Corrected the figure for Rt across sp
Switched to Rt (mistakenly did arima ratio before)

@@
 
-**IM I have done this (with arima ratio), see below. Assuming this should be SI plot?**
-![Rt (Arima ratio) across species](tables_figures/rt_across_species.png)
+**IM I have done this for Rt, see below.
+I notice we don't have negative at all, just above or below 0.
+I'm assuming this is what the reviewer is referring to?
+Assuming this should be SI plot?
+Please let me know if you think this figure should be altered at all.
+**
+![Rt across species](tables_figures/rt_across_species.png)
 

### 2020-07-05 scripts/climate_sensitivity_analysis.R

@@ -586,7 +586,7 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #3. mixed effects model for output of #2.
 
 ##start here if just re-running model runs ####
-trees_all_subar <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
+# trees_all_subar <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
 # trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arima.csv", stringsAsFactors = FALSE)
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv",
                           stringsAsFactors = FALSE)
@@ -1099,7 +1099,7 @@ write.csv(res_full, "manuscript/tables_figures/publication/tableS7_top_residual_
 ##3e. Compare Rt values across species ####
 library(ggplot2)
 png("manuscript/tables_figures/rt_across_species.png", width=960)
-ggplot(trees_all_subar) +
+ggplot(trees_all_sub) +
   aes(x = sp, y = resist.value) +
   geom_boxplot(fill = "#0c4c8a") +
   theme_minimal()

### 2020-07-05 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Wider xylem conduits make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they are exposed to higher solar radiation, greater wind speeds, and lower relative humidity (e.g., @koike_leaf_2001; @kunert_revised_2017).
+Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].

### 2020-07-05 non_publication/original_bookdown_files/04-results.Rmd
Moved files to archive folder

--- 2019-07-31 manuscript/04-results.Rmd
@@
 cof_bio %>%
-  kable("html", align = 'clc', caption = 'Biophysical Model') %>%
+  kable("latex", align = 'clc', caption = 'Biophysical Model') %>%
     kable_styling(full_width = F, position = "float_left")
@@
 cof_full %>%
-  kable("html", align = 'clc', caption = 'Full Model') %>%
+  kable("latex", align = 'clc', caption = 'Full Model') %>%
     kable_styling(full_width = F, position = "right")

### 2020-07-05 scripts/graphs_for_manuscript.R
Fixed Figure 1 as per #87

@@ -41,8 +41,8 @@ Mode <- function(x) {
    ux[which.max(tabulate(match(x, ux)))]
 }
 
-apply(d, 2, Mode)
-test[,.(mode = Mode(resist.value)), by=.(year)]
+# apply(d, 2, Mode)
+# test[,.(mode = Mode(resist.value)), by=.(year)]
 
 #save as png so that way you're arranging only png
 png("manuscript/tables_figures/publication/Figure1b_density_plot.png", res = 300, width = 150, height = 50, units = "mm", pointsize = 10)
@@ -52,14 +52,20 @@ ggplot(trees_all_sub) +
    geom_vline(xintercept=1) +
    scale_fill_hue(labels=c("1966", "1977", "1999")) +
    labs(x="resistance value") +
-   theme_minimal()
+   theme_minimal() +
+   annotate(geom="text", x=0.1, y=1.5, 
+            label = "(b)", fontface="bold", size=7)
 dev.off()
 
 plot1 <- readPNG("manuscript/tables_figures/publication/Figure1a_Time_series_for_each_species.png")
 plot2 <- readPNG("manuscript/tables_figures/publication/Figure1b_density_plot.png")
 
 png("manuscript/tables_figures/publication/Figure1.png", res=300, height=200, width=150, units="mm", pointsize=10)
-plot_grid(rasterGrob(plot1), rasterGrob(plot2), align = "v", nrow = 2, rel_heights = c(3/4, 1/4), axis = "b")
+g <- plot_grid(rasterGrob(plot1), rasterGrob(plot2), align = "v", nrow = 2, rel_heights = c(3/4, 1/4), axis = "b")
+g + annotate(geom="text", x=0.25, y=0.99, 
+             label = "(a)", fontface="bold", size=3) +
+   annotate(geom="text", x=0.25, y=0.23, 
+             label = "(b)", fontface="bold", size=3)
 dev.off()
 
 #########################################################################

### 2020-07-05 scripts/plot_time_series.R
Moved file out of archive, updated script

@@ -30,7 +30,7 @@ colors.species <- colorRampPalette(c("purple", "cadetblue", "darkorange", "red",
 
 clim <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/data/climate/Formated_CRU_SCBI_1901_2016.csv")
 
-clim_PDSI <- read.csv("data/pdsi/pdsi_timeseries_winchester_1949-2012.csv", stringsAsFactors = F)
+clim_PDSI <- read.csv("data/climate/pdsi_timeseries_winchester_1949-2012.csv", stringsAsFactors = F)
 clim_PDSI$year <- sapply(strsplit(clim_PDSI$Date, "/"), "[[", 3)
 clim_PDSI$month <- sapply(strsplit(clim_PDSI$Date, "/"), "[[", 1)
 names(clim_PDSI) <- c("Date", "pdsi", "year", "month")

### 2020-07-05 scripts/graphs_for_manuscript.R
Fully fixed figure lettering

@@ -62,9 +62,9 @@ plot2 <- readPNG("manuscript/tables_figures/publication/Figure1b_density_plot.pn
 
 png("manuscript/tables_figures/publication/Figure1.png", res=300, height=200, width=150, units="mm", pointsize=10)
 g <- plot_grid(rasterGrob(plot1), rasterGrob(plot2), align = "v", nrow = 2, rel_heights = c(3/4, 1/4), axis = "b")
-g + annotate(geom="text", x=0.25, y=0.99, 
+g + annotate(geom="text", x=0.13, y=0.99, 
              label = "(a)", fontface="bold", size=3) +
-   annotate(geom="text", x=0.25, y=0.23, 
+   annotate(geom="text", x=0.13, y=0.23, 
              label = "(b)", fontface="bold", size=3)
 dev.off()
 

### 2020-07-05 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
#95

@@
 In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence).
-Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 1 unit relative to the corresponding null model lacking that variable (Table 4).
+**Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 1 unit relative to the corresponding null model lacking that variable (Table 4).
+**
 
 We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year and had a consistent direction of response across all droughts (Table 4).
 We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".

### 2020-07-05 manuscript/climate_sensitivity_draft.Rmd
fixed AIC significance error (not sure how this came in)

@@
 In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence).
-**Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 1 unit relative to the corresponding null model lacking that variable (Table 4).
-**
+Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2.0 relative to the corresponding null model lacking that variable (Table 4).
 

### 2020-07-05 manuscript/response_to_reviews.Rmd
update candidate variable criteria in manuscript, respond to related reviewer concern
#95

@@
 
-**I have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **
+**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **
 
@@
 See also my other comment above.
+
+**We have modified the text here to read:
+"Candidate variables were selected, based on the single-variable tests, as those that had a consistent direction of response across all droughts and whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of signficance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models."**
 

### 2020-07-05 manuscript/climate_sensitivity_draft.Rmd
update candidate variable criteria in manuscript, respond to related reviewer concern
#95

@@
 We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year and had a consistent direction of response across all droughts (Table 4).
+Candidate variables were selected, based on the single-variable tests, as those that had a consistent direction of response across all droughts and whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of signficance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
 We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".

### 2020-07-05 scripts/climate_sensitivity_analysis.R
Added notes and did more for VIF

@@ -586,8 +586,7 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #3. mixed effects model for output of #2.
 
 ##start here if just re-running model runs ####
-# trees_all_subar <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
-# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arima.csv", stringsAsFactors = FALSE)
+# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv",
                           stringsAsFactors = FALSE)
 x1966 <- trees_all_sub[trees_all_sub$year == 1966, ]
@@ -613,7 +612,11 @@ library(data.table)
 # using a predefined null model and test model (i.e. all test models have height).
 # After these are tested from a defined data frame (sum_mod_traits), the "candidate"
 # variables (cand_full) are those variables that appear in one of the 4 scenarios'
-# top models. These candidates are then used in ##6b.
+# top models. These candidates are then used in ##3b.
+
+# The "reform" is from Issue #95 in Github, where we decide to include the TWI*height 
+# interaction plus crown position in the base model that we compare against the other 
+# traits. Prior to this, the base model only had height.
 
 sum_mod_traits <- data.frame(
   "prediction" = c(3.1, 3.2, 3.3, 3.4, 3.5),
@@ -791,12 +794,15 @@ write.csv(cand_full, "manuscript/tables_figures/publication/S3_candidate_traits_
 # with the results AND identifies the top models for each scenario that are <=1 dAIC 
 # (top_models) from the best (best_mod_traits).
 
+# For an explanation on what "reform" means, see the description under ##3a. 
+
 best_mod_traits <- data.frame("best_model" = NA,
                               "scenario" = c("all droughts", "1966", "1977", "1999")
 )
-top_vars <- c(unique(cand_full$variable), 
-              paste0(unique(cand_full$variable)[1], "+",
-                     unique(cand_full$variable)[2]))
+
+## ONLY KEEP PLA and TLP as top variables! See Issue #95 on github.
[…]

### 2020-07-07 manuscript/tables&figures.Rmd
Updated tables&figures

@@
   add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
-                     "xylem porosity" = 2," ")) %>%
+                     " "," ")) %>%
   pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
@@
   pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 4, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 4, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%

### 2020-07-07 scripts/climate_sensitivity_analysis.R
Update climate_sensitivity_analysis.R

@@ -529,8 +529,8 @@ trees_all$TWI.ln <- log(trees_all$TWI)
 ##fram 140939 has been mislabeled. It is recorded as having a small dbh when that is the second stem. In terms of canopy position, though, it fell between time of coring and when positions were recorded, thus we do not know its position.
 trees_all <- trees_all[!trees_all$tree == 140939, ]
 
-##2i. if using resistance value, constrain values to be <=2
-# trees_all <- trees_all[trees_all$resist.value <=2]
+##2i. if using resistance value, constrain values to be <=2 ####
+trees_all <- trees_all[trees_all$resist.value <=2]
 
 # write.csv(trees_all, "manuscript/tables_figures/trees_all.csv", row.names=FALSE)
 ##2i. prepare dataset for running regression models ####
@@ -586,9 +586,9 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #3. mixed effects model for output of #2.
 
 ##start here if just re-running model runs ####
-# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
-trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv",
-                          stringsAsFactors = FALSE)
+trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
+# trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv", stringsAsFactors = FALSE)
+
 x1966 <- trees_all_sub[trees_all_sub$year == 1966, ]
 x1977 <- trees_all_sub[trees_all_sub$year == 1977, ]
 x1999 <- trees_all_sub[trees_all_sub$year == 1999, ]
@@ -783,8 +783,8 @@ for (i in seq(along=sum_mod_traits[,c(8,11,14,17)])){
 
 cand_full <- cand_full[complete.cases(cand_full), ]
 
-write.csv(sum_mod_traits, "manuscript/tables_figures/tested_traits_all_reform.csv", row.names=FALSE)
-write.csv(cand_full, "manuscript/tables_figures/publication/S3_candidate_traits_reform.csv", row.names=FALSE)
+write.csv(sum_mod_traits, "manuscript/tables_figures/tested_traits_all_reform_arimaratio.csv", row.names=FALSE)
[…]

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
remove current table S4
#92

@@
 \newpage
-```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
-tableS4 <- read.csv("tables_figures/publication/tableS4_Rt_arima.csv", stringsAsFactors = FALSE)
-tableS4 <- tableS4[1:21, ]
-colnames(tableS4) <- c("Year", "Tree", "Rt", "Rt_ARIMA")
-kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Comparison of Rt and ARIMA Rt results, showing a sample of 7 trees. Full table can be found at THIS WEBSITE.") %>%
-  # pack_rows("", 1, 3, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 4, 6, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 7, 9, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 10, 12, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 13, 15, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 16, 18, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("", 19, 21, latex_gap_space = ".4em", colnum = 4, hline_before = FALSE) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-\newpage
 ![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/publication/figureS1_location_cored_trees.png){width=500px}

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
added tables S4, S5
#92

@@
 \newpage
+```{r table S4, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
+Single-variable tests of hypothesized drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
+Models including each variable were compared to corresponding null models.
+Delta AICc is the AICc of the null model minus that of the model including the variable (thus, Delta AICc > 1 indicates that the variable significantly improves the model).
+Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
+  kable_styling()
+```
+*$\Delta$AICc > 1: variable qualified for inclusion in full model
+
+\newpage
+```{r table S5, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+multivar_table <- read.csv("tables_figures/publication/tableS5_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+kable(multivar_table, booktabs = TRUE, caption = "Table S5.
+Summary of top full models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
+Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (Delta AICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
+                     " "," ")) %>%
+  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 4, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  row_spec(1, bold = T) %>%
+  row_spec(2, bold = T) %>%
+  row_spec(3, bold = T) %>%
+  row_spec(4, bold = T) %>%
+  kable_styling()
+```
+
+\newpage
 ![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/publication/figureS1_location_cored_trees.png){width=500px}

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
minor updates to tables 4-5 and S4-S5

@@
 kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
-Single-variable tests of hypothesized drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
+Individual trait tests of hypothesized drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
 Models including each variable were compared to corresponding null models.
@@
 ```
-*$\Delta$AICc > 1: variable qualified for inclusion in full model
+**$\Delta$AICc > 2: variable considered significant as an individual predictor
 

### 2020-07-09 manuscript/tables&figures.Rmd
minor updates to tables 4-5 and S4-S5

@@
 kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
-Single-variable tests of hypothesized drivers of drought resistance.
+Individual trait tests of hypothesized drivers of drought resistance ($Rt$ metric).
 Models including each variable were compared to corresponding null models.
@@
 *$\Delta$AICc > 1: variable qualified for inclusion in full model
+**$\Delta$AICc > 2: variable qualified for inclusion in full model and considered significant as an individual predictor
 

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
fixed row packing/ bold

@@
   pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 4, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 3, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
@@
   row_spec(3, bold = T) %>%
-  row_spec(4, bold = T) %>%
+  row_spec(5, bold = T) %>%
   kable_styling()

### 2020-07-09 manuscript/climate_sensitivity_draft.Rmd
update model description in text to GLMM
#94

@@
 Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
-The general statistical model for hypothesis testing was a mixed effects model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-Mixed effects models were implemented in the lme4 package in R [@R-lme4].
-We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implmented in the AICcmodavg package in R [@R-AICcmodavg].
+The general statistical model for hypothesis testing was a generalized linear mixed model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
+Generalized linear mixed models were implemented in the **XX** package in R [**REF**].
+We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].
 AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016]
@@
 Candidate variables were selected, based on the single-variable tests, as those that had a consistent direction of response across all droughts and whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
-We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of signficance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
+We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
 We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
fixes to Tables 4, S4
- removed null model column (not needed)
- fixed/ clarified meaning of dAICc thresholds
- improved table captions

@@
 kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
-Individual trait tests of hypothesized drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
-Models including each variable were compared to corresponding null models.
-Delta AICc is the AICc of the null model minus that of the model including the variable (thus, Delta AICc > 1 indicates that the variable significantly improves the model).
+Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
+Delta AICc is the AICc of a model excluding the trait minus that of the model including it.
 Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
   pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%

### 2020-07-09 manuscript/tables&figures.Rmd
fixes to Tables 4, S4
- removed null model column (not needed)
- fixed/ clarified meaning of dAICc thresholds
- improved table captions

--- 2020-07-07 manuscript/tables&figures.Rmd
@@
 kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
-Single-variable tests of hypothesized drivers of drought resistance.
-Models including each variable were compared to corresponding null models.
-Delta AICc is the AICc of the null model minus that of the model including the variable (thus, Delta AICc > 1 indicates that the variable significantly improves the model).
+Individual tests of species traits as drivers of drought resistance ($Rt$ metric).
+Delta AICc is the AICc of a model excluding the trait minus that of the model including it.
 Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  # pack_rows("Tree size and microenvironment", 1, 12, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
+  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
   pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
+  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doesn't work
   kable_styling()
 ```
-*$\Delta$AICc > 1: variable qualified for inclusion in full model
+*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
+
+**$\Delta$AICc > 2: variable is considered significant as an individual predictor (and meets $\Delta$AICc criterion for inclusion in full model)
 

### 2020-07-09 manuscript/climate_sensitivity_SI.Rmd
replaced dAICc with ∆AICc in tables and table legends (now footnotes)

@@
 tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. Rank refers to ") %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. ") %>%
   pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
@@
 ```
+Rank refers to ...**XX**
 
@@
 Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
-Delta AICc is the AICc of a model excluding the trait minus that of the model including it.
-Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
+", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 ```
+Variable abbreviations are as in Table 2. 
+$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
+
 **$\Delta$AICc > 2: variable considered significant as an individual predictor
@@
 Summary of top full models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (Delta AICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
+", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 ```
+Models are ranked by AICc. Shown are all models whose AICc value falls within 1.0 ($\Delta$AICc<1) of the best model (bold).
 

### 2020-07-09 manuscript/tables&figures.Rmd
replaced dAICc with ∆AICc in tables and table legends (now footnotes)

@@
 singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
-kable(singlevar_table, booktabs = TRUE, caption = "Table 4.
-Individual tests of species traits as drivers of drought resistance ($Rt$ metric).
-Delta AICc is the AICc of a model excluding the trait minus that of the model including it.
-Variable abbreviations are as in Table 2.", format = 'latex', escape = FALSE) %>%
+kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Individual tests of species traits as drivers of drought resistance ($Rt$ metric).", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 ```
+Variable abbreviations are as in Table 2. 
+$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
+
 *$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
@@
 multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(multivar_table, booktabs = TRUE, caption = "Table 5.
-Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (Delta AICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>%
+kable(multivar_table, booktabs = TRUE, caption = "Table 5. Summary of top full models for each drought instance. ", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 ```
-
-<!-- \clearpage -->
-<!-- ```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE} -->
-<!-- # *****this is for if doing arimaratio!!!!***** -->
-<!-- library(knitr) -->
-<!-- library(kableExtra) -->
-<!-- multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
--->
-<!-- kable(multivar_table, booktabs = TRUE, caption = "Table 5.
-Summary of top full models for each drought instance.
-Models are ranked by AICc, and we show all models whose AICc value falls within 1.0 (dAICc<1) of the best model (bold).", format = 'latex', escape = FALSE) %>% -->
-<!--   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>% -->
-<!--   add_header_above(c(" "," "," "," "," ", " ", " ", "xylem porosity" = 2," ", " ")) %>% -->
-<!--   pack_rows("", 1, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
-<!--   pack_rows("", 3, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
-<!--   pack_rows("", 6, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
-<!--   pack_rows("", 8, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>% -->
-<!--   row_spec(1, bold = T) %>% -->
-<!--   row_spec(3, bold = T) %>% -->
-<!--   row_spec(6, bold = T) %>% -->
-<!--   row_spec(8, bold = T) %>% -->
-<!--   kable_styling() -->
-<!-- ``` -->
+Models are ranked by AICc. Shown are all models whose AICc value falls within 1.0 ($\Delta$AICc<1) of the best model (bold).
 

### 2020-07-09 manuscript/climate_sensitivity_draft.Rmd
minor updates to analysis methods

@@
 
-Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable.
-The general statistical model for hypothesis testing was a generalized linear mixed model with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects.
-Generalized linear mixed models were implemented in the **XX** package in R [**REF**].
+Analyses focused on testing the predictions presented in Table 1, with $Rt$ (or $Rt_{ARIMA}$) as the response variable.
+Models were run for all drought years combined and for each drought year individually.
+The general statistical model for hypothesis testing was a generalized linear mixed model (GLMM), implemented  in the **XX** package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables including drought year (multi-drought model only), $ln[H]$*$TWI$, crown position, and 1-2 species traits (see below).
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
-AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016]
+AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].
 
-Models were run for all drought years combined and for each drought year individually.
-In order to determine the relative importance of each predictor variable individually, we first implemented models with the variable in question as a fixed effect, along with drought year (for model with all drought years combined) and $ln[H]$ (included in null models because of it's substantial influence).
-Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2.0 relative to the corresponding null model lacking that variable (Table 4).
+To avoid over-fitting models with five species traits (Table 2) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
+These tests followed the model structure specified above.
+Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
 
-We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined.
-Candidate variables were selected, based on the single-variable tests, as those that had a consistent direction of response across all droughts and whose addition to a corresponding null model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
-We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
-We compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models".
+We then determined the best full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
+To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc). 
 When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models.

### 2020-07-09 manuscript/tables&figures.Rmd
update table 2 to include Rt_ARIMA
@mcgregorian1 , please confirm that sample size for Rt_ARIMA is correct ( I put same as Rt, but that might not be true)

#92

@@
   column_spec(4, width = "7cm") %>%
-  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Dependent variables", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Independent variables", 3, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("tree size", 6, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("microhabitat", 8, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("species' traits", 13, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()

### 2020-07-09 manuscript/tables&figures.Rmd
table 2 header

--- 2020-07-09 manuscript/tables&figures.Rmd
@@
 table2 <- read.csv("tables_figures/publication/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2. Summary of variables.", 
+kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2. Summary of dependent and independent variables examined here, along with units, definitions, and sample sizes.", 
         escape=FALSE) %>%

### 2020-07-09 manuscript/tables&figures.Rmd
worked on Table 1
This cannot be finalized until we resolve issue #99

@@
 
+\blandscape
 ```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1.
-Summary of hypotheses, corresponding specific predictions, and results.
-We count predictions as fully supported ('yes') when the response is signficant in single-variable tests (Table 4) and included in all top full models and as partially supported ('(yes)') or rejected ('(no)') when the direction of response consistently matched the prediction but the effect was not significant in all models.
-A dash symbol indicates that the response was not significant in any of the top full models.") %>%
+kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
@@
   column_spec(2:5, width = "1.2cm") %>%
-  column_spec(6, width = "2.4cm") %>%
-  kableExtra::landscape()
+  column_spec(6, width = "2.4cm") 
+ # kableExtra::landscape()
 ```
+
+Parentheses indicate that the prediction was supported by one but not all of the top full models (Table **5**).
+Dash symbols indicate that the response was not significant, or not represented in any of the top full models.
+Asterisks indicate that the $Rt_{ARIMA}$ model yielded a different result: 
+*different significance in $Rt_{ARIMA}$ model
+**$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no signficant trend in $Rt$ model).
+
+**TILDAS (~) indicate that we don't currently have a (full) test of this prediction ([issue 99](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)**)
+
+\elandscape
 

### 2020-07-10 manuscript/response_to_reviews.Rmd
worked on reviewer responses
all of R1 (although upen issues remain)

#97

--- 2020-07-04 manuscript/response_to_reviews.Rmd
@@
 
-**We switched to use of Generalized linear mixed models. This did not change the outcome of any analyses.**
+**We switched to use of generalized linear mixed models. This did not change the outcome of any analyses.**
 
@@
 
+[*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
 Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models.
@@
 
-*This is an interesting suggestion.
-Depending on where we end up going with some other issues, it may make sense to de-emphasize the individual models--perhaps move Table 5 to SI and replace with figure ([issue #71](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71))?
-However, I think we could also just make an argument for leaving this as is.*
+*How about this: *
+
+*(1) keep a version of table 5 in the main text, showing just the top model for Rt, and maybe also that of $Rt_{ARIMA}$.  *
+
+*(2) put full versions of table 2 (for Rt and $Rt_{ARIMA}$), including all models within the top dAICc=2, in the appendix*
 
 related to [*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+
+related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)
+
+**We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations.
+However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
+This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
+(Removed was ring porosity, which had opposite effects in different droughts.)**
 
@@
 
+*This table is definitely not just exploratory, and is the source for supporting/ rejecting predictions in Table 1.
+Thus, keeping it in the main text could be justified.
+That said, it presents mainly negative results, and I'd like to reduce the number of tables, so moving it to the appedix would be fine.*
+
 [*issue #95*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/95)
+
+related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)
 
@@
 
-*Ugghh, this doesn't even change anything--we'd still be analyzing the same 3 droughts!
-The question is whether to remove the growth reduction criteria.
-I personally think that analyzing 2 criteria is a strength, as our goal was to identify droughts that were both climatologically and biologically signficant.
-In any case, this is just an issue of changing the wording.*
-
-*IM - I've updated the language in the manuscript, I think it's better.*
+**We have made the suggested change: we now present only the PDSI criteria in the methods, and mention the fact that the selected years were also the ones with largest growth reductions during the study period in the results section.
+Using just the PDSI criterion does not change the years analyzed.**
 
@@
 
-- Line 113: correct ‘positively’. **done**
+- Line 113: correct ‘positively’.
 
-- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’. **done**
+**done**
+
+- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.
+
+**done**
 
@@
 should be rather straightforward and the authors would increase accuracy across species in height estimation.
-  - *IM I'm confused about how to answer this.
-There's no direct response to the first question, as the measurements I got (along with Atticus and previous measurements) were all lumped into the pot to make the regression equations, which we explain.
-Otherwise, I did go out and confirm measurements.
-I guess my only response (and question to you @teixeirak) is it worth explaining how disparate our data was and that's why we have different methods?
-Along with me going and getting new ones?*
 
-- Lines 166-167: please explain further the hypothesis and why. *IM this is now line 170. I'm a little confused myself.*
+**We have specified how many species used the interspecific allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be possible to go collect more data for these species.**
 
-- Line 222: what about climate in years before 1991? *IM I don't get this. At the beginning of the paragraph we say we looked at droughts 1950-2009...*
+- Lines 166-167: please explain further the hypothesis and why.
 
-- Line 231: define AICc. **This is done (with added reference)**
+**We have reworded this sentence to read, "While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative."**
 
-- *IM I'm a bit confused by the next three comments.*
+- Line 222: what about climate in years before 1991?
+
+**We no longer mention 1991, as it does not meet the PDSI criteria.**
+
+- Line 231: define AICc.
+
+**This is done (with added reference)**
+
 - Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).
+
+**This section has been re-written.**
 
@@
 
+**We have modified this text to read as follows:**
+
+**"Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models."**
+
 - Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.
+
+**As stated above, we have removed the $Rt$ criterion from the drought identification process, and now present this information for the first time in this section, adding the following statement: "Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$."**
 
@@
 
+**It's not unprecedented to see some trees exhibit increased growth during drought, and not surprising.
+Stand dynamics and other factors result in a lot of scatter in the relationship between tree growth and climate, so while the distribution of $Rt$ in Fig.
+1b was shifted left (normal year should be normally distributed around 1), there still are some trees with increased growth.
+Our analysis helps to work out which trees these were (often shorter trees).**
+
+**We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
+1b), potentially due to decreased leaf area of competitors during the drought, and consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al.
+2015).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**
+
 - Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.
 
-*this is true of crown position, not height*
-- *IM how do we want to officially respond to this?*
+**It is true that crown position estimates become more accurate in later dates, but height should be equally reliable across years, having been estimated based on reconstructed DBH.
+This was not previously clearly specified in the methods.
+To clarify, we have added the following statement: "We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$."**
+
+**We agree that the assessment of crown position is somewhat problematic, both because of its increasing inaccuracy going back through time, and because of collinearity with $H$.
+However, given that it is both biologically interesting and comes out in our top models, we have chosen to include it.
+We note that this makes the test of our hypothesis conservative.
+We are transparent about this issue, to which we devote a paragraph in the discussion (par.
+3).**
 
@@
 
+*Regarding Fig.
+2, my current inclination is to keep it While I think its true that such profiles are common, I'm not currently aware of a lot of studies showing such profiles.
+I just prepared a grant proposal related to this, and current intern Nidhi Vinod is working on a review on vertical profiles in temperature, so I'd hope I'd be relatively up to speed...
+I'll check into this more and come back to it.*
+
 [*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
@@
 
-**IM I have done this (with arima ratio), see below. Assuming this should be SI plot?**
-![Rt (Arima ratio) across species](tables_figures/rt_across_species.png)
+*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
+1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
 
@@
 It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
-  - *IM Krista do you know more about this?*
 
-- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated). *IM does this need a response?*
+*Tree-ring data inherently have a lot of variation. Neil could probably write a better response here, with refs.*
 
-- Line 348: non-drought. *IM do we have the original submitted draft? I'm not sure what this is referring to*
+- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).
+
+**Please see response above.**
+
+- Line 348: non-drought.
+
+**Fixed.**
 
 - Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
-  - *IM well, in our original course of thinking about it as based on Lloret, yes, Rt is a measure of a tree's particular resistance, and it fluctuates from year to year.
-Maybe the reviewer would have preferred if we had said that lower radial growth shows the tree was more vulnerable *in that year*?
-And then we reiterate that the focal drought years all had low Rt?*
-  - *That being said, we're now using arima ratio and all values of Rt are positive. Regardless, I don't see where we make the full claim the reviewer is referring to.*
+
+**We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here).
+We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019]."**
 
 - Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.
-  - *this explains the previous comment*
+
+**See above comment.
+We have reworded this sentence to read, "In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought."**
 
 Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
-- *IM I added in a thing to the caption*
+
+**This is now clarified in the table footnote.**
 
 Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
-- *IM Krista, isn't the number of cores synonymous for us with the sample size for each species?*
+
+**In this study, we have one core per tree, now specified in the methods. We have changed the table header to "n trees".**
 
 Table 3: the authors could show some test to see if the traits presented are different among species.
-- *IM Krista, do you think this is necessary?
-This would require extra stats to say something like "x traits did not differ significantly across species." I'm not sure how that result would affect our conclusions/writing.*
 
-Table S1 and S2: correct R2. *IM I added the correct Latex notation but it's not rendering it when I knit the file...*
+**maybe include these in fig with $Rt$ across species? [*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)*
 
-Table S3: please, explain what ‘rank’ means. *IM Krista, do you remember? Is this from the pointer year exercise? Or was this ranking the months in the context of each year's PDSI?*
+Table S1 and S2: correct R2.
+
+*IM I added the correct Latex notation but it's not rendering it when I knit the file... (KAT: I just tried and wasn't able to get it working.*
+
+Table S3: please, explain what ‘rank’ means.
+
+**Done.**
 
@@
 The question is whether there was any stand dynamics involved in those height series that had to be included in the models.
-- *IM seems like this comment is meant for the paper than for Figure S3. What do you think?*
+
+**We recognize the importance of stand dynamics, and have added specific mention to the discussion: "It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation,...
+"**
 
@@
 
+**Thank you.**
+
 105 this sounds great, but was a little unclear.
@@
 173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.
+
+**Thank you.**
 

### 2020-07-10 manuscript/climate_sensitivity_SI.Rmd
worked on reviewer responses
all of R1 (although upen issues remain)

#97

@@
 bark[,2] <- NULL
-colnames(bark) <- c("Species", "Equations", "$R^{2}$")
+colnames(bark) <- c("Species", "Equations", "$R^2$")
 kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
@@
 \newpage
-```{r Table S2, eval=TRUE, echo=FALSE, message=FALSE}
+```{r Table S2, eval=TRUE, echo=FALSE,  warning=FALSE}
 library(knitr)
@@
 heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
-heights[,2] <- NULL
-colnames(heights) <- c("Species", "Equations", "$R^{2}$")
-kable(heights, booktabs=TRUE, format = "latex", caption = "Table S2: Species-specific height regression equations") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S2: Species-specific height regression equations", escape = FALSE) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
 ```
@@
 tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts. ") %>%
-  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
+") %>%
+  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
   pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("other", 13, 16, latex_gap_space = ".4em", colnum = 3, hline_before = FALSE) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
 ```
-Rank refers to ...**XX**
 

### 2020-07-10 manuscript/climate_sensitivity_draft.Rmd
minor text rearrangement in methods
worked on reviewer responses
all of R1 (although upen issues remain)

#97

@@
 **S1**).
-Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017.
+Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017.
 In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013].
@@
 
-For each cored tree, we combined tree-ring records and allometric equations of bark thickness to retroactively calculate DBH for the years 1950-2009. 
+For each cored tree, we combined tree-ring records and allometric equations of bark thickness to reconstruct DBH for the years 1950-2009. 
 Prior DBH was estimated using the following equation:
@@
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression on log-transformed data to relate bark thickness to diameter inside bark from 2008 data (Table S1), which were then used to determine bark thickness in the retroactive calculation of DBH.
+Specifically, we used linear regression on log-transformed data to relate $r_{bark}$ to diameter inside bark from 2008 data (Table S1), which were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
 
@@
 Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] \sim ln[DBH]$).
-For species with insufficient height data to create reliable species-specific allometries, heights were calculated from an equation developed by combining the height measurements across all species.
+For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species.
+We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$.
 
 Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
-While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative.
+While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative.
 An analysis of crown position relative to height (Fig.
@@
 
-We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community.
-Using the pointRes package in R [@R-pointRes], we identified three drought years - 1966, 1977, and 1999 (Figs.
-**1**, **S2**, Table S3) - for two reasons.
-First, these were the three years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions. 
+Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Second, these were also years with widespread tree growth reduction, here defined as when >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years.
-This follows the drought resistance ($Rt$) metric of [@lloret_components_2011].
-In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3).
-Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp.
-in the area from approximately 1988 through 1995 [@twery_effects_1991].
+Based on this, we identified three drought years - 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3).
 
@@
 
-For each drought year, we calculated drought resistance ($Rt$) as the ratio of $BAI$ during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
@@
 All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. 
+Other R-packages aside from those already listed were very helpful in conducting analyses. These are listed in the Supplementary Information. 
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively.
-
-Other R-packages aside from those already listed were very helpful in conducting analyses. These are listed in the Supplementary Information.
 
@@
 **1b**).
-In each drought, roughly 30% of the cored trees had growth reductions of >30% ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
+Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$.
+Specifically, in each drought, roughly 30% of the cored trees had growth reductions of $\ge$ 30% ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999.
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
@@
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-**1b**), potentially due to decreased leaf area of competitors during the drought.
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).
+**1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation: *11-13%* for all droughts combined, and *21-26%* for each individual drought (Table 5).
 
@@
 **2d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
-Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
+Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
 
@@
 @rodriguez-caton_long-term_2015].
-In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
+In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
 

### 2020-07-10 manuscript/tables&figures.Rmd
worked on reviewer responses
all of R1 (although upen issues remain)

#97

@@
 singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
-kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Individual tests of species traits as drivers of drought resistance ($Rt$ metric).", format = 'latex', escape = FALSE) %>%
+kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Individual tests of species traits as drivers of drought resistance ($Rt$).", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
   add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doesn't work
   kable_styling()

### 2020-07-10 manuscript/response_to_reviews.Rmd
add citations
worked on reviewer responses
Started on Olson review (reviewer 2). That one's fun!

#97

@@
 output: pdf_document
+
+bibliography: ["book.bib",  "packages.bib"]
+biblio-style: "apalike"
 ---
@@
 
-*What a nice review! We can easily add some language on the importance.*
+**Thank you for this wonderful review.**
+
+*What a nice review! We can easily add some language on the importance. We should also add citations to many of the studies mentioned.*
 
 first paragraph—this setup is very compelling, and the point is urgently true.
+
+**Thank you**
 
@@
 
+**Thank you for this interesting information.
+We've modified the sentence to read, "Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (Zak et al.
+2010, Olson et al.
+2014,2018).**
+
 67 Appealing to the vulnerability-diameter link might get blowback from some readers because there is as yet no satisfactory theory linking vulnerability to diameter in an entirely satisfactory way, so maybe just say “wider conduits are likely” or that they are “plausibly” or “possibly” more vulnerable, or that many “data are consistent” with such a link.
@@
 
+**We've added the word "plausibly". We agree that this linkage makes a lot of sense.**
+
 69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability.
@@
 There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.
+
+**Thank you.**
 
@@
 Figure S3: please specify that height values are estimated from reconstructed DBH.
+
+## Citations

### 2020-07-10 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses
Started on Olson review (reviewer 2). That one's fun!

#97

@@
 Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity,more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Indeed, tall trees require xylem of greater hydraulic efficiency in their basal portions, such that xylem conduit diameters are wider in taller trees within and across species [@olson_plant_2018; @liu_hydraulic_2019].
-Wider xylem conduits make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al.
+2010, Olson et al.
+2014,2018*).
+Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
 Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).

### 2020-07-11 manuscript/tables&figures.Rmd
update Table 2, add new Fig. 2
#100 , #101

@@
 ```
+
+*Sample sizes are prior to removal of outliers. Prior to analysis, we discarded # records with $Rt$ >2 and # records with $Rt_{ARIMA}$ >2.
+
 \clearpage
@@
 **Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
@@
 
-**Figure 2.
+**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
+
+**Figure 3.
 Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
@@
 ![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
@@
 ![**Figure 2.
+Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig.
+1.
+Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
+
+![**Figure 3.
 Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
@@
 Also shown is (d) 2018 tree heights by canopy position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2.png){width=500px}
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}

### 2020-07-12 manuscript/climate_sensitivity_SI.Rmd
update row packing in table S5
#99

@@
                      " "," ")) %>%
-  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 1, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("", 3, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 9, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(2, bold = T) %>%
   row_spec(3, bold = T) %>%
   row_spec(5, bold = T) %>%
+  row_spec(9, bold = T) %>%
   kable_styling()

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
update table 1 and text to remove DBH test
#99

@@
 
-- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
-- Individual-level drought resistance decreased with tree height, which was the dominant size-related variable affecting drought response.
+- Individual-level drought resistance decreased with tree height.
 Resistance was greater among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon desiccation.
@@
 
-Here, we examine how tree size, microenvironment characteristics, and species' traits collectively shape drought responses.
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses.
 We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
-First, we focus on the role of tree size and its interaction with microenvironment.
-We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of @bennett_larger_2015.
-We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
+First, we focus on the role of tree height and its interaction with microenvironment.
+We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
 Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$.
@@
 
-Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1).
-The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019].
+Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1).
+The greater susceptibility of taller trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height with modest additional affect of crown exposure [@liu_effect_1993; @stovall_tree_2019].
 We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.

### 2020-07-12 manuscript/tables&figures.Rmd
update table 1 and text to remove DBH test
#99

@@
   pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Taller trees have lower Rt.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Trees with more exposed crowns have lower Rt.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Species traits", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-07-12 manuscript/response_to_reviews.Rmd
work in new species variation

@@
 
-[*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+**We have *also* added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
 
@@
 
+**We have also added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
+
 - Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.
@@
 1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
+
+**To address the question about species, we have added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
 

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
work in new species variation

@@
 
-*Community-level drought responses*
+*Community-level and species' drought responses*
 
@@
 
+Responses varied across species and by drought (Fig. **2**). 
+Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **~0.6**) and highest in *Fagus grandifolia* (mean $Rt$ = **~0.9**).
+
 *Tree size, microenvironment, and drought resistance*
 
-Larger-diameter trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4).
-The same held true for $ln[H]$ in single-variable tests (Tables 1, 4).
+Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although $ln[H]$ was not significant during 1977 or 1999 individually (Tables 1, 4).
 When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).
@@
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
-**2d**).
+**3d**).
 When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$.
@@
 In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-**2**).
+**3**).
 Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-**2a**).
+**3a**).
 Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
-**2b**).
+**3b**).
 Air temperature did not vary across the vertical profile (Fig.
-**2c**).
+**3c**).
 
@@
 
-The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015].
+The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig.
+**2**) and associated traits at a single site [see also @elliott_forest_2015].
 Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.

### 2020-07-12 manuscript/response_to_reviews.Rmd
worked on reviewer responses (misc small changes)

@@
 
+**Fixed.**
+
 L189: fresh volume?
+
+**Yes, now specified.**
 
@@
 
+**We have made changes to the analysis that resulted in dropping ring porosity from the full models.
+While we agree that it is problematic that we only have two ring-porous species, we believe that it's worth including because this engages with, and rejects, the existing understanding that in this biome, ring-porous species are more drought tolerant.**
+
 L297: TLP was significant for 1966 according to Table 4.
@@
 
+**Unfortunately, no. All we could say is that the 1999 drought didn't result in a big loss of live biomass.**
+
 Table 3: are DBH units correct? I suspect they should be mm instead of cm.
+
+*These are incorrect. Ian, please convert to cm.*
 
@@
 
+**We now specify that these codes are given in Table 3.**
+
 Figure 2: where are the leaf hydraulic traits shown (cf. caption)?
+
+**It was an error that this was mentioned in the caption; now removed.**
 
@@
 
+**Bark thickness measurements were drawn from an existing data set, and we did not make additional measurements because the trait is unlikely to have meaningful influence on results.
+This study already goes above and beyond the existing norm for accounting for bark thickness in DBH reconstructions; typically there is no correction for changes in bark thickness as the tree ages.**
+
 Table S3: what does ‘rank’ refer to?
+
+**This has been clarified.**
 
@@
 
+*IAN, please do this*
+
 Figure S3: please specify that height values are estimated from reconstructed DBH.
 
+**Done.**
+
 ## Citations

### 2020-07-12 manuscript/climate_sensitivity_SI.Rmd
worked on reviewer responses (misc small changes)

@@
 \newpage
-![Figure S3: Height by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
+![Figure S3: Height (from reconstructed DBH) by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
 

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses (misc small changes)

@@
 $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves.
-wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to volume, which was estimated using Archimedes’ displacement.
+Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement.
 We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$).
@@
 
-The leaf hydraulic traits $PLA_{dry}$ and $\pi_{tlp}$ were linked to drought responses, whereas the other traits considered had insignficant and/or inconsistent correlations to $Rt$ (Tables 1,4,5).
-In the single-variable tests, $LMA$ and wood density were never significantly associated with $Rt$ (Table 4) and were excluded from the full models.
-In contrast, xylem porosity, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table 4).
-Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).
-$PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4).
-Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5).
-$\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+Wood density, $LMA$, and xylem porosity were all poor predictors of $Rt$ (Tables 1,4,5). 
+Wood density and $LMA$ were never significantly associated with $Rt$ in the single-variable tests (Table 4) and were therefore excluded from the full models. 
+Xylem porosity was also excluded from the full models, as it had no significant influence for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Table **4**).
+It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+**2**), further refuting the idea that xylem porosity is a useful predictor of $Rt$.
+
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Tables 1,4,5). 
+Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table 4), qualifying them as candidate variables for the full model.
+$PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). 
+$\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
 
@@
 
-Figure S3: Height by canopy position across the three focal droughts and in the year of
+Figure S3: Height (from reconstructed DBH) by canopy position across the three focal droughts and in the year of
 measurement (2018)

### 2020-07-12 manuscript/tables&figures.Rmd
worked on reviewer responses (misc small changes)

@@
 ![**Figure 3.
-Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
 In these plots, heights are slightly offset for visualization purposes.

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
xylem porosity-related changes

@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category.
-Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
+*It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category.* Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
 However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent.
@@
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
+Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
update Rt by species
#100

@@
 Responses varied across species and by drought (Fig. **2**). 
-Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **~0.6**) and highest in *Fagus grandifolia* (mean $Rt$ = **~0.9**).
+Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
 

### 2020-07-12 manuscript/climate_sensitivity_SI.Rmd
move tables 4 and 5 to appendix

@@
 
-\newpage
+\clearpage
 ```{r table S4, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 library(kableExtra)
+singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+kable(singlevar_table, booktabs = TRUE, caption = "Table S4. Individual tests of species traits as drivers of drought resistance ($Rt$).", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  kable_styling()
+```
+Variable abbreviations are as in Table 2. 
+$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
+
+*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
+
+**$\Delta$AICc > 2: variable is considered significant as an individual predictor (and meets $\Delta$AICc criterion for inclusion in full model)
+
+\newpage
+```{r table S5, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
 singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
-kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
+kable(singlevar_table, booktabs = TRUE, caption = "Table S5.
 Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
@@
 \newpage
-```{r table S5, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table S6, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
+                     " "," ")) %>%
+  pack_rows("", 1, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 9, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 11, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 21, 33, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  row_spec(1, bold = T) %>%
+  row_spec(9, bold = T) %>%
+  row_spec(11, bold = T) %>%
+  row_spec(21, bold = T) %>%
+  kable_styling()
+```
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold).
+
+\newpage
+```{r table S7, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
@@
 multivar_table <- read.csv("tables_figures/publication/tableS5_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
-kable(multivar_table, booktabs = TRUE, caption = "Table S5.
+kable(multivar_table, booktabs = TRUE, caption = "Table S7.
 Summary of top full models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
@@
                      " "," ")) %>%
-  pack_rows("", 1, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 3, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 9, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 8, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 20, 30, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(3, bold = T) %>%
   row_spec(5, bold = T) %>%
-  row_spec(9, bold = T) %>%
+  row_spec(8, bold = T) %>%
+  row_spec(20, bold = T) %>%
   kable_styling()
 ```
-Models are ranked by AICc. Shown are all models whose AICc value falls within 1.0 ($\Delta$AICc<1) of the best model (bold).
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold).
 

### 2020-07-12 manuscript/tables&figures.Rmd
move tables 4 and 5 to appendix

@@
 
-\clearpage
-```{r table 4, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
-kable(singlevar_table, booktabs = TRUE, caption = "Table 4. Individual tests of species traits as drivers of drought resistance ($Rt$).", format = 'latex', escape = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  kable_styling()
-```
-Variable abbreviations are as in Table 2. 
-$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
-
-*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
-
-**$\Delta$AICc > 2: variable is considered significant as an individual predictor (and meets $\Delta$AICc criterion for inclusion in full model)
-
-\clearpage
-```{r table 5, eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
-kable(multivar_table, booktabs = TRUE, caption = "Table 5. Summary of top full models for each drought instance. ", format = 'latex', escape = FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
-                     " "," ")) %>%
-  pack_rows("", 1, 1, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 2, 2, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 3, 3, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 4, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  row_spec(1, bold = T) %>%
-  row_spec(2, bold = T) %>%
-  row_spec(3, bold = T) %>%
-  row_spec(4, bold = T) %>%
-  kable_styling()
-```
-Models are ranked by AICc. Shown are all models whose AICc value falls within 1.0 ($\Delta$AICc<1) of the best model (bold).
-
 \newpage

### 2020-07-13 manuscript/tables&figures.Rmd
worked on table 1
add draft model visualization figure
#71

@@
 Parentheses indicate that the prediction was supported by one but not all of the top full models (Table **5**).
-Dash symbols indicate that the response was not significant, or not represented in any of the top full models.
+Dash symbols indicate that the response was not significant/ not represented in any of the top full models.
 Asterisks indicate that the $Rt_{ARIMA}$ model yielded a different result: 
@@
 
+**Figure 4. ...**
+
 \newpage
@@
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
+
+![**Figure 4.
+Visualization of top models for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
+**THIS IS NOT YET CORRECT--CANOPY POSITION NEEDS TO BE CHANGED TO CATEGORICAL VARIABLE, AND THE POSITIVE HEIGHT EFFECT FOR 1999 IS ALSO WRONG (PROBABLY ARTIFACT OF CP).
+FINAL FIG WILL MATCH MODELS IN TABLE S6.** ](tables_figures/publication/regr_visuals.png){width=500px}

### 2020-07-13 manuscript/climate_sensitivity_SI.Rmd
update SI tables (renamed files and changed in code)
#99

@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/table4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
 kable(singlevar_table, booktabs = TRUE, caption = "Table S4. Individual tests of species traits as drivers of drought resistance ($Rt$).", format = 'latex', escape = FALSE) %>%
@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+singlevar_table <- read.csv("tables_figures/publication/tableS5_singlevariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(singlevar_table, booktabs = TRUE, caption = "Table S5.
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS5_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.

### 2020-07-13 manuscript/response_to_reviews.Rmd
xylem porosity-related changes
worked on response to reviews

@@
 
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+**We created a new figure (*Fig. 4*) visualizing the top model results.**  ([*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)()
 
@@
 
-*How about this: *
-
-*(1) keep a version of table 5 in the main text, showing just the top model for Rt, and maybe also that of $Rt_{ARIMA}$.  *
-
-*(2) put full versions of table 2 (for Rt and $Rt_{ARIMA}$), including all models within the top dAICc=2, in the appendix*
-
-related to [*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
-
-related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)
+**In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig.
+4*) visualizing the top model results.
+We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**
 
@@
 
-*This table is definitely not just exploratory, and is the source for supporting/ rejecting predictions in Table 1.
-Thus, keeping it in the main text could be justified.
-That said, it presents mainly negative results, and I'd like to reduce the number of tables, so moving it to the appedix would be fine.*
-
-[*issue #95*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/95)
-
-related to [*issue #99*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)
+**We have moved this table to the SI.
+We modified the analysis approach such that separate models for individual variable tests (as previously show in this table) are now limited to the species traits.
+While this table is (and was) the only display item (besides Table 1) giving these results, the results are either mostly negative (wood density, SLA, xylem porosity) or preliminary tests qualifying variables for inclusion in full model ($PLA_{dry}$ and $\pi_{tlp}$).**
 
@@
 
-*Regarding Fig.
-2, my current inclination is to keep it While I think its true that such profiles are common, I'm not currently aware of a lot of studies showing such profiles.
-I just prepared a grant proposal related to this, and current intern Nidhi Vinod is working on a review on vertical profiles in temperature, so I'd hope I'd be relatively up to speed...
-I'll check into this more and come back to it.*
+[*issue #102*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/10)
 
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+**We did create two new figures showing the results of the tree-ring analysis (and moved two tables to the SI).**
 
@@
 
-*good point. hypothesis refers to temperate deciduous biome.*
+*good point. hypothesis refers to temperate deciduous biome, and is not supported here.*
 

### 2020-07-13 manuscript/response_to_reviews.Rmd
added response on keeping vertical profiles fig in main doc
closes #102

@@
 
-[*issue #102*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/10)
+**Our preference is to keep this figure in the main manuscript.
+We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing similar height profiles for a temperate forest.
+Such patterns are not shown some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*.
+If we're missing studies that show such vertical profiles, we'd appreciate the references.**
 

### 2020-07-13 manuscript/tables&figures.Rmd
added response on keeping vertical profiles fig in main doc
closes #102
update tables & figures

@@
 Asterisks indicate that the $Rt_{ARIMA}$ model yielded a different result: 
-*different significance in $Rt_{ARIMA}$ model
-**$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no signficant trend in $Rt$ model).
-
-**TILDAS (~) indicate that we don't currently have a (full) test of this prediction ([issue 99](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/99)**)
+*$Rt_{ARIMA}$ had a difference in significance/ inclusion in top full models.
+**$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no significant trend in $Rt$ model).
 
@@
 Visualization of top models for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
-**THIS IS NOT YET CORRECT--CANOPY POSITION NEEDS TO BE CHANGED TO CATEGORICAL VARIABLE, AND THE POSITIVE HEIGHT EFFECT FOR 1999 IS ALSO WRONG (PROBABLY ARTIFACT OF CP).
+**THIS IS NOT YET CORRECT (having trouble making gg-plot work.
+THE POSITIVE HEIGHT EFFECT FOR 1999 IS WRONG.
 FINAL FIG WILL MATCH MODELS IN TABLE S6.** ](tables_figures/publication/regr_visuals.png){width=500px}

### 2020-07-13 manuscript/tables&figures.Rmd
minor figure file updates

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Taller trees have lower Rt.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Trees with more exposed crowns have lower Rt.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Tree size and microenvironment", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Taller height makes trees less drought resistant.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Trees with more exposed crowns have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
   column_spec(2:5, width = "1.2cm") %>%
-  column_spec(6, width = "2.4cm") 
+  column_spec(6, width = "2cm") 
  # kableExtra::landscape()
@@
 
-Parentheses indicate that the prediction was supported by one but not all of the top full models (Table **5**).
-Dash symbols indicate that the response was not significant/ not represented in any of the top full models.
-Asterisks indicate that the $Rt_{ARIMA}$ model yielded a different result: 
+Parentheses indicate that the prediction was supported by one but not all of the top full models (Table S6).
+Dash symbols indicate that the response was not significant (Table S4)/ not represented in any of the top full models (Table S6).
+Asterisks indicate that the $Rt_{ARIMA}$ models (Table S5, S7) yielded a different result: 
 *$Rt_{ARIMA}$ had a difference in significance/ inclusion in top full models.

### 2020-07-13 manuscript/climate_sensitivity_draft.Rmd
update abstract

@@
 
-- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in a broadleaf deciduous forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 
-- Individual-level drought resistance decreased with tree height.
-Resistance was greater among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon desiccation.
-However, there was substantial variation in the best predictor variables across the three drought periods.
+- Individual-level drought resistance decreased with tree height and was not significantly linked to canopy position.
+The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topgraphic wetness index.
+Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.
 
-- We conclude that hydraulic traits and tree height influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
+- We conclude that tree height and hydraulic traits influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
 Thus, these factors can be useful for predicting future drought responses under climate change.

### 2020-07-13 manuscript/climate_sensitivity_SI.Rmd
minor SI updates

@@
 singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
-kable(singlevar_table, booktabs = TRUE, caption = "Table S4. Individual tests of species traits as drivers of drought resistance ($Rt$).", format = 'latex', escape = FALSE) %>%
+kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
+Individual tests of species traits as drivers of drought resistance, where $Rt$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
   add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
-  pack_rows("Species traits", 1, 6, latex_gap_space = "1em", colnum = 3, hline_before = FALSE) %>%
-  #mutate(dAICc = cell_spec(dAICc, "html", bold = ifelse(dAICc > 1.999, TRUE, FALSE))) %>%  #trying to put sig values in bold, but this doen't work
   kable_styling()

### 2020-07-13 manuscript/climate_sensitivity_draft.Rmd
update methods & results

@@
 We then determined the best full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=1 of the very top model (that with lowest AICc). 
-When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models.
-If the variable appeared in only some of the models, we considered this partial support/rejection.
+To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=2 of the very top model (that with lowest AICc). 
+When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1).
+If the variable appeared in some but not all of these models, and its sign we was consistent across models, we considered this partial support/rejection. 
+In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.
 
@@
 
-Responses varied across species and by drought (Fig. **2**). 
+Responses varied across species and by drought (Fig. 2). 
 Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
@@
 
-Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually, although $ln[H]$ was not significant during 1977 or 1999 individually (Tables 1, 4).
-When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).
+Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. 4;). 
+Specifically, $ln[H]$ appeared, with negative coefficient, in the top models and all models within $\Delta$AICc=2 of these (Tables S6-S7).
+For the 1977 and 1999 droughts, $ln[H]$ did not appear in the top models, but was included, with negative coefficient, in some of the models that were statistically indistinguishable ($\Delta$AICc<2) from the top model (Tables 1, S6-S7).
 
-Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig.
-**3d**).
-When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$.
-Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).
-When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees.
-Crown position was included in some of the full models (Table 5).
-In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest.
-In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.
+Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
+Crown position was never statistically significant (Table 1). 
+It did, however, come out in the top models for 1977 and 1999 (Fig. 4), and was included in some of the models within $\Delta$AICc=2 of the top model for 1966 and all droughts combined (Table S6).
+When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a top model, and was included among the models within $\Delta$AICc=2 of the top model only for 1977.
+Canopy position did not have a consistent influence, when included in models within $\Delta$AICc=2 of the top model, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
+4; Tables S6-S7).
 
@@
 
-$Rt$ was negatively correlated with $ln[TWI]$ (Tables 4-5), rejecting the idea that trees in moist microsites would be less affected by drought.
+$Rt$ had a signficantly negative response to $ln[TWI]$ in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in models within $\Delta$AICc=2 of the top model in 1999 and the $Rt_{ARIMA}$ model for 1966 (Fig.
+4, Tables 4-5).
+This negates the idea that trees in moist microsites would be less affected by drought. 
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected as the $ln[H] *ln[TWI]$ interaction was never significant (Table 4).
+This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never the significant, and had a positive sign in any models within $\Delta$AICc=2 of the top model in which it appeared (Tables 1, S6-S7).
+This term did appear, with positive coefficient, in the top $Rt_{ARIMA}$ model for all years combined (Table S7).
 
@@
 
-Wood density, $LMA$, and xylem porosity were all poor predictors of $Rt$ (Tables 1,4,5). 
-Wood density and $LMA$ were never significantly associated with $Rt$ in the single-variable tests (Table 4) and were therefore excluded from the full models. 
-Xylem porosity was also excluded from the full models, as it had no significant influence for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Table **4**).
+Wood density, $LMA$, and xylem porosity were all poor predictors of $Rt$ (Tables 1,S4-S5). 
+Wood density and $LMA$ were never significantly associated with $Rt$ in the single-variable tests and were therefore excluded from the full models. 
+Xylem porosity was also excluded from the full models, as it had no significant influence for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Table S4).
 It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-**2**), further refuting the idea that xylem porosity is a useful predictor of $Rt$.
+2), further refuting the idea that xylem porosity is a useful predictor of $Rt$.
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Tables 1,4,5). 
-Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table 4), qualifying them as candidate variables for the full model.
-$PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). 
-$\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5).
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. 4; Tables 1,S4-S7). 
+Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
+$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. 4; Tables S6-S7). 
+For 1977 and 1999, it was included, with negative coefficient, in some of the models within $\Delta$AICc=2 of the top model (Tables S6-S7).
+$\pi_{tlp}$ was included, with negative coefficient, in the top full model for all droughts combined and for the 1977 drought individually (Fig. 4; Table 5). 
+It was included in some of the models within $\Delta$AICc=2 of the top model in 1999 (Tables S6-S7).
 

### 2020-07-13 manuscript/climate_sensitivity_draft.Rmd
update discussion

@@
 
-Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1).
-The greater susceptibility of taller trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height with modest additional affect of crown exposure [@liu_effect_1993; @stovall_tree_2019].
+Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig.
+4).
+The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
 We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
@@
 Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts.
+diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
 This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
-The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms.
-However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+The directions of these responses were consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms.
+However, the strengths of each predictor varied across droughts (Fig.
+4; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
@@
 The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
-**1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Fig.
-**1b**).
+**1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs.
+**1b, 2**).
 These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
@@
 **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation: *11-13%* for all droughts combined, and *21-26%* for each individual drought (Table 5).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-26% for each individual drought (Table S6).
 
-Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5).
+Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Fig.
+4; Tables 1, S6-S7).
 This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
 However, the collinearity between the two variables (Fig.
-**2d**) makes it impossible to confidently partition causality.
+**3d**) makes it impossible to confidently partition causality.
 Taller trees are more likely to be in dominant canopy positions (Fig.
-**2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
-**2a-b**).
-Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves.
+**3d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
+**3a-b**).
+Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*).
 However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
-**2d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).
+**3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig.
+4; Tables 1, S6-S7).
 Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
@@
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
-The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019].
+The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
 It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016].
@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position.
+Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height, potentially with some contributions from canopy position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable.

### 2020-07-13 manuscript/climate_sensitivity_SI.Rmd
add draft S5

@@
 \newpage
-![Figure S4: Comparison of Rt and ARIMA results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+
+\newpage
+![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-13 manuscript/climate_sensitivity_SI.Rmd
update to new phytologist citation style (.csl)

@@
 biblio-style: apalike
+csl: new-phytologist.csl
 header-includes:

### 2020-07-13 manuscript/climate_sensitivity_draft.Rmd
update to new phytologist citation style (.csl)

@@
 biblio-style: "apalike"
-csl: apa.csl
+csl: new-phytologist.csl
 header-includes:
@@
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past **10** years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.

### 2020-07-13 manuscript/response_to_reviews.Rmd
update to new phytologist citation style (.csl)
started response letter to editor

@@
 biblio-style: "apalike"
+csl: new-phytologist.csl
 ---
@@
 **bold - response**
+
+~~~~
+
+Dear Editor:
+
+... Major changes included the following:
+
+* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years).
+Results were similar, and we now present that as a secondary model (mostly in SI).
+
+* We slightly modified the method of determining which trait variables go into the full model.
+Specifically, we added the criteria of consistent direction across droughts, and as a result dropped xylem porosity.
+This also makes sense because the two diffuse porous species (Liriodendron tulipifera and Fagus grandifolia) are at opposite ends of the spectrum in terms of drought resistance.
+(Side note: this contradicts the current understanding that diffuse porous species in the temperate deciduous biome are more drought sensitive, but not the bigger global picture, as R2 points out.).
+
+* We moved former tables 4 and 5 to the SI.
+
+* We added two figures: one showing Rt differences across species, and one visualizing the model results  (the latter is not yet right, as gg-plot is being difficult, but will eventually match  the model results.)
+
+The result is an easier to read paper with more robust support.
+
+\newpage
 

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
adjusted top/best model terminology
I've implemented your suggestion, @mcgregorian1 . This closes #103

@@
 
-We then determined the best full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=2 of the very top model (that with lowest AICc). 
+We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
+To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
 When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1).
@@
 Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. 4;). 
-Specifically, $ln[H]$ appeared, with negative coefficient, in the top models and all models within $\Delta$AICc=2 of these (Tables S6-S7).
-For the 1977 and 1999 droughts, $ln[H]$ did not appear in the top models, but was included, with negative coefficient, in some of the models that were statistically indistinguishable ($\Delta$AICc<2) from the top model (Tables 1, S6-S7).
+Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all models within $\Delta$AICc=2 of these (Tables S6-S7).
+For the 1977 and 1999 droughts, $ln[H]$ did not appear in the best models, but was included, with negative coefficient, among some of the top models, which were statistically indistinguishable ($\Delta$AICc<2) from the top model (Tables 1, S6-S7).
 
@@
 Crown position was never statistically significant (Table 1). 
-It did, however, come out in the top models for 1977 and 1999 (Fig. 4), and was included in some of the models within $\Delta$AICc=2 of the top model for 1966 and all droughts combined (Table S6).
-When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a top model, and was included among the models within $\Delta$AICc=2 of the top model only for 1977.
-Canopy position did not have a consistent influence, when included in models within $\Delta$AICc=2 of the top model, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
+It did, however, come out in the best models for 1977 and 1999 (Fig. 4), and was included in some of the top models for 1966 and all droughts combined (Table S6).
+When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a best model, and was included among the  top models only for 1977.
+Canopy position did not have a consistent influence, when included in top models, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
 4; Tables S6-S7).
@@
 
-$Rt$ had a signficantly negative response to $ln[TWI]$ in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in models within $\Delta$AICc=2 of the top model in 1999 and the $Rt_{ARIMA}$ model for 1966 (Fig.
+$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1999 and the $Rt_{ARIMA}$ model for 1966 (Fig.
 4, Tables 4-5).
@@
 Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never the significant, and had a positive sign in any models within $\Delta$AICc=2 of the top model in which it appeared (Tables 1, S6-S7).
-This term did appear, with positive coefficient, in the top $Rt_{ARIMA}$ model for all years combined (Table S7).
+This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never the significant, and had a positive sign in any top models  in which it appeared (Tables 1, S6-S7).
+This term did appear, with positive coefficient, in the best $Rt_{ARIMA}$ model for all years combined (Table S7).
 
@@
 It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-2), further refuting the idea that xylem porosity is a useful predictor of $Rt$.
+2), further refuting the idea that xylem porosity is a useful predictor of $Rt$ at this site.
 
@@
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. 4; Tables S6-S7). 
-For 1977 and 1999, it was included, with negative coefficient, in some of the models within $\Delta$AICc=2 of the top model (Tables S6-S7).
-$\pi_{tlp}$ was included, with negative coefficient, in the top full model for all droughts combined and for the 1977 drought individually (Fig. 4; Table 5). 
-It was included in some of the models within $\Delta$AICc=2 of the top model in 1999 (Tables S6-S7).
+For 1977 and 1999, it was included, with negative coefficient, in some of the top models (Tables S6-S7).
+$\pi_{tlp}$ was included, with negative coefficient, in the best model for all droughts combined and for the 1977 drought individually (Fig. 4; Table 5). 
+It was included in some of the top models for 1999 (Tables S6-S7).
 

### 2020-07-14 manuscript/tables&figures.Rmd
adjusted top/best model terminology
I've implemented your suggestion, @mcgregorian1 . This closes #103

@@
 
-Parentheses indicate that the prediction was supported by one but not all of the top full models (Table S6).
-Dash symbols indicate that the response was not significant (Table S4)/ not represented in any of the top full models (Table S6).
+Parentheses indicate that the prediction was supported by one but not all of the top  models (Table S6).
+Dash symbols indicate that the response was not significant (Table S4)/ not represented in any of the top models (Table S6).
 Asterisks indicate that the $Rt_{ARIMA}$ models (Table S5, S7) yielded a different result: 
-*$Rt_{ARIMA}$ had a difference in significance/ inclusion in top full models.
+*$Rt_{ARIMA}$ had a difference in significance/ inclusion in top  models.
 **$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no significant trend in $Rt$ model).
@@
 ![**Figure 4.
-Visualization of top models for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
+Visualization of best models for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
 **THIS IS NOT YET CORRECT (having trouble making gg-plot work.

### 2020-07-14 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd
started working on ring- vs diffuse-porous

@@
 
-*good point. hypothesis refers to temperate deciduous biome, and is not supported here.*
+**This is a great point. The hypothesis refers to temperate deciduous biome, which we now clarify in the sentence in question:**
+
+**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
+2007, Olson et al.
+2020*) and does not resolve differences among the many species within each category."**
+
+**We dug into the ring/diffuse porous distinction more, and ended up making a few substantive changes that relate to this theme: (1) We added a new figure showing $Rt$ across species (Fig.
+2).
+This illustrates that the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum.
+(2) We modified the criteria of selecting candidate variables for the full models to include the criterion that directions of responses must be consistent across droughts for inclusion in the full model.
+This resulted in exclusion of xylem porosity (Table S4)--a change that is also warrented by the fact that we only have two diffuse-porous species.
+(3) We added the following content to the discussion: **
+
+*insert text*
 
@@
 
+*from NObby: "Regarding the few individuals I would argue that intraspecific SD was low (less than 10% of the mean values). This justifies - in my opinion - only sampling three individuals. "*
+
 L188: ‘Wood’.

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
started working on ring- vs diffuse-porous

@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-*It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category.* Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
+Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
+2007, Olson et al.
+2020*) and does not resolve differences among the many species within each category.
+Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
 However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent.
@@
 
-We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig.
+We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig.
 **S1**).
@@
 
-Hydraulic traits were collected in August 2018 (Table 3).
+Hydraulic traits were collected in August 2018 (Tables 2-3).
 We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot.
@@
 
+Species traits... (Table 3, Fig. S#)
+
 Wood density, $LMA$, and xylem porosity were all poor predictors of $Rt$ (Tables 1,S4-S5). 
@@
 It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-2), further refuting the idea that xylem porosity is a useful predictor of $Rt$ at this site.
+2), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
 
@@
 **2**) and associated traits at a single site [see also @elliott_forest_2015].
-Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+2).
+While the low $Rt$ of *L.
+tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F.
+grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
+There are two potential explanations for this discrepancy.
+First, other traits can and do override the influence of xylem porosity on drought resistance. 
+Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S#**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
+A second reason is microclimate...
+
+*diffuse-porous vs ring-porous:*
+*-Elliot et al. do not include beech as one of their species*
+*-Friedrichs was study of 3 species in Europe. beech was the most drought-sensitive*
+*-Kannenberg did not include beech*
+
+Concerted measurement of tree-rings and leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
@@
 Thompson, and Victoria Meakem for original collection and processing of cores.
-Thanks also to Camila D.
-Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
+Thanks also to Camila D. Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection. 
+This manuscript was improved based on helpful reviews by Mark Olson and three anonymous reviewers.
 Funding for the establishment of the SCBI ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led Forest Global Earth Observatory (ForestGEO), the Smithsonian Institution, and the HSBC Climate Partnership.

### 2020-07-14 manuscript/climate_sensitivity_SI.Rmd
adjusted top/best model terminology
I've implemented your suggestion, @mcgregorian1 . This closes #103
figure/table numbering
#104

@@
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.
-Summary of top full models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
+Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
 ", format = 'latex', escape = FALSE) %>%
@@
 \newpage
-![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+![Figure S4: **PLACEHOLDER** ](tables_figures/publication/independent_variables_by_species.png)
 
 \newpage
-![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)
+![Figure S5: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+
+\newpage
+![Figure S6: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
moved subsection on identifying drought years up in methods
figure/table numbering
#104

@@
 Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2).
 First, we focus on the role of tree height and its interaction with microenvironment. 
@@
 
-Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
+Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W; Fig.
+**S1**) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
 SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park.
@@
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
-Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.).
+Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 3).
+
+*Identifying drought years*
+
+We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions. 
+Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
+Based on this, we identified three drought years - 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3).
+
+The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
+The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
+The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).
 
@@
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019].
-The resulting chronologies were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.
+The resulting chronologies (Fig. **1a**) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.
 
 For each cored tree, we combined tree-ring records and allometric equations of bark thickness to reconstruct DBH for the years 1950-2009. 
-Prior DBH was estimated using the following equation:
+Prior $DBH$ was estimated using the following equation:
 
@@
 
-Hydraulic traits were collected in August 2018 (Tables 2-3).
+Hydraulic traits were collected in August 2018 (Tables 2-3; Fig.
+**S4**).
 We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot.
@@
 
-*Identifying drought years*
-
-We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions. 
-Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
-PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified three drought years - 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3).
-
-The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
-The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
-The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).
-
 *Statistical Analysis*
@@
 Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+The two metrics were strongly correlated (Fig.
+**S5**).
 Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
@@
 Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S#**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
 A second reason is microclimate...

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
worked on discussion of traits

@@
 **2**) and associated traits at a single site [see also @elliott_forest_2015].
+Our study reinforced the findings of previous studies (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
 Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
@@
 Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to it's drought resistance (Fig. 4; see discussion below). 
-A second reason is microclimate...
-
-*diffuse-porous vs ring-porous:*
-*-Elliot et al. do not include beech as one of their species*
-*-Friedrichs was study of 3 species in Europe. beech was the most drought-sensitive*
-*-Kannenberg did not include beech*
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to it's drought resistance (Fig. **4**; see discussion below). 
+A second explanation of why *F.
+grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less dominant canopy positions compared to most other species (Fig.
+**S4**).
+The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig.
+**3**).
+Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
 

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
clarify single trait tests
#105

@@
 Models were run for all drought years combined and for each drought year individually. 
-The general statistical model for hypothesis testing was a generalized linear mixed model (GLMM), implemented  in the **XX** package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables including drought year (multi-drought model only), $ln[H]$*$TWI$, crown position, and 1-2 species traits (see below).
+The general statistical model for hypothesis testing was a generalized linear mixed model (GLMM), implemented  in the **XX** package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables as specified below.
+In the multi-year model only, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 To avoid over-fitting models with five species traits (Table 2) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
-These tests followed the model structure specified above.
-Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
+These tests followed the model structure specified above, including $ln[H]$, $TWI$, and crown position in the null model.
+Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to this null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
 We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
@@
 We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables and identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
+To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$, crown position, and species traits as specified above.
+We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
 When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1).

### 2020-07-14 scripts/graphs_for_manuscript.R
Update graphs_for_manuscript.R

@@ -159,7 +159,7 @@ summary(bleh[[4]]) #1999
 library(data.table)
 rt <- fread("manuscript/tables_figures/trees_all_sub.csv")
 traits_hydr <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_indvidual_level.csv?token=AJNRBELJEQJZPPXAC4GV4KS7CZAC2")
-traits_hydr <- traits[,.(sp,PLA_dry_percent, mean_TLP_Mpa)]
+traits_hydr <- traits_hydr[,.(sp,PLA_dry_percent, mean_TLP_Mpa)]
 
 rt <- rt[,year := as.character(year)]
 traits <- c("height.ln.m", "TWI.ln")

### 2020-07-14 manuscript/climate_sensitivity_draft.Rmd
minor changes
#105

@@
 
-- Individual-level drought resistance decreased with tree height and was not significantly linked to canopy position.
-The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topgraphic wetness index.
+- Individual-level drought resistance decreased with tree height, which was highly correlated with canopy position.
+The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
 Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.
@@
 
-We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions. 
+We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with anomalously dry peak growing season climatic conditions. 
 Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
@@
 
-Species traits... (Table 3, Fig. S#)
+Species traits... (Table 3, Fig. **S4**)
 

### 2020-07-16 manuscript/tables&figures.Rmd
update table 1 on height/CP (results not yet final)
#106

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  pack_rows("Tree size and microenvironment", 1, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Tree size and microenvironment", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("Taller height makes trees less drought resistant.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Trees with more exposed crowns have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Species traits", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Trees with more exposed crowns have lower drought resistance.", 2, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 4, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Species traits", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-07-16 manuscript/response_to_reviews.Rmd
switched back to LMM
#107 #105

@@
 
-**We switched to use of generalized linear mixed models. This did not change the outcome of any analyses.**
+**A GLMM would only avoid transformation of the dependent variable (Rt), but that was (and is) not transformed.
+We tried the analysis with generalized linear mixed models, and confirmed that this gave very similar results, but are sticking with LMM because...**
 
 [*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
+[*issue #107*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/107)
 

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
switched back to LMM
#107 #105

@@
 
-- Individual-level drought resistance decreased with tree height, which was highly correlated with canopy position.
+- Individual-level drought resistance decreased with tree height, *which was highly correlated with canopy position*.
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
@@
 Models were run for all drought years combined and for each drought year individually. 
-The general statistical model for hypothesis testing was a generalized linear mixed model (GLMM), implemented  in the **XX** package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables as specified below.
+The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables as specified below.
 In the multi-year model only, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))

### 2020-07-16 manuscript/climate_sensitivity_SI.Rmd
update SI with lmer results
#107, #105

CP (#106) removed from S4-S5 , still in S6-S7 (I'll return to this later)

@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
 kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/tableS5_singlevariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+singlevar_table <- read.csv("tables_figures/publication/tableS5_singlevariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(singlevar_table, booktabs = TRUE, caption = "Table S5.
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_reform.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_reform_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.

### 2020-07-16 manuscript/climate_sensitivity_SI.Rmd
manually create S6 and S7
#106
I manually deleted the rows with both height and CP, after checking that results would match those posted in issue #106.

@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer_CP+HmodelsDeleted.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
@@
                      " "," ")) %>%
-  pack_rows("", 1, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 9, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 11, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 21, 33, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 6, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 15, 21, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(9, bold = T) %>%
-  row_spec(11, bold = T) %>%
-  row_spec(21, bold = T) %>%
+  row_spec(5, bold = T) %>%
+  row_spec(6, bold = T) %>%
+  row_spec(15, bold = T) %>%
   kable_styling()
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio_CP+HmodelsDeleted.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.
@@
   pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 8, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 20, 30, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 8, 17, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 18, 28, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
@@
   row_spec(8, bold = T) %>%
-  row_spec(20, bold = T) %>%
+  row_spec(18, bold = T) %>%
   kable_styling()

### 2020-07-16 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd

--- 2020-07-14 manuscript/response_to_reviews.Rmd
@@
 
-**We switched to use of generalized linear mixed models. This did not change the outcome of any analyses.**
-
-[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
+**A GLMM would only avoid transformation of the dependent variable (Rt), but that was (and is) not transformed.
+We tried the analysis with generalized linear mixed models with a Gaussian distribution, and confirmed that this gave very similar results, so we are sticking with LMM.**
 

### 2020-07-16 manuscript/climate_sensitivity_SI.Rmd
consistent use of "crown position" instead of canopy position
#105

@@
 \newpage
-![Figure S3: Height (from reconstructed DBH) by canopy position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
+![Figure S3: Height (from reconstructed DBH) by crown position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
 

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
consistent use of "crown position" instead of canopy position
#105

@@
 
-- Individual-level drought resistance decreased with tree height, *which was highly correlated with canopy position*.
+- Individual-level drought resistance decreased with tree height, which was overall a stronger predictor than crown exposure.
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
@@
 
-*Key words*: annual growth; canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring
+*Key words*: annual growth; crown exposure; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring
 
@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because they tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
+Second, larger trees may have lower drought resistance because their crowns tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
@@
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019].
-Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
+Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, crown position, root water access, and species' traits.
 
@@
 An analysis of crown position relative to height (Fig.
-**2d**) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig.
+**2d**) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among crown positions (Fig.
 **S3**), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018.
@@
 When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a best model, and was included among the  top models only for 1977.
-Canopy position did not have a consistent influence, when included in top models, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
+Crown position did not have a consistent influence, when included in top models, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
 4; Tables S6-S7).
@@
 
-Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Fig.
+Our analysis indicates that tree height has a stronger influence on drought response than does crown position (Fig.
 4; Tables 1, S6-S7).
@@
 Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*).
-However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig.
+However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig.
 **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig.
@@
 
-Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts.
-However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis.
+Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts.
+However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis.
 The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height, potentially with some contributions from canopy position.
+Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height, potentially with some contributions from crown position.
 The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable.
@@
 
-Figure S3: Height (from reconstructed DBH) by canopy position across the three focal droughts and in the year of
+Figure S3: Height (from reconstructed DBH) by crown position across the three focal droughts and in the year of
 measurement (2018)

### 2020-07-16 manuscript/tables&figures.Rmd
consistent use of "crown position" instead of canopy position
#105

@@
 In these plots, heights are slightly offset for visualization purposes.
-Also shown is (**d**) 2018 tree heights by canopy position (see Table 2 for codes).
+Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
@@
 In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) 2018 tree heights by canopy position (see Table 2 for codes).
+Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
#105

@@
 This negates the idea that trees in moist microsites would be less affected by drought. 
-Nevertheless, we tested for a negative $ln[H] *ln[TWI]$ interaction, which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never the significant, and had a positive sign in any top models  in which it appeared (Tables 1, S6-S7).
+Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
+This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models  in which it appeared (Tables 1, S6-S7).
 This term did appear, with positive coefficient, in the best $Rt_{ARIMA}$ model for all years combined (Table S7).

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
worked on height & CP fixes
#106 #111

@@
 We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$, crown position, and species traits as specified above.
+To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$ or crown position and $TWI$, and species traits as specified above.
+Crown position and $ln[H]$ were never considered in the same model because of high correlation (0.73) between the two variables. 
 We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
@@
 
-Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. 4;). 
-Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all models within $\Delta$AICc=2 of these (Tables S6-S7).
-For the 1977 and 1999 droughts, $ln[H]$ did not appear in the best models, but was included, with negative coefficient, among some of the top models, which were statistically indistinguishable ($\Delta$AICc<2) from the top model (Tables 1, S6-S7).
+Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. **4**). 
+Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all top models (Tables S6-S7).
+For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7).
+For the 1999 drought, $ln[H]$ appeared in no top models for $Rt$ and some top models, with positive coefficient, for $Rt_{ARIMA}$.
 
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
-Crown position was never statistically significant (Table 1). 
-It did, however, come out in the best models for 1977 and 1999 (Fig. 4), and was included in some of the top models for 1966 and all droughts combined (Table S6).
-When $Rt_{ARIMA}$ was used as the predictor variable, crown position was never in a best model, and was included among the  top models only for 1977.
-Crown position did not have a consistent influence, when included in top models, across years: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in all other years suppressed trees had the lowest $Rt$ (Fig.
-4; Tables S6-S7).
+Crown position was not included in any of the top models for all droughts or for 1966 individually (Table 1). 
+It did, however, come out in the best models for 1977 and 1999 (Fig. **4**; Table S6).
+When $Rt_{ARIMA}$ was used as the response variable, crown position was never in a best model, and was included among the top models only for 1977 (Table S7).
+Crown position did not have a consistent influence: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in 1999 intermediate and suppressed trees had lower $Rt$ (Fig. 4; Tables S6-S7).
 
@@
 
-$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1999 and the $Rt_{ARIMA}$ model for 1966 (Fig.
-4, Tables 4-5).
+$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. 4, Table S6). 
+When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1966 and 1999 (Table S7).
 This negates the idea that trees in moist microsites would be less affected by drought. 
@@
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
-We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.
 The negative effect of height on $Rt$ held after accounting for species' traits.
+We found only a marginal effect of crown exposure with inconsistent direction across two individual droughts.
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-26% for each individual drought (Table S6).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S6).
 
-Our analysis indicates that tree height has a stronger influence on drought response than does crown position (Fig.
-4; Tables 1, S6-S7).
+Our analysis indicates that tree height has an overall stronger influence on drought response than does crown position (Table 1).
 This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
@@
 Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*).
-However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig.
+*Previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019], and ...* However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig.
 **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig.
-4; Tables 1, S6-S7).
+4; Tables 1).
 Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
 
-Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts.
+*Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts.
 However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis.
@@
 Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
+*
 

### 2020-07-16 manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 
-**maybe include these in fig with $Rt$ across species? [*issue #100*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)*
+**We now show differences in traits among species in Table S4.**
+
+*need to add ANOVA, prsent reults in paper*
 
@@
 
+**We agree that this could be a concern, but do not believe it is applicable here, for two reasons.
+First, this would be more likely to introduce a bias if the trees cored dead were killed by drought.
+We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
+However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
+Second, quoting our previous publication using these cores (Helcoski et al.
+2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with ≥ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+
+**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**
+
 L165-167: I think this needs to be explained more clearly.

### 2020-07-16 manuscript/response_to_reviews.Rmd
removed CP from full model, focus on just main results
issue #111 , #112

@@
 
-**We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations.
+*We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations.
 However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
 This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
-(Removed was ring porosity, which had opposite effects in different droughts.)**
+(Removed was ring porosity, which had opposite effects in different droughts.) *
 
@@
 3).**
+
+*see issue on CP!!!!*
 

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses
removed CP from full model, focus on just main results
issue #111 , #112

@@
 
-- Individual-level drought resistance decreased with tree height, which was overall a stronger predictor than crown exposure.
+- Individual-level drought resistance decreased with tree height, which was strongly correlated with crown exposure.
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
@@
 We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2).
-First, we focus on the role of tree height and its interaction with microenvironment. 
-We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites.
-Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
+First, we focus on the role of tree size and its interaction with microenvironment. 
+We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
+(2015).
+We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
+Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
@@
 ### Materials and Methods
-*Study site*
+*Study site and microclimate*
 
@@
 Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 3).
+
+*To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities].
+We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
+After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.* **(work on organization of this.)**
 
@@
 In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. 
+We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
@@
 
-To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities].
-We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
-After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.
-
 *Statistical Analysis*
@@
 
+*Tree height and microenvironment*
+
+In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
+**3**).
+Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
+**3a**).
+Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
+**3b**).
+Air temperature did not vary across the vertical profile (Fig.
+**3c**).
+
+Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
+Correlation between height and canopy position was 0.73.
+
 *Community-level and species' drought responses*
@@
 
-*Tree size, microenvironment, and drought resistance*
+*Tree height, microenvironment, and drought resistance*
 
@@
 For the 1999 drought, $ln[H]$ appeared in no top models for $Rt$ and some top models, with positive coefficient, for $Rt_{ARIMA}$.
-
-Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
-Crown position was not included in any of the top models for all droughts or for 1966 individually (Table 1). 
-It did, however, come out in the best models for 1977 and 1999 (Fig. **4**; Table S6).
-When $Rt_{ARIMA}$ was used as the response variable, crown position was never in a best model, and was included among the top models only for 1977 (Table S7).
-Crown position did not have a consistent influence: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in 1999 intermediate and suppressed trees had lower $Rt$ (Fig. 4; Tables S6-S7).
-
-In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-**3**).
-Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-**3a**).
-Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
-**3b**).
-Air temperature did not vary across the vertical profile (Fig.
-**3c**).
 

### 2020-07-16 manuscript/tables&figures.Rmd
removed CP from full model, focus on just main results
issue #111 , #112

@@
   pack_rows("Tree size and microenvironment", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Taller height makes trees less drought resistant.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Trees with more exposed crowns have lower drought resistance.", 2, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+ pack_rows("Trees with more exposed crowns have lower drought resistance.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Taller trees are less drought resistant.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 4, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(1, width = "12cm") %>%
-  column_spec(2:5, width = "1.2cm") %>%
-  column_spec(6, width = "2cm") 
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
+  #add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
+  #column_spec(1, width = "12cm") %>%
+  #column_spec(2:5, width = "1.2cm") %>%
+  #column_spec(6, width = "2cm") 
  # kableExtra::landscape()
@@
 Parentheses indicate that the prediction was supported by one but not all of the top  models (Table S6).
-Dash symbols indicate that the response was not significant (Table S4)/ not represented in any of the top models (Table S6).
-Asterisks indicate that the $Rt_{ARIMA}$ models (Table S5, S7) yielded a different result: 
-*$Rt_{ARIMA}$ had a difference in significance/ inclusion in top  models.
-**$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no significant trend in $Rt$ model).
+Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S6).
 
@@
 **Figure 1.
+Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
+
+**Figure 2.
 Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
@@
 
-**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
-
-**Figure 3.
-Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
+**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
 
@@
 ![**Figure 1.
+Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
+
+![**Figure 2.
 Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
@@
 
-![**Figure 2.
+![**Figure 3.
 Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig.
@@
 Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
-
-![**Figure 3.
-Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
 

### 2020-07-16 manuscript/climate_sensitivity_SI.Rmd

@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer_CP+HmodelsDeleted.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
-                     " "," ")) %>%
   pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("", 5, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 6, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 15, 21, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 6, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 11, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
@@
   row_spec(6, bold = T) %>%
-  row_spec(15, bold = T) %>%
+  row_spec(11, bold = T) %>%
   kable_styling()
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio_CP+HmodelsDeleted.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," "," "," ", " ", " ", "crown position" = 4, 
-                     " "," ")) %>%
   pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 8, 17, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("", 18, 28, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
@@
   row_spec(8, bold = T) %>%
-  row_spec(18, bold = T) %>%
+  row_spec(15, bold = T) %>%
   kable_styling()

### 2020-07-16 manuscript/tables&figures.Rmd
return fig 3 to orginal position--at least for now
#112
update hypothesis table
#115

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  pack_rows("Tree size and microenvironment", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
- pack_rows("Trees with more exposed crowns have lower drought resistance.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Taller trees are less drought resistant.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 4, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Species traits", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
+  pack_rows("Tree size and microenvironment", 1, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+ pack_rows("Across the forest vertical profile, taller trees are exposed to higher evaporative demand.", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("At least within the forest setting, taller trees are less drought resistant.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Species traits", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
   #add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
@@
 **Figure 1.
-Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
-
-**Figure 2.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
@@
 
-**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
+**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
+
+**Figure 3.
+Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
 
@@
 ![**Figure 1.
-Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
-
-![**Figure 2.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
@@
 
-![**Figure 3.
+\newpage
+
+![**Figure 2.
 Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig.
@@
 Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
+
+\newpage
+
+![**Figure 3.
+Height profiles in growing season climatic conditions and tree heights by crown position.** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
+
+\newpage
 

### 2020-07-16 manuscript/response_to_reviews.Rmd
worked on crown position
#112
removed Fig. S3

--- 2020-07-16 manuscript/response_to_reviews.Rmd
@@
 **In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig.
-4*) visualizing the top model results.
+4*) visualizing the results of the best model.
 We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**
 
-**We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations.
-However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
+**We have also dramatically decreased the emphasis on differences across drought years, as follows:* **
+
+**(1) Table 1 now shows just the results from the model with all drought years combined.**
+
+**(2) The results and discussion now focus on the results from the three drought years combined.**
+
+**(3) In order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
 This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
 (Removed was ring porosity, which had opposite effects in different droughts.)**
+
+*We do plot the three droughts separately in figures **#,#**, as we do fee that this makes the story richer.*
 
@@
 
+**We no longer attempt to use crown position to describe variation in drought resistance.
+Instead, we now use it to characterize *contemporary* differences in microenvironmental conditions faced by trees of different heights.**
+
 L217-226: this is problematic because drought resistance (Rt) as defined by Lloret et al.

### 2020-07-16 manuscript/response_to_reviews.tex
worked on crown position
#112
removed Fig. S3

--- 2020-07-16 manuscript/response_to_reviews.Rmd
@@
----
-title: "Response to reviews"
-output: pdf_document
-
-bibliography: ["book.bib",  "packages.bib"]
-biblio-style: "apalike"
-csl: new-phytologist.csl
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-*italics - temporary comments for points that remain to be addressed*
-
-**bold - response**
-
-~~~~
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Response to reviews},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+
+\title{Response to reviews}
+\author{}
+\date{\vspace{-2.5em}}
+
+\begin{document}
+\maketitle
+
+\emph{italics - temporary comments for points that remain to be
+addressed}
+
+\textbf{bold - response}
+
+\textasciitilde\textasciitilde\textasciitilde\textasciitilde{}
 
@@
 
-... Major changes included the following:
-
-* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years).
-Results were similar, and we now present that as a secondary model (mostly in SI).
-
-* We slightly modified the method of determining which trait variables go into the full model.
-Specifically, we added the criteria of consistent direction across droughts, and as a result dropped xylem porosity.
-This also makes sense because the two diffuse porous species (Liriodendron tulipifera and Fagus grandifolia) are at opposite ends of the spectrum in terms of drought resistance.
-(Side note: this contradicts the current understanding that diffuse porous species in the temperate deciduous biome are more drought sensitive, but not the bigger global picture, as R2 points out.).
-
-* We moved former tables 4 and 5 to the SI.
-
-* We added two figures: one showing Rt differences across species, and one visualizing the model results  (the latter is not yet right, as gg-plot is being difficult, but will eventually match  the model results.)
+\ldots{} Major changes included the following:
+
+\begin{itemize}
+\item
+  At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric
+  of drought resistance (using an ARIMA model to project what growth
+  would have been in drought year based on the trend over the past 10
+  years). Results were similar, and we now present that as a secondary
+  model (mostly in SI).
+\item
+  We slightly modified the method of determining which trait variables
+  go into the full model. Specifically, we added the criteria of
+  consistent direction across droughts, and as a result dropped xylem
+  porosity. This also makes sense because the two diffuse porous species
+  (Liriodendron tulipifera and Fagus grandifolia) are at opposite ends
+  of the spectrum in terms of drought resistance. (Side note: this
+  contradicts the current understanding that diffuse porous species in
+  the temperate deciduous biome are more drought sensitive, but not the
+  bigger global picture, as R2 points out.).
+\item
+  We moved former tables 4 and 5 to the SI.
+\item
+  We added two figures: one showing Rt differences across species, and
+  one visualizing the model results (the latter is not yet right, as
+  gg-plot is being difficult, but will eventually match the model
+  results.)
+\end{itemize}
 
@@
 
-## Referee: 1
+\hypertarget{referee-1}{%
+\subsection{Referee: 1}\label{referee-1}}
 
@@
 
-The study addresses the relationship between drought resistance (Rt) calculated from radial growth, and different traits and microenvironmental characteristics on 12 tree species from one forest in the Eastern US.
-Both the dataset and the study are meaningful, and the relationship found between Rt and height of biological importance.
-However, I think there are a number of issues that need to be worked out.
-
-It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought.
-This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript).
-In the paper, the authors use Rt similarly to most studies in the literature, which is fine.
-Yet, Rt (and particularly the other resilience indices based on radial growth) does not only depend on the individual response to drought events, but also on stand dynamics around target trees.
-For this reason, if stands around trees are not comparable or show changes after some disturbance (e.g.
-the drought events analysed) differently, this may bias comparisons of Resilience indices (although it is true that Rt should be less influenced than the other indices).
-The problem is that there is not such stand competition data for old droughts.
-This does not invalidate the study but I think the authors should discuss the issue, particularly for the analysis and comparisons of single droughts.
-I would suggest including also other Resilience indices, not only Rt, to enrich the discussion.
-
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
-**To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA).
-Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
-In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
-While this unquestionably leads to biased estimates for some trees, we would not expect it to bias our analysis; rather, it is one of the reasons that the amount of variation that can be explained by analyses such as these is relatively low.
-We now include the following content on ARIMA in the manuscript:**
-
-(*insert text*)
-
-**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
-While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) [DeSoto et al.
-(2020, Nature)](https://www.nature.com/articles/s41467-020-14300-5) used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
-In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
-In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
-These findings would argue that resistance is a more important variable for angiosperms.
-(3) As the reviewer notes, resilience indices would probably tend to be more sensitive to a tree's neighborhood and prior conditions.
-We now state the logic for focusing on $Rt$ in the methods section:**
-
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
-"**
-
-Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis.
-Then I missed a further discussion on the role of individual species (e.g.
-see my comment below on Rt>1).
-
-**A GLMM would only avoid transformation of the dependent variable (Rt), but that was (and is) not transformed.
-We tried the analysis with generalized linear mixed models with a Gaussian distribution, and confirmed that this gave very similar results, so we are sticking with LMM.**
-
-I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.
-
-**We created a new figure (*Fig. 4*) visualizing the top model results.**  ([*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)()
-
-**We have *also* added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
-
-Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models.
-LL ratio tests or deviance-anova tests can be used in the case of nested models with LMM or GLMM respectively.
-Along the text: dAIC is generally referred as $\Delta$AIC, please change accordingly.
-
-[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
-
-**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **
-
-Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included.
-And I suggest discussing only the model with all years together.
-This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation.
-This would help to ensure a better use of the Resilience indices.
-
-**In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig.
-4*) visualizing the top model results.
-We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**
-
-**We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations.
-However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
-This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
-(Removed was ring porosity, which had opposite effects in different droughts.)**
-
-The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis.
-If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.
-
-**We have moved this table to the SI.
-We modified the analysis approach such that separate models for individual variable tests (as previously show in this table) are now limited to the species traits.
-While this table is (and was) the only display item (besides Table 1) giving these results, the results are either mostly negative (wood density, SLA, xylem porosity) or preliminary tests qualifying variables for inclusion in full model ($PLA_{dry}$ and $\pi_{tlp}$).**
-
-Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al.
-included).
-Droughts must be selected based only on some climatic/drought index (PDSI, SPEI,…) and independently of growth.
-Otherwise the analysis is biased towards years with low growth, and different questions would arise with a difficult answer.
-If the authors select only those droughts where there is a growth decrease: is that decrease evident in all species and similar across species?
-What happens with more intense droughts that those selected (if any), why they do not force a decrease in growth if drought is a main forcing agent?
-I see that 1966 and 1999 are the driest (as in PDSI) in Figure 1, but there is at least one year around 1955 that it is as dry or drier than 1977.
-In line 204 it is stated that they are the driest.
-If this is the case, the analysis should be sound but it should be deleted what is stated in lines 202-203 where authors say that they selected years also based on low growth.
-In fact, that should be a result after having selected the drought years only based on dryness (as in lines 208-209).
-Line 211 then they search for pointer years… a little bit messy.
-Please, explain better, and it should follow: 1) selection of dry years without regarding growth data; 2) characterise growth those dry years; 3) If the authors want to compare with other pointer years, fine, but that should be a third independent step properly justified and explained.
-Following this rationale, then in line 221 it should be difficult that Rt>1, had we selected those years with a negative growth anomaly.
-Please, explain clearly and consistently.
-
-**We have made the suggested change: we now present only the PDSI criteria in the methods, and mention the fact that the selected years were also the ones with largest growth reductions during the study period in the results section.
-Using just the PDSI criterion does not change the years analyzed.**
+The study addresses the relationship between drought resistance (Rt)
+calculated from radial growth, and different traits and
+microenvironmental characteristics on 12 tree species from one forest in
+the Eastern US. Both the dataset and the study are meaningful, and the
+relationship found between Rt and height of biological importance.
+However, I think there are a number of issues that need to be worked
+out.
+
+It is important to better discuss the proper use of Rt, particularly
+when comparing single drought years, to assess tree resistance to
+drought. This metric has been much used lately as one of the so-called
+resilience indices (I wonder why other Rs indices where not included in
+the manuscript). In the paper, the authors use Rt similarly to most
+studies in the literature, which is fine. Yet, Rt (and particularly the
+other resilience indices based on radial growth) does not only depend on
+the individual response to drought events, but also on stand dynamics
+around target trees. For this reason, if stands around trees are not
+comparable or show changes after some disturbance (e.g.~the drought
+events analysed) differently, this may bias comparisons of Resilience
+indices (although it is true that Rt should be less influenced than the
+other indices). The problem is that there is not such stand competition
+data for old droughts. This does not invalidate the study but I think
+the authors should discuss the issue, particularly for the analysis and
+comparisons of single droughts. I would suggest including also other
+Resilience indices, not only Rt, to enrich the discussion.
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92}{\emph{issue
+\#92}}
+
+\textbf{To account for potential trends in individual tree growth prior
+to the drought, which could be influenced by stand dynamics, we tried an
+intervention time series analysis (ARIMA). Details on the comparison
+between \(Rt\) and \(Rt_{ARIMA}\) are presented in the responses to
+reviewer 4. In short, the ARIMA method yielded broadly similar results,
+but tended to produce less reasonable estimates when the difference
+between the two metrics was large, so we decided to retain \(Rt\) as the
+focal metric. While this unquestionably leads to biased estimates for
+some trees, we would not expect it to bias our analysis; rather, it is
+one of the reasons that the amount of variation that can be explained by
+analyses such as these is relatively low. We now include the following
+content on ARIMA in the manuscript:}
+
+(\emph{insert text})
+
+\textbf{Regarding the resilience indices, we considered analyzing these
+and decided it would be best to stick with just resistance for three
+reasons: (1) Most importantly, we would expect resilience to be governed
+by fundamentally different mechanisms than resistance. While it would be
+interesting to examine resilience, this would be beyond the scope of the
+current paper. (2)
+\href{https://www.nature.com/articles/s41467-020-14300-5}{DeSoto et
+al.~(2020, Nature)} used an extensive dataset of trees sampled after
+severe droughts and compared the responses to previous, less-severe
+droughts between trees that survived and trees that were killed by the
+severe drought. In angiosperms, trees that died from a severe drought
+had lower resistance to previous droughts, compared to trees that
+survived the severe drought. In gymnosperms, trees that died during
+severe droughts had lower recovery following previous droughts, compared
+to trees that survived the severe drought. These findings would argue
+that resistance is a more important variable for angiosperms. (3) As the
+reviewer notes, resilience indices would probably tend to be more
+sensitive to a tree's neighborhood and prior conditions. We now state
+the logic for focusing on \(Rt\) in the methods section:}
+
+\textbf{``We focus exclusively on drought resistance (\(Rt\) or
+\(Rt_{ARIMA}\)), and not on the resilience metrics described in Lloret
+\emph{et al.} (2011), because (1) we would expect resilience to be
+controlled by a different set of mechanisms, and (2) the findings of
+\emph{\href{https://www.nature.com/articles/s41467-020-14300-5}{DeSoto
+et al.~(2020)}} suggest that \(Rt\) is a more important drought response
+metric for angiosperms.''}
+
+Statistical analysis: the analysis would be more precise if the authors
+used Generalized linear mixed models to avoid data transformation and
+take into account (with random effects, as they do for the linear mixed
+model, LMM, presented now) the influence of species in the analysis.
+Then I missed a further discussion on the role of individual species
+(e.g.~see my comment below on Rt\textgreater1).
+
+\textbf{A GLMM would only avoid transformation of the dependent variable
+(Rt), but that was (and is) not transformed. We tried the analysis with
+generalized linear mixed models with a Gaussian distribution, and
+confirmed that this gave very similar results, so we are sticking with
+LMM.}
+
+I suggest to show some relationship, e.g.~between Rt and traits in
+figures. The study only presents two figures and there is plenty of room
+to present results more clearly with extra figures.
+
+\textbf{We created a new figure (\emph{Fig. 4}) visualizing the top
+model results.}
+(\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71}{\emph{issue
+\#71}}()
+
+\textbf{We have \emph{also} added a new figure (\emph{Fig. 2}) showing
+how \(Rt\) varied across species in each of the three droughts.}
+
+Only assessing models with AIC is not enough to prove sound
+relationships in multidimensional and complex data-models. LL ratio
+tests or deviance-anova tests can be used in the case of nested models
+with LMM or GLMM respectively. Along the text: dAIC is generally
+referred as \(\Delta\)AIC, please change accordingly.
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94}{\emph{issue
+\#94}}
+
+\textbf{We have gone through the manuscript / tables and figures and
+updated where it said ``dAICc'' to be \(\Delta\)AICc. }
+
+Following the two previous comments, I suggest to leave in Table 5 just
+the best model (rather that a multimodel inference as is presented now)
+to discuss more clearly the covariates included. And I suggest
+discussing only the model with all years together. This would be more
+robust and avoid ramble differences between years that mostly look
+spurious and with a difficult physiological explanation. This would help
+to ensure a better use of the Resilience indices.
+
+\textbf{In the spirit of this comment (simplifying presentation), we
+have created a new figure (\emph{Fig. 4}) visualizing the results of the
+best model. We moved Table 5 to the Appendix (now Table S6, plus a
+parallel Table S7 for \(Rt_{ARIMA}\)).}
+
+\textbf{We have also dramatically decreased the emphasis on differences
+across drought years, as follows:* }
+
+\textbf{(1) Table 1 now shows just the results from the model with all
+drought years combined.}
+
+\textbf{(2) The results and discussion now focus on the results from the
+three drought years combined.}
+
+\textbf{(3) In order to make models more consistent across years and
+reduce spurious correlations, we now include the criteria that for a
+trait to be included as a candidate variable in the full model, the sign
+of coefficients must be consistent across all droughts (in addition to
+meeting the AIC criteria, as before). This resulted in only two
+candidate variables for the main model (\(PLA_{dry}\) and
+\(\pi_{tlp}\)). (Removed was ring porosity, which had opposite effects
+in different droughts.)}
+
+\emph{We do plot the three droughts separately in figures
+\textbf{\#,\#}, as we do fee that this makes the story richer.}
+
+The authors might like to consider including Table 4 as a supplement,
+because as it stands it looks like an exploratory analysis. If the
+`single-variable approach' is not just exploratory, state it clearly and
+why it is needed besides the full models with several parameters.
+
+\textbf{We have moved this table to the SI. We modified the analysis
+approach such that separate models for individual variable tests (as
+previously show in this table) are now limited to the species traits.
+While this table is (and was) the only display item (besides Table 1)
+giving these results, the results are either mostly negative (wood
+density, SLA, xylem porosity) or preliminary tests qualifying variables
+for inclusion in full model (\(PLA_{dry}\) and \(\pi_{tlp}\)).}
+
+Importantly, the drought selection would be biased if years were
+selected as from line 202, which is a common mistake in similar studies
+(the original of Lloret et al.~included). Droughts must be selected
+based only on some climatic/drought index (PDSI, SPEI,\ldots) and
+independently of growth. Otherwise the analysis is biased towards years
+with low growth, and different questions would arise with a difficult
+answer. If the authors select only those droughts where there is a
+growth decrease: is that decrease evident in all species and similar
+across species? What happens with more intense droughts that those
+selected (if any), why they do not force a decrease in growth if drought
+is a main forcing agent? I see that 1966 and 1999 are the driest (as in
+PDSI) in Figure 1, but there is at least one year around 1955 that it is
+as dry or drier than 1977. In line 204 it is stated that they are the
+driest. If this is the case, the analysis should be sound but it should
+be deleted what is stated in lines 202-203 where authors say that they
+selected years also based on low growth. In fact, that should be a
+result after having selected the drought years only based on dryness (as
+in lines 208-209). Line 211 then they search for pointer years\ldots{} a
+little bit messy. Please, explain better, and it should follow: 1)
+selection of dry years without regarding growth data; 2) characterise
+growth those dry years; 3) If the authors want to compare with other
+pointer years, fine, but that should be a third independent step
+properly justified and explained. Following this rationale, then in line
+221 it should be difficult that Rt\textgreater1, had we selected those
+years with a negative growth anomaly. Please, explain clearly and
+consistently.
+
+\textbf{We have made the suggested change: we now present only the PDSI
+criteria in the methods, and mention the fact that the selected years
+were also the ones with largest growth reductions during the study
+period in the results section. Using just the PDSI criterion does not
+change the years analyzed.}
 
@@
 
-- Line 113: correct ‘positively’.
-
-**done**
-
-- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.
-
-**done**
-
-- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
-should be rather straightforward and the authors would increase accuracy across species in height estimation.
-
-**We have specified how many species used the interspecific allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be possible to go collect more data for these species.**
-
-- Lines 166-167: please explain further the hypothesis and why.
-
-**We have reworded this sentence to read, "While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative."**
-
-- Line 222: what about climate in years before 1991?
-
-**We no longer mention 1991, as it does not meet the PDSI criteria.**
-
-- Line 231: define AICc.
-
-**This is done (with added reference)**
-
-- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).
-
-**This section has been re-written.**
-
-- Line 241-242: enough to say once that models were selected based on AIC (line 237-238) and then one criterion, why dAIC<1 here and dAIC<2 before?
-The classic assumption is 2 (e.g.
-Burnham & Anderson 2004).
-See also my other comment above.
-
-**We have modified this text to read as follows:**
-
-**"Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
-We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models."**
-
-- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.
-
-**As stated above, we have removed the $Rt$ criterion from the drought identification process, and now present this information for the first time in this section, adding the following statement: "Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$."**
-
-- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?
-
-**It's not unprecedented to see some trees exhibit increased growth during drought, and not surprising.
-Stand dynamics and other factors result in a lot of scatter in the relationship between tree growth and climate, so while the distribution of $Rt$ in Fig.
-1b was shifted left (normal year should be normally distributed around 1), there still are some trees with increased growth.
-Our analysis helps to work out which trees these were (often shorter trees).**
-
-**We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), potentially due to decreased leaf area of competitors during the drought, and consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al.
-2015).
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**
-
-**We have also added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
-
-- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.
-
-**It is true that crown position estimates become more accurate in later dates, but height should be equally reliable across years, having been estimated based on reconstructed DBH.
-This was not previously clearly specified in the methods.
-To clarify, we have added the following statement: "We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$."**
-
-**We agree that the assessment of crown position is somewhat problematic, both because of its increasing inaccuracy going back through time, and because of collinearity with $H$.
-However, given that it is both biologically interesting and comes out in our top models, we have chosen to include it.
-We note that this makes the test of our hypothesis conservative.
-We are transparent about this issue, to which we devote a paragraph in the discussion (par.
-3).**
-
-- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest.
-They are indeed interesting to characterise the forest, but they do not add to individual tree variability.
-You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.
-
-**Our preference is to keep this figure in the main manuscript.
-We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing similar height profiles for a temperate forest.
-Such patterns are not shown some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*.
-If we're missing studies that show such vertical profiles, we'd appreciate the references.**
-
-**We did create two new figures showing the results of the tree-ring analysis (and moved two tables to the SI).**
-
-- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?
-
-*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
-
-**To address the question about species, we have added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
-
-- Line 331-333: not so sure about this.
-It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
-
-*Tree-ring data inherently have a lot of variation. Neil could probably write a better response here, with refs.*
-
-- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).
-
-**Please see response above.**
-
-- Line 348: non-drought.
-
-**Fixed.**
-
-- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
-
-**We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here).
-We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019]."**
-
-- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.
-
-**See above comment.
-We have reworded this sentence to read, "In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought."**
-
-Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
-
-**This is now clarified in the table footnote.**
-
-Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
-
-**In this study, we have one core per tree, now specified in the methods. We have changed the table header to "n trees".**
-
-Table 3: the authors could show some test to see if the traits presented are different among species.
-
-**We now show differences in traits among species in Table S4.**
-
-*need to add ANOVA, prsent reults in paper*
+\begin{itemize}
+\tightlist
+\item
+  Line 113: correct `positively'.
+\end{itemize}
+
+\textbf{done}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 134: cambial growth also includes that of phloem; so better refer
+  to `xylem growth'.
+\end{itemize}
+
+\textbf{done}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 161: how many species were estimated for height like this?
+  Wouldn't it be possible for authors to go to the field and perform
+  some height estimations? This should be rather straightforward and the
+  authors would increase accuracy across species in height estimation.
+\end{itemize}
+
+\textbf{We have specified how many species used the interspecific
+allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be
+possible to go collect more data for these species.}
+
+\begin{itemize}
+\tightlist
+\item
+  Lines 166-167: please explain further the hypothesis and why.
+\end{itemize}
+
+\textbf{We have reworded this sentence to read, ``While some tree crowns
+undoubtedly changed position over the past several decades, in this case
+the bias would be unlikely to result in false acceptance of the
+prediction that dominant trees have the lowest \(Rt\) (\emph{i.e.}, type
+I error unlikely, type II error possible), making our hypothesis test
+conservative.''}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 222: what about climate in years before 1991?
+\end{itemize}
+
+\textbf{We no longer mention 1991, as it does not meet the PDSI
+criteria.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 231: define AICc.
+\end{itemize}
+
+\textbf{This is done (with added reference)}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 234-247: it is not clear what the authors did, please explain
+  clearer this paragraph and the next and why you did two different (yet
+  similar) analyses (single-variable and full model).
+\end{itemize}
+
+\textbf{This section has been re-written.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 241-242: enough to say once that models were selected based on
+  AIC (line 237-238) and then one criterion, why dAIC\textless1 here and
+  dAIC\textless2 before? The classic assumption is 2 (e.g.~Burnham \&
+  Anderson 2004). See also my other comment above.
+\end{itemize}
+
+\textbf{We have modified this text to read as follows:}
+
+\textbf{``Trait variables were considered appropriate for inclusion in
+the main model if they had a consistent direction of response across all
+droughts and if their addition to a corresponding null model lacking the
+trait improved fit (at \(\Delta\)AICc \(\ge\) 1.0) in at least one
+drought year (Table 4). We note that the \(\Delta\)AICc \(\ge\) 1.0
+criterion is not a test of significance, but of whether the variable has
+enough influence to be considered as a \emph{candidate} variable in full
+models.''}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 255-256: but this was already an assumption in M-M (line
+  202-203). Please, see my comment below and be consistent along the
+  text.
+\end{itemize}
+
+\textbf{As stated above, we have removed the \(Rt\) criterion from the
+drought identification process, and now present this information for the
+first time in this section, adding the following statement: ``Across the
+entire study period (1950-2009), the focal drought years were the three
+years with the largest fraction of trees exhibiting \(Rt \le 0.7\).''}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 258: so this means that they are no sensitive to the drought?
+  Which trees, any species in particular? Is this a consequence of stand
+  dynamics shadowing the impact of drought events?
+\end{itemize}
+
+\textbf{It's not unprecedented to see some trees exhibit increased
+growth during drought, and not surprising. Stand dynamics and other
+factors result in a lot of scatter in the relationship between tree
+growth and climate, so while the distribution of \(Rt\) in Fig. 1b was
+shifted left (normal year should be normally distributed around 1),
+there still are some trees with increased growth. Our analysis helps to
+work out which trees these were (often shorter trees).}
+
+\textbf{We address this question in the discussion, where we have added
+a statement about how this is not unprecedented: ``Across all three
+droughts, the majority of trees experienced reduced growth, but a
+substantial portion had increased growth (Fig. 1b), potentially due to
+decreased leaf area of competitors during the drought, and consistent
+with prior observations that smaller trees can exhibit increased growth
+rates during drought (Bennett et al.~2015). It is likely because of the
+moderate impact of these droughts, along with other factors influencing
+tree growth (e.g., stand dynamics), that our best models characterize
+only a modest amount of variation''}
+
+\textbf{We have also added a new figure (\emph{Fig. 2}) showing how
+\(Rt\) varied across species in each of the three droughts.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 266-267: why only in 1966? I think this shows that this variable
+  is not properly assessed. The older the date, the more accurate the
+  estimation of crown position and height were.
+\end{itemize}
+
+\textbf{It is true that crown position estimates become more accurate in
+later dates, but height should be equally reliable across years, having
+been estimated based on reconstructed DBH. This was not previously
+clearly specified in the methods. To clarify, we have added the
+following statement: ``We then used these allometries to estimate \(H\)
+for each drought year, \(Y\), based on reconstructed \(DBH_Y\).''}
+
+\textbf{We agree that the assessment of crown position is somewhat
+problematic, both because of its increasing inaccuracy going back
+through time, and because of collinearity with \(H\). However, given
+that it is both biologically interesting and comes out in our top
+models, we have chosen to include it. We note that this makes the test
+of our hypothesis conservative. We are transparent about this issue, to
+which we devote a paragraph in the discussion (par. 3).}
+
+\emph{see issue on CP!!!!}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 275-280: those profiles were assumed a priori in the introduction
+  and are common to any forest. They are indeed interesting to
+  characterise the forest, but they do not add to individual tree
+  variability. You could think to send Figure 2 to an appendix, in
+  consequence, and use figures to show more specific results related to
+  your individual tree analyses.
+\end{itemize}
+
+\textbf{Our preference is to keep this figure in the main manuscript. We
+agree that similar patterns are common to any forest, and are not trying
+to claim this as any kind of novel result, but we're not aware of
+studies showing similar height profiles for a temperate forest. Such
+patterns are not shown some relevant textbooks: Gordan Bonan's 2016
+\emph{Ecological Climatology} and Norman \& Campbell's
+\emph{Environmental Biophysics}. If we're missing studies that show such
+vertical profiles, we'd appreciate the references.}
+
+\textbf{We did create two new figures showing the results of the
+tree-ring analysis (and moved two tables to the SI).}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 330-331: any evidence to support this? It looks rather
+  speculative. As suggested above, does any species particularly show
+  positive Rt or positive Rt are randomly distributed in your trees?
+\end{itemize}
+
+\emph{Current text, with phrase of concern in bold: ``Across all three
+droughts, the majority of trees experienced reduced growth, but a
+substantial portion had increased growth (Fig. 1b), potentially due to
+decreased leaf area of competitors during the drought (\textbf{REF--if
+we can find one}), and consistent with prior observations that smaller
+trees can exhibit increased growth rates during drought (Bennett
+\emph{et al.}, 2015).'' }
+
+\textbf{To address the question about species, we have added a new
+figure (\emph{Fig. 2}) showing how \(Rt\) varied across species in each
+of the three droughts.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 331-333: not so sure about this. It can also mean that there is
+  no causal relationship between Rt and covariates, mostly because
+  radial growth Rt should not be used as an integral proxy of tree
+  resistance to drought, such as it is often used now in the literature.
+\end{itemize}
+
+\emph{Tree-ring data inherently have a lot of variation. Neil could
+probably write a better response here, with refs.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 334-339: indeed, assessment of these two variables is not
+  conclusive, as suggested above (just data from 2018, from the past
+  estimated and correlated).
+\end{itemize}
+
+\textbf{Please see response above.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 348: non-drought.
+\end{itemize}
+
+\textbf{Fixed.}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 349: do authors really think that lower radial growth as from Rt
+  on a specific year means that the tree is more vulnerable? I think
+  this is over conclusive.
+\end{itemize}
+
+\textbf{We did not intend the meaning here as vulnerability \emph{to
+mortality} (assuming that's the objection here). We have reworded this
+sentence to read, ``Belowground, taller trees would tend to have larger
+root systems, but the potentially greater access to water did not
+override the disadvantage conferred by height--and, in fact, greater
+moisture access in non-drought years (here, higher TWI) appears to make
+trees more sensitive to drought (Zuleta \emph{et al.}, 2017; Stovall
+\emph{et al.}, 2019).''}
+
+\begin{itemize}
+\tightlist
+\item
+  Line 400: I do not agree we can directly link Rt with tree
+  vulnerability to drought; we might be currently overinterpreting these
+  resilience indices extrapolating to physiological conclusions that do
+  not necessary need to hold.
+\end{itemize}
+
+\textbf{See above comment. We have reworded this sentence to read, ``In
+the meantime, the results of this study advance our knowledge of the
+factors conferring drought resistance in a mature forest, opening the
+door for more accurate forecasting of forest responses to future
+drought.''}
+
+Table 1: do `-` signs in the table correspond to `no' (non-significant)?
+Please explain.
+
+\textbf{This is now clarified in the table footnote.}
+
+Table 3: please, add also the number of trees, it is more informative
+than the number of cores to check sample size in your study.
+
+\textbf{In this study, we have one core per tree, now specified in the
+methods. We have changed the table header to ``n trees''.}
+
+Table 3: the authors could show some test to see if the traits presented
+are different among species.
+
+\textbf{We now show differences in traits among species in Table S4.}
+
+\emph{need to add ANOVA, prsent reults in paper}
 
@@
 
-*IM I added the correct Latex notation but it's not rendering it when I knit the file... (KAT: I just tried and wasn't able to get it working.*
-
-Table S3: please, explain what ‘rank’ means.
-
-**Done.**
-
-Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time.
-The question is whether there was any stand dynamics involved in those height series that had to be included in the models.
-
-**We recognize the importance of stand dynamics, and have added specific mention to the discussion: "It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation,...
-"**
-
-## Referee: 2
+\emph{IM I added the correct Latex notation but it's not rendering it
+when I knit the file\ldots{} (KAT: I just tried and wasn't able to get
+it working.}
+
+Table S3: please, explain what `rank' means.
+
+\textbf{Done.}
+
+Figure S3: I understand that authors only have height data from 2018 and
+height for the other dates looks trivial: indeed there will be an
+increase in height with an increase in DBH in time. The question is
+whether there was any stand dynamics involved in those height series
+that had to be included in the models.
+
+\textbf{We recognize the importance of stand dynamics, and have added
+specific mention to the discussion: ``It is likely because of the
+moderate impact of these droughts, along with other factors influencing
+tree growth (e.g., stand dynamics), that our best models characterize
+only a modest amount of variation,\ldots{}''}
+
+\hypertarget{referee-2}{%
+\subsection{Referee: 2}\label{referee-2}}
 
@@
 
-This manuscript examines a truly urgent and important issue.
-The general issue is the clear vulnerability of forests to ongoing climate change induced drought worldwide.
-The issue is particularly vexing because, while it seems clear that plant hydraulics is involved in vulnerability to drought, the exact mechanisms connecting hydraulics to vulnerability are not clear, and just what indices biologists should use to gauge vulnerability is as a result unclear.
-Within the broad pattern of global forest decline with climate change induced drought is the particularly mystifying and horrific preferential loss of the largest individuals all over the world.
-In addition to their esthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.
-
-The issue of the preferential death of large trees is not only important because of its effects on forest function and culture.
-It is important because it seems to require novel theory.
-Under current expectations, the largest individuals, with their greater root reach, etc., should be more rather than less resistant.
-That they might not be demands a novel explanatory framework.
-Yet the very existence of the tendency is debated, with some studies reporting greater mortality among small rather than large individuals (Camarero et al., 2015; Colangelo et al., 2017).
-These studies, however, were not designed to take into account the microsite variation that presumably should lead to marked differences in survivorship even within the same height class with drought.
-One approach, which we have used (Soriano et al., in press; Olson et al., 2018) is to conduct experiments under controlled conditions to detect the effect of height, and our results are consistent with the greater vulnerability of taller individuals.
-However, under natural conditions, a study that takes this variation into account is needed to test the notion that taller individuals are preferentially vulnerable is needed, and this study provides exactly such a design.
-
-I found this paper not only to address a very important point but also to be exceptionally well structured and written.
-My main suggestion is that it could use a bit more emphasis on its importance, and I would be interested in hearing a bit more about what the authors feel their results imply regarding the cause of the preferential vulnerability of large individuals.
-I give line-by-line suggestions (and sometimes just reactions):
-
-**Thank you for this wonderful review.**
-
-*What a nice review! We can easily add some language on the importance. We should also add citations to many of the studies mentioned.*
-
-first paragraph—this setup is very compelling, and the point is urgently true.
-
-**Thank you**
-
-66- “in their basal portions” seems fine for this level of intro, but note that in angiosperms, taller trees also have wider vessels at the stem tip (and thus throughout the conductive system) (Zach et al., 2010; Olson et al., 2014, 2018).
-This seems plausibly the result of selection keeping conduit carbon costs down with height growth.
-As a tree grows taller, if terminal twig vessel diameters become wider, then given a constant rate of tip-to-base vessel widening, vessel diameters will be wider along the entire stem.
-Vessels that are wider throughout would have lower per-conduit resistance than those that, all else being equal, are narrower.
-So, conceivably as a tree gets taller, if vessels widen at the twig tip, then lower per-conduit resistance could mean that the same leaf area is supplied by fewer vessels.
-This could potentially contribute to keeping per-leaf area carbon costs constant with height growth.
-This prediction seems confirmed by the experiment of (Echeverría et al., 2019), who found that as Moringa trees grew taller, there were wider vessels at the stem tip and fewer vessels per unit leaf area, but that conductance per unit leaf area should remain the same.
-Terminal twig tracheids don’t seem to widen in gymnosperms, e.g.
-(Williams et al., 2019).
-(Sack et al., 2012) have shown very clearly that petiole base vessel diameter scales predictably with leaf size, as would be expected; it is not clear how leaf size variation is involved in terminal twig vessel diameter scaling with height, though in wide samplings leaf size doesn’t seem to scale with height (Jensen & Zwieniecki, 2013).
-
-**Thank you for this interesting information.
-We've modified the sentence to read, "Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (Zak et al.
-2010, Olson et al.
-2014,2018).**
-
-67 Appealing to the vulnerability-diameter link might get blowback from some readers because there is as yet no satisfactory theory linking vulnerability to diameter in an entirely satisfactory way, so maybe just say “wider conduits are likely” or that they are “plausibly” or “possibly” more vulnerable, or that many “data are consistent” with such a link.
-My view is that there very well might not be a link, but in the absence of a link between vulnerability and conduit diameter, it is very hard to explain a very large number of well-established observations.
-These include the following: why growth rings always proceed from wide to narrow conduits as drought ensues; why maximum community conduit diameter is narrower in drylands than in wet places; why lianas have wider vessel diameter variances (i.e.
-not only wider but also numerous narrower vessels) than self-supporting plants (Rosell & Olson, 2014); why plants virtually always produce vessels narrower than the maximum possible; when internal helical sculpture in conduits varies in a growth ring it is always associated with narrow latewood vessels; when simple and scalariform plates vary across growth rings sclariform plates are always found in the narrow latewood vessels; and many other observations.
-If there is no link between drought induced embolism vulnerability and conduit diameter, and there might not be, all of these observations, which are currently simultaneously and parsimoniously explained by the link, will require alternative explanations.
-As a result, while there is as yet no well accepted mechanistic link between vulnerability and diameter, it seems to me irresponsible to reject the link without providing a satisfactory and similarly parsimonious and general replacement explanation for all of these phenomena.
-So, I think the authors are justified in cautiously invoking the link here.
-
-**We've added the word "plausibly". We agree that this linkage makes a lot of sense.**
-
-69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability.
-The generalization that on a global level large trees seem disproportionately vulnerable is often greeted with the skeptical observation that in specific cases a small tree died next to a large one that survived.
-However, the prediction is that *all else being equal* large individuals should be more vulnerable than smaller *conspecifics*.
-There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.
-
-**Thank you.**
-
-73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.
-
-73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.
-
-77 In the concluding sentence, “greater growth reductions” came as a surprise because I thought that the focus was greater vulnerability/ mortality.
-Also, there are convincing theoretical and empirical reasons to think that large trees don’t slow down e.g.
-(Sillett et al., 2010; Selaya & Anten, 2010; Michaletz et al., 2014).
-
-96 this all seems compelling.
-To make the point stronger, either justify that the leaf traits are causally more important, or that, while they make up part of a unified hydraulic system all of whose components are important, leaf traits are simply easier to measure and this justifies their study.
-The latter position seems most plausible to me.
-
-102 great questions, though within-species comparisons are likely to be most informative here, and your mixed model design is in accord with this; just being cautious because sometimes it’s seductive to regard selection acting between rather than within species
-
-**Thank you.**
-
-105 this sounds great, but was a little unclear.
-When you say “first, we test…” I thought that you would examine a relationship between microsite and some tree characteristic, e.g.
-that the largest trees are in moist microsites.
-But the next sentence didn’t refer to a tree-microsite correlation but vulnerability-size.
-It would be good to clarify the prose here because these are very important passages.
-The third sentence did seem to be about tree-microsite correlations.
-
-112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from.
-Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007).
-They are virtually absent in drylands, especially warm tropical drylands.
-Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).
-
-**This is a great point. The hypothesis refers to temperate deciduous biome, which we now clarify in the sentence in question:**
-
-**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
-2007, Olson et al.
-2020*) and does not resolve differences among the many species within each category."**
-
-**We dug into the ring/diffuse porous distinction more, and ended up making a few substantive changes that relate to this theme: (1) We added a new figure showing $Rt$ across species (Fig.
-2).
-This illustrates that the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum.
-(2) We modified the criteria of selecting candidate variables for the full models to include the criterion that directions of responses must be consistent across droughts for inclusion in the full model.
-This resulted in exclusion of xylem porosity (Table S4)--a change that is also warrented by the fact that we only have two diffuse-porous species.
-(3) We added the following content to the discussion: **
-
-*insert text*
-
-113 These correlations seem reasonable, e.g. see (Méndez-Alonzo et al., 2012)
-
-124 and Table 3.
-This seems like a good palette of species because they are very different in their hydraulic construction.
-In addition to libriform fibers, oaks have very abundant vasicentric tracheids, especially in latewood.
-This means that the oaks will likely maintain some conductance even under highly negative tensions.
-This is not by virtue of their ring porosity (note that most oaks globally aren’t ring porous!), but because of the high proportion of vasicentric tracheids in their latewood.
-Liriodendron (diffuse porous, scalariform perforation plates) and Carya (likely ring porous at your locality; simple perforation plates) both have libriform fibers and lack vasicentric tracheids.
-See the data in (Olson et al., in press).
-There is relatively low family-level diversity represented here; our phylogenetic analyses (Olson et al., in press) suggest that there is very little phylogenetic signal in the conduit diameter-stem length relationship, so this could help justify if necessary not including phylogeny in the analyses.
-
-*this is new terminology to me. Look into Olson et al. in press. Would these wood traits be worth including as covariates?*
-
-127 “tree size”= height?
-
-138 you might discuss the adequacy of the height range studied.
-If it is true that conduit diameter predicts vulnerability, then the greatest differences in vulnerability should be over the height range over which conduit diameter changes most rapidly.
-The relationship between conduit diameter and plant height turns out to follow a power-law like distribution, see for example the SI in (Olson et al., 2018).
-This means that differences in vulnerability should be very marked across smaller size classes (say below 7 m or so and very marked below 3 m) and be less and less visible across larger plants.
-So, studies of changes in conduit diameter/ vulnerability with height would ideally include plants smaller than those included here, so maybe comment on how the height range might affect the ability to detect the patterns of interest.
-It strikes me that it is appropriate because it is the height range over which the putative preferential susceptibility of large trees to drought is being observed.
-
-173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.
-
-**Thank you.**
-
-209 It couldn’t hurt to define basal area increment
-
-227 the nesting seems appropriate and sufficient for taking species/ “phylogeny” into account
+This manuscript examines a truly urgent and important issue. The general
+issue is the clear vulnerability of forests to ongoing climate change
+induced drought worldwide. The issue is particularly vexing because,
+while it seems clear that plant hydraulics is involved in vulnerability
+to drought, the exact mechanisms connecting hydraulics to vulnerability
+are not clear, and just what indices biologists should use to gauge
+vulnerability is as a result unclear. Within the broad pattern of global
+forest decline with climate change induced drought is the particularly
+mystifying and horrific preferential loss of the largest individuals all
+over the world. In addition to their esthetic appeal, large individuals
+make disproportionate contributions to forest function, making the issue
+examined in this ms extremely timely and important.
+
+The issue of the preferential death of large trees is not only important
+because of its effects on forest function and culture. It is important
+because it seems to require novel theory. Under current expectations,
+the largest individuals, with their greater root reach, etc., should be
+more rather than less resistant. That they might not be demands a novel
+explanatory framework. Yet the very existence of the tendency is
+debated, with some studies reporting greater mortality among small
+rather than large individuals (Camarero et al., 2015; Colangelo et al.,
+2017). These studies, however, were not designed to take into account
+the microsite variation that presumably should lead to marked
+differences in survivorship even within the same height class with
+drought. One approach, which we have used (Soriano et al., in press;
+Olson et al., 2018) is to conduct experiments under controlled
+conditions to detect the effect of height, and our results are
+consistent with the greater vulnerability of taller individuals.
+However, under natural conditions, a study that takes this variation
+into account is needed to test the notion that taller individuals are
+preferentially vulnerable is needed, and this study provides exactly
+such a design.
+
+I found this paper not only to address a very important point but also
+to be exceptionally well structured and written. My main suggestion is
+that it could use a bit more emphasis on its importance, and I would be
+interested in hearing a bit more about what the authors feel their
+results imply regarding the cause of the preferential vulnerability of
+large individuals. I give line-by-line suggestions (and sometimes just
+reactions):
+
+\textbf{Thank you for this wonderful review.}
+
+\emph{What a nice review! We can easily add some language on the
+importance. We should also add citations to many of the studies
+mentioned.}
+
+first paragraph---this setup is very compelling, and the point is
+urgently true.
+
+\textbf{Thank you}
+
+66- ``in their basal portions'' seems fine for this level of intro, but
+note that in angiosperms, taller trees also have wider vessels at the
+stem tip (and thus throughout the conductive system) (Zach et al., 2010;
+Olson et al., 2014, 2018). This seems plausibly the result of selection
+keeping conduit carbon costs down with height growth. As a tree grows
+taller, if terminal twig vessel diameters become wider, then given a
+constant rate of tip-to-base vessel widening, vessel diameters will be
+wider along the entire stem. Vessels that are wider throughout would
+have lower per-conduit resistance than those that, all else being equal,
+are narrower. So, conceivably as a tree gets taller, if vessels widen at
+the twig tip, then lower per-conduit resistance could mean that the same
+leaf area is supplied by fewer vessels. This could potentially
+contribute to keeping per-leaf area carbon costs constant with height
+growth. This prediction seems confirmed by the experiment of (Echeverría
+et al., 2019), who found that as Moringa trees grew taller, there were
+wider vessels at the stem tip and fewer vessels per unit leaf area, but
+that conductance per unit leaf area should remain the same. Terminal
+twig tracheids don't seem to widen in gymnosperms, e.g.~(Williams et
+al., 2019). (Sack et al., 2012) have shown very clearly that petiole
+base vessel diameter scales predictably with leaf size, as would be
+expected; it is not clear how leaf size variation is involved in
+terminal twig vessel diameter scaling with height, though in wide
+samplings leaf size doesn't seem to scale with height (Jensen \&
+Zwieniecki, 2013).
+
+\textbf{Thank you for this interesting information. We've modified the
+sentence to read, "Indeed, tall trees require xylem of greater hydraulic
+efficiency, such that xylem conduit diameters are wider in the basal
+portions of taller trees, both within and across species (Olson \emph{et
+al.}, 2018; Liu \emph{et al.}, 2019), and throughout the conductive
+systems of angiosperms (Zak et al.~2010, Olson et al.~2014,2018).}
+
+67 Appealing to the vulnerability-diameter link might get blowback from
+some readers because there is as yet no satisfactory theory linking
+vulnerability to diameter in an entirely satisfactory way, so maybe just
+say ``wider conduits are likely'' or that they are ``plausibly'' or
+``possibly'' more vulnerable, or that many ``data are consistent'' with
+such a link. My view is that there very well might not be a link, but in
+the absence of a link between vulnerability and conduit diameter, it is
+very hard to explain a very large number of well-established
+observations. These include the following: why growth rings always
+proceed from wide to narrow conduits as drought ensues; why maximum
+community conduit diameter is narrower in drylands than in wet places;
+why lianas have wider vessel diameter variances (i.e.~not only wider but
+also numerous narrower vessels) than self-supporting plants (Rosell \&
+Olson, 2014); why plants virtually always produce vessels narrower than
+the maximum possible; when internal helical sculpture in conduits varies
+in a growth ring it is always associated with narrow latewood vessels;
+when simple and scalariform plates vary across growth rings sclariform
+plates are always found in the narrow latewood vessels; and many other
+observations. If there is no link between drought induced embolism
+vulnerability and conduit diameter, and there might not be, all of these
+observations, which are currently simultaneously and parsimoniously
+explained by the link, will require alternative explanations. As a
+result, while there is as yet no well accepted mechanistic link between
+vulnerability and diameter, it seems to me irresponsible to reject the
+link without providing a satisfactory and similarly parsimonious and
+general replacement explanation for all of these phenomena. So, I think
+the authors are justified in cautiously invoking the link here.
+
+\textbf{We've added the word ``plausibly''. We agree that this linkage
+makes a lot of sense.}
+
+69 this is a very important point---microsite is crucial both for the
+ultimate height an individual grows to as well as its vulnerability. The
+generalization that on a global level large trees seem
+disproportionately vulnerable is often greeted with the skeptical
+observation that in specific cases a small tree died next to a large one
+that survived. However, the prediction is that \emph{all else being
+equal} large individuals should be more vulnerable than smaller
+\emph{conspecifics}. There is so much variation in microsite and other
+factors that the greater vulnerability of larger individuals should
+emerge clearly only in carefully controlled experiments where all else
+is indeed (virtually) equal, or with high sample sizes where variation
+in microsite etc should random out and the signal of generally higher
+vulnerability should come through.
+
+\textbf{Thank you.}
+
+73 The ``greater root reach'' point is almost universally appealed
+to---cites for these counterarguments would be useful.
+
+73 Also, larger trees would necessarily seem to have greater stem
+storage capacity and thus, if capacitance really is important, more
+water volume available for buffering transient drought events, so this
+is another reason to expect larger individuals to be more rather than
+less resistant.
+
+77 In the concluding sentence, ``greater growth reductions'' came as a
+surprise because I thought that the focus was greater vulnerability/
+mortality. Also, there are convincing theoretical and empirical reasons
+to think that large trees don't slow down e.g.~(Sillett et al., 2010;
+Selaya \& Anten, 2010; Michaletz et al., 2014).
+
+96 this all seems compelling. To make the point stronger, either justify
+that the leaf traits are causally more important, or that, while they
+make up part of a unified hydraulic system all of whose components are
+important, leaf traits are simply easier to measure and this justifies
+their study. The latter position seems most plausible to me.
+
+102 great questions, though within-species comparisons are likely to be
+most informative here, and your mixed model design is in accord with
+this; just being cautious because sometimes it's seductive to regard
+selection acting between rather than within species
+
+\textbf{Thank you.}
+
+105 this sounds great, but was a little unclear. When you say ``first,
+we test\ldots{}'' I thought that you would examine a relationship
+between microsite and some tree characteristic, e.g.~that the largest
+trees are in moist microsites. But the next sentence didn't refer to a
+tree-microsite correlation but vulnerability-size. It would be good to
+clarify the prose here because these are very important passages. The
+third sentence did seem to be about tree-microsite correlations.
+
+112 wasn't clear where the ``ring porous more drought resistant''
+hypothesis came from. Global data show that ring porous species are
+almost entirely restricted to the north temperate zone, especially
+deciduous broadleaf forests (Wheeler et al., 2007). They are virtually
+absent in drylands, especially warm tropical drylands. Species in
+drylands are almost overwhelmingly diffuse porous; the high wood density
+species in chaparrals and in some desert plants can be semi ring porous
+(Olson et al., in press).
+
+\textbf{This is a great point. The hypothesis refers to temperate
+deciduous biome, which we now clarify in the sentence in question:}
+
+\textbf{``Studies in temperate broadleaf forests have observed that
+ring-porous species showing higher drought tolerance than diffuse-porous
+species (Kannenberg \emph{et al.}, 2019; Friedrichs \emph{et al.}, 2009;
+Elliott \emph{et al.}, 2015), but this distinction would not hold in the
+global context (\emph{Wheeler et al.~2007, Olson et al.~2020}) and does
+not resolve differences among the many species within each category.''}
+
+\textbf{We dug into the ring/diffuse porous distinction more, and ended
+up making a few substantive changes that relate to this theme: (1) We
+added a new figure showing \(Rt\) across species (Fig. 2). This
+illustrates that the two diffuse-porous species, \emph{Liriodendron
+tulipifera} and \emph{Fagus grandifolia}, were at opposite ends of the
+\(Rt\) spectrum. (2) We modified the criteria of selecting candidate
+variables for the full models to include the criterion that directions
+of responses must be consistent across droughts for inclusion in the
+full model. This resulted in exclusion of xylem porosity (Table S4)--a
+change that is also warrented by the fact that we only have two
+diffuse-porous species. (3) We added the following content to the
+discussion: }
+
+\emph{insert text}
+
+113 These correlations seem reasonable, e.g.~see (Méndez-Alonzo et al.,
+2012)
+
+124 and Table 3. This seems like a good palette of species because they
+are very different in their hydraulic construction. In addition to
+libriform fibers, oaks have very abundant vasicentric tracheids,
+especially in latewood. This means that the oaks will likely maintain
+some conductance even under highly negative tensions. This is not by
+virtue of their ring porosity (note that most oaks globally aren't ring
+porous!), but because of the high proportion of vasicentric tracheids in
+their latewood. Liriodendron (diffuse porous, scalariform perforation
+plates) and Carya (likely ring porous at your locality; simple
+perforation plates) both have libriform fibers and lack vasicentric
+tracheids. See the data in (Olson et al., in press). There is relatively
+low family-level diversity represented here; our phylogenetic analyses
+(Olson et al., in press) suggest that there is very little phylogenetic
+signal in the conduit diameter-stem length relationship, so this could
+help justify if necessary not including phylogeny in the analyses.
+
+\emph{this is new terminology to me. Look into Olson et al.~in press.
+Would these wood traits be worth including as covariates?}
+
+127 ``tree size''= height?
+
+138 you might discuss the adequacy of the height range studied. If it is
+true that conduit diameter predicts vulnerability, then the greatest
+differences in vulnerability should be over the height range over which
+conduit diameter changes most rapidly. The relationship between conduit
+diameter and plant height turns out to follow a power-law like
+distribution, see for example the SI in (Olson et al., 2018). This means
+that differences in vulnerability should be very marked across smaller
+size classes (say below 7 m or so and very marked below 3 m) and be less
+and less visible across larger plants. So, studies of changes in conduit
+diameter/ vulnerability with height would ideally include plants smaller
+than those included here, so maybe comment on how the height range might
+affect the ability to detect the patterns of interest. It strikes me
+that it is appropriate because it is the height range over which the
+putative preferential susceptibility of large trees to drought is being
+observed.
+
+173 TWI seems like an appropriate index for characterizing microsite,
+with the expectation that individuals in moister areas should be more
+buffered from drought.
+
+\textbf{Thank you.}
+
+209 It couldn't hurt to define basal area increment
+
+227 the nesting seems appropriate and sufficient for taking species/
+``phylogeny'' into account
 
@@
 
-**Thank you.**
-
-275—some people think that the increase in terminal twig vessels with height growth keeps pace with the changes in VPD from understory to canopy
+\textbf{Thank you.}
+
+275---some people think that the increase in terminal twig vessels with
+height growth keeps pace with the changes in VPD from understory to
+canopy
 
@@
 
-**Thank you.**
-
-301 this intro paragraph to the discussion ranks very well the results from most significant to least
-
-**Thank you.**
-
-327 I agree that this is the point that most deserves discussion in paragraph 2
-
-334 You might go into detail regarding what you think this says about the causal hypothesis behind the global decline of large individuals.
-To me it helps reject notions that somehow larger trees are more exposed to wind throw or other damaging agents.
-It is consistent with the hypothesis that larger individuals have wider conduits, and (assuming that wider conduits are more vulnerable to drought induced embolism) thus larger individuals are (all else being equal…) more vulnerable to embolism than smaller conspecifics.
-
-336 beautifully said “impossible for trees to e ciently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”
-
-**Thank you.**
-
-349 fascinating; it seems plausible that current climate changed induced large-individual mortality involves trees growing to conduit diameters, and thus heights, permitted under previously more even and moister climates, and now finding themselves in more extreme climates that make their heights and conduits excessively wide given current conditions.
-
-350 if you wanted to expand your discussion of causal mechanisms or the implications of your findings, I think you could jettison this paragraph.
-I admire your thoroughness in offering caveats, but I agree that this does not seem like a serious problem for your study.
-
-382 an effort that seems in harmony with these concerns is (Christoffersen et al., 2016)
+\textbf{Thank you.}
+
+301 this intro paragraph to the discussion ranks very well the results
+from most significant to least
+
+\textbf{Thank you.}
+
+327 I agree that this is the point that most deserves discussion in
+paragraph 2
+
+334 You might go into detail regarding what you think this says about
+the causal hypothesis behind the global decline of large individuals. To
+me it helps reject notions that somehow larger trees are more exposed to
+wind throw or other damaging agents. It is consistent with the
+hypothesis that larger individuals have wider conduits, and (assuming
+that wider conduits are more vulnerable to drought induced embolism)
+thus larger individuals are (all else being equal\ldots) more vulnerable
+to embolism than smaller conspecifics.
+
+336 beautifully said ``impossible for trees to e ciently transport water
+to great heights and simultaneously maintain strong resistance and
+resilience to drought-induced embolism''
+
+\textbf{Thank you.}
+
+349 fascinating; it seems plausible that current climate changed induced
+large-individual mortality involves trees growing to conduit diameters,
+and thus heights, permitted under previously more even and moister
+climates, and now finding themselves in more extreme climates that make
+their heights and conduits excessively wide given current conditions.
+
+350 if you wanted to expand your discussion of causal mechanisms or the
+implications of your findings, I think you could jettison this
+paragraph. I admire your thoroughness in offering caveats, but I agree
+that this does not seem like a serious problem for your study.
+
+382 an effort that seems in harmony with these concerns is
+(Christoffersen et al., 2016)
 
@@
 
-**Thank you.**
-
-In conclusion, this ms addresses a very important issue with a well-designed study and appropriate analyses. It is well written and interesting and I hope my comments are useful.
-
-**Thank you.**
-
-Mark Olson molson@ib.unam.mx
+\textbf{Thank you.}
+
+In conclusion, this ms addresses a very important issue with a
+well-designed study and appropriate analyses. It is well written and
+interesting and I hope my comments are useful.
+
+\textbf{Thank you.}
+
+Mark Olson
+\href{mailto:molson@ib.unam.mx}{\nolinkurl{molson@ib.unam.mx}}
 
 Camarero JJ, Gazol A, Sangüesa-Barreda G, Oliva J, Vicente-Serrano SM.
-2015.
-To die or not to die: early warnings of tree dieback in response to a severe drought (D Gibson, Ed.).
-Journal of Ecology 103: 44–57.
-
-Christoffersen BO, Gloor M, Fauset S, Fyllas NM, Galbraith DR, Baker TR, Kruijt B, Rowland L, Fisher RA, Binks OJ, et al.
-2016.
-Linking hydraulic traits to tropical forest function in a size-structured and trait-driven model (TFS v.1-Hydro).
-Geoscientific Model Development 9: 4227–4255.
-
-Colangelo M, Camarero JJ, Borghetti M, Gazol A, Gentilesca T, Ripullone F.
-2017.
-Size Matters a Lot: Drought-Affected Italian Oaks Are Smaller and Show Lower Growth Prior to Tree Death.
-Frontiers in Plant Science 8.
-
-Echeverría A, Anfodillo T, Soriano D, Rosell JA, Olson ME.
-2019.
-Constant theoretical conductance, changes in vessel diameter and number with height growth in Moringa oleifera.
-Journal of Experimental Botany.
-
-Jensen KH, Zwieniecki MA. 2013. Physical Limits to Leaf Size in Tall Trees. Physical Review Letters 110.
-
-Méndez-Alonzo R, Paz H, Zuluaga RC, Rosell JA, Olson ME. 2012. Coordinated evolution of leaf and stem economics in tropical dry forest trees. Ecology 93: 2397–2406.
-
-Michaletz ST, Cheng D, Kerkhoff AJ, Enquist BJ. 2014. Convergence of terrestrial plant production across global climate gradients. Nature 512: 39–43.
-
-Olson ME, Anfodillo T, Rosell JA, Petit G, Crivellaro A, Isnard S, León-Gómez C, Alvarado-Cárdenas LO, Castorena M.
-2014.
-Universal hydraulics of the flowering plants: vessel diameter scales with stem length across angiosperm lineages, habits and climates.
-Ecology Letters 17: 988–997.
-
-Olson ME, Rosell JA, Martínez-Pérez C, León-Gómez C, Fajardo A, Isnard S, Cervantes-Alcayde M-A, Echeverría A, Figueroa-Abúndiz, Segovia-Rivas A, et al.
-in press.
-Xylem vessel diameter-shoot length scaling: ecological significance of porosity types and other traits.
-Ecological Monographs.
-
-Olson ME, Soriano D, Rosell JA, Anfodillo T, Donoghue MJ, Edwards EJ, León-Gómez C, Dawson T, Camarero Martínez JJ, Castorena M, et al.
-2018.
+2015. To die or not to die: early warnings of tree dieback in response
+to a severe drought (D Gibson, Ed.). Journal of Ecology 103: 44--57.
+
+Christoffersen BO, Gloor M, Fauset S, Fyllas NM, Galbraith DR, Baker TR,
+Kruijt B, Rowland L, Fisher RA, Binks OJ, et al.~2016. Linking hydraulic
+traits to tropical forest function in a size-structured and trait-driven
+model (TFS v.1-Hydro). Geoscientific Model Development 9: 4227--4255.
+
+Colangelo M, Camarero JJ, Borghetti M, Gazol A, Gentilesca T, Ripullone
+F. 2017. Size Matters a Lot: Drought-Affected Italian Oaks Are Smaller
+and Show Lower Growth Prior to Tree Death. Frontiers in Plant Science 8.
+
+Echeverría A, Anfodillo T, Soriano D, Rosell JA, Olson ME. 2019.
+Constant theoretical conductance, changes in vessel diameter and number
+with height growth in Moringa oleifera. Journal of Experimental Botany.
+
+Jensen KH, Zwieniecki MA. 2013. Physical Limits to Leaf Size in Tall
+Trees. Physical Review Letters 110.
+
+Méndez-Alonzo R, Paz H, Zuluaga RC, Rosell JA, Olson ME. 2012.
+Coordinated evolution of leaf and stem economics in tropical dry forest
+trees. Ecology 93: 2397--2406.
+
+Michaletz ST, Cheng D, Kerkhoff AJ, Enquist BJ. 2014. Convergence of
+terrestrial plant production across global climate gradients. Nature
+512: 39--43.
+
+Olson ME, Anfodillo T, Rosell JA, Petit G, Crivellaro A, Isnard S,
+León-Gómez C, Alvarado-Cárdenas LO, Castorena M. 2014. Universal
+hydraulics of the flowering plants: vessel diameter scales with stem
+length across angiosperm lineages, habits and climates. Ecology Letters
+17: 988--997.
+
+Olson ME, Rosell JA, Martínez-Pérez C, León-Gómez C, Fajardo A, Isnard
+S, Cervantes-Alcayde M-A, Echeverría A, Figueroa-Abúndiz, Segovia-Rivas
+A, et al.~in press. Xylem vessel diameter-shoot length scaling:
+ecological significance of porosity types and other traits. Ecological
+Monographs.
+
+Olson ME, Soriano D, Rosell JA, Anfodillo T, Donoghue MJ, Edwards EJ,
+León-Gómez C, Dawson T, Camarero Martínez JJ, Castorena M, et al.~2018.
 Plant height and hydraulic vulnerability to drought and cold.
-Proceedings of the National Academy of Sciences 115: 7551–7556.
-
-Rosell JA, Olson ME. 2014. Do lianas really have wide vessels? Vessel diameter–stem length scaling in non-self-supporting plants. Perspectives in Plant Ecology, Evolution and Systematics 16: 288–295.
-
-Sack L, Scoffoni C, McKown AD, Rawls M, Havran JC, Tran H, Tran T.
-2012.
-Developmentally based scaling of leaf venation architecture explains global ecological patterns.
-Nature Communications 3: ncomms1835.
-
-Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional trees have similar lifetime carbon gain in tropical secondary forest. Ecology 91: 1102–1113.
-Sillett SC, Van Pelt R, Koch GW, Ambrose AR, Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production through old age in tall trees. Forest Ecology and Management 259: 976–994.
-
-Soriano D, Echeverría A, Anfodillo T, Rosell JA, Olson ME. in press. Hydraulic traits vary as the result of tip-to-base conduit widening in vascular plants. Journal of Experimental Botany.
-
-Wheeler EA, Baas P, Rodgers S. 2007. Variations In Dicot Wood Anatomy: A Global Analysis Based on the Insidewood Database. IAWA Journal 28: 229–258.
-
-Williams CB, Anfodillo T, Crivellaro A, Lazzarin M, Dawson TE, Koch GW. 2019. Axial variation of xylem conduits in the Earth’s tallest trees. Trees 33: 1299–1311.
-
-Zach A, Schuldt B, Brix S, Culmsee H, Culmsee H.
-2010.
-Vessel diameter and xylem hydraulic conductivity increase with tree height in tropical rainforest trees in Sulawesi, Indonesia.
-Flora 205: 506–512.
-
-## Referee: 3
+Proceedings of the National Academy of Sciences 115: 7551--7556.
+
+Rosell JA, Olson ME. 2014. Do lianas really have wide vessels? Vessel
+diameter--stem length scaling in non-self-supporting plants.
+Perspectives in Plant Ecology, Evolution and Systematics 16: 288--295.
+
+Sack L, Scoffoni C, McKown AD, Rawls M, Havran JC, Tran H, Tran T. 2012.
+Developmentally based scaling of leaf venation architecture explains
+global ecological patterns. Nature Communications 3: ncomms1835.
+
+Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional
+trees have similar lifetime carbon gain in tropical secondary forest.
+Ecology 91: 1102--1113. Sillett SC, Van Pelt R, Koch GW, Ambrose AR,
+Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production
+through old age in tall trees. Forest Ecology and Management 259:
+976--994.
+
+Soriano D, Echeverría A, Anfodillo T, Rosell JA, Olson ME. in press.
+Hydraulic traits vary as the result of tip-to-base conduit widening in
+vascular plants. Journal of Experimental Botany.
+
+Wheeler EA, Baas P, Rodgers S. 2007. Variations In Dicot Wood Anatomy: A
+Global Analysis Based on the Insidewood Database. IAWA Journal 28:
+229--258.
+
+Williams CB, Anfodillo T, Crivellaro A, Lazzarin M, Dawson TE, Koch GW.
+2019. Axial variation of xylem conduits in the Earth's tallest trees.
+Trees 33: 1299--1311.
+
+Zach A, Schuldt B, Brix S, Culmsee H, Culmsee H. 2010. Vessel diameter
+and xylem hydraulic conductivity increase with tree height in tropical
+rainforest trees in Sulawesi, Indonesia. Flora 205: 506--512.
+
+\hypertarget{referee-3}{%
+\subsection{Referee: 3}\label{referee-3}}
 
@@
 
-McGregor and colleagues use a combination of long-term tree records and functional trait data to explore the key drivers that shape the growth resistance of different temperate tree species to extreme drought.
-They find that that leaf hydraulic traits and tree size are strong predictors of growth responses to drought – with taller trees most affected.
-
-I enjoyed the paper very much.
-It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology.
-I could definitely see the paper being of interest to a broad readership.
-However, I do have some important questions and suggestions about the analysis which I think need more careful thought.
-Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.
+McGregor and colleagues use a combination of long-term tree records and
+functional trait data to explore the key drivers that shape the growth
+resistance of different temperate tree species to extreme drought. They
+find that that leaf hydraulic traits and tree size are strong predictors
+of growth responses to drought -- with taller trees most affected.
+
+I enjoyed the paper very much. It's clearly and succinctly written, and
+brings together a range of complementary datasets to address some very
+topical questions in forest ecology. I could definitely see the paper
+being of interest to a broad readership. However, I do have some
+important questions and suggestions about the analysis which I think
+need more careful thought. Below I've listed my main concerns, following
+which there are some other comments that the authors may want to take
+into consideration in revising their paper.
 
@@
 
-L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!).
-But doesn’t this introduce a bias in your analysis?
-Trees that died may have had very different responses to drought that those which survived.
-Do you get similar results if you analyse trees that lived and died separately?
-
-**We agree that this could be a concern, but do not believe it is applicable here, for two reasons.
-First, this would be more likely to introduce a bias if the trees cored dead were killed by drought.
-We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
-However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores (Helcoski et al.
-2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with ≥ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
-
-**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**
-
-L165-167: I think this needs to be explained more clearly.
-Why are type I errors unlikely and type II error possible?
-I also think there are real issue with the analysis of crown position.
-Presumably on average the crown illumination index of trees would have increased over the past 50 or so years (from suppressed to dominant) as trees grew and neighbouring trees died.
-So I’m not convinced you can include this in your models going back half a century.
-It’s just not sensible to assume that crown position would not have changed over such a long time period.
-
-L217-226: this is problematic because drought resistance (Rt) as defined by Lloret et al.
-(2011) assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time.
-But this is rarely/never the case with tree growth, which shows clear temporal trends related to tree size/ontogeny.
-If you don’t account for these directional trends then your estimates of Rt will be biased.
-For instance, if a tree is on a long-term trajectory of decreasing BAI over time, when you calculate Rt you will overestimate the impact of the drought (and vice versa for trees on an upward growth trajectory).
-So you need to first detrend the tree ring series to remove long-term growth trends (unless you can show that your trees exhibited no long-term temporal trends in BAI).
-The fact that pre-drought conditions were different for the three droughts is also a problem, but not much you can do about this except acknowledge as you have done.
-
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
-**We do not agree that the $Rt$ metric "assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time." If this was true, $Rt$ would be calculated by dividing growth (BAI) in the drought year by the long-term mean growth increment.
-Instead, $Rt$ is calculated by dividing growth in the focal year by the mean growth over a narrow window of preceding years (5 years in this study).
-Even though there may be a long-term directional trend in BAI, if you were to calculate the trend in BAI over each 5-year increment, the trend could go in any direction.
-To account for such trends, we tried an intervention time series analysis (ARIMA).
-Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
-In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
-We now include the following content on ARIMA in the manuscript:**
-
-(*insert text*)
-
-L260-261: This is rather strange.
-So for the same individual tree, size was important in determining response to drought in one year but not in the others?
-And that’s for the years further back in time when the tree would have been smallest?
-Couldn’t that be interpreted as trees becoming less susceptible to drought as they grow larger/older?
-It’s also a little concerning that results are very different across the three droughts.
-Of these, the 1999 drought is the one I would tend to trust most as stand structure and composition would have been most similar to today.
+L139-140: makes sense to not core live trees in the plot (surprised you
+got permission to do this in 2011!). But doesn't this introduce a bias
+in your analysis? Trees that died may have had very different responses
+to drought that those which survived. Do you get similar results if you
+analyse trees that lived and died separately?
+
+\textbf{We agree that this could be a concern, but do not believe it is
+applicable here, for two reasons. First, this would be more likely to
+introduce a bias if the trees cored dead were killed by drought. We are
+aware of studies showing differences in climate sensitivity of trees
+killed by drought versus those that survived. However,these trees were
+not killed by drought; there was no significant drought in the years in
+which dead trees were cored or in the 3 previous years. Second, quoting
+our previous publication using these cores (Helcoski et al.~2019): "
+Chronologies of trees cored live and dead were pooled following analyses
+showing similar climate sensitivity at least up to 2009 (that is
+excluding 7--8 yr before death) for all four species with ≥ 10 cores in
+each category (LITU, QURU, QUVE, FRAM; comparison plots available at
+\url{https://github.c}
+om/SCBI-ForestGEO/climate\_sensitivity\_cores/tree/master/
+results/live\_vs\_dead).".}
+
+\textbf{To clarify that this shouldn't be an issue, we have added the
+following sentence (methods section): ``We note that drought was
+probably not a cause of mortality for these trees, as monthly May-Aug
+\(PDSI\) did not drop below -1.75 in these years or the three years
+prior (2013-2017), and that trees cored dead displayed similar climate
+sensitivity to trees cored live (Helcoski \emph{et al.}, 2019).''}
+
+L165-167: I think this needs to be explained more clearly. Why are type
+I errors unlikely and type II error possible? I also think there are
+real issue with the analysis of crown position. Presumably on average
+the crown illumination index of trees would have increased over the past
+50 or so years (from suppressed to dominant) as trees grew and
+neighbouring trees died. So I'm not convinced you can include this in
+your models going back half a century. It's just not sensible to assume
+that crown position would not have changed over such a long time period.
+
+\textbf{We no longer attempt to use crown position to describe variation
+in drought resistance. Instead, we now use it to characterize
+\emph{contemporary} differences in microenvironmental conditions faced
+by trees of different heights.}
+
+L217-226: this is problematic because drought resistance (Rt) as defined
+by Lloret et al.~(2011) assumes that tree growth (BAI in this case)
+fluctuates around a long term mean that is stable through time. But this
+is rarely/never the case with tree growth, which shows clear temporal
+trends related to tree size/ontogeny. If you don't account for these
+directional trends then your estimates of Rt will be biased. For
+instance, if a tree is on a long-term trajectory of decreasing BAI over
+time, when you calculate Rt you will overestimate the impact of the
+drought (and vice versa for trees on an upward growth trajectory). So
+you need to first detrend the tree ring series to remove long-term
+growth trends (unless you can show that your trees exhibited no
+long-term temporal trends in BAI). The fact that pre-drought conditions
+were different for the three droughts is also a problem, but not much
+you can do about this except acknowledge as you have done.
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92}{\emph{issue
+\#92}}
+
+\textbf{We do not agree that the \(Rt\) metric ``assumes that tree
+growth (BAI in this case) fluctuates around a long term mean that is
+stable through time.'' If this was true, \(Rt\) would be calculated by
+dividing growth (BAI) in the drought year by the long-term mean growth
+increment. Instead, \(Rt\) is calculated by dividing growth in the focal
+year by the mean growth over a narrow window of preceding years (5 years
+in this study). Even though there may be a long-term directional trend
+in BAI, if you were to calculate the trend in BAI over each 5-year
+increment, the trend could go in any direction. To account for such
+trends, we tried an intervention time series analysis (ARIMA). Details
+on the comparison between \(Rt\) and \(Rt_{ARIMA}\) are presented in the
+responses to reviewer 4. In short, the ARIMA method yielded broadly
+similar results, but tended to produce less reasonable estimates when
+the difference between the two metrics was large, so we decided to
+retain \(Rt\) as the focal metric. We now include the following content
+on ARIMA in the manuscript:}
+
+(\emph{insert text})
+
+L260-261: This is rather strange. So for the same individual tree, size
+was important in determining response to drought in one year but not in
+the others? And that's for the years further back in time when the tree
+would have been smallest? Couldn't that be interpreted as trees becoming
+less susceptible to drought as they grow larger/older? It's also a
+little concerning that results are very different across the three
+droughts. Of these, the 1999 drought is the one I would tend to trust
+most as stand structure and composition would have been most similar to
+today.
 
@@
 
-L42-54: I think it would be worth making the link between growth resilience and probability of mortality.
-Below are links to a couple of papers that have recently shown that trees that are less resilient to drought are also the ones most likely to die.
-https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41467-020-14300-5.epdf%3Fshared_access_token%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw%253D%253D&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=6%2B7kb11Bzq4LivdOhnR4frg1MEFGaX%2FiIMUqJBG9gsM%3D&amp;reserved=0
-https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0
-
-L80-101 & L112-115: This is a nice summary of how traits can help us understand species responses to drought.
-The only thing that is a little strange is that you start by discussing commonly measured traits that have been used to explore susceptibility to drought (e.g., WD and LMA).
-Then you say that the low predictive power of these traits is because they are poor surrogates of what we should be measuring – i.e.
-hydraulic traits.
-But then you end by saying that these are too hard/expensive to measure, so you’re going to look at other surrogates.
-
-L102 & L106: the introduction gives a nice overview of size and trait influences on resistance to drought.
-But the third focus of your study – microenvironment characteristics – gets little or no attention.
-In fact it’s not very clear what you mean by ‘microenvironment’ for most of the paper.
-
-L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?
-
-L173: this is the first mention of what I’m guessing is your ‘microenvironment’ descriptor. Worth briefly mentioning why you are measuring TWI.
-
-L230: Somewhere here it would be good to remind the reader what these explanatory variables are.
-
-L234-237: Not very clear how these models were constructed.
-Why did you fit each predictor independently to test its importance as opposed to standardizing them and having them together in a multiple regression and then compare their coefficients?
-And why was height retained in all models?
-Testing one predictor at a time might be sensible if you have lots of collinearity between predictors.
-But it could also mean you miss important conditional effects (e.g., the effect of variable x only emerges when you first account for the effect of variable y).
-
-L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular).
-So given that you’ve already mentioned how you defined drought in the methods maybe you don’t need to repeat this here?
-
-L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?
+L42-54: I think it would be worth making the link between growth
+resilience and probability of mortality. Below are links to a couple of
+papers that have recently shown that trees that are less resilient to
+drought are also the ones most likely to die.
+\url{https://nam02.safelinks.protection.outlook.com/?url=https\%3A\%2F\%2Fwww.nature.com\%2Farticles\%2Fs41467-020-14300-5.epdf\%3Fshared_access_token\%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw\%253D\%253D\&data=02\%7C01\%7CTeixeiraK\%40si.edu\%7Cda67e36e2ccc4277d4aa08d7ed356305\%7C989b5e2a14e44efe93b78cdd5fc5d11c\%7C0\%7C0\%7C637238685875974418\&sdata=6\%2B7kb11Bzq4LivdOhnR4frg1MEFGaX\%2FiIMUqJBG9gsM\%3D\&reserved=0}
+\url{https://nam02.safelinks.protection.outlook.com/?url=https\%3A\%2F\%2Fwww.nature.com\%2Farticles\%2Fs41558-019-0583-9\&data=02\%7C01\%7CTeixeiraK\%40si.edu\%7Cda67e36e2ccc4277d4aa08d7ed356305\%7C989b5e2a14e44efe93b78cdd5fc5d11c\%7C0\%7C0\%7C637238685875974418\&sdata=DF9lSNfAFTjKmSWfb7FI8\%2FW4GDN4ljKxZK5NNS8eCxc\%3D\&reserved=0}
+
+L80-101 \& L112-115: This is a nice summary of how traits can help us
+understand species responses to drought. The only thing that is a little
+strange is that you start by discussing commonly measured traits that
+have been used to explore susceptibility to drought (e.g., WD and LMA).
+Then you say that the low predictive power of these traits is because
+they are poor surrogates of what we should be measuring --
+i.e.~hydraulic traits. But then you end by saying that these are too
+hard/expensive to measure, so you're going to look at other surrogates.
+
+L102 \& L106: the introduction gives a nice overview of size and trait
+influences on resistance to drought. But the third focus of your study
+-- microenvironment characteristics -- gets little or no attention. In
+fact it's not very clear what you mean by `microenvironment' for most of
+the paper.
+
+L150-152: Given that height is estimated and not measured, wouldn't it
+be better to use DBH as a (measured) dimension of tree size?
+
+L173: this is the first mention of what I'm guessing is your
+`microenvironment' descriptor. Worth briefly mentioning why you are
+measuring TWI.
+
+L230: Somewhere here it would be good to remind the reader what these
+explanatory variables are.
+
+L234-237: Not very clear how these models were constructed. Why did you
+fit each predictor independently to test its importance as opposed to
+standardizing them and having them together in a multiple regression and
+then compare their coefficients? And why was height retained in all
+models? Testing one predictor at a time might be sensible if you have
+lots of collinearity between predictors. But it could also mean you miss
+important conditional effects (e.g., the effect of variable x only
+emerges when you first account for the effect of variable y).
+
+L255-258: good to have this summary, but it's sort of expected that you
+have a decline in growth as this is how you defined drought (a little
+circular). So given that you've already mentioned how you defined
+drought in the methods maybe you don't need to repeat this here?
+
+L266-267: Don't think you can test the effects of crown dominance going
+back this far. How can you be sure a tree was dominant/suppressed back
+in 1966?
 
@@
 
-Tables and Figures: there are a large number of tables.
-I wonder if Table 2 and 3 would best fit in supporting information (with key information integrated directly into the text).
-I also feel that it would be nice to represent some of the key results (e.g., Rt and its relationship to height and leaf hydraulic traits) in a figure.
-
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
-
-## Referee: 4
+Tables and Figures: there are a large number of tables. I wonder if
+Table 2 and 3 would best fit in supporting information (with key
+information integrated directly into the text). I also feel that it
+would be nice to represent some of the key results (e.g., Rt and its
+relationship to height and leaf hydraulic traits) in a figure.
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71}{\emph{issue
+\#71}}
+
+\hypertarget{referee-4}{%
+\subsection{Referee: 4}\label{referee-4}}
 
@@
 
-In this paper, McGregor et al.
-study the effect of tree height, crown position, ring-porosity, wood density, LMA and two leaf water relations traits (percent loss of leaf area upon desiccation, PLA; and the turgor loss point, TLP) on the drought resistance of 12 temperate species from an oak-hickory forest of northern Virginia (USA).
-Drought resistance is assessed using series of radial growth data and for the three most severe droughts in the period 1950 - 2009 in the study area.
-The authors find consistent negative effects of tree height on drought resistance; effects of crown position, ring-porosity and water relation traits for some droughts; and no effect of wood density and LMA.
-The paper is well written and nicely organized around clear hypotheses.
-The results are intriguing but, in my opinion, the experimental design is not strong enough to clearly advance our understanding of the determinants of tree drought responses.
-
-Firstly, the way drought resistance (Rt) is estimated, as the ratio of BAI during drought to the mean BAI over the five years preceding the drought can be problematic, as growth dynamics during this pre-drought period were quite variable across species (Figure 1; incidentally, why do you show ring width indices in this figure if you then use BAI?).
-This approach has also advantages, particularly when applied in a resilience framework (as in the original Lloret et al.
-(2011)).
-However, if you focus specificaly on resistance it seems that directly modelling the growth reduction caused by the drought (e.g., using intervention time series analysis or other methods) would be a more robust approach.
-At the very least, if you continue using your simpler approach I would strongly suggest including some measure of pre-drought growth in the models to account for differences in growth dynamics among species.
-It would also nice to clarify why you do not assess recovery and resilience, unlike Lloret et al.
-(2011).
-
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
-*add specifics (changes to text; SI table/fig #s) to text below*
-
-**In response to Reviewer 4's comments, we tried an intervention time series analysis using the 'auto.arima' function from the 'forecast' package [@R-forecast].
-Specifically, we modeled BAI  based on the timeseries for the 10 years prior to the drought year (itself acting as the intervention).
-We then recorded the ratio between the mean predicted BAI and the observed BAI during the drought year, and compared these results to $Rt$.
-These yielded similar outcomes, which we now present in the Supplementary Info.
-The original $Rt$ and $Rt_{ARIMA}$ were fairly well correlated, but with some scatter (see figs below).
-To get a sense of which metric was performing better, we visually reviewed the records of the 18 trees with largest differences between $Rt$ and $Rt_{ARIMA}$.
-$Rt$ estimates were better for the majority of these, and generally were not hugely unrealistic, whereas $Rt_{ARIMA}$ could be way off (e.g., for several trees that were undergoing growth declines prior to drought and hit minima at or near the drought, $Rt_{ARIMA}$ was much greater than 1, suggesting a substantial growth increase during the drought.).
-Thus, we decided to stick with $Rt$ as the primary metric.**
-
-![Figure S4](tables_figures/publication/figureS4_Rt_arima_comparison.png)
-
-**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
-While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) [DeSoto et al.
-(2020, Nature)](https://www.nature.com/articles/s41467-020-14300-5) used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
-In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
-In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
-These findings would argue that resistance is a more important variable for angiosperms.
-We now state the logic for focusing on $Rt$ in the methods section:**
-
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
-"**
-
-Secondly, the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance, and it mixes different types of hypotheses operating at different levels.
-On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses.
-On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height.
-In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability.
-In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results.
-Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.
-
-Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required.
-The paper assumes that all species were subjected to the same drought levels.
-However, this is unlikely to be true, as different species are likely to occupy different microhabitats in the plot, and hence be subjected to different drought conditions.
-The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.
-
-*It seems from these comments they think this shouldn't be published at all due to its limited scope and not enough explanatory power.
-The resulting study that they suggest, if completely thorough, would take a long time to create.
-I'm not sure how to address this comment, other than to make a note in the manuscript acknowledging that a more thorough study would be great but this is what we found with what we had.*
-
-*I agree.... I wouldn't change analysis for this. We can come back to wording a response later.*
+In this paper, McGregor et al.~study the effect of tree height, crown
+position, ring-porosity, wood density, LMA and two leaf water relations
+traits (percent loss of leaf area upon desiccation, PLA; and the turgor
+loss point, TLP) on the drought resistance of 12 temperate species from
+an oak-hickory forest of northern Virginia (USA). Drought resistance is
+assessed using series of radial growth data and for the three most
+severe droughts in the period 1950 - 2009 in the study area. The authors
+find consistent negative effects of tree height on drought resistance;
+effects of crown position, ring-porosity and water relation traits for
+some droughts; and no effect of wood density and LMA. The paper is well
+written and nicely organized around clear hypotheses. The results are
+intriguing but, in my opinion, the experimental design is not strong
+enough to clearly advance our understanding of the determinants of tree
+drought responses.
+
+Firstly, the way drought resistance (Rt) is estimated, as the ratio of
+BAI during drought to the mean BAI over the five years preceding the
+drought can be problematic, as growth dynamics during this pre-drought
+period were quite variable across species (Figure 1; incidentally, why
+do you show ring width indices in this figure if you then use BAI?).
+This approach has also advantages, particularly when applied in a
+resilience framework (as in the original Lloret et al.~(2011)). However,
+if you focus specificaly on resistance it seems that directly modelling
+the growth reduction caused by the drought (e.g., using intervention
+time series analysis or other methods) would be a more robust approach.
+At the very least, if you continue using your simpler approach I would
+strongly suggest including some measure of pre-drought growth in the
+models to account for differences in growth dynamics among species. It
+would also nice to clarify why you do not assess recovery and
+resilience, unlike Lloret et al.~(2011).
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92}{\emph{issue
+\#92}}
+
+\emph{add specifics (changes to text; SI table/fig \#s) to text below}
+
+\textbf{In response to Reviewer 4's comments, we tried an intervention
+time series analysis using the `auto.arima' function from the `forecast'
+package ({\textbf{???}}). Specifically, we modeled BAI based on the
+timeseries for the 10 years prior to the drought year (itself acting as
+the intervention). We then recorded the ratio between the mean predicted
+BAI and the observed BAI during the drought year, and compared these
+results to \(Rt\). These yielded similar outcomes, which we now present
+in the Supplementary Info. The original \(Rt\) and \(Rt_{ARIMA}\) were
+fairly well correlated, but with some scatter (see figs below). To get a
+sense of which metric was performing better, we visually reviewed the
+records of the 18 trees with largest differences between \(Rt\) and
+\(Rt_{ARIMA}\). \(Rt\) estimates were better for the majority of these,
+and generally were not hugely unrealistic, whereas \(Rt_{ARIMA}\) could
+be way off (e.g., for several trees that were undergoing growth declines
+prior to drought and hit minima at or near the drought, \(Rt_{ARIMA}\)
+was much greater than 1, suggesting a substantial growth increase during
+the drought.). Thus, we decided to stick with \(Rt\) as the primary
+metric.}
+
+\begin{figure}
+\centering
+\includegraphics{tables_figures/publication/figureS4_Rt_arima_comparison.png}
+\caption{Figure S4}
+\end{figure}
+
+\textbf{Regarding the resilience indices, we considered analyzing these
+and decided it would be best to stick with just resistance for three
+reasons: (1) Most importantly, we would expect resilience to be governed
+by fundamentally different mechanisms than resistance. While it would be
+interesting to examine resilience, this would be beyond the scope of the
+current paper. (2)
+\href{https://www.nature.com/articles/s41467-020-14300-5}{DeSoto et
+al.~(2020, Nature)} used an extensive dataset of trees sampled after
+severe droughts and compared the responses to previous, less-severe
+droughts between trees that survived and trees that were killed by the
+severe drought. In angiosperms, trees that died from a severe drought
+had lower resistance to previous droughts, compared to trees that
+survived the severe drought. In gymnosperms, trees that died during
+severe droughts had lower recovery following previous droughts, compared
+to trees that survived the severe drought. These findings would argue
+that resistance is a more important variable for angiosperms. We now
+state the logic for focusing on \(Rt\) in the methods section:}
+
+\textbf{``We focus exclusively on drought resistance (\(Rt\) or
+\(Rt_{ARIMA}\)), and not on the resilience metrics described in Lloret
+\emph{et al.} (2011), because (1) we would expect resilience to be
+controlled by a different set of mechanisms, and (2) the findings of
+\emph{\href{https://www.nature.com/articles/s41467-020-14300-5}{DeSoto
+et al.~(2020)}} suggest that \(Rt\) is a more important drought response
+metric for angiosperms.''}
+
+Secondly, the choice of traits does not appear well justified, as it
+seems to be driven more by easiness of measurement than by relevance,
+and it mixes different types of hypotheses operating at different
+levels. On one hand there is tree height and crown position, which are
+measured at the level of individual trees and correspond to closely
+related hypotheses. On the other hand there are ring-porosity, wood
+density, LMA and water relations traits, which are measured at the
+species level and seem to be totally disconnected from the hypotheses
+regarding tree height. In addition, it is very hard to compare the
+explanatory power of these variables, as done in the paper, as they are
+based on very different sample sizes (571 trees vs 12 species) and
+respond to different sources of variability. In fact the way height
+effects are analyzed mixes the within- and between-species levels, which
+makes it difficult to interpret the results. Finally, I think it would
+be more correct to refer to PLA and TLP as water relations traits rather
+than hydraulic traits.
+
+Thirdly, a more thorough analysis of the distribution of the study
+species in the sample plot would be required. The paper assumes that all
+species were subjected to the same drought levels. However, this is
+unlikely to be true, as different species are likely to occupy different
+microhabitats in the plot, and hence be subjected to different drought
+conditions. The authors account for this, to some extent, by including
+individual-level TWI in some models, but a more thorough analysis of the
+distribution of the study species as a function of TWI and other
+relevant micro-environmental variables would be needed to fully
+interpret the results.
+
+\emph{It seems from these comments they think this shouldn't be
+published at all due to its limited scope and not enough explanatory
+power. The resulting study that they suggest, if completely thorough,
+would take a long time to create. I'm not sure how to address this
+comment, other than to make a note in the manuscript acknowledging that
+a more thorough study would be great but this is what we found with what
+we had.}
+
+\emph{I agree\ldots. I wouldn't change analysis for this. We can come
+back to wording a response later.}
 
@@
 
-L59-79: the arguments presented in this paragraph mix different scales, from within-individual to within- or among-communities.
-I think the arguments could be better structured to reflect the sources of variability involved and, ideally, focus on the scale that will be addressed in the paper.
-
-L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.
-
-L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.
-
-L97: ‘greater variation’ that what?
-
-L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…
-
-L137-140: mixing cores from dying and surviving individuals could be problematic. Did you find any association between survival and previous drought resistance? Was mortality different across species?
+L59-79: the arguments presented in this paragraph mix different scales,
+from within-individual to within- or among-communities. I think the
+arguments could be better structured to reflect the sources of
+variability involved and, ideally, focus on the scale that will be
+addressed in the paper.
+
+L67: I suggest toning down the sentence to something like `Wider xylem
+conduits are associated to higher vulnerability to embolism during
+drought', as the mechanism linking wider conduits with embolism
+vulnerability is far from clear and probably indirect.
+
+L81: you seem to use drought resistance and drought tolerance
+interchangeably, which may be confusing. I think it is important to
+clarify these definitions early on in the paper.
+
+L97: `greater variation' that what?
+
+L100-101: that is not true really. See for example Zhu et al.~(2018) in
+Tree Phys, Farrell et al.~(2017) in PCE\ldots{}
+
+L137-140: mixing cores from dying and surviving individuals could be
+problematic. Did you find any association between survival and previous
+drought resistance? Was mortality different across species?
 
@@
 
-L181: N=3 individuals per species is really quite low.
-In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.
-
-*from NObby: "Regarding the few individuals I would argue that intraspecific SD was low (less than 10% of the mean values). This justifies - in my opinion - only sampling three individuals. "*
-
-L188: ‘Wood’.
-
-**Fixed.**
+L181: N=3 individuals per species is really quite low. In addition, by
+sampling up to 8 m height you would probably be sampling less exposed
+branches from taller trees, which could confound your results.
+
+\emph{from NObby: ``Regarding the few individuals I would argue that
+intraspecific SD was low (less than 10\% of the mean values). This
+justifies - in my opinion - only sampling three individuals.''}
+
+L188: `Wood'.
+
+\textbf{Fixed.}
 
@@
 
-**Yes, now specified.**
+\textbf{Yes, now specified.}
 
@@
 
-L228-234: please clarify model structure.
-I understand you included tree as a random factor only when the three droughts were analyzed together, right?
-It would be nice to know what was the contribution of random factors to model fit (only one R2 is shown in Table 5, and it is unclear whether it corresponds to marginal or conditional values).
-
-L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.
-
-L287-299: it would be very useful to use scatterplots to show the effects of traits on Rt in a figure.
-
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
-
-L291-294: but this is problematic considering that you only had two diffuse-porous species.
-
-**We have made changes to the analysis that resulted in dropping ring porosity from the full models.
-While we agree that it is problematic that we only have two ring-porous species, we believe that it's worth including because this engages with, and rejects, the existing understanding that in this biome, ring-porous species are more drought tolerant.**
+L228-234: please clarify model structure. I understand you included tree
+as a random factor only when the three droughts were analyzed together,
+right? It would be nice to know what was the contribution of random
+factors to model fit (only one R2 is shown in Table 5, and it is unclear
+whether it corresponds to marginal or conditional values).
+
+L238, 242, 244: it is unclear to me why in some cases you use a dAICc of
+1 and sometimes of 2.
+
+L287-299: it would be very useful to use scatterplots to show the
+effects of traits on Rt in a figure.
+
+\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71}{\emph{issue
+\#71}}
+
+L291-294: but this is problematic considering that you only had two
+diffuse-porous species.
+
+\textbf{We have made changes to the analysis that resulted in dropping
+ring porosity from the full models. While we agree that it is
+problematic that we only have two ring-porous species, we believe that
+it's worth including because this engages with, and rejects, the
+existing understanding that in this biome, ring-porous species are more
+drought tolerant.}
 
@@
 
-L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.
+L301-302: I would tone down this, as your models explained a rather low
+\% of the variation in Rt.
 
@@
 
-L314-316: or simply that your sample size was rather low. See also main comments above.
-
-L329-331: do you have any information on the demographic effects (e.g., mortality) of the three droughts studied?
-
-**Unfortunately, no. All we could say is that the 1999 drought didn't result in a big loss of live biomass.**
-
-Table 3: are DBH units correct? I suspect they should be mm instead of cm.
-
-*These are incorrect. Ian, please convert to cm.*
+L314-316: or simply that your sample size was rather low. See also main
+comments above.
+
+L329-331: do you have any information on the demographic effects (e.g.,
+mortality) of the three droughts studied?
+
+\textbf{Unfortunately, no. All we could say is that the 1999 drought
+didn't result in a big loss of live biomass.}
+
+Table 3: are DBH units correct? I suspect they should be mm instead of
+cm.
+
+\emph{These are incorrect. Ian, please convert to cm.}
 
@@
 
-**We now specify that these codes are given in Table 3.**
-
-Figure 2: where are the leaf hydraulic traits shown (cf. caption)?
-
-**It was an error that this was mentioned in the caption; now removed.**
-
-Table S1: why only 11 species? Some of the fits are really low…
-
-**Bark thickness measurements were drawn from an existing data set, and we did not make additional measurements because the trait is unlikely to have meaningful influence on results.
-This study already goes above and beyond the existing norm for accounting for bark thickness in DBH reconstructions; typically there is no correction for changes in bark thickness as the tree ages.**
-
-Table S3: what does ‘rank’ refer to?
-
-**This has been clarified.**
+\textbf{We now specify that these codes are given in Table 3.}
+
+Figure 2: where are the leaf hydraulic traits shown (cf.~caption)?
+
+\textbf{It was an error that this was mentioned in the caption; now
+removed.}
+
+Table S1: why only 11 species? Some of the fits are really low\ldots{}
+
+\textbf{Bark thickness measurements were drawn from an existing data
+set, and we did not make additional measurements because the trait is
+unlikely to have meaningful influence on results. This study already
+goes above and beyond the existing norm for accounting for bark
+thickness in DBH reconstructions; typically there is no correction for
+changes in bark thickness as the tree ages.}
+
+Table S3: what does `rank' refer to?
+
+\textbf{This has been clarified.}
 
@@
 
-*IAN, please do this*
-
-Figure S3: please specify that height values are estimated from reconstructed DBH.
-
-**Done.**
-
-## Citations
+\emph{IAN, please do this}
+
+Figure S3: please specify that height values are estimated from
+reconstructed DBH.
+
+\textbf{Done.}
+
+\hypertarget{citations}{%
+\subsection*{Citations}\label{citations}}
+\addcontentsline{toc}{subsection}{Citations}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-bennett_larger_2015}{}%
+\textbf{\textnormal{Bennett AC}, \textnormal{McDowell NG},
+\textnormal{Allen CD}, \textnormal{Anderson-Teixeira KJ}}.
+\textbf{2015}. Larger trees suffer most during drought in forests
+worldwide. \emph{Nature Plants} \textbf{1}: 15139.
+
+\leavevmode\hypertarget{ref-elliott_forest_2015}{}%
+\textbf{\textnormal{Elliott KJ}, \textnormal{Miniat CF},
+\textnormal{Pederson N}, \textnormal{Laseter SH}}. \textbf{2015}. Forest
+tree growth response to hydroclimate variability in the southern
+Appalachians. \emph{Global Change Biology} \textbf{21}: 4627--4641.
+
+\leavevmode\hypertarget{ref-friedrichs_species-specific_2009}{}%
+\textbf{\textnormal{Friedrichs DA}, \textnormal{Trouet V},
+\textnormal{Büntgen U}, \textnormal{Frank DC}, \textnormal{Esper J},
+\textnormal{Neuwirth B}, \textnormal{Löffler J}}. \textbf{2009}.
+Species-specific climate sensitivity of tree growth in Central-West
+Germany. \emph{Trees} \textbf{23}: 729.
+
+\leavevmode\hypertarget{ref-helcoski_growing_2019}{}%
+\textbf{\textnormal{Helcoski R}, \textnormal{Tepley AJ},
+\textnormal{Pederson N}, \textnormal{McGarvey JC}, \textnormal{Meakem
+V}, \textnormal{Herrmann V}, \textnormal{Thompson JR},
+\textnormal{Anderson‐Teixeira KJ}}. \textbf{2019}. Growing season
+moisture drives interannual variation in woody productivity of a
+temperate deciduous forest. \emph{New Phytologist} \textbf{0}.
+
+\leavevmode\hypertarget{ref-kannenberg_linking_2019}{}%
+\textbf{\textnormal{Kannenberg SA}, \textnormal{Novick KA},
+\textnormal{Alexander MR}, \textnormal{Maxwell JT}, \textnormal{Moore
+DJP}, \textnormal{Phillips RP}, \textnormal{Anderegg WRL}}.
+\textbf{2019}. Linking drought legacy effects across scales: From leaves
+to tree rings to ecosystems. \emph{Global Change Biology} \textbf{0}.
+
+\leavevmode\hypertarget{ref-liu_hydraulic_2019}{}%
+\textbf{\textnormal{Liu H}, \textnormal{Gleason SM}, \textnormal{Hao G},
+\textnormal{Hua L}, \textnormal{He P}, \textnormal{Goldstein G},
+\textnormal{Ye Q}}. \textbf{2019}. Hydraulic traits are coordinated with
+maximum plant height at the global scale. \emph{Science Advances}
+\textbf{5}: eaav1332.
+
+\leavevmode\hypertarget{ref-lloret_components_2011}{}%
+\textbf{\textnormal{Lloret F}, \textnormal{Keeling EG}, \textnormal{Sala
+A}}. \textbf{2011}. Components of tree resilience: Effects of successive
+low-growth episodes in old ponderosa pine forests. \emph{Oikos}
+\textbf{120}: 1909--1920.
+
+\leavevmode\hypertarget{ref-olson_plant_2018}{}%
+\textbf{\textnormal{Olson ME}, \textnormal{Soriano D},
+\textnormal{Rosell JA}, \textnormal{Anfodillo T}, \textnormal{Donoghue
+MJ}, \textnormal{Edwards EJ}, \textnormal{León-Gómez C},
+\textnormal{Dawson T}, \textnormal{Martínez JJC}, \textnormal{Castorena
+M} \emph{et al.}} \textbf{2018}. Plant height and hydraulic
+vulnerability to drought and cold. \emph{Proceedings of the National
+Academy of Sciences} \textbf{115}: 7551--7556.
+
+\leavevmode\hypertarget{ref-stovall_tree_2019}{}%
+\textbf{\textnormal{Stovall AEL}, \textnormal{Shugart H},
+\textnormal{Yang X}}. \textbf{2019}. Tree height explains mortality risk
+during an intense drought. \emph{Nature Communications} \textbf{10}:
+1--6.
+
+\leavevmode\hypertarget{ref-zuleta_drought-induced_2017}{}%
+\textbf{\textnormal{Zuleta D}, \textnormal{Duque A},
+\textnormal{Cardenas D}, \textnormal{Muller‐Landau HC},
+\textnormal{Davies SJ}}. \textbf{2017}. Drought-induced mortality
+patterns and rapid biomass recovery in a terra firme forest in the
+Colombian Amazon. \emph{Ecology} \textbf{98}: 2538--2546.
+
+\end{document}

### 2020-07-16 manuscript/climate_sensitivity_SI.Rmd
worked on crown position
#112
removed Fig. S3

@@
 \newpage
-![Figure S3: Height (from reconstructed DBH) by crown position across the three focal droughts and in the year of measurement (2018)](tables_figures/publication/figureS3_height_plot_analysis.png){width=600px}
+![Figure S3: **PLACEHOLDER** ](tables_figures/publication/independent_variables_by_species.png)
 
 \newpage
-![Figure S4: **PLACEHOLDER** ](tables_figures/publication/independent_variables_by_species.png)
+![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
 
 \newpage
-![Figure S5: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
-
-\newpage
-![Figure S6: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)
+![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
worked on crown position
#112
removed Fig. S3

@@
 Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
-While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative.
-An analysis of crown position relative to height (Fig.
-**2d**) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among crown positions (Fig.
-**S3**), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018.
-However, dominant and co-dominant trees were similar in height (Figs.
-**2d**, **S3**).
 
@@
 We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$ or crown position and $TWI$, and species traits as specified above.
-Crown position and $ln[H]$ were never considered in the same model because of high correlation (0.73) between the two variables. 
+To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$ and species traits as specified above.
 We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 

### 2020-07-17 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd

@@
 
+*from Lawren: "Re the trait values, I believe we would see species differences, even with n=3, looking at the means and SD presented.
+Suggest to run an ANOVA on each trait (and/or a MANOVA on all traits) and put the significance values in a new row in Table 3.
+(BTW also suggest limiting values in Table 3 to 3 sig figs, and presenting SE in that Table with the means rather than SD)."*
+
 *from NObby: "Regarding the few individuals I would argue that intraspecific SD was low (less than 10% of the mean values). This justifies - in my opinion - only sampling three individuals. "*

### 2020-07-17 manuscript/tables&figures.Rmd
added two Fig.4 options to figures

@@
 
-![**Figure 4.
-Visualization of best models for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
-**THIS IS NOT YET CORRECT (having trouble making gg-plot work.
-THE POSITIVE HEIGHT EFFECT FOR 1999 IS WRONG.
-FINAL FIG WILL MATCH MODELS IN TABLE S6.** ](tables_figures/publication/regr_visuals.png){width=500px}
+![**Figure 4_option1. Visualization of best model for all droughts combined.** Model coefficients are given in Table S6.** ](tables_figures/publication/best_model_with_data.png){width=500px}
+
+\newpage
+
+![**Figure 4_option 2.
+Visualization of best model for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
+**confidence intervals would be added.** ](tables_figures/publication/best_models_ind_years.png){width=500px}

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
remove crown position from table 2

@@
 
-Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
+Crown position--a categorical variable classifying trees based on exposure to sunlight--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
+Trees were classified as follows: *dominant* trees were defined as those with crowns above the general level of the canopy, *codominant* trees as those with crowns within the the canopy; *intermediate* trees as those with crowns below the canopy level, but illuminated from above; and *suppressed* as those below the canopy and receiving minimal direct illumination from above.
 

### 2020-07-17 manuscript/tables&figures.Rmd
remove crown position from table 2

@@
 table2 <- read.csv("tables_figures/publication/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2. Summary of dependent and independent variables examined here, along with units, definitions, and sample sizes.", 
+kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2.
+Summary of dependent and independent variables in our statistical models fo drought resistance, along with units, definitions, and sample sizes.",
         escape=FALSE) %>%
@@
   pack_rows("Dependent variables", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 3, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("tree size", 6, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("microhabitat", 8, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("species' traits", 13, 19, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Independent variables", 3, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("species' traits", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
switched tables 2 & 3

@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Second, larger trees may have lower drought resistance because their crowns tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
+Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
-Third, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure.
+Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water.
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019].
-Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height, crown position, root water access, and species' traits.
+Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.
 
@@
 Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2).
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
 First, we focus on the role of tree size and its interaction with microenvironment. 
@@
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
-Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 3).
+Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 2).
 
@@
 
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 2).
+Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 3).
 The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
@@
 
-We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig.
+We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 2; Fig.
 **S1**).
@@
 
-To avoid over-fitting models with five species traits (Table 2) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
+To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
 These tests followed the model structure specified above, including $ln[H]$, $TWI$, and crown position in the null model.
@@
 
-Species traits... (Table 3, Fig. **S4**)
+Species traits... (Table 2, Fig. **S4**)
 
@@
 It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-2), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
+**2**), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. 4; Tables 1,S4-S7). 
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. **4**; Tables 1,S4-S7). 
 Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
@@
 For 1977 and 1999, it was included, with negative coefficient, in some of the top models (Tables S6-S7).
-$\pi_{tlp}$ was included, with negative coefficient, in the best model for all droughts combined and for the 1977 drought individually (Fig. 4; Table 5). 
+$\pi_{tlp}$ was included, with negative coefficient, in the best model for all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
 It was included in some of the top models for 1999 (Tables S6-S7).
@@
 Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig.
-4).
+**4**).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
-The negative effect of height on $Rt$ held after accounting for species' traits.
-We found only a marginal effect of crown exposure with inconsistent direction across two individual droughts.
+*Height was asscociated with...* The negative effect of height on $Rt$ held after accounting for species' traits.
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 However, the strengths of each predictor varied across droughts (Fig.
-4; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+**4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
@@
 **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig.
-4; Tables 1).
+**4**; Tables 1).
 Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
@@
 Concerted measurement of tree-rings and leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
-The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
+The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig.
+**4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
 Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses.
@@
 This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Early- to mid- successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+Early- to mid- successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].

### 2020-07-17 manuscript/tables&figures.Rmd
update table 2 sample sizes
closes #101
switched tables 2 & 3

@@
 
+```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+table3 <- read.csv("tables_figures/publication/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
+kable(table3, format = "latex", booktabs=TRUE, caption = "Table 2.
+Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
+Variable abbreviations are as in Table 3.", escape=FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
+  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
+  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
+kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse.
+We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
+```
+
+\clearpage
+
 ```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
@@
 table2 <- read.csv("tables_figures/publication/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(table2, format = "latex", booktabs = TRUE, caption = "Table 2.
+kable(table2, format = "latex", booktabs = TRUE, caption = "Table 3.
 Summary of dependent and independent variables in our statistical models fo drought resistance, along with units, definitions, and sample sizes.",
@@
 
-*Sample sizes are prior to removal of outliers. Prior to analysis, we discarded # records with $Rt$ >2 and # records with $Rt_{ARIMA}$ >2.
-
-\clearpage
-
-```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
-library(knitr)
-library(kableExtra)
-table3 <- read.csv("tables_figures/publication/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
-kable(table3, format = "latex", booktabs=TRUE, caption = "Table 3.
-Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
-Variable abbreviations are as in Table 2.", escape=FALSE) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
-  #kable_styling(bootstrap_options = "striped", full_width = FALSE)
-kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse.
-We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)
-```
+*Sample sizes are after removal of outliers, and refer to the $Rt$ model. Dashes indicate that the variable was available for all records.
 

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
updating discussion
#112

@@
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
-*Height was asscociated with...* The negative effect of height on $Rt$ held after accounting for species' traits.
+Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig.
+**3**).
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs.
+The negative effect of height on $Rt$ held after accounting for species' traits.
+Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
 This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
-The directions of these responses were consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms.
+The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms.
 However, the strengths of each predictor varied across droughts (Fig.
@@
 
-Our analysis indicates that tree height has an overall stronger influence on drought response than does crown position (Table 1).
-This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
-However, the collinearity between the two variables (Fig.
-**3d**) makes it impossible to confidently partition causality.
-Taller trees are more likely to be in dominant canopy positions (Fig.
-**3d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig.
-**3a-b**).
-Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*).
-*Previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019], and ...* However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig.
-**3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig.
-**4**; Tables 1).
+Consistent with studies in other forests worldwide (**REFS**), taller trees in this forest exhibited lower drought resistance. 
+Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
+Taller trees also face dramatically different microenvironments (Fig. **3**). 
+They are exposed to higher wind speeds and lower humidity (Fig. **3a-b**), resulting in higher evaporative demand (**REFS**). 
+Unlike other temperate forests where modestly cooler understory conditions have been documented (**REFS**), particularly under drier conditions (**REFS**), we observed no significant variation in air temperatures across the vertical profile (Fig.
+**3c**).
+More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman REF*).
+Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
+Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
+Unfortunately, collinearity between height and crown exposure in this study (Fig. **3d**) makes it impossible to confidently partition causality. 
+Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
 Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
+The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
+Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms.
 
-*Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts.
-However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis.
-The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
-It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016].
-The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
-Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
-*
+*Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves .*
 

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
switched figs 2 & 3
#115

@@
 In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig.
-**3**).
+**2**).
 Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-**3a**).
+**2a**).
 Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
-**3b**).
+**2b**).
 Air temperature did not vary across the vertical profile (Fig.
-**3c**).
+**2c**).
 
-Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
+Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). 
 Correlation between height and canopy position was 0.73.
@@
 At the community level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
-**1b**).
+**2b**).
 Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$.
@@
 
-Responses varied across species and by drought (Fig. 2). 
+Responses varied across species and by drought (Fig. **3**). 
 Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
@@
 It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-**2**), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
+**3**), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
 
@@
 Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
-$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. 4; Tables S6-S7). 
+$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
 For 1977 and 1999, it was included, with negative coefficient, in some of the top models (Tables S6-S7).
@@
 Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig.
-**3**).
+**2**).
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 **1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs.
-**1b, 2**).
+**1b**).
 These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
@@
 Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
-Taller trees also face dramatically different microenvironments (Fig. **3**). 
-They are exposed to higher wind speeds and lower humidity (Fig. **3a-b**), resulting in higher evaporative demand (**REFS**). 
+Taller trees also face dramatically different microenvironments (Fig. **2**). 
+They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand (**REFS**). 
 Unlike other temperate forests where modestly cooler understory conditions have been documented (**REFS**), particularly under drier conditions (**REFS**), we observed no significant variation in air temperatures across the vertical profile (Fig.
-**3c**).
+**2c**).
 More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman REF*).
@@
 Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
-Unfortunately, collinearity between height and crown exposure in this study (Fig. **3d**) makes it impossible to confidently partition causality. 
+Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 
 Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
@@
 
-*Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves .*
+*Saving this for references: Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017] .*
 
 The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig.
-**2**) and associated traits at a single site [see also @elliott_forest_2015].
+**3**) and associated traits at a single site [see also @elliott_forest_2015].
 Our study reinforced the findings of previous studies (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
 Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-2).
+**3**).
 While the low $Rt$ of *L.
@@
 The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig.
-**3**).
+**2**).
 Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
@@
 ### Supplementary Information
+
+*redo this list!!*
+
 Table S1: Species-specific bark thickness regression equations

### 2020-07-17 manuscript/tables&figures.Rmd
switched figs 2 & 3
#115

@@
 
-**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.
+**Figure 2.
+Contemporary height profiles in sun exposure and growing season microclimate under non-drought conditions.** Shown are average ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (d) tree heights by 2018 crown position.
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
 
-**Figure 3.
-Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
+**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.
 
@@
 ![**Figure 2.
-Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig.
-1.
-Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
+Contemporary height profiles in sun exposure and growing season microclimate under non-drought conditions.** Shown are average ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Also shown is (d) tree heights by 2018 crown position.
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
 
@@
 
-![**Figure 3.
-Height profiles in growing season climatic conditions and tree heights by crown position.** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes).
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
+![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
 

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
references

@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, where they experience higher solar radiation, greater wind speeds, and lower relative humidity (**REFS--KAT**).
+Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand (**REFS--KAT**)[@kunert_revised_2017].
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
@@
 
-Consistent with studies in other forests worldwide (**REFS**), taller trees in this forest exhibited lower drought resistance. 
+Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance. 
 Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
 Taller trees also face dramatically different microenvironments (Fig. **2**). 
-They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand (**REFS**). 
-Unlike other temperate forests where modestly cooler understory conditions have been documented (**REFS**), particularly under drier conditions (**REFS**), we observed no significant variation in air temperatures across the vertical profile (Fig.
+They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
+Unlike other temperate forests where modestly cooler understory conditions have been documented (*Zellweger et al.
+2019*), particularly under drier conditions (*Davis et al.
+2019*), we observed no significant variation in air temperatures across the vertical profile (Fig.
 **2c**).
-More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman REF*).
+More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman; Rey-Sanchez et al.
+2016*).
 Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
@@
 The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
-Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms.
-
-*Saving this for references: Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017] .*
+Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees fo different size.
 

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
small updates

@@
 
-*To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities].
-We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
-After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.* **(work on organization of this.)**
-
 *Identifying drought years*
@@
 
+To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities].
+We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
+After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.*
+
 Crown position--a categorical variable classifying trees based on exposure to sunlight--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
@@
 The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables as specified below.
-In the multi-year model only, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))
+In the multi-year model, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). 
-Correlation between height and canopy position was 0.73.
+*Correlation between height and canopy position was 0.73.*
 
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
-
-Responses varied across species and by drought (Fig. **3**). 
-Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
 
@@
 Species traits... (Table 2, Fig. **S4**)
+Responses varied across species and by drought (Fig. **3**). 
+Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
 

### 2020-07-17 manuscript/response_to_reviews.Rmd
minor work, mostly on response to reviews

@@
 
-* We slightly modified the method of determining which trait variables go into the full model.
-Specifically, we added the criteria of consistent direction across droughts, and as a result dropped xylem porosity.
-This also makes sense because the two diffuse porous species (Liriodendron tulipifera and Fagus grandifolia) are at opposite ends of the spectrum in terms of drought resistance.
-(Side note: this contradicts the current understanding that diffuse porous species in the temperate deciduous biome are more drought sensitive, but not the bigger global picture, as R2 points out.).
+* We dropped crown position as a factor in the statistical models of drought resistance, and instead include it as part of an enhanced focus on the differences in microenviroment experienced by trees of different height (now in Table 1; Fig.
+2).
+This change was prompted by comments by two reviewers (**R1, R3**) that canopy position could not be reliably projected back to 1966 or disentagled from height.
+
+* We added two figures: one showing Rt differences across species, and one visualizing the model results  (the latter is not yet right, as gg-plot is being difficult, but will eventually match  the model results.)
 
 * We moved former tables 4 and 5 to the SI.
-
-* We added two figures: one showing Rt differences across species, and one visualizing the model results  (the latter is not yet right, as gg-plot is being difficult, but will eventually match  the model results.)
 
@@
 
-**It is true that crown position estimates become more accurate in later dates, but height should be equally reliable across years, having been estimated based on reconstructed DBH.
-This was not previously clearly specified in the methods.
-To clarify, we have added the following statement: "We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$."**
-
-**We agree that the assessment of crown position is somewhat problematic, both because of its increasing inaccuracy going back through time, and because of collinearity with $H$.
-However, given that it is both biologically interesting and comes out in our top models, we have chosen to include it.
-We note that this makes the test of our hypothesis conservative.
-We are transparent about this issue, to which we devote a paragraph in the discussion (par.
-3).**
-
-*see issue on CP!!!!*
+**We no longer attempt to use crown position to describe variation in drought resistance.
+Instead, we now use it to characterize *contemporary* differences in microenvironmental conditions faced by trees of different heights.**
 
@@
 
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+**We have moved two tables (former tables 4 and 5 to the SI). The best models are now visualized in a figure (Fig. 4). We have also added a figure showing drought response by species.**
 
@@
 
+**We agree that this could be a concern, but do not believe it is applicable here, for two reasons.
+First, this would be more likely to introduce a bias if the trees cored dead were killed by drought.
+We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
+However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
+Second, quoting our previous publication using these cores (Helcoski et al.
+2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with ≥ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+
+**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**
+
 L160: you mean a log-log regression, right?

### 2020-07-17 manuscript/response_to_reviews.tex
minor work, mostly on response to reviews

@@
 \item
-  We slightly modified the method of determining which trait variables
-  go into the full model. Specifically, we added the criteria of
-  consistent direction across droughts, and as a result dropped xylem
-  porosity. This also makes sense because the two diffuse porous species
-  (Liriodendron tulipifera and Fagus grandifolia) are at opposite ends
-  of the spectrum in terms of drought resistance. (Side note: this
-  contradicts the current understanding that diffuse porous species in
-  the temperate deciduous biome are more drought sensitive, but not the
-  bigger global picture, as R2 points out.).
-\item
-  We moved former tables 4 and 5 to the SI.
+  We dropped crown position as a factor in the statistical models of
+  drought resistance, and instead include it as part of an enhanced
+  focus on the differences in microenviroment experienced by trees of
+  different height (now in Table 1; Fig. 2). This change was prompted by
+  comments by two reviewers (\textbf{R1, R3}) that canopy position could
+  not be reliably projected back to 1966 or disentagled from height.
 \item
@@
   results.)
+\item
+  We moved former tables 4 and 5 to the SI.
 \end{itemize}
@@
 
-\textbf{It is true that crown position estimates become more accurate in
-later dates, but height should be equally reliable across years, having
-been estimated based on reconstructed DBH. This was not previously
-clearly specified in the methods. To clarify, we have added the
-following statement: ``We then used these allometries to estimate \(H\)
-for each drought year, \(Y\), based on reconstructed \(DBH_Y\).''}
-
-\textbf{We agree that the assessment of crown position is somewhat
-problematic, both because of its increasing inaccuracy going back
-through time, and because of collinearity with \(H\). However, given
-that it is both biologically interesting and comes out in our top
-models, we have chosen to include it. We note that this makes the test
-of our hypothesis conservative. We are transparent about this issue, to
-which we devote a paragraph in the discussion (par. 3).}
-
-\emph{see issue on CP!!!!}
+\textbf{We no longer attempt to use crown position to describe variation
+in drought resistance. Instead, we now use it to characterize
+\emph{contemporary} differences in microenvironmental conditions faced
+by trees of different heights.}
 
@@
 
-\href{https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71}{\emph{issue
-\#71}}
+\textbf{We have moved two tables (former tables 4 and 5 to the SI). The
+best models are now visualized in a figure (Fig. 4). We have also added
+a figure showing drought response by species.}
 
@@
 
+\textbf{We agree that this could be a concern, but do not believe it is
+applicable here, for two reasons. First, this would be more likely to
+introduce a bias if the trees cored dead were killed by drought. We are
+aware of studies showing differences in climate sensitivity of trees
+killed by drought versus those that survived. However,these trees were
+not killed by drought; there was no significant drought in the years in
+which dead trees were cored or in the 3 previous years. Second, quoting
+our previous publication using these cores (Helcoski et al.~2019): "
+Chronologies of trees cored live and dead were pooled following analyses
+showing similar climate sensitivity at least up to 2009 (that is
+excluding 7--8 yr before death) for all four species with ≥ 10 cores in
+each category (LITU, QURU, QUVE, FRAM; comparison plots available at
+\url{https://github.c}
+om/SCBI-ForestGEO/climate\_sensitivity\_cores/tree/master/
+results/live\_vs\_dead).".}
+
+\textbf{To clarify that this shouldn't be an issue, we have added the
+following sentence (methods section): ``We note that drought was
+probably not a cause of mortality for these trees, as monthly May-Aug
+\(PDSI\) did not drop below -1.75 in these years or the three years
+prior (2013-2017), and that trees cored dead displayed similar climate
+sensitivity to trees cored live (Helcoski \emph{et al.}, 2019).''}
+
 L160: you mean a log-log regression, right?
@@
 branches from taller trees, which could confound your results.
+
+\emph{from Lawren: ``Re the trait values, I believe we would see species
+differences, even with n=3, looking at the means and SD presented.
+Suggest to run an ANOVA on each trait (and/or a MANOVA on all traits)
+and put the significance values in a new row in Table 3. (BTW also
+suggest limiting values in Table 3 to 3 sig figs, and presenting SE in
+that Table with the means rather than SD).''}
 

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
minor work, mostly on response to reviews

@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand (**REFS--KAT**)[@kunert_revised_2017].
+Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017].
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].

### 2020-07-18 manuscript/tables&figures.Rmd
new fig 4
#71

@@
 
-![**Figure 4_option1. Visualization of best model for all droughts combined.** Model coefficients are given in Table S6.** ](tables_figures/publication/best_model_with_data.png){width=500px}
-
-\newpage
-
-![**Figure 4_option 2.
-Visualization of best model for all droughts combined and for each drought separately.** Model coefficients are given in Table S6.
-**confidence intervals would be added.** ](tables_figures/publication/best_models_ind_years.png){width=500px}
+![**Figure 4_option1. Visualization of best model for all droughts combined.** Model coefficients are given in Table S6.** ](tables_figures/publication/Figure4_model_vis.png){width=500px}

### 2020-07-18 manuscript/tables&figures.Rmd
update table/fig numbers and file names
#98

@@
 
-```{r, Table 3, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table 2, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-table3 <- read.csv("tables_figures/publication/table3_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-table3[8,8] <- paste0(table3[8,8], footnote_marker_symbol(1))
-kable(table3, format = "latex", booktabs=TRUE, caption = "Table 2.
+table2 <- read.csv("tables_figures/publication/table2_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(1))
+kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2.
 Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
@@
 
-```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r Table 3, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-table2 <- read.csv("tables_figures/publication/table2_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(table2, format = "latex", booktabs = TRUE, caption = "Table 3.
+table3 <- read.csv("tables_figures/publication/table3_variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(table3, format = "latex", booktabs = TRUE, caption = "Table 3.
 Summary of dependent and independent variables in our statistical models fo drought resistance, along with units, definitions, and sample sizes.",
@@
 Also shown is (d) tree heights by 2018 crown position.
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2_NEON_vars_height_profile.png){width=500px}
 
@@
 
-![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
+![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_rt_across_sp.png){width=500px}
 
@@
 
-![**Figure 4_option1. Visualization of best model for all droughts combined.** Model coefficients are given in Table S6.** ](tables_figures/publication/Figure4_model_vis.png){width=500px}
+![**Figure 4.
+Visualization of best models for all droughts combined and for each individual drought year.** Model coefficients are given in Table S6.
+](tables_figures/publication/Figure4_model_vis.png){width=500px}

### 2020-07-18 manuscript/climate_sensitivity_draft.Rmd
update last paragraph of discussion
closes #111

@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Large trees have been disproportionately impacted by strong drought in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height, potentially with some contributions from crown position.
-The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable.
-This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age.
-Early- to mid- successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
+We show that, at least at this site, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system in drier microenvironments.  
+This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
+The same may not be true of short-statured, early-successional forests, which appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits.
+The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance.
+Further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].

### 2020-07-18 manuscript/climate_sensitivity_draft.Rmd
more emphasis in combined-years model in results (started, not complete)
#113

@@
 
-Taller trees showed stronger growth reductions during drought when evaluating the three drought years together and for 1966 individually (Table 1; Fig. **4**). 
-Specifically, $ln[H]$ appeared, with negative coefficient, in the best models and all top models (Tables S6-S7).
+Taller trees showed stronger growth reductions during drought (Table 1; Fig. **4**). 
+Specifically, $ln[H]$ appeared, with negative coefficient, in the best model and all top models when evaluating the three drought years together (Tables S6-S7). 
+The same held true for 1966 individually.
 For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7).
-For the 1999 drought, $ln[H]$ appeared in no top models for $Rt$ and some top models, with positive coefficient, for $Rt_{ARIMA}$.
+For the 1999 drought, $ln[H]$ had no signficant effect.
 

### 2020-07-18 manuscript/climate_sensitivity_draft.Rmd
Stovall's feedback
#110

@@
 There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held after accounting for species' traits.
+The negative effect of height on $Rt$ held after accounting for species' traits, consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al.
+2020*](https://doi.org/10.1038/s41467-020-17213-5)).
 Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
@@
 We show that, at least at this site, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system in drier microenvironments.  
-This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
+This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than belowground environment. 
 The same may not be true of short-statured, early-successional forests, which appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits.

### 2020-07-18 manuscript/climate_sensitivity_draft.Rmd
Zailaa comments
#110
also rearranged folder

@@
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
-Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin (**DEFINE**) correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
@@
 Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
-We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.
+We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011], positively correlated with $LMA$, that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher drought resistance.
 
@@
 Elevations range from 273 to 338 m above sea level with a topographic relief of 65m [@bourg_initial_2013].
-Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm during our study period (1960-2009; source: CRU TS v.4.01; @harris_updated_2014).
+Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm yr^-1^ during our study period [1960-2009; source: CRU TS v.4.01; @harris_updated_2014].
 Dominant tree taxa within this secondary forest include *Liriodendron tulipifera*,  oaks (*Quercus* spp.), and hickories (*Carya* spp.; Table 2).

### 2020-07-18 manuscript/climate_sensitivity_draft.Rmd
started Helcoski comments
#110

@@
 
-Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006].
+Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the forest-dominated terrestrial carbon sink will respond to climate change [@friedlingstein_climatecarbon_2006].
 An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019].
@@
 Many studies have shown that within species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016).
-It has yet to be resolved which of several potential underlying mechanisms most strongly shape size trends in drought response.
-First, tree height may be a primary driver.
+Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016).
+It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response.
+First, tree height itself may be a primary driver.
 Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].

### 2020-07-19 manuscript/tables&figures.Rmd
update figures /legends
#71, #115

@@
 
-**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
-PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
-2014).
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
-Figure modified from Helcoski *et al.* (2019).
-Density plots (**b**) show the distribution of resistance values for each drought.
-
-**Figure 2.
-Contemporary height profiles in sun exposure and growing season microclimate under non-drought conditions.** Shown are average ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
-In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) tree heights by 2018 crown position.
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
-
-**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.
-
-**Figure 4. ...**
+* copy legends here at the end*
 
@@
 In these plots, heights are slightly offset for visualization purposes.
-Also shown is (d) tree heights by 2018 crown position.
+Asterisks indicate signficant differences between the top and bottom of the height profile.
+Also shown is (d) tree heights by 2018 crown position, with letters indicating significance groupings.
 In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2_NEON_vars_height_profile.png){width=500px}
@@
 ![**Figure 4.
-Visualization of best models for all droughts combined and for each individual drought year.** Model coefficients are given in Table S6.
+Visualization of best models for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
+Model coefficients are given in Table S6.
 ](tables_figures/publication/Figure4_model_vis.png){width=500px}

### 2020-07-19 manuscript/climate_sensitivity_draft.Rmd
Helcoski comments
#110

@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-Larger trees may also have lower drought resistance because their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017].
+
+Larger trees may also have lower drought resistance because of microenvironmental and ecological factors.
+Their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017].
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018].
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
+Studies in temperate broadleaf forests have observed that ring-porous species show higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
 2007, Olson et al.
@@
 However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent.
-For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
 Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011].
+
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
@@
 Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
-We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011], positively correlated with $LMA$, that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher drought resistance.
+We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
+We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher drought resistance.
 
@@
 
-We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 2; Fig.
-**S1**).
+We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve dominant species (Table 2; Fig. **S1**). 
+Selected species were those with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$) and together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019].
 Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. 
@@
 Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees).
-Measurement methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018].
+Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018].
 Rangefinders used either the tangent method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon ForestryPro) for calculating heights.
@@
 All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. 
-Other R-packages aside from those already listed were very helpful in conducting analyses. These are listed in the Supplementary Information. 
+Other R-packages used in analyses are listed in the Supplementary Information (*Appendix S1*).
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively.
@@
 **2**).
-Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig.
-**2a**).
-Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig.
-**2b**).
-Air temperature did not vary across the vertical profile (Fig.
-**2c**).
+Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. **2a**). 
+Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. **2b**). 
+Air temperature did not vary across the vertical profile (Fig. **2c**).
 
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). 
+There were significant differences in height across all crown position classes (Fig. **2d**).
 *Correlation between height and canopy position was 0.73.*
@@
 **3**) and associated traits at a single site [see also @elliott_forest_2015].
-Our study reinforced the findings of previous studies (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
+Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
 Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.

### 2020-07-19 manuscript/climate_sensitivity_draft.Rmd

@@
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but rarely consider the roles of tree size and microenvironment.
+However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.
 
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies in temperate broadleaf forests have observed that ring-porous species show higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
+Studies in temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
 2007, Olson et al.
@@
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses.
-Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin (**DEFINE**) correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+Hydraulic traits including water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) and hydraulic safety margin (**DEFINE**) correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
@@
 
-Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
 We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
@@
 We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
-We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
 (2015).
@@
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
-We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher drought resistance.
+We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.
 
@@
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified three drought years - 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3).
+Based on this, we identified the three strongest droughts during the study period (Figs. **1**, **S2**, Table S3).
 
 The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
-The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts.
+The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. 
+In August 1966, $PDSI$ reached its lowest monthly value (-4.82) of the three droughts. 
 The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
-The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).
+The 1999 drought was preceded by wetter than average conditions until the previous June, but $PDSI$ plummeted below -3.0 in October 1998 and remained below this threshold through August 1999.
 

### 2020-07-19 scripts/tables_for_manuscript.R
Converted DBH to cm #116

@@ -3,7 +3,7 @@
 # Developed by Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
 # First created November 2019
 ################################################################
-#Table 3 ####
+#Table 2 ####
 
 # necessary packages
 library(RCurl)
@@ -53,9 +53,9 @@ dbh_cored <- species[,c(1:9)]
 
 dbh_cored <- dbh_cored %>%
   group_by(sp) %>%
-  summarise(mean = mean(dbh2018),
-            min = min(dbh2018),
-            max = max(dbh2018))
+  summarise(mean = mean(dbh2018/10),
+            min = min(dbh2018/10),
+            max = max(dbh2018/10))
 dbh_cored$mean <- round(dbh_cored$mean, 2)
 dbh_cored$range <- paste0(dbh_cored$min, " - ", dbh_cored$max)
 dbh_cored[,c("min", "max")] <- NULL
@@ -81,7 +81,7 @@ table2$percent.ANPP <- c(2, 3.7, 1.1,
 
 
 #get standard deviation
-org_traits <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEKYJX6WD25ZOZTZOFK7CI7AK", stringsAsFactors = FALSE)
+org_traits <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_species_level.csv?token=AJNRBEKXLSB4H62QPT3RZQC7DWCPE", stringsAsFactors = FALSE)
 
 table2$`$PLA$ (\\%)_sd` <- org_traits$PLA_dry_percent_sd[match(table2$sp, org_traits$sp)]
 table2$`$LMA$ ($\frac{g}{cm^2}$)_sd` <- org_traits$LMA_g_per_m2_sd[match(table2$sp, org_traits$sp)]
@@ -103,7 +103,7 @@ table2[8,6] <- "semi-ring*" #change juni ring porosity
 
 
 
-write.csv(table2, "manuscript/tables_figures/publication/table3_species_table.csv", row.names=FALSE)
+write.csv(table2, "manuscript/tables_figures/publication/table2_species_table.csv", row.names=FALSE)
 # write.csv(table2_SI, "manuscript/tables_figures/publication/TableS2.csv", row.names=FALSE)
 
 

### 2020-07-19 manuscript/climate_sensitivity_SI.Rmd
Fig. S3 caption
#100

@@
 \newpage
-![Figure S3: **PLACEHOLDER** ](tables_figures/publication/independent_variables_by_species.png)
+![Figure S3: **Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
+](tables_figures/publication/independent_variables_by_species.png)
 

### 2020-07-20 manuscript/tables&figures.Rmd
minor updates to tables & figs

@@
 library(kableExtra)
-table2 <- read.csv("tables_figures/publication/table2_species_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+table2 <- read.csv("tables_figures/publication/table2_species_table_hand_formatted.csv", stringsAsFactors = FALSE, check.names = FALSE)
 table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(1))
@@
 
-* copy legends here at the end*
+*copy legends here at the end*
 
@@
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png){width=500px}
+Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png)
 
@@
 Also shown is (d) tree heights by 2018 crown position, with letters indicating significance groupings.
-In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2_NEON_vars_height_profile.png){width=500px}
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure2_NEON_vars_height_profile.png)
 
@@
 
-![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_rt_across_sp.png){width=500px}
+![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_rt_across_sp.png)
 
@@
 Model coefficients are given in Table S6.
-](tables_figures/publication/Figure4_model_vis.png){width=500px}
+](tables_figures/publication/Figure4_model_vis.png)

### 2020-07-20 manuscript/climate_sensitivity_draft.Rmd
finished Tepley comments
#110

@@
 
-- Individual-level drought resistance decreased with tree height, which was strongly correlated with crown exposure.
+- Individual-level drought resistance decreased with tree height, which was strongly correlated with *crown exposure*.
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
@@
 In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. 
-In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. 
+Annual tree mortality censuses were initiated in 2014 [@gonzalezakre_patterns_2016], and in 2016-2017, cores were collected from all trees found to have died since the previous year’s. 
 We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
@@
 
-$$DBH_Y  = DBH_{2008} - 2*\left[\sum_{year=Y}^{2008} (r_{ring, Y}) - r_{bark,Y} + r_{bark,2008}\right]$$
+$$DBH_Y  = DBH_{2008} - 2*\left[r_{bark,2008} - r_{bark,Y} + \sum_{year=Y}^{2008} r_{ring, Y} \right]$$
 
@@
 
-Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 to 2019 (n=1,518 trees).
+Tree heights ($H$) were measured by several researchers for a variety of purposes between 2012 and 2019 (n=1,518 trees).
 Methods included direct measurements using a collapsible measurement rod on small trees [@neon_national_2018] or a tape measure on recently fallen trees (this study); geometric calculations using clinometer and tape measure [@stovall_assessing_2018] or digital rangefinders [@andersonteixeira_size-related_2015; @neon_national_2018]; and ground-based LiDAR [@stovall_terrestrial_2018].
@@
 
-For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during drought to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the past ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 The two metrics were strongly correlated (Fig.
 **S5**).
-Because $Rt$ tended to produce more reasonable estimates than $Rt_{ARIMA}$ when there was a large difference between these metrics, we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
+Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
 We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
 
-Analyses focused on testing the predictions presented in Table 1, with $Rt$ (or $Rt_{ARIMA}$) as the response variable. 
+Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
 Models were run for all drought years combined and for each drought year individually. 
-The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [**REF**], with $Rt$ (or $Rt_{ARIMA}$) as the response variable, tree nested within species as a random effect, and independent variables as specified below.
+The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [**REF**], using tree nested within species as a random effect, and independent variables as specified below.
 In the multi-year model, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))
@@
 To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
-These tests followed the model structure specified above, including $ln[H]$, $TWI$, and crown position in the null model.
+These tests followed the model structure specified above, including $ln[H]$ and $TWI$ in the null model.
 Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to this null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
@@
 When a variable appeared in all of these models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction (Table 1).
-If the variable appeared in some but not all of these models, and its sign we was consistent across models, we considered this partial support/rejection. 
+If the variable appeared in some but not all of these models, and its sign was consistent across models, we considered this partial support/rejection. 
 In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.
@@
 Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. **2b**). 
-Air temperature did not vary across the vertical profile (Fig. **2c**).
+Air temperature did not vary consistently across the vertical profile (Fig. **2c**).
 
@@
 This negates the idea that trees in moist microsites would be less affected by drought. 
-Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
+Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
 This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
@@
 In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. **4**; Tables 1,S4-S7). 
-Both had consistent signs across all droughts and explained modest amounts of variation ($\Delta$AICc > 1.0) during at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
+Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
@@
 **2**).
-There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+There was no evidence that greater availability of, or access to soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
 The negative effect of height on $Rt$ held after accounting for species' traits, consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al.
@@
 diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
-This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
+This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
 The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms.
@@
 Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.
-Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
+
+Belowground, taller trees would tend to have larger root systems (**SCALING REF**), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought (**REFS- PANAMA ISOTOPES STUDY**).
+Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (**PANAMA ISOTOPES STUDY**).
+In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
 The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.

### 2020-07-20 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd

@@
 
-**We created a new figure (*Fig. 4*) visualizing the top model results.**  ([*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)()
-
-**We have *also* added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
+**We created two new figures: one showing how $Rt$ varied across species in each of the three droughts (Fig. 3) and one visualizing the best models (Fig. 4).**
 

### 2020-07-20 manuscript/climate_sensitivity_draft.Rmd
Bourg comments
#110

@@
 Many studies have shown that within species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016).
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016).
 It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response.
@@
 The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
-From this census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
+From these census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
 Census data are available through the ForestGEO data portal (www.forestgeo.si.edu).
@@
 **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
-These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
+These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought -- and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
 
@@
 Tree-ring chronologies were developed by RH under guidance of AJT and NP.
-Trait data was collected by IM, JZ under guidance of NK and LS.
+Trait data were collected by IM, JZ under guidance of NK and LS.
 Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM.

### 2020-07-20 manuscript/response_to_reviews.Rmd
Sack comments
closes #110!

@@
 **"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
-"**
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics."**
 
@@
 
-**We have also dramatically decreased the emphasis on differences across drought years, as follows:* **
+**We have also decreased the emphasis on differences across drought years, as follows:* **
 
@@
 
-**(2) The results and discussion now focus on the results from the three drought years combined.**
-
-**(3) In order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
+**(2) To make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
 This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
 (Removed was ring porosity, which had opposite effects in different droughts.)**
+
+**(3) The results and discussion now focus more on the results from the three drought years combined.**
 
@@
 
-**Our preference is to keep this figure in the main manuscript.
-We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing similar height profiles for a temperate forest.
-Such patterns are not shown some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*.
-If we're missing studies that show such vertical profiles, we'd appreciate the references.**
+**Unless the reviewers and editor strongly disagree, we would request to keep this figure to clarify the main manuscript.
+We have revised the manuscript such that this figure is better integrated with the hypothesis testing (Table 1) and interpretation of results.
+We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing this depiction of height profiles for a temperate forest.
+(Such patterns are not shown in some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*.
+If we're missing studies that show such vertical profiles, we'd appreciate the references.)**
 
@@
 227 the nesting seems appropriate and sufficient for taking species/ “phylogeny” into account
+
+**Thank you.**
 

### 2020-07-20 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd

@@
 
-**We now show differences in traits among species in Table S4.**
-
-*need to add ANOVA, prsent reults in paper*
+**We now show differences in traits among species (with signficance groupings) in Figure S4.**
 
@@
 
-*IM I added the correct Latex notation but it's not rendering it when I knit the file... (KAT: I just tried and wasn't able to get it working.*
+**This is an issue with rendering latex in R Markdown tables, and we have some bug that is preventing it from rendering correctly.
+We have noted the problem and will ensure that it's fixed before publication.
+We note that, for *New Phytologist*, SI materials would be transferred to the journal template rather than be in a file generated by R Markdown.**
 

### 2020-07-20 manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 
+**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems (Enquist and Niklas 2001), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
+Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (Meinzer et al.
+1999)."**
+
 73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.
+
+*We now mention this: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (**REF**)."*
 

### 2020-07-20 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses

@@
 Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (**REF**).
 Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al.
@@
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. 
-Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems, which potentially counteracts some of the biophysical challenges they face by allowing greater access to water.
+Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), potentially counteracting some of the biophysical challenges they face by allowing greater access to water.
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. 
@@
 
-Belowground, taller trees would tend to have larger root systems (**SCALING REF**), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought (**REFS- PANAMA ISOTOPES STUDY**).
-Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (**PANAMA ISOTOPES STUDY**).
+Belowground, taller trees would tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
+Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (*Meinzer et al. 1999; DOI: 10.1007/s004420050931*).
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
@@
 Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
-We show that, at least at this site, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system in drier microenvironments.  
-This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than belowground environment. 
-The same may not be true of short-statured, early-successional forests, which appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits.
-The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance.
+We show that, at least within the mature forest studied here, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system, even in drier microenvironments.
+This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
+The same may not be true of systems where short trees exist outside of a buffered understory environment--*i.e.*, open grown trees or short-statured, early-successional forests.
+The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
+The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
 Further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.

### 2020-07-20 manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 
+*We've ensured that language at the start of the paragraph refers to growth: "Many studies have shown that within and across species, large trees tend to be more affected by drought.
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; **REF**].
+"*
+
+**Regarding the issue of whether large trees slow down or not as they age, that's another (very interesting) issue.
+The cited references apply to non-drought situations, and I'd see drought response versus age-related growth patterns as largely independent.
+In this study, it shouldn't matter because our resistance metric takes the ratio of growth during drought relative to what would be expected based on recent years.**
+
 96 this all seems compelling.
@@
 The latter position seems most plausible to me.
+
+**We have revised this paragraph to make the latter point: "Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016]."**
 
@@
 The third sentence did seem to be about tree-microsite correlations.
+
+**We have considerably resivesd this paragraph. It now reads as follows: **
+
+*insert text*
 
@@
 
+**Thank you.**
+
 124 and Table 3.
@@
 
-*this is new terminology to me. Look into Olson et al. in press. Would these wood traits be worth including as covariates?*
+**Thank you.**
 
 127 “tree size”= height?
+
+**Yes; clarified.**
 
@@
 
+**Our trees were mostly above the 3 or 7m threshold.
+It's important to note that the DBH and height at the time of coring did not correspond to that at the time of drought.
+Still, the minimum height was about 5m, so we'd presumably be mostly missing that range of most dramatic changes in conduit diameter.
+It could be interesting to try to align the size range over which Rt changes most dramatically with those over which conduit diameter and microclimate change most rapidly, but unfortunately that's beyond the scope here.**
+
+*need to put something in the text here. Line 138 was where we talked about the sizes of cored trees.*
+
+*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/121*
+
 173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.
@@
 209 It couldn’t hurt to define basal area increment
+
+**Done.**
 
@@
 Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional trees have similar lifetime carbon gain in tropical secondary forest. Ecology 91: 1102–1113.
+
 Sillett SC, Van Pelt R, Koch GW, Ambrose AR, Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production through old age in tall trees. Forest Ecology and Management 259: 976–994.
@@
 
+**This sentence has been re-written: "The 1999 drought was preceded by wetter than average conditions until the previous June, but $PDSI$ plummeted below -3.0 in October 1998 and remained below this threshold through August 1999."**
+
 L228-234: please clarify model structure.

### 2020-07-20 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses

@@
 
-Many studies have shown that within species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding (*e.g.*, @hacket-pain_consistent_2016).
+Many studies have shown that within and across species, large trees tend to be more affected by drought. 
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; **REF**].
 It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
@@
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. 
-Hydraulic traits including water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) and hydraulic safety margin (**DEFINE**) correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation ($PLA_{dry}$) [@scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+Hydraulic traits including water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) and hydraulic safety margin between typical minimum water potentials and P50 or P88 correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More easily measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
+These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
@@
 Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA.
-First, we focus on the role of tree size and its interaction with microenvironment. 
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
+First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
 We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
@@
 
-Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree size, microenvironment characteristics, and species traits (Table 3).
+Within or just outside the ForestGEO plot, we collected data on a suite of variables including tree heights, microenvironment characteristics, and species traits (Table 3).
 The SCBI ForestGEO plot was censused in 2008, 2013, and 2018 following standard ForestGEO protocols, whereby all free-standing woody stems $\ge$ 1cm diameter at breast height (DBH) were mapped, tagged, measured at DBH, and identified to species [@condit_tropical_1998].
@@
 
-For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
@@
 
-*Community-level and species' drought responses*
+*Community-level drought responses*
 
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
+
+In the context of the multivariate model, $Rt$ did not vary across drougth years.
+That is, drought year did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model -- and therefore was excluded from the final model.
+(**IAN, CONFIRM THIS**))
 

### 2020-07-21 scripts/climate_sensitivity_analysis.R
Updated notes in code

@@ -532,7 +532,7 @@ trees_all <- trees_all[!trees_all$tree == 140939, ]
 trees_all <- trees_all[trees_all$resist.value <=2]
 
 # write.csv(trees_all, "manuscript/tables_figures/trees_all.csv", row.names=FALSE)
-##2i. prepare dataset for running regression models ####
+##2j. prepare dataset for running regression models ####
 ##take out columns that are unnecessary for model runs
 trees_all_sub <- trees_all[, !colnames(trees_all) %in% c("p50.MPa", "p80.MPa", "dbh_old.mm",  "dbh_old.cm", "sap_ratio", "height.m")]
 
@@ -542,7 +542,7 @@ trees_all_sub <- trees_all_sub[complete.cases(trees_all_sub), ]
 # write.csv(trees_all_sub, "manuscript/tables_figures/trees_all_sub_arima.csv", row.names=FALSE)
 # write.csv(trees_all_sub, "manuscript/tables_figures/trees_all_sub_arimaratio.csv", row.names=FALSE)
 
-##2j. make subsets for individual years, combine all to list ####
+##2k. make subsets for individual years, combine all to list ####
 x1966 <- trees_all_sub[trees_all_sub$year == 1966, ]
 x1977 <- trees_all_sub[trees_all_sub$year == 1977, ]
 x1999 <- trees_all_sub[trees_all_sub$year == 1999, ]
@@ -551,7 +551,7 @@ model_df <- list(trees_all_sub, x1966, x1977, x1999)
 names(model_df) <- c("trees_all_sub", "x1966", "x1977", "x1999")
 
 #
-##2k. correlation plot ####
+##2l. correlation plot ####
 library(corrplot)
 cr <- model_df[[1]]
 cr$rp <- ifelse(cr$rp == "ring", 1, 2)
@@ -563,7 +563,7 @@ cr[,c("year", "rp", "position_all")] <- sapply(cr[,c("year", "rp", "position_all
 correw <- cor(cr[,-c(2,4,14)])
 corrplot(correw, method="number", type="lower")
 
-##2l. get base stats for everything from trees_all_sub ####
+##2m. get base stats for everything from trees_all_sub ####
 trees_all_sub$dbh.cm <- exp(trees_all_sub$dbh.ln.cm)
 trees_all_sub$height.m <- exp(trees_all_sub$height.ln.m)
 
@@ -884,7 +884,7 @@ for (i in seq(along=c(1:4))){
             delta$model_var <- "dAICc"
             
[…]

### 2020-07-21 manuscript/tables&figures.Rmd
Updated per #119

@@
 Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
-Variable abbreviations are as in Table 3.", escape=FALSE) %>%
+Variable abbreviations are as in Table 3.
+DBH measurements are from the most recent ForestGEO census in 2018.", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
Updated slightly per #118

@@
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. 
-Counteracting the liabilities of associated with tall height, large trees tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), potentially counteracting some of the biophysical challenges they face by allowing greater access to water.
+Counteracting the liabilities associated with tall height, large trees tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. 
-Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought will require sorting out the interactive effects of height and associated exposure, root water access, and species' traits.
+Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought requires sorting out the interactive effects of height and associated exposure, root water access, and species' traits.
 
@@
 Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. 
-Hydraulic traits including water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) and hydraulic safety margin between typical minimum water potentials and P50 or P88 correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
+Hydraulic traits, such as water potentials - thresholds representing a certain percent loss of conductivity (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) - and hydraulic safety margin - between typical minimum water potentials and P50 or P88 - correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
 These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
@@
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
-Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
+Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict drought resistance. 
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
@@
 Crown position--a categorical variable classifying trees based on exposure to sunlight--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999.
-Trees were classified as follows: *dominant* trees were defined as those with crowns above the general level of the canopy, *codominant* trees as those with crowns within the the canopy; *intermediate* trees as those with crowns below the canopy level, but illuminated from above; and *suppressed* as those below the canopy and receiving minimal direct illumination from above.
+Trees were classified as follows: *dominant* trees were defined as those with crowns above the general level of the canopy, *co-dominant* trees as those with crowns within the the canopy; *intermediate* trees as those with crowns below the canopy level, but illuminated from above; and *suppressed* as those below the canopy and receiving minimal direct illumination from above.
 
@@
 **S4**).
-We sampled small sun-exposed branches up to eight meters above ground from three individuals of each species in and around the ForestGEO plot.
+We sampled small, sun-exposed branches up to eight meters above the ground from three individuals of each species in and around the ForestGEO plot.
 Sampled branches were re-cut under water at least two nodes above the original cut and re-hydrated overnight in covered buckets under opaque plastic bags before measurements were taken.
@@
 
-For each drought year, we calculated drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated a metric drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
@@
 Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
-We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
+In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
 (2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
@@
 Models were run for all drought years combined and for each drought year individually. 
-The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [**REF**], using tree nested within species as a random effect, and independent variables as specified below.
-In the multi-year model, we also included a fixed effect of drought year. (**IAN, CONFIRM THIS**))
+The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below.
+In the multi-year model, we also included a fixed effect of drought year to represent the combined effect of each constituent drought scenario.
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 
-To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single GLMM, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
-These tests followed the model structure specified above, including $ln[H]$ and $TWI$ in the null model.
-Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to this null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
-We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models.
+To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
+These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
+Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
+We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models.
 
 We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
-To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$TWI$ and species traits as specified above.
+To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$ln[TWI]$ and species traits as specified above.
 We identified the full set of models within $\Delta$AICc=2 of the best model (that with lowest AICc). 
@@
 
-All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. 
+All analysis beyond basic data collection was performed using R version 3.6.2 [@R-base]. 
 Other R-packages used in analyses are listed in the Supplementary Information (*Appendix S1*).
@@
 Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). 
-There were significant differences in height across all crown position classes (Fig. **2d**).
-*Correlation between height and canopy position was 0.73.*
+There were significant differences in height across all crown position classes (Fig.
+**2d**).
+Despite this, a comparison test between height and crown position data from the most recent ForestGEO census (2018) revealed a correlation of 0.73.
+As early models showed $ln[H]$ had a much stronger effect than crown position, we dropped crown position from later analysis.
 
@@
 
-In the context of the multivariate model, $Rt$ did not vary across drougth years.
-That is, drought year did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model -- and therefore was excluded from the final model.
-(**IAN, CONFIRM THIS**))
+In the context of the multivariate model, $Rt$ did not vary across drought years.
+That is, "drought year" as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model -- and therefore was excluded from the final model as a testing variable.
 
@@
 Taller trees showed stronger growth reductions during drought (Table 1; Fig. **4**). 
-Specifically, $ln[H]$ appeared, with negative coefficient, in the best model and all top models when evaluating the three drought years together (Tables S6-S7). 
+Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 
 The same held true for 1966 individually.
-For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7).
-For the 1999 drought, $ln[H]$ had no signficant effect.
+For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7).
+For the 1999 drought, $ln[H]$ had no significant effect.
 
@@
 This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
-This term did appear, with positive coefficient, in the best $Rt_{ARIMA}$ model for all years combined (Table S7).
+This term did appear, however, with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7).
 
@@
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
-For 1977 and 1999, it was included, with negative coefficient, in some of the top models (Tables S6-S7).
-$\pi_{tlp}$ was included, with negative coefficient, in the best model for all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
-It was included in some of the top models for 1999 (Tables S6-S7).
+For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
+$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
+It was also included in some of the top models for 1999 (Tables S6-S7).
 
@@
 **2**).
-There was no evidence that greater availability of, or access to soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held after accounting for species' traits, consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al.
+There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
+The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al.
 2020*](https://doi.org/10.1038/s41467-020-17213-5)).
 Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
+diffuse porous), and was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
 This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
@@
 **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
-These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought -- and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
+These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
 
@@
 **1b**).
-These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
+These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019].
+However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
 Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
-The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
+The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
 Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees fo different size.
@@
 First, other traits can and do override the influence of xylem porosity on drought resistance. 
-Ring porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to it's drought resistance (Fig. **4**; see discussion below). 
+Ring-porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
 A second explanation of why *F.
-grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less dominant canopy positions compared to most other species (Fig.
+grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig.
 **S4**).
@@
 Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses.
-Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017].
+Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type.
+Further, it adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017].
 Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests.

### 2020-07-21 manuscript/response_to_reviews.Rmd
worked on R2

@@
 
+**Interesting, thanks.**
+
 281 nice
@@
 
+**Thank you.**
+
 334 You might go into detail regarding what you think this says about the causal hypothesis behind the global decline of large individuals.
@@
 
-336 beautifully said “impossible for trees to e ciently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”
+*KAT, come back to this*
+
+336 beautifully said “impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”
 
@@
 
+*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/122*
+
 350 if you wanted to expand your discussion of causal mechanisms or the implications of your findings, I think you could jettison this paragraph.
@@
 
+**We have changed our treatment of canopy position and removed this paragraph.**
+
 382 an effort that seems in harmony with these concerns is (Christoffersen et al., 2016)
+
+**Agreed. We have added this citation.**
 

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
worked on R2

@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
+*This is likely because moister habitats would tend to support species and individuals with more mesophytic traits (e.g., higher $\pi_{tlp}$ and low $PLA_{dry}$, wider conduits), potentially growing to greater heights, and these are then more vulnerable when drought hits.*
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@powell_differences_2017].
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al.
+2016](https://pearl.plymouth.ac.uk/handle/10026.1/12918)) [@powell_differences_2017].
 

### 2020-07-21 manuscript/response_to_reviews.Rmd
sentence on traits and TWI
#122

@@
 
-*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/122*
+**We have added the following text:**
+
+*This may be because moister habitats would tend to support species and individuals with more mesophytic traits (*Bartlett et al.
+2016 Drought.
+DOI: 10.1890/15-0468.1*; *Mencuccini 2003, DOI: 10.1046/j.1365-3040.2003.00991.x*) [@medeiros_extensive_2019], potentially growing to greater heights (e.g., *Detto et al.
+DOI: 10.1371/journal.pone.0076296*), and these are then more vulnerable when drought hits.*
 

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
sentence on traits and TWI
#122

--- 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
+This may be because moister habitats would tend to support species and individuals with more mesophytic traits (*Bartlett et al.
+2016 Drought.
+DOI: 10.1890/15-0468.1*; *Mencuccini 2003, DOI: 10.1046/j.1365-3040.2003.00991.x*) [@medeiros_extensive_2019], potentially growing to greater heights (e.g., *Detto et al.
+DOI: 10.1371/journal.pone.0076296*), and these are then more vulnerable when drought hits.
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.

### 2020-07-21 manuscript/tables&figures.Rmd
DBH in Table 2 heading
#119

@@
 Variable abbreviations are as in Table 3.
-DBH measurements are from the most recent ForestGEO census in 2018.", escape=FALSE) %>%
+DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
+  add_header_above(c(" "," "," ", "contemporary DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
   #kable_styling(bootstrap_options = "striped", full_width = FALSE)

### 2020-07-21 manuscript/climate_sensitivity_SI.Rmd
update R2, Tables S6/S7
#119 , #98

@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/table5_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
@@
 ```
-Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold).
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). $R^2$ refers to conditional $R^2$.
 
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
#119

@@
 **1b**).
-These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019].
+These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019].
 However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
@@
 Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
-Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
+Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
 Unfortunately, collinearity between height and crown exposure in this study (Fig. **2d**) makes it impossible to confidently partition causality. 

### 2020-07-21 manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 
-*KAT, come back to this*
+**Unfortunately, we ended up concluding that we can't reliably parse the effects of height *per se* and canopy position in this forest, due to collinearity of height and canopy position, and also because of increasing uncertainty in canopy position as we go back in time.
+Thus, the paragraph is now modified to discuss both height and associated microenvironment.**
 
@@
 Of these, the 1999 drought is the one I would tend to trust most as stand structure and composition would have been most similar to today.
+
+**We don't find it surprising that height could be more important in some droughts than in others.
+We address this in the second paragraph of the discussion: "Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
+**S2**), severely stressed a larger portion of trees (Fig.
+**1b**).
+The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]."**
+
+**It's important to keep in mind that we're using height estimates specific to each drought year, and these estimates should be fairly reliable (we've dropped CP, which we decided was too unreliable).
+We now clarify that height is specific to the drought year: "Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig.
+**4**)." While the distribution of heights did shift upwards with progressive droughts, this shift is modest relative to the total height range (a few meters relative to a height range of ~40 m).**
+
+*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/121*
+
+**There are a number of factors-- including height distribution and also atmospheric CO2 -- that differ across the droughts.
+To account for this as best as possible, we included year as a fixed effect in all statistical models (even though it did not come out significant).
+This is now clearly specified in the methods: "In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics (Fig.
+**S2**)." There were no signficant differences across years, which we now specify in the results.**
 

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses

@@
 The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below.
-In the multi-year model, we also included a fixed effect of drought year to represent the combined effect of each constituent drought scenario.
+In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics (Fig. **S2**).
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 In the context of the multivariate model, $Rt$ did not vary across drought years.
-That is, "drought year" as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model -- and therefore was excluded from the final model as a testing variable.
+That is, "drought year" as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model.
 
@@
 
-Taller trees showed stronger growth reductions during drought (Table 1; Fig. **4**). 
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
 Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 

### 2020-07-21 manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 Second, quoting our previous publication using these cores (Helcoski et al.
-2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with ≥ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
 
@@
 
-**There are a number of factors-- including height distribution and also atmospheric CO2 -- that differ across the droughts.
+**There are a number of factors-- including height distribution and also atmospheric CO2 -- that differ across the droughts and may bias the mean response.
 To account for this as best as possible, we included year as a fixed effect in all statistical models (even though it did not come out significant).
 This is now clearly specified in the methods: "In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics (Fig.
-**S2**)." There were no signficant differences across years, which we now specify in the results.**
+**S2**)." There were no significant differences across years, which we now specify in the results.**
 
@@
 https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41467-020-14300-5.epdf%3Fshared_access_token%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw%253D%253D&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=6%2B7kb11Bzq4LivdOhnR4frg1MEFGaX%2FiIMUqJBG9gsM%3D&amp;reserved=0
+
+ *[DeSoto et al. (2020)](https://www.nature.com/articles/s41467-020-14300-5)*
+
 https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0
+
+*Liu et al. 2019; DOI: 10.1038/s41558-019-0583-9*
+
+**Thank you. We have added the following statement:**
+
+*Because the resistance and resilience of growth to drought is linked to their probability of surviving drought (*[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)*; *Liu et al.
+2019; DOI: 10.1038/s41558-019-0583-9*), understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.*
 
@@
 
+**We have revised this section, and believe that it now makes our points more clearly: (1) the traditionally measured traits are poor surrogates, (2) hydraulic traits are useful but time-consuming, (3) $PLA_{dry}$ and $\pi_{tlp}$ are promising alternatives that are both relatively easy to measure and have direct physiological linkages to tree growth and mortality responses to drought.
+The text now reads as follows:**
+
+*insert 2 paragraphs on traits in intro*
+
 L102 & L106: the introduction gives a nice overview of size and trait influences on resistance to drought.
@@
 L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?
+
+**We prefer to use height because it is more meaningful biologically. DBH would give the same result.**
 
@@
 
-*These are incorrect. Ian, please convert to cm.*
+**These were incorrect (in mm). Now converted to cm.**
 
@@
 
-*IAN, please do this*
+**Done.**
 

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses

@@
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
+Because the resistance and resilience of growth to drought is linked to their probability of surviving drought (*[DeSoto et al.
+(2020)](https://www.nature.com/articles/s41467-020-14300-5)*; *Liu et al.
+2019; DOI: 10.1038/s41558-019-0583-9*), understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
 However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.
@@
 
-Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. 
-Hydraulic traits, such as water potentials - thresholds representing a certain percent loss of conductivity (e.g., P50, P80, P88, representing 50, 80, and 88% loss of conductivity, respectively) - and hydraulic safety margin - between typical minimum water potentials and P50 or P88 - correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
+For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance [@anderegg_hydraulic_2018].
+However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 

### 2020-07-21 manuscript/response_to_reviews.Rmd
worked on R3

@@
 
+**We have placed the focus primarily on tree size and traits because that's where we see the most interesting story.
+We are interested in microenvironment primarily in how it interacts with tree size--i.e., in how microclimate varies across the vertical forest gradient, and in whether larger root systems are an advantage in drier microsites.
+We have made a number of changes that should make this clearer.
+For instance, we now have a paragraph in the introduction starting as follows "Larger trees may also have lower drought resistance because of microenvironmental and ecological factors." We then discuss how larger trees have more exposed crowns, and how larger root systems could be advantageous in drier microsites during drought.**
+
 L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?
@@
 
+**We have revised this sentence to read, "Topographic wetness index (TWI), *used here as a metric of long-term mean moisture availability*,..**
+
 L230: Somewhere here it would be good to remind the reader what these explanatory variables are.
+
+**We have revised this sentence to read, "The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below." We later specify the list of variables.**
 
@@
 
+*(just need to write a response here-- no further changes needed to manuscript)*
+
 L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular).
@@
 
+**We have dropped the criteria of decline in growth as part of how we defined drought. This does not change the drought years that qualify for inclusion.**
+
 L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?
+
+**We have dropped crown position from the model, and now use it only to give a picture of how microenvironmental conditions vary with tree height under contemporary conditions (Fig. 2).**
 

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
worked on R3

@@
 Counteracting the liabilities associated with tall height, large trees tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
+Larger root systems--if they grant access to deeper water sources--would be particularly advantageous in drier microenvironments (e.g., hilltops, as compared to valleys and streambeds) during drought.
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. 
@@
 
-Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Fig.
+Topographic wetness index (TWI), used here as a metric of long-term mean moisture availability, was calculated using the dynatopmodel package in R (Fig.
 **S1**) [@R-dynatopmodel].

### 2020-07-21 manuscript/response_to_reviews.Rmd
finished R3!
worked on R4 comments

@@
 
+**This has been revised to read, "In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7)."**
+
 Tables and Figures: there are a large number of tables.
@@
 
+**We respond to this one point at a time:**
+
+*the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance*
+
+**Ease of measurement is part of what makes $PLA_{dry}$ and $\pi_{tlp}$ interesting. As we now clarify in the introduction:**
+
+**"However, [traits such as P50, P88, and hydraulic safety margin] are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
+These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].*The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.*"**
+
+*it mixes different types of hypotheses operating at different levels.
+On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses.
+On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height.
+In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability.
+In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results.*
+
+**There is nothing fundamentally wrong with a model comparing individual- and species- level effects, but we understand why it was confusing and believe that the revised presentation of the analysis should make all of this easier to follow.
+Improvements related to this comment include: (1) revision to how traits are selected for inclusion as candidate variables for the full model, which provides a more appropriate test of individual traits and guards against over-fitting; (2) moving of former tables 4 and 5 to the SI, and replacing fig.
+5 with a figure visualizing the model results; (3) adding a couple new figures showing variation in $Rt$ (Fig.
+3) and predictor variables (Fig.
+S3) across species; and (4) dropping canopy position from the model for predicting $Rt$.
+
+*Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.*
+
+*RESPONSE NEEDED HERE!*
+
 Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required.
@@
 
-*It seems from these comments they think this shouldn't be published at all due to its limited scope and not enough explanatory power.
-The resulting study that they suggest, if completely thorough, would take a long time to create.
-I'm not sure how to address this comment, other than to make a note in the manuscript acknowledging that a more thorough study would be great but this is what we found with what we had.*
-
-*I agree.... I wouldn't change analysis for this. We can come back to wording a response later.*
+**We have added a new figure to the SI showing variation in predictor variables across species (Fig.
+S3).
+While there was some variation across species in $H$ and $TWI$, there were few pronounced differences across species.
+We now discuss the case of the most anomalous species, *Fagus grandifolia*, in the discussion section.
+We also determined that there is little systematic variation in $PLA_{dry}$ and $\pi_{tlp}$ with respect to $H$ or $TWI$ (analysis not shown).
+This, together with the fact that our statistical models do account for $TWI$, should minimize bias related to distribution of the study species within the plot**
 
@@
 
+**We have revised this paragraph to provide some more specifics on the scale and setting of these tests.**
+
 L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.
+
+**We have revised this to read, "Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018]..."**
 
@@
 
+*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/124*
+
 L97: ‘greater variation’ that what?
 
+**This was removed.
+The text here now reads, "More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
+These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.
+2018*].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al.
+2018*; but see *Farrell et al.
+2017*].
+The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
+"**
+
 L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…
+
+**We have added citations to both these papers and revised the sentence so that it is true: "The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**
 
@@
 
+**Yes. fixed.**
+
 L181: N=3 individuals per species is really quite low.
@@
 
-*from Lawren: "Re the trait values, I believe we would see species differences, even with n=3, looking at the means and SD presented.
-Suggest to run an ANOVA on each trait (and/or a MANOVA on all traits) and put the significance values in a new row in Table 3.
-(BTW also suggest limiting values in Table 3 to 3 sig figs, and presenting SE in that Table with the means rather than SD)."*
+**More samples would be ideal, but was not feasible given the time and resources available, and have no way to sample more now (in part because of COVID-19).
+We note that intraspecific SD was generally low with respect to mean values and interspecific variation, and that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019].
+We have added a figure to the SI (Fig.
+S3) showing variation in traits across species.**
 
-*from NObby: "Regarding the few individuals I would argue that intraspecific SD was low (less than 10% of the mean values). This justifies - in my opinion - only sampling three individuals. "*
+*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/125*
 

### 2020-07-21 manuscript/climate_sensitivity_SI.Rmd
update S3
worked on R4 comments

@@
 \newpage
+
 ![Figure S3: **Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
-](tables_figures/publication/independent_variables_by_species.png)
+](tables_figures/publication/FigureS3_traits_signif.png)
 
 \newpage
+
 ![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
@@
 \newpage
+
 ![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)

### 2020-07-21 manuscript/climate_sensitivity_draft.Rmd
finished R3!
worked on R4 comments

@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies in temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
 2007, Olson et al.
 2020*) and does not resolve differences among the many species within each category.
-Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent.
+Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
+However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent.
 For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
@@
 In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
-For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance [@anderegg_hydraulic_2018].
+For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests (*[Anderegg et al.
+2016](www.pnas.org/cgi/doi/10.1073/pnas.1525678113)*).
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
-These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].
-The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.
+These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.
+2018*].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al.
+2018*; but see *Farrell et al.
+2017*].
+The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
 
@@
 Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
-Species-specific height allometries were developed (Table S2) using logarithmic regression ($ln[H] \sim ln[DBH]$).
+Species-specific height allometries were developed (Table S2) using log-log regression ($ln[H] \sim ln[DBH]$).
 For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species.
@@
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were linked to drought responses (Fig. **4**; Tables 1,S4-S7). 
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7). 
 Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
@@
 **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
-Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019], and our findings indicate that these traits also influence drought responses.
+Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; *Zhu et al.
+2018*], and our findings indicate that these traits also influence drought responses.
 Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type.

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
added Fig. S3, rearranged SI figures

@@
 heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
-kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S2: Species-specific height regression equations", escape = FALSE) %>%
+kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S2. Species-specific height regression equations", escape = FALSE) %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
@@
 tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
+kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3.
+Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
 ") %>%
@@
 \newpage
-![Figure S1: Map of ForestGEO plot showing TWI and location of cored trees](tables_figures/publication/figureS1_location_cored_trees.png){width=500px}
+![**Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought**](tables_figures/publication/figureS1_drought_plot.jpg)
 
 \newpage
-![Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought](tables_figures/publication/figureS2_drought_plot.jpg){width=600px}
+![**Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees**](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}
 
@@
 
-![Figure S3: **Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
-](tables_figures/publication/FigureS3_traits_signif.png)
+![**Figure S3. Distribution of reconstructed tree heights across drought years.** ](tables_figures/publication/FigureS3_height_histogram.png)
 
@@
 
-![Figure S4: Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+![**Figure S4.
+Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
+](tables_figures/publication/FigureS4_traits_signif.png)
 
@@
 
-![Figure S5: **PLACEHOLDER** ](tables_figures/publication/figureS5_species Rt differences.png)
+![**Figure S5. Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario**](tables_figures/publication/figureS5_Rt_arima_comparison.png)

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
added Fig. S3, rearranged SI figures

@@
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified the three strongest droughts during the study period (Figs. **1**, **S2**, Table S3).
+Based on this, we identified the three strongest droughts during the study period (Figs. **1**, **S1**, Table S3).
 
-The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
+The droughts differed in intensity and antecedent moisture conditions (Fig. **S1**, Table S3). 
 The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. 
@@
 
-We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve dominant species (Table 2; Fig. **S1**). 
+We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve dominant species (Table 2; Fig. **S2**). 
 Selected species were those with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$) and together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019].
@@
 We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$.
+The distribution of $H$ across drought years is shown in Fig. **S3**.
 
@@
 Topographic wetness index (TWI), used here as a metric of long-term mean moisture availability, was calculated using the dynatopmodel package in R (Fig.
-**S1**) [@R-dynatopmodel].
+**S2**) [@R-dynatopmodel].
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
@@
 The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below.
-In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics (Fig. **S2**).
+In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 
-$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. 4, Table S6). 
+$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. **4**, Table S6). 
 When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1966 and 1999 (Table S7).
@@
 Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
-**S2**), severely stressed a larger portion of trees (Fig.
+**S1**), severely stressed a larger portion of trees (Fig.
 **1b**).

### 2020-07-22 manuscript/tables&figures.Rmd
update Fig. 2

--- 2020-07-21 manuscript/tables&figures.Rmd
@@
 Variable abbreviations are as in Table 3.
-DBH measurements are from the most recent ForestGEO census in 2018.", escape=FALSE) %>%
+DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
-  add_header_above(c(" "," "," ", "DBH (cm)" = 2, "species traits (mean +/- sd)" = 5)) %>%
+  add_header_above(c(" "," "," ", "contemporary DBH (cm)" = 2, "species traits (mean +/- se)" = 5)) %>%
   #kable_styling(bootstrap_options = "striped", full_width = FALSE)

### 2020-07-22 manuscript/response_to_reviews.Rmd
Update response_to_reviews.Rmd
mostly figure numbering

@@
 
-*We do plot the three droughts separately in figures **#,#**, as we do fee that this makes the story richer.*
+**We do still present results from the three droughts separately (e.g., Figs. 1,3,4; Tables S4-S7), as we do feel that this makes the story richer.**
 
@@
 
-**We have also added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
+**We have also added a new figure (*Fig. 3*) showing how $Rt$ varied across species in each of the three droughts.**
 
@@
 
-**To address the question about species, we have added a new figure (*Fig. 2*) showing how $Rt$ varied across species in each of the three droughts.**
+**To address the question about species, we have added a new figure (*Fig. 3*) showing how $Rt$ varied across species in each of the three droughts.**
 
@@
 
-**We now show differences in traits among species (with signficance groupings) in Figure S4.**
+**We now show differences in traits among species (with significance groupings) in Figure S4.**
 
@@
 **We dug into the ring/diffuse porous distinction more, and ended up making a few substantive changes that relate to this theme: (1) We added a new figure showing $Rt$ across species (Fig.
-2).
+3).
 This illustrates that the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum.
@@
 
-**Our trees were mostly above the 3 or 7m threshold.
+**Our trees were mostly above the 3 or 7m threshold (see Fig.
+S3).
 It's important to note that the DBH and height at the time of coring did not correspond to that at the time of drought.
@@
 We address this in the second paragraph of the discussion: "Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
-**S2**), severely stressed a larger portion of trees (Fig.
+**S1**), severely stressed a larger portion of trees (Fig.
 **1b**).
@@
 
-**It's important to keep in mind that we're using height estimates specific to each drought year, and these estimates should be fairly reliable (we've dropped CP, which we decided was too unreliable).
+**It's important to keep in mind that we're using height estimates specific to each drought year, and these estimates should be fairly reliable (we've dropped crown position).
 We now clarify that height is specific to the drought year: "Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig.
-**4**)." While the distribution of heights did shift upwards with progressive droughts, this shift is modest relative to the total height range (a few meters relative to a height range of ~40 m).**
-
-*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/121*
+**4**)." While the distribution of heights did shift upwards with progressive droughts, this shift is modest relative to the total height range (a few meters relative to a height range of ~40 m; Fig.
+S3).**
 
@@
 To account for this as best as possible, we included year as a fixed effect in all statistical models (even though it did not come out significant).
-This is now clearly specified in the methods: "In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics (Fig.
-**S2**)." There were no significant differences across years, which we now specify in the results.**
+This is now clearly specified in the methods: "In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics." There were no significant differences across years, which we now specify in the results.**
 
@@
 
-![Figure S4](tables_figures/publication/figureS4_Rt_arima_comparison.png)
+![Figure S5](tables_figures/publication/figureS5_Rt_arima_comparison.png)
 
@@
 3) and predictor variables (Fig.
-S3) across species; and (4) dropping canopy position from the model for predicting $Rt$.
+S4) across species; and (4) dropping canopy position from the model for predicting $Rt$.**
 
@@
 **We have added a new figure to the SI showing variation in predictor variables across species (Fig.
-S3).
+S4).
 While there was some variation across species in $H$ and $TWI$, there were few pronounced differences across species.
@@
 We have added a figure to the SI (Fig.
-S3) showing variation in traits across species.**
+S4) showing variation in traits across species.**
 
 *https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/125*
+
+**The sampling strategy was consistent across species, and also consistent with general methods for leaf trait sampling.
+Because we were not testing species-level trait response (as opposed to sampling traits on every cored tree), we don't expect any bias related to sampling height.**
 
@@
 
-**We now specify that these codes are given in Table 3.**
+**We now specify that these codes are given in Table 2.**
 

### 2020-07-22 manuscript/response_to_reviews.Rmd
"leaf hydraulic traits" --> "leaf drought tolerance traits"
I think I got everything. Closes #126

@@
 **"However, [traits such as P50, P88, and hydraulic safety margin] are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
+More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
 These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
@@
 
-*RESPONSE NEEDED HERE!*
+**We now refer to these as "leaf drought tolerance traits", including in the manuscript title.**
 
@@
 **This was removed.
-The text here now reads, "More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
+The text here now reads, "More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
 These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
"leaf hydraulic traits" --> "leaf drought tolerance traits"
I think I got everything. Closes #126

@@
 
-**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
+**Title:** Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest
 
@@
 
-- We conclude that tree height and hydraulic traits influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
+- We conclude that tree height and leaf drought tolerance traits influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
 Thus, these factors can be useful for predicting future drought responses under climate change.
 
-*Key words*: annual growth; crown exposure; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree height; tree-ring
+*Key words*: annual growth; crown exposure; drought; Forest Global Earth Observatory (ForestGEO); leaf drought tolerance traits; temperate broadleaf deciduous forest; tree height; tree-ring
 
@@
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable leaf hydraulic traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
+More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
 These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.
@@
 Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
 First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
@@
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
-Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict drought resistance. 
+Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf leaf drought tolerance traits-–predict drought resistance. 
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
@@
 
-Hydraulic traits were collected in August 2018 (Tables 2-3; Fig.
+Species' trait data were collected in August 2018 (Tables 2-3; Fig.
 **S4**).
@@
 
-Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig.
+Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig.
 **4**).
@@
 Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), and was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$).
+diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$).
 This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
@@
 
-Concerted measurement of tree-rings and leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response.
+Concerted measurement of tree-rings and leaf drought tolerance traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of drought tolerance traits in shaping drought response.
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig.
@@
 The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al.
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al.
 2016](https://pearl.plymouth.ac.uk/handle/10026.1/12918)) [@powell_differences_2017].
@@
 The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
-Further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].
@@
 Thompson, and Victoria Meakem for original collection and processing of cores.
-Thanks also to Camila D. Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection. 
+Thanks also to Camila D.
+Medeiros for guidance on leaf drought tolerance and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection.
 This manuscript was improved based on helpful reviews by Mark Olson and three anonymous reviewers.

### 2020-07-22 manuscript/tables&figures.Rmd
"leaf hydraulic traits" --> "leaf drought tolerance traits"
I think I got everything. Closes #126

@@
   pack_rows("Species traits", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
+  pack_rows("Species' traits--particularly leaf drought tolerance traits--predict drought reistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
   #add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%

### 2020-07-22 manuscript/response_to_reviews.Rmd
finished first pass at reviewer responses!!!!

@@
 
-*Tree-ring data inherently have a lot of variation. Neil could probably write a better response here, with refs.*
+**Our analysis of $Rt_{ARIMA}$ (described above) confirms the findings based on analysis of $Rt$.**
 
@@
 
-**Ease of measurement is part of what makes $PLA_{dry}$ and $\pi_{tlp}$ interesting. As we now clarify in the introduction:**
+**Ease of measurement is part of what makes $PLA_{dry}$ and $\pi_{tlp}$ interesting. We now clarify this in the introduction:**
 
@@
 
-**More samples would be ideal, but was not feasible given the time and resources available, and have no way to sample more now (in part because of COVID-19).
-We note that intraspecific SD was generally low with respect to mean values and interspecific variation, and that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019].
+**There were significant differences among species for all sampled traits, and intraspecific SD was generally low with respect to mean values and interspecific variation.
 We have added a figure to the SI (Fig.
-S4) showing variation in traits across species.**
-
-*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/125*
+S4) showing variation in traits across species.
+We also note that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019].
+More samples would be ideal, but was not feasible given the time and resources available.
+**
 
@@
 
+**We have substantially revised this section and expect that it should be far more clear. The text in question now reads:**
+
+**"The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
+In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
+Individual year models included a random effect of species.
+All models included fixed effects of independent variables of interest (Tables 1,3) as specified below."**
+
+**We have specified that R2 corresponds to conditional values (footnotes of former table 5/ current table S6, and table S7).**
+
 L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.
+
+**dAICc = 2 was consistently used to test for significance, whereas dAICc=1 was used as a preliminary threshhold for variable selection.
+We now clearly specify this in the footnote of table S4 (former table 4) and in the text: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
+We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models."**
 
@@
 
-[*issue #71*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/71)
+**We have added the specified figure (Fig.
+S6).
+In addition, we have added 3 related figures illustrate how traits and $Rt$ vary across species (Figs.
+S4 and 3, respectively) and how $Rt$ responds to  $PLA_{dry}$ and $\pi_{tlp}$.**
 
@@
 
+**This text was removed.**
+
 L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.
+
+**Tree-ring data are inherently noisy, particularly in more mesic forests.
+In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree.
+These factors contribute to the low R^2^, but do not negate the significance of our findings.**
 
@@
 
+**This text was removed.**
+
 L314-316: or simply that your sample size was rather low. See also main comments above.
+
+**Our sample size of tree-ring data is quite good, and we believe the traits sample size to be sufficient (see above).
+We have now tried numerous iterations of this analysis, and the result described here holds.
+Biologically, it also makes sense that the importance of height, TWI, and traits would vary somewhat across droughts, given that every drought is different.**
 
@@
 
+\newpage
 ## Citations

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
"leaf hydraulic traits" --> "leaf drought tolerance traits"
I think I got everything. Closes #126
finished first pass at reviewer responses!!!!

@@
 ### Supplementary Information
-## Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest
+## Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest
 
@@
 ![**Figure S5. Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario**](tables_figures/publication/figureS5_Rt_arima_comparison.png)
+
+\newpage
+
+![**Figure S6. Visualization of best model, with data, for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
finished first pass at reviewer responses!!!!

@@
 
-- Individual-level drought resistance decreased with tree height, which was strongly correlated with *crown exposure*.
+- Individual-level drought resistance decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
@@
 Models were run for all drought years combined and for each drought year individually. 
-The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below.
-In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
+The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
+In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
+Individual year models included a random effect of species.  
+All models included fixed effects of independent variables of interest (Tables 1,3) as specified below. 
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 
-Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Figs. **4, S6**). 
 Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 
@@
 
-$Rt$ had a significantly negative response to $ln[TWI]$ in all drought years combined and in 1977 and 1999 individually (Fig. **4**, Table S6). 
-When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in all drought years combined and in 1977, and a negative effect of $ln[TWI]$ was included in some of the models in 1966 and 1999 (Table S7).
+$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined and in 1977 and 1999 individually (Figs. **4, S6**, Table S6-S7). 
+The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S6).
+When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in 1977, and included in some of the top models in 1966 and 1999 (Table S7).
 This negates the idea that trees in moist microsites would be less affected by drought. 
@@
 
-Species traits... (Table 2, Fig. **S4**)
-Responses varied across species and by drought (Fig. **3**). 
+Species traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$) varied significantly among species (all p<0.05 in ANOVA; Table 2, Fig. **S4**).
+Drought resistance also varied across species and by drought (Fig. **3**). 
 Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
@@
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7). 
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Figs. **4, S6**; Tables 1,S4-S7). 
 Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
@@
 **1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S6).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Fig.
+**S6**; Table S6).
 

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
misc cleanup

@@
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-Because the resistance and resilience of growth to drought is linked to their probability of surviving drought (*[DeSoto et al.
+Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought (*[DeSoto et al.
 (2020)](https://www.nature.com/articles/s41467-020-14300-5)*; *Liu et al.
@@
 Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
-We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
+We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
 First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
@@
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified the three strongest droughts during the study period (Figs. **1**, **S1**, Table S3).
+Based on this, we identified the three strongest droughts during the study period (Figs. **1, S1**; Table S3).
 
@@
 In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. 
-Annual tree mortality censuses were initiated in 2014 [@gonzalezakre_patterns_2016], and in 2016-2017, cores were collected from all trees found to have died since the previous year’s. 
+Annual tree mortality censuses were initiated in 2014 [@gonzalezakre_patterns_2016], and in 2016-2017, cores were collected from all trees found to have died since the previous year’s census. 
 We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
@@
 We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018].
-After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.*
+After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.
 
@@
 **2d**).
-Despite this, a comparison test between height and crown position data from the most recent ForestGEO census (2018) revealed a correlation of 0.73.
-As early models showed $ln[H]$ had a much stronger effect than crown position, we dropped crown position from later analysis.
+A comparison test between height and crown position data from the most recent ForestGEO census (2018) revealed a correlation of 0.73.
 
@@
 In the context of the multivariate model, $Rt$ did not vary across drought years.
-That is, "drought year" as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model.
+That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model.
 
@@
 
-$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined and in 1977 and 1999 individually (Figs. **4, S6**, Table S6-S7). 
+$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Figs. **4, S6**, Table S6-S7). 
 The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S6).
@@
 This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
-This term did appear, however, with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7).
+This term did appear with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7), indicating that the negative effect of height on  $Rt$ was significantly stronger in wetter microhabitats.
 
@@
 
-The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig.
-**1a**, @helcoski_growing_2019) and had substantial but not devastating impacts on tree growth (Figs.
+The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig.
+**1a**, @helcoski_growing_2019] and had substantial but not devastating impacts on tree growth (Figs.
 **1b**).
 These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019].
-However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
+However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019] have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
 Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
@@
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
-Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees fo different size.
+Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms across trees of different size.
 

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
added footnote on fixed effect of year to tables S6 ans S7
closes #123

@@
 ```
-Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). $R^2$ refers to conditional $R^2$.
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
+$R^2$ refers to conditional $R^2$. 
+Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small (1966: 0, 1977: -0.019, 1999: -0.005; same values in all top models).
 
@@
 ```
-Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). $R^2$ refers to conditional $R^2$.
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
+$R^2$ refers to conditional $R^2$. 
+Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
 

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
added footnote on fixed effect of year to tables S6 ans S7
closes #123

@@
 All models included fixed effects of independent variables of interest (Tables 1,3) as specified below.
+All variables had variance inflation factors <1.5 (1 +/- 0.2). 
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 
-Species traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$) varied significantly among species (all p<0.05 in ANOVA; Table 2, Fig. **S4**).
-Drought resistance also varied across species and by drought (Fig. **3**). 
+Species, as a factor in ANOVA, had significant influence (p<0.05) on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
+**S4**).
+Drought resistance also varied across species, overall and in each drought year (Fig. **3**). 
 Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
references
#128

@@
 Many studies have shown that within and across species, large trees tend to be more affected by drought. 
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; **REF**].
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016]. 
 It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
@@
 Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (**REF**).
+Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (*Phillips et al. 2003; DOI:10.1093/treephys/23.4.237; Scholz et al. 2011*).
 Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al.

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
changed order of tables S1-S3
#104

@@
 \newpage
-```{r Table S1, eval=TRUE, echo=FALSE, message=FALSE}
+```{r, Table S1, eval=TRUE, echo=FALSE, message=FALSE}
 library(knitr)
 library(kableExtra)
-
-bark <- read.csv("tables_figures/publication/tableS1_bark_regression.csv", stringsAsFactors = FALSE)
-bark[,2] <- NULL
-colnames(bark) <- c("Species", "Equations", "$R^2$")
-kable(bark, booktabs=TRUE, format = "latex", caption = "Table S1: Species-specific bark thickness regression equations") %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
-```
-
-\newpage
-```{r Table S2, eval=TRUE, echo=FALSE,  warning=FALSE}
-library(knitr)
-library(kableExtra)
-
-heights <- read.csv("tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
-kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S2. Species-specific height regression equations", escape = FALSE) %>%
-  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
-```
-
-\newpage
-```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
-tableS3 <- read.csv("tables_figures/publication/tables3_drought_stats.csv", stringsAsFactors = FALSE)
-kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3.
+tableS1 <- read.csv("tables_figures/publication/tableS1_drought_stats.csv", stringsAsFactors = FALSE)
+kable(tableS1, booktabs=TRUE, format = "latex", caption = "Table S1.
 Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
@@
   pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)
+```
+
+\newpage
+```{r Table S2, eval=TRUE, echo=FALSE, message=FALSE}
+library(knitr)
+library(kableExtra)
+
+bark <- read.csv("tables_figures/publication/tableS2_bark_regression.csv", stringsAsFactors = FALSE)
+bark[,2] <- NULL
+colnames(bark) <- c("Species", "Equations", "$R^2$")
+kable(bark, booktabs=TRUE, format = "latex", caption = "Table S2. Species-specific bark thickness regression equations") %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
+```
+
+\newpage
+```{r Table S3, eval=TRUE, echo=FALSE,  warning=FALSE}
+library(knitr)
+library(kableExtra)
+
+heights <- read.csv("tables_figures/publication/tableS3_height_regression.csv", stringsAsFactors = FALSE)
+kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S3. Species-specific height regression equations", escape = FALSE) %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
changed order of tables S1-S3
#104

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  5,365	 |	No. of figures | 	2 (both colour)
-Summary|  	198 | 	No. of Tables | 	5  
-Introduction|	1,034 |	No of Supporting Information files | 	6
+Total word count (excluding summary, references and legends) |  5,365	 |	No. of figures | 	4 (all colour)
+Summary|  	198 | 	No. of Tables | 	3  
+Introduction|	1,034 |	No of Supporting Information files | 	#
 Materials and Methods |	1,945	||	 
@@
 We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with anomalously dry peak growing season climatic conditions. 
-Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S1), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified the three strongest droughts during the study period (Figs. **1, S1**; Table S3).
+Based on this, we identified the three strongest droughts during the study period (Figs. **1, S1**; Table S1).
 
-The droughts differed in intensity and antecedent moisture conditions (Fig. **S1**, Table S3). 
+The droughts differed in intensity and antecedent moisture conditions (Fig. **S1**, Table S1). 
 The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall. 
@@
 Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression on log-transformed data to relate $r_{bark}$ to diameter inside bark from 2008 data (Table S1), which were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
+Specifically, we used linear regression on log-transformed data to relate $r_{bark}$ to diameter inside bark from 2008 data (Table S2), which were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
 
@@
 Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
-Species-specific height allometries were developed (Table S2) using log-log regression ($ln[H] \sim ln[DBH]$). 
+Species-specific height allometries were developed (Table S3) using log-log regression ($ln[H] \sim ln[DBH]$). 
 For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species.

### 2020-07-22 manuscript/climate_sensitivity_draft.Rmd
Added language for describing Fig. 3 #127

@@
 
-```{r include=FALSE}
+```{r include=FALSE, eval=FALSE}
 # automatically create a bib database for R packages
@@
 knitr::write_bib(c(
-  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea', 'forecast'
+  .packages(), 'knitr', 'rmarkdown', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'raster', 'elevatr', 'dynatopmodel', 'stringi', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'purrr', 'pointRes', 'dplR', 'data.table', 'tools', 'dplyr', 'reshape2', 'png', 'grid', 'gridExtra', 'cowplot', 'ggpubr', 'extrafont', 'rasterVis', 'tidyr', 'neonUtilities', 'lubridate', 'LeafArea', 'forecast', 'visreg', 'agricolae'
 ), file = 'packages.bib')
@@
 All models included fixed effects of independent variables of interest (Tables 1,3) as specified below.
-All variables had variance inflation factors <1.5 (1 +/- 0.2). 
+All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
 We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
@@
 **S4**).
-Drought resistance also varied across species, overall and in each drought year (Fig. **3**). 
+Drought resistance also varied across species, overall and in each drought year (Fig.
+**3**).
+Significant differences in $Rt$ across species were most pronounced in 1966 with a total of seven distinct groupings, while 1977 had four and 1999 had two.
 Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).

### 2020-07-22 manuscript/climate_sensitivity_SI.Rmd
Fixed placement of package citations #129

@@
     fig_caption: yes       
+    includes:  
+      in_header: preamble-latex.tex
 bibliography: ["packages.bib"]
@@
 
-### Supplementary Information
+# Supplementary Information
 ## Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest
@@
 \listoffigures
-
 \newpage
-While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes.
-As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others.
-While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.
-
-R-packages not already cited in the main manuscript include the following, listed alphabetically by corresponding package name:
-
-[@R-base; @R-broom; @R-car; @R-cowplot; @R-data.table; @R-devtools; @R-dplR; @R-dplyr; @R-extrafont; @R-ggplot2; @R-ggpubr; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lubridate; @R-MuMIn; @R-piecewiseSEM; @R-png; @R-purrr; @R-raster; @R-rasterVis; @R-RCurl; @R-readxl; @R-reshape2; @R-rgdal; @R-rgeos; @R-rmarkdown; @R-sf; @R-stringi; @R-stringr; @R-tidyr]
 
@@
 ![**Figure S6. Visualization of best model, with data, for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)
+
+\newpage
+## Further Package Citations
+While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes.
+As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others.
+While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.
+
+R-packages not already cited in the main manuscript include the following, listed alphabetically by corresponding package name:
+
+[@R-base; @R-broom; @R-car; @R-cowplot; @R-data.table; @R-devtools; @R-dplR; @R-dplyr; @R-extrafont; @R-ggplot2; @R-ggpubr; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lubridate; @R-MuMIn; @R-piecewiseSEM; @R-png; @R-purrr; @R-raster; @R-rasterVis; @R-RCurl; @R-readxl; @R-reshape2; @R-rgdal; @R-rgeos; @R-rmarkdown; @R-sf; @R-stringi; @R-stringr; @R-tidyr]

### 2020-07-22 scripts/climate_sensitivity_analysis.R
Irrelevant fix

@@ -785,7 +785,7 @@ for (i in seq(along=sum_mod_traits[,c(8,11,14,17)])){
 cand_full <- cand_full[complete.cases(cand_full), ]
 
 #The info in this table is used to update table 4 (Rt) or S4 (arimaratio)
-write.csv(sum_mod_traits, "manuscript/tables_figures/tested_traits_all_lmer_arimaratio_CPout.csv", row.names=FALSE)
+write.csv(sum_mod_traits, "manuscript/tables_figures/tested_traits_all_lmer_CPout.csv", row.names=FALSE)
 write.csv(cand_full, "manuscript/tables_figures/candidate_traits_lmer_arimaratioCPout.csv", row.names=FALSE)
 
 ##3b reform. determine the best full model (expand for fuller explanation) ####

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
Updated references per #128

@@
 bibliography: ["book.bib",  "packages.bib"]
+link-citations: yes
 biblio-style: "apalike"
@@
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
-Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought (*[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)*; *Liu et al.
-2019; DOI: 10.1038/s41558-019-0583-9*), understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
+Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
 However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.
@@
 Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (*Phillips et al. 2003; DOI:10.1093/treephys/23.4.237; Scholz et al. 2011*).
-Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (*Zak et al.
-2010, Olson et al.
-2014,2018*).
+Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].
+Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
@@
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. 
-Counteracting the liabilities associated with tall height, large trees tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
+Counteracting the liabilities associated with tall height, large trees tend to have larger root systems [@enquist_global_2002], potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
 Larger root systems--if they grant access to deeper water sources--would be particularly advantageous in drier microenvironments (e.g., hilltops, as compared to valleys and streambeds) during drought.
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
-2007, Olson et al.
-2020*) and does not resolve differences among the many species within each category.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
@@
 In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
-For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests (*[Anderegg et al.
-2016](www.pnas.org/cgi/doi/10.1073/pnas.1525678113)*).
+For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
-These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.
-2018*].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al.
-2018*; but see *Farrell et al.
-2017*].
+These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
@@
 Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
-In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
+In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
 
@@
 Other R-packages used in analyses are listed in the Supplementary Information (*Appendix S1*).
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively.
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
 
@@
 There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought ([*Stovall et al.
-2020*](https://doi.org/10.1038/s41467-020-17213-5)).
+The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020].
 Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
@@
 They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
-Unlike other temperate forests where modestly cooler understory conditions have been documented (*Zellweger et al.
-2019*), particularly under drier conditions (*Davis et al.
-2019*), we observed no significant variation in air temperatures across the vertical profile (Fig.
+Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig.
 **2c**).
-More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman; Rey-Sanchez et al.
-2016*).
+More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016].
 Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
@@
 
-Belowground, taller trees would tend to have larger root systems (*Enquist and Niklas 2001; DOI: 10.1126/science.1066360*), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (*Meinzer et al. 1999; DOI: 10.1007/s004420050931*).
+Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
+Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
-This may be because moister habitats would tend to support species and individuals with more mesophytic traits (*Bartlett et al.
-2016 Drought.
-DOI: 10.1890/15-0468.1*; *Mencuccini 2003, DOI: 10.1046/j.1365-3040.2003.00991.x*) [@medeiros_extensive_2019], potentially growing to greater heights (e.g., *Detto et al.
-DOI: 10.1371/journal.pone.0076296*), and these are then more vulnerable when drought hits.
+This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 First, other traits can and do override the influence of xylem porosity on drought resistance. 
-Ring-porous species are restricted mainly to temperate deciduous forests (*Wheeler et al. 2007*), while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
+Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
 *Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
@@
 **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
-Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; *Zhu et al.
-2018*], and our findings indicate that these traits also influence drought responses.
+Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; @zhu_leaf_2018], and our findings indicate that these traits also influence drought responses.
 Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type.
@@
 The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts ([Christoffersen et al.
-2016](https://pearl.plymouth.ac.uk/handle/10026.1/12918)) [@powell_differences_2017].
+Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@christoffersen_linking_2016; @powell_differences_2017].
 

### 2020-07-23 manuscript/climate_sensitivity_SI.Rmd
Rewrote SI list at end of main text

@@
 kable(singlevar_table, booktabs = TRUE, caption = "Table S5.
-Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
-", format = 'latex', escape = FALSE) %>%
+Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
-kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable. ", format = 'latex', escape = FALSE) %>%
+kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.
-Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
-", format = 'latex', escape = FALSE) %>%
+Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
Rewrote SI list at end of main text

@@
 
-*redo this list!!*
+Table S1: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
 
-Table S1: Species-specific bark thickness regression equations
+Table S2: Species-specific bark thickness regression equations.
 
-Table S2: Species-specific height regression equations
+Table S3: Species-specific height regression equations.
 
-Table S3: Palmer drought severity index (PDSI) by month for focal droughts
+Table S4. Individual tests of species traits as drivers of drought resistance, where $Rt$ is used as the response variable.
 
-Figure S1: Map of ForestGEO plot showing TWI and location of cored trees
+Table S5. Individual tests of species traits as drivers of drought resistance, where $Rt_{ARIMA}$ is used as the response variable.
 
-Figure S2: Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to
-each focal drought
+Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.
 
-Figure S3: Height (from reconstructed DBH) by crown position across the three focal droughts and in the year of
-measurement (2018)
+Table S7. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
+
+Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought
+
+Figure S2: Map of ForestGEO plot showing topographic wetness index and location of cored trees
+
+Figure S3: Distribution of reconstructed tree heights across drought years.
+
+Figure S4. Distribution of independent variables by species.
+
+Figure S5. Comparison of $Rt$ and $Rt_{ARIMA}$ results, with residuals, for each drought scenario
+
+Figure S6. Visualization of best model, with data, for all droughts combined.

### 2020-07-23 scripts/climate_sensitivity_analysis.R
Updated file names in script per #98

@@ -287,7 +287,7 @@ ggplot(data = bark, aes(x = log(diam_nobark_2008.mm), y = log(bark.depth.mm))) +
   theme_minimal()
 
 #no total regression equation at bottom because all species are accounted for in dataset.
-##these equations are also found in tableS1_bark_regression.csv
+##these equations are also found in tableS2_bark_regression.csv
 bark$predict_barkthick.ln.mm <- NA
 bark$predict_barkthick.ln.mm <- 
   ifelse(bark$species == "caco", -1.56+0.416*log(bark$diam_nobark_2008.mm),
@@ -432,7 +432,7 @@ trees_all$dbh.ln.cm <- log(trees_all$dbh_old.cm)
 ##2e. add in tree heights ####
 ## taken from the canopy_heights script
 #the full equation is using all points for which we have data to create the equation, despite that for several species we don't have enough data to get a sp-specific equation
-height_regr <- read.csv("manuscript/tables_figures/publication/tableS2_height_regression.csv", stringsAsFactors = FALSE)
+height_regr <- read.csv("manuscript/tables_figures/publication/tableS3_height_regression.csv", stringsAsFactors = FALSE)
 
 height_regr$Equations <- gsub("^.*= ", "", height_regr$Equations)
 height_regr$Equations <- gsub("[[:alpha:]].*$", "x", height_regr$Equations)
@@ -583,7 +583,6 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 
 #########################################################################################
 #3. mixed effects model for output of #2.
-
 ##start here if just re-running model runs ####
 trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub.csv", stringsAsFactors = FALSE); arima_vals=FALSE
 # trees_all_sub <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE); arima_vals=TRUE
@@ -784,7 +783,8 @@ for (i in seq(along=sum_mod_traits[,c(8,11,14,17)])){
 
 cand_full <- cand_full[complete.cases(cand_full), ]
 
-#The info in this table is used to update table 4 (Rt) or S4 (arimaratio)
+#The info in sum_mod_traits is used to update 
[…]

### 2020-07-23 scripts/graphs_for_manuscript.R

@@ -615,7 +615,7 @@ ggarrange(NEON_list$wind_plot, NEON_list$RH_plot, NEON_list$SAAT_plot, heights_b
 dev.off()
 
 #########################################################################
-#3. Figure S1: Export map of plot using TWI and cored tree locations
+#3. Figure S2: Export map of plot using TWI and cored tree locations
 ## necessary packages ####
 library(raster)
 library(rasterVis)
@@ -660,7 +660,7 @@ t <- t +
                 default.units='native')
    })
 
-png("manuscript/tables_figures/publication/figureS1_location_cored_trees.png", width=5, height=7, units="in", res=300)
+png("manuscript/tables_figures/publication/figureS2_location_cored_trees.png", width=5, height=7, units="in", res=300)
 
 t + #add north arrow
    layer({
@@ -715,7 +715,7 @@ heights_allplot$tree <- as.character(heights_allplot$tree)
 heights_allplot$position_all_abb <- as.character(heights_allplot$position_all_abb)
 
 ########################################################################
-#5. Visualizing regression output ####
+#5. Figure 4. Visualizing regression output ####
 library(lme4)
 library(ggpubr)
 # remotes::install_github("pbreheny/visreg")
@@ -878,7 +878,7 @@ grid.arrange(arr, legend, ncol=2, widths=c(4,1))
 dev.off()
 
 
-## only plot the all-years model with visreg ####
+## Figure S6. only plot the all-years model with visreg ####
 xl <- c("ln[H]", "ln[TWI]", "PLA", "TLP")
 vars <- c("height.ln.m", "TWI.ln", "PLA_dry_percent", "mean_TLP_Mpa")
 
@@ -1125,6 +1125,63 @@ hist(exp(x1966$height.ln.m))
 
 
 ########################################################################
+#7. compare Rt values with arima_ratio ####
+library(data.table)
+arima_ratio <- read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv", stringsAsFactors = FALSE)
+
+rt <- read.csv("manuscript/tables_figures/trees_all_sub.csv", stringsAsFactors = FALSE)
+
+years <- c(1966, 1977, 1999)
+
+layout(matrix(1:8, nrow=2, byrow=TRUE))
+res_full <- NULL
+for(i in 1:4){
+   cols <- c("tree", "resist.value")
+   
+   if(i==1){
+      cols <- c("year", cols)
+      compare <- rt[,cols]
[…]

### 2020-07-23 manuscript/climate_sensitivity_SI.Rmd
Updated figure and table captions / fixed colnames #117

@@
 colnames(bark) <- c("Species", "Equations", "$R^2$")
-kable(bark, booktabs=TRUE, format = "latex", caption = "Table S2. Species-specific bark thickness regression equations") %>%
+kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific bark thickness regression equations") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
@@
 heights <- read.csv("tables_figures/publication/tableS3_height_regression.csv", stringsAsFactors = FALSE)
+colnames(heights)[3] <- "$R^2$"
 kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S3. Species-specific height regression equations", escape = FALSE) %>%
@@
 singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+colnames(singlevar_table)[1] <- "variable"
 kable(singlevar_table, booktabs = TRUE, caption = "Table S4.
@@
 singlevar_table <- read.csv("tables_figures/publication/tableS5_singlevariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(singlevar_table)[1] <- "variable"
 kable(singlevar_table, booktabs = TRUE, caption = "Table S5.
@@
 
+*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
+
 **$\Delta$AICc > 2: variable considered significant as an individual predictor
@@
 multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(multivar_table)[1] <- "drought"
 kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
@@
 multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(multivar_table)[1] <- "drought"
 kable(multivar_table, booktabs = TRUE, caption = "Table S7.
@@
 \newpage
-![**Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees**](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}
+![**Figure S2.
+Map of ForestGEO plot showing topographic wetness index and location of cored trees.** Scale units are in meters](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}
 
@@
 
-![**Figure S6. Visualization of best model, with data, for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)
+![**Figure S6.
+Visualization of the best model ($\Delta$AICc=0) with data for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)
 

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
Gillerot citation

@@
 Many studies have shown that within and across species, large trees tend to be more affected by drought. 
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016]. 
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020].
 It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
more on rooting depth

@@
 Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
+While tree size can correlate with the depth of water extraction [@brum_2018_hydrological], the linkage is not consistent.
+Shorter trees can vary broadly in the depth of water uptake [@stahl_2013_depth], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

### 2020-07-23 manuscript/response_to_reviews.Rmd
more rooting depth refs

@@
 
-**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems (Enquist and Niklas 2001), but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms (Meinzer et al.
-1999)."**
+**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
+While tree size can correlate with the depth of water extraction [@brum_2018_hydrological], the linkage is not consistent.
+Shorter trees can vary broadly in the depth of water uptake [@stahl_2013_depth], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
+Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018]."**
 

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
more rooting depth refs

--- 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
@@
 Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-Rather, larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
+While tree size can correlate with the depth of water extraction [@brum_hydrological_2018], the linkage is not consistent.
+Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
+Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018].
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
@@
 First, other traits can and do override the influence of xylem porosity on drought resistance. 
-Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (*REFS*). 
+Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (**REFS**). 
 *Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 

### 2020-07-23 manuscript/climate_sensitivity_draft.tex
more rooting depth refs

--- 2019-08-14 manuscript/climate_sensitivity_draft.tex
@@
-\documentclass[]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
 \usepackage{lmodern}
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Tree size and hydraulic traits shape drought responses in a temperate broadleaf forest},
-            pdfauthor={Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack, Krista Anderson-Teixeira},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{natbib}
-\bibliographystyle{apalike}
+\usepackage{longtable,booktabs}
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx,grffile}
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-
-  \title{Tree size and hydraulic traits shape drought responses in a temperate
-broadleaf forest}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{Ian McGregor, Ryan Helcoski, Norbert Kunert, Alan Tepley, Valentine
-Herrmann, Joseph Zailaa, Atticus Stovall, Neil Pederson, Lawren Sack,
-Krista Anderson-Teixeira}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
-    \date{}
-    \predate{}\postdate{}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{setspace}\onehalfspacing
+\usepackage{lineno}
+\linenumbers
 \usepackage{float}
@@
 \captionsetup[table]{labelformat=empty}
+\captionsetup[figure]{labelformat=empty}
 \usepackage{dcolumn}
+\usepackage[]{natbib}
+\bibliographystyle{apalike}
+
+\author{}
+\date{\vspace{-2.5em}}
 
 \begin{document}
-\maketitle
 
-\subsubsection{Summary}\label{summary}
+\raggedright
 
-Predicting forest responses to drought is an increasingly critical task
-under climate change effects. Part of the problem is due to the lack of
-studies analyzing the confluence of leaf hydraulic traits with
-biophysical parameters. In this study, we analyze the interaction
-between these two trait groups using forest census data from a 25.6-ha
-ForestGEO plot in Virginia (USA). Drought periods were defined by both
-Palmer Drought Severity Indices (PDSI) and their identification from
-tree-ring records for 12 species representing 97\% of woody
-productivity. Each drought scenario (1964-66, 1977, 1999), along with
-the overall trend, was then tested against leaf hyraulic trait
-measurements and microhabitat biophysical data. Individual-level growth
-responses to the three individual droughts were stronger among taller
-trees in dominant canopy positions, those in wetter microsites, and for
-more drought-sensitive species as assessed by leaf traits (turgor loss
-at less negative leaf water potential, greater shrinkage with leaf
-dehydration), with substantial variation in the best predictor variables
-across given droughts. We conclude that when droughts occur, large
-dominant trees, drought-sensitive species, and individuals in wetter
-microhabitats are likely to be most strongly affected.
+\textbf{Title:} Tree height and leaf drought tolerance traits shape
+growth responses across droughts in a temperate broadleaf forest
 
-\emph{The Summary for research papers, which must be usable as a stand-
-alone document, must not exceed 200 words and should be organized using
-four bullet points to indicate: (1) the research conducted, including
-the rationale, (2) methods, (3) key results, and (4) the main
-conclusion, including the key points of discussion. It should not
-contain citations of other papers.}
+\textbf{Authors:} Ian R. McGregor\textsuperscript{1,2}, Ryan
+Helcoski\textsuperscript{1}, Norbert Kunert\textsuperscript{1,3}, Alan
+J. Tepley\textsuperscript{1,4}, Erika B.
+Gonzalez-Akre\textsuperscript{1}, Valentine Herrmann\textsuperscript{1},
+Joseph Zailaa\textsuperscript{1,5}, Atticus E.L.
+Stovall\textsuperscript{1,6,7}, Norman A. Bourg\textsuperscript{1},
+William J. McShea\textsuperscript{1}, Neil Pederson\textsuperscript{8},
+Lawren Sack\textsuperscript{9,10}, Kristina J.
+Anderson-Teixeira\textsuperscript{1,3}*
 
-\subsubsection{Introduction}\label{introduction}
+\textbf{Author Affiliations:}
 
-Understanding how and why trees respond to drought is critical to
-predicting forest drought responses and climate change feedbacks. (1
-paragraph on this- I have lots of content on this and will add later.)
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Conservation Ecology Center; Smithsonian Conservation Biology
+  Institute; National Zoological Park, Front Royal, VA 22630, USA
+\item
+  Center for Geospatial Analytics; North Carolina State University;
+  Raleigh, NC 27607, USA
+\item
+  Center for Tropical Forest Science-Forest Global Earth Observatory;
+  Smithsonian Tropical Research Institute; Panama, Republic of Panama
+\item
+  Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta,
+  Canada
+\item
+  Biological Sciences Department; California State University; Los
+  Angeles, CA 90032, USA
+\item
+  Department of Environmental Sciences, University of Virginia,
+  Charlottesville, VA 22903, USA
+\item
+  NASA Goddard Space Flight Center; Greenbelt, MD 20771, USA
+\item
+  Harvard Forest, Petersham, MA 01366, USA
+\item
+  Department of Ecology and Evolutionary Biology; University of
+  California, Los Angeles; Los Angeles, CA 90095, USA
+\item
+  Institute of the Environment and Sustainability; University of
+  California, Los Angeles; Los Angeles, CA 90095, USA
+\end{enumerate}
 
-{[}Understanding forest responses to drought requires increased
-functional understanding of the factors that confer individual-level
-vulnerability or resistance.{]} Forests are diverse in terms of tree
-sizes and functional traits, and it is known that trees varying in size
-and functional traits respond differently to drought (e.g.,
-\citep{bennett_larger_2015}; REFS). Therefore, in order to understand
-whole-forest response to drought, we need a functional understanding of
-how responses vary by tree size, microhabitat, and species. There are 3
-fundamental questions that must be addressed:
+*corresponding author:
+\href{mailto:teixeirak@si.edu}{\nolinkurl{teixeirak@si.edu}}; +1 540 635
+6546
 
-\emph{First, what drives the observed tendency for large trees to suffer
-more during drought?}\\
-\citet{bennett_larger_2015} showed that in forests globally, large trees
-suffer greater growth reductions during drought. However, this analysis
-quantified tree size based on DBH, which has no direct mechanistic
-meaning. This study proposed two major mechanisms (besides insects) for
-the observed greater drougth growth reductions of large trees: (1)
-inherently greater biophysical challenge of being tall; (2) greater
-exposure of the crowns of large trees. Counteracting these effects, (3)
-the larger root systems of larger trees may confer an advantage in terms
-of allowing greater access to water, but it appears that this effect is
-usually insufficient to offset the costs of height and/or crown
-exposure.
+\begin{longtable}[]{@{}llll@{}}
+\toprule
+\begin{minipage}[b]{0.35\columnwidth}\raggedright
+Text\strut
+\end{minipage} & \begin{minipage}[b]{0.16\columnwidth}\raggedright
+word count\strut
+\end{minipage} & \begin{minipage}[b]{0.22\columnwidth}\raggedright
+other\strut
+\end{minipage} & \begin{minipage}[b]{0.15\columnwidth}\raggedright
+n\strut
+\end{minipage}\tabularnewline
+\midrule
+\endhead
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Total word count (excluding summary, references and legends)\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+5,365\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No.~of figures\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+4 (all colour)\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Summary\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+198\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No.~of Tables\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+3\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Introduction\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+1,034\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+No of Supporting Information files\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\#\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Materials and Methods\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+1,945\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Results\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+697\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Discussion\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+1467\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\begin{minipage}[t]{0.35\columnwidth}\raggedright
+Acknowledgements\strut
+\end{minipage} & \begin{minipage}[t]{0.16\columnwidth}\raggedright
+125\strut
+\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
+\strut
+\end{minipage} & \begin{minipage}[t]{0.15\columnwidth}\raggedright
+\strut
+\end{minipage}\tabularnewline
+\bottomrule
+\end{longtable}
 
-Canopy trees have lower drought resistance because they are exposed to
-higher solar radiation, greater wind speeds, and lower humidity.
-Alternatively, the generally supressed status of subcanopy trees may be
-insufficient to override the benefits of their buffered environment
-during drought.
+\newpage
 
-\emph{Second, how do species' traits - alone and in interaction with
-tree size - influence drought response?} Analyzing drought responses on
-the species level does not fully explain mechanisms and is not feasible
-in diverse forests. The solution is a trait-based approach. Leaf
-hydraulic traits hold more promise than more
-commonly/traditionally-measured traits such as wood density and specific
-leaf area (SLA)
-(\href{https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/1365-2435.13229}{Medeiros
-et al. 2019}).
-
-Commonly measured traits including wood density (REFS), leaf mass per
-area \citep{abrams_adaptations_1990} \citep{guerfel_impacts_2009}, and
-ring porosity (Elliot et al. 2015, Friedrichs et al. 2009) have been
-linked to drought responses and likely correlated with drought
-resistance in this forest. However, we hypothesize that leaf hydraulic
-traits such as leaf area shrinkage upon dessication (PLA) and turgor
-loss point (TLP), which are emerging as potentially more informative
-traits but whose effect on drought resistance has never been tested
-(CONFIRM), will prove better predictors.
-
-Moreover, there may be an interaction between traits and tree size. It
-is possible that the pattern observed by \citet{bennett_larger_2015}
-could be caused by smaller trees being more drought resistant.
-Alternatively, larger trees may have more drought-resistant traits as
-adpatations to greater biophysical challenges.
-
-\emph{Third, are responses similar or variable across individual drought
-years?} Droughts are rarely explicitly defined in ecological studies
-\citep{slette_how_2019}, yet no two droughts are the same.
-
-We need to understand the the factors confirming drought vulnerability
-or resistance not only for extreme droughts with dramatic impacts on
-tree growth and mortality, which tend to dominate the literature
-\citep{bennett_larger_2015} (REFS), but also for more modest but
-frequent droughts--e.g., those with historical return intervals on the
-order of a decade.
-
-Here, we combine tree-ring records covering three droughts (1964-66,
-1977, 1999), species functional and hydraulic trait measurments, and
-forest census data from a 25.6-ha ForestGEO plot in Virginia (USA) to
-test a series of hypotheses and associated specific predictions (Table
-1) designed to yield functional understanding of how tree size,
-microenrivonment, and species' traits collectively shape drought
-responses. First, we focus on the role of tree size and its interaction
-with microenvironement. We confirm that, consistent with most forests
-globally \citep{bennett_larger_2015}, larger-diameter trees have lower
-drought resistance in this forest, which is in an ecoregion represented
-by only one study in Bennett et al (2015) (H1.0). We then test
-hypotheses designed to disentangle the relative importance of tree
-height (H1.1), crown exposure (H1.2), and root water access, which
-should be greater for larger trees in dry but not perpetually wet
-microsites (H1.3). Second, we focus on the role of species' functional
-and hydraulic traits and their interaction with tree height. We
-hypothesize that drought resistance will follow predicted and observed
-patterns in relation to wood density, specific leaf area, and ring
-porosity (P2.1a-c), but that leaf hydraulic traits such as leaf area
-shrinkage upon dehydration and turgor loss point will prove better
-predictors (P2.1d-e). We then test whether these traits correlate with
-tree height (P2.2), potentially driving the observed tendency for taller
-trees to suffer more during drought (P2.3). Finally, we focused on
-variability among droughts, asking how community resistance varied
-across droughts (H3.1) and whether the factors confirming vulnerability
-or resistance varied across droughts (H3.2).
-
-\emph{Let's put Table 1 here. We need to give the hypotheses/
-predictions here. It seems counterintuitive to put the results right up
-front like that, but it actually tends to be a higher-impact style of
-writing, and recommended
-\href{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005619}{here}.}
-
-\textbf{Table 1. Summary of hypotheses, predictions and results}
-\includegraphics[width=5.20833in]{tables_figures/Table1.png}
-
-We count predictions as fully supported (or rejected) when the direction
-of response matches (or contradicts) the prediction. H2.3-H3.2 are based
-on if models containing species traits in H2.1 had dAIC\textgreater{}=2
-relative to the appropriate null model or to any alternative
-multivariate model within 2 dAIC. Parentheses indicate that predictions
-were partially supported (or rejected). In other words, parentheses
-indicate when the direction of response matched (or contradicted) the
-prediction in some but not all models with dAIC\textless{}2 relative to
-the appropriate null model. With categorical variables such as crown
-position, a ``(yes)'' notation describes when the trend matched the
-prediction, but it wasn't significant A ``(yes/no) or (''no/yes``)''
-indicates tendencies in opposite directions in univariate tests versus
-the best full models, respectively. ``(yes)
-
-\subsubsection{Materials and Methods}\label{materials-and-methods}
-
-\emph{Study site} Research was conducted at the 25.6 ha ForestGEO
-(Global Earth Observatory) study plot at the Smithsonian Conservation
-Biology Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°
-08'43.4''W) \citep{andersonteixeira_ctfs-forestgeo:_2015}. SCBI is
-located in the central Appalachian Mountains at the northern edge of
-Shenandoah National Park. Elevations range from 273-338m above sea level
-\citep{gonzalezakre_patterns_2016} with a topographic relief of 65m
-\citep{bourg_initial_2013}. Dominant species include \emph{Liriodendron
-tulipifera}, oaks (\emph{Quercus} spp.), and hickories (\emph{Carya}
-spp.).
-
-\emph{Data collection and preparation} The SCBI ForestGEO plot was
-censused in 2008, 2013, and 2018 following standard ForestGEO protocols,
-whereby all free-standing woody stems \textgreater{}= 1cm diameter at
-breast height (DBH) were mapped, tagged, measured at DBH, and identified
-to species \citep{condit_tropical_1998}. From this census data, we used
-measurements of DBH from 2008 to calculate historical DBH (described
-below), along with tree location in the plot to determine the
-topographic wetness index. Furthermore, we used data for all stems
-\textgreater{}= 10cm to analyze functional trait composition relative to
-tree height. \emph{I added the last sentence-- not sure if you've
-described that below yet. I don't think you need to note here which data
-we use, so long as that's described in the portions where you describe
-the analysis.}
-
-We analyzed tree-ring data from the twelve species contributing most to
-woody aboveground net primary productivity (ANPP), which together
-comprised 97\% of study plot ANPP between 2008 and 2013
-\citep{helcoski_growing_2019}. Cores were obtained in 2010-2011 or
-2016-2017 from a breast height of 1.3m using a 5mm increment borer. In
-2010-2011, cores were collected from randomly selected live trees of
-species with at least 30 individuals of DBH \textgreater{}= 10cm
-\citep{bourg_initial_2013}. In 2016-2017, cores were collected from all
-trees found dead in the annual mortality census
-\citep{gonzalezakre_patterns_2016}. Cores were sanded, measured, and
-cross-dated using standard procedures, as detailed in
-\citep{helcoski_growing_2019}. The resulting chronologies have been
-published in association with \citet{helcoski_growing_2019}: (ITRDB;
-GitHub/Zenodo). \emph{is this a database? Yes. Ryan submitted the data
-but I don't think its posted yet. We should also cite GitHub/Zenodo
-here. I'll come back to that. }
-
-Height measurements (n=\# trees) were taken by several researchers
-between 2012 to 2019, and are archived in a public
-\href{https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/tree/master/tree_dimensions/tree_heights}{GitHub
-repository}. Measurement methods included manual
-\citep[NEON]{stovall_assessing_2018}, digital rangefinders
-\citep{andersonteixeira_size-related_2015}, and automatic LiDAR
-\citep{stovall_terrestrial_2018}. Rangefinders either used the tangent
-method (Impulse 200LR, TruPulse 360R) or the sine method (Nikon
-ForestryPro) for calculating heights. The associated errors for using
-either method were acknowledged \citep{larjavaara_measuring_2013}.
-Species-specific height allometries were developed (Table S\# -
-\textbf{ADD THIS TABLE TO SI}). For species that didn't have enough
-height measurements, heights were calculated from equations derived from
-all species in the study.
-
-For each tree, we combined tree-ring records and allometric equations of
-height and bark thickness to retroactively calculate DBH and estimate
-height for the years 1950-2009. Prior DBH was estimated using the
-following equation, using 2008 as the earliest year for having reliable
-DBH measurements:
-
-\[ diamYEAR = dbh2008 - 2*(bark.depth2008) - 2*\Sigma(ring.widthYEAR:ring.width2008) + 2*(bark.depthYEAR) \]
-
-Here, \emph{ring.width} was measured from cores. Bark thickness was
-estimated from species-specific allometries based on the bark thickness
-data of \citep{andersonteixeira_size-related_2015}. Specifically, we
-used linear regression equations on log-transformed data to relate bark
-thickness to DBH (Table S\#- \textbf{create table to give these
-equations in SI}) and then used these to estimate bark thickness based
-on DBH.
-
-Crown positions were recorded in the field during the growing season of
-2018 following the crown position protocol from
-\citep{jennings_assessing_1999}, whereby positions were ranked as
-dominant, codominant, intermediate, or suppressed. As there was no way
-to retroactively estimate crown position, we assumed that 2018 crown
-position was reflective of each tree's position over the past 60 years.
-While some trees undoubtedly changed position, an analysis of crown
-position relative to height (Fig. XX) and height change since
-\emph{1959} indicated that change was likely slow. {[}\textbf{work on
-this-- provide details?}{]}
-
-Topographic wetness index (TWI) was calculated using the
-\citep{R-dynatopmodel} package in R.
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=5.20833in]{tables_figures/Figure3.png}
-\caption{Location of cored trees}
-\end{figure}
-
-Hydraulic traits were collected from SCBI and are summarized in Table 1.
-In August 2018, we collected leaf samples from three individuals of each
-species \ldots{} (\textbf{Nobby's description of methods for the
-following (see word document}) 1. PLA 2. LMA 4. Wood density 5. TLP
-
-\textbf{Table 2. Species analyzed here, listed in descending order of
-ANPP\_stem. n cores and DBH range represented, and species traits}
-{[}*This replaces/combines the two remaining tables in this section.
-Suggested columns, with those to include only if they fit in
-parentheses: species, (stems \textgreater{}=10 cm per ha in plot),
-(ANPP\_stem), n cores, DBH range of cores, (n cores in each crown
-position) species means for each trait{]}
-
-\begin{verbatim}
-## Warning: package 'kableExtra' was built under R version 3.4.4
-\end{verbatim}
-
-\textbackslash{}begin\{table\}{[}!h{]}
-
-\textbackslash{}caption\{\label{tab:Table2}Overview of analyzed species,
-detailing DBH mean and range of cored trees, the number of cores
-represented by each crown position of each species, and mean hydraulic
-trait measurements. Units of measurements are in mm (DBH), \% (PLA),
-g/m2 (LMA), MPa (TLP), and g/cm3 (WD).\} \centering
-
-\begin{tabular}{lrrlrrrr}
-\toprule
-sp & mean\_DBH & range\_DBH & RP & PLA & LMA & TLP & WD\\
-\midrule
-caco & 271.87 & 508.0 & ring & 17.22 & 45.86 & -2.13 & 0.83\\
-cagl & 313.89 & 887.0 & ring & 21.09 & 42.76 & -2.13 & 0.62\\
-caovl & 352.87 & 511.0 & ring & 14.80 & 47.60 & -2.48 & 0.96\\
-cato & 209.74 & 201.1 & ring & 16.56 & 45.36 & -2.20 & 0.83\\
-fagr & 235.11 & 960.0 & diffuse & 9.45 & 30.68 & -2.57 & 0.62\\
-\addlinespace
-fram & 353.63 & 883.3 & ring & 13.06 & 43.28 & -2.10 & 0.56\\
-juni & 481.42 & 628.0 & semi-ring & 24.64 & 72.13 & -2.76 & 1.09\\
-litu & 368.54 & 904.0 & diffuse & 19.56 & 46.92 & -1.92 & 0.40\\
-qual & 471.51 & 677.0 & ring & 8.52 & 75.80 & -2.58 & 0.61\\
-qupr & 422.48 & 767.0 & ring & 11.75 & 71.77 & -2.36 & 0.61\\
-\addlinespace
-quru & 548.79 & 1369.3 & ring & 11.01 & 71.13 & -2.64 & 0.62\\
-quve & 541.38 & 981.8 & ring & 13.42 & 48.69 & -2.39 & 0.65\\
-\bottomrule
-\end{tabular}
-
-\textbackslash{}end\{table\}
-
-\emph{Climate and drought years} {[}\textbf{add description of climate
-data used in Fig. 1, NEON vertical profiles}{]}
-
-To accurately understand climate sensitivity, this study used a specific
-definition of drought, which is not a common practice
-\citep{slette_how_2019}. We used the pointRes package \citep{R-pointRes}
-in R (version 3.5.3) to determine drought periods based on trees'
-drought resistance, which is defined by \citep{lloret_components_2011}
-as the ratio between the performance during and before the disturbance.
-Candidate drought years were defined if \textgreater{}50\% of the cored
-trees experienced \textless{}30\% growth in a year compared to the
-previous 5 years. These were then cross-validated with the regional
-Palmer Drought Severity Index (PDSI) values for each year, which yielded
-a set of three periods that were consistently shown as drought:
-1964-1966, 1977, and 1999.
-
-\subsubsection{Results}\label{results}
-
-Results outline
-
-Descriptions of Droughts
+\hypertarget{summary}{%
+\subsubsection{Summary}\label{summary}}
 
 \begin{itemize}
-\tightlist
 \item
-  physical characteristics of drought: duration, min PDSI (and when
-  occurred), see slette et al 2019 to see how they suggest reporting
-  each one
+  As climate change is driving increased drought frequency and severity
+  in many forested regions around the world, mechanistic understanding
+  of the factors conferring drought resistance in trees is increasingly
+  important. The dendrochronological record provides a window through
+  which we can understand how tree size and species' traits shape tree
+  growth responses during droughts.
 \item
-  sum nb.series for each scenario, mult by perc.neg (pointers df),
-  define thresholds (e.g.~75\% of all trees experienced the growth
-  reduction of 40\% or more)
+  We analyzed tree-ring records for twelve species that comprise 97\% of
+  the woody productivity of the 25.6-ha ForestGEO plot in a broadleaf
+  deciduous forest of northern Virginia (USA) to test hypotheses on how
+  tree height, microenvironment characteristics, and species' traits
+  shaped drought responses across the three strongest regional droughts
+  over a 60-year period (1950 - 2009).
 \item
-  make table for this
+  Individual-level drought resistance decreased with tree height, which
+  was strongly correlated with exposure to higher evaporative demand and
+  solar radiation. The potentially greater rooting volume of larger
+  trees did not confer an advantage in sites with low topographic
+  wetness index. Resistance was greater among species whose leaves
+  experienced less shrinkage upon desiccation and lost turgor (wilted)
+  at more negative water potentials.
 \item
-  citing table 3 (the coefficient for individual year) - maybe mention
-  how 77 and 99 have neg coefficients, indicating potential height
-  effect: as time goes on, trees are more affected by drought anyways
-  because they grow taller
-\item
-  this is followed by Figure 2
+  We conclude that tree height and leaf drought tolerance traits
+  influence growth responses during drought, as recorded in the
+  tree-ring record spanning historical droughts. Thus, these factors can
+  be useful for predicting future drought responses under climate
+  change.
 \end{itemize}
 
-Hypothesis results - one header for each main question - present results
-for each one, walk through via order and logic of Table 1 - no citations
-here
+\emph{Key words}: annual growth; crown exposure; drought; Forest Global
+Earth Observatory (ForestGEO); leaf drought tolerance traits; temperate
+broadleaf deciduous forest; tree height; tree-ring
 
-Once the data was collected, linear mixed models were run following the
-order of the hypotheses as seen in Figure ???
-{[}individual\_tested\_traits{]}. Using the \citep{R-pointRes} package,
-we set up models with the resistance value as the response variable, and
-each prediction's variable as the independent variable. Variables'
-importance in predicting drought tolerance was calculated from
-mixed-effects models and the lowest AICc
-\citep[\citet{R-AICcmodavg}]{R-lme4}.Null models were determined in
-order of the predictions. First, we analyzed the combined scenario to
-determine if ``year'' was significant. Upon establishing this, we tested
-height and DBH as size parameters. Although both were significant,
-height was kept due to its larger delta AICc compared with the null
-model. We then tested the remaining biophysical and hydraulic traits
-individually against a null model containing height and year. This
-yielded Figure ??? (cand\_full). All variables with dAICc
-\textgreater{}2 were used as candidates for each scenario's best model
-(figure ???? (tested\_traits\_best))
+\newpage
 
-\textbf{Describe each individual drought scenario } 1964-1966 1977 1999
+\hypertarget{introduction}{%
+\subsubsection{Introduction}\label{introduction}}
 
-\textbf{Figure 1. Time series of peak growing season (May-August)
-climate conditions and residual chronologies for each species.} Droughts
-analyzed here are indicated by dashed lines, and shading indicates the
-pre-drought period used in calculations of the resistance metric. Figure
-modified from \citep{helcoski_growing_2019}.
+Forests play a critical global role in climate regulation
+\citep{bonan_forests_2008}, yet there remains enormous uncertainty as to
+how the forest-dominated terrestrial carbon sink will respond to climate
+change \citep{friedlingstein_climatecarbon_2006}. An important aspect of
+this uncertainty lies with physiological responses of trees to drought
+\citep{kennedy_implementing_2019}. In many forested regions around the
+world, the risk of severe drought is increasing
+\citep{trenberth_global_2014, dai_climate_2018}, often despite
+increasing precipitation
+\citep{intergovernmental_panel_on_climate_change_climate_2015, cook_unprecedented_2015}.
+Droughts, intensified by climate change, have been affecting forests
+worldwide and are expected to continue as one of the most important
+drivers of forest change in the future
+\citep{allen_global_2010, allen_underestimation_2015}. Understanding
+forest responses to drought requires elucidation of how tree size,
+microenvironment, and species' traits jointly influence individual-level
+drought resistance, and the extent to which their influence is
+consistent across droughts. Because the resistance and resilience of
+growth to drought is linked to trees' probability of surviving drought
+\citep{desoto_low_2020, liu_hydraulic_2019}, understanding growth
+responses can also help elucidate which trees are most vulnerable to
+drought-induced mortality. However, it has proven difficult to resolve
+the many factors affecting tree growth during drought with available
+forest census data, which only rarely captures extreme drought, and with
+tree-ring records, which capture multiple droughts but usually only
+sample a subset of a forest community, typically focusing on a single
+species or the largest individuals.
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=5.20833in]{tables_figures/Figure1.jpg}
-\caption{Time series of combined tree cores by species}
-\end{figure}
+Many studies have shown that within and across species, large trees tend
+to be more affected by drought. Greater growth reductions for larger
+trees were first shown on a global scale by \citet{bennett_larger_2015},
+and subsequent studies have reinforced this finding
+\citep[\emph{e.g.},][]{hacket-pain_consistent_2016, gillerot_tree_2020}.
+It has yet to be resolved which of several potential underlying
+mechanisms most strongly shape these trends in drought response. First,
+tree height itself may be a primary driver. Taller trees face the
+biophysical challenge of lifting water greater distances against the
+effects of gravity and friction
+\citep{mcdowell_relationships_2011, mcdowell_darcys_2015, ryan_hydraulic_2006, couvreur_water_2018}.
+Vertical gradients in stem and leaf traits--including smaller and
+thicker leaves (higher leaf mass per area, LMA), greater resistance to
+hydraulic dysfunction (\emph{i.e.}, more negative water potential at
+50\% loss of hydraulic conductivity, more negative P50), and lower
+hydraulic conductivity at greater heights
+\citep{couvreur_water_2018, koike_leaf_2001, mcdowell_relationships_2011}--enable
+trees to become tall \citep{couvreur_water_2018}. Greater stem
+capacitance (\emph{i.e.}, water storage capacity) of larger trees may
+also confer resistance to transient droughts
+{[}\citet{phillips_reliance_2003}*; \citet{scholz_hydraulic_2011}{]}.
+Indeed, tall trees require xylem of greater hydraulic efficiency, such
+that xylem conduit diameters are wider in the basal portions of taller
+trees, both within and across species
+\citep{olson_plant_2018, liu_hydraulic_2019}, and throughout the
+conductive systems of angiosperms
+\citep{zach_vessel_2010, olson_universal_2014, olson_plant_2018}. Wider
+xylem conduits plausibly make large trees more vulnerable to embolism
+during drought \citep{olson_plant_2018}, and traits conducive to
+efficient water transport may also lead to poor ability to recover from
+or re-route water around embolisms \citep{roskilly_conflicting_2019}.
 
-\textbf{Results for first main question: what drives the observed
-tendency for large trees to suffer more during drought?} H1.0, H1.1,
-H1.2, H1.3 DBH, height, crown position, and TWI
+Larger trees may also have lower drought resistance because of
+microenvironmental and ecological factors. Their crowns tend to occupy
+more exposed canopy positions, which are associated with higher
+evaporative demand \citep{kunert_revised_2017}. Subcanopy trees tend to
+fare better specifically due to the benefits of a buffered environment
+\citep{pretzsch_drought_2018}. Counteracting the liabilities associated
+with tall height, large trees tend to have larger root systems
+\citep{enquist_global_2002}, potentially mitigating some of the
+biophysical challenges they face by allowing greater access to water.
+Larger root systems--if they grant access to deeper water sources--would
+be particularly advantageous in drier microenvironments (e.g., hilltops,
+as compared to valleys and streambeds) during drought. Finally, tree
+size-related responses to drought can be modified by species' traits and
+their distribution across size classes
+\citep{meakem_role_2018, liu_hydraulic_2019}. Understanding the
+mechanisms driving the greater relative growth reductions of larger
+trees during drought requires sorting out the interactive effects of
+height and associated exposure, root water access, and species' traits.
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=5.20833in]{tables_figures/Figure2.png}
-\caption{Height profile graphs}
-\end{figure}
+Debates have also arisen regarding the traits influencing tree growth
+responses to drought. Studies within temperate broadleaf forests have
+observed ring-porous species showing higher drought tolerance than
+diffuse-porous species
+\citep{friedrichs_species-specific_2009, elliott_forest_2015, kannenberg_linking_2019},
+but this distinction would not hold in the global context
+\citep{wheeler_variations_2007, olson_xylem_2020} and does not resolve
+differences among the many species within each category.
+Commonly-measured traits including wood density and leaf mass per area
+(\(LMA\)) have been linked to drought responses within some temperate
+deciduous forests
+\citep{abrams_adaptations_1990, guerfel_impacts_2009, hoffmann_hydraulic_2011, martinbenito_convergence_2015}
+and across forests worldwide \citep{greenwood_tree_2017}. However, in
+other cases these traits could not explain drought tolerance
+\citep[e.g., in a tropical rainforest;][]{marechaux_leaf_2019}, or the
+direction of response was not always consistent. For instance, higher
+wood density has been associated with greater drought resistance at a
+global scale \citep{greenwood_tree_2017}, but correlated negatively with
+tree performance during drought in a broadleaf deciduous forest in the
+southeastern United States \citep{hoffmann_hydraulic_2011}. Thus, the
+perceived influence of these traits on drought resistance may actually
+reflect indirect correlations with other traits that more directly drive
+drought responses \citep{hoffmann_hydraulic_2011}.
 
-\textbf{Results for second main question: how do species' traits - alone
-and in interaction with tree size - influence drought response? } H2.1,
-H2.2, H2.3 Hydraulic traits alone, traits with height
+In contrast, hydraulic traits have direct physiological linkages to tree
+growth and mortality responses to drought. For instance, water
+potentials at which percent the loss of conductivity surpasses a certain
+threshold (e.g., P50 and P88, representing 50 and 88\% loss of
+conductivity, respectively) and hydraulic safety margin (\emph{i.e.},
+difference between typical minimum water potentials and P50 or P88)
+correlate with drought performance across global forests
+\citep{anderegg_meta-analysis_2016}. However, these are time-consuming
+to measure and therefore infeasible for predicting or modeling drought
+responses in highly diverse forests (\emph{e.g.}, in the tropics). More
+easily-measurable leaf drought tolerance traits that have direct linkage
+to plant hydraulic function can explain variation in plant distribution
+and function \citep{medeiros_extensive_2019}. These include leaf area
+shrinkage upon desiccation \citep[\(PLA_{dry}\);][]{scoffoni_leaf_2014}
+and the leaf water potential at turgor loss point (\(\pi_{tlp}\)),
+\emph{i.e.,} the water potential at which leaf wilting occurs
+\citep{bartlett_correlations_2016, zhu_leaf_2018}. Both traits correlate
+with hydraulic vulnerability and drought tolerance as part of unified
+plant hydraulic systems
+\citep{scoffoni_leaf_2014, bartlett_correlations_2016, zhu_leaf_2018, farrell_does_2017}.
+The abilities of both \(PLA_{dry}\) and \(\pi_{tlp}\) to explain tree
+drought resistance remains untested.
 
-\includegraphics{tables_figures/Table4.png} \#this is
-tested\_traits\_all
+Here, we examine how tree height, microenvironment characteristics, and
+species' traits collectively shape drought resistance, defined as the
+ratio of annual growth in a drought year to that which would be expected
+in the absence of drought based on previous years' growth. We test a
+series of hypotheses and associated specific predictions (Table 1) based
+on the combination of tree-ring records from the three strongest
+droughts over a 60-year period (1950 - 2009), species trait
+measurements, and census and microenvironmental data from a large forest
+dynamics plot in Virginia, USA. First, we focus on how tree size, alone
+and in its interaction with microenvironmental gradients, influences
+drought resistance. We examine the contemporary relationship between
+tree height and microenvironment, including growing season
+meteorological conditions and crown exposure. We then test whether,
+consistent with most forests globally, larger-diameter, taller trees
+tend to have lower drought resistance in this forest, which is in a
+region (eastern North America) represented by only two studies in the
+global review of Bennett et al.~(2015). We also test for an influence of
+potential access to available soil water, which should be greater for
+larger trees in dry but not in perpetually wet microsites. Finally, we
+focus on the role of species' traits, testing the hypothesis that
+species' traits-\/--particularly leaf leaf drought tolerance
+traits-\/--predict drought resistance. We test predictions that drought
+resistance is higher in ring-porous than semi-ring and diffuse-porous
+species and that it is correlated with wood density--either positively
+\citep{greenwood_tree_2017} or negatively
+\citep{hoffmann_hydraulic_2011} and positively correlated with \(LMA\).
+We further test predictions that species with low \(PLA_{dry}\) have
+higher drought resistance, and that species whose leaves lose turgor
+lower water potentials (more negative \(\pi_{tlp}\)) have higher
+resistance.
 
-\textbf{Results for third main question: are responses similar or
-variable across individual drought years?} H3.1 and H3.2 Combining
-biophysical with hydraulic traits, which come out as candidates for best
-model?
+\hypertarget{materials-and-methods}{%
+\subsubsection{Materials and Methods}\label{materials-and-methods}}
 
-\subsubsection{Discussion}\label{discussion}
+\emph{Study site and microclimate}
 
-Discussion outline
+Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth
+Observatory) study plot at the Smithsonian Conservation Biology
+Institute (SCBI) in Virginia, USA (38°53'36.6``N, 78°08'43.4''W; Fig.
+\textbf{S1})
+\citep{bourg_initial_2013, andersonteixeira_ctfs-forestgeo:_2015}. SCBI
+is located in the central Appalachian Mountains near the northern
+boundary of Shenandoah National Park. Elevations range from 273 to 338 m
+above sea level with a topographic relief of 65m
+\citep{bourg_initial_2013}. Climate is humid temperate, with mean annual
+temperature of 12.7\textsuperscript{\(\circ\)}C and precipitation of
+1005 mm yr\textsuperscript{-1} during our study period \citep[1960-2009;
+source: CRU TS v.4.01;][]{harris_updated_2014}. Dominant tree taxa
+within this secondary forest include \emph{Liriodendron tulipifera},
+oaks (\emph{Quercus} spp.), and hickories (\emph{Carya} spp.; Table 2).
 
-\begin{itemize}
-\tightlist
-\item
-  first paragraph is summary of main findings, not re-presenting the
-  results
-\item
-  ``we supported this hypothesis, but not this one'' (following same
-  order as Table 1 still)
-\item
-  direction of responses seems to be relatively consistent but the
-  individual responses vary
-\item
-  tie things in (see here
-\item
-  main thing is that we're now better understanding what confers
-  vulnerability or resilience on trees during drought - how forests
-  respond to future droughts/climate change
-\item
-  ``we filled the gap'' = 1-2 sentences
-\item
-  limitations at our site:
-\item
-  aren't able to analyze historical forest community, nor trees that
-  were killed by these droughts {[}aka we don't have data for
-  individuals that were most severely affected{]} (though we found that
-  there's little variation in climate sensitivity for trees that were
-  cored dead vs cored alive).
-\item
-  p50/p80
-\item
-  We used crown position despite its uncertainty. However, height and
-  crown position change relatively slowly so it shouldn't be that far
-  off.
-\item
-  We did not use crowding index because it has much more uncertainty
-\item
-  limitations for extrapolating other sites:
-\item
-  forests are different from place to place, but we've seen how in
-  forests around the world, forests tend to suffer more. We've
-  identified some facets for why this happens. (cite other studies)
-\item
-  has been observed elsewhere that individuals in more moist habitats
-  are more susceptible to drought (cite other studies)
-\item
-  the species may be different from other sites but the hydraulic traits
-  are general across species;
-\item
-  next paragraph saying that our study advances understanding of
-  droughts, e.g.~bennett et al 2015 doesn't go into mechanisms but we do
-\item
-  and we show that height is more important than exposure, but doesn't
-  eliminate the effect of it
-\item
-  further contribution is we show that two leaf hydraulic traits -
-  relatively easy to measure - are good predictors of drought response,
-  can be helpful for scaling up (e.g.~from our site to eastern deciduous
-  biome)
-\item
-  biophysical mechanisms are things that should be seen as universal but
-  relative importance of each can vary within each drought
-\item
-  first study to show how these traits affect woody growth response to
-  drought \textbf{confirmed by Lawren}
-\item
-  science is better now. by advancing our understanding of the
-  mechanisms for individual-level responses to drought, this opens door
-  for better predictions (elaborate from above). ``This is absolutely
-  critical to predicting forest responses to future droughts, which
-  we're likely to see more of in the future as a result to climate
-  change. Forecasting forest responses to these droughts is a huge and
-  important challenge'', science is better.
-\end{itemize}
+\emph{Identifying drought years}
 
-\emph{1. paragraph summarizing main results--\textgreater{} primary
-conclusions} When including only biophysical traits, trees' resistance
-value (on a per-species basis) is explained best by crown position and
-height, with codominant trees being the most resistant to drought. This
-follows on work done by \citep{bennett_larger_2015} {[}and others?{]}
-which show that larger trees suffer more during drought, and confirms
-that this susceptibility can be seen in tree ring analyses. Adding in
-crown position with the leaf hydraulic traits yields a slightly worse
-predictive model for drought tolerance, with height remaining as the
-only significant biophysical variable.
+We identified the three largest droughts within the time period
+1950-2009, defining drought \citep{slette_how_2019} as events with
+anomalously dry peak growing season climatic conditions. Specifically,
+we used the metric of Palmer Drought Severity Index (PDSI) during
+May-August (MJJA; Table S1), which were identified by
+\citet{helcoski_growing_2019} as the months of the current year to which
+annual tree growth was most sensitive at this site. PDSI divisional data
+for Northern Virginia were obtained from NOAA
+(\url{https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp}) in
+December 2017. Based on this, we identified the three strongest droughts
+during the study period (Figs. \textbf{1, S1}; Table S1).
 
-We partially supported the hypothesis that crown exposure makes trees
-more vulnerable to drought. Co-dominant trees had the highest drought
-resistance. Dominant trees had lower resistance, likely because they are
-the most exposed. Other studies have found clear evidence of greater
-drought sensitivity in trees with exposed crowns (e.g.,
-\citep{suarez_factors_2004}; \citep{scharnweber_confessions_2019}). At
-the same time, intermediate and suppressed trees had even lower
-resistance. This indicates that other mechanisms such as competition or
-rooting depth were important. (Also note that our study design was not
-ideal for testing the role of canopy position. Current canopy position
-is a conservative separator of canopy position: trees may currently be
-in more dominant positions than they were at the time, but backwards
-movement is unlikely. This would bias against finding a signficant
-effect for H1.2. Height may be a more reliable predictor of past canopy
-position than is current canopy position, and explains a portion of
-variation in canopy position.)
+The droughts differed in intensity and antecedent moisture conditions
+(Fig. \textbf{S1}, Table S1). The 1966 drought was preceded by two years
+of moderate drought during the growing season and severe to extreme
+drought starting the previous fall. In August 1966, \(PDSI\) reached its
+lowest monthly value (-4.82) of the three droughts. The 1977 drought was
+the least intense throughout the growing season, and it was preceded by
+2.5 years of near-normal conditions, making it the mildest of the three
+droughts. The 1999 drought was preceded by wetter than average
+conditions until the previous June, but \(PDSI\) plummeted below -3.0 in
+October 1998 and remained below this threshold through August 1999.
 
-Proximity to stream--either vertical (elev) or horizontal (distance)--
-did not increase drought resistance; rather, it tended to decrease
-resistance (H1.3a). This may be because individuals growing further from
-water are acclimatized to drier conditions. However, the increase in
-drought resistance with distance from stream was less for small than
-large trees (H1.3b), indicating a potential importance of root
-depth/volume in conferring drought resistance.
+\emph{Data collection and preparation}
 
-\textbf{misc content to integrate} From \citep{kannenberg_linking_2019},
-species with diffuse porous wood anatomy (\emph{Liriodendron}) are more
-sensitive to drought, whereas ring-porous are not as affected because
-they more easily rebuild structures for hydraulic conductivity. This
-paper mentions it would be good to have this data with respect to latent
-affects from drought. \#\#\# Acknowledgements
+Within or just outside the ForestGEO plot, we collected data on a suite
+of variables including tree heights, microenvironment characteristics,
+and species traits (Table 3). The SCBI ForestGEO plot was censused in
+2008, 2013, and 2018 following standard ForestGEO protocols, whereby all
+free-standing woody stems \(\ge\) 1cm diameter at breast height (DBH)
+were mapped, tagged, measured at DBH, and identified to species
+\citep{condit_tropical_1998}. From these census data, we used
+measurements of DBH from 2008 to calculate historical DBH and data for
+all stems \(\ge\) 10cm to analyze functional trait composition relative
+to tree height (all analyses described below). Census data are available
+through the ForestGEO data portal (www.forestgeo.si.edu).
 
-words
+We analyzed tree-ring data (xylem growth increment) from 571 trees
+representing the twelve dominant species (Table 2; Fig. \textbf{S2}).
+Selected species were those with the greatest contributions to woody
+aboveground net primary productivity (\(ANPP_{stem}\)) and together
+comprised 97\% of study plot \(ANPP_{stem}\) between 2008 and 2013
+\citep{helcoski_growing_2019}. Cores (one per tree) were collected
+within the ForestGEO plot at breast height (1.3m) in 2010-2011 or
+2016-2017. In 2010-2011, cores were collected from randomly selected
+live trees of each species that had at least 30 individuals \(\ge\) 10
+cm DBH \citep{bourg_initial_2013}. Annual tree mortality censuses were
+initiated in 2014 \citep{gonzalezakre_patterns_2016}, and in 2016-2017,
+cores were collected from all trees found to have died since the
+previous year's census. We note that drought was probably not a cause of
+mortality for these trees, as monthly May-Aug \(PDSI\) did not drop
+below -1.75 in these years or the three years prior (2013-2017), and
+that trees cored dead displayed similar climate sensitivity to trees
+cored live \citep{helcoski_growing_2019}. Cores were sanded, measured,
+and crossdated using standard procedures, as detailed in
+\citep{helcoski_growing_2019}. The resulting chronologies (Fig.
+\textbf{1a}) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in
+association with \citet{helcoski_growing_2019}.
 
-\subsubsection{Author Contribution}\label{author-contribution}
+For each cored tree, we combined tree-ring records and allometric
+equations of bark thickness to reconstruct DBH for the years 1950-2009.
+Prior \(DBH\) was estimated using the following equation:
 
-words
+\[DBH_Y  = DBH_{2008} - 2*\left[r_{bark,2008} - r_{bark,Y} + \sum_{year=Y}^{2008} r_{ring, Y} \right]\]
 
+Here, \(Y\) denotes the year of interest, \(r_{ring}\) denotes ring
+width derived from cores, and \(r_{bark}\) denotes bark thickness. Bark
+thickness was estimated from species-specific allometries based on the
+bark thickness data from the site
+\citep{andersonteixeira_size-related_2015}. Specifically, we used linear
+regression on log-transformed data to relate \(r_{bark}\) to diameter
+inside bark from 2008 data (Table S2), which were then used to determine
+\(r_{bark}\) in the \(DBH\) reconstruction.
+
+Tree heights (\(H\)) were measured by several researchers for a variety
+of purposes between 2012 and 2019 (n=1,518 trees). Methods included
+direct measurements using a collapsible measurement rod on small trees
+\citep{neon_national_2018} or a tape measure on recently fallen trees
+(this study); geometric calculations using clinometer and tape measure
+\citep{stovall_assessing_2018} or digital rangefinders
+\citep{andersonteixeira_size-related_2015, neon_national_2018}; and
+ground-based LiDAR \citep{stovall_terrestrial_2018}. Rangefinders used
+either the tangent method (Impulse 200LR, TruPulse 360R) or the sine
+method (Nikon ForestryPro) for calculating heights. Both methods are
+associated with some error \citep{larjavaara_measuring_2013}, but in
+this instance there was no clear advantage of one or the other.
+Measurements from the National Ecological Observatory Network (NEON)
+were collected near the ForestGEO plot following standard NEON protocol,
+whereby vegetation of short stature was measured with a collapsible
+measurement rod, and taller trees with a rangefinder
+\citep{neon_national_2018}. Species-specific height allometries were
+developed (Table S3) using log-log regression (\(ln[H] \sim ln[DBH]\)).
+For species with insufficient height data to create reliable
+species-specific allometries (n=2, JUNI and FRAM), heights were
+calculated from an equation developed by combining the height
+measurements across all species. We then used these allometries to
+estimate \(H\) for each drought year, \(Y\), based on reconstructed
+\(DBH_Y\). The distribution of \(H\) across drought years is shown in
+Fig. \textbf{S3}.
+
+To characterize how environmental conditions vary with height, data were
+obtained from the NEON tower located \textless1km from the study area
+via the neonUtilities package \citep{R-neonUtilities}. We used wind
+speed, relative humidity, and air temperature data, all measured over a
+vertical profile spanning heights from 7.2 m to above the top of the
+tree canopy (31.0 or 51.8m, depending on censor), for the years
+2016-2018 \citep{neon_national_2018}. After filtering for missing and
+outlier values, we determined the daily minima and maxima, which we then
+aggregated at the monthly scale.
+
+Crown position--a categorical variable classifying trees based on
+exposure to sunlight--was recorded for all cored trees that remained
+standing during the growing season of 2018 following the protocol of
+\citet{jennings_assessing_1999}. Trees were classified as follows:
+\emph{dominant} trees were defined as those with crowns above the
+general level of the canopy, \emph{co-dominant} trees as those with
+crowns within the the canopy; \emph{intermediate} trees as those with
+crowns below the canopy level, but illuminated from above; and
+\emph{suppressed} as those below the canopy and receiving minimal direct
+illumination from above.
+
+Topographic wetness index (TWI), used here as a metric of long-term mean
+moisture availability, was calculated using the dynatopmodel package in
+R (Fig. \textbf{S2}) \citep{R-dynatopmodel}. Originally developed by
+\citet{beven_physically_1979}, TWI was part of a hydrological run-off
+model and has since been used for a number of purposes in hydrology and
+ecology \citep{sorensen_calculation_2006}. TWI calculation depends on an
+input of a digital elevation model (DEM; \textasciitilde3.7 m resolution
+from the elevatr package \citep{R-elevatr}), and from this yields a
+quantitative assessment defined by how ``wet'' an area is, based on
+areas where run-off is more likely. From our observations in the plot,
+TWI performed better at categorizing wet areas than the Euclidean
+distance from the stream.
+
+Species' trait data were collected in August 2018 (Tables 2-3; Fig.
+\textbf{S4}). We sampled small, sun-exposed branches up to eight meters
+above the ground from three individuals of each species in and around
+the ForestGEO plot. Sampled branches were re-cut under water at least
+two nodes above the original cut and re-hydrated overnight in covered
+buckets under opaque plastic bags before measurements were taken.
+Rehydrated leaves taken towards the apical end of the branch (n=3 per
+individual: small, medium, and large) were scanned, weighed, dried at
+60\(^\circ\) C for \(\ge\) 48 hours, and then re-scanned and weighed.
+Leaf area was calculated from scanned images using the LeafArea R
+package \citep{R-LeafArea}. \(LMA\) was calculated as the ratio of leaf
+dry mass to fresh area. \(PLA_{dry}\) was calculated as the percent loss
+of area between fresh and dry leaves. Wood density was calculated for
+\textasciitilde1cm diameter stem samples (bark and pith removed) as the
+ratio of dry weight to fresh volume, which was estimated using
+Archimedes' displacement. We used the rapid determination method of
+\citet{bartlett_rapid_2012} to estimate osmotic potential at turgor loss
+point (\(\pi_{tlp}\)). Briefly, two 4 mm diameter leaf discs were cut
+from each leaf, tightly wrapped in foil, submerged in liquid nitrogen,
+perforated 10-15 times with a dissection needle, and then measured using
+a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
+Osmotic potential (\(\pi_{osm}\)) given by the osmometer was used to
+estimate (\(\pi_{tlp}\)) using the equation
+\(\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}\) \citep{bartlett_rapid_2012}.
+
+\emph{Statistical Analysis}
+
+For each drought year, we calculated a metric drought resistance
+(\(Rt\)) as the ratio of basal area increment (\(BAI\); \emph{i.e.},
+change in cross-sectional area) during the drought year to the mean
+\(BAI\) over the five years preceding the drought
+\citep{lloret_components_2011}. Thus, \(Rt\) values \textless1 and
+\textgreater1 indicate growth reductions and increases, respectively.
+Because the \(Rt\) metric could be biased by directional pre-drought
+growth trends, we also tried an intervention time series analysis
+(ARIMA, \citep{R-forecast}) that predicted mean drought-year growth
+based on trends over the preceding ten years and used this value in
+place of the five-year mean in calculations of resistance
+(\(Rt_{ARIMA}\)= observed \(BAI\)/ predicted \(BAI\)). The two metrics
+were strongly correlated (Fig. \textbf{S5}). Visual review of the
+individual tree-ring sequences with the largest discrepancies between
+these metrics revealed that \(Rt\) was less prone to unreasonable
+estimates than \(Rt_{ARIMA}\), so we selected \(Rt\) as our focal
+metric, presenting parallel results for \(Rt_{ARIMA}\) in the
+Supplementary Info. In this study we focus exclusively on drought
+resistance metrics (\(Rt\) or \(Rt_{ARIMA}\)), and not on the resilience
+metrics described in \citet{lloret_components_2011}, because (1) we
+would expect resilience to be controlled by a different set of
+mechanisms, and (2) the findings of \citep{desoto_low_2020} suggest that
+\(Rt\) is a more important drought response metric for angiosperms in
+that low resistance to moderate droughts was a better predictor of
+mortality during subsequent severe droughts than the resilience metrics.
+
+Analyses focused on testing the predictions presented in Table 1 with
+\(Rt\) as the response variable, and then repeated using \(Rt_{ARIMA}\)
+as the response variable. Models were run for all drought years combined
+and for each drought year individually. The general statistical model
+for hypothesis testing was a mixed effects model, implemented in the
+lme4 package in R \citep{R-lme4}. In the multi-year model, we included a
+random effect of tree nested within species and a fixed effect of
+drought year to represent the combined effects of differences in drought
+characteristics. Individual year models included a random effect of
+species. All models included fixed effects of independent variables of
+interest (Tables 1,3) as specified below. All variables across all best
+models had variance inflation factors \textless1.2 (1 +/- 0.019). We
+used AICc to assess model selection, and conditional/marginal R-squared
+to assess model fit as implemented in the AICcmodavg package in R
+\citep{R-AICcmodavg}. AICc refers to a corrected version of AICc, and is
+best suited for small data sizes \citep[see][]{brewer_relative_2016}.
+
+To avoid over-fitting models with five species traits (Table 3) across
+only 12 species, we did not include all traits as fixed effects in a
+single linear mixed model, but rather conducted individual tests of each
+species trait to determine the relative importance and appropriateness
+for inclusion in the main model. These tests followed the model
+structure specified above, then added \(ln[H]\) and \(ln[TWI]\) to
+create a base model against which we tested traits. Trait variables were
+considered appropriate for inclusion in the main model if they had a
+consistent direction of response across all droughts and if their
+addition to the base model improved fit (at \(\Delta\)AICc \(\ge\) 1.0)
+in at least one drought year (Table S4). We note that we did not use the
+\(\Delta\)AICc \(\ge\) 1.0 criterion as a test of significance, but
+rather of whether the variable had enough influence to be considered as
+a \emph{candidate} variable in full models.
+
+We then determined the top full models for predicting \(Rt\) (or
+\(Rt_{ARIMA}\)). To do so, we compared models with all possible
+combinations of candidate variables, including \(ln[H]\)*\(ln[TWI]\) and
+species traits as specified above. We identified the full set of models
+within \(\Delta\)AICc=2 of the best model (that with lowest AICc). When
+a variable appeared in all of these models and the sign of the
+coefficient was consistent across models, we viewed this as support for
+the acceptance/rejection of the associated prediction (Table 1). If the
+variable appeared in some but not all of these models, and its sign was
+consistent across models, we considered this partial support/rejection.
+In presentation of the results below, we note instances where the
+\(Rt_{ARIMA}\) model disagreed with the \(Rt\) model, but otherwise do
+not discuss the \(Rt_{ARIMA}\) model.
+
+All analysis beyond basic data collection was performed using R version
+3.6.2 \citep{R-base}. Other R-packages used in analyses are listed in
+the Supplementary Information (\emph{Appendix S1}). All data, code, and
+results are available through the SCBI-ForestGEO organization on GitHub
+(\url{https://github.com/SCBI-ForestGEO}: SCBI-ForestGEO-Data and
+McGregor\_climate-sensitivity-variation repositories), with static
+versions corresponding to data and analyses presented here archived in
+Zenodo (DOIs: 10.5281/zenodo.3604993 and \emph{{[}TBD{]}}, respectively.
+
+\hypertarget{results}{%
+\subsubsection{Results}\label{results}}
+
+\emph{Tree height and microenvironment}
+
+In the years for which we have vertical profiles in climate data
+(2016-2018), taller trees--or those in dominant crown positions-- were
+generally exposed to higher evaporative demand during the peak growing
+season months (May-August; Fig. \textbf{2}). Specifically, maximum daily
+wind speeds were significantly higher above the top of the canopy
+(40-50m) than within and below (10-30m) (Fig. \textbf{2a}). Relative
+humidity was also somewhat lower during June-August, ranging from
+\textasciitilde50-80\% above the canopy and \textasciitilde60-90\% in
+the understory (Fig. \textbf{2b}). Air temperature did not vary
+consistently across the vertical profile (Fig. \textbf{2c}).
+
+Crown position varied as expected with height (dominant \textgreater{}
+co-dominant \textgreater{} intermediate \textgreater{} suppressed), but
+with substantial variation (Fig. \textbf{2d}). There were significant
+differences in height across all crown position classes (Fig.
+\textbf{2d}). A comparison test between height and crown position data
+from the most recent ForestGEO census (2018) revealed a correlation of
+0.73.
+
+\emph{Community-level drought responses}
+
+At the community level, cored trees showed substantial growth reductions
+in all three droughts, with a mean \(Rt\) of 0.86 in 1966 and 1999, and
+0.84 in 1977 (Fig. \textbf{2b}). Across the entire study period
+(1950-2009), the focal drought years were the three years with the
+largest fraction of trees exhibiting \(Rt \le 0.7\). Specifically, in
+each drought, roughly 30\% of the cored trees had growth reductions of
+\(\ge\) 30\% (\(Rt \le 0.7\)): 29\% in 1966, 32\% in 1977, and 27\% in
+1999. However, some individuals exhibited increased growth, \emph{i.e.},
+\(Rt > 1.0\): 26\% of trees in 1966, 22\% in 1977, and 26\% in 1999.
+
+In the context of the multivariate model, \(Rt\) did not vary across
+drought years. That is, drought year as a variable did not appear in any
+of the top models -- \emph{i.e.}, models that were statistically
+indistinguishable (\(\Delta\)AICc\textless2) from the best model.
+
+\emph{Tree height, microenvironment, and drought resistance}
+
+Taller trees (based on \(H\) in the drought year) showed stronger growth
+reductions during drought (Table 1; Figs. \textbf{4, S6}). Specifically,
+\(ln[H]\) appeared, with a negative coefficient, in the best model
+((\(\Delta\)AICc=0) and all top models when evaluating the three drought
+years together (Tables S6-S7). The same held true for 1966 individually.
+For the 1977 drought, \(ln[H]\) did not appear in the best model, but
+was included, with a negative coefficient, among the top
+models--\emph{i.e.}, models that were statistically indistinguishable
+(\(\Delta\)AICc\textless2) from the best model (Tables 1, S6-S7). For
+the 1999 drought, \(ln[H]\) had no significant effect.
+
+\(Rt\) had a significantly negative response to \(ln[TWI]\) across all
+drought years combined (Figs. \textbf{4, S6}, Table S6-S7). The effect
+was also significant for 1977 and 1999 individually (Fig. \textbf{4},
+Table S6). When \(Rt_{ARIMA}\) was used as the response variable, the
+effect was significant in 1977, and included in some of the top models
+in 1966 and 1999 (Table S7). This negates the idea that trees in moist
+microsites would be less affected by drought. Nevertheless, we tested
+for a \(ln[H] *ln[TWI]\) interaction, a negative sign of which could
+indicate that smaller trees (presumably with smaller rooting volume) are
+more susceptible to drought in drier microenvironments with a deeper
+water table. This hypothesis was rejected, as the \(ln[H] *ln[TWI]\)
+interaction was never significant, and had a positive sign in any top
+models in which it appeared (Tables 1, S6-S7). This term did appear with
+a positive coefficient in the best \(Rt_{ARIMA}\) model for all years
+combined (Table S7), indicating that the negative effect of height on
+\(Rt\) was significantly stronger in wetter microhabitats.
+
+\emph{Species' traits and drought resistance}
+
+Species, as a factor in ANOVA, had significant influence
+(p\textless0.05) on all traits (wood density, \(LMA\), \(PLA_{dry}\),
+and \(\pi_{tlp}\)), with more significant pairwise differences for wood
+density and \(PLA_{dry}\) than for \(LMA\) and \(\pi_{tlp}\) (Table 2,
+Fig. \textbf{S4}). Drought resistance also varied across species,
+overall and in each drought year (Fig. \textbf{3}). Significant
+differences in \(Rt\) across species were most pronounced in 1966 with a
+total of seven distinct groupings, while 1977 had four and 1999 had two.
+Averaged across all droughts, \(Rt\) was lowest in \emph{Liriodendron
+tulipifera} (mean \(Rt\) = \textbf{0.66}) and highest in \emph{Fagus
+grandifolia} (mean \(Rt\) = \textbf{0.99}).
+
+Wood density, \(LMA\), and xylem porosity were all poor predictors of
+\(Rt\) (Tables 1,S4-S5). Wood density and \(LMA\) were never
+significantly associated with \(Rt\) in the single-variable tests and
+were therefore excluded from the full models. Xylem porosity was also
+excluded from the full models, as it had no significant influence for
+all droughts combined and had contrasting effects in the individual
+droughts: whereas ring-porous species had higher \(Rt\) than diffuse-
+and semi-ring- porous species in the 1966 and 1999 droughts, they had
+lower \(Rt\) in 1977 (Table S4). It is noteworthy that the two
+diffuse-porous species in our study, \emph{Liriodendron tulipifera} and
+\emph{Fagus grandifolia}, were at opposite ends of the \(Rt\) spectrum
+(Fig. \textbf{3}), further refuting the idea that xylem porosity is a
+useful predictor of \(Rt\) in the context of this study.
+
+In contrast, \(PLA_{dry}\), and \(\pi_{tlp}\) were both negatively
+correlated to drought resistance (Figs. \textbf{4, S6}; Tables 1,S4-S7).
+Both had consistent signs across all droughts, and their inclusion at
+least marginally improved the model (\(\Delta\)AICc \textgreater{} 1.0)
+for at least one of the three droughts (Table S4), qualifying them as
+candidate variables for the full model. \(PLA_{dry}\) had a significant
+influence, with negative coefficient, in full models for the three
+droughts combined and for the 1966 drought individually (Fig.
+\textbf{4}; Tables S6-S7). For 1977 and 1999, it was included with a
+negative coefficient in some of the top models (Tables S6-S7).
+\(\pi_{tlp}\) was included with a negative coefficient in the best model
+for both all droughts combined and for the 1977 drought individually
+(Fig. \textbf{4}; Table S6). It was also included in some of the top
+models for 1999 (Tables S6-S7).
+
+\hypertarget{discussion}{%
+\subsubsection{Discussion}\label{discussion}}
+
+Tree height, microenvironment, and leaf drought tolerance traits shaped
+tree growth responses across three droughts at our study site (Table 1,
+Fig. \textbf{4}). The greater susceptibility of larger trees to drought,
+similar to forests worldwide \citep{bennett_larger_2015}, was driven
+primarily by their height \citep{stovall_tree_2019}. Taller height was
+likely a liability in itself, and was also associated with greater
+exposure to conditions that would promote water loss and heat damage
+during drought (Fig. \textbf{2}). There was no evidence that greater
+availability of, or access to, soil water availability increased drought
+resistance; in contrast, trees in wetter topographic positions had lower
+\(Rt\) \citep{zuleta_drought-induced_2017, stovall_tree_2019}, and the
+larger potential rooting volume of large trees provided no advantage in
+the drier microenvironments. The negative effect of height on \(Rt\)
+held after accounting for species' traits, which is consistent with
+recent work finding height had a stronger influence on mortality risk
+than forest type during drought \citep{stovall_reply_2020}. Drought
+resistance was not consistently linked to species' \(LMA\), wood
+density, or xylem type (ring- vs.~diffuse porous), but was negatively
+correlated with leaf drought tolerance traits (\(PLA_{dry}\),
+\(\pi_{tlp}\)). This is the first study to our knowledge linking
+\(PLA_{dry}\) and \(\pi_{tlp}\) to growth reduction during drought. The
+directions of these responses were consistent across droughts (Table
+S6), supporting the premise that they were driven by fundamental
+physiological mechanisms. However, the strengths of each predictor
+varied across droughts (Fig. \textbf{4}; Tables S6-S7), indicating that
+drought characteristics interact with tree size, microenvironment, and
+traits to shape which individuals are most affected. These findings
+advance our knowledge of the factors that make trees vulnerable to
+growth declines during drought and, by extension, likely make them more
+vulnerable to mortality \citep{sapes_plant_2019}.
+
+The droughts considered here were of a magnitude that has occurred with
+an average frequency of approximately once every 10-15 years \citep[Fig.
+\textbf{1a},][]{helcoski_growing_2019} and had substantial but not
+devastating impacts on tree growth (Figs. \textbf{1b}). These droughts
+were classified as severe (\(PDSI\) \textless{} -3.0; 1977) or extreme
+(\(PDSI\) \textless{} -4.0; 1966, 1999) at our site and have been linked
+to tree mortality in the eastern United States
+\citep{druckenbrod_redefining_2019}. However, extreme, multiannual
+droughts such as the so-called ``megadroughts'' of this type that have
+triggered massive tree die-off in other regions
+\citep[e.g.,][]{allen_global_2010, stovall_tree_2019} have not occurred
+in the Eastern United States within the past several decades
+\citep{clark_impacts_2016}. Of the droughts considered here, the 1966
+drought, which was preceded by two years of dry conditions (Fig.
+\textbf{S1}), severely stressed a larger portion of trees (Fig.
+\textbf{1b}). The tendency for large trees to have lowest resistance was
+most pronounced in this drought, consistent with other findings that
+this physiological response increases with drought severity
+\citep{bennett_larger_2015, stovall_tree_2019}. Across all three
+droughts, the majority of trees experienced reduced growth, but a
+substantial portion had increased growth (Fig. \textbf{1b}), potentially
+due to decreased leaf area of competitors during the drought
+(\textbf{REF--if we can find one}), and consistent with prior
+observations that smaller trees can exhibit increased growth rates
+during drought \citep{bennett_larger_2015}. It is likely because of the
+moderate impact of these droughts, along with other factors influencing
+tree growth (e.g., stand dynamics), that our best models characterize
+only a modest amount of variation in \(Rt\): 11-12\% for all droughts
+combined, and 18-25\% for each individual drought (Fig. \textbf{S6};
+Table S6).
+
+Consistent with studies in other forests worldwide
+\citep{bennett_larger_2015}, taller trees in this forest exhibited lower
+drought resistance. Mechanistically, this is consistent with, and
+reinforces, previous findings that biophysical constraints make it
+impossible for trees to efficiently transport water to great heights and
+simultaneously maintain strong resistance and resilience to
+drought-induced embolism
+\citep{olson_plant_2018, couvreur_water_2018, roskilly_conflicting_2019}.
+Taller trees also face dramatically different microenvironments (Fig.
+\textbf{2}). They are exposed to higher wind speeds and lower humidity
+(Fig. \textbf{2a-b}), resulting in higher evaporative demand. Unlike
+other temperate forests where modestly cooler understory conditions have
+been documented \citep{zellweger_seasonal_2019}, particularly under
+drier conditions \citep{davis_microclimatic_2019}, we observed no
+significant variation in air temperatures across the vertical profile
+(Fig. \textbf{2c}). More critically for tree physiology, leaf
+temperatures can become significantly elevated over air temperature
+under conditions of high solar radiation and low stomatal conductance
+\citep{campbell_introduction_1998, rey-sanchez_spatial_2016}. Under
+drought, when air temperatures tend to be warmer, direct solar radiation
+tends to be higher (because of less cloud cover), and less water is
+available for evaporative cooling of the leaves, trees with sun-exposed
+crowns may not be able to simultaneously maintain leaf temperatures
+below damaging extremes and avoid drought-induced embolism. Indeed,
+previous studies have shown lower drought resistance in more exposed
+trees
+\citep{liu_effect_1993, suarez_factors_2004, scharnweber_confessions_2019}.
+Unfortunately, collinearity between height and crown exposure in this
+study (Fig. \textbf{2d}) makes it impossible to confidently partition
+causality. Additional research comparing drought responses of early
+successional and mature forest stands, along with short and tall
+isolated trees, would be valuable for more clearly disentangling the
+roles of tree height and crown exposure.
+
+Belowground, taller trees would tend to have larger root systems
+\citep{enquist_global_2002}, but this does not necessarily imply that
+they have greater access to or reliance on deep soil-water resources
+that may be critical during drought. While tree size can correlate with
+the depth of water extraction \citep{brum_hydrological_2018}, the
+linkage is not consistent. Shorter trees can vary broadly in the depth
+of water uptake \citep{stahl_depth_2013}, and larger trees may allocate
+more to abundant shallow roots that are beneficial for taking up water
+from rainstorms \citep{meinzer_partitioning_1999}. Moreover, reliance on
+deep soil-water resources can actually prove a liability during severe
+and prolonged drought, as these can experience more intense water
+scarcity relative to non-drought conditions
+\citep{chitra-tarak_roots_2018}. In any case, the potentially greater
+access to water did not override the disadvantage conferred by
+height--and, in fact, greater moisture access in non-drought years
+(here, higher TWI) appears to make trees more sensitive to drought
+\citep{zuleta_drought-induced_2017, stovall_tree_2019}. This may be
+because moister habitats would tend to support species and individuals
+with more mesophytic traits
+\citep{bartlett_drought_2016, mencuccini_ecological_2003, medeiros_extensive_2019},
+potentially growing to greater heights (e.g.,
+\citet{detto_hydrological_2013}), and these are then more vulnerable
+when drought hits. The observed height-sensitivity of \(Rt\), together
+with the lack of conferred advantage to large stature in drier
+topographic positions, agrees with the concept that physiological
+limitations to transpiration under drought shift from soil water
+availability to the plant-atmosphere interface as forests age
+\citep{bretfeld_plant_2018}, such that tall, dominant trees are the most
+sensitive in mature forests. Again, additional research comparing
+drought responses across forests with different tree heights and water
+availability would be valuable for disentangling the relative importance
+of above- and belowground mechanisms across trees of different size.
+
+The development of tree-ring chronologies for the twelve most dominant
+tree species at our site
+\citep{helcoski_growing_2019, bourg_initial_2013} gave us the sample
+size to compare historical drought responses across species (Fig.
+\textbf{3}) and associated traits at a single site \citep[see
+also][]{elliott_forest_2015}. Our study reinforced current understanding
+(see Introduction) that wood density and \(LMA\) are not reliably linked
+to drought resistance (Table 1). Contrary to previous studies in
+temperate deciduous forests, we did not find an association between
+xylem porosity and drought tolerance, as the two diffuse-porous species,
+\emph{Liriodendron tulipifera} and \emph{Fagus grandifolia}, were at
+opposite ends of the \(Rt\) spectrum (Fig. \textbf{3}). While the low
+\(Rt\) of \emph{L. tulipifera} is consistent with other studies
+\citep{elliott_forest_2015}, the high \(Rt\) of \emph{F. grandifolia}
+contrasts with studies identifying diffuse porous species in general
+\citep{elliott_forest_2015, kannenberg_linking_2019}, and the genus
+\emph{Fagus} in particular \citep{friedrichs_species-specific_2009}, as
+drought sensitive. There are two potential explanations for this
+discrepancy. First, other traits can and do override the influence of
+xylem porosity on drought resistance. Ring-porous species are restricted
+mainly to temperate deciduous forests \citep{wheeler_variations_2007},
+while highly drought-tolerant diffuse-porous species exist in other
+biomes (\textbf{REFS}). \emph{Fagus grandifolia} had intermediate
+\(\pi_{tlp}\) and low \(PLA_{dry}\) (Fig. \textbf{S4}), which would have
+contributed to its drought resistance (Fig. \textbf{4}; see discussion
+below). A second explanation of why \emph{F. grandifolia} trees at this
+particular site had higher \(Rt\) is that the sampled individuals,
+reflective of the population within the plot, are generally shorter and
+in less-dominant canopy positions compared to most other species (Fig.
+\textbf{S4}). The species, which is highly shade-tolerant, also has deep
+crowns \citep{andersonteixeira_size-related_2015}, implying that a lower
+proportion of leaves would be affected by harsher microclimatic
+conditions at the top of the canopy under drought (Fig. \textbf{2}).
+Thus, the high \(Rt\) of the sampled \emph{F. grandifolia} population
+can be explained by a combination of fairly drought-resistant leaf
+traits, shorter stature, and a buffered microenvironment.
+
+Concerted measurement of tree-rings and leaf drought tolerance traits of
+emerging importance
+\citep{scoffoni_leaf_2014, bartlett_correlations_2016, medeiros_extensive_2019}
+allowed novel insights into the role of drought tolerance traits in
+shaping drought response. The finding that \(PLA_{dry}\) and
+\(\pi_{tlp}\) can be useful for predicting drought responses of tree
+growth (Fig. \textbf{4}; Table 1) is both novel and consistent with
+previous studies linking these traits to habitat and drought tolerance.
+Previous studies have demonstrated that \(\pi_{tlp}\) and \(PLA_{dry}\)
+are physiologically meaningful traits linked to species distribution
+along moisture gradients
+\citep{marechaux_drought_2015, fletcher_evolution_2018, medeiros_extensive_2019, simeone_coupled_2019, rosas_adjustments_2019, zhu_leaf_2018},
+and our findings indicate that these traits also influence drought
+responses. Furthermore, the observed linkage of \(\pi_{tlp}\) to \(Rt\)
+in this forest aligns with observations in the Amazon that \(\pi_{tlp}\)
+is higher in drought-intolerant than drought-tolerant plant functional
+type. Further, it adds support to the idea that this trait is useful for
+categorizing and representing species' drought responses in models
+\citep{powell_differences_2017}. Because both \(PLA_{dry}\) and
+\(\pi_{tlp}\) can be measured relatively easily
+\citep{bartlett_rapid_2012, scoffoni_leaf_2014}, they hold promise for
+predicting drought growth responses across diverse forests. The
+importance of predicting drought responses from species traits increases
+with tree species diversity; whereas it is feasible to study drought
+responses for all dominant species in most boreal and temperate forests
+(e.g., this study), this becomes difficult to impossible for species
+that do not form annual rings, and for diverse tropical forests.
+Although progress is being made for the tropics
+\citep{schongart_dendroecological_2017}, a full linkage of drought
+tolerance traits to drought responses would be invaluable for
+forecasting how little-known species and whole forests will respond to
+future droughts
+\citep{christoffersen_linking_2016, powell_differences_2017}.
+
+As climate change drives increasing drought in many of the world's
+forests
+\citep{trenberth_global_2014, intergovernmental_panel_on_climate_change_climate_2015},
+the fate of forests and their climate feedbacks will be shaped by the
+biophysical and physiological drivers observed here. Our results,
+consistent with other observations around the world, imply that the
+tallest, most exposed trees will be most affected
+\citep{bennett_larger_2015, stovall_tree_2019}. We show that, at least
+within the mature forest studied here, the vulnerability conferred by
+tall height and associated crown exposure outweigh any advantage of a
+larger root system, even in drier microenvironments. This would suggest
+that the drought responses of trees in mature forests are more strongly
+differentiated along the size spectrum by their above- than below-ground
+environment. The same may not be true of systems where short trees exist
+outside of a buffered understory environment--\emph{i.e.}, open grown
+trees or short-statured, early-successional forests. The latter appear
+to be limited more strongly by root water access during drought
+\citep{bretfeld_plant_2018}, and would also be dominated by species with
+different traits. The earlier-successional species at our site
+(\emph{Liriodendron tulipifera}, \emph{Quercus spp.}, \emph{Fraxinus
+americana}) display a mix of traits conferring drought tolerance and
+resistance (Table 2), while the late-successional \emph{Fagus
+grandifolia} displayed high drought resistance, in part because it
+exists primarily within a buffered microenvironment. Further research on
+how leaf drought tolerance traits and drought vulnerability change over
+the course of succession would be valuable for addressing how drought
+tolerance changes as forests age
+\citep[e.g.~][]{rodriguez-caton_long-term_2015}. In the meantime, the
+results of this study advance our knowledge of the factors conferring
+drought resistance in a mature forest, opening the door for more
+accurate forecasting of forest responses to future drought.
+
+\hypertarget{acknowledgements}{%
+\subsubsection{Acknowledgements}\label{acknowledgements}}
+
+We especially thank the numerous researchers who helped to collect the
+data used here, in particular Jennifer C. McGarvey, Jonathan R.
+Thompson, and Victoria Meakem for original collection and processing of
+cores. Thanks also to Camila D. Medeiros for guidance on leaf drought
+tolerance and functional trait measurements, Edward Brzostek's lab for
+collaboration on leaf sampling, and Maya Prestipino for data collection.
+This manuscript was improved based on helpful reviews by Mark Olson and
+three anonymous reviewers. Funding for the establishment of the SCBI
+ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led
+Forest Global Earth Observatory (ForestGEO), the Smithsonian
+Institution, and the HSBC Climate Partnership. This study was funded by
+ForestGEO, a Virginia Native Plant Society grant to KAT and AJT, and
+support from the Harvard Forest and National Science Foundation which
+supports the PalEON project (NSF EF-1241930) for NP.
+
+\hypertarget{author-contribution}{%
+\subsubsection{Author Contribution}\label{author-contribution}}
+
+KAT, IM, and AJT designed the research. Tree-ring chronologies were
+developed by RH under guidance of AJT and NP. Trait data were collected
+by IM, JZ under guidance of NK and LS. Other plot data were collected by
+IM, AS, EGA, and NB under guidance of EGA and WM. Data analyses were
+performed by IM under guidance of KAT and VH. KAT and IM interpreted the
+results. IM and KAT wrote the first draft of manuscript, and all authors
+contributed to revisions.
+
+\hypertarget{supplementary-information}{%
 \subsubsection{Supplementary
-Information}\label{supplementary-information}
+Information}\label{supplementary-information}}
 
-\begin{table}[!h]
+Table S1: Monthly Palmer Drought Severity Index (PDSI), and its rank
+among all years between 1950 and 2009 (driest=1), for focal droughts.
 
-\caption{\label{tab:Table S1}Species-specific height regression equations}
-\centering
-\begin{tabular}{llr}
-\toprule
-Species & Equations & r.2\\
-\midrule
-Carya cordiformis & 0.348+0.808*x & 0.879\\
-Carya glabra & 0.681+0.704*x & 0.855\\
-Carya ovalis & 0.621+0.722*x & 0.916\\
-Carya tomentosa & 0.776+0.701*x & 0.894\\
-Fagus grandifolia & 0.708+0.662*x & 0.857\\
-\addlinespace
-Liriodendron tulipifera & 1.32+0.524*x & 0.761\\
-Quercus alba & 1.14+0.548*x & 0.647\\
-Quercus prinus & 0.44+0.751*x & 0.869\\
-Quercus rubra & 1.17+0.533*x & 0.773\\
-all & 0.879+0.634*x & 0.857\\
-\bottomrule
-\end{tabular}
-\end{table}
+Table S2: Species-specific bark thickness regression equations.
 
-\emph{p50 and p80} We decided to include values of P50 and P80 in the
-leaf traits model, defined by \citep{anderegg_meta-analysis_2016} as the
-water potentials at which a species loses 50\% and 88\% {[}80\% by
-proxy{]}, respectively, of hydraulic conductivity. Values were
-calculated by (\textbf{insert new methods here??}), and were only
-available for six species (\emph{C. glabra}, \emph{L. tulipifera},
-\emph{Q. alba}, \emph{Q. prinus}, \emph{Q. rubra}, and \emph{Q.
-velutina}). Because of this, the model runs were considered to be
-incomplete due to the exclusion of the other 8 species. Results revealed
-neither p50 nor p80 to be significant, thus for the full analysis we
-decided to drop the two traits in order to include all species in the
-full analysis.
+Table S3: Species-specific height regression equations.
 
-\begin{table}[!h]
+Table S4. Individual tests of species traits as drivers of drought
+resistance, where \(Rt\) is used as the response variable.
 
-\caption{\label{tab:Table S2}Candidate variables for best model}
-\centering
-\begin{tabular}{rlll}
-\toprule
-prediction & variable & variable\_description & top\_model\\
-\midrule
-1.2 & position\_all & crown.position w/height & 1999\\
-2.2 & height.ln.m & ln[height] & all\\
-2.2 & height.ln.m & ln[height] & 1966\\
-2.3 & position\_all & crown.position alone & 1966\\
-2.4 & TWI.ln & ln[topographic.wetness.index] & all\\
-\addlinespace
-2.4 & TWI.ln & ln[topographic.wetness.index] & 1977\\
-2.4 & TWI.ln & ln[topographic.wetness.index] & 1999\\
-3.1 & year & drought.year & all\\
-3.1 & rp & ring.porosity & 1999\\
-3.2 & PLA\_dry\_percent & percent.loss.area & all\\
-\addlinespace
-3.2 & PLA\_dry\_percent & percent.loss.area & 1966\\
-3.4 & mean\_TLP\_Mpa & mean.turgor.loss.point & all\\
-\bottomrule
-\end{tabular}
-\end{table}
+Table S5. Individual tests of species traits as drivers of drought
+resistance, where \(Rt_{ARIMA}\) is used as the response variable.
 
-\begin{figure}[ht]
-\includegraphics[width=1\linewidth]{tables_figures/ForestGEO_plot} \caption{Map of ForestGEO plot}\label{fig:Plot}
-\end{figure}
+Table S6. Summary of top full models for each drought instance, where
+\(Rt\) is used as the response variable.
 
-\emph{how do we want to present Table S3? Would it be better as an image
-of an excel file, since it's so large? Did we want to keep all
-coefficients here?}
+Table S7. Summary of top models for each drought instance, where
+\(Rt_{ARIMA}\) is used as the response variable.
 
-\begin{table}[!h]
+Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the
+2.5 years prior to each focal drought
 
-\caption{\label{tab:Table S3}Top model variations for each drought scenario, with dAICc values <= 2}
-\centering
-\begin{tabular}{lrll}
-\toprule
-Modnames & Delta\_AICc & scenario & coef\\
-\midrule
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.00 & trees\_all\_sub & (Intercept) (1.101), position\_alldominant (-0.044), position\_allintermediate (-0.039), position\_allsuppressed (-0.049), height.ln.m (-0.092), TWI.ln (-0.058), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.148)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 0.84 & trees\_all\_sub & (Intercept) (0.953), height.ln.m (-0.06), TWI.ln (-0.065), year1977 (-0.089), year1999 (-0.076), PLA\_dry\_percent (-0.008), mean\_TLP\_Mpa (-0.164)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+year+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp/tree) & 1.57 & trees\_all\_sub & (Intercept) (1.021), position\_alldominant (-0.046), position\_allintermediate (-0.042), position\_allsuppressed (-0.053), height.ln.m (-0.097), year1977 (-0.088), year1999 (-0.073), PLA\_dry\_percent (-0.009), mean\_TLP\_Mpa (-0.146)\\
-resist.value \textasciitilde{} height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1966 & (Intercept) (1.576), height.ln.m (-0.145), rpring (0.091), rpsemi-ring (0.232), PLA\_dry\_percent (-0.021)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.37 & x1966 & (Intercept) (1.505), height.ln.m (-0.146), TWI.ln (0.043), rpring (0.092), rpsemi-ring (0.227), PLA\_dry\_percent (-0.021)\\
-\addlinespace
-resist.value \textasciitilde{} height.ln.m+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.39 & x1966 & (Intercept) (1.177), height.ln.m (-0.135), PLA\_dry\_percent (-0.014), mean\_TLP\_Mpa (-0.148)\\
-resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.00 & x1977 & (Intercept) (0.386), TWI.ln (-0.136), rpring (-0.233), rpsemi-ring (-0.328), mean\_TLP\_Mpa (-0.37)\\
-resist.value \textasciitilde{} position\_all+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.28 & x1977 & (Intercept) (0.341), position\_alldominant (-0.079), position\_allintermediate (-0.025), position\_allsuppressed (0.039), TWI.ln (-0.128), rpring (-0.23), rpsemi-ring (-0.335), mean\_TLP\_Mpa (-0.385)\\
-resist.value \textasciitilde{} height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.06 & x1977 & (Intercept) (0.469), height.ln.m (-0.032), TWI.ln (-0.133), rpring (-0.223), rpsemi-ring (-0.318), mean\_TLP\_Mpa (-0.368)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 0.00 & x1999 & (Intercept) (1.31), position\_alldominant (0.003), position\_allintermediate (-0.077), position\_allsuppressed (-0.1), height.ln.m (-0.09), TWI.ln (-0.083), rpring (0.191), rpsemi-ring (0.224), PLA\_dry\_percent (-0.009)\\
-\addlinespace
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 0.05 & x1999 & (Intercept) (0.855), position\_alldominant (0.001), position\_allintermediate (-0.078), position\_allsuppressed (-0.1), height.ln.m (-0.092), TWI.ln (-0.08), rpring (0.183), rpsemi-ring (0.06), mean\_TLP\_Mpa (-0.146)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+(1|sp) & 0.55 & x1999 & (Intercept) (1.19), position\_alldominant (0.002), position\_allintermediate (-0.08), position\_allsuppressed (-0.104), height.ln.m (-0.09), TWI.ln (-0.085), rpring (0.199), rpsemi-ring (0.134)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+mean\_TLP\_Mpa+(1|sp) & 0.94 & x1999 & (Intercept) (0.722), position\_alldominant (-0.002), position\_allintermediate (-0.083), position\_allsuppressed (-0.105), height.ln.m (-0.1), rpring (0.183), rpsemi-ring (0.046), mean\_TLP\_Mpa (-0.154)\\
-resist.value \textasciitilde{} TWI.ln+rp+PLA\_dry\_percent+(1|sp) & 1.07 & x1999 & (Intercept) (1.044), TWI.ln (-0.097), rpring (0.187), rpsemi-ring (0.265), PLA\_dry\_percent (-0.01)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+PLA\_dry\_percent+(1|sp) & 1.15 & x1999 & (Intercept) (1.191), position\_alldominant (0.001), position\_allintermediate (-0.081), position\_allsuppressed (-0.106), height.ln.m (-0.098), rpring (0.192), rpsemi-ring (0.215), PLA\_dry\_percent (-0.009)\\
-\addlinespace
-resist.value \textasciitilde{} TWI.ln+rp+mean\_TLP\_Mpa+(1|sp) & 1.41 & x1999 & (Intercept) (0.516), TWI.ln (-0.094), rpring (0.178), rpsemi-ring (0.074), mean\_TLP\_Mpa (-0.166)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+TWI.ln+rp+PLA\_dry\_percent+mean\_TLP\_Mpa+(1|sp) & 1.82 & x1999 & (Intercept) (1.084), position\_alldominant (0.002), position\_allintermediate (-0.077), position\_allsuppressed (-0.099), height.ln.m (-0.091), TWI.ln (-0.081), rpring (0.186), rpsemi-ring (0.147), PLA\_dry\_percent (-0.005), mean\_TLP\_Mpa (-0.076)\\
-resist.value \textasciitilde{} position\_all+height.ln.m+rp+(1|sp) & 1.88 & x1999 & (Intercept) (1.065), position\_alldominant (-0.001), position\_allintermediate (-0.085), position\_allsuppressed (-0.11), height.ln.m (-0.098), rpring (0.201), rpsemi-ring (0.123)\\
-\bottomrule
-\end{tabular}
-\end{table}
+Figure S2: Map of ForestGEO plot showing topographic wetness index and
+location of cored trees
 
-\includegraphics{tables_figures/FigureS1.png} (see Issue \#32)
+Figure S3: Distribution of reconstructed tree heights across drought
+years.
+
+Figure S4. Distribution of independent variables by species.
+
+Figure S5. Comparison of \(Rt\) and \(Rt_{ARIMA}\) results, with
+residuals, for each drought scenario
+
+Figure S6. Visualization of best model, with data, for all droughts
+combined.
 

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
fixed refs

@@
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
-Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf leaf drought tolerance traits-–predict drought resistance. 
+Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
@@
 Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-While tree size can correlate with the depth of water extraction [@brum_hydrological_2018], the linkage is not consistent.
+While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent.
 Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
-Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018].
+Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitratarak_roots_2018].
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

### 2020-07-23 manuscript/response_to_reviews.Rmd
removed claim lacking ref

@@
 **We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), potentially due to decreased leaf area of competitors during the drought, and consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al.
+1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al.
 2015).
@@
 *Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
+1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
 

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
removed claim lacking ref

@@
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-**1b**), potentially due to decreased leaf area of competitors during the drought (**REF--if we can find one**), and consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
+**1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Fig.

### 2020-07-23 manuscript/response_to_reviews.Rmd
clarify drought tolerance vs resistance
#124

@@
 
-*Current text, with phrase of concern in bold: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]." *
+**We have removed the claim that increased growth may be due to reduced leaf area.**
 
@@
 It could be interesting to try to align the size range over which Rt changes most dramatically with those over which conduit diameter and microclimate change most rapidly, but unfortunately that's beyond the scope here.**
-
-*need to put something in the text here. Line 138 was where we talked about the sizes of cored trees.*
-
-*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/121*
 
@@
 
+**We now define drought resistance in the first paragraph: " Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...".
+We then use this term when we wish to speak specifically about resistance.
+We use "tolerance" as a more general term, consistent with what seems to be a fairly nebulous definition in the literature.**
+
 *https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/124*

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
clarify drought tolerance vs resistance
#124

@@
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015].
-Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts.
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts.
 Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
@@
 Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
-Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
 **3**).
@@
 First, other traits can and do override the influence of xylem porosity on drought resistance. 
-Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (**REFS**). 
+Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
 *Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
@@
 The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
-The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
-Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g.
+The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
+Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought resistance changes as forests age [e.g.
 @rodriguez-caton_long-term_2015].

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
Added sample text to final paragraph and added ref #120

@@
 The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
+**This has potentially far-reaching implications for future forests.
+If we accept that climate change will cause more intense droughts, it is possible we will see a shift from mature forests with buffering, tall trees to forests with a shorter overall height (e.g.** see @mcdowell_pervasive_2020).
+**At this point, species that had high resistance due to buffering (such as F.
+grandifolia) may in turn become more susceptible.
+This shift, in addition to the susceptiblity of large trees to fire and pests/pathogens, could be highly detrimental to temperate forests.
+That being said,**
 Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought resistance changes as forests age [e.g.

### 2020-07-24 manuscript/response_to_reviews.Rmd
Updated per #117

@@
 
-(*insert text*)
+**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+The two metrics were strongly correlated (Fig.
+**S5**).
+Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**
 
@@
 
-*What a nice review! We can easily add some language on the importance. We should also add citations to many of the studies mentioned.*
-
 first paragraph—this setup is very compelling, and the point is urgently true.
@@
 
-*We now mention this: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts (**REF**)."*
+**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].**
 
@@
 
-*We've ensured that language at the start of the paragraph refers to growth: "Many studies have shown that within and across species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; **REF**].
-"*
+**We've ensured that language at the start of the paragraph refers to growth: "Many studies have shown that within and across species, large trees tend to be more affected by drought.
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020]."**
 
@@
 
-**We have considerably resivesd this paragraph. It now reads as follows: **
-
-*insert text*
+**We have considerably revised this paragraph.
+It now reads as follows: "...First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance.
+We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure.
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
+(2015).
+We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites.
+Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance.
+We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
+We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance."**
 
@@
 
-**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context (*Wheeler et al.
-2007, Olson et al.
-2020*) and does not resolve differences among the many species within each category."**
+**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007, @olson_xylem_2020] and does not resolve differences among the many species within each category."**
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
revamped last paragraph
#120

--- 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
@@
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts.
-Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
+Because the resistance and resilience (i.e., post-drought recovery) of growth to drought are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
 However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.
@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
-We show that, at least within the mature forest studied here, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system, even in drier microenvironments.
-This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
-The same may not be true of systems where short trees exist outside of a buffered understory environment--*i.e.*, open grown trees or short-statured, early-successional forests.
-The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
-The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment.
-Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought resistance changes as forests age [e.g.
-@rodriguez-caton_long-term_2015].
-In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
+Our results show that taller, more exposed trees and species with less drought tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
+Resilience and survival are both linked to resistance (**REFS**), implying that the same factors may influence these. 
+Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].
+As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].
+At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F. grandifolia*) could in turn become more susceptible. 
+Here, the relative importance of tree height *per se* versus crown exposure becomes critical, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
+Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 
+Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature (**REFS**) and be dominated by species with more drought-tolerant traits (**REFS**).
+Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought.
 

### 2020-07-24 manuscript/response_to_reviews.Rmd
Fixed references and slightly edited final paragraph #117; #120

@@
 **We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-While tree size can correlate with the depth of water extraction [@brum_2018_hydrological], the linkage is not consistent.
-Shorter trees can vary broadly in the depth of water uptake [@stahl_2013_depth], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
-Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018]."**
+While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent.
+Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
+Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitratarak_roots_2018]."**
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
Fixed references and slightly edited final paragraph #117; #120

@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Our results show that taller, more exposed trees and species with less drought tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
+Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
 Resilience and survival are both linked to resistance (**REFS**), implying that the same factors may influence these. 
@@
 At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F. grandifolia*) could in turn become more susceptible. 
-Here, the relative importance of tree height *per se* versus crown exposure becomes critical, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
+Here, the relative importance of tree height *per se* versus crown exposure becomes crucial, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
 Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 

### 2020-07-24 manuscript/response_to_reviews.Rmd
Updated responses per issue #117

@@
 2).
-This change was prompted by comments by two reviewers (**R1, R3**) that canopy position could not be reliably projected back to 1966 or disentagled from height.
+This change was prompted by comments by two reviewers (**R1, R3**) that canopy position could not be reliably projected back to 1966 or disentangled from height.
 
@@
 
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
 **To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA).
@@
 While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) [DeSoto et al.
-(2020, Nature)](https://www.nature.com/articles/s41467-020-14300-5) used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
+(2) [@desoto_low_2020] used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
 In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
@@
 
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics."**
+**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics."**
 
@@
 
-[*issue #94*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/94)
+**We have run deviance-anova tests and can be seen in Figs. 3 and S4**
 
@@
 **We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought (Bennett et al.
-2015).
+1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**
@@
 We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing this depiction of height profiles for a temperate forest.
-(Such patterns are not shown in some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*.
+(Such patterns are not shown in some relevant textbooks: Gordon Bonan's 2016 *Ecological Climatology* [@bonan_ecological_2015] and Norman & Campbell's *Environmental Biophysics* [@campbell_introduction_1998].
 If we're missing studies that show such vertical profiles, we'd appreciate the references.)**
@@
 Within the broad pattern of global forest decline with climate change induced drought is the particularly mystifying and horrific preferential loss of the largest individuals all over the world.
-In addition to their esthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.
+In addition to their aesthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.
 
@@
 **Thank you for this interesting information.
-We've modified the sentence to read, "Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms (Zak et al.
-2010, Olson et al.
-2014,2018).**
+We've modified the sentence to read, "Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].**
 
@@
 We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure.
-We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
-(2015).
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites.
@@
 (2) We modified the criteria of selecting candidate variables for the full models to include the criterion that directions of responses must be consistent across droughts for inclusion in the full model.
-This resulted in exclusion of xylem porosity (Table S4)--a change that is also warrented by the fact that we only have two diffuse-porous species.
-(3) We added the following content to the discussion: **
-
-*insert text*
+This resulted in exclusion of xylem porosity (Table S4)--a change that is also warranted by the fact that we only have two diffuse-porous species.
+(3) We added the following content to the discussion: "Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
+diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$)...Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+**3**).
+While the low $Rt$ of *L.
+tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F.
+grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
+There are two potential explanations for this discrepancy.
+First, other traits can and do override the influence of xylem porosity on drought resistance.
+Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]." **
 
@@
 
-*This may be because moister habitats would tend to support species and individuals with more mesophytic traits (*Bartlett et al.
-2016 Drought.
-DOI: 10.1890/15-0468.1*; *Mencuccini 2003, DOI: 10.1046/j.1365-3040.2003.00991.x*) [@medeiros_extensive_2019], potentially growing to greater heights (e.g., *Detto et al.
-DOI: 10.1371/journal.pone.0076296*), and these are then more vulnerable when drought hits.*
+**This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.**
 
@@
 However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores (Helcoski et al.
-2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+Second, quoting our previous publication using these cores [@helcoski_growing_2019]: " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
 
@@
 
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
 **We do not agree that the $Rt$ metric "assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time." If this was true, $Rt$ would be calculated by dividing growth (BAI) in the drought year by the long-term mean growth increment.
@@
 
- *[DeSoto et al. (2020)](https://www.nature.com/articles/s41467-020-14300-5)*
-
 https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0
-
-*Liu et al. 2019; DOI: 10.1038/s41558-019-0583-9*
 
@@
 
-*Because the resistance and resilience of growth to drought is linked to their probability of surviving drought (*[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)*; *Liu et al.
-2019; DOI: 10.1038/s41558-019-0583-9*), understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.*
+**"Because the resistance and resilience of growth to drought is linked to their probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality."**
 
@@
 
-*insert 2 paragraphs on traits in intro*
+**"Many studies have shown that within and across species, large trees tend to be more affected by drought.
+Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020].
+It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response.
+First, tree height itself may be a primary driver.
+Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].
+Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
+Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]."**
+
+**"...Debates have also arisen regarding the traits influencing tree growth responses to drought.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
+Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
+However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent.
+For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
+Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]."**
+
+**"In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought.
+For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
+However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
+These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**
 
@@
 
-[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)
-
-*add specifics (changes to text; SI table/fig #s) to text below*
-
 **In response to Reviewer 4's comments, we tried an intervention time series analysis using the 'auto.arima' function from the 'forecast' package [@R-forecast].
@@
 
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of *[DeSoto et al.
-(2020)](https://www.nature.com/articles/s41467-020-14300-5)* suggest that $Rt$ is a more important drought response metric for angiosperms.
+**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms.
 "**
@@
 We also determined that there is little systematic variation in $PLA_{dry}$ and $\pi_{tlp}$ with respect to $H$ or $TWI$ (analysis not shown).
-This, together with the fact that our statistical models do account for $TWI$, should minimize bias related to distribution of the study species within the plot**
+This, together with the fact that our statistical models do account for $TWI$, should minimize bias related to distribution of the study species within the plot.**
 
@@
 
-*https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/124*
-
 L97: ‘greater variation’ that what?
@@
 The text here now reads, "More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; *Zhu et al.
-2018*].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; *Zhu et al.
-2018*; but see *Farrell et al.
-2017*].
+These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
+Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].
 The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
@@
 In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree.
-These factors contribute to the low R^2^, but do not negate the significance of our findings.**
+These factors contribute to the low R2, but do not negate the significance of our findings.**
 

### 2020-07-24 manuscript/tables&figures.Rmd

@@
 options(knitr.table.format = "latex")
-```
-
-```{r include=FALSE}
-# automatically create a bib database for R packages
-knitr::write_bib(c(
-  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'lme4', 'AICcmodavg', 'car', 'piecewiseSEM', 'MuMIn', 'stringr', 'dplyr', 'ggplot2', 'devtools', 'rgdal', 'broom', 'sf', 'ggthemes', 'rgeos', 'RCurl', 'readxl', 'pointRes', 'dplR', 'data.table', 'tools', 'reshape2', 'kableExtra', 'raster', 'elevatr', 'dynatopmodel', 'grid', 'gridExtra', 'ggstance', 'Hmisc', 'magrittr', 'ggpubr'
-), 'packages.bib')
 ```

### 2020-07-24 manuscript/response_to_reviews.Rmd
Updated per #117

@@
 
-**We have run deviance-anova tests and can be seen in Figs. 3 and S4**
+**We implemented the suggestions by running both LL ratio tests and anova tests on the top models.
+They both gave similar ranking to $\Delta$AICc, thus we continued with our original route.
+However, to give a more comprehensive view of the covariates, we now show significant differences between species compared to $Rt$ (for each drought scenario; Fig.
+3), as well as significant differences between species for each individual trait tested in the multivariate models (Fig.
+S4).**
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd

@@
 Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].
+Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
 Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015] and would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].

### 2020-07-24 manuscript/response_to_reviews.Rmd
drafted letter to Editor
#117

@@
 
-... Major changes included the following:
+We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest” (previously “Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest”), according to the recommendations of reviewers.
+This has resulted in an easier to read manuscript with more robust support.
+
+We addressed all of the reviewers’ points, as detailed below, resulting in several major changes:
 
 * At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years).
-Results were similar, and we now present that as a secondary model (mostly in SI).
+Results were similar, and conclusions the same.
+We now present those results as a secondary model (mostly in SI).
 
@@
 2).
-This change was prompted by comments by two reviewers (**R1, R3**) that canopy position could not be reliably projected back to 1966 or disentangled from height.
+This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966 or disentangled from height.
 
-* We added two figures: one showing Rt differences across species, and one visualizing the model results  (the latter is not yet right, as gg-plot is being difficult, but will eventually match  the model results.)
+* We shifted to more figures and fewer tables in the main article (as requested by R1, R3, and R4).
+We added two figures to the main text: one showing Rt differences across species (Fig.
+3) and one visualizing the model results (Fig.
+4).
+We moved former tables 4 and 5 to the SI.
+We also added several new figures and tables to the SI to provide information requested by the reviewers.
 
-* We moved former tables 4 and 5 to the SI.
+We hope that this manuscript will now be found suitable for publication in New Phytologist, and we look forward to your response.
 
-The result is an easier to read paper with more robust support.
+Best regards,
+
+Kristina Anderson-Teixeira (on behalf of all authors)
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
Updated with ian final pass #117

@@
 We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
-We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al.
-(2015).
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
@@
 We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$).
-Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapour pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
+Briefly, two 4 mm diameter leaf discs were cut from each leaf, tightly wrapped in foil, submerged in liquid nitrogen, perforated 10-15 times with a dissection needle, and then measured using a vapor pressure osmometer (VAPRO 5520, Wescor, Logan, UT, USA).
 Osmotic potential ($\pi_{osm}$) given by the osmometer was used to estimate ($\pi_{tlp}$) using the equation $\pi_{tlp}=0.832 \pi_{osm} ^{-0.631}$ [@bartlett_rapid_2012].
@@
 Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
-In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
+In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of @desoto_low_2020 suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
 
@@
 In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.
+Visualization of the best mixed effects model per drought scenario (Fig.
+**4**) was created by the visreg package [@R-visreg].
 
@@
 Species, as a factor in ANOVA, had significant influence (p<0.05) on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
-**S4**).
+**S4** as characterized by the agricolae package @R-agricolae).
 Drought resistance also varied across species, overall and in each drought year (Fig.

### 2020-07-24 manuscript/response_to_reviews.Rmd
worked on response to reviews
(also found refs for final paragraph, quoted here)

@@
 ```
-
-*italics - temporary comments for points that remain to be addressed*
-
-**bold - response**
-
-~~~~
 
@@
 
-**Thank you for this wonderful review.**
+**Thank you for this wonderful review.
+We have revised the final paragraph to put more emphasis on the importance and discuss implications for forests under a changing climate.
+This paragraph now reads:**
+
+**"As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
+Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected, at least in terms of growth during the drought year.
+Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these.
+Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].
+As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].
+At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F.
+grandifolia*) could in turn become more susceptible.
+Here, the relative importance of tree height *per se* versus crown exposure becomes crucial, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
+Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question.
+Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature and be dominated by species with more drought-tolerant traits [@liu_hydraulic_2019; @bartlett_correlations_2016; @zhu_leaf_2018].
+Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought."**
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
added refs to final paragraph
closes #120

--- 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
@@
 Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
-Resilience and survival are both linked to resistance (**REFS**), implying that the same factors may influence these. 
+Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these. 
 Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].
@@
 Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 
-Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature (**REFS**) and be dominated by species with more drought-tolerant traits (**REFS**).
+Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature and be dominated by species with more drought-tolerant traits [@liu_hydraulic_2019; @bartlett_correlations_2016; @zhu_leaf_2018].
 Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought.

### 2020-07-24 manuscript/response_to_reviews.Rmd
final check of reviewer responses (up to line 268 in .rmd file)

@@
 2).
-This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966 or disentangled from height.
+This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966, and our own conclusion that it could not be reliably disentangled from height.
 
@@
 
-We hope that this manuscript will now be found suitable for publication in New Phytologist, and we look forward to your response.
+We hope that this manuscript will now be found suitable for publication in *New Phytologist*, and we look forward to your response.
 
@@
 **To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA).
-Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
+Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4, and in the revised manuscript (Fig.
+S5).
 In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
@@
 While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) [@desoto_low_2020] used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
+(2) @desoto_low_2020 used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
 In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
@@
 
-**We have also decreased the emphasis on differences across drought years, as follows:* **
+**We have also decreased the emphasis on differences across drought years, as follows: **
 
@@
 This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
-(Removed was ring porosity, which had opposite effects in different droughts.)**
+(Removed was xylem porosity, which had opposite effects in different droughts.)**
 
@@
 **We have moved this table to the SI.
-We modified the analysis approach such that separate models for individual variable tests (as previously show in this table) are now limited to the species traits.
+We modified the analysis approach such that separate models for individual variable tests (as previously shown in this table) are now limited to the species traits.
 While this table is (and was) the only display item (besides Table 1) giving these results, the results are either mostly negative (wood density, SLA, xylem porosity) or preliminary tests qualifying variables for inclusion in full model ($PLA_{dry}$ and $\pi_{tlp}$).**
@@
 
-**We have reworded this sentence to read, "While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of the prediction that dominant trees have the lowest $Rt$ (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative."**
+**Crown position, and this hypothesis, have been removed from the analysis of $Rt$.**
 
@@
 
-**Unless the reviewers and editor strongly disagree, we would request to keep this figure to clarify the main manuscript.
+**Unless the reviewers and editor strongly disagree, we would request to keep this figure in the main manuscript.
 We have revised the manuscript such that this figure is better integrated with the hypothesis testing (Table 1) and interpretation of results.
@@
 **We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here).
-We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019]."**
+We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought."**
 
@@
 
-**See above comment.
-We have reworded this sentence to read, "In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought."**
+**This text was removed, and we now clearly distinguish between drought resistance and vulnerablity to mortality throughout the manuscript.**
 
@@
 
-**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].**
+**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011]."**
 
@@
 
-**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007, @olson_xylem_2020] and does not resolve differences among the many species within each category."**
+**"Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015] and would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category."**
 
@@
 (3) We added the following content to the discussion: "Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$)...Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$)...Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
 **3**).

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
final check of reviewer responses (up to line 268 in .rmd file)

@@
 Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
-Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
 **3**).

### 2020-07-24 manuscript/response_to_reviews.Rmd
response to reviews--should be final

@@
 
-**This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.**
+**This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013], and these are then more vulnerable when drought hits.**
 
@@
 We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
-However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores [@helcoski_growing_2019]: " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+However, these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
+Second, quoting our previous publication using these cores [@helcoski_growing_2019]: "Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.com/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/results/live_vs_dead).".**
 
@@
 
-**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 The two metrics were strongly correlated (Fig.
@@
 
-**"Many studies have shown that within and across species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020].
-It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response.
-First, tree height itself may be a primary driver.
-Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].
-Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
-Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]."**
-
-**"...Debates have also arisen regarding the traits influencing tree growth responses to drought.
+**"Debates have also arisen regarding the traits influencing tree growth responses to drought.
 Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
@@
 
-*(just need to write a response here-- no further changes needed to manuscript)*
+**We have revised our method, and now present a single main model with all variables at once.
+However, we did conduct preliminary independent tests of each species trait to avoid over-fitting in the model.
+The most relevant text now reads:**
+
+**"To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
+These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
+Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
+We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models."**
 
@@
 
-**We have dropped the criteria of decline in growth as part of how we defined drought. This does not change the drought years that qualify for inclusion.**
+**We note that the intent was never to test for a decline in growth, but to examine how it varied across trees; hence, we do not see it as circular.
+However, we have dropped the criteria of decline in growth as part of how we defined drought.
+This does not change the drought years that qualify for inclusion.**
 
@@
 
-**We now define drought resistance in the first paragraph: " Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...".
+**We now define drought resistance in the first paragraph: "Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...".
 We then use this term when we wish to speak specifically about resistance.
@@
 However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores [@helcoski_growing_2019]: " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+Second, quoting our previous publication using these cores [@helcoski_growing_2019]: "Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
 

### 2020-07-24 manuscript/climate_sensitivity_SI.Rmd
checked SI

@@
 \listoffigures
+
+**Appendix S1. Further Package Citations**
+
 \newpage
@@
 $\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
-
-*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
 
@@
 \newpage
-## Further Package Citations
+## Appendix S1. Further Package Citations
 While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes.

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
checked SI

@@
 All analysis beyond basic data collection was performed using R version 3.6.2 [@R-base]. 
-Other R-packages used in analyses are listed in the Supplementary Information (*Appendix S1*).
+Other R-packages used in analyses are listed in the Supplementary Information (Appendix S1).
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
@@
 
-Table S1: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
+Table S1. Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
 
-Table S2: Species-specific bark thickness regression equations.
+Table S2. Species-specific bark thickness regression equations.
 
-Table S3: Species-specific height regression equations.
+Table S3. Species-specific height regression equations.
 
@@
 
-Figure S2: Map of ForestGEO plot showing topographic wetness index and location of cored trees
+Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees
 
-Figure S3: Distribution of reconstructed tree heights across drought years.
+Figure S3. Distribution of reconstructed tree heights across drought years.
 
@@
 Figure S6. Visualization of best model, with data, for all droughts combined.
+
+Appendix S1. Further Package Citations

### 2020-07-24 manuscript/tables&figures.Rmd
final check tables-figures
#117

@@
 
-*copy legends here at the end*
+**Figure 1.
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
+2014).
+Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Figure modified from Helcoski *et al.* (2019).
+Density plots (b) show the distribution of resistance values for each drought.
+
+**Figure 2.
+Contemporary height profiles in sun exposure and growing season microclimate under non-drought conditions.** Shown are average ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018.
+In these plots, heights are slightly offset for visualization purposes.
+Asterisks indicate signficant differences between the top and bottom of the height profile.
+Also shown is (d) tree heights by 2018 crown position, with letters indicating significance groupings.
+In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.
+
+**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.
+
+**Figure 4.
+Visualization of best models for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
+Model coefficients are given in Table S6.
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
shortened/ checked abstract

@@
 Summary|  	198 | 	No. of Tables | 	3  
-Introduction|	1,034 |	No of Supporting Information files | 	#
+Introduction|	1,034 |	No of Supporting Information files | 	14
 Materials and Methods |	1,945	||	 
@@
 ### Summary
-- As climate change is driving increased drought frequency and severity in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important.
-The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape tree growth responses during droughts.
+- As climate change drives increased drought in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important.
+The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape growth responses during droughts.
 
-- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in a broadleaf deciduous forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity in a broadleaf deciduous forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
 

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
manuscript final check

@@
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
-Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015].
+Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
 Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts.
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015] and would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015], would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
@@
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 The two metrics were strongly correlated (Fig.
@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
-This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.
+This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013, and these are then more vulnerable when drought hits.
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 
-Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature and be dominated by species with more drought-tolerant traits [@liu_hydraulic_2019; @bartlett_correlations_2016; @zhu_leaf_2018].
+Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will shift towards shorter stature and be dominated by species with more drought-tolerant traits [@liu_hydraulic_2019; @bartlett_correlations_2016; @zhu_leaf_2018].
 Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought.

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
word count, minor shortening in results
#117

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) |  5,365	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,490	 |	No. of figures | 	4 (all colour)
 Summary|  	198 | 	No. of Tables | 	3  
-Introduction|	1,034 |	No of Supporting Information files | 	14
-Materials and Methods |	1,945	||	 
-Results |	697	 ||
-Discussion |	1467	||
-Acknowledgements | 125	||
+Introduction|	1,264 |	No of Supporting Information files | 	14
+Materials and Methods |	2,228	||	 
+Results |	929	 ||
+Discussion |	1927	||
+Acknowledgements | 142	||
 
@@
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Figs. **4, S6**; Tables 1,S4-S7). 
-Both had consistent signs across all droughts, and their inclusion at least marginally improved the model ($\Delta$AICc > 1.0) for at least one of the three droughts (Table S4), qualifying them as candidate variables for the full model.
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Figs.
+**4, S6**; Tables 1,S4-S7).
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 

### 2020-08-31 manuscript/2020_4_30_NewPhyt_review1/response_to_reviews.Rmd
added decision letter, set up response to reviews

--- 2020-07-24 manuscript/response_to_reviews.Rmd
@@
 
-**This is an issue with rendering latex in R Markdown tables, and we have some bug that is preventing it from rendering correctly.
-We have noted the problem and will ensure that it's fixed before publication.
-We note that, for *New Phytologist*, SI materials would be transferred to the journal template rather than be in a file generated by R Markdown.**
+**Done.**
 

### 2020-08-31 manuscript/response_to_reviews.Rmd
added decision letter, set up response to reviews

--- 2020-07-24 manuscript/response_to_reviews.Rmd
@@
 We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest” (previously “Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest”), according to the recommendations of reviewers.
-This has resulted in an easier to read manuscript with more robust support.
 
-We addressed all of the reviewers’ points, as detailed below, resulting in several major changes:
-
-* At the request of 3 reviewers (R1,R3,R4), we tried an alternate metric of drought resistance (using an ARIMA model to project what growth would have been in drought year based on the trend over the past 10 years).
-Results were similar, and conclusions the same.
-We now present those results as a secondary model (mostly in SI).
-
-* We dropped crown position as a factor in the statistical models of drought resistance, and instead include it as part of an enhanced focus on the differences in microenviroment experienced by trees of different height (now in Table 1; Fig.
-2).
-This change was prompted by comments by two reviewers (R1, R3) that canopy position could not be reliably projected back to 1966, and our own conclusion that it could not be reliably disentangled from height.
-
-* We shifted to more figures and fewer tables in the main article (as requested by R1, R3, and R4).
-We added two figures to the main text: one showing Rt differences across species (Fig.
-3) and one visualizing the model results (Fig.
-4).
-We moved former tables 4 and 5 to the SI.
-We also added several new figures and tables to the SI to provide information requested by the reviewers.
+...
 
@@
 
+Decision: Accept subject to revision
+
 ## Referee: 1
@@
 
-The study addresses the relationship between drought resistance (Rt) calculated from radial growth, and different traits and microenvironmental characteristics on 12 tree species from one forest in the Eastern US.
-Both the dataset and the study are meaningful, and the relationship found between Rt and height of biological importance.
-However, I think there are a number of issues that need to be worked out.
+The authors have done a thorough job of addressing my concerns and I found the revised manuscript much improved. Well done!
 
-It is important to better discuss the proper use of Rt, particularly when comparing single drought years, to assess tree resistance to drought.
-This metric has been much used lately as one of the so-called resilience indices (I wonder why other Rs indices where not included in the manuscript).
-In the paper, the authors use Rt similarly to most studies in the literature, which is fine.
-Yet, Rt (and particularly the other resilience indices based on radial growth) does not only depend on the individual response to drought events, but also on stand dynamics around target trees.
-For this reason, if stands around trees are not comparable or show changes after some disturbance (e.g.
-the drought events analysed) differently, this may bias comparisons of Resilience indices (although it is true that Rt should be less influenced than the other indices).
-The problem is that there is not such stand competition data for old droughts.
-This does not invalidate the study but I think the authors should discuss the issue, particularly for the analysis and comparisons of single droughts.
-I would suggest including also other Resilience indices, not only Rt, to enrich the discussion.
-
-**To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA).
-Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4, and in the revised manuscript (Fig.
-S5).
-In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
-While this unquestionably leads to biased estimates for some trees, we would not expect it to bias our analysis; rather, it is one of the reasons that the amount of variation that can be explained by analyses such as these is relatively low.
-We now include the following content on ARIMA in the manuscript:**
-
-**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
-The two metrics were strongly correlated (Fig.
-**S5**).
-Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**
-
-**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
-While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) @desoto_low_2020 used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
-In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
-In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
-These findings would argue that resistance is a more important variable for angiosperms.
-(3) As the reviewer notes, resilience indices would probably tend to be more sensitive to a tree's neighborhood and prior conditions.
-We now state the logic for focusing on $Rt$ in the methods section:**
-
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics."**
-
-Statistical analysis: the analysis would be more precise if the authors used Generalized linear mixed models to avoid data transformation and take into account (with random effects, as they do for the linear mixed model, LMM, presented now) the influence of species in the analysis.
-Then I missed a further discussion on the role of individual species (e.g.
-see my comment below on Rt>1).
-
-**A GLMM would only avoid transformation of the dependent variable (Rt), but that was (and is) not transformed.
-We tried the analysis with generalized linear mixed models with a Gaussian distribution, and confirmed that this gave very similar results, so we are sticking with LMM.**
-
-I suggest to show some relationship, e.g. between Rt and traits in figures. The study only presents two figures and there is plenty of room to present results more clearly with extra figures.
-
-**We created two new figures: one showing how $Rt$ varied across species in each of the three droughts (Fig. 3) and one visualizing the best models (Fig. 4).**
-
-Only assessing models with AIC is not enough to prove sound relationships in multidimensional and complex data-models.
-LL ratio tests or deviance-anova tests can be used in the case of nested models with LMM or GLMM respectively.
-Along the text: dAIC is generally referred as $\Delta$AIC, please change accordingly.
-
-**We implemented the suggestions by running both LL ratio tests and anova tests on the top models.
-They both gave similar ranking to $\Delta$AICc, thus we continued with our original route.
-However, to give a more comprehensive view of the covariates, we now show significant differences between species compared to $Rt$ (for each drought scenario; Fig.
-3), as well as significant differences between species for each individual trait tested in the multivariate models (Fig.
-S4).**
-
-**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **
-
-Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included.
-And I suggest discussing only the model with all years together.
-This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation.
-This would help to ensure a better use of the Resilience indices.
-
-**In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig.
-4*) visualizing the results of the best model.
-We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**
-
-**We have also decreased the emphasis on differences across drought years, as follows: **
-
-**(1) Table 1 now shows just the results from the model with all drought years combined.**
-
-**(2) To make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before).
-This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$).
-(Removed was xylem porosity, which had opposite effects in different droughts.)**
-
-**(3) The results and discussion now focus more on the results from the three drought years combined.**
-
-**We do still present results from the three droughts separately (e.g., Figs. 1,3,4; Tables S4-S7), as we do feel that this makes the story richer.**
-
-The authors might like to consider including Table 4 as a supplement, because as it stands it looks like an exploratory analysis.
-If the ‘single-variable approach’ is not just exploratory, state it clearly and why it is needed besides the full models with several parameters.
-
-**We have moved this table to the SI.
-We modified the analysis approach such that separate models for individual variable tests (as previously shown in this table) are now limited to the species traits.
-While this table is (and was) the only display item (besides Table 1) giving these results, the results are either mostly negative (wood density, SLA, xylem porosity) or preliminary tests qualifying variables for inclusion in full model ($PLA_{dry}$ and $\pi_{tlp}$).**
-
-Importantly, the drought selection would be biased if years were selected as from line 202, which is a common mistake in similar studies (the original of Lloret et al.
-included).
-Droughts must be selected based only on some climatic/drought index (PDSI, SPEI,…) and independently of growth.
-Otherwise the analysis is biased towards years with low growth, and different questions would arise with a difficult answer.
-If the authors select only those droughts where there is a growth decrease: is that decrease evident in all species and similar across species?
-What happens with more intense droughts that those selected (if any), why they do not force a decrease in growth if drought is a main forcing agent?
-I see that 1966 and 1999 are the driest (as in PDSI) in Figure 1, but there is at least one year around 1955 that it is as dry or drier than 1977.
-In line 204 it is stated that they are the driest.
-If this is the case, the analysis should be sound but it should be deleted what is stated in lines 202-203 where authors say that they selected years also based on low growth.
-In fact, that should be a result after having selected the drought years only based on dryness (as in lines 208-209).
-Line 211 then they search for pointer years… a little bit messy.
-Please, explain better, and it should follow: 1) selection of dry years without regarding growth data; 2) characterise growth those dry years; 3) If the authors want to compare with other pointer years, fine, but that should be a third independent step properly justified and explained.
-Following this rationale, then in line 221 it should be difficult that Rt>1, had we selected those years with a negative growth anomaly.
-Please, explain clearly and consistently.
-
-**We have made the suggested change: we now present only the PDSI criteria in the methods, and mention the fact that the selected years were also the ones with largest growth reductions during the study period in the results section.
-Using just the PDSI criterion does not change the years analyzed.**
-
-Other minor comments:
-
-- Line 113: correct ‘positively’.
-
-**done**
-
-- Line 134: cambial growth also includes that of phloem; so better refer to ‘xylem growth’.
-
-**done**
-
-- Line 161: how many species were estimated for height like this? Wouldn’t it be possible for authors to go to the field and perform some height estimations? This 
-should be rather straightforward and the authors would increase accuracy across species in height estimation.
-
-**We have specified how many species used the interspecific allometry (n=2, JUNI and FRAM). Because of COVID-19, it would not be possible to go collect more data for these species.**
-
-- Lines 166-167: please explain further the hypothesis and why.
-
-**Crown position, and this hypothesis, have been removed from the analysis of $Rt$.**
-
-- Line 222: what about climate in years before 1991?
-
-**We no longer mention 1991, as it does not meet the PDSI criteria.**
-
-- Line 231: define AICc.
-
-**This is done (with added reference)**
-
-- Line 234-247: it is not clear what the authors did, please explain clearer this paragraph and the next and why you did two different (yet similar) analyses (single-variable and full model).
-
-**This section has been re-written.**
-
-- Line 241-242: enough to say once that models were selected based on AIC (line 237-238) and then one criterion, why dAIC<1 here and dAIC<2 before?
-The classic assumption is 2 (e.g.
-Burnham & Anderson 2004).
-See also my other comment above.
-
-**We have modified this text to read as follows:**
-
-**"Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to a corresponding null model lacking the trait improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table 4).
-We note that the $\Delta$AICc $\ge$ 1.0 criterion is not a test of significance, but of whether the variable has enough influence to be considered as a *candidate* variable in full models."**
-
-- Line 255-256: but this was already an assumption in M-M (line 202-203). Please, see my comment below and be consistent along the text.
-
-**As stated above, we have removed the $Rt$ criterion from the drought identification process, and now present this information for the first time in this section, adding the following statement: "Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$."**
-
-- Line 258: so this means that they are no sensitive to the drought? Which trees, any species in particular? Is this a consequence of stand dynamics shadowing the impact of drought events?
-
-**It's not unprecedented to see some trees exhibit increased growth during drought, and not surprising.
-Stand dynamics and other factors result in a lot of scatter in the relationship between tree growth and climate, so while the distribution of $Rt$ in Fig.
-1b was shifted left (normal year should be normally distributed around 1), there still are some trees with increased growth.
-Our analysis helps to work out which trees these were (often shorter trees).**
-
-**We address this question in the discussion, where we have added a statement about how this is not unprecedented: "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
-1b), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation"**
-
-**We have also added a new figure (*Fig. 3*) showing how $Rt$ varied across species in each of the three droughts.**
-
-- Line 266-267: why only in 1966? I think this shows that this variable is not properly assessed. The older the date, the more accurate the estimation of crown position and height were.
-
-**We no longer attempt to use crown position to describe variation in drought resistance.
-Instead, we now use it to characterize *contemporary* differences in microenvironmental conditions faced by trees of different heights.**
-
-- Line 275-280: those profiles were assumed a priori in the introduction and are common to any forest.
-They are indeed interesting to characterise the forest, but they do not add to individual tree variability.
-You could think to send Figure 2 to an appendix, in consequence, and use figures to show more specific results related to your individual tree analyses.
-
-**Unless the reviewers and editor strongly disagree, we would request to keep this figure in the main manuscript.
-We have revised the manuscript such that this figure is better integrated with the hypothesis testing (Table 1) and interpretation of results.
-We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing this depiction of height profiles for a temperate forest.
-(Such patterns are not shown in some relevant textbooks: Gordon Bonan's 2016 *Ecological Climatology* [@bonan_ecological_2015] and Norman & Campbell's *Environmental Biophysics* [@campbell_introduction_1998].
-If we're missing studies that show such vertical profiles, we'd appreciate the references.)**
-
-**We did create two new figures showing the results of the tree-ring analysis (and moved two tables to the SI).**
-
-- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?
-
-**We have removed the claim that increased growth may be due to reduced leaf area.**
-
-**To address the question about species, we have added a new figure (*Fig. 3*) showing how $Rt$ varied across species in each of the three droughts.**
-
-- Line 331-333: not so sure about this.
-It can also mean that there is no causal relationship between Rt and covariates, mostly because radial growth Rt should not be used as an integral proxy of tree resistance to drought, such as it is often used now in the literature.
-
-**Our analysis of $Rt_{ARIMA}$ (described above) confirms the findings based on analysis of $Rt$.**
-
-- Line 334-339: indeed, assessment of these two variables is not conclusive, as suggested above (just data from 2018, from the past estimated and correlated).
-
-**Please see response above.**
-
-- Line 348: non-drought.
-
-**Fixed.**
-
-- Line 349: do authors really think that lower radial growth as from Rt on a specific year means that the tree is more vulnerable? I think this is over conclusive.
-
-**We did not intend the meaning here as vulnerability *to mortality* (assuming that's the objection here).
-We have reworded this sentence to read, "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought."**
-
-- Line 400: I do not agree we can directly link Rt with tree vulnerability to drought; we might be currently overinterpreting these resilience indices extrapolating to physiological conclusions that do not necessary need to hold.
-
-**This text was removed, and we now clearly distinguish between drought resistance and vulnerablity to mortality throughout the manuscript.**
-
-Table 1: do ‘-‘ signs in the table correspond to ‘no’ (non-significant)? Please explain.
-
-**This is now clarified in the table footnote.**
-
-Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.
-
-**In this study, we have one core per tree, now specified in the methods. We have changed the table header to "n trees".**
-
-Table 3: the authors could show some test to see if the traits presented are different among species.
-
-**We now show differences in traits among species (with significance groupings) in Figure S4.**
-
-Table S1 and S2: correct R2.
-
-**This is an issue with rendering latex in R Markdown tables, and we have some bug that is preventing it from rendering correctly.
-We have noted the problem and will ensure that it's fixed before publication.
-We note that, for *New Phytologist*, SI materials would be transferred to the journal template rather than be in a file generated by R Markdown.**
-
-Table S3: please, explain what ‘rank’ means.
-
-**Done.**
-
-Figure S3: I understand that authors only have height data from 2018 and height for the other dates looks trivial: indeed there will be an increase in height with an increase in DBH in time.
-The question is whether there was any stand dynamics involved in those height series that had to be included in the models.
-
-**We recognize the importance of stand dynamics, and have added specific mention to the discussion: "It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation,...
-"**
+**Thank you.**
 
@@
 
-This manuscript examines a truly urgent and important issue.
-The general issue is the clear vulnerability of forests to ongoing climate change induced drought worldwide.
-The issue is particularly vexing because, while it seems clear that plant hydraulics is involved in vulnerability to drought, the exact mechanisms connecting hydraulics to vulnerability are not clear, and just what indices biologists should use to gauge vulnerability is as a result unclear.
-Within the broad pattern of global forest decline with climate change induced drought is the particularly mystifying and horrific preferential loss of the largest individuals all over the world.
-In addition to their aesthetic appeal, large individuals make disproportionate contributions to forest function, making the issue examined in this ms extremely timely and important.
+Thank you very much for sending me this revised version of the ms by McGregor et al.
+It is a very important study with a clever and appropriate sampling design, and is exceptionally well written.
+I enjoyed reading it again and find it very valuable.
+Thanks very much to the authors for tweaking the text here and there and adding detail, which made the ms even more compelling.
+I had a few reactions that might be useful:
 
-The issue of the preferential death of large trees is not only important because of its effects on forest function and culture.
-It is important because it seems to require novel theory.
-Under current expectations, the largest individuals, with their greater root reach, etc., should be more rather than less resistant.
-That they might not be demands a novel explanatory framework.
-Yet the very existence of the tendency is debated, with some studies reporting greater mortality among small rather than large individuals (Camarero et al., 2015; Colangelo et al., 2017).
-These studies, however, were not designed to take into account the microsite variation that presumably should lead to marked differences in survivorship even within the same height class with drought.
-One approach, which we have used (Soriano et al., in press; Olson et al., 2018) is to conduct experiments under controlled conditions to detect the effect of height, and our results are consistent with the greater vulnerability of taller individuals.
-However, under natural conditions, a study that takes this variation into account is needed to test the notion that taller individuals are preferentially vulnerable is needed, and this study provides exactly such a design.
+72 One suggestion is that it would be better not to say that larger individuals have xylem of “greater efficiency.” Instead, Red Queen like, conduits widen allowing whole-pathlength resistance to remain constant over much of the height range of trees.
+Presumably selection favors variants in which leaves maintain their productivity constant insofar as is possible as an individual grows taller, and resistance as constant as possible would contribute to this.
+So, taller trees don’t have more efficient conduits; they widen in a way that maintains the same whole-pathlength efficiency.
+Here you could say something like “taller trees have wider conduits, which help maintain constant the resistance that would otherwise increase as trees grow taller.”
 
-I found this paper not only to address a very important point but also to be exceptionally well structured and written.
-My main suggestion is that it could use a bit more emphasis on its importance, and I would be interested in hearing a bit more about what the authors feel their results imply regarding the cause of the preferential vulnerability of large individuals.
-I give line-by-line suggestions (and sometimes just reactions):
+392 “Constraint” is always such a vague term in biology (meaning everything from factors opposing selection to selection itself), with every biologist understanding something different.
+Replacing the term always seems to me to lead to increased clarity.
+Here, it could be removed entirely “previous findings that it is impossible for trees…”, I think profitably.
 
-**Thank you for this wonderful review.
-We have revised the final paragraph to put more emphasis on the importance and discuss implications for forests under a changing climate.
-This paragraph now reads:**
+395 if taller individuals are exposed to greater evaporative demand, then for a given photosynthetic productivity they would require greater conductance than they would need for a given height if they were sheltered.
+This would favor even wider conduits, making taller individuals even more vulnerable!
+You rightly highlight the need to separate height per se and canopy position.
 
-**"As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected, at least in terms of growth during the drought year.
-Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these.
-Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].
-As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].
-At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F.
-grandifolia*) could in turn become more susceptible.
-Here, the relative importance of tree height *per se* versus crown exposure becomes crucial, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
-Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question.
-Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature and be dominated by species with more drought-tolerant traits [@liu_hydraulic_2019; @bartlett_correlations_2016; @zhu_leaf_2018].
-Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought."**
+With regard to root mass, this might be useful: https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fdoi.org%2F10.1371%2Fjournal.pone.0086550&amp;data=02%7C01%7CTeixeiraK%40si.edu%7C04ca4a8deb534237b5a508d84dd84e13%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637344938706540440&amp;sdata=tiYufg0%2F%2BuRBThcGeUHuRjVDhWPmbK0xP%2Fb2moH%2F210%3D&amp;reserved=0 seem to show above-below ground biomass isometry.
 
-first paragraph—this setup is very compelling, and the point is urgently true.
+Thanks again and congratulations to the authors.
 
-**Thank you**
-
-66- “in their basal portions” seems fine for this level of intro, but note that in angiosperms, taller trees also have wider vessels at the stem tip (and thus throughout the conductive system) (Zach et al., 2010; Olson et al., 2014, 2018).
-This seems plausibly the result of selection keeping conduit carbon costs down with height growth.
-As a tree grows taller, if terminal twig vessel diameters become wider, then given a constant rate of tip-to-base vessel widening, vessel diameters will be wider along the entire stem.
-Vessels that are wider throughout would have lower per-conduit resistance than those that, all else being equal, are narrower.
-So, conceivably as a tree gets taller, if vessels widen at the twig tip, then lower per-conduit resistance could mean that the same leaf area is supplied by fewer vessels.
-This could potentially contribute to keeping per-leaf area carbon costs constant with height growth.
-This prediction seems confirmed by the experiment of (Echeverría et al., 2019), who found that as Moringa trees grew taller, there were wider vessels at the stem tip and fewer vessels per unit leaf area, but that conductance per unit leaf area should remain the same.
-Terminal twig tracheids don’t seem to widen in gymnosperms, e.g.
-(Williams et al., 2019).
-(Sack et al., 2012) have shown very clearly that petiole base vessel diameter scales predictably with leaf size, as would be expected; it is not clear how leaf size variation is involved in terminal twig vessel diameter scaling with height, though in wide samplings leaf size doesn’t seem to scale with height (Jensen & Zwieniecki, 2013).
-
-**Thank you for this interesting information.
-We've modified the sentence to read, "Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].**
-
-67 Appealing to the vulnerability-diameter link might get blowback from some readers because there is as yet no satisfactory theory linking vulnerability to diameter in an entirely satisfactory way, so maybe just say “wider conduits are likely” or that they are “plausibly” or “possibly” more vulnerable, or that many “data are consistent” with such a link.
-My view is that there very well might not be a link, but in the absence of a link between vulnerability and conduit diameter, it is very hard to explain a very large number of well-established observations.
-These include the following: why growth rings always proceed from wide to narrow conduits as drought ensues; why maximum community conduit diameter is narrower in drylands than in wet places; why lianas have wider vessel diameter variances (i.e.
-not only wider but also numerous narrower vessels) than self-supporting plants (Rosell & Olson, 2014); why plants virtually always produce vessels narrower than the maximum possible; when internal helical sculpture in conduits varies in a growth ring it is always associated with narrow latewood vessels; when simple and scalariform plates vary across growth rings sclariform plates are always found in the narrow latewood vessels; and many other observations.
-If there is no link between drought induced embolism vulnerability and conduit diameter, and there might not be, all of these observations, which are currently simultaneously and parsimoniously explained by the link, will require alternative explanations.
-As a result, while there is as yet no well accepted mechanistic link between vulnerability and diameter, it seems to me irresponsible to reject the link without providing a satisfactory and similarly parsimonious and general replacement explanation for all of these phenomena.
-So, I think the authors are justified in cautiously invoking the link here.
-
-**We've added the word "plausibly". We agree that this linkage makes a lot of sense.**
-
-69 this is a very important point—microsite is crucial both for the ultimate height an individual grows to as well as its vulnerability.
-The generalization that on a global level large trees seem disproportionately vulnerable is often greeted with the skeptical observation that in specific cases a small tree died next to a large one that survived.
-However, the prediction is that *all else being equal* large individuals should be more vulnerable than smaller *conspecifics*.
-There is so much variation in microsite and other factors that the greater vulnerability of larger individuals should emerge clearly only in carefully controlled experiments where all else is indeed (virtually) equal, or with high sample sizes where variation in microsite etc should random out and the signal of generally higher vulnerability should come through.
-
-**Thank you.**
-
-73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.
-
-**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
-While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent.
-Shorter trees can vary broadly in the depth of water uptake [@stahl_depth_2013], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999].
-Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitratarak_roots_2018]."**
-
-73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.
-
-**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011]."**
-
-77 In the concluding sentence, “greater growth reductions” came as a surprise because I thought that the focus was greater vulnerability/ mortality.
-Also, there are convincing theoretical and empirical reasons to think that large trees don’t slow down e.g.
-(Sillett et al., 2010; Selaya & Anten, 2010; Michaletz et al., 2014).
-
-**We've ensured that language at the start of the paragraph refers to growth: "Many studies have shown that within and across species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020]."**
-
-**Regarding the issue of whether large trees slow down or not as they age, that's another (very interesting) issue.
-The cited references apply to non-drought situations, and I'd see drought response versus age-related growth patterns as largely independent.
-In this study, it shouldn't matter because our resistance metric takes the ratio of growth during drought relative to what would be expected based on recent years.**
-
-96 this all seems compelling.
-To make the point stronger, either justify that the leaf traits are causally more important, or that, while they make up part of a unified hydraulic system all of whose components are important, leaf traits are simply easier to measure and this justifies their study.
-The latter position seems most plausible to me.
-
-**We have revised this paragraph to make the latter point: "Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016]."**
-
-102 great questions, though within-species comparisons are likely to be most informative here, and your mixed model design is in accord with this; just being cautious because sometimes it’s seductive to regard selection acting between rather than within species
-
-**Thank you.**
-
-105 this sounds great, but was a little unclear.
-When you say “first, we test…” I thought that you would examine a relationship between microsite and some tree characteristic, e.g.
-that the largest trees are in moist microsites.
-But the next sentence didn’t refer to a tree-microsite correlation but vulnerability-size.
-It would be good to clarify the prose here because these are very important passages.
-The third sentence did seem to be about tree-microsite correlations.
-
-**We have considerably revised this paragraph.
-It now reads as follows: "...First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance.
-We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure.
-We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
-We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites.
-Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance.
-We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
-We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance."**
-
-112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from.
-Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007).
-They are virtually absent in drylands, especially warm tropical drylands.
-Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).
-
-**This is a great point. The hypothesis refers to temperate deciduous biome, which we now clarify in the sentence in question:**
-
-**"Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015] and would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category."**
-
-**We dug into the ring/diffuse porous distinction more, and ended up making a few substantive changes that relate to this theme: (1) We added a new figure showing $Rt$ across species (Fig.
-3).
-This illustrates that the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum.
-(2) We modified the criteria of selecting candidate variables for the full models to include the criterion that directions of responses must be consistent across droughts for inclusion in the full model.
-This resulted in exclusion of xylem porosity (Table S4)--a change that is also warranted by the fact that we only have two diffuse-porous species.
-(3) We added the following content to the discussion: "Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
-diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$)...Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-**3**).
-While the low $Rt$ of *L.
-tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F.
-grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
-There are two potential explanations for this discrepancy.
-First, other traits can and do override the influence of xylem porosity on drought resistance.
-Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]." **
-
-113 These correlations seem reasonable, e.g. see (Méndez-Alonzo et al., 2012)
-
-**Thank you.**
-
-124 and Table 3.
-This seems like a good palette of species because they are very different in their hydraulic construction.
-In addition to libriform fibers, oaks have very abundant vasicentric tracheids, especially in latewood.
-This means that the oaks will likely maintain some conductance even under highly negative tensions.
-This is not by virtue of their ring porosity (note that most oaks globally aren’t ring porous!), but because of the high proportion of vasicentric tracheids in their latewood.
-Liriodendron (diffuse porous, scalariform perforation plates) and Carya (likely ring porous at your locality; simple perforation plates) both have libriform fibers and lack vasicentric tracheids.
-See the data in (Olson et al., in press).
-There is relatively low family-level diversity represented here; our phylogenetic analyses (Olson et al., in press) suggest that there is very little phylogenetic signal in the conduit diameter-stem length relationship, so this could help justify if necessary not including phylogeny in the analyses.
-
-**Thank you.**
-
-127 “tree size”= height?
-
-**Yes; clarified.**
-
-138 you might discuss the adequacy of the height range studied.
-If it is true that conduit diameter predicts vulnerability, then the greatest differences in vulnerability should be over the height range over which conduit diameter changes most rapidly.
-The relationship between conduit diameter and plant height turns out to follow a power-law like distribution, see for example the SI in (Olson et al., 2018).
-This means that differences in vulnerability should be very marked across smaller size classes (say below 7 m or so and very marked below 3 m) and be less and less visible across larger plants.
-So, studies of changes in conduit diameter/ vulnerability with height would ideally include plants smaller than those included here, so maybe comment on how the height range might affect the ability to detect the patterns of interest.
-It strikes me that it is appropriate because it is the height range over which the putative preferential susceptibility of large trees to drought is being observed.
-
-**Our trees were mostly above the 3 or 7m threshold (see Fig.
-S3).
-It's important to note that the DBH and height at the time of coring did not correspond to that at the time of drought.
-Still, the minimum height was about 5m, so we'd presumably be mostly missing that range of most dramatic changes in conduit diameter.
-It could be interesting to try to align the size range over which Rt changes most dramatically with those over which conduit diameter and microclimate change most rapidly, but unfortunately that's beyond the scope here.**
-
-173 TWI seems like an appropriate index for characterizing microsite, with the expectation that individuals in moister areas should be more buffered from drought.
-
-**Thank you.**
-
-209 It couldn’t hurt to define basal area increment
-
-**Done.**
-
-227 the nesting seems appropriate and sufficient for taking species/ “phylogeny” into account
-
-**Thank you.**
-
-260 amazing
-
-**Thank you.**
-
-275—some people think that the increase in terminal twig vessels with height growth keeps pace with the changes in VPD from understory to canopy
-
-**Interesting, thanks.**
-
-281 nice
-
-**Thank you.**
-
-301 this intro paragraph to the discussion ranks very well the results from most significant to least
-
-**Thank you.**
-
-327 I agree that this is the point that most deserves discussion in paragraph 2
-
-**Thank you.**
-
-334 You might go into detail regarding what you think this says about the causal hypothesis behind the global decline of large individuals.
-To me it helps reject notions that somehow larger trees are more exposed to wind throw or other damaging agents.
-It is consistent with the hypothesis that larger individuals have wider conduits, and (assuming that wider conduits are more vulnerable to drought induced embolism) thus larger individuals are (all else being equal…) more vulnerable to embolism than smaller conspecifics.
-
-**Unfortunately, we ended up concluding that we can't reliably parse the effects of height *per se* and canopy position in this forest, due to collinearity of height and canopy position, and also because of increasing uncertainty in canopy position as we go back in time.
-Thus, the paragraph is now modified to discuss both height and associated microenvironment.**
-
-336 beautifully said “impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism”
-
-**Thank you.**
-
-349 fascinating; it seems plausible that current climate changed induced large-individual mortality involves trees growing to conduit diameters, and thus heights, permitted under previously more even and moister climates, and now finding themselves in more extreme climates that make their heights and conduits excessively wide given current conditions.
-
-**We have added the following text:**
-
-**This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights (e.g., @detto_hydrological_2013), and these are then more vulnerable when drought hits.**
-
-350 if you wanted to expand your discussion of causal mechanisms or the implications of your findings, I think you could jettison this paragraph.
-I admire your thoroughness in offering caveats, but I agree that this does not seem like a serious problem for your study.
-
-**We have changed our treatment of canopy position and removed this paragraph.**
-
-382 an effort that seems in harmony with these concerns is (Christoffersen et al., 2016)
-
-**Agreed. We have added this citation.**
-
-385 very nice paragraph
-
-**Thank you.**
-
-In conclusion, this ms addresses a very important issue with a well-designed study and appropriate analyses. It is well written and interesting and I hope my comments are useful.
-
-**Thank you.**
-
-Mark Olson molson@ib.unam.mx
-
-Camarero JJ, Gazol A, Sangüesa-Barreda G, Oliva J, Vicente-Serrano SM.
-2015.
-To die or not to die: early warnings of tree dieback in response to a severe drought (D Gibson, Ed.).
-Journal of Ecology 103: 44–57.
-
-Christoffersen BO, Gloor M, Fauset S, Fyllas NM, Galbraith DR, Baker TR, Kruijt B, Rowland L, Fisher RA, Binks OJ, et al.
-2016.
-Linking hydraulic traits to tropical forest function in a size-structured and trait-driven model (TFS v.1-Hydro).
-Geoscientific Model Development 9: 4227–4255.
-
-Colangelo M, Camarero JJ, Borghetti M, Gazol A, Gentilesca T, Ripullone F.
-2017.
-Size Matters a Lot: Drought-Affected Italian Oaks Are Smaller and Show Lower Growth Prior to Tree Death.
-Frontiers in Plant Science 8.
-
-Echeverría A, Anfodillo T, Soriano D, Rosell JA, Olson ME.
-2019.
-Constant theoretical conductance, changes in vessel diameter and number with height growth in Moringa oleifera.
-Journal of Experimental Botany.
-
-Jensen KH, Zwieniecki MA. 2013. Physical Limits to Leaf Size in Tall Trees. Physical Review Letters 110.
-
-Méndez-Alonzo R, Paz H, Zuluaga RC, Rosell JA, Olson ME. 2012. Coordinated evolution of leaf and stem economics in tropical dry forest trees. Ecology 93: 2397–2406.
-
-Michaletz ST, Cheng D, Kerkhoff AJ, Enquist BJ. 2014. Convergence of terrestrial plant production across global climate gradients. Nature 512: 39–43.
-
-Olson ME, Anfodillo T, Rosell JA, Petit G, Crivellaro A, Isnard S, León-Gómez C, Alvarado-Cárdenas LO, Castorena M.
-2014.
-Universal hydraulics of the flowering plants: vessel diameter scales with stem length across angiosperm lineages, habits and climates.
-Ecology Letters 17: 988–997.
-
-Olson ME, Rosell JA, Martínez-Pérez C, León-Gómez C, Fajardo A, Isnard S, Cervantes-Alcayde M-A, Echeverría A, Figueroa-Abúndiz, Segovia-Rivas A, et al.
-in press.
-Xylem vessel diameter-shoot length scaling: ecological significance of porosity types and other traits.
-Ecological Monographs.
-
-Olson ME, Soriano D, Rosell JA, Anfodillo T, Donoghue MJ, Edwards EJ, León-Gómez C, Dawson T, Camarero Martínez JJ, Castorena M, et al.
-2018.
-Plant height and hydraulic vulnerability to drought and cold.
-Proceedings of the National Academy of Sciences 115: 7551–7556.
-
-Rosell JA, Olson ME. 2014. Do lianas really have wide vessels? Vessel diameter–stem length scaling in non-self-supporting plants. Perspectives in Plant Ecology, Evolution and Systematics 16: 288–295.
-
-Sack L, Scoffoni C, McKown AD, Rawls M, Havran JC, Tran H, Tran T.
-2012.
-Developmentally based scaling of leaf venation architecture explains global ecological patterns.
-Nature Communications 3: ncomms1835.
-
-Selaya NG, Anten NPR. 2010. Leaves of pioneer and later-successional trees have similar lifetime carbon gain in tropical secondary forest. Ecology 91: 1102–1113.
-
-Sillett SC, Van Pelt R, Koch GW, Ambrose AR, Carroll AL, Antoine ME, Mifsud BM. 2010. Increasing wood production through old age in tall trees. Forest Ecology and Management 259: 976–994.
-
-Soriano D, Echeverría A, Anfodillo T, Rosell JA, Olson ME. in press. Hydraulic traits vary as the result of tip-to-base conduit widening in vascular plants. Journal of Experimental Botany.
-
-Wheeler EA, Baas P, Rodgers S. 2007. Variations In Dicot Wood Anatomy: A Global Analysis Based on the Insidewood Database. IAWA Journal 28: 229–258.
-
-Williams CB, Anfodillo T, Crivellaro A, Lazzarin M, Dawson TE, Koch GW. 2019. Axial variation of xylem conduits in the Earth’s tallest trees. Trees 33: 1299–1311.
-
-Zach A, Schuldt B, Brix S, Culmsee H, Culmsee H.
-2010.
-Vessel diameter and xylem hydraulic conductivity increase with tree height in tropical rainforest trees in Sulawesi, Indonesia.
-Flora 205: 506–512.
+Mark Olson
 
@@
 
-McGregor and colleagues use a combination of long-term tree records and functional trait data to explore the key drivers that shape the growth resistance of different temperate tree species to extreme drought.
-They find that that leaf hydraulic traits and tree size are strong predictors of growth responses to drought – with taller trees most affected.
+This is a resubmission of a previously revised manuscript.
+I have enjoyed reading the manuscript.
+The paper is well written and the methods are clearly presented.
+The hypotheses are sound and well framed within the physiological implications of tree height.
 
-I enjoyed the paper very much.
-It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology.
-I could definitely see the paper being of interest to a broad readership.
-However, I do have some important questions and suggestions about the analysis which I think need more careful thought.
-Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.
+The authors have satisfactorily answered most previous comments.
+For instance, I appreciated the discussion on the two indices Rt and Rtarima.
+I agree that results comparing them are pretty similar (Table S6 vs Table S7) and that authors can use Rt.
 
-Main comments
+Yet, I still believe there are a couple of details that need revision:
 
-L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!).
-But doesn’t this introduce a bias in your analysis?
-Trees that died may have had very different responses to drought that those which survived.
-Do you get similar results if you analyse trees that lived and died separately?
+1) I still think that adding the same analysis on the two other resilience indices (resilience and recovery) from Lloret et al.
+(2011) would increase the interest of the manuscript (as the authors acknowledge in line 476).
+Using DeSoto et al.
+2020 as a justification is not enough.
+There are potential caveats in that paper and even if we accepted the general trends in growth resilience proposed in that study, studies like that of McGregor et al.
+apply at a different and more detailed scale, where those global relationships might not hold.
+This adds interest to the present ms, but for that same reason most readers will like to see a similar discussion on the other complementary growth resilience indices together in the same paper.
 
-**We agree that this could be a concern, but do not believe it is applicable here, for two reasons.
-First, this would be more likely to introduce a bias if the trees cored dead were killed by drought.
-We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
-However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores [@helcoski_growing_2019]: " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
+2) In Line 263 it is stated that both marginal and conditional R2 are reported, but then I could only find conditional R2s in Table S6, S7.
+Please, report also marginal R2 in those Tables to show how much variability it is explained by the fixed effects (i.e.
+the traits and independent variables in Table 3).
+This is important so we can judge how robust the tested covariates are.
+I can accept what authors argue in their response (‘Tree-ring data are inherently noisy, particularly in more mesic forests.
+In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree.
+These factors contribute to the low R2, but do not negate the significance of our findings’).
+However, still I think that, even if (or particularly because) we see many papers today with very low variability explained (see for instance some of the analyses in the previously mentioned DeSoto et al.
+2020), we should demand a minimum level of explanatory power in models in ecophysiological papers.
+Otherwise we cannot assure that inferences are robust enough and relationships worth to be discussed.
 
-**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**
+Other comments:
 
-L165-167: I think this needs to be explained more clearly.
-Why are type I errors unlikely and type II error possible?
-I also think there are real issue with the analysis of crown position.
-Presumably on average the crown illumination index of trees would have increased over the past 50 or so years (from suppressed to dominant) as trees grew and neighbouring trees died.
-So I’m not convinced you can include this in your models going back half a century.
-It’s just not sensible to assume that crown position would not have changed over such a long time period.
+- As mentioned on a previous comment, in the manuscript the authors sampled dead and live trees, but then I did not see any further mentioning to this later in the manuscript (lines 175-178 and Table 2).
+This would merit some discussion and further analysis, how many trees were died and how many trees were alive of those in Table 2?
 
-**We no longer attempt to use crown position to describe variation in drought resistance.
-Instead, we now use it to characterize *contemporary* differences in microenvironmental conditions faced by trees of different heights.**
+- Figure S4 and line 333: are differences expressed by different letters correct in Figure S4?
+Particularly for $\pi_{tlp}$ (S4d) the boxplot suggests differences (more than, e.g.
+in b) among different species but then letters are similar (as discussed in line 333).
 
-L217-226: this is problematic because drought resistance (Rt) as defined by Lloret et al.
-(2011) assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time.
-But this is rarely/never the case with tree growth, which shows clear temporal trends related to tree size/ontogeny.
-If you don’t account for these directional trends then your estimates of Rt will be biased.
-For instance, if a tree is on a long-term trajectory of decreasing BAI over time, when you calculate Rt you will overestimate the impact of the drought (and vice versa for trees on an upward growth trajectory).
-So you need to first detrend the tree ring series to remove long-term growth trends (unless you can show that your trees exhibited no long-term temporal trends in BAI).
-The fact that pre-drought conditions were different for the three droughts is also a problem, but not much you can do about this except acknowledge as you have done.
+- LMA is not shown in Figure S4, and in Table 2 there are no tests so we cannot judge whether there are differences among species for that leaf trait. Please, add also LMA to Figure S4.
 
-**We do not agree that the $Rt$ metric "assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time." If this was true, $Rt$ would be calculated by dividing growth (BAI) in the drought year by the long-term mean growth increment.
-Instead, $Rt$ is calculated by dividing growth in the focal year by the mean growth over a narrow window of preceding years (5 years in this study).
-Even though there may be a long-term directional trend in BAI, if you were to calculate the trend in BAI over each 5-year increment, the trend could go in any direction.
-To account for such trends, we tried an intervention time series analysis (ARIMA).
-Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4.
-In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric.
-We now include the following content on ARIMA in the manuscript:**
+- Line 135: leave turgor ‘at’ lower ….
 
-**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
-The two metrics were strongly correlated (Fig.
-**S5**).
-Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**
+- Line 272: please define dAIC. I think it should be better and more robust to stick in all cases to dAIC>2, there are only two covariates selected with <1 in Table S4.
 
-L260-261: This is rather strange.
-So for the same individual tree, size was important in determining response to drought in one year but not in the others?
-And that’s for the years further back in time when the tree would have been smallest?
-Couldn’t that be interpreted as trees becoming less susceptible to drought as they grow larger/older?
-It’s also a little concerning that results are very different across the three droughts.
-Of these, the 1999 drought is the one I would tend to trust most as stand structure and composition would have been most similar to today.
+- Line 312: please, refer to the corresponding Table or Supplement where this statement is supported.
 
-**We don't find it surprising that height could be more important in some droughts than in others.
-We address this in the second paragraph of the discussion: "Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
-**S1**), severely stressed a larger portion of trees (Fig.
-**1b**).
-The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]."**
+- Line 371-372: actually it is radial growth that is affected; whether that can be extrapolated to the tree level is something that needs to be demonstrated.
 
-**It's important to keep in mind that we're using height estimates specific to each drought year, and these estimates should be fairly reliable (we've dropped crown position).
-We now clarify that height is specific to the drought year: "Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig.
-**4**)." While the distribution of heights did shift upwards with progressive droughts, this shift is modest relative to the total height range (a few meters relative to a height range of ~40 m; Fig.
-S3).**
+- Line 385 vs lines 308-309 and line 338: does that mean that there was a correlation between canopy position and positive Rt?
+Can it also be related to moisture sensitivity of different species, or to niche (microenvironment) partitioning among species?
+(as nicely discussed in lines 420-422).
+I see from lines 445-450 that sampled FAGR were small, which is coherent with the authors’ statement.
 
-**There are a number of factors-- including height distribution and also atmospheric CO2 -- that differ across the droughts and may bias the mean response.
-To account for this as best as possible, we included year as a fixed effect in all statistical models (even though it did not come out significant).
-This is now clearly specified in the methods: "In the multi-year model, we also included a fixed effect of drought year to represent the combined effects of differences in drought characteristics." There were no significant differences across years, which we now specify in the results.**
+Table S5: no need to specify dAIC>1.
 
-Other comments
-
-L42-54: I think it would be worth making the link between growth resilience and probability of mortality.
-Below are links to a couple of papers that have recently shown that trees that are less resilient to drought are also the ones most likely to die.
-https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41467-020-14300-5.epdf%3Fshared_access_token%3DD9UngngFedreJeMMIQaYctRgN0jAjWel9jnR3ZoTv0PoKBL3IktPXOnnO0jnQ18EsS6uIPgICnN1RqVtsCGbwDx3K-1u5B0F1T9z3onm3gq-9mknjXKtuc0p7xxY3JipBCqwvtnrdqVNpsLdyzkNRw%253D%253D&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=6%2B7kb11Bzq4LivdOhnR4frg1MEFGaX%2FiIMUqJBG9gsM%3D&amp;reserved=0
-
-https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.nature.com%2Farticles%2Fs41558-019-0583-9&amp;data=02%7C01%7CTeixeiraK%40si.edu%7Cda67e36e2ccc4277d4aa08d7ed356305%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637238685875974418&amp;sdata=DF9lSNfAFTjKmSWfb7FI8%2FW4GDN4ljKxZK5NNS8eCxc%3D&amp;reserved=0
-
-**Thank you. We have added the following statement:**
-
-**"Because the resistance and resilience of growth to drought is linked to their probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality."**
-
-L80-101 & L112-115: This is a nice summary of how traits can help us understand species responses to drought.
-The only thing that is a little strange is that you start by discussing commonly measured traits that have been used to explore susceptibility to drought (e.g., WD and LMA).
-Then you say that the low predictive power of these traits is because they are poor surrogates of what we should be measuring – i.e.
-hydraulic traits.
-But then you end by saying that these are too hard/expensive to measure, so you’re going to look at other surrogates.
-
-**We have revised this section, and believe that it now makes our points more clearly: (1) the traditionally measured traits are poor surrogates, (2) hydraulic traits are useful but time-consuming, (3) $PLA_{dry}$ and $\pi_{tlp}$ are promising alternatives that are both relatively easy to measure and have direct physiological linkages to tree growth and mortality responses to drought.
-The text now reads as follows:**
-
-**"Many studies have shown that within and across species, large trees tend to be more affected by drought.
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020].
-It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response.
-First, tree height itself may be a primary driver.
-Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
-Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].
-Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
-Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019]."**
-
-**"...Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
-Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent.
-For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
-Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]."**
-
-**"In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought.
-For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
-However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**
-
-L102 & L106: the introduction gives a nice overview of size and trait influences on resistance to drought.
-But the third focus of your study – microenvironment characteristics – gets little or no attention.
-In fact it’s not very clear what you mean by ‘microenvironment’ for most of the paper.
-
-**We have placed the focus primarily on tree size and traits because that's where we see the most interesting story.
-We are interested in microenvironment primarily in how it interacts with tree size--i.e., in how microclimate varies across the vertical forest gradient, and in whether larger root systems are an advantage in drier microsites.
-We have made a number of changes that should make this clearer.
-For instance, we now have a paragraph in the introduction starting as follows "Larger trees may also have lower drought resistance because of microenvironmental and ecological factors." We then discuss how larger trees have more exposed crowns, and how larger root systems could be advantageous in drier microsites during drought.**
-
-L150-152: Given that height is estimated and not measured, wouldn’t it be better to use DBH as a (measured) dimension of tree size?
-
-**We prefer to use height because it is more meaningful biologically. DBH would give the same result.**
-
-L173: this is the first mention of what I’m guessing is your ‘microenvironment’ descriptor. Worth briefly mentioning why you are measuring TWI.
-
-**We have revised this sentence to read, "Topographic wetness index (TWI), *used here as a metric of long-term mean moisture availability*,..**
-
-L230: Somewhere here it would be good to remind the reader what these explanatory variables are.
-
-**We have revised this sentence to read, "The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below." We later specify the list of variables.**
-
-L234-237: Not very clear how these models were constructed.
-Why did you fit each predictor independently to test its importance as opposed to standardizing them and having them together in a multiple regression and then compare their coefficients?
-And why was height retained in all models?
-Testing one predictor at a time might be sensible if you have lots of collinearity between predictors.
-But it could also mean you miss important conditional effects (e.g., the effect of variable x only emerges when you first account for the effect of variable y).
-
-*(just need to write a response here-- no further changes needed to manuscript)*
-
-L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular).
-So given that you’ve already mentioned how you defined drought in the methods maybe you don’t need to repeat this here?
-
-**We have dropped the criteria of decline in growth as part of how we defined drought. This does not change the drought years that qualify for inclusion.**
-
-L266-267: Don’t think you can test the effects of crown dominance going back this far. How can you be sure a tree was dominant/suppressed back in 1966?
-
-**We have dropped crown position from the model, and now use it only to give a picture of how microenvironmental conditions vary with tree height under contemporary conditions (Fig. 2).**
-
-L287: specify how they were related to drought. Positive? Negative?
-
-**This has been revised to read, "In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance (Fig. **4**; Tables 1,S4-S7)."**
-
-Tables and Figures: there are a large number of tables.
-I wonder if Table 2 and 3 would best fit in supporting information (with key information integrated directly into the text).
-I also feel that it would be nice to represent some of the key results (e.g., Rt and its relationship to height and leaf hydraulic traits) in a figure.
-
-**We have moved two tables (former tables 4 and 5 to the SI). The best models are now visualized in a figure (Fig. 4). We have also added a figure showing drought response by species.**
-
-## Referee: 4
-
-Comments to the Author
-
-In this paper, McGregor et al.
-study the effect of tree height, crown position, ring-porosity, wood density, LMA and two leaf water relations traits (percent loss of leaf area upon desiccation, PLA; and the turgor loss point, TLP) on the drought resistance of 12 temperate species from an oak-hickory forest of northern Virginia (USA).
-Drought resistance is assessed using series of radial growth data and for the three most severe droughts in the period 1950 - 2009 in the study area.
-The authors find consistent negative effects of tree height on drought resistance; effects of crown position, ring-porosity and water relation traits for some droughts; and no effect of wood density and LMA.
-The paper is well written and nicely organized around clear hypotheses.
-The results are intriguing but, in my opinion, the experimental design is not strong enough to clearly advance our understanding of the determinants of tree drought responses.
-
-Firstly, the way drought resistance (Rt) is estimated, as the ratio of BAI during drought to the mean BAI over the five years preceding the drought can be problematic, as growth dynamics during this pre-drought period were quite variable across species (Figure 1; incidentally, why do you show ring width indices in this figure if you then use BAI?).
-This approach has also advantages, particularly when applied in a resilience framework (as in the original Lloret et al.
-(2011)).
-However, if you focus specificaly on resistance it seems that directly modelling the growth reduction caused by the drought (e.g., using intervention time series analysis or other methods) would be a more robust approach.
-At the very least, if you continue using your simpler approach I would strongly suggest including some measure of pre-drought growth in the models to account for differences in growth dynamics among species.
-It would also nice to clarify why you do not assess recovery and resilience, unlike Lloret et al.
-(2011).
-
-**In response to Reviewer 4's comments, we tried an intervention time series analysis using the 'auto.arima' function from the 'forecast' package [@R-forecast].
-Specifically, we modeled BAI  based on the timeseries for the 10 years prior to the drought year (itself acting as the intervention).
-We then recorded the ratio between the mean predicted BAI and the observed BAI during the drought year, and compared these results to $Rt$.
-These yielded similar outcomes, which we now present in the Supplementary Info.
-The original $Rt$ and $Rt_{ARIMA}$ were fairly well correlated, but with some scatter (see figs below).
-To get a sense of which metric was performing better, we visually reviewed the records of the 18 trees with largest differences between $Rt$ and $Rt_{ARIMA}$.
-$Rt$ estimates were better for the majority of these, and generally were not hugely unrealistic, whereas $Rt_{ARIMA}$ could be way off (e.g., for several trees that were undergoing growth declines prior to drought and hit minima at or near the drought, $Rt_{ARIMA}$ was much greater than 1, suggesting a substantial growth increase during the drought.).
-Thus, we decided to stick with $Rt$ as the primary metric.**
-
-![Figure S5](tables_figures/publication/figureS5_Rt_arima_comparison.png)
-
-**Regarding the resilience indices, we considered analyzing these and decided it would be best to stick with just resistance for three reasons: (1) Most importantly, we would expect resilience to be governed by fundamentally different mechanisms than resistance.
-While it would be interesting to examine resilience, this would be beyond the scope of the current paper.
-(2) [@desoto_low_2020] used an extensive dataset of trees sampled after severe droughts and compared the responses to previous, less-severe droughts between trees that survived and trees that were killed by the severe drought.
-In angiosperms, trees that died from a severe drought had lower resistance to previous droughts, compared to trees that survived the severe drought.
-In gymnosperms, trees that died during severe droughts had lower recovery following previous droughts, compared to trees that survived the severe drought.
-These findings would argue that resistance is a more important variable for angiosperms.
-We now state the logic for focusing on $Rt$ in the methods section:**
-
-**"We focus exclusively on drought resistance ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of [@desoto_low_2020] suggest that $Rt$ is a more important drought response metric for angiosperms.
-"**
-
-Secondly, the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance, and it mixes different types of hypotheses operating at different levels.
-On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses.
-On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height.
-In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability.
-In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results.
-Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.
-
-**We respond to this one point at a time:**
-
-*the choice of traits does not appear well justified, as it seems to be driven more by easiness of measurement than by relevance*
-
-**Ease of measurement is part of what makes $PLA_{dry}$ and $\pi_{tlp}$ interesting. We now clarify this in the introduction:**
-
-**"However, [traits such as P50, P88, and hydraulic safety margin] are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation ($PLA_{dry}$) and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016].*The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.*"**
-
-*it mixes different types of hypotheses operating at different levels.
-On one hand there is tree height and crown position, which are measured at the level of individual trees and correspond to closely related hypotheses.
-On the other hand there are ring-porosity, wood density, LMA and water relations traits, which are measured at the species level and seem to be totally disconnected from the hypotheses regarding tree height.
-In addition, it is very hard to compare the explanatory power of these variables, as done in the paper, as they are based on very different sample sizes (571 trees vs 12 species) and respond to different sources of variability.
-In fact the way height effects are analyzed mixes the within- and between-species levels, which makes it difficult to interpret the results.*
-
-**There is nothing fundamentally wrong with a model comparing individual- and species- level effects, but we understand why it was confusing and believe that the revised presentation of the analysis should make all of this easier to follow.
-Improvements related to this comment include: (1) revision to how traits are selected for inclusion as candidate variables for the full model, which provides a more appropriate test of individual traits and guards against over-fitting; (2) moving of former tables 4 and 5 to the SI, and replacing fig.
-5 with a figure visualizing the model results; (3) adding a couple new figures showing variation in $Rt$ (Fig.
-3) and predictor variables (Fig.
-S4) across species; and (4) dropping canopy position from the model for predicting $Rt$.**
-
-*Finally, I think it would be more correct to refer to PLA and TLP as water relations traits rather than hydraulic traits.*
-
-**We now refer to these as "leaf drought tolerance traits", including in the manuscript title.**
-
-Thirdly, a more thorough analysis of the distribution of the study species in the sample plot would be required.
-The paper assumes that all species were subjected to the same drought levels.
-However, this is unlikely to be true, as different species are likely to occupy different microhabitats in the plot, and hence be subjected to different drought conditions.
-The authors account for this, to some extent, by including individual-level TWI in some models, but a more thorough analysis of the distribution of the study species as a function of TWI and other relevant micro-environmental variables would be needed to fully interpret the results.
-
-**We have added a new figure to the SI showing variation in predictor variables across species (Fig.
-S4).
-While there was some variation across species in $H$ and $TWI$, there were few pronounced differences across species.
-We now discuss the case of the most anomalous species, *Fagus grandifolia*, in the discussion section.
-We also determined that there is little systematic variation in $PLA_{dry}$ and $\pi_{tlp}$ with respect to $H$ or $TWI$ (analysis not shown).
-This, together with the fact that our statistical models do account for $TWI$, should minimize bias related to distribution of the study species within the plot.**
-
-Minor or more technical suggestions:
-
-L59-79: the arguments presented in this paragraph mix different scales, from within-individual to within- or among-communities.
-I think the arguments could be better structured to reflect the sources of variability involved and, ideally, focus on the scale that will be addressed in the paper.
-
-**We have revised this paragraph to provide some more specifics on the scale and setting of these tests.**
-
-L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.
-
-**We have revised this to read, "Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018]..."**
-
-L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.
-
-**We now define drought resistance in the first paragraph: " Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...".
-We then use this term when we wish to speak specifically about resistance.
-We use "tolerance" as a more general term, consistent with what seems to be a fairly nebulous definition in the literature.**
-
-L97: ‘greater variation’ that what?
-
-**This was removed.
-The text here now reads, "More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019].
-These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
-Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].
-The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
-"**
-
-L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…
-
-**We have added citations to both these papers and revised the sentence so that it is true: "The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**
-
-L137-140: mixing cores from dying and surviving individuals could be problematic. Did you find any association between survival and previous drought resistance? Was mortality different across species?
-
-**We agree that this could be a concern, but do not believe it is applicable here, for two reasons.
-First, this would be more likely to introduce a bias if the trees cored dead were killed by drought.
-We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived.
-However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years.
-Second, quoting our previous publication using these cores [@helcoski_growing_2019]: " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**
-
-**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**
-
-L160: you mean a log-log regression, right?
-
-**Yes. fixed.**
-
-L181: N=3 individuals per species is really quite low.
-In addition, by sampling up to 8 m height you would probably be sampling less exposed branches from taller trees, which could confound your results.
-
-**There were significant differences among species for all sampled traits, and intraspecific SD was generally low with respect to mean values and interspecific variation.
-We have added a figure to the SI (Fig.
-S4) showing variation in traits across species.
-We also note that a sample size of n=3 is not unprecendented for this type of analysis [@medeiros_extensive_2019].
-More samples would be ideal, but was not feasible given the time and resources available.
-**
-
-**The sampling strategy was consistent across species, and also consistent with general methods for leaf trait sampling.
-Because we were not testing species-level trait response (as opposed to sampling traits on every cored tree), we don't expect any bias related to sampling height.**
-
-L188: ‘Wood’.
-
-**Fixed.**
-
-L189: fresh volume?
-
-**Yes, now specified.**
-
-L223: clarify that the -4.53 value corresponds to July.
-
-**This sentence has been re-written: "The 1999 drought was preceded by wetter than average conditions until the previous June, but $PDSI$ plummeted below -3.0 in October 1998 and remained below this threshold through August 1999."**
-
-L228-234: please clarify model structure.
-I understand you included tree as a random factor only when the three droughts were analyzed together, right?
-It would be nice to know what was the contribution of random factors to model fit (only one R2 is shown in Table 5, and it is unclear whether it corresponds to marginal or conditional values).
-
-**We have substantially revised this section and expect that it should be far more clear. The text in question now reads:**
-
-**"The general statistical model for hypothesis testing was a mixed effects model, implemented in the lme4 package in R [@R-lme4].
-In the multi-year model, we included a random effect of tree nested within species and a fixed effect of drought year to represent the combined effects of differences in drought characteristics.
-Individual year models included a random effect of species.
-All models included fixed effects of independent variables of interest (Tables 1,3) as specified below."**
-
-**We have specified that R2 corresponds to conditional values (footnotes of former table 5/ current table S6, and table S7).**
-
-L238, 242, 244: it is unclear to me why in some cases you use a dAICc of 1 and sometimes of 2.
-
-**dAICc = 2 was consistently used to test for significance, whereas dAICc=1 was used as a preliminary threshhold for variable selection.
-We now clearly specify this in the footnote of table S4 (former table 4) and in the text: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
-We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models."**
-
-L287-299: it would be very useful to use scatterplots to show the effects of traits on Rt in a figure.
-
-**We have added the specified figure (Fig.
-S6).
-In addition, we have added 3 related figures illustrate how traits and $Rt$ vary across species (Figs.
-S4 and 3, respectively) and how $Rt$ responds to  $PLA_{dry}$ and $\pi_{tlp}$.**
-
-L291-294: but this is problematic considering that you only had two diffuse-porous species.
-
-**We have made changes to the analysis that resulted in dropping ring porosity from the full models.
-While we agree that it is problematic that we only have two ring-porous species, we believe that it's worth including because this engages with, and rejects, the existing understanding that in this biome, ring-porous species are more drought tolerant.**
-
-L297: TLP was significant for 1966 according to Table 4.
-
-**This text was removed.**
-
-L301-302: I would tone down this, as your models explained a rather low % of the variation in Rt.
-
-**Tree-ring data are inherently noisy, particularly in more mesic forests.
-In large part, this is due to strong neighborhood effects, along with variable allocation to ring growth around the circumference of the tree.
-These factors contribute to the low R2, but do not negate the significance of our findings.**
-
-L305-306: not for the 1977 and 1999 droughts, according to Table 5.
-
-**This text was removed.**
-
-L314-316: or simply that your sample size was rather low. See also main comments above.
-
-**Our sample size of tree-ring data is quite good, and we believe the traits sample size to be sufficient (see above).
-We have now tried numerous iterations of this analysis, and the result described here holds.
-Biologically, it also makes sense that the importance of height, TWI, and traits would vary somewhat across droughts, given that every drought is different.**
-
-L329-331: do you have any information on the demographic effects (e.g., mortality) of the three droughts studied?
-
-**Unfortunately, no. All we could say is that the 1999 drought didn't result in a big loss of live biomass.**
-
-Table 3: are DBH units correct? I suspect they should be mm instead of cm.
-
-**These were incorrect (in mm). Now converted to cm.**
-
-Figure 1: please explain all acronyms.
-
-**We now specify that these codes are given in Table 2.**
-
-Figure 2: where are the leaf hydraulic traits shown (cf. caption)?
-
-**It was an error that this was mentioned in the caption; now removed.**
-
-Table S1: why only 11 species? Some of the fits are really low…
-
-**Bark thickness measurements were drawn from an existing data set, and we did not make additional measurements because the trait is unlikely to have meaningful influence on results.
-This study already goes above and beyond the existing norm for accounting for bark thickness in DBH reconstructions; typically there is no correction for changes in bark thickness as the tree ages.**
-
-Table S3: what does ‘rank’ refer to?
-
-**This has been clarified.**
-
-Figure S1: please add spatial scale to the figure.
-
-**Done.**
-
-Figure S3: please specify that height values are estimated from reconstructed DBH.
-
-**Done.**
-
-\newpage
 ## Citations

### 2020-09-06 manuscript/response_to_reviews.Rmd
dealt with Mark Olson comments

@@
 
+**We have reworded the sentence to read, "Taller trees have wider conduits in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller."**
+
 392 “Constraint” is always such a vague term in biology (meaning everything from factors opposing selection to selection itself), with every biologist understanding something different.
@@
 Here, it could be removed entirely “previous findings that it is impossible for trees…”, I think profitably.
+
+**We have made this change.
+It now reads, "Mechanistically, this is consistent with, and reinforces, previous findings that it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]."**
 
@@
 
+**Yes, we agree. Thank you.**
+
 With regard to root mass, this might be useful: https://nam02.safelinks.protection.outlook.com/?url=https%3A%2F%2Fdoi.org%2F10.1371%2Fjournal.pone.0086550&amp;data=02%7C01%7CTeixeiraK%40si.edu%7C04ca4a8deb534237b5a508d84dd84e13%7C989b5e2a14e44efe93b78cdd5fc5d11c%7C0%7C0%7C637344938706540440&amp;sdata=tiYufg0%2F%2BuRBThcGeUHuRjVDhWPmbK0xP%2Fb2moH%2F210%3D&amp;reserved=0 seem to show above-below ground biomass isometry.
 
+**Thank you. We have added this reference in two places (intro and discussion).**
+
 Thanks again and congratulations to the authors.
+
+**Thank you.**
 

### 2020-09-06 manuscript/climate_sensitivity_draft.Rmd
dealt with Mark Olson comments

@@
 Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
-Indeed, tall trees require xylem of greater hydraulic efficiency, such that xylem conduit diameters are wider in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019], and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018].
+Taller trees have wider conduits in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller.
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
+**(here would be a good place to comment on resilience.)**
 
@@
 Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. 
-Counteracting the liabilities associated with tall height, large trees tend to have larger root systems [@enquist_global_2002], potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
+Counteracting the liabilities associated with tall height, large trees tend to have larger root systems [@enquist_global_2002; @hui_near_2014], potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
 Larger root systems--if they grant access to deeper water sources--would be particularly advantageous in drier microenvironments (e.g., hilltops, as compared to valleys and streambeds) during drought.
@@
 Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance. 
-Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
+Mechanistically, this is consistent with, and reinforces, previous findings that it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
 Taller trees also face dramatically different microenvironments (Fig. **2**). 
@@
 
-Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
+Belowground, taller trees would tend to have larger root systems [@enquist_global_2002; @hui_near_2014], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought.
 While tree size can correlate with the depth of water extraction [@brum_hydrological_2019], the linkage is not consistent.

### 2020-09-06 manuscript/response_to_reviews.Rmd
worked on R3 comments

@@
 
+*([issue #131](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/131))*
+
 2) In Line 263 it is stated that both marginal and conditional R2 are reported, but then I could only find conditional R2s in Table S6, S7.
@@
 
+*This is straightforward: need to report the marginal R2. Maybe also add some commentary on how a lot of noise doesn't negate significant signals.*
+
 Other comments:
@@
 
+*I think our argument is pretty strong, but we need to make some change to respond to this comment.
+Looking at resilience and recovery should help; presuming that they are high, we have a further argument that death was not caused by drought, and hence that status at coring doesn't matter.*
+
+*All trees included in the analysis lived at least 17-18 years past the most recent drought. *
+
 - Figure S4 and line 333: are differences expressed by different letters correct in Figure S4?
@@
 
+*Ian, please check this and respond.*
+
 - LMA is not shown in Figure S4, and in Table 2 there are no tests so we cannot judge whether there are differences among species for that leaf trait. Please, add also LMA to Figure S4.
+
+*Ian, we should add both LMA and WD here.*
 
@@
 
+**Thanks for catching this; fixed.
+Sentence now reads, "We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher resistance."**
+
 - Line 272: please define dAIC. I think it should be better and more robust to stick in all cases to dAIC>2, there are only two covariates selected with <1 in Table S4.
+
+**We now define AICc at first mention, which is in the paragraph above.
+The sentence now reads "We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection,..." In the sentence referenced here, we define $\Delta$AIC: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4)."**
+
+**We have given careful thought to the trait variable selection criteria.
+It is important to note that selection of traits is not a test of their significance in the final model; it is simply a way of objectively reducing the number of candidate traits in these models.
+While we fully agree that $\Delta$AICc $\ge$ 2 is the appropriate criterion for assessing significance, it is not necessarily appropriate for candidate variable selection because variable significance changes when considered independently versus in the context of a full model.
+That is, the significance of variables--particularly those with marginal significance-- cannot be reliably assessed outside the context of a model that simultaneously accounts for other relevant variables.
+As a case in point, $\pi_{tlp}$ would be removed by a $\Delta$AICc $\ge$ 2 criterion for variable selection, but is included in the best model for all droughts combined, as well as the 1977 drought.
+While it's effects are marginal (i.e., included in some but not all top models), we feel that it is important to include it as a candidate variable both for the statistical considerations mentioned above, and because we expect that a lot of readers will be very interested to see a full assessment of this trait, which is currently of great interest to the research community.**
 
@@
 
+**We have added this reference: "That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S6-S7)."**
+
 - Line 371-372: actually it is radial growth that is affected; whether that can be extrapolated to the tree level is something that needs to be demonstrated.
 
-- Line 385 vs lines 308-309 and line 338: does that mean that there was a correlation between canopy position and positive Rt?
+**This is true.
+Of course, tree biomass growth is almost always inferred from radial or diameter growth, sometimes in conjunction with height, using allometries.
+The issue of whether radial stem growth (or BAI, as examined here) actually translates directly to biomass growth goes far beyond the present study.
+To make sure this is clearly defined here, we've specified that we're considering stem growth in a couple key places:**
+
+* **in the introduction, where we define resistance: "Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth."**
+
+* **in the sentence identified above: "These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019]."**
+
+Line 385 vs lines 308-309 and line 338: does that mean that there was a correlation between canopy position and positive Rt?
 Can it also be related to moisture sensitivity of different species, or to niche (microenvironment) partitioning among species?
@@
 
+*(need to return here.)*
+
 Table S5: no need to specify dAIC>1.
 
+**This is fixed.**
+
 ## Citations

### 2020-09-06 manuscript/climate_sensitivity_draft.Rmd
worked on R3 comments

@@
 
-Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
 We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
@@
 We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
-We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance.
+We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher resistance.
 
@@
 All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
-We used AICc to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg]. 
+We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].
 AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].
@@
 These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
-Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4).
+Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4).
 We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models.
@@
 In the context of the multivariate model, $Rt$ did not vary across drought years.
-That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model.
+That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S6-S7).
 
@@
 For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
-$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig. **4**; Table S6). 
+$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig.
+**4**; Table S6), although its influence was not significant at $\Delta$AICc<2.
 It was also included in some of the top models for 1999 (Tables S6-S7).
@@
 **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
-These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
+These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
 

### 2020-09-06 manuscript/response_to_reviews.Rmd
finished all I currently can on R3 (and all revierwers)

@@
 
-*(need to return here.)*
+**Yes, although we do not include canopy position in our statistical models.
+To make the link between high $Rt$ and increased growth, we have modified the sentence previously on line 385 to read, "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits; Fig.
+**4**) had increased growth (Fig.
+**1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]."**
+
+*(for reference, to be deleted:)*
+
+*385: Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b),*
+
+*208-309: However, some individuals exhibited increased growth, i.e., Rt > 1.0: 26% of trees in 1966, 22% in 1977, and 26% in 1999.*
+
+*338: Averaged across all droughts, Rt was lowest in Liriodendron tulipifera (mean Rt = 0.66) and highest in Fagus grandifolia (mean Rt = 0.99).*
 

### 2020-09-06 manuscript/climate_sensitivity_draft.Rmd
finished all I currently can on R3 (and all revierwers)

@@
 The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
-Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig.
+Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits; Fig.
+**4**) had increased growth (Fig.
 **1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].

### 2020-09-06 manuscript/climate_sensitivity_draft.Rmd
TLP significance
#134

@@
 The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
-Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.
+Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and*, marginally, those whose leaves* lost turgor (wilted) at more negative water potentials.
 

### 2020-09-07 manuscript/response_to_reviews.Rmd
TLP significance
#134
more on R3 comment on live/dead

@@
 
-*All trees included in the analysis lived at least 17-18 years past the most recent drought. *
+**We consider it extremely unlikely that the status at coring would affect drought response.
+We have modified the text at (former) lines 175-178 to make this more clear: "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
+Lagged drought-induced mortality would be unlikely, given that the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare."**
+
+**We have added a footnote to Table 2 indicating where the information on numbers cored live vs dead may be found (Table S1 of Helcoski et al. 2019).**
 
@@
 While it's effects are marginal (i.e., included in some but not all top models), we feel that it is important to include it as a candidate variable both for the statistical considerations mentioned above, and because we expect that a lot of readers will be very interested to see a full assessment of this trait, which is currently of great interest to the research community.**
+
+*([issue #134](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/134))*
 

### 2020-09-07 manuscript/climate_sensitivity_draft.Rmd
more on R3 comment on live/dead

@@
 Discussion |	1927	||
-Acknowledgements | 142	||
+Acknowledgments | 142	||
 
@@
 We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
+Lagged drought-induced mortality would be unlikely, given that the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare.
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 

### 2020-09-07 manuscript/tables&figures.Rmd
more on R3 comment on live/dead

@@
 table2 <- read.csv("tables_figures/publication/table2_species_table_hand_formatted.csv", stringsAsFactors = FALSE, check.names = FALSE)
-table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(1))
+table2[0,3] <- paste0(table2[0,3], footnote_marker_symbol(1))
+table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(2))
 kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2.
@@
   #kable_styling(bootstrap_options = "striped", full_width = FALSE)
-kableExtra::footnote(symbol = c("Semi-ring porosity is intermediate between ring and diffuse.
+kableExtra::footnote(symbol = c("Numbers cored live versus dead are given in Table S1 of Helcoski et al.
+(2019).", "Semi-ring porosity is intermediate between ring and diffuse.
 We group it with diffuse-porous species for more even division of species between categories."), footnote_as_chunk = FALSE)

### 2020-09-07 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
-Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these. 
-Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].
+Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected in terms of growth during the drought year. 
+**Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these.
+Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].**
 As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].

### 2020-09-08 manuscript/tables&figures.Rmd
revise Table 1 to include Rc an Rs
#131

@@
 library(kableExtra)
-hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table3.csv", stringsAsFactors = FALSE, check.names = FALSE)
 kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
@@
  pack_rows("Across the forest vertical profile, taller trees are exposed to higher evaporative demand.", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("At least within the forest setting, taller trees are less drought resistant.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("At least within the forest setting, taller trees are less drought tolerance", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought tolerance", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Species' traits--particularly leaf drought tolerance traits--predict drought reistance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
-  #add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  #column_spec(1, width = "12cm") %>%
+  pack_rows("Species' traits--particularly leaf drought tolerance traits--predict drought tolerance.", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  add_header_above(c(" ", " ", "tree-ring drought records" = 3, " ")) %>%
+  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
+  column_spec(2, width = "1.7cm") 
   #column_spec(2:5, width = "1.2cm") %>%
@@
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png)
+Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1_recovery.png)
 
@@
 
-![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_rt_across_sp.png)
+![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_resistance_across_sp.png)
 
@@
 Model coefficients are given in Table S6.
-](tables_figures/publication/Figure4_model_vis.png)
+](tables_figures/publication/Figure4_model_vis_all.png)

### 2020-09-08 manuscript/tables&figures.Rmd
fix up Table1
#131

@@
   add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
-  column_spec(2, width = "1.7cm") 
-  #column_spec(2:5, width = "1.2cm") %>%
-  #column_spec(6, width = "2cm") 
+  column_spec(2, width = "1.7cm")  %>%
+  column_spec(3:5, width = "1.2cm") %>%
+  column_spec(6, width = "1.8cm") 
  # kableExtra::landscape()

### 2020-09-08 manuscript/tables&figures.Rmd
add Rc and Rs to table 3
#131 , #139

@@
   column_spec(4, width = "7cm") %>%
-  pack_rows("Dependent variables", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Independent variables", 3, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("species' traits", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Dependent variables", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Independent variables", 5, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("species' traits", 10, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   kable_styling()

### 2020-09-08 manuscript/tables&figures.Rmd
update tables&figures
#131

@@
  pack_rows("Across the forest vertical profile, taller trees are exposed to higher evaporative demand.", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("At least within the forest setting, taller trees are less drought tolerance", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought tolerance", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("At least within the forest setting, taller trees are less drought tolerance.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought tolerance.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Species traits", 7, 11, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
@@
 
+![3b](tables_figures/publication/Figure3_resilience_across_sp.png)
+
 \newpage

### 2020-09-08 manuscript/climate_sensitivity_draft.Rmd
changed abstract
#131

@@
 ### Summary
-- As climate change drives increased drought in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important.
+- As climate change drives increased drought in many forested regions around the world, mechanistic understanding of the factors conferring drought tolerance in trees is increasingly important.
 The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape growth responses during droughts.
@@
 
-- Individual-level drought resistance decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
-The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
-Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and*, marginally, those whose leaves* lost turgor (wilted) at more negative water potentials.
+- Individual-level drought tolerance (resistance, recovery, and resilience) decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
+The potentially greater rooting volume of larger trees did not confer a resistance advantage in sites with low topographic wetness index, but marginally increased recovery and resilience.
+*Resistance and resilience were greater among species whose leaves experienced less shrinkage upon desiccation and, while recovery was greatest among those whose leaves lost turgor (wilted) at more negative water potentials.*
 

### 2020-09-09 manuscript/tables&figures.Rmd
update table 1 with final results

@@
  pack_rows("Across the forest vertical profile, taller trees are exposed to higher evaporative demand.", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
-  pack_rows("At least within the forest setting, taller trees are less drought tolerance.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
+  pack_rows("At least within the forest setting, taller trees are less drought tolerant.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
   pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought tolerance.", 6, 6, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
@@
 
-Parentheses indicate that the prediction was supported by one but not all of the top  models (Table S6).
+Parentheses indicate that the prediction was supported by at least one but not all of the top  models (Table S6).
 Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S6).

### 2020-09-09 manuscript/climate_sensitivity_draft.Rmd
update summary
#131
word count exactly at limit; I'd consider this "final"

--- 2020-09-07 manuscript/climate_sensitivity_draft.Rmd
@@
 Total word count (excluding summary, references and legends) | 6,490	 |	No. of figures | 	4 (all colour)
-Summary|  	198 | 	No. of Tables | 	3  
+Summary|  	200 | 	No. of Tables | 	3  
 Introduction|	1,264 |	No of Supporting Information files | 	14
@@
 ### Summary
-- As climate change drives increased drought in many forested regions around the world, mechanistic understanding of the factors conferring drought resistance in trees is increasingly important.
-The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape growth responses during droughts.
+- As climate change drives increased drought in many forested regions, mechanistic understanding of the factors conferring drought tolerance in trees is increasingly important.
+The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape growth responses to droughts.
 
-- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity in a broadleaf deciduous forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).
+- We analyzed tree-ring records for twelve species in a broadleaf deciduous forest in Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period.
 
-- Individual-level drought resistance decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
-The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index.
-Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and*, marginally, those whose leaves* lost turgor (wilted) at more negative water potentials.
+- Drought tolerance (resistance, recovery, and resilience) decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
+The potentially greater rooting volume of larger trees did not confer a resistance advantage, but marginally increased recovery and resilience, in sites with low topographic wetness index.
+Drought tolerance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.
 
-- We conclude that tree height and leaf drought tolerance traits influence growth responses during drought, as recorded in the tree-ring record spanning historical droughts.
-Thus, these factors can be useful for predicting future drought responses under climate change.
+- Tree height and leaf drought tolerance traits influence growth responses during and after drought, as recorded in the tree-ring record spanning historical droughts.
+As climate change-induced droughts intensify, tall trees with drought-sensitive leaves will be most vulnerable to immediate and longer-term growth reductions.
 

### 2020-09-09 manuscript/climate_sensitivity_draft.Rmd
worked on revising intro

@@
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
-Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts.
-Because the resistance and resilience (i.e., post-drought recovery) of growth to drought are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
-However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (resistance) and to recover to its pre-drought growth rate (resilience) [@lloret_components_2011].
+Because the resistance and resilience of growth to drought are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
+However, it has proven difficult to resolve the many factors affecting tree growth during drought and the extent to which their influence is consistent across droughts.
+This is because available forest census data only rarely captures extreme drought, whereas tree-ring records capture multiple droughts but typically focus on only the largest individuals of  one or a few species.
 
 Many studies have shown that within and across species, large trees tend to be more affected by drought. 
-Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020].
-It has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
+Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @gillerot_tree_2020].
+Although lower recovery and resilience of larger trees have also been observed [@hacket-pain_consistent_2016; @gillerot_tree_2020], in general we have much more limited understanding of how and why these scale with tree size.
+
+*physiological model of @trugman_tree_2018 predicts that recovery and resilience would not necessarily be lower, even in trees that will ultimately die.*
+
+Moreover, it has yet to be resolved which of several potential underlying mechanisms most strongly shape these trends in drought response. 
 First, tree height itself may be a primary driver. 
@@
 
-Larger trees may also have lower drought resistance because of microenvironmental and ecological factors. 
+Larger trees may also have lower drought tolerance because of microenvironmental and ecological factors. 
 Their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017]. 
@@
 Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].
-The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested.
+The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain the drought tolerance of tree growth remains untested.
 
-Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth.
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape three metrics of drought tolerance: 
+(1) resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth;
+(2) recovery, defined the ratio of post-drought growth to growth during the drought year;
+and (3) resilience, defined as the ratio of post-drought to pre-drought growth [@lloret_components_2011].
 We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from the three strongest droughts over a 60-year period (1950 - 2009), species trait measurements, and census and microenvironmental data from a large forest dynamics plot in Virginia, USA.
-First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. 
+First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought tolerance. 
 We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
-We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
+We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought tolerance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
 We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
-Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. 
-We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
-We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher resistance.
+Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought tolerance. 
+We test predictions that drought tolerance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$.
+We further test predictions that species with low $PLA_{dry}$ and those whose leaves lose turgor at lower water potentials (more negative $\pi_{tlp}$) have higher tolerance.
 

### 2020-09-09 manuscript/climate_sensitivity_draft.Rmd
update abstract
per Neil's suggestion

@@
 - As climate change drives increased drought in many forested regions, mechanistic understanding of the factors conferring drought tolerance in trees is increasingly important.
-The dendrochronological record provides a window through which we can understand how tree size and species’ traits shape growth responses to droughts.
+The dendrochronological record provides a window through which we can understand how tree size and traits shape growth responses to droughts.
 
@@
 
-- Tree height and leaf drought tolerance traits influence growth responses during and after drought, as recorded in the tree-ring record spanning historical droughts.
+- The tree-ring record reveals that tree height and leaf drought tolerance traits influenced growth responses during and after significant droughts in the meteorological record.
 As climate change-induced droughts intensify, tall trees with drought-sensitive leaves will be most vulnerable to immediate and longer-term growth reductions.

### 2020-09-10 manuscript/tables&figures.Rmd
update figure captions
#131

@@
 ![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
 2014).
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric.
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1_recovery.png)
+Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought.](tables_figures/publication/Figure1.png)
 
@@
 
-![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_resistance_across_sp.png)
+![**Figure 3.
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_resistance_across_sp.png)
 
@@
 ![**Figure 4.
-Visualization of best models for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
+Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
 Model coefficients are given in Table S6.

### 2020-09-10 manuscript/response_to_reviews.Rmd
worked on response to reviews (minor)

@@
 
-*([issue #131](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/131))*
+**Thank you for this suggestion.
+Although we were initially hesitant to include these other indices, we found that their addition greatly enriched the manuscript.
+We have added analyses parallel to those for resistance for both recovery and resilience.
+Because we feel that resilience is the more interesting and informative of these two, we have given it precedence for inclusion in main manuscript figures, where space is limiting.
+Specifically, we have added a density plot of resilience to Fig.1 (+ parallel plot for recovery in the SI), a plot of resilience by species to Fig.
+3 (+ parallel plot for recovery in the SI), and plots of the best models for both recovery and resilience to Fig.
+4.
+We have also added recovery and resilience to the hypothesis table (Table 1), and their definitions to Table 3.
+We have also modified the text throughout and included relevant tables presenting statistics in the supplementary info.**
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
Renamed files and updated SI code for that #131

@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/tableS4_singlevariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
+singlevar_table <- read.csv("tables_figures/publication/tableS4_Rt_singlevariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_all.csv`.
 colnames(singlevar_table)[1] <- "variable"
@@
 library(kableExtra)
-singlevar_table <- read.csv("tables_figures/publication/tableS5_singlevariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+singlevar_table <- read.csv("tables_figures/publication/tableS5_Rt_singlevariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 colnames(singlevar_table)[1] <- "variable"
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_multivariable_models_lmer_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS6_RRt_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 colnames(multivar_table)[1] <- "drought"
@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS7_multivariable_models_lmer_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS7_Rt_multivariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 colnames(multivar_table)[1] <- "drought"

### 2020-09-11 scripts/climate_sensitivity_analysis.R

@@ -1066,7 +1066,7 @@ for (i in seq(along=c(1:4))){
   top_models <- rbind(top_models, top)
 }
 
-#the data in top_models is used to update tables S6 (Rt) and S7 (Rt_arima)
+#the data in top_models is used for creating Fig. 4
 file_bestmod <- paste0("manuscript/tables_figures/tested_traits_best_",
                        metric, "_CPout")
 file_top <- paste0("manuscript/tables_figures/top_models_dAIC_",
@@ -1234,7 +1234,7 @@ for(i in seq(along=patterns)){
   coeff_new$rank <- gsub(patterns[[i]], "", coeff_new$rank)
 }
 
-#this table is used to fill in Table 5 (Rt) or S5 (arimaratio)
+#this table is used to fill in Table S6 
 write.csv(coeff_new, 
           paste0("manuscript/tables_figures/tested_traits_best_coeff_",
           metric, "_CPout.csv"), row.names=FALSE)

### 2020-09-11 scripts/graphs_for_manuscript.R
Updated plot code

@@ -26,27 +26,42 @@ for(i in 1:length(mt)){
       paste0("manuscript/tables_figures/trees_all_sub_", metric, ".csv"))
    trees_all_sub$year <- as.character(trees_all_sub$year)
    
-   plt <- ggplot(trees_all_sub) +
-      aes(x = metric.value, fill = year) +
-      geom_density(adjust = 1L, alpha=.5) +
-      geom_vline(xintercept=1) +
-      ylim(0,2) +
-      scale_fill_hue(labels=c("1966", "1977", "1999")) +
-      labs(x=paste0(metric, " value")) +
-      theme_minimal(base_size=10) +
-      theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))
-   # annotate(geom="text", x=0.1, y=1.5, 
-   #          label = "(b)", fontface="bold", size=7)
-   
-   #we want recovery (i==3) to be a full plot on its own for appendix
-   if(i==1){
-      plt <- plt  +
-         theme(legend.position="none")
-   } else if(i==2) {
-      plt <- plt + 
-         theme(axis.title.y = element_blank(),
-               legend.text = element_text(size=10)) +
-         guides(shape = guide_legend(override.aes = list(size = 3)))
+   # library(data.table)
+   # trees_all_sub <- fread(
+   #    paste0("manuscript/tables_figures/trees_all_sub_", metric, ".csv"))
+   if(i %in% c(1,2)){
+      plt <- ggplot(trees_all_sub) +
+         aes(x = metric.value, fill = year) +
+         geom_density(adjust = 1L, alpha=.5) +
+         geom_vline(xintercept=1) +
+         ylim(0,2) +
+         scale_fill_hue(labels=c("1966", "1977", "1999")) +
+         labs(x=paste0(metric, " value")) +
+         theme_minimal(base_size=10) +
+         theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))
+      # annotate(geom="text", x=0.1, y=1.5, 
+      #          label = "(b)", fontface="bold", size=7)
+      
+      #we want recovery (i==3) to be a full plot on its own for appendix
+      if(i==1){
+         plt <- plt  +
+            theme(legend.position="none")
+      } else if(i==2) {
+         plt <- plt + 
+            theme(axis.title.y = element_blank(),
[…]

### 2020-09-11 manuscript/tables&figures.Rmd
update Table 3
#139
update tables and figures file (should be done!)

@@
 
-*Sample sizes are after removal of outliers, and refer to the $Rt$ model. Dashes indicate that the variable was available for all records.
+*Sample sizes are after removal of outliers. Dashes indicate that the variable was available for all records. For resistance, sample sizes of independent variables to the $Rt$ model.
 
@@
 ![**Figure 3.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_resistance_across_sp.png)
-
-![3b](tables_figures/publication/Figure3_resilience_across_sp.png)
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_Rt-Rs_across_sp.png)
 

### 2020-09-11 scripts/drought_plot.m
added post-drought years to PDSI plot

@@ -13,31 +13,36 @@ PDSI=NOAA_table2(:,5);
 
 dr=[1966 1977 1999];
 colors= [ 1 .4 .7; .2 .7 .2 ; .4 .4 1];
+figure(1)
 for d=1:3
     fyr=dr(d);
-    index=find(year>(fyr-3) & year<(fyr+1));
+    index=find(year>(fyr-5) & year<(fyr+5));
     date_rel=date(index)-datenum([fyr 1 1]).* (1+0*(index));
     plot(date_rel,PDSI(index), 'LineWidth',2, 'Color', colors(d,:)); hold on;
 end
 
-xlim([-729 365]);
-set(gca, 'XTick', (-729 : 365/4 : 365 ));
+xlim([-365*2+1 365*3+1]);
+set(gca, 'XTick', (-365.25*2+2 : 365.25/3 : 365.25*3+2 ));
 datetick('x','mmm','keepticks')
 
 xline(121,'-r',{'peak growing season','of drought year'},'LineWidth',4);
 xline(243,'-r','LineWidth',4);
-xline(0,'--');
-xline(-365,'--');
-xline(-730,'--');
+xline(-365+1,'--');
+xline(0+1,'--');
+xline(365+1,'--');
+xline(365*2+1,'--');
+
+
 
 xlabel('month')
 ylabel('PDSI')
-legend({'1966' '1977' '1999'},'Location','Southwest')
+legend({'1966' '1977' '1999'},'Location','Southeast')
 legend('Boxoff')
 
+set(gcf, 'Position',[200 200 1000 400])
 
 shg
 
-cd '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/manuscript/tables_figures'
-print('drought_plot', '-djpeg')
+cd '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/manuscript/tables_figures/publication'
+print('figureS1_drought_plot', '-djpeg')
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
renumbered tables, inserted into SI (still needs some work)
#131

@@
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS6_RRt_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+singlevar_table <- read.csv("tables_figures/publication/tableS6_Rc_singlevariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(singlevar_table)[1] <- "variable"
+kable(singlevar_table, booktabs = TRUE, caption = "Table S6. Individual tests of species traits as drivers of drought recovery ($Rc$).", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  kable_styling()
+```
+Variable abbreviations are as in Table 2. 
+$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
+
+**$\Delta$AICc > 2: variable considered significant as an individual predictor
+
+\newpage
+```{r table S7, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+singlevar_table <- read.csv("tables_figures/publication/tableS7_Rs_singlevariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(singlevar_table)[1] <- "variable"
+kable(singlevar_table, booktabs = TRUE, caption = "Table S7. Individual tests of species traits as drivers of drought resilience ($Rs$).", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  add_header_above(c(" "," ","all droughts"=2,"1966"=2, "1977" = 2, "1999" = 2)) %>%
+  kable_styling()
+```
+Variable abbreviations are as in Table 2. 
+$\Delta$AICc is the AICc of a model excluding the trait minus that of the model including it.
+
+**$\Delta$AICc > 2: variable considered significant as an individual predictor
+
+\newpage
+```{r table S8, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+multivar_table <- read.csv("tables_figures/publication/tableS8_Rt_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 colnames(multivar_table)[1] <- "drought"
-kable(multivar_table, booktabs = TRUE, caption = "Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
+kable(multivar_table, booktabs = TRUE, caption = "Table S8. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
@@
 \newpage
-```{r table S7, eval = TRUE, echo=FALSE, warning=FALSE}
+```{r table S9, eval = TRUE, echo=FALSE, warning=FALSE}
 library(knitr)
 library(kableExtra)
-multivar_table <- read.csv("tables_figures/publication/tableS7_Rt_multivariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+multivar_table <- read.csv("tables_figures/publication/tableS9_Rt_multivariable_models_arimaratio_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
 colnames(multivar_table)[1] <- "drought"
-kable(multivar_table, booktabs = TRUE, caption = "Table S7.
+kable(multivar_table, booktabs = TRUE, caption = "Table S9.
 Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
@@
 Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
+
+\newpage
+```{r table S10, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+multivar_table <- read.csv("tables_figures/publication/tableS10_Rc_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(multivar_table)[1] <- "drought"
+kable(multivar_table, booktabs = TRUE, caption = "Table S10. Summary of top models for each drought instance, where $Rc$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  #pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  row_spec(1, bold = T) %>%
+  row_spec(5, bold = T) %>%
+  row_spec(8, bold = T) %>%
+  row_spec(15, bold = T) %>%
+  kable_styling()
+```
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
+$R^2$ refers to conditional $R^2$. 
+Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
+
+\newpage
+```{r table S11, eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+multivar_table <- read.csv("tables_figures/publication/tableS11_Rs_multivariable_models_CPout.csv", stringsAsFactors = FALSE, check.names = FALSE) 
+colnames(multivar_table)[1] <- "drought"
+kable(multivar_table, booktabs = TRUE, caption = "Table S11. Summary of top models for each drought instance, where $Rs$ is used as the response variable.", format = 'latex', escape = FALSE) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
+  #pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  row_spec(1, bold = T) %>%
+  row_spec(5, bold = T) %>%
+  row_spec(8, bold = T) %>%
+  row_spec(15, bold = T) %>%
+  kable_styling()
+```
+Models are ranked by AICc. Shown are all models whose AICc value falls within 2.0 ($\Delta$AICc<1) of the best model (bold). 
+$R^2$ refers to conditional $R^2$. 
+Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
 

### 2020-09-11 manuscript/response_to_reviews.Rmd
SI cleanup

@@
 
-*This is straightforward: need to report the marginal R2. Maybe also add some commentary on how a lot of noise doesn't negate significant signals.*
+**We now report both marginal and conditional R2 in tables S8-S12**.
+
+*Maybe also add some commentary on how a lot of noise doesn't negate significant signals.*
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
SI cleanup

@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  #pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 5, 9, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 10, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 21, 26, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
   row_spec(5, bold = T) %>%
-  row_spec(8, bold = T) %>%
-  row_spec(15, bold = T) %>%
+  row_spec(10, bold = T) %>%
+  row_spec(21, bold = T) %>%
   kable_styling()
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
-  #pack_rows("", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 5, 7, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 8, 14, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
-  #pack_rows("", 15, 25, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 1, 5, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 6, 10, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 11, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("", 17, 21, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
   row_spec(1, bold = T) %>%
-  row_spec(5, bold = T) %>%
-  row_spec(8, bold = T) %>%
-  row_spec(15, bold = T) %>%
+  row_spec(6, bold = T) %>%
+  row_spec(11, bold = T) %>%
+  row_spec(17, bold = T) %>%
   kable_styling()
@@
 \newpage
-![**Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought**](tables_figures/publication/figureS1_drought_plot.jpg)
+![**Figure S1. Time series of Palmer Drought Severity Index (PDSI) foreach focal drought year $\pm$ 2 years**](tables_figures/publication/figureS1_drought_plot.jpg)
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
update table numbers

@@
 ![**Figure S6.
-Visualization of the best model ($\Delta$AICc=0) with data for all droughts combined.** Model coefficients are given in Table S6.](tables_figures/publication/FigureS6_best_model_with_data.png)
+Visualization of the best model ($\Delta$AICc=0) with data for all droughts combined.** Model coefficients are given in Table S8.](tables_figures/publication/FigureS6_best_model_with_data.png)
 

### 2020-09-11 manuscript/climate_sensitivity_draft.Rmd
update table numbers

@@
 In the context of the multivariate model, $Rt$ did not vary across drought years.
-That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S6-S7).
+That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9).
 
@@
 Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Figs. **4, S6**). 
-Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S6-S7). 
+Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9). 
 The same held true for 1966 individually.
-For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S7).
+For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S9).
 For the 1999 drought, $ln[H]$ had no significant effect.
 
-$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Figs. **4, S6**, Table S6-S7). 
-The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S6).
-When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in 1977, and included in some of the top models in 1966 and 1999 (Table S7).
+$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Figs. **4, S6**, Table S8-S9). 
+The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S8).
+When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in 1977, and included in some of the top models in 1966 and 1999 (Table S9).
 This negates the idea that trees in moist microsites would be less affected by drought. 
 Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S6-S7).
-This term did appear with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S7), indicating that the negative effect of height on  $Rt$ was significantly stronger in wetter microhabitats.
+This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S8-S9).
+This term did appear with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S9), indicating that the negative effect of height on  $Rt$ was significantly stronger in wetter microhabitats.
 
@@
 In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Figs.
-**4, S6**; Tables 1,S4-S7).
-$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S6-S7). 
-For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S6-S7).
+**4, S6**; Tables 1,S4-S5,S8-S9).
+$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). 
+For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S8-S9).
 $\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig.
-**4**; Table S6), although its influence was not significant at $\Delta$AICc<2.
-It was also included in some of the top models for 1999 (Tables S6-S7).
+**4**; Table S8), although its influence was not significant at $\Delta$AICc<2.
+It was also included in some of the top models for 1999 (Tables S8-S9).
 
@@
 This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
-The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms.
+The directions of these responses were consistent across droughts (Table S8), supporting the premise that they were driven by fundamental physiological mechanisms.
 However, the strengths of each predictor varied across droughts (Fig.
-**4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+**4**; Tables S8-S9), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
@@
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Fig.
-**S6**; Table S6).
+**S6**; Table S8).
 
@@
 
+**NEED TO UPDATE !!****
 Table S1. Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.

### 2020-09-11 manuscript/tables&figures.Rmd
update table numbers

@@
 
-Parentheses indicate that the prediction was supported by at least one but not all of the top  models (Table S6).
-Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S6).
+Parentheses indicate that the prediction was supported by at least one but not all of the top  models (Table S8).
+Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S8).
 
@@
 Visualization of best models for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
-Model coefficients are given in Table S6.
+Model coefficients are given in Tables S8 and S10-11.
 
@@
 Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
-Model coefficients are given in Table S6.
+Model coefficients are given in Table S8.
 ](tables_figures/publication/Figure4_model_vis_all.png)

### 2020-09-11 manuscript/tables&figures.Rmd
fixes to fig captions

@@
 kable(table3, format = "latex", booktabs = TRUE, caption = "Table 3.
-Summary of dependent and independent variables in our statistical models fo drought resistance, along with units, definitions, and sample sizes.",
+Summary of dependent and independent variables in our statistical models of drought tolerance, along with units, definitions, and sample sizes.",
         escape=FALSE) %>%
@@
 
-*Sample sizes are after removal of outliers. Dashes indicate that the variable was available for all records. For resistance, sample sizes of independent variables to the $Rt$ model.
+Sample sizes are after removal of outliers. Dashes indicate that the variable was available for all records. 
+*Sample sizes of independent variables refer to the $Rt$ model.
 
@@
 **Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses (b).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
 2014).
-Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric.
+Focal droughts are indicated by dashed lines, and shading indicates the pre- and post- drought periods used in calculations of the resistance metric.
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b) show the distribution of resistance values for each drought.
+Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought.
 
@@
 In these plots, heights are slightly offset for visualization purposes.
-Asterisks indicate signficant differences between the top and bottom of the height profile.
+Asterisks indicate significant differences between the top and bottom of the height profile.
 Also shown is (d) tree heights by 2018 crown position, with letters indicating significance groupings.
@@
 
-**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Species codes are given in Table 2.
+**Figure 3. Figure 3. Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
 
 **Figure 4.
-Visualization of best models for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
+Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
 Model coefficients are given in Tables S8 and S10-11.
@@
 Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
-Model coefficients are given in Table S8.
+Model coefficients are given in Tables S8 and S10-11.
 ](tables_figures/publication/Figure4_model_vis_all.png)

### 2020-09-11 manuscript/response_to_reviews.Rmd
update
remove FIg. S6
closes #141

@@
 
+*Note that we dropped former Fig.
+S6, which was not essential and not added in direct response to any reviewer comments.
+Keeping this figure would have required adding parallel figures for recovery and resilience, which we felt would be adding too many marginally valuable appendix figures.*
 ...
@@
 
+https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/142
+
 - LMA is not shown in Figure S4, and in Table 2 there are no tests so we cannot judge whether there are differences among species for that leaf trait. Please, add also LMA to Figure S4.
@@
 *Ian, we should add both LMA and WD here.*
+
+https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/142
 

### 2020-09-11 manuscript/climate_sensitivity_draft.Rmd
remove FIg. S6
closes #141

@@
 
-Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Figs. **4, S6**). 
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
 Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9). 
@@
 
-$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Figs. **4, S6**, Table S8-S9). 
+$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Fig. **4**, Table S8-S9). 
 The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S8).
@@
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Figs.
-**4, S6**; Tables 1,S4-S5,S8-S9).
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Fig.
+**4**; Tables 1,S4-S5,S8-S9).
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). 
@@
 **1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Fig.
-**S6**; Table S8).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S8).
 

### 2020-09-11 manuscript/response_to_reviews.Rmd
adjust AIC criteria

--- 2020-09-11 manuscript/response_to_reviews.Rmd
@@
 
-We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest” (previously “Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest”), according to the recommendations of reviewers.
+We have revised our manuscript, “Tree height and leaf drought tolerance traits shape growth responses across droughts in a temperate broadleaf forest”, according to the recommendations of reviewers.
+These changes have further strengthened the manuscript.
 
+A complete list of substantive changes is as follows:
+
+* Most importantly, we have added analyses parallel to those for resistance for both recovery and resilience.
+Because we feel that resilience is the more interesting and informative of these two, we have given it precedence for inclusion in main manuscript figures, where space is limiting.
+Specifically, we have added a density plot of resilience to Fig.1 (+ parallel plot for recovery in the SI), a plot of resilience by species to Fig.
+3 (+ parallel plot for recovery in the SI), and plots of the best models for both recovery and resilience to Fig.
+4.
+We have also added recovery and resilience to the hypothesis table (Table 1), and their definitions to Table 3.
+We have also modified the text throughout and included relevant tables presenting statistics in the supplementary info.
+
+* To account for the fact that we now consider three main response variables, and in response to a comment from R3, we slightly modified the criteria for inclusion of traits in the main statistical models.
+As now stated in the methods, "Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
+While we tested the significance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below)."
+
+* To keep the number of SI items to a reasonable number, we dropped former Fig.
+S6, which was not essential and not added in direct response to any reviewer comments.
+Keeping this figure would have required adding parallel figures for recovery and resilience, which we felt would be adding too many marginally valuable appendix figures.
 ...
@@
 
-**We now define AICc at first mention, which is in the paragraph above.
-The sentence now reads "We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection,..." In the sentence referenced here, we define $\Delta$AIC: "Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4)."**
+**We now clearly define AICc at first mention, which is in the paragraph above.
+The sentence now reads "R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].
+Individual model terms were considered significant when their addition to a model improved fit at $\Delta$AICc $\ge$ 2.0, where $\Delta$AICc is the difference in AICc between models with and without the trait."**
 
-**We have given careful thought to the trait variable selection criteria.
-It is important to note that selection of traits is not a test of their significance in the final model; it is simply a way of objectively reducing the number of candidate traits in these models.
-While we fully agree that $\Delta$AICc $\ge$ 2 is the appropriate criterion for assessing significance, it is not necessarily appropriate for candidate variable selection because variable significance changes when considered independently versus in the context of a full model.
-That is, the significance of variables--particularly those with marginal significance-- cannot be reliably assessed outside the context of a model that simultaneously accounts for other relevant variables.
-As a case in point, $\pi_{tlp}$ would be removed by a $\Delta$AICc $\ge$ 2 criterion for variable selection, but is included in the best model for all droughts combined, as well as the 1977 drought.
-While it's effects are marginal (i.e., included in some but not all top models), we feel that it is important to include it as a candidate variable both for the statistical considerations mentioned above, and because we expect that a lot of readers will be very interested to see a full assessment of this trait, which is currently of great interest to the research community.**
-
-*([issue #134](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/134))*
+**ALthough the $\Delta$AICc $\ge$ 1.0 criteria was used only for variable selection and not as a significance test, we have dropped it and now use only $\Delta$AICc $\ge$ 2.0.
+To account for the fact that we now consider three main response variables, we modified the criteria for inclusion of traits in the main statistical models.
+As now stated in the methods, "Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
+While we tested the significance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below)."**
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
remove FIg. S6
closes #141
adjust AIC criteria

@@
 
-*$\Delta$AICc > 1: variable meets $\Delta$AICc criterion for inclusion in full model
-
-**$\Delta$AICc > 2: variable is considered significant as an individual predictor (and meets $\Delta$AICc criterion for inclusion in full model)
+**$\Delta$AICc > 2: variable considered significant as an individual predictor
 
@@
 \newpage
-
-![**Figure S6.
-Visualization of the best model ($\Delta$AICc=0) with data for all droughts combined.** Model coefficients are given in Table S8.](tables_figures/publication/FigureS6_best_model_with_data.png)
-
-\newpage
 ## Appendix S1. Further Package Citations

### 2020-09-11 manuscript/climate_sensitivity_draft.Rmd
adjust AIC criteria

@@
 All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
-We used Akaike information criterion with correction for small sample sizes (AICc) to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].
-AICc refers to a corrected version of AICc, and is best suited for small data sizes [see @brewer_relative_2016].
+We used Akaike information criterion with correction for small sample sizes [AICc; see @brewer_relative_2016] to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].
+Individual model terms were considered significant when their addition to a model improved fit at $\Delta$AICc $\ge$ 2.0, where $\Delta$AICc is the difference in AICc between models with and without the trait.
 
@@
 These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
-Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0, where $\Delta$AICc is the difference in AICc between models with and without the trait) in at least one drought year (Table S4).
-We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models.
+Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
+While we tested the signficance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below).
 
@@
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ - the traits that qualified for inclusion in the full model (Table S4) - were both negatively correlated to drought resistance (Fig.
-**4**; Tables 1,S4-S5,S8-S9).
+In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance and resilience (Fig. **4**; Tables 1, S4-S11). 
 $PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). 

### 2020-09-11 manuscript/climate_sensitivity_draft.Rmd
renumber & add new figures to SI, adjust elsewhere accordingly
#131

@@
 In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.
-Visualization of the best mixed effects model per drought scenario (Fig.
-**4**) was created by the visreg package [@R-visreg].
+Visualization of the best mixed effects model per drought scenario was created by the visreg package [@R-visreg].
 
@@
 At the community level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
-**2b**).
+**1b**).
 Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$.
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
+
+*here, cite Figs. 1c, S6*
 
@@
 
-*Species' traits and drought resistance*
+*Species' traits and drought tolerance*
 
@@
 **S4** as characterized by the agricolae package @R-agricolae).
-Drought resistance also varied across species, overall and in each drought year (Fig.
-**3**).
+Drought tolerance also varied across species, overall and in each drought year (Figs.
+**3, S7**).
 Significant differences in $Rt$ across species were most pronounced in 1966 with a total of seven distinct groupings, while 1977 had four and 1999 had two.

### 2020-09-11 manuscript/tables&figures.Rmd
Update tables&figures.Rmd
renumber & add new figures to SI, adjust elsewhere accordingly
#131

@@
 Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought.
+See Fig.
+S6 for parallel plot for recovery ($Rc$).
 
@@
 
-**Figure 3. Figure 3. Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
+**Figure 3.
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
+See Fig.
+S7 for parallel plot for recovery ($Rc$).
 
@@
 Figure modified from Helcoski *et al.* (2019).
-Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought.](tables_figures/publication/Figure1.png)
+Density plots (b-c) show the distribution of $Rt$ and $Rs$ values for each drought.
+See Fig.
+S6 for parallel plot for recovery ($Rc$).
+](tables_figures/publication/Figure1.png)
 
@@
 ![**Figure 3.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.](tables_figures/publication/Figure3_Rt-Rs_across_sp.png)
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
+See Fig.
+S7 for parallel plot for recovery ($Rc$).](tables_figures/publication/Figure3_Rt-Rs_across_sp.png)
 

### 2020-09-11 manuscript/climate_sensitivity_SI.Rmd
renumber & add new figures to SI, adjust elsewhere accordingly
#131
Updated "year" coefficients per #140

@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
+Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
+Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small **(1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models)**.
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
+Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small 
+**(1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172)**.
 
@@
 \newpage
+
+![**Figure S6. Drought recovery, $Rc$, across species for the three focal droughts.**](tables_figures/publication/FigureS6_density_plot_recovery.png)
+
+\newpage
+
+![**Figure S7. Drought recovery, $Rc$, across species for the three focal droughts.**](tables_figures/publication/FigureS7_Rc_across_sp.png)
+
+\newpage
 ## Appendix S1. Further Package Citations

### 2020-09-11 scripts/climate_sensitivity_analysis.R
Update climate_sensitivity_analysis.R

@@ -622,18 +622,19 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #########################################################################################
 #3. mixed effects model for output of #2.
 ##start here if just re-running model runs ####
-metric <- "resilience" #resistance, recovery, or resilience
-arima_vals <- FALSE #TRUE or FALSE
+metric <- "resistance" #resistance, recovery, or resilience
+arima_vals <- TRUE #TRUE or FALSE
 
 trees_all_sub <- read.csv(
   paste0("manuscript/tables_figures/trees_all_sub_",
          metric, ".csv"), stringsAsFactors = FALSE)
 
 if(metric=="resistance" & arima_vals){
-  trees_all_sub <- read.csv(
-    paste0("manuscript/tables_figures/trees_all_sub_", metric, 
-           "arimaratio.csv"), 
-           stringsAsFactors = FALSE)
+  trees_all_sub <- 
+    read.csv("manuscript/tables_figures/trees_all_sub_arimaratio.csv",
+             stringsAsFactors = FALSE)
+  library(data.table)
+  setnames(trees_all_sub, old=c("resist.value"), new=c("metric.value"))
 }
 
 trees_all_sub$year <- as.character(trees_all_sub$year)
@@ -1235,9 +1236,16 @@ for(i in seq(along=patterns)){
 }
 
 #this table is used to fill in Table S6 
-write.csv(coeff_new, 
-          paste0("manuscript/tables_figures/tested_traits_best_coeff_",
-          metric, "_CPout.csv"), row.names=FALSE)
+if(metric == "resistance" & arima_vals){
+  write.csv(coeff_new, 
+            paste0("manuscript/tables_figures/tested_traits_best_coeff_",
+                   metric, "_CPout_arimaratio.csv"), row.names=FALSE)
+} else {
+  write.csv(coeff_new, 
+            paste0("manuscript/tables_figures/tested_traits_best_coeff_",
+                   metric, "_CPout.csv"), row.names=FALSE)
+}
+
 
 ## END OF NORMAL ANALYSIS. 3d and 3e are extra
 

### 2020-09-11 scripts/climate_sensitivity_analysis.R
Update climate_sensitivity_analysis.R

@@ -622,8 +622,8 @@ for(i in seq(along=trees_all_sub[,c(5:9,16)])){
 #########################################################################################
 #3. mixed effects model for output of #2.
 ##start here if just re-running model runs ####
-metric <- "resistance" #resistance, recovery, or resilience
-arima_vals <- TRUE #TRUE or FALSE
+metric <- "resilience" #resistance, recovery, or resilience
+arima_vals <- FALSE #TRUE or FALSE
 
 trees_all_sub <- read.csv(
   paste0("manuscript/tables_figures/trees_all_sub_",
@@ -869,18 +869,22 @@ best_mod_traits <- data.frame("best_model" = NA,
 ## signs across all drought years. See Issues #95 + #133 on github.
 top_vars_poss <- c(unique(cand_full$variable))
 
-sum_mod_traits <- data.table(sum_mod_traits)
+#originally we were taking directly from the candidate traits, but
+#we then decided to only use TLP and PLA across all metrics in order to
+#be consistent.
+# sum_mod_traits <- data.table(sum_mod_traits)
+# 
+# top_vars <- c()
+# for(h in 1:length(top_vars_poss)){
+#   b <- sum_mod_traits[variable == top_vars_poss[h], 
+#                       .(coef_all, coef_1966, coef_1977, coef_1999)]
+#   issame <- as.logical(b[,1] == b[,2] & b[,2] == b[,3] & b[,3] == b[,4])
+#   
+#   if(issame == TRUE){
+#     top_vars <- c(top_vars, top_vars_poss[h])
+#   }
+# }
 
-top_vars <- c()
-for(h in 1:length(top_vars_poss)){
-  b <- sum_mod_traits[variable == top_vars_poss[h], 
-                      .(coef_all, coef_1966, coef_1977, coef_1999)]
-  issame <- as.logical(b[,1] == b[,2] & b[,2] == b[,3] & b[,3] == b[,4])
-  
-  if(issame == TRUE){
-    top_vars <- c(top_vars, top_vars_poss[h])
-  }
-}
 
 top_vars <- c("mean_TLP_Mpa", "PLA_dry_percent")
 

### 2020-09-12 manuscript/tables&figures.Rmd
Updated Table 3 and updated .Rmd #139

@@
 
-Sample sizes are after removal of outliers. Dashes indicate that the variable was available for all records. 
-*Sample sizes of independent variables refer to the $Rt$ model.
+Sample sizes are after removal of outliers.
+Dashes under **Independent variables** indicate that the variable was available for all records.
+Xylem porosity sample sizes by vessel arrangement are for all drought scenarios combined.
 

### 2020-09-12 scripts/climate_sensitivity_analysis.R

@@ -637,6 +637,13 @@ if(metric=="resistance" & arima_vals){
   setnames(trees_all_sub, old=c("resist.value"), new=c("metric.value"))
 }
 
+#determine N trees for all years with semi-ring.
+# library(data.table)
+# g <- as.data.table(trees_all_sub)
+# g <- g[,rp := ifelse(sp=="juni", "semi-ring", rp)]
+# g[,.N,by=.(rp)]
+# trees_all_sub <- as.data.frame(g)
+
 trees_all_sub$year <- as.character(trees_all_sub$year)
 
 x1966 <- trees_all_sub[trees_all_sub$year == 1966, ]

### 2020-09-13 manuscript/climate_sensitivity_SI.Rmd
a couple corrections

@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small **(1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models)**.
+Year was included in the model for all drought years and appeared in all its top models **(1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models)**.
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small 
+Year was included in the model for all drought years and apperaed in all its top models
 **(1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172)**.
@@
 \newpage
-![**Figure S1. Time series of Palmer Drought Severity Index (PDSI) foreach focal drought year $\pm$ 2 years**](tables_figures/publication/figureS1_drought_plot.jpg)
+![**Figure S1. Time series of Palmer Drought Severity Index (PDSI) for each focal drought year $\pm$ 2 years**](tables_figures/publication/figureS1_drought_plot.jpg)
 

### 2020-09-13 manuscript/climate_sensitivity_draft.Rmd
a bit of work on incorporating Rc and Rs
#131

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) | 6,490	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,490 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
 Summary|  	200 | 	No. of Tables | 	3  
-Introduction|	1,264 |	No of Supporting Information files | 	14
+Introduction|	1,264 |	No of Supporting Information files | 	19
 Materials and Methods |	2,228	||	 
@@
 
-For each drought year, we calculated a metric drought resistance ($Rt$) as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought [@lloret_components_2011].
+For each drought year, we calculated metrics of drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs$), following @lloret_components_2011.
+We defined $Rt$ as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought.
 Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
@@
 Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
-In this study we focus exclusively on drought resistance metrics ($Rt$ or $Rt_{ARIMA}$), and not on the resilience metrics described in @lloret_components_2011, because (1) we would expect resilience to be controlled by a different set of mechanisms, and (2) the findings of @desoto_low_2020 suggest that $Rt$ is a more important drought response metric for angiosperms in that low resistance to moderate droughts was a better predictor of mortality during subsequent severe droughts than the resilience metrics.
+*describe Rc, Rs*
 
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
-
-*here, cite Figs. 1c, S6*
+*here, discuss $Rc$ and $Rs$, cite Figs. 1c, S6*
 
@@
 That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9).
+*results for Rc, Rs*
 
-*Tree height, microenvironment, and drought resistance*
+*Tree height, microenvironment, and drought tolerance*
 
 Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
-Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model (($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9). 
+Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model ($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9). 
 The same held true for 1966 individually.
@@
 The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020].
-Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
+Drought tolerance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$).
@@
 
-Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance. 
+Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*. 
 Mechanistically, this is consistent with, and reinforces, previous findings that it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
@@
 Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected in terms of growth during the drought year. 
-**Resilience and survival are both linked to resistance [@desoto_low_2020; @gessler_way_2020], implying that the same factors may influence these.
-Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019].**
+Survival is linked to resistance and resilience [**CHECK REFS**] [@desoto_low_2020; @gessler_way_2020], implying it may be influenced by the same factors. 
+Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought survival remains to be tested, taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019]. 
 As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].
-At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F. grandifolia*) could in turn become more susceptible. 
-Here, the relative importance of tree height *per se* versus crown exposure becomes crucial, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
+At this point, species whose drought tolerance relies in part on existence within a buffered microenvironment (e.g., *F. grandifolia*) could in turn become more susceptible. 
+Here, the relative importance of tree height *per se* versus crown exposure becomes crucial, shaping whether the dominant trees of shorter canopies are significantly more drought tolerant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
 Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 

### 2020-09-13 scripts/graphs_for_manuscript.R
Updated per #142

@@ -286,8 +286,10 @@ ggplot(full) +
 library(data.table)
 library(ggplot2)
 library(agricolae)
-rt <- fread("manuscript/tables_figures/trees_all_sub.csv")
-traits_hydr <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_indvidual_level.csv?token=AJNRBEPQVCRRCIDQSBGNC2S7EC5A2")
+rt <- fread("manuscript/tables_figures/trees_all_sub_resistance.csv")
+
+##this is the INDIVIDUAL LEVEL DATA, not species
+traits_hydr <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_indvidual_level.csv?token=AJNRBEIWYZQPW2RYIPESJDS7M5SLQ")
 traits_hydr <- traits_hydr[,.(sp,PLA_dry_percent, mean_TLP_Mpa, 
                               WD_g_per_cm3, LMA_g_per_m2)]
 
@@ -297,9 +299,9 @@ var <- c("height.ln.m", "TWI.ln", "PLA_dry_percent", "mean_TLP_Mpa",
 for(i in 1:6){
    
    if(i %in% c(1,2)){
-      anovout <- aov(get(var[i]) ~ sp, data=rt)
+      anovout <- aov(rt[,get(var[i])] ~ sp, data=rt)
    } else {
-      anovout <- aov(get(var[i]) ~ sp, data=traits_hydr)
+      anovout <- aov(traits_hydr[,get(var[i])] ~ sp, data=traits_hydr)
    }
    hsdout <- HSD.test(anovout, "sp", group=TRUE)
    grouptab <- hsdout$groups
@@ -329,11 +331,14 @@ for(i in 1:2){
          ylab(ylabs[i]) +
          xlab("Species") +
          theme_minimal() +
-         theme(axis.text = element_text(size=12),
-               axis.text.x=element_text(angle=90),
-               axis.title=element_text(size=12, face="bold"),
-               legend.text=element_text(size=12),
-               legend.title=element_text(size=12))
+         theme(legend.text=element_text(size=12),
+               legend.title=element_text(size=12),
+               axis.text.x=element_blank(),
+               axis.title.x=element_blank()
+               # axis.text = element_text(size=12),
[…]

### 2020-09-14 manuscript/response_to_reviews.Rmd
revise methods
#131

@@
 I agree that results comparing them are pretty similar (Table S6 vs Table S7) and that authors can use Rt.
+
+**Thank you.
+Because we had to cut content in order to present the recovery and resilience metrics (see below) while keeping within word limits, we have cut back on the reporting of $Rt_{ARIMA} results in the main text, but still show the full comparisons in the Appendix.
+Specifically, we now state in the following in the methods: "$Rt$ and $Rt_{ARIMA}$ were strongly correlated (Fig.
+**S5**), and showed similar responses to the independent variables of interest (cf.
+Tables S4-55, S8-S9).
+Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$.
+We therefore determined that use of 5-year means, as described above, were more appropriate metrics than those based on ARIMA projections." We then drop presentation of $Rt_{ARIMA}$ results in the results section.**
 

### 2020-09-14 manuscript/climate_sensitivity_draft.Rmd
Added quick thing to part of intro
revise methods
#131

@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-**(here would be a good place to comment on resilience.)**
+**(here would be a good place to comment on resilience. "What may help the tree in competition, therefore, can also be a deteriment to drought resilience.")**
 
@@
 The 1999 drought was preceded by wetter than average conditions until the previous June, but $PDSI$ plummeted below -3.0 in October 1998 and remained below this threshold through August 1999. 
+Following all three droughts, $PDSI$ rebounded to near-normal conditions in September or October (Fig. **S1**).
 
@@
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
-The resulting chronologies (Fig. **1a**) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.
+The resulting chronologies (Fig. **1a**) were published in Zenodo [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2019].
 
@@
 
-Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness. 
-Bark thickness was estimated from species-specific allometries based on the bark thickness data from the site [@andersonteixeira_size-related_2015].
-Specifically, we used linear regression on log-transformed data to relate $r_{bark}$ to diameter inside bark from 2008 data (Table S2), which were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
+Here, $Y$ denotes the year of interest, $r_{ring}$ denotes ring width derived from cores, and $r_{bark}$ denotes bark thickness, which was estimated from species-specific allometries based on the bark thickness data from the site [Table S2; @andersonteixeira_size-related_2015].
 
@@
 Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
-Species-specific height allometries were developed (Table S3) using log-log regression ($ln[H] \sim ln[DBH]$). 
+Species-specific height allometries were developed using log-log regression ($ln[H] \sim ln[DBH]$; Table S3). 
 For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species.
-We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$.
-The distribution of $H$ across drought years is shown in Fig. **S3**.
+We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$ (Fig. **S3**).
 
@@
 
-For each drought year, we calculated metrics of drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs$), following @lloret_components_2011.
+For each drought year, we calculated metrics of drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs$), following @lloret_components_2011 (Table 3).
 We defined $Rt$ as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought. 
-Thus, $Rt$ values <1 and >1 indicate growth reductions and increases, respectively.
-Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
-The two metrics were strongly correlated (Fig.
-**S5**).
-Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info.
-*describe Rc, Rs*
+$Rc$ was defined as the ratio of mean $BAI$ over the five years following the drought to $BAI$ of the drought year.
+$Rs$ was defined as the ratio of mean $BAI$ over the five years following the drought to that of the five years preceding the drought. 
+Thus, for all metrics, values <1 and >1 indicate growth reductions and increases, respectively.
 
-Analyses focused on testing the predictions presented in Table 1 with $Rt$ as the response variable, and then repeated using $Rt_{ARIMA}$ as the response variable. 
+Because these metric could potentially be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+$Rt$ and $Rt_{ARIMA}$ were strongly correlated (Fig. **S5**), and showed similar responses to the independent variables of interest (cf. Tables S4-55, S8-S9).
+Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$.
+We therefore determined that use of 5-year means, as described above, were more appropriate metrics than those based on ARIMA projections.
+
+Analyses focused on testing the predictions presented in Table 1 with $Rt$ (or $Rt_{ARIMA}$), $Rc$, or $Rs$ as the response variable. 
 Models were run for all drought years combined and for each drought year individually. 
@@
 Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
-While we tested the signficance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below).
+While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below).
 
-We then determined the top full models for predicting $Rt$ (or $Rt_{ARIMA}$). 
+We then determined the top full models for predicting each dependent variable. 
 To do so, we compared models with all possible combinations of candidate variables, including $ln[H]$*$ln[TWI]$ and species traits as specified above.
@@
 If the variable appeared in some but not all of these models, and its sign was consistent across models, we considered this partial support/rejection. 
-In presentation of the results below, we note instances where the $Rt_{ARIMA}$ model disagreed with the $Rt$ model, but otherwise do not discuss the $Rt_{ARIMA}$ model.
 Visualization of the best mixed effects model per drought scenario was created by the visreg package [@R-visreg].

### 2020-09-14 manuscript/climate_sensitivity_SI.Rmd
typo

@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and apperaed in all its top models
+Year was included in the model for all drought years and appeared in all its top models
 **(1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172)**.

### 2020-09-14 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
#131, #144

@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
-*here, discuss $Rc$ and $Rs$, cite Figs. 1c, S6*
+Recovery was generally strong, with $Rc$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **S6**), and $Rs$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **1c**).
 
@@
 That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9).
-*results for Rc, Rs*
+In contrast, both $Rc$ and $Rs$ varied across years, being highest in all of the top models (Tables S10-S11), in which both were highest in 1966, intermediate in 1977, and lowest in 1999.
 

### 2020-09-14 manuscript/climate_sensitivity_draft.Rmd
working on results
#131

@@
 If the variable appeared in some but not all of these models, and its sign was consistent across models, we considered this partial support/rejection.
-Visualization of the best mixed effects model per drought scenario was created by the visreg package [@R-visreg].
 
@@
 
-Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (Table 1; Fig. **4**). 
-Specifically, $ln[H]$ appeared, with a negative coefficient, in the best model ($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9). 
-The same held true for 1966 individually.
-For the 1977 drought, $ln[H]$ did not appear in the best model, but was included, with a negative coefficient, among the top models--*i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (Tables 1, S6-S9).
-For the 1999 drought, $ln[H]$ had no significant effect.
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (i.e., lower $Rt$) and less resilience following drought (i.e., lower $Rc$ and $Rs$; Table 1; Fig.
+**4**).
+Specifically, for $Rt$, $ln[H]$ appeared, with negative coefficient, in the best model ($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9).
+The same held true for 1966 individually, but there was no significant effect of $ln[H]$ for 1977 or 1999 individually.
+For $Rc$, $ln[H]$ appeared, with negative coefficient, in the best model without a $ln[H] *ln[TWI]$ interaction, for the three drought years together and for 1977, but not for 1966 or 1999.
+For $Rs$, again considering the best models without a $ln[H] *ln[TWI]$ interaction, there was a negative effect of $ln[H]$ for the three drought years together and for 1966 and 1977, and a non-significant negative trend in 1999.
 
+**RESUME HERE**
 $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Fig. **4**, Table S8-S9). 

### 2020-09-14 manuscript/tables&figures.Rmd
remove column for Rt_arima
@mcgregorian1 , I think it's overkill to list Rt_arima sample sizes in a separate column, so reverted back to former system for Rt.
#139
working on results
#131

--- 2020-09-11 manuscript/tables&figures.Rmd
@@
 
-Sample sizes are after removal of outliers. Dashes indicate that the variable was available for all records. 
+Sample sizes are after removal of outliers.
+Dashes for sample sizes of independent variables indicate that the variable was available for all records.
+Xylem porosity sample sizes are sums across all drought years.
+
 *Sample sizes of independent variables refer to the $Rt$ model.
@@
 ![**Figure 4.
-Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the bootpredictlme4 package.
+Visualization of top statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** For cases where the best model includes a DBH x TWI interaction ($Rc$ in all droughts and 1966, $Rs$ in all droughts and 1977), we plot the best model without the interaction.
+Visualization of the best mixed effects model per drought scenario was created by the visreg package [@R-visreg], and confidence intervals were defined via bootsrapping in the bootpredictlme4 package.
 Model coefficients are given in Tables S8 and S10-11.

### 2020-09-14 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 
-**RESUME HERE**
-$Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined (Fig. **4**, Table S8-S9). 
-The effect was also significant for 1977 and 1999 individually (Fig. **4**, Table S8).
-When $Rt_{ARIMA}$ was used as the response variable, the effect was significant in 1977, and included in some of the top models in 1966 and 1999 (Table S9).
-This negates the idea that trees in moist microsites would be less affected by drought. 
-Nevertheless, we tested for a $ln[H] *ln[TWI]$ interaction, a negative sign of which could indicate that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in drier microenvironments with a deeper water table.
-This hypothesis was rejected, as the $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top models in which it appeared (Tables 1, S8-S9).
-This term did appear with a positive coefficient in the best $Rt_{ARIMA}$ model for all years combined (Table S9), indicating that the negative effect of height on  $Rt$ was significantly stronger in wetter microhabitats.
+Trees in drier microsites showed greater growth declines during drought; i.e., $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig.
+**4**, Table S8-S9).
+The $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top $Rt$ models in which it appeared (Tables 1, S8-S9), rejecting the hypothesis that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in microenvironments with a deeper water table.
+In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. 4, Tables S10-S11).
+Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
+This implies that small trees had greater recovery and resilience in wetter microhabitats, whereas large trees had greater recovery and resilience in dry microhabitats.
 

### 2020-09-15 manuscript/climate_sensitivity_draft.Rmd
Added mean Rc and Rs #144

@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
-Recovery was generally strong, with $Rc$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **S6**), and $Rs$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **1c**).
+Recovery was generally strong, with $Rc$ averaging **1.55** in 1966, **1.42**in 1977, and **1.34** in 1999 (Fig.
+**S6**), and $Rs$ averaging **1.28** in 1966, **1.19**in 1977, and **1.12** in 1999 (Fig.
+**1c**).
 

### 2020-09-15 manuscript/climate_sensitivity_SI.Rmd
Added details to table s2, also updaed bark_regression table

@@
 colnames(bark) <- c("Species", "Equations", "$R^2$")
-kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific bark thickness regression equations") %>%
+kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific bark thickness regression equations. Units for r_bark and DBH are in mm.") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
 ```
+We used linear regression on log-transformed data to relate $r_{bark}$ to the diameter inside bark from 2008 data.
+These were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
+The equation for *Fagus grandifolia* because it functionally has negligble bark.
 

### 2020-09-15 scripts/climate_sensitivity_analysis.R
Update climate_sensitivity_analysis.R

@@ -309,7 +309,7 @@ bark$diam_nobark_2008.mm <- bark$DBH.mm.2008 - 2*bark$bark.depth.mm
 #3. Fit a linear model, and use model to predict log(bark.depth.mm)
 source_gist("524eade46135f6348140")
 ggplot(data = bark, aes(x = log(diam_nobark_2008.mm), y = log(bark.depth.mm))) +
-  stat_smooth_func(geom="text",method="lm",hjust=0.16, vjust=-0.5,parse=TRUE) +
+  stat_smooth_func(geom="text",method="lm",hjust=0.8, vjust=-0.5,parse=TRUE) +
   geom_smooth(method="lm", se=FALSE, color="black") +
   geom_point(color = "#0c4c8a") +
   theme_minimal() +

### 2020-09-15 manuscript/climate_sensitivity_draft.Rmd
update results
#131, #144

--- 2020-09-14 manuscript/climate_sensitivity_draft.Rmd
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.
-Recovery was generally strong, with $Rc$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **S6**), and $Rs$ averaging **##** in 1966, **##**in 1977, and **##** in 1999 (Fig. **1c**).
+Recovery was generally strong and complete, with $Rc$ averaging 1.55 in 1966, 1.42 in 1977, and 1.34 in 1999 (Fig.
+**S6**) and $Rs$ averaging 1.28 in 1966, 1.19 in 1977, and 1.12 in 1999 (Fig.
+**1c**).
 
@@
 **S4** as characterized by the agricolae package @R-agricolae).
-Drought tolerance also varied across species, overall and in each drought year (Figs.
-**3, S7**).
-Significant differences in $Rt$ across species were most pronounced in 1966 with a total of seven distinct groupings, while 1977 had four and 1999 had two.
-Averaged across all droughts, $Rt$ was lowest in *Liriodendron tulipifera* (mean $Rt$ = **0.66**) and highest in *Fagus grandifolia* (mean $Rt$ = **0.99**).
+Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 
+Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *Liriodendron tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
+These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **3**).
 
-Wood density, $LMA$, and xylem porosity were all poor predictors of $Rt$ (Tables 1,S4-S5). 
-Wood density and $LMA$ were never significantly associated with $Rt$ in the single-variable tests and were therefore excluded from the full models. 
-Xylem porosity was also excluded from the full models, as it had no significant influence for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Table S4).
-It is noteworthy that the two diffuse-porous species in our study, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-**3**), further refuting the idea that xylem porosity is a useful predictor of $Rt$ in the context of this study.
+Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1,S4-S5). 
+Wood density and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 
+Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
+In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. 3).
 
-In contrast, $PLA_{dry}$, and $\pi_{tlp}$ were both negatively correlated to drought resistance and resilience (Fig. **4**; Tables 1, S4-S11). 
-$PLA_{dry}$ had a significant influence, with negative coefficient, in full models for the three droughts combined and for the 1966 drought individually (Fig. **4**; Tables S8-S9). 
-For 1977 and 1999, it was included with a negative coefficient in some of the top models (Tables S8-S9).
+Drought resistance and resilience, but not recovery, were negatively correlated with $PLA_{dry}$ and $\pi_{tlp}$ (Fig. **4**; Tables 1, S4-S11). 
+For $Rt$, $PLA_{dry}$ had a significant influence, with negative coefficient, in top models for the three droughts combined and for the 1966 drought individually (Fig.
+**4**; Tables S8-S9).
+It was also included in some of the top models for 1999 (Tables S8-S9).
 $\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig.
@@
 It was also included in some of the top models for 1999 (Tables S8-S9).
+
+Recovery was not significantly correlated with either $PLA_{dry}$ or $\pi_{tlp}$. 
+There was only one best $Rc$ model containing one of these terms ($\pi_{tlp}$ in 1977 drought), but in no instance was one of these terms included in all top models (i.e., at $\Delta$AICc<2).
+
+For $Rs$, $PLA_{dry}$ was in the best models for the three droughts combined and for the 1966 drought individually, and some of the top models for 1977 and 1999 (Fig.
+**4**; Table S11); however, its effects were not significant atat $\Delta$AICc<2.
+$\pi_{tlp}$ was in the best models for the three droughts combined and for 1966 and 1999 individually, and in one of the top models for 1977 (Fig.
+**4**; Table S11).
+It's effects were significant at $\Delta$AICc<2 for 1999 only.
 

### 2020-09-15 manuscript/climate_sensitivity_SI.Rmd
fixed tables 2-3 units specification
#143

--- 2020-09-14 manuscript/climate_sensitivity_SI.Rmd
@@
 colnames(bark) <- c("Species", "Equations", "$R^2$")
-kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2. Species-specific bark thickness regression equations") %>%
+kable(bark, booktabs=TRUE, format = "latex", escape=FALSE, caption = "Table S2.
+Species-specific regression equations for bark thickness (mm) as a function of diameter at breast height without bark (mm).") %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
 ```
+We used linear regression on log-transformed data to relate $r_{bark}$ to the diameter inside bark from 2008 data.
+These were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
+No bark correction was applied for *Fagus grandifolia*, which has thin bark.
 
@@
 colnames(heights)[3] <- "$R^2$"
-kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S3. Species-specific height regression equations", escape = FALSE) %>%
+kable(heights, booktabs=TRUE, format = 'latex', caption = "Table S3. Species-specific regression equations for height (m) as a function of DBH (cm)", escape = FALSE) %>%
   kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position", protect_latex = TRUE)

### 2020-09-16 manuscript/climate_sensitivity_draft.Rmd
Added small updates to text and table

@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought.
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015], would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015], nor does it hold in the global context [@wheeler_variations_2007; @olson_xylem_2020], and it does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
@@
 All models included fixed effects of independent variables of interest (Tables 1,3) as specified below.
-All variables across all best models had variance inflation factors <1.2 (1 +/- 0.019). 
+All variables across all best models had variance inflation factors between 1 and 1.045.
 We used Akaike information criterion with correction for small sample sizes [AICc; see @brewer_relative_2016] to assess model selection, and conditional/marginal R-squared to assess model fit as implemented in the AICcmodavg package in R [@R-AICcmodavg].

### 2020-09-16 scripts/climate_sensitivity_analysis.R
Reran variance inflation factors for all vars

@@ -1102,48 +1102,70 @@ write.csv(top_models, file_top, row.names=FALSE)
 # In a way, this is a companion to a correlation plot.
 # https://www.statisticshowto.com/variance-inflation-factor/
 
+if(metric=="resistance" & arima_vals==TRUE){
+  best_mod_traits <- read.csv(paste0("manuscript/tables_figures/tested_traits_best_", metric, "_CPout_arimaratio.csv"), stringsAsFactors = FALSE)
+} else {
+  best_mod_traits <- read.csv(paste0("manuscript/tables_figures/tested_traits_best_", metric, "_CPout.csv"), stringsAsFactors = FALSE)
+}
+
+#we have to remove the interaction as per Github issue #
+##leaving it in gives us VIFs of >10 and 40
 input <- best_mod_traits$best_model
 best_mod_traits$best_model <- gsub("\\*", "\\+", input)
 
 hazel_vif <- NULL
 for(i in 1:nrow(best_mod_traits)){
-  mod <- lmer(best_mod_traits$best_model[i], 
-                       data = model_df[[i]],
-                        REML=TRUE,
-                       control = 
-                 lmerControl(optimizer ="Nelder_Mead"))
-  
-  hazel <- as.data.frame(car::vif(mod))
-  hazel$scen <- best_mod_traits$scenario[i]
-  if(ncol(hazel)>2){
-    hazel <- hazel[,c(3,4)]
+  if(str_count(best_mod_traits$best_model[i], "\\+") > 1){
+    mod <- lmer(best_mod_traits$best_model[i], 
+                data = model_df[[i]],
+                REML=TRUE,
+                control = 
+                  lmerControl(optimizer ="Nelder_Mead"))
+    
+    hazel <- as.data.frame(car::vif(mod))
+    hazel$scen <- best_mod_traits$scenario[i]
+    if(ncol(hazel)>2){
+      hazel <- hazel[,c(3,4)]
+    } else {
+      hazel <- hazel[,c(1,2)]
+    }
+    colnames(hazel) <- c("metric", "scen")
+    
   } else {
-    hazel <- hazel[,c(1,2)]
+    hazel <- data.frame(metric=NA, scen=best_mod_traits$scenario[i])
   }
-  colnames(hazel) <- c("metric", "scen")
+  
   hazel_vif <- rbind(hazel_vif, hazel)
 }
+hazel_vif$var <- rownames(hazel_vif)
+hazel_vif$var <- ifelse(hazel_vif$var=="1", "<2 model terms, can't do VIF",
[…]

### 2020-09-16 scripts/climate_sensitivity_analysis.R
Update climate_sensitivity_analysis.R

@@ -1108,8 +1108,10 @@ if(metric=="resistance" & arima_vals==TRUE){
   best_mod_traits <- read.csv(paste0("manuscript/tables_figures/tested_traits_best_", metric, "_CPout.csv"), stringsAsFactors = FALSE)
 }
 
-#we have to remove the interaction as per Github issue #
-##leaving it in gives us VIFs of >10 and 40
+#we have to remove the interaction as per Github issue #97
+##leaving it in gives us VIFs of >10 and 40. This is because a model with
+##an interaction includes its consitutent covariates, so by default those 
+##two will be correlated (e.g. correlate "height*TWI" with "height")
 input <- best_mod_traits$best_model
 best_mod_traits$best_model <- gsub("\\*", "\\+", input)
 

### 2020-09-17 manuscript/climate_sensitivity_SI.Rmd
Added small updates to text and table
Small updates to text and table captions

@@
 We used linear regression on log-transformed data to relate $r_{bark}$ to the diameter inside bark from 2008 data.
-These were then used to determine $r_{bark}$ in the $DBH$ reconstruction.
+These were then used to determine $r_{bark}$ in the $DBH_{Y}$ reconstruction (DBH in year Y).
 No bark correction was applied for *Fagus grandifolia*, which has thin bark.
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small **(1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models)**.
+Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and appeared in all its top models **(1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models)**.
+Year was included in the model for all drought years and appeared in all its top models (1966: 0, 1977: -0.14, 1999: -0.217; same values in all top models).
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and appeared in all its top models
-**(1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172)**.
+Year was included in the model for all drought years and appeared in all its top models (1966: 0, 1977: -0.099, -0.099, -0.099, -0.097, -0.097; 1999: -0.174, -0.174, -0.174, -0.173, -0.172).
 

### 2020-09-17 manuscript/climate_sensitivity_draft.Rmd
Small updates to text and table captions

@@
 That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9).
-In contrast, both $Rc$ and $Rs$ varied across years, being highest in all of the top models (Tables S10-S11), in which both were highest in 1966, intermediate in 1977, and lowest in 1999.
+In contrast, "drought year" varied for both $Rc$ and $Rs$, and were present in all of the top models for the combined scenario (Tables S10-S11).
+Between both metrics, coefficients for "drought year" were highest in 1966, intermediate in 1977, and lowest in 1999.
 
@@
 It was also included in some of the top models for 1999 (Tables S8-S9).
-$\pi_{tlp}$ was included with a negative coefficient in the best model for both all droughts combined and for the 1977 drought individually (Fig.
+$\pi_{tlp}$ was included with a negative coefficient in the best model for the combined droughts scenario and for the 1977 drought individually (Fig.
 **4**; Table S8), although its influence was not significant at $\Delta$AICc<2.
@@
 For $Rs$, $PLA_{dry}$ was in the best models for the three droughts combined and for the 1966 drought individually, and some of the top models for 1977 and 1999 (Fig.
-**4**; Table S11); however, its effects were not significant atat $\Delta$AICc<2.
+**4**; Table S11); however, its effects were not significant at $\Delta$AICc<2.
 $\pi_{tlp}$ was in the best models for the three droughts combined and for 1966 and 1999 individually, and in one of the top models for 1977 (Fig.
 **4**; Table S11).
-It's effects were significant at $\Delta$AICc<2 for 1999 only.
+However, its effects were significant at $\Delta$AICc<2 for 1999 only.
 
@@
 Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*. 
-Mechanistically, this is consistent with, and reinforces, previous findings that it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
+Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
 Taller trees also face dramatically different microenvironments (Fig. **2**). 

### 2020-09-18 manuscript/climate_sensitivity_SI.Rmd
Added citation names for R packages

@@
 
-[@R-base; @R-broom; @R-car; @R-cowplot; @R-data.table; @R-devtools; @R-dplR; @R-dplyr; @R-extrafont; @R-ggplot2; @R-ggpubr; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lubridate; @R-MuMIn; @R-piecewiseSEM; @R-png; @R-purrr; @R-raster; @R-rasterVis; @R-RCurl; @R-readxl; @R-reshape2; @R-rgdal; @R-rgeos; @R-rmarkdown; @R-sf; @R-stringi; @R-stringr; @R-tidyr]
+R base [@R-base]; broom [@R-broom]; car [@R-car]; cowplot [@R-cowplot]; data.table [@R-data.table]; devtools [@R-devtools]; dplR [@R-dplR]; dplyr [@R-dplyr]; extrafont [@R-extrafont]; ggplot2 [@R-ggplot2]; ggpubr [@R-ggpubr]; ggthemes [@R-ggthemes]; gridExtra [@R-gridExtra]; knitr [@R-knitr]; lubridate [@R-lubridate]; MuMIn [@R-MuMIn]; piecewiseSEM [@R-piecewiseSEM]; png [@R-png]; purrr [@R-purrr]; raster [@R-raster]; rasterVis [@R-rasterVis]; RCurl [@R-RCurl]; readxl [@R-readxl]; reshape2 [@R-reshape2]; rgdal [@R-rgdal]; rgeos [@R-rgeos]; rmarkdown [@R-rmarkdown]; sf [@R-sf]; stringi [@R-stringi]; stringr [@R-stringr]; tidyr [@R-tidyr]

### 2020-09-20 manuscript/response_to_reviews.Rmd
Added content to respones to reviews

@@
 
-*Ian, please check this and respond.*
-
-https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/142
+**In Figure S4, letters express signficiantly different groupings per variable.
+For example, the PLA values for a species in group "b" are significantly different from the PLA values for a species in group "c".
+We have added the following context to the caption: "Letter groupings do not transfer across variables (boxplots)."**
 
@@
 
-*Ian, we should add both LMA and WD here.*
-
-https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/142
+**Thank you for addressing this. We have now added LMA and WD to the figure, similar to the other variables.**
 

### 2020-09-20 manuscript/climate_sensitivity_draft.Rmd
Added small clarifier on small trees

@@
 
-Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*. 
+Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*--when compared to smaller trees.
 Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
-This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013, and these are then more vulnerable when drought hits.
+This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013], and these are then more vulnerable when drought hits.
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.

### 2020-09-20 manuscript/response_to_reviews.Rmd
Added clarification for response to reviewer 3

@@
 For example, the PLA values for a species in group "b" are significantly different from the PLA values for a species in group "c".
+However, group "b" for PLA has no bearing on group "b" for LMA, e.g., or any other variable.
 We have added the following context to the caption: "Letter groupings do not transfer across variables (boxplots)."**
@@
 
-**ALthough the $\Delta$AICc $\ge$ 1.0 criteria was used only for variable selection and not as a significance test, we have dropped it and now use only $\Delta$AICc $\ge$ 2.0.
+**Although the $\Delta$AICc $\ge$ 1.0 criteria was used only for variable selection and not as a significance test, we have dropped it and now use only $\Delta$AICc $\ge$ 2.0.
 To account for the fact that we now consider three main response variables, we modified the criteria for inclusion of traits in the main statistical models.

### 2020-09-21 manuscript/climate_sensitivity_SI.Rmd
Added content to respones to reviews
update year coefficients
#145

@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years, but its effect was not included in any top models, and coefficients were small (1966: 0, 1977: -0.019, 1999: -0.005; same values in all top models).
+Year was included in the model for all drought years and appeared in all its top models, but coefficients were small (1966: 0, 1977: -0.019, 1999: -0.005; same values in all top models).
 
@@
 $R^2$ refers to conditional $R^2$. 
-Year was included in the model for all drought years and apperaed in all its top models, but coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
+Year was included in the model for all drought years and appeared in all its top models, but coefficients were small (1966: 0, 1977: -0.03, 1999: 0.008; same values in all top models).
 
@@
 Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
-](tables_figures/publication/FigureS4_traits_signif.png)
+Letter groupings do not transfer between variables.](tables_figures/publication/FigureS4_traits_signif.png)
 

### 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
update year coefficients
#145

@@
 
-Because these metric could potentially be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
+Because these metrics could potentially be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$).
 $Rt$ and $Rt_{ARIMA}$ were strongly correlated (Fig. **S5**), and showed similar responses to the independent variables of interest (cf. Tables S4-55, S8-S9).
@@
 
-At the community level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig.
-**1b**).
+At the community level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig. **1b**). 
 Across the entire study period (1950-2009), the focal drought years were the three years with the largest fraction of trees exhibiting $Rt \le 0.7$. 
@@
 
-In the context of the multivariate model, $Rt$ did not vary across drought years.
-That is, drought year as a variable did not appear in any of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S9).
-In contrast, "drought year" varied for both $Rc$ and $Rs$, and were present in all of the top models for the combined scenario (Tables S10-S11).
-Between both metrics, coefficients for "drought year" were highest in 1966, intermediate in 1977, and lowest in 1999.
+In the context of the multivariate models, all response variables varied across drought years.
+That is, in models with all drought years combined, year was present in all of the top models -- *i.e.*, models that were statistically indistinguishable ($\Delta$AICc<2) from the best model (see footnotes on Tables S8-S11).
+For $Rt$, differences among drought years were small (<0.02; Table S8). 
+In contrast, differences among years were larger for $Rc$ and $Rs$, with coefficients for "drought year" highest in 1966, intermediate in 1977, and lowest in 1999.
 

### 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
minor fixes to results

@@
 
-Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (i.e., lower $Rt$) and less resilience following drought (i.e., lower $Rc$ and $Rs$; Table 1; Fig.
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (i.e., lower $Rt$) and less rebound following drought (i.e., lower $Rc$ and $Rs$; Table 1; Fig.
 **4**).
@@
 
-Species, as a factor in ANOVA, had significant influence (p<0.05) on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
+Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
 **S4** as characterized by the agricolae package @R-agricolae).
@@
 
-Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1,S4-S5). 
+Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1, S4-S5). 
 Wood density and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 

### 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 **2**).
-There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
-The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020].
+There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.**need something here on Rs** The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020].
 Drought tolerance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.

### 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
add Powers ref

@@
 Both traits correlate with hydraulic vulnerability and drought tolerance as part of unified plant hydraulic systems [@scoffoni_leaf_2014; @bartlett_correlations_2016; @zhu_leaf_2018; @farrell_does_2017].
-The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain the drought tolerance of tree growth remains untested.
+The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain the drought tolerance of tree growth remains untested [but see @powers_catastrophic_2020 for $\pi_{tlp}$ link to mortality].
 
@@
 Survival is linked to resistance and resilience [**CHECK REFS**] [@desoto_low_2020; @gessler_way_2020], implying it may be influenced by the same factors. 
-Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought survival remains to be tested, taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019]. 
+Indeed, while no link between $PLA_{dry}$ or $\pi_{tlp}$ on drought survival has been established [but see @powers_catastrphic_2020], taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019].
 As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].

### 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
revised discussion
#131

--- 2020-09-21 manuscript/climate_sensitivity_draft.Rmd
@@
 
-Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig.
-**4**).
+Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**).
 The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
-Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig.
-**2**).
+Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). 
 There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 **1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$: 11-12% for all droughts combined, and 18-25% for each individual drought (Table S8).
+It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).
 
-Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--*and also recovery and resilience*--when compared to smaller trees.
+Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--and also recovery and resilience--when compared to smaller trees.
 Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
@@
 **3**) and associated traits at a single site [see also @elliott_forest_2015].
-Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
-Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought tolerance (Table 1).
+Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
 **3**).
@@
 Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought tolerance (Fig. **4**; see discussion below). 
 A second explanation of why *F.
-grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig.
+grandifolia* trees at this particular site had higher $Rt$ and $Rs$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig.
 **S4**).
@@
 **2**).
-Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
+Thus, the high $Rt$ and $Rs$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.
 
@@
 As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here.
-Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected in terms of growth during the drought year. 
-Survival is linked to resistance and resilience [**CHECK REFS**] [@desoto_low_2020; @gessler_way_2020], implying it may be influenced by the same factors. 
-Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought survival remains to be tested, taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019]. 
+Our results show that taller, more exposed trees and species with less drought-tolerant leaf traits will be most affected in terms of both growth during the drought year and subsequent growth.
+Survival is linked to resistance and resilience [@desoto_low_2020; @gessler_way_2020], implying it may be influenced by the same factors. 
+Indeed, while no link between $PLA_{dry}$ or $\pi_{tlp}$ on drought survival has been established [but see @powers_catastrphic_2020], taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019].
 As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].

### 2020-09-22 manuscript/tables&figures.Rmd
Update tables&figures.Rmd

@@
 
-Parentheses indicate that the prediction was supported by at least one but not all of the top  models (Table S8).
-Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S8).
+Parentheses indicate that the prediction was supported by at least one but not all of the top  models (Tables S8, S10, S11).
+Dash symbols indicate that the response was not significant (Tables S4, S6, S7).
 

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
new word count, trying to shorten

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) | 6,490 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,654 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
 Summary|  	200 | 	No. of Tables | 	3  
 Introduction|	1,264 |	No of Supporting Information files | 	19
-Materials and Methods |	2,228	||	 
-Results |	929	 ||
-Discussion |	1927	||
+Materials and Methods |	2,178	||	 
+Results |	1126	 ||
+Discussion |	1952	||
 Acknowledgments | 142	||
@@
 In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015].
-Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
-Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (resistance) and to recover to its pre-drought growth rate (resilience) [@lloret_components_2011].
-Because the resistance and resilience of growth to drought are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
+Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as an important driver of forest change  [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (resistance) and recover to its pre-drought growth rate (resilience) [@lloret_components_2011].
+Because resistance and resilience are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
 However, it has proven difficult to resolve the many factors affecting tree growth during drought and the extent to which their influence is consistent across droughts.
@@
 Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @gillerot_tree_2020].
-Although lower recovery and resilience of larger trees have also been observed [@hacket-pain_consistent_2016; @gillerot_tree_2020], in general we have much more limited understanding of how and why these scale with tree size.
+Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], in general we have much more limited understanding of how and why these scale with tree size.
 
@@
 Other R-packages used in analyses are listed in the Supplementary Information (Appendix S1).
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
 
@@
 The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig.
-**1a**, @helcoski_growing_2019] and had substantial but not devastating impacts on tree growth (Figs.
-**1b**).
-These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019].
-However, extreme, multiannual droughts such as the so-called “megadroughts” of this type that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019] have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016].
-Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig.
-**S1**), severely stressed a larger portion of trees (Fig.
-**1b**).
+**1a**, @helcoski_growing_2019] and had substantial but short-lived impacts on tree growth (Fig.
+**1**).
+These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], but were modest compared to the so-called “megadroughts” that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019; @clark_impacts_2016].
+Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). 
 The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
@@
 **1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
+Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).
@@
 
+### Data and code availability
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
+
+### ORCID
+Kristina J. Anderson-Teixeira: https://orcid.org/0000-0001-8461-9713
+
 ### Supplementary Information

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd

@@
 Many studies have shown that within and across species, large trees tend to be more affected by drought. 
-Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @gillerot_tree_2020].
-Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], in general we have much more limited understanding of how and why these scale with tree size.
-
-*physiological model of @trugman_tree_2018 predicts that recovery and resilience would not necessarily be lower, even in trees that will ultimately die.*
+Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @pretzsch_drought_2018; @gillerot_tree_2020].
+Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], a recent physiological model suggests that large trees  destined to die following drought may still exhibit high recovery and resilience [@trugman_tree_2018].
+Thus, in general we have much more limited understanding of how and why drought resilience scales with tree size.
 
@@
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
-**(here would be a good place to comment on resilience. "What may help the tree in competition, therefore, can also be a deteriment to drought resilience.")**
 
@@
 Their crowns tend to occupy more exposed canopy positions, which are associated with higher evaporative demand [@kunert_revised_2017]. 
-Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. 
 Counteracting the liabilities associated with tall height, large trees tend to have larger root systems [@enquist_global_2002; @hui_near_2014], potentially mitigating some of the biophysical challenges they face by allowing greater access to water.
@@
 Debates have also arisen regarding the traits influencing tree growth responses to drought. 
-Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction does not always hold [@martin-benito_convergence_2015], nor does it hold in the global context [@wheeler_variations_2007; @olson_xylem_2020], and it does not resolve differences among the many species within each category.
+Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019].
+However, this distinction does not always hold within the biome [@martin-benito_convergence_2015] or in the global context [@wheeler_variations_2007; @olson_xylem_2020], and it does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
@@
 More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance [@campbell_introduction_1998; @rey-sanchez_spatial_2016].
-Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
+Under drought, when direct solar radiation tends to be higher (because of less cloud cover) and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism.
 Indeed, previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].

### 2020-09-22 manuscript/response_to_reviews.Rmd
update response to reviews

@@
 
-*I think our argument is pretty strong, but we need to make some change to respond to this comment.
-Looking at resilience and recovery should help; presuming that they are high, we have a further argument that death was not caused by drought, and hence that status at coring doesn't matter.*
-
-**We consider it extremely unlikely that the status at coring would affect drought response.
-We have modified the text at (former) lines 175-178 to make this more clear: "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
-Lagged drought-induced mortality would be unlikely, given that the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare."**
+**We consider it extremely unlikely that the status at coring would affect drought response, particularly given the high resilience observed in the most recent round of analysis, and addressing this would add complexity and length not afforded by the journal's  length requirements.
+We have modified the text at (former) lines 175-178 to make this more clear: "It is unlikely that drought was a factor in the death of any of these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 (near-normal) in these years or the three years prior (2013-2017).
+Moreover, the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare.
+Having found that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019], we pooled the samples for this analysis.."**
 
@@
 
-**In Figure S4, letters express signficiantly different groupings per variable.
+**In Figure S4, letters express significantly different groupings per variable.
 For example, the PLA values for a species in group "b" are significantly different from the PLA values for a species in group "c".

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
word-smithing/ shortening
wording change in #146 (closes #146)

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) | 6,654 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,603 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
 Summary|  	200 | 	No. of Tables | 	3  
@@
 Results |	1126	 ||
-Discussion |	1952	||
+Discussion |	1953	||
 Acknowledgments | 142	||
@@
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as an important driver of forest change  [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
-Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (resistance) and recover to its pre-drought growth rate (resilience) [@lloret_components_2011].
-Because resistance and resilience are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (*resistance*) and recover to its pre-drought growth rate [*resilience*; @lloret_components_2011].
+Survival has been shown to be linked to resistance, recovery, and resilience [@desoto_low_2020; @gessler_way_2020], implying they may be influenced by the same factors.
 However, it has proven difficult to resolve the many factors affecting tree growth during drought and the extent to which their influence is consistent across droughts.
@@
 
-In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
+In contrast, *hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
 For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
+More easily-measurable* leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
 These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].
@@
 
-Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W; Fig.
+Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W, elevation 273-338 m.a.s.l.; Fig.
 **S1**) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
-SCBI is located in the central Appalachian Mountains near the northern boundary of Shenandoah National Park.
-Elevations range from 273 to 338 m above sea level with a topographic relief of 65m [@bourg_initial_2013].
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm yr^-1^ during our study period [1960-2009; source: CRU TS v.4.01; @harris_updated_2014].
@@
 
-We identified the three largest droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with anomalously dry peak growing season climatic conditions. 
-Specifically, we used the metric of Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S1), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site.
+We identified the three largest droughts within the time period 1960-2009, defining drought [@slette_how_2019] based on Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S1), which were identified by @helcoski_growing_2019 as the months to which annual tree growth was most sensitive at this site.
 PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, we identified the three strongest droughts during the study period (Figs. **1, S1**; Table S1).
+Based on this, the three strongest droughts during the study period occurred in 1966, 1977, and 1999 (Figs. **1, S1**; Table S1).
 
@@
 From these census data, we used measurements of DBH from 2008 to calculate historical DBH and data for all stems $\ge$ 10cm to analyze functional trait composition relative to tree height (all analyses described below).
-Census data are available through the ForestGEO data portal (www.forestgeo.si.edu).
 
@@
 In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. 
-Annual tree mortality censuses were initiated in 2014 [@gonzalezakre_patterns_2016], and in 2016-2017, cores were collected from all trees found to have died since the previous year’s census. 
-We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
-Lagged drought-induced mortality would be unlikely, given that the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare.
+In summers of 2016 and 2017, cores were collected from all trees found to have died within the preceding year based on annual tree mortality censuses [@gonzalezakre_patterns_2016]. 
+It is unlikely that drought was a factor in the death of any of these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 (near-normal) in these years or the three years prior (2013-2017). 
+Moreover, the trees analyzed here lived at least 17-18 years past the most recent major drought (1999), whereas the meta-analysis of @trugman_tree_2018 indicates that >10-year lags in drought-attributed mortality are rare.
+Having found that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019], we pooled the samples for this analysis. 
 Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
@@
 All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
+Full ForestGEO census data for SCBI are available through the ForestGEO data portal (www.forestgeo.si.edu).
 

### 2020-09-22 manuscript/response_to_reviews.Rmd
moved content from results to fig caption

@@
 As now stated in the methods, "Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
-While we tested the significance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below)."
+While we tested the significance of xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (see Results)."
 
@@
 Keeping this figure would have required adding parallel figures for recovery and resilience, which we felt would be adding too many marginally valuable appendix figures.
+
+* To present the new results on recovery and resilience while keeping within the journal's word limits, we condensed text / eliminated unnecessary content throughout.
+
 ...

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
moved content from results to fig caption

@@
 Both methods are associated with some error [@larjavaara_measuring_2013], but in this instance there was no clear advantage of one or the other.
-Measurements from the National Ecological Observatory Network (NEON) were collected near the ForestGEO plot following standard NEON protocol, whereby vegetation of short stature was measured with a collapsible measurement rod, and taller trees with a rangefinder [@neon_national_2018].
 Species-specific height allometries were developed using log-log regression ($ln[H] \sim ln[DBH]$; Table S3). 
@@
 
-For each drought year, we calculated metrics of drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs$), following @lloret_components_2011 (Table 3).
-We defined $Rt$ as the ratio of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) during the drought year to the mean $BAI$ over the five years preceding the drought. 
-$Rc$ was defined as the ratio of mean $BAI$ over the five years following the drought to $BAI$ of the drought year.
-$Rs$ was defined as the ratio of mean $BAI$ over the five years following the drought to that of the five years preceding the drought. 
-Thus, for all metrics, values <1 and >1 indicate growth reductions and increases, respectively.
+For each drought year, we calculated metrics of drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs$), following @lloret_components_2011. 
+These metrics compare ratios of basal area increment ($BAI$; *i.e.*, change in cross-sectional area) before, during, and after the drought year, as specified in Table 3.  
+For all metrics, values <1 and >1 indicate growth reductions and increases, respectively.
 
@@
 Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
-While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (detailed below).
+While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (see Results).
 
@@
 For $Rt$, differences among drought years were small (<0.02; Table S8). 
-In contrast, differences among years were larger for $Rc$ and $Rs$, with coefficients for "drought year" highest in 1966, intermediate in 1977, and lowest in 1999.
+In contrast, differences among years were larger for $Rc$ and $Rs$, with coefficients for year highest in 1966, intermediate in 1977, and lowest in 1999.
 
@@
 
-Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (i.e., lower $Rt$) and less rebound following drought (i.e., lower $Rc$ and $Rs$; Table 1; Fig.
+Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (*i.e.*, lower $Rt$) and less rebound following drought (*i.e.*, lower $Rc$ and $Rs$; Table 1; Fig.
 **4**).
@@
 
-Trees in drier microsites showed greater growth declines during drought; i.e., $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig.
+Trees in drier microsites showed greater growth declines during drought; *i.e.*, $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig.
 **4**, Table S8-S9).
@@
 Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
-This implies that small trees had greater recovery and resilience in wetter microhabitats, whereas large trees had greater recovery and resilience in dry microhabitats.
+This implies a non-significant tendency for small trees to have greater recovery and resilience in wetter microhabitats, but for large trees to have greater recovery and resilience in dry microhabitats.
 
@@
 Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
-**S4** as characterized by the agricolae package @R-agricolae).
+**S4**).
 Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
finished reviewing/ condensing pass

@@
 Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**).
-The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
-Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). 
+Taller trees had greater exposure to conditions that would promote water loss and heat damage during drought (Fig.
+**2**), which is one plausible mechanism for their lower drought resistance, recovery, and resilience (Fig.
+**4**).
 There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019].
-Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits; Fig.
-**4**) had increased growth (Fig.
-**1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
+Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits) had increased growth (Figs.
+**1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
 Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
@@
 Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests.
-The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests.
-Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@christoffersen_linking_2016; @powell_differences_2017].
+The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests where most species do not form annual rings [but see @schongart_dendroecological_2017 for a review of progress in tropical dendroecology].
+A full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@christoffersen_linking_2016; @powell_differences_2017].
 

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
reviewed/ integrated refs from Alan
#135

@@
 Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @pretzsch_drought_2018; @gillerot_tree_2020].
-Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], a recent physiological model suggests that large trees  destined to die following drought may still exhibit high recovery and resilience [@trugman_tree_2018].
+Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], results are mixed [@merlin_effects_2015], and a recent physiological model suggests that large trees destined to die following drought may still exhibit high recovery and resilience [@trugman_tree_2018].
 Thus, in general we have much more limited understanding of how and why drought resilience scales with tree size.
@@
 Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought tolerance (Fig. **4**; see discussion below). 
+*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig.
+**S4**), which would have contributed to its drought tolerance (Fig.
+**4**; see discussion below), in concordance with studies identifying *Fagus* species as intermediate in drought tolerance [@vitasse_contrasting_2019; @pretzsch_drought_2018].
 A second explanation of why *F.

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
word count and word version for coauthors

@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) | 6,603 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,458 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
 Summary|  	200 | 	No. of Tables | 	3  
-Introduction|	1,264 |	No of Supporting Information files | 	19
-Materials and Methods |	2,178	||	 
-Results |	1126	 ||
-Discussion |	1953	||
+Introduction|	1,340 |	No of Supporting Information files | 	19
+Materials and Methods |	2,036	||	 
+Results |	1124	 ||
+Discussion |	1952	||
 Acknowledgments | 142	||

### 2020-09-22 manuscript/response_to_reviews.Rmd
update response to reviews

@@
 * To present the new results on recovery and resilience while keeping within the journal's word limits, we condensed text / eliminated unnecessary content throughout.
-
-...
 
@@
 **Thank you.
-Because we had to cut content in order to present the recovery and resilience metrics (see below) while keeping within word limits, we have cut back on the reporting of $Rt_{ARIMA} results in the main text, but still show the full comparisons in the Appendix.
+Because we had to cut content in order to present the recovery and resilience metrics (see below) while keeping within word limits, we have cut back on the reporting of $Rt_{ARIMA}$ results in the main text, but still show the full comparisons in the Appendix.
 Specifically, we now state in the following in the methods: "$Rt$ and $Rt_{ARIMA}$ were strongly correlated (Fig.
@@
 **We now report both marginal and conditional R2 in tables S8-S12**.
-
-*Maybe also add some commentary on how a lot of noise doesn't negate significant signals.*
 
@@
 **Yes, although we do not include canopy position in our statistical models.
-To make the link between high $Rt$ and increased growth, we have modified the sentence previously on line 385 to read, "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits; Fig.
-**4**) had increased growth (Fig.
-**1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]."**
-
-*(for reference, to be deleted:)*
-
-*385: Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b),*
-
-*208-309: However, some individuals exhibited increased growth, i.e., Rt > 1.0: 26% of trees in 1966, 22% in 1977, and 26% in 1999.*
-
-*338: Averaged across all droughts, Rt was lowest in Liriodendron tulipifera (mean Rt = 0.66) and highest in Fagus grandifolia (mean Rt = 0.99).*
+To make the link between high $Rt$ and increased growth, we have modified the sentence previously on line 385 to read, "Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits*) had increased growth (Figs.
+**1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]."**
 

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
Ryan's fixes
we're missing a few citations. I created Exported Items.bib with the refs, but for some reason its breaking the system when I add them to book.bib. That needs to be figured out.

@@
 
-In contrast, *hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
+In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
 For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
-More easily-measurable* leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
+More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
 These include leaf area shrinkage upon desiccation [$PLA_{dry}$; @scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016; @zhu_leaf_2018].

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
Updated to make sure citations worked #147

@@
   pdf_document:
-    citation_package: natbib
     fig_caption: yes
@@
 Research was conducted at the 25.6-ha ForestGEO (Forest Global Earth Observatory) study plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA (38\°53'36.6"N, 78\°08'43.4"W, elevation 273-338 m.a.s.l.; Fig.
-**S1**) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo:_2015].
+**S1**) [@bourg_initial_2013; @andersonteixeira_ctfs-forestgeo_2015].
 Climate is humid temperate, with mean annual temperature of 12.7^$\circ$^C and precipitation of 1005 mm yr^-1^ during our study period [1960-2009; source: CRU TS v.4.01; @harris_updated_2014].
@@
 Survival is linked to resistance and resilience [@desoto_low_2020; @gessler_way_2020], implying it may be influenced by the same factors. 
-Indeed, while no link between $PLA_{dry}$ or $\pi_{tlp}$ on drought survival has been established [but see @powers_catastrphic_2020], taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019].
+Indeed, while no link between $PLA_{dry}$ or $\pi_{tlp}$ on drought survival has been established [but see @powers_catastrophic_2020], taller trees have lower survival [@bennett_larger_2015; @stovall_tree_2019].
 As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020].

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
Updated word_doc for #147 to include references

@@
 author: ""
-output:
-  pdf_document:
-    fig_caption: yes
+output: word_document
+  # pdf_document:
+    # fig_caption: yes
 bibliography: ["book.bib",  "packages.bib"]

### 2020-09-25 scripts/graphs_for_manuscript.R
Updated figure S4 with correct axis units

@@ -289,7 +289,7 @@ library(agricolae)
 rt <- fread("manuscript/tables_figures/trees_all_sub_resistance.csv")
 
 ##this is the INDIVIDUAL LEVEL DATA, not species
-traits_hydr <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_indvidual_level.csv?token=AJNRBEIWYZQPW2RYIPESJDS7M5SLQ")
+traits_hydr <- fread("https://raw.githubusercontent.com/EcoClimLab/HydraulicTraits/master/data/SCBI/processed_trait_data/SCBI_all_traits_table_indvidual_level.csv?token=AJNRBEIC4363KJGA6P7S2727OZ2CS")
 traits_hydr <- traits_hydr[,.(sp,PLA_dry_percent, mean_TLP_Mpa, 
                               WD_g_per_cm3, LMA_g_per_m2)]
 
@@ -387,7 +387,9 @@ posplot <- posplot +
    
 
 hydr <- c("PLA_dry_percent", "mean_TLP_Mpa", "LMA_g_per_m2", "WD_g_per_cm3")
-ylabs <- c(expression(PLA[dry]), expression(pi[TLP]), "LMA [g/m^2]", "WD [g/cm^3]")
+ylabs <- c(expression(PLA[dry]~"[%]"), 
+           expression(pi[TLP]~"[Mpa]"), 
+           "LMA [g/m^2]", "WD [g/cm^3]")
 ggtraits <- list()
 for(i in 1:4){
    varnm <- hydr[[i]]

### 2020-09-25 manuscript/climate_sensitivity_SI.Rmd
Updated for New Phyt guidelines #148

@@
 ![**Figure S2.
-Map of ForestGEO plot showing topographic wetness index and location of cored trees.** Scale units are in meters](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}
+Map of ForestGEO plot showing topographic wetness index (color scale) and location of cored trees.** Scale units are in meters](tables_figures/publication/figureS2_location_cored_trees.png){width=500px}
 

### 2020-09-25 manuscript/climate_sensitivity_draft.Rmd
Updated for New Phyt guidelines #148

@@
 Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area, LMA), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
 Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
@@
 Species-specific height allometries were developed using log-log regression ($ln[H] \sim ln[DBH]$; Table S3). 
-For species with insufficient height data to create reliable species-specific allometries (n=2, JUNI and FRAM), heights were calculated from an equation developed by combining the height measurements across all species.
+For species with insufficient height data to create reliable species-specific allometries (n=2, *Juglans nigra* (JUNI) and *Fraxinus americana* (FRAM)), heights were calculated from an equation developed by combining the height measurements across all species.
 We then used these allometries to estimate $H$ for each drought year, $Y$, based on reconstructed $DBH_Y$ (Fig. **S3**).
@@
 The $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top $Rt$ models in which it appeared (Tables 1, S8-S9), rejecting the hypothesis that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in microenvironments with a deeper water table.
-In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. 4, Tables S10-S11).
+In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. **4**, Tables S10-S11).
 Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
@@
 Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 
-Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *Liriodendron tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
+Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *L. tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
 These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **3**).
@@
 Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
-In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. 3).
+In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. **3**).
 
@@
 **1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
-Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
+Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. **1**), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
 It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).
@@
 Taller trees also face dramatically different microenvironments (Fig. **2**). 
-They are exposed to higher wind speeds and lower humidity (Fig. **2a-b**), resulting in higher evaporative demand. 
+They are exposed to higher wind speeds and lower humidity (Fig. **2a,b**), resulting in higher evaporative demand. 
 Unlike other temperate forests where modestly cooler understory conditions have been documented [@zellweger_seasonal_2019], particularly under drier conditions [@davis_microclimatic_2019], we observed no significant variation in air temperatures across the vertical profile (Fig.
@@
 Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought tolerance (Table 1).
-Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
+Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *L.* and *F.
+grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
 **3**).
@@
 Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
-*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig.
+*F.
+grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig.
 **S4**), which would have contributed to its drought tolerance (Fig.

### 2020-09-25 manuscript/tables&figures.Rmd
moved content from results to fig caption
Updated for New Phyt guidelines #148

@@
 kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2.
-Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.
-Variable abbreviations are as in Table 3.
-DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).", escape=FALSE) %>%
+Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
@@
 ```
+Variable abbreviations are as in Table 3. DBH measurements are from the most recent ForestGEO census in 2018 (live trees) or tree mortality censuses in 2016 and 2017 (trees cored dead).
 
@@
 **Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance, $Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.
@@
 S7 for parallel plot for recovery ($Rc$).
+Analysis conducted using *agricolae* package [@R-agricolae].
 
@@
 ![**Figure 1.
-Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance,$Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
+Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide growth resistance, $Rt$ (b), and resilience, $Rs$ (c).** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes).
 PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al.

### 2020-09-25 manuscript/climate_sensitivity_draft.Rmd
minor change from Norm
update author contributions

@@
 KAT, IM, and AJT designed the research.
-Tree-ring chronologies were developed by RH under guidance of AJT and NP.
-Trait data were collected by IM, JZ under guidance of NK and LS.
-Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM.
+Tree-ring chronologies were developed by RH under guidance of AJT, KAT, and NP.
+Trait data were collected by IM and JZ under guidance of NK, KAT, and LS.
+Other plot data were collected by NB in coordination with WM and by IM and AS under guidance of EGA and KAT.
 Data analyses were performed by IM under guidance of KAT and VH.

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
Updated with AT comments #149

@@
 3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
-4. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+4. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada T6H 3S5
 5. Biological Sciences Department; California State University; Los Angeles, CA 90032, USA
@@
 However, this distinction does not always hold within the biome [@martin-benito_convergence_2015] or in the global context [@wheeler_variations_2007; @olson_xylem_2020], and it does not resolve differences among the many species within each category.
-Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
+Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
 However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent.
@@
 In contrast, hydraulic traits have direct physiological linkages to tree growth and mortality responses to drought. 
-For instance, water potentials at which percent the loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
+For instance, water potentials at which the percent loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
 However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
@@
 
-Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape three metrics of drought tolerance: 
-(1) resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought based on previous years' growth;
-(2) recovery, defined the ratio of post-drought growth to growth during the drought year;
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape three metrics of drought tolerance as compared with five years' pre-drought growth: 
+(1) resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought from previous growth;
+(2) recovery, defined as the ratio of post-drought growth to growth during the drought year;
 and (3) resilience, defined as the ratio of post-drought to pre-drought growth [@lloret_components_2011].
@@
 $PLA_{dry}$ was calculated as the percent loss of area between fresh and dry leaves.
-Wood density was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement.
+$WD$ was calculated for ~1cm diameter stem samples (bark and pith removed) as the ratio of dry weight to fresh volume, which was estimated using Archimedes’ displacement.
 We used the rapid determination method of @bartlett_rapid_2012 to estimate osmotic potential at turgor loss point ($\pi_{tlp}$).
@@
 Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
-While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2) and opposite drought responses of the only two diffuse-porous species (see Results).
+While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2).
+In addition, we observed opposite drought responses of the only two diffuse-porous species (see Results), themselves likely representing the most and least shade-tolerant species in the study area.
 
@@
 However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999. 
-Recovery was generally strong and complete, with $Rc$ averaging 1.55 in 1966, 1.42 in 1977, and 1.34 in 1999 (Fig.
+Recovery was generally strong and complete within five years following each of the drought years, with $Rc$ averaging 1.55 in 1966, 1.42 in 1977, and 1.34 in 1999 (Fig.
 **S6**) and $Rs$ averaging 1.28 in 1966, 1.19 in 1977, and 1.12 in 1999 (Fig.
@@
 
-Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits (wood density, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for wood density and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
+Species, as a factor in ANOVA, had significant (p<0.05) influence on all traits ($WD$, $LMA$, $PLA_{dry}$, and $\pi_{tlp}$), with more significant pairwise differences for $WD$ and $PLA_{dry}$ than for $LMA$ and $\pi_{tlp}$ (Table 2, Fig.
 **S4**).
@@
 
-Wood density, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1, S4-S5). 
-Wood density and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 
+$WD$, $LMA$, and xylem porosity were all poor predictors of drought tolerance (Tables 1, S4-S5). 
+$WD$ and $LMA$ were never significantly associated with $Rt$, $Rc$, or $Rs$ in the single-variable tests and were therefore excluded from the full models. 
 Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
@@
 The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020].
-Drought tolerance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs.
+Drought tolerance was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs.
 diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$).
@@
 **3**) and associated traits at a single site [see also @elliott_forest_2015].
-Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought tolerance (Table 1).
+Our study reinforced current understanding (see Introduction) that $WD$ and $LMA$ are not reliably linked to drought tolerance (Table 1).
 Contrary to several previous studies in temperate deciduous forests [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], we did not find an association between xylem porosity and drought resistance or resilience, as the two diffuse-porous species, *L.* and *F.

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
Updated SI table and figure names

@@
 ### Supplementary Information
-
-**NEED TO UPDATE !!****
-Table S1. Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.
+Table S1. Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 for focal droughts.
 
@@
 
-Table S6. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.
+Table S6. Individual tests of species traits as drivers of drought recovery ($Rc$).
 
-Table S7. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
+Table S7. Individual tests of species traits as drivers of drought resilience ($Rs$).
 
-Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the 2.5 years prior to each focal drought
+Table S8. Summary of top full models for each drought instance, where $Rt$ is used as the response variable.
 
-Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees
+Table S9. Summary of top  models for each drought instance, where $Rt_{ARIMA}$ is used as the response variable.
+
+Table S10. Summary of top models for each drought instance, where $Rc$ is used as the response variable.
+
+Table S11. Summary of top models for each drought instance, where $Rs$ is used as the response variable.
+
+Figure S1. Time series of Palmer Drought Severity Index (PDSI) for the 2 years prior and after each focal drought.
+
+Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees.
 
@@
 
-Figure S6. Visualization of best model, with data, for all droughts combined.
+Figure S6. Density plot of Recovery ($Rc$) values for each focal year.
+
+Figure S7. Drought recovery, $Rc$, across species for the three focal droughts.
 

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
Addressed LS comments #149

@@
 
-- We analyzed tree-ring records for twelve species in a broadleaf deciduous forest in Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period.
+- We analyzed tree-ring records for twelve species in a broadleaf deciduous forest in Virginia (USA) to test hypotheses for how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period.
 
-- Drought tolerance (resistance, recovery, and resilience) decreased with tree height, which was strongly correlated with exposure to higher evaporative demand and solar radiation.
+- Drought tolerance (resistance, recovery, and resilience) decreased with tree height, which was strongly correlated with exposure to higher solar radiation and evaporative demand.
 The potentially greater rooting volume of larger trees did not confer a resistance advantage, but marginally increased recovery and resilience, in sites with low topographic wetness index.
-Drought tolerance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.
+Drought tolerance was greater among species whose leaves lost turgor (wilted) at more negative water potentials and experienced less shrinkage upon desiccation.
 
@@
 Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as an important driver of forest change  [@allen_global_2010; @allen_underestimation_2015; @mcdowell_pervasive_2020].
-Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (*resistance*) and recover to its pre-drought growth rate [*resilience*; @lloret_components_2011].
+Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought tolerance, defined here as a tree's ability to maintain growth during drought (*resistance*), increase growth relative to drought minimum (*recovery*), and re-establish its pre-drought growth rate [*resilience*; @lloret_components_2011].
 Survival has been shown to be linked to resistance, recovery, and resilience [@desoto_low_2020; @gessler_way_2020], implying they may be influenced by the same factors.
@@
 Greater growth reductions (*i.e.*, lower drought resistance) in larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @pretzsch_drought_2018; @gillerot_tree_2020].
-Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], results are mixed [@merlin_effects_2015], and a recent physiological model suggests that large trees destined to die following drought may still exhibit high recovery and resilience [@trugman_tree_2018].
+Although lower recovery and resilience of larger trees have also been observed [@gillerot_tree_2020], results were mixed [@merlin_effects_2015], and a recent physiological model suggests that large trees destined to die following drought may still exhibit high recovery and resilience [@trugman_tree_2018].
 Thus, in general we have much more limited understanding of how and why drought resilience scales with tree size.
@@
 Taller trees face the biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018].
-Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
+Vertical gradients in stem and leaf traits--including smaller and thicker leaves (higher leaf mass per area), greater resistance to hydraulic dysfunction (*i.e.*, more negative water potential at 50% loss of hydraulic conductivity, more negative P50), and the tapering of hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]--enable trees to become tall [@couvreur_water_2018].
 Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
@@
 Finally, tree size-related responses to drought can be modified by species' traits and their distribution across size classes [@meakem_role_2018; @liu_hydraulic_2019]. 
-Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought requires sorting out the interactive effects of height and associated exposure, root water access, and species' traits.
+Understanding the mechanisms driving the greater relative growth reductions of larger trees during drought requires disentangling the interactive effects of height and associated exposure, root water access, and species' traits.
 
@@
 For instance, water potentials at which the percent loss of conductivity surpasses a certain threshold (e.g., P50 and P88, representing 50 and 88% loss of conductivity, respectively) and hydraulic safety margin (*i.e.*, difference between typical minimum water potentials and P50 or P88) correlate with drought performance across global forests [@anderegg_meta-analysis_2016].
-However, these are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
+However, these are time-consuming to measure and therefore often infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics).
 More easily-measurable leaf drought tolerance traits that have direct linkage to plant hydraulic function can explain variation in plant distribution and function [@medeiros_extensive_2019]. 
@@
 We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought tolerance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of [@bennett_larger_2015].
-We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
+We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in consistently wet microsites. 
 Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought tolerance. 
@@
 We identified the three largest droughts within the time period 1960-2009, defining drought [@slette_how_2019] based on Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S1), which were identified by @helcoski_growing_2019 as the months to which annual tree growth was most sensitive at this site.
-PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. 
-Based on this, the three strongest droughts during the study period occurred in 1966, 1977, and 1999 (Figs. **1, S1**; Table S1).
+PDSI divisional data for Northern Virginia were obtained in December 2017 from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp), from which we determined the three strongest droughts during the study period occurred in 1966, 1977, and 1999 (Figs.
+**1, S1**; Table S1).
 
@@
 To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model.
-These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits.
+These tests followed the model structure specified above, with $ln[H]$ and $ln[TWI]$ added to create a base model against which we tested traits.
 Trait variables were considered appropriate for inclusion in the main model if their addition to the base model significantly improved fit for at least one metric of drought tolerance ($Rt$, $Rc$, or $Rs$; Tables S4,S6-S7).
-While we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of highly uneven distribution of species across categories (Table 2).
+Although we tested xylem porosity as a predictor (Table 1), we did not consider it appropriate for inclusion in the main model because of its highly uneven distribution of species across categories (Table 2).
 In addition, we observed opposite drought responses of the only two diffuse-porous species (see Results), themselves likely representing the most and least shade-tolerant species in the study area.
@@
 Consistent with studies in other forests worldwide [@bennett_larger_2015], taller trees in this forest exhibited lower drought resistance--and also recovery and resilience--when compared to smaller trees.
-Mechanistically, this is consistent with, and reinforces, previous findings that it is impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
-Taller trees also face dramatically different microenvironments (Fig. **2**). 
+Mechanistically, this is consistent with, and reinforces, previous findings for a trade-off between the ability of trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@liu_hydraulic_2019; @olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019].
+Taller trees also face dramatically distinctive microenvironments (Fig. **2**). 
 They are exposed to higher wind speeds and lower humidity (Fig. **2a,b**), resulting in higher evaporative demand. 
@@
 In any case, the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
-This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013], and these are then more vulnerable when drought hits.
+This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013], and these are then more vulnerable when drought occurs.
 The observed height-sensitivity of $Rt$, together with the lack of conferred advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests.
@@
 
-The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig.
+The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] provided a sufficient sample size to compare historical drought responses across species (Fig.
 **3**) and associated traits at a single site [see also @elliott_forest_2015].
@@
 
-Concerted measurement of tree-rings and leaf drought tolerance traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of drought tolerance traits in shaping drought response.
+Concerted measurement of tree-rings and leaf drought tolerance traits of emerging importance in published literature [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of drought tolerance traits in shaping drought response.
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig.
@@
 
+Ian R. McGregor: https://orcid.org/0000-0002-5763-021X
+
 ### Supplementary Information

### 2020-09-26 manuscript/tables&figures.Rmd
Added authorities and updated table 2 #148

@@
 Overview of analyzed species, listed in order of their relative contributions to woody stem productivity ($ANPP_{stem}$) in the plot, along with numbers and sizes sampled, and species traits.", escape=FALSE) %>%
+  column_spec(1, italic=TRUE) %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
reviewed changes from AT, made one small change
#149

@@
 
-Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape three metrics of drought tolerance as compared with five years' pre-drought growth: 
+Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape three metrics of drought tolerance: 
 (1) resistance, defined as the ratio of annual stem growth in a drought year to that which would be expected in the absence of drought from previous growth;
@@
 ### ORCID
+Ian R. McGregor: https://orcid.org/0000-0002-5763-021X
+
 Kristina J. Anderson-Teixeira: https://orcid.org/0000-0001-8461-9713
-
-Ian R. McGregor: https://orcid.org/0000-0002-5763-021X
 

### 2020-09-26 manuscript/tables&figures.Rmd

@@
 table2[8,8] <- paste0(table2[8,8], footnote_marker_symbol(2))
+table2$species <- paste0(table2$species, " ", table2$authority)
+table2$authority <- NULL
 kable(table2, format = "latex", booktabs=TRUE, caption = "Table 2.

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
final read/ minor changes

@@
 Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019].
-However, this distinction does not always hold within the biome [@martin-benito_convergence_2015] or in the global context [@wheeler_variations_2007; @olson_xylem_2020], and it does not resolve differences among the many species within each category.
+However, this differentiation is not universal within the biome [@martin-benito_convergence_2015], does not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020], and does not resolve differences among the many species within each category.
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
@@
 Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006].
-TWI calculation depends on an input of a digital elevation model (DEM; ~3.7 m resolution from the elevatr package [@R-elevatr]), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
+TWI calculation depends on an input of a digital elevation model [DEM; ~3.7 m resolution from the elevatr package [@R-elevatr], and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely.
 From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.
@@
 Appendix S1. Further Package Citations
+
+\newpage
+### References

### 2020-09-26 manuscript/response_to_reviews.Rmd
finalized response to reviews

@@
 
-**We consider it extremely unlikely that the status at coring would affect drought response, particularly given the high resilience observed in the most recent round of analysis, and addressing this would add complexity and length not afforded by the journal's  length requirements.
+**We consider it extremely unlikely that the status at coring would affect drought response, particularly given the high resilience observed in the most recent round of analysis, and addressing this in the manuscript would add complexity and length not afforded by the journal's length requirements.
 We have modified the text at (former) lines 175-178 to make this more clear: "It is unlikely that drought was a factor in the death of any of these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 (near-normal) in these years or the three years prior (2013-2017).

### 2020-09-26 manuscript/climate_sensitivity_SI.Rmd
Updated SI table and figure names
minor modification of appendix s1
rename methods s1, cut some text

@@
 
-**Appendix S1. Further Package Citations**
+**Methods S1. Further Package Citations**
 
@@
 
-![**Figure S6. Drought recovery, $Rc$, across species for the three focal droughts.**](tables_figures/publication/FigureS6_density_plot_recovery.png)
+![**Figure S6. Density plot of Recovery ($Rc$) values for each focal year.**](tables_figures/publication/FigureS6_density_plot_recovery.png)
 
@@
 \newpage
-## Appendix S1. Further Package Citations
+## Methods S1. Further Package Citations
 While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes.
-As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others.
-While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.
-
 R-packages not already cited in the main manuscript include the following, listed alphabetically by corresponding package name:
@@
 R base [@R-base]; broom [@R-broom]; car [@R-car]; cowplot [@R-cowplot]; data.table [@R-data.table]; devtools [@R-devtools]; dplR [@R-dplR]; dplyr [@R-dplyr]; extrafont [@R-extrafont]; ggplot2 [@R-ggplot2]; ggpubr [@R-ggpubr]; ggthemes [@R-ggthemes]; gridExtra [@R-gridExtra]; knitr [@R-knitr]; lubridate [@R-lubridate]; MuMIn [@R-MuMIn]; piecewiseSEM [@R-piecewiseSEM]; png [@R-png]; purrr [@R-purrr]; raster [@R-raster]; rasterVis [@R-rasterVis]; RCurl [@R-RCurl]; readxl [@R-readxl]; reshape2 [@R-reshape2]; rgdal [@R-rgdal]; rgeos [@R-rgeos]; rmarkdown [@R-rmarkdown]; sf [@R-sf]; stringi [@R-stringi]; stringr [@R-stringr]; tidyr [@R-tidyr]
+
+### References

### 2020-09-26 manuscript/climate_sensitivity_draft.Rmd
minor modification of appendix s1
rename methods s1, cut some text
word count, pdf for submission
#148

@@
 author: ""
-output: word_document
-  # pdf_document:
+output: 
+  #  word_document
+   pdf_document:
     # fig_caption: yes
@@
 ---------------------------------------------|---------------------|-----------------------------|--------------------
-Total word count (excluding summary, references and legends) | 6,458 **(Limit 6500)**	 |	No. of figures | 	4 (all colour)
+Total word count (excluding summary, references and legends) | 6,494	 |	No. of figures | 	4 (all colour)
 Summary|  	200 | 	No. of Tables | 	3  
-Introduction|	1,340 |	No of Supporting Information files | 	19
-Materials and Methods |	2,036	||	 
-Results |	1124	 ||
-Discussion |	1952	||
+Introduction|	1,347 |	No of Supporting Information files | 	19
+Materials and Methods |	2,057	||	 
+Results |	1129	 ||
+Discussion |	1955	||
 Acknowledgments | 142	||
@@
 All analysis beyond basic data collection was performed using R version 3.6.2 [@R-base]. 
-Other R-packages used in analyses are listed in the Supplementary Information (Appendix S1).
+Other R-packages used in analyses are listed in the Supplementary Information (Methods S1).
 
@@
 
-Appendix S1. Further Package Citations
+Methods S1. Further Package Citations
 

### 2020-09-27 manuscript/climate_sensitivity_SI.Rmd
Updated SI Rmd but did not knit

@@
 
-![**Figure S6. Density plot of Recovery ($Rc$) values for each focal year.**](tables_figures/publication/FigureS6_density_plot_recovery.png)
+![**Figure S6. Density plot of drought recovery ($Rc$) values for each focal drought year.**](tables_figures/publication/FigureS6_density_plot_recovery.png)
 
@@
 
-![**Figure S7. Drought recovery, $Rc$, across species for the three focal droughts.**](tables_figures/publication/FigureS7_Rc_across_sp.png)
+![**Figure S7.
+Drought recovery ($Rc$) across species for the three focal droughts.
+Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
+Letter groupings do not transfer between variables.**](tables_figures/publication/FigureS7_Rc_across_sp.png)
 

### 2020-10-05 manuscript/climate_sensitivity_draft.Rmd
knit a word copy
setup for production edits

@@
 output: 
-  #  word_document
-   pdf_document:
+    word_document
+    #pdf_document:
     # fig_caption: yes
@@
 
+Norbert Kunert: https://orcid.org/0000-0002-5602-6221
+
 Kristina J. Anderson-Teixeira: https://orcid.org/0000-0001-8461-9713

### 2020-10-06 manuscript/tables&figures.Rmd
Swapped Figs 3-4, added text to legends #150

@@
 **Figure 3.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
+Visualization of top statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** For cases where the best model includes a DBH x TWI interaction ($Rc$ in all droughts and 1966, $Rs$ in all droughts and 1977), we plot the best model without the interaction.
+Visualization of the best mixed effects model per drought scenario was created by the *visreg* package in R, and confidence intervals were defined via bootsrapping in the *bootpredictlme4* package.
+Model coefficients are given in Tables S8 and S10-11.
+Descriptions of variables (e.g.
+$ln[H]$ can be found in Table 3.
+
+**Figure 4.
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) value ranges across species for the three focal droughts.** Species codes are given in Table 2, and closed dots represent outliers from the main boxplots.
+The dotted line at y=1 represents no change in $Rt$ or $Rs$ from the previous year.
+Letters illustrate significance groupings per year (colored and ordered, top to bottom, 1966, 1977, 1999).
+That is, a group of species with the same letter above their boxplot (e.g.
+"b") are statistically different from species in another group (e.g.
+"a").
+Letter groupings do not transfer across variables $Rt$ and $Rs$.
 See Fig.
 S7 for parallel plot for recovery ($Rc$).
-Analysis conducted using *agricolae* package [@R-agricolae].
-
-**Figure 4.
-Visualization of best statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** Confidence intervals were defined via bootsrapping in the  bootpredictlme4 package.
-Model coefficients are given in Tables S8 and S10-11.
+Analysis conducted using *agricolae* package in R.
+Descriptions of variables (e.g.
+$ln[H]$ can be found in Table 3.
 
@@
 ![**Figure 3.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b), across species for the three focal droughts.** Species codes are given in Table 2.
-See Fig.
-S7 for parallel plot for recovery ($Rc$).](tables_figures/publication/Figure3_Rt-Rs_across_sp.png)
+Visualization of top statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** For cases where the best model includes a DBH x TWI interaction ($Rc$ in all droughts and 1966, $Rs$ in all droughts and 1977), we plot the best model without the interaction.
+Visualization of the best mixed effects model per drought scenario was created by the *visreg* package in R, and confidence intervals were defined via bootsrapping in the *bootpredictlme4* package.
+Model coefficients are given in Tables S8 and S10-11.
+Descriptions of variables (e.g.
+$ln[H]$ can be found in Table 3.)
+](tables_figures/publication/Figure3_model_vis_all.png)
 
@@
 ![**Figure 4.
-Visualization of top statistical models for drought resistance ($Rt$), recovery ($Rc$), and resilience ($Rs)$ for all droughts combined and for each individual drought year.** For cases where the best model includes a DBH x TWI interaction ($Rc$ in all droughts and 1966, $Rs$ in all droughts and 1977), we plot the best model without the interaction.
-Visualization of the best mixed effects model per drought scenario was created by the visreg package [@R-visreg], and confidence intervals were defined via bootsrapping in the bootpredictlme4 package.
-Model coefficients are given in Tables S8 and S10-11.
-](tables_figures/publication/Figure4_model_vis_all.png)
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) value ranges across species for the three focal droughts.** Species codes are given in Table 2, and closed dots represent outliers from the main boxplots.
+The dotted line at y=1 represents no change in $Rt$ or $Rs$ from the previous year.
+Letters illustrate significance groupings per year (colored and ordered, top to bottom, 1966, 1977, 1999).
+That is, a group of species with the same letter above their boxplot (e.g.
+"b") are statistically different from species in another group (e.g.
+"a").
+Letter groupings do not transfer across variables $Rt$ and $Rs$.
+See Fig.
+S7 for parallel plot for recovery ($Rc$).
+Analysis conducted using *agricolae* package in R.](tables_figures/publication/Figure4_Rt-Rs_across_sp.png)

### 2020-10-06 manuscript/climate_sensitivity_SI.Rmd
Added edits from #150

@@
 output:
-  pdf_document:
-    fig_caption: yes       
-    includes:  
-      in_header: preamble-latex.tex
+  word_document:
+  #pdf_document:
+   # fig_caption: yes       
+  #  includes:  
+   #   in_header: preamble-latex.tex
 bibliography: ["packages.bib"]
 biblio-style: apalike
-csl: new-phytologist.csl
 header-includes:

### 2020-10-06 manuscript/climate_sensitivity_draft.Rmd
added some orcid/ social media
Update climate_sensitivity_draft.Rmd
Added edits from #150

@@
 link-citations: yes
-biblio-style: "apalike"
-csl: new-phytologist.csl
 header-includes:
@@
 ), file = 'packages.bib')
+
+knitr::write_bib(x = c("knitr", "rmarkdown") , file = ".bib")
 ```
@@
 *corresponding author: teixeirak@si.edu; +1 540 635 6546
+
+**Social Media:**
+
+Joseph Zailaa: 
+https://twitter.com/ZailaaJ
+
+Neil Pederson:
+https://twitter.com/YellowBuckeye,
+https://www.facebook.com/neil.pederson.5,
+https://www.instagram.com/1yellowbuckeye/
+
+Kristina Anderson-Teixiera:
+https://twitter.com/K_A_Teixeira
+
+Ian R. McGregor:
+https://twitter.com/realclimatian
 
@@
 Commonly-measured traits including wood density ($WD$) and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017].
-However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent.
+However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2020], or the direction of response was not always consistent.
 For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011].
@@
 Taller trees (based on $H$ in the drought year) showed stronger growth reductions during drought (*i.e.*, lower $Rt$) and less rebound following drought (*i.e.*, lower $Rc$ and $Rs$; Table 1; Fig.
-**4**).
+**3**).
 Specifically, for $Rt$, $ln[H]$ appeared, with negative coefficient, in the best model ($\Delta$AICc=0) and all top models when evaluating the three drought years together (Tables S8-S9).
@@
 Trees in drier microsites showed greater growth declines during drought; *i.e.*, $Rt$ had a significantly negative response to $ln[TWI]$ across all drought years combined, and in 1977 and 1999 individually (Fig.
-**4**, Table S8-S9).
+**3**, Table S8-S9).
 The $ln[H] *ln[TWI]$ interaction was never significant, and had a positive sign in any top $Rt$ models in which it appeared (Tables 1, S8-S9), rejecting the hypothesis that smaller trees (presumably with smaller rooting volume) are more susceptible to drought in microenvironments with a deeper water table.
-In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. **4**, Tables S10-S11).
+In contrast, $ln[TWI]$ did not appear in any of the best models for $Rc$ or $Rs$ (combined of for individual years), except in interaction with $ln[H]$ (Fig. **3**, Tables S10-S11).
 Negative $ln[H] *ln[TWI]$ interactions appeared in the best models for both $Rc$ and $Rs$ for all years combined, as well as in one individual year for each (1966 for $Rc$, 1977 for $Rs$). 
@@
 **S4**).
-Drought tolerance also varied across species, overall and in each drought year (Figs. **3, S7**). 
+Drought tolerance also varied across species, overall and in each drought year (Figs. **4, S7**). 
 Species with overall lowest and highest $Rt$ and $Rs$ were, respectively, *L. tulipifera* (mean $Rt$ = 0.66, mean $Rs$ = 1.04) and *Fagus grandifolia* (mean $Rt$ = 0.99; mean $Rs$ = 1.65).
-These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **3**).
+These two species--notably the only two diffuse-porous species in our study--differed significantly from one another in $Rt$ and $Rs$ in each drought year (Fig. **4**).
 
@@
 Xylem porosity had no significant influence on $Rt$ or $Rs$ in models for all droughts combined (Tables S4, S7). 
-In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. **3**).
+In contrast, $Rc$ was significantly higher in diffuse- and semi-ring porous species than in ring-porous species (Table S6, Fig. **4**).
 
-Drought resistance and resilience, but not recovery, were negatively correlated with $PLA_{dry}$ and $\pi_{tlp}$ (Fig. **4**; Tables 1, S4-S11). 
+Drought resistance and resilience, but not recovery, were negatively correlated with $PLA_{dry}$ and $\pi_{tlp}$ (Fig. **3**; Tables 1, S4-S11). 
 For $Rt$, $PLA_{dry}$ had a significant influence, with negative coefficient, in top models for the three droughts combined and for the 1966 drought individually (Fig.
-**4**; Tables S8-S9).
+**3**; Tables S8-S9).
 It was also included in some of the top models for 1999 (Tables S8-S9).
 $\pi_{tlp}$ was included with a negative coefficient in the best model for the combined droughts scenario and for the 1977 drought individually (Fig.
-**4**; Table S8), although its influence was not significant at $\Delta$AICc<2.
+**3**; Table S8), although its influence was not significant at $\Delta$AICc<2.
 It was also included in some of the top models for 1999 (Tables S8-S9).
@@
 For $Rs$, $PLA_{dry}$ was in the best models for the three droughts combined and for the 1966 drought individually, and some of the top models for 1977 and 1999 (Fig.
-**4**; Table S11); however, its effects were not significant at $\Delta$AICc<2.
+**3**; Table S11); however, its effects were not significant at $\Delta$AICc<2.
 $\pi_{tlp}$ was in the best models for the three droughts combined and for 1966 and 1999 individually, and in one of the top models for 1977 (Fig.
-**4**; Table S11).
+**3**; Table S11).
 However, its effects were significant at $\Delta$AICc<2 for 1999 only.
@@
 
-Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**).
+Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **3**).
 Taller trees had greater exposure to conditions that would promote water loss and heat damage during drought (Fig.
 **2**), which is one plausible mechanism for their lower drought resistance, recovery, and resilience (Fig.
-**4**).
+**3**).
 There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
@@
 However, the strengths of each predictor varied across droughts (Fig.
-**4**; Tables S8-S9), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
+**3**; Tables S8-S9), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected.
 These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
@@
 Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits) had increased growth (Figs.
-**1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
+**1b, 3**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015].
 Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. **1**), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
@@
 The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] provided a sufficient sample size to compare historical drought responses across species (Fig.
-**3**) and associated traits at a single site [see also @elliott_forest_2015].
+**4**) and associated traits at a single site [see also @elliott_forest_2015].
 Our study reinforced current understanding (see Introduction) that $WD$ and $LMA$ are not reliably linked to drought tolerance (Table 1).
@@
 grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig.
-**3**).
+**4**).
 While the low $Rt$ of *L.
@@
 **S4**), which would have contributed to its drought tolerance (Fig.
-**4**; see discussion below), in concordance with studies identifying *Fagus* species as intermediate in drought tolerance [@vitasse_contrasting_2019; @pretzsch_drought_2018].
+**3**; see discussion below), in concordance with studies identifying *Fagus* species as intermediate in drought tolerance [@vitasse_contrasting_2019; @pretzsch_drought_2018].
 A second explanation of why *F.
@@
 The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig.
-**4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
+**3**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance.
 Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; @zhu_leaf_2018], and our findings indicate that these traits also influence drought responses.
@@
 ### Author Contribution 
-KAT, IM, and AJT designed the research.
+KJA-T, IRM, and AJT designed the research.
 Tree-ring chronologies were developed by RH under guidance of AJT, KAT, and NP.
-Trait data were collected by IM and JZ under guidance of NK, KAT, and LS.
-Other plot data were collected by NB in coordination with WM and by IM and AS under guidance of EGA and KAT.
-Data analyses were performed by IM under guidance of KAT and VH.
-KAT and IM interpreted the results.
-IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
+Trait data were collected by IRM and JZ under guidance of NK, KAT, and LS.
+Other plot data were collected by NAB in coordination with WJM and by IRM and AELS under guidance of EBG-A and KJA-T.
+Data analyses were performed by IRM under guidance of KJA-T and VH.
+KJA-T and IRM interpreted the results.
+IRM and KJA-T wrote the first draft of manuscript, and all authors contributed to revisions.
 
@@
 ### ORCID
+
+**These need complete names, ordering**
+
 Ian R. McGregor: https://orcid.org/0000-0002-5763-021X
@@
 Norbert Kunert: https://orcid.org/0000-0002-5602-6221
+
+Alan Tepley: https://orcid.org/0000-0002-5701-9613
+
+Neil Pederson: https://orcid.org/0000-0003-3830-263X
+
+Bourg: 0000-0002-7443-1992
+
+McShea: 0000-0002-8102-0200
+
+Joseph: https://orcid.org/0000-0001-9103-190X
 

### 2020-10-06 scripts/graphs_for_manuscript.R
Added 300dpi conversion to ppi

@@ -5,6 +5,8 @@
 ######################################################
 library(ggplot2)
 
+#118.11 ppi = 300dpi. Argument in `png()` is res = <integer>
+
 ## DEFINE THIS BEFORE PLOTTING!
 metric <- "resistance" #resistance, recovery, or resilience
 
@@ -124,7 +126,7 @@ Mode <- function(x) {
 # apply(d, 2, Mode)
 # test[,.(mode = Mode(resist.value)), by=.(year)]
 
-##1b Figure 3: Distribution by species and anova of Rt by sp ####
+##1b Figure 4: Distribution by species and anova of Rt by sp ####
 library(data.table)
 library(agricolae)
 library(ggpubr)
@@ -842,7 +844,7 @@ heights_allplot$tree <- as.character(heights_allplot$tree)
 heights_allplot$position_all_abb <- as.character(heights_allplot$position_all_abb)
 
 ########################################################################
-#5. Figure 4. Visualizing regression output ####
+#5. Figure 3. Visualizing regression output ####
 library(lme4)
 library(ggpubr)
 # remotes::install_github("pbreheny/visreg")
@@ -1146,7 +1148,8 @@ for(w in 1:3){
    }
 }
 
-png("manuscript/tables_figures/publication/Figure4_model_vis_all.png", height=600, width=900)
+png("manuscript/tables_figures/publication/Figure3_model_vis_all.png", 
+    height=600, width=900, res=118) #118.11 ppi = 300dpi
 arr_full <- ggarrange(arr_rt, arr_rc, arr_rs, nrow=3)
 print(arr_full)
 dev.off()

### 2020-10-06 manuscript/tables&figures.Rmd
adjust Fig. 4 legend

@@
 **Figure 4.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) value ranges across species for the three focal droughts.** Species codes are given in Table 2, and closed dots represent outliers from the main boxplots.
-The dotted line at y=1 represents no change in $Rt$ or $Rs$ from the previous year.
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) across species for the three focal droughts.** Species codes are given in Table 2.
+Shaded boxes represent the interquartile range, with horizontal line at median, whiskers represent the range within 2.7 SD, and dots represent outliers.
+The horizontal dotted line at y=1 represents no change in growth between the five years prior to drought and the drought year ($Rt$) or the five years following the drought ($Rs$).
 Letters illustrate significance groupings per year (colored and ordered, top to bottom, 1966, 1977, 1999).
@@
 Descriptions of variables (e.g.
-$ln[H]$ can be found in Table 3.
+$ln[H]$) can be found in Table 3.
 
@@
 Descriptions of variables (e.g.
-$ln[H]$ can be found in Table 3.)
-](tables_figures/publication/Figure3_model_vis_all.png)
+$ln[H]$ can be found in Table 3.)](tables_figures/publication/Figure3_model_vis_all.png)
 
@@
 ![**Figure 4.
-Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) value ranges across species for the three focal droughts.** Species codes are given in Table 2, and closed dots represent outliers from the main boxplots.
-The dotted line at y=1 represents no change in $Rt$ or $Rs$ from the previous year.
+Drought resistance, $Rt$ (a), and resilience, $Rs$ (b) across species for the three focal droughts.** Species codes are given in Table 2.
+Shaded boxes represent the interquartile range, with horizontal line at median, whiskers represent the range within 2.7 SD, and dots represent outliers.
+The horizontal dotted line at y=1 represents no change in growth between the five years prior to drought and the drought year ($Rt$) or the five years following the drought ($Rs$).
 Letters illustrate significance groupings per year (colored and ordered, top to bottom, 1966, 1977, 1999).
@@
 S7 for parallel plot for recovery ($Rc$).
-Analysis conducted using *agricolae* package in R.](tables_figures/publication/Figure4_Rt-Rs_across_sp.png)
+Analysis conducted using *agricolae* package in R.
+Descriptions of variables (e.g.
+$ln[H]$) can be found in Table 3.](tables_figures/publication/Figure4_Rt-Rs_across_sp.png)

### 2020-10-06 manuscript/climate_sensitivity_draft.Rmd
social media/ orcid updates
more social media/ orcid

@@
 
+**Brief Heading:** Tree-rings reveal how tree height and leaf traits shape drought tolerance in a broadleaf forest
+
+Ian R. McGregor:
+
+https://twitter.com/realclimatian
+
 Joseph Zailaa:
+
 https://twitter.com/ZailaaJ
 
+Atticus E.L. Stovall:
+
+https://twitter.com/StovallAtticus
+
 Neil Pederson:
+
 https://twitter.com/YellowBuckeye,
+
 https://www.facebook.com/neil.pederson.5,
+
 https://www.instagram.com/1yellowbuckeye/
@@
 Kristina Anderson-Teixiera:
+
 https://twitter.com/K_A_Teixeira
-
-Ian R. McGregor:
-https://twitter.com/realclimatian
 
@@
 
-**These need complete names, ordering**
+Ian R. McGregor: 0000-0002-5763-021X
 
-Ian R. McGregor: https://orcid.org/0000-0002-5763-021X
+Norbert Kunert: 0000-0002-5602-6221
 
-Norbert Kunert: https://orcid.org/0000-0002-5602-6221
+Alan J. Tepley: 0000-0002-5701-9613
 
-Alan Tepley: https://orcid.org/0000-0002-5701-9613
+Erika B. Gonzalez-Akre: 0000-0001-8305-6672
 
-Neil Pederson: https://orcid.org/0000-0003-3830-263X
+Valentine Herrmann: 0000-0002-4519-481X
 
-Bourg: 0000-0002-7443-1992
+Joseph Zailaa: 0000-0001-9103-190X
 
-McShea: 0000-0002-8102-0200
+Atticus E.L. Stovall: 0000-0001-9512-3318
 
-Joseph: https://orcid.org/0000-0001-9103-190X
+Norman A. Bourg: 0000-0002-7443-1992
 
-Kristina J. Anderson-Teixeira: https://orcid.org/0000-0001-8461-9713
+William J. McShea: 0000-0002-8102-0200
+
+Neil Pederson: 0000-0003-3830-263X
+
+Lawren Sack: 0000-0002-7009-7202
+
+Kristina J. Anderson-Teixeira: 0000-0001-8461-9713
 

### 2020-10-07 manuscript/climate_sensitivity_SI.Rmd
Updated Fig S7 caption (and Fig S4) #150

@@
 ![**Figure S4.
-Distribution of independent variables by species.** Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
-Letter groupings do not transfer between variables.](tables_figures/publication/FigureS4_traits_signif.png)
+Distribution of independent variable values by species.
+Species codes are given in Table 2.
+Boxes in plots (a) – (f) represent the interquartile range, with the horizontal line at the median, whiskers representing the range within 2.7 SD, and dots representing outliers.
+For plots (a) – (f), species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
+Similarly, letter groupings do not transfer between variables.
+Meanwhile, plot (g) shows the number of trees in each crown position per species.
+Descriptions of variables (e.g.
+$ln[H]$) can be found in Table 3.](tables_figures/publication/FigureS4_traits_signif.png)
 
@@
 Drought recovery ($Rc$) across species for the three focal droughts.
-Species that are assigned the same letter are not significantly different from each other with regard to the tested variable.
-Letter groupings do not transfer between variables.**](tables_figures/publication/FigureS7_Rc_across_sp.png)
+Species codes are given in Table 2.
+Shaded boxes represent the interquartile range, with horizontal line at median, whiskers represent the range within 2.7 SD, and dots represent outliers.
+The dotted line at y=1 represents no change in Rc from the five years prior to drought.
+Letters illustrate significance groupings per year (colored and ordered, top to bottom, 1966, 1977, 1999).
+That is, a group of species with the same letter above their boxplot (e.g.
+"b") are statistically different from species in another group (e.g.
+"a").
+See Fig.
+4 for parallel plot for resistance ($Rt$) and resilience ($Rs$).
+Analysis conducted using agricolae package in R.
+**](tables_figures/publication/FigureS7_Rc_across_sp.png)
 

### 2020-10-07 manuscript/climate_sensitivity_draft.Rmd
add Lawren's twitter
updated DOI for data release including hydraulic traits
#79

@@
 https://www.instagram.com/1yellowbuckeye/
+
+Lawren Sack:
+
+https://twitter.com/lawrensack
 
@@
 ### Data and code availability
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.4070038 and *[TBD]*, respectively.
 Full ForestGEO census data for SCBI are available through the ForestGEO data portal (www.forestgeo.si.edu).

### 2020-10-07 non_publication/scripts/archive_climate_sensitivity.R
delete files not for publication

@@ -1,1421 +1,0 @@
-######################################################
-# Purpose: Original code for McGregor et al, kept here for reference
-# Created by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
-# R version 3.5.2 - First created February 2020
-######################################################
-
-# All of this code corresponds to "canopy_position_analysis.R" and contains original
-# code written for the analysis, but was not used for the final analyses.
-# Instead of throwing it out, I have moved it here in case we ever
-# want to reference our train of thought, or the original methods we decided to use.
-
-# for ease of reading, collapse all the sections ahead of time
-
-#########################################################################################
-# this section is not used at all in the current analysis script. This is from when
-# I was first starting out understanding mixed effects models in R and the difference
-# between AIC and BIC, for example. These mini chunks use CRU data to walk through
-# the process (the accompanying plots are at the end of the graphs_for_manuscript
-# script).
-
-## necessary packages ####
-library(car)
-library(MASS)
-library(lme4)
-library(AICcmodavg)
-library(data.table)
-library(dplyr)
-##3a. explaining mixed effects model in R ####
-cru1901 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/results/canopy_vs_subcanopy/1901_2009/tables/monthly_correlation/correlation_with_CRU_SCBI_1901_2016_climate_data.csv"), stringsAsFactors = FALSE)
-
-#subset out caco, cato, and frni because they don't have pair of canopy and subcanopy
-cru1901 <- cru1901[!(cru1901$Species %in% c("CACO_subcanopy", "CATO_subcanopy", "FRNI_subcanopy")), ]
-
-cru1901_loop <- cru1901
-
-#create separate identifier
-cru1901_loop$position <- ifelse(grepl("subcanopy", cru1901$Species), "subcanopy", "canopy")
-cru1901_loop$Species <- gsub("_[[:alpha:]]+$", "", cru1901$Species)
-
-#response variable is coef (continuous) (renamed corr)
[…]

### 2020-10-07 non_publication/scripts/archive_graphs_for_manuscript.R

@@ -1,478 +1,0 @@
-###########################################################
-# Purpose: Archive figure and table script from "graphs_for_manuscript.R"
-# Developed by: Ian McGregor, contact Anderson-Teixeira (teixeirak@si.edu)
-# First written: July 2019
-###########################################################
-#5. Archive: Original Figure 2 with PLA and TLP plotted against height
-# These PLA and TLP plots were originally plots e and f of Figure 2. We have chosen
-# to not include these.
-
-## necessary packages ####
-library(ggstance)
-library(raster)
-library(dplyr)
-library(Hmisc)
-library(magrittr)
-
-
-##5a. get height data for all trees >10cm dbh in census ####
-scbi.stem3 <- read.csv(text=getURL("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"), stringsAsFactors = FALSE)
-
-scbi.stem3$dbh <- as.numeric(scbi.stem3$dbh)
-scbi.stem3$dbh <- ifelse(is.na(scbi.stem3$dbh), 0, scbi.stem3$dbh)
-
-scbi18_ht <- scbi.stem3
-scbi18_ht <- scbi18_ht[scbi18_ht$dbh>=100, ]
-scbi18_ht <- scbi18_ht[,c(2:5,11,14:15)]
-
-scbi18_ht$dbh_old.cm <- scbi18_ht$dbh/10
-scbi18_ht$dbh.ln.cm <- log(scbi18_ht$dbh_old.cm)
-
-#linear log-log regression
-#use height_regr from above
-
-scbi18_ht$height.ln.m <- NA
-for(w in seq(along=height_regr$sp)){
-  sp_foc <- height_regr$sp[[w]]
-  ht_eq <- height_regr[height_regr$sp == sp_foc, ]
-  num <- gsub("\\*x", "", ht_eq$Equations)
-  num1 <- as.numeric(stri_extract_first_regex(num, "[[:digit:]].[[:digit:]]+"))
-  num2 <- as.numeric(stri_extract_last_regex(num, "[[:digit:]].[[:digit:]]+"))
-  
-  if(sp_foc %in% scbi18_ht[,"sp"]){
-    scbi18_ht$height.ln.m <- 
-      ifelse(scbi18_ht$sp == sp_foc,
-             num1 + num2*scbi18_ht$dbh.ln.cm,
[…]

### 2020-10-07 non_publication/scripts/compare_pdsi_values.R

@@ -1,168 +1,0 @@
-#compare pdsi values
-
-library(raster)
-library(ncdf4)
-library(ncdf4.helpers)
-library(ncdf.tools)
-library(lubridate)
-
-#National drought monitor ####
-# https://droughtatlas.unl.edu/Data.aspx
-win_pdsi <- read.csv("C:/Users/mcgregori/Dropbox (Smithsonian)/Github_Ian/tree-growth-and-traits/data/pdsi_timeseries_winchester_1949-2012.csv")
-colnames(win_pdsi) <- c("date", "win_pdsi")
-win_pdsi$date <- as.POSIXct(win_pdsi$date, format = "%m/%d/%Y")
-win_pdsi$date <- as.Date(win_pdsi$date)
-
-win_scpdsi <- read.csv("C:/Users/mcgregori/Dropbox (Smithsonian)/Github_Ian/tree-growth-and-traits/data/scpdsi_timeseries_winchester_1949-2012.csv")
-colnames(win_scpdsi) <- c("date", "win_scpdsi")
-win_scpdsi$date <- as.POSIXct(win_scpdsi$date, format = "%m/%d/%Y")
-win_scpdsi$date <- as.Date(win_scpdsi$date)
-
-
-
-winchester <- merge(win_pdsi, win_scpdsi)
-
-#CRU ####
-# https://crudata.uea.ac.uk/cru/data/drought/
-fname <- "I:/Krista Lab/Netcdf PDSI values/scPDSI.cru_ts3.26early.bams2018.GLOBAL.1901.2017.nc" #from CRU
-cru <- nc_open("I:/Krista Lab/Netcdf PDSI values/scPDSI.cru_ts3.26early.bams2018.GLOBAL.1901.2017.nc")
-time <- ncvar_get(cru, varid = "time")
-
-#CRU data shows time as days since 1900-01-01.
-timetrue <- convertDateNcdf2R(time, units = "days", origin = as.POSIXct("1900-01-01"), time.format = c("%Y-%m-%d"))
-
-cru_nc <- brick(fname)
-scbi <- data.frame("lon" = 78.1653, "lat" = 38.8871)
-cru_scpdsi <- extract(cru_nc, scbi)
-cru_scpdsi <- data.frame(cru_scpdsi)
-cru_scpdsi[2, ] <- colnames(cru_scpdsi)
-
-library(data.table)
[…]

### 2020-10-07 non_publication/scripts/core_list_by_canopy_subcanopy.R

@@ -1,53 +1,0 @@
-# sort cores for Neil
-
-## location of cores is now in SCBI-ForestGEO-Data repo and is public.
-
-core_list <- read.csv("data/core_list_for_neil.csv")
-
-library(dplR)
-
-species <- core_list$sp[!duplicated(core_list$sp)]
-species <- as.vector(species)
-
-for (sp in species){
-  sp1 <- read.rwl(paste0(sp, "_drop.rwl"))
-  
-  sp1_can <- core_list$tag[core_list$sp %in% sp & core_list$crown.position=="canopy"]
-  sp1_can <- ifelse(sp1_can %in% colnames(sp1), sp1_can, paste0(sp1_can,"A"))
-  sp1_can <- ifelse(sp1_can %in% colnames(sp1), sp1_can, paste0("0", sp1_can))
-  sp1_canopy <- sp1[,names(sp1) %in% sp1_can, drop=F] #drop=F only for frni_canopy
-  write.rwl(sp1_canopy, paste0(sp, "_drop_canopy.rwl"), long.names=TRUE)
-  
-  sp1_sub <- core_list$tag[core_list$sp %in% sp & core_list$crown.position=="sub-canopy"]
-  sp1_sub <- ifelse(sp1_sub %in% colnames(sp1), sp1_sub, paste0(sp1_sub,"A"))
-  sp1_sub <- ifelse(sp1_sub %in% colnames(sp1), sp1_sub, paste0("0", sp1_sub))
-  sp1_subcan <- sp1[, colnames(sp1) %in% sp1_sub]
-  write.rwl(sp1_subcan, paste0(sp, "_drop_subcanopy.rwl"), long.names=TRUE)
-}
-
-
-
-
-
-
-
-#####################################################################################
-## for troubleshooting
-frni <- read.rwl("C:/Users/mcgregori/Dropbox (Smithsonian)/Github_Ian/SCBI-ForestGEO-Data_private/tree_cores/chronologies/current_chronologies/frni_drop.rwl")
-
-
-  frni_can <- core_list$tag[core_list$sp=="frni" & core_list$crown.position=="canopy"]
-  frni_can <- ifelse(frni_can %in% colnames(frni), frni_can, paste0(frni_can, "A"))
-  frni_can <- ifelse(frni_can %in% colnames(frni), frni_can, paste0("0", frni_can))
-  frni_canopy <- frni[,names(frni) %in% frni_can, drop=F] 
[…]

### 2020-10-07 manuscript/climate_sensitivity_draft.Rmd
add DOI
#86

@@
 ### Data and code availability
-All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.4070038 and *[TBD]*, respectively.
+All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.4070038 and 10.5281/zenodo.4070059, respectively.
 Full ForestGEO census data for SCBI are available through the ForestGEO data portal (www.forestgeo.si.edu).

### 2020-10-30 manuscript/climate_sensitivity_draft.Rmd
Added some changes caught by NewPhyt editors

@@
 Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003; @scholz_hydraulic_2011].
-Taller trees have wider conduits in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller.
+Wider conduits are found in the basal portions of taller trees, both within and across species [@olson_plant_2018; @liu_hydraulic_2019] and throughout the conductive systems of angiosperms [@zach_vessel_2010; @olson_universal_2014; @olson_plant_2018], which help maintain constant the resistance that would otherwise increase as trees grow taller.
 Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018], and traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms [@roskilly_conflicting_2019].
